### 2020-02-19 paper.tex
Initial draft of text.

\documentclass{article}
\usepackage[round]{natbib}

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\tskit}[0]{\texttt{tskit}}
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\scrm}[0]{\texttt{scrm}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}

\begin{document}

\title{msprime 1.0: an efficient and extensible coalescent simulation framework}
\author{Author list to be filled in
}
% \address{

% \section{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}

\maketitle

% Abstract: ~ 150 words.  Concise summary of the results.

% Key points in abstract.
% 1) Coalescent simulation is a vital tool
% 2) Coalescent with recombination is hard
% 3) Large sample sizes are crucial today
% 4) Msprime has lots of features and uses a community development model

\begin{abstract}
Coalescent simulation is a key tool in population genetics and
and has been integral to coalescent theory since its earliest days.
Because of the ease at which the basic model can be simulated,
a large number of different simulators have been developed. However,
the coalescent with recombination is far more challenging to simulate,
and, until recently, it was not possible to simulate efficiently.
The \msprime\ software has revolutionised
coalescent simulation, making it possible to simulate millions
of whole genomes for the first time. We summarise the many features
of \msprime\ 1.0, which is built around a core model of efficiently
implementing recombination via the succinct tree sequence data
structure. We advocate a community oriented open-source development
process as a way to reduce duplication of effort and increase
the quality of simulation software.
\end{abstract}

% Keywords: For software, include the relevant models, algorithms and language;
\textbf{Keywords:} Coalescent simulation, Python

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction: Limit of 500 words. The introduction should describe the
% significance of the software or resource presented, discuss novelty, provide an
% overview of the software or resource, and describe how researchers can access it.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Introduction}

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating these to each other. Because of its
elegance and simplicity, the coalescent is now central to much of
population genetics and
genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Two developments of recent years, however, have made the majority
of these simulators of little relevance to modern datasets.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing technology---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent can be simulated in constant
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging to simulate.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, we had to resort to
approximate models of recombination~\citep{mcvean2005approximating,staab2014scrm}
if we wished to simulate whole chromosomes.

The second development that has made the majority of
present day simulators inapplicable to modern data sets is the
exponential increase in sample size. Human datasets such
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2019variation} now contain hundreds of
thousands of genomes, and there is every reason to believe that
datasets of millions will soon be commonplace~\citep{stephens2015big}.
Classical simulators such as \ms\ and even fast approximate simulators
such as \scrm\ simply cannot cope with this scale. Even if we
spend months of CPU time on running simulations with millions of samples,
the output would require terabytes of storage space to store and
days of CPU time to parse~\citep{kelleher2016efficient}.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has profoundly changed
what is possible. Through the use of efficient data structures, it is
now possible to simulate millions of whole chromosomes, without resorting
to approximations. The ``succinct tree
sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} formats. This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring}
and calculation of population genetic statistics~\citep{ralph2019efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

% JK: Earlier notes on this para
% - Msprime 1.0 combines the features of many different simulators and is
%   far more efficient than all of them. The combination of a Python API,
%   an extremely efficient storage format, and a well-engineered and
%   extensible codebase changes the game. Stdpopsim builds on this base
%   to provide easy access to standard models of population genetics.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Methods: For software, the authors should describe key elements of the algorithms
% implemented, and are expected to provide both the source code as well as
% examples illustrating the use of the software.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Methods}

In this section we describe the main features of \msprime\ 1.0.

\subsection*{Recombination and gene conversion}
Recombination is implemented in \msprime\ via
Hudson's algorithm, which
works backwards in time tracking the
through the effects of common ancestor and recombination events
~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}. Common
ancestor events combine the ancestral material of two lineages, and may
 result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.

We also support variable recombination rates along a chromosome.
This allows us to model both recombination hotspots like
\texttt{msHOT}~\citep{hellenthal2007mshot} and also empirical
recombination maps like those supported by X Y Z.
(What about ~\cite{wang2014new}?)
We have also implemented gene
conversion following the model defined by~\cite{wiuf2000coalescent}
and used in \ms. Further features are planned to model bacterial
evolution [Can you elaborate here please Franz?]

\subsection*{SMC approximations}

The SMC~\citep{mcvean2005approximating} and SMC'~\citep{marjoram2006fast}
models are supported.
The performance of msprime has made the SMC
approximation largely unnecessary, at least when simulating human-like
recombination rates. The SMC is an important analytical approximation
forming the basis for several important inference methods
methods~\citep{
li2011inference,
harris2013inferring,
sheehan2013estimating,
schiffels2014inferring,
carmi2014renewal,
rasmussen2014genome,
zheng2014bayesian,
terhorst2017robust},
so simulations are still useful to examine  how good the approximation
is.

The SMC has been shown to be a good approximation to the
coalescent with recombination for samples of less than
four~\citep{hobolth2014markovian,wilton2015smc}
and in simulations, produces patterns of LD indistinguishable from the
full coalescent~\citep{mcvean2005approximating,marjoram2006fast}. However,
LD is a crude approximation of the correlation structure of trees
and it is not clear how good the SMC is for larger sample sizes.

\subsection*{Population structure}
We support all the stuff that ms does.
Demographic events where we change
the properties of the population.

We provide a debugger interface which allows the user to see the
various population sizes and growth rates over time. This includes
a way to compute coalescence rate trajectory, which provides a useful
ground truth when comparing with methods that estimate population
size over time~\citep{adrion2019community}.

The interface
is very flexible and general. We also reading in population tree
descriptions as produced by programs such as
StarBeast~\citep{heled2009bayesian}.

We support tracking lineages through time by associating a population
with each node, and having migration records. We also have
census events, which will record the location of every lineage
at a particular time.

\subsection*{Selection}
Selection was added to the coalescent ... Lots of simulators exist:
SelSim~\citep{spencer2004selsim}
mbs~\citep{teshima2009mbs},
msms~\citep{ewing2010msms},
% What about Cosi, did it simulate selection?
Cosi2~\citep{shlyakhter2014cosi2},
and discoal~\citep{kern2016discoal}.

We have implemented the basic infrastructure exists for the structured
coalescent, which makes adding support for, e.g.,
inversions straightforward~\citep{peischl2013sequential}

\subsection*{Discrete time Wright-Fisher}
The coalescent is a good approximation for ancient history, but
is bad for the recent past and in particular when we have large
sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Describe basic
approach [DOM, can you fill in please].
See ~\cite{nelson2020accounting} for more details.
We compare the peformance of this simulation in Fig X
with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.

We support changing the simulation model, so that we can easily
run hybrid simulations like described by~\cite{bhaskar2014distortion}
This ia a powerful approach because it lets us combine any
number of different simulation models.

\subsection*{Integration with forwards-time simulations}
Forward simulations can also use tree
sequences~\citep{kelleher2018efficient,haller2018tree}. In particular,
SLiM 3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output tree sequences.
Msprime can take tree sequences simulated by these forward simulators
as input and complete the simulation (``recapitation'').

\subsection*{Multiple merger coalescents}
Kingman's coalescent assumes that all mergers are binary. This
assumption can be violated in some situations. Lambda coalescents
~\citep{donnelly1999particle,schweinsberg2000coalescents}
HybridLambda~\citep{zhu2015hybrid}
We support X Y Z.

Mention~\citep{becheler2020occupancy} somewhere.

\subsection*{Ancestral recombination graphs}
The definition of ARGs is vague and confusing.
\cite{minichiello2006mapping} define it as the data structure
rather than the original stochastic
process~\citep{griffiths1991two,griffiths1997ancestral}.
\cite{rasmussen2014genome} have a slightly different approach.
Sometimes the ARG is useful.
The tree sequence output for \msprime\ can be turned into
an ARG using the \texttt{record\_full\_arg} option, where we store
recombination nodes.

We also provide an interface to compute the likelihood of a given ARG
under the~\cite{kuhner2000maximum} model.

\subsection*{Simultaneous bottlenecks}
Overview of what simulataneous bottlenecks literature. What we
do [Konrad].

\subsection*{Seedbank coalescent}
Seedbank coalescent. [If we implement this fill in please Thibaud]

\subsection*{Simulation interface}
The primary interface for \msprime\ is through a Python application
programming interface (API). This has many advantages. One is
that we can grow the interface over time. Another is that
downstream applications such as \texttt{stdpopsim}~\citep{adrion2019community}
can be developed.

\subsection*{Output}
The output of \msprime\ is a ``succinct tree sequence`` (or tree sequence,
for short) using the \tskit\ library. [Quick overview of the output formats
of other simulators, and the perf advantages of tree sequences.]

\subsection*{Development model}
The community development process for \msprime\ is enabled by following an
open source development practise with a strong emphasis on code quality
and correctness. To ensure a consistent style, we require that all code
follows the PEP8 guidelines. Unit tests for all additions are required, and
are automatically run on each pull request on a variety of continuous
integration testing platforms. For new simulation features, statistical tests
must also be added, comparing the distribution of results to either analytical
results or existing simulators.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Results and Discussion: For software, the authors are expected to present a
% few examples that demonstrate the use of the software and benchmarks. Ideally,
% the examples should be simple. We encourage the use of boxes with scripts
% supported by a narrative. Benchmarks carried under more realistic situations
% than the one presented in the examples are also encouraged.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Results and Discussion}
\subsection*{Performance}
We compare \msprime\ with some existing simulators that have a subset of its
features.

\subsection*{Examples}
We show a few examples of running \msprime.

\subsection*{Discussion}
% Simulation is useful, and it's really easy for a single locus.
% This has lead to a culture of developing lots of simulators
% rather than focusing on one
Simulation has played a key role in the development of coalescent theory
and its use in understanding observed patterns of genetic variation.
The basic model can be simulated for $n$ samples in $O(n)$ time
and is straightforward to implement---for example, Hudson provided a fully
functional and highly efficient simulator as a C source code listing in an
appendix to his seminal review paper~\citep{hudson1990gene}. Because of
this ease, the dominant approach to producing simulators for extensions
to this basic model has been to develop them independently. This is
often seen as a learning opportunity, since those developing extensions
to the coalescent must have an excellent understanding of the basic
model in order to develop extensions to the coalescent.
The coalescent with recombination, however, is \emph{not} straightforward
to implement, and numerous different approaches have been
proposed over the
years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
mcvean2005approximating}. It has only recently become clear that it is
possible to implement the model efficiently, and only through the use
of sophisticated data structures and an intricate and subtle
algorithm~\citep{kelleher2016efficient}.
This efficiency, and the ability to handle recombination in a principled
way is vital with today's vast numbers of whole genome samples.

% Making lots of simulators has given us a fragmented and poort
% quality ecosystem.
The approach of developing ``in-house'' simulators
has lead to a huge proliferation of different programs, with
slightly different feature sets and often incompatible interfaces.
It is a fragmented ecosystem with significant issues with software
quality (see, e.g.,~\cite{yang2014critical});
few simulators are actively maintained after publication.
There seems to be little point
in continuing the huge duplication of effort we have seen.
It is not possible to achieve the performance levels of
\msprime\ without investing substantially in software engineering.
As demonstrated here, \msprime\ is extensible and we would argue
that developing this simulator as an open-source community asset
is of more benefit to all.

There is also the welcome development of libraries such as
Quetzal~\citep{becheler2019quetzal} and libsequence~\citep{thornton2014cpp}
which break this pattern of monolithic simulation software.

Thus, development of \msprime\ is not completed and much remains
to be done. For example, one key aim of the 1.0 series is to
improve the mutation models supported which are limited to
simple infinite-sites currently. The ability to simulate
sophisticated mutation models like Seq-Gen~\citep{rambaut1997seq}
and Dawg~\citep{cartwright2005dna} at the megasample scale would be
useful in many applications. Another key focus of development
will be [WHAT?]

% Finish up with a paragraph saying that there loads of exciting
% new things that simulation can do. We need MORE sims!!!
The ability to infer trees at scale for the first time
in the presence of
recombination~\citep{harris2019database,kelleher2019inferring,
speidel2019method,tang2019genealogy}
opens the possibility for new applications of coalescent simulations.
Also \stdpopsim\ is cool~\citep{adrion2019community}

\section*{Acknowledgments}
JK is supported by the Robertson Foundation. [Others fill in their ACKs please]

\bibliographystyle{plainnat}
\bibliography{paper}

\end{document}

### 2020-02-24 paper.tex

@@
 conversion following the model defined by~\cite{wiuf2000coalescent}
-and used in \ms. Further features are planned to model bacterial
-evolution [Can you elaborate here please Franz?]
+and used in \ms. 
+% % Uncomment this sentence, when the corresponding issue in the msprime github repo is resolved.
+% Furthermore, msprime is now able to simulate gene conversion events that are 
+% initiated proportional to the local recombination rate specified by the recombination map.
+
+Gene conversion is particularly useful to model homologous recombination in bacterial evolution.
+We compare the performance of \msprime\ with gene conversion to \ms\ and
+the whole bacterial genome simulator SimBac~\citep{brown2016simbac}.
+Further features are planned to model bacterial evolution. In particular, we
+aim to include bacterial gene transfer and utilize the succinct tree sequence structure
+to represent the resulting ancestral gene transfer graph~\citep{baumdicker2014AGTG}.
 
@@
 See ~\cite{nelson2020accounting} for more details.
-We compare the peformance of this simulation in Fig X
+We compare the performance of this simulation in Fig X
 with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.

### 2020-02-26 paper.tex
Citations for pyvolve.

@@
 simple infinite-sites currently. The ability to simulate
-sophisticated mutation models like Seq-Gen~\citep{rambaut1997seq}
-and Dawg~\citep{cartwright2005dna} at the megasample scale would be
+sophisticated mutation models like Seq-Gen~\citep{rambaut1997seq},
+Dawg~\citep{cartwright2005dna} and Pyvolve~\citep{spielman2015pyvolve} at the megasample scale would be
 useful in many applications. Another key focus of development

### 2020-05-18 lineage_probabilities/lineage_probs.tex
draft of information about N(t) and lineage probs

Many popular approaches in population genetics use the distribution of coalescence rates between 
pairs of lineages in one or more populations to infer effective population sizes over time 
\cite{li2011inference,sheehan2013estimating,schiffels2014inferring}
or split times 
and subsequent migration rates between populations 
\cite{wang2020tracking}.
Given a demographic model, the 
\texttt{msprime} DemographyDebugger is able to compute coalescence rates for two or more lineages drawn 
from one or more populations at specified times in the past, providing a ground truth for coalescent rate-based 
inference methods. To compute coalescence rates for a set of times $Ts=(t_0, t_1, \ldots)$ in the 
past, we write
\begin{lstlisting}[frame=single]
dbg = demography.debug()
rates = dbg.coalescence_rate_trajectory(Ts, num_samples)
\end{lstlisting}
where \texttt{num_samples} is a list with length equal to the number of populations that
specifies the number of sampled lineages from each population. For example, 
to get pairwise rates within the same 
population, this list would contain zeros for all populations except for the population index of interest, 
which would have a $2$. To get the rate of coalescence for two lineages sampled from different 
populations, we would set \texttt{num_samples} to have ones in those population indices.

More simply, \texttt{msprime} can also compute the probability of a lineage that was in a given
population at some time to be found in other populations (through continuous or mass migration events). 
For a given sample time $t\geq0$ and any list of times $Ts$ in the past, we write
\begin{lstlisting}[frame=single]
prob = dbg.lineage_probabilities(Ts, sample_time=t)
\end{lstlisting}
This function is also used to determine the time intervals that lineages can be found in each population,
given a list of contemporary and/or ancient samples. For a given sampling configuration, by writing
\begin{lstlisting}[frame=single]
lineage_dict = dbg.possible_lineage_locations(samples=samples)
\end{lstlisting}
we get a dictionary with epoch intervals as keys whose values are a list that indicates True or False for 
whether each population could contain ancestral material to our samples.
If no sampling configuration is provided, we assume we sample lineages from each population at time zero.

### 2020-05-20 paper.tex
move text to paper

@@
 \usepackage[round]{natbib}
+\usepackage{listings}
 
@@
 
+\subsection*{Computing coalescence rates and lineage location probabilities}
+Many popular approaches in population genetics use the distribution of coalescence rates between 
+pairs of lineages in one or more populations to infer effective population sizes over time 
+\cite{li2011inference,sheehan2013estimating,schiffels2014inferring}
+or split times 
+and subsequent migration rates between populations 
+\cite{wang2020tracking}.
+Given a demographic model, the 
+\texttt{msprime} DemographyDebugger is able to compute coalescence rates for two or more lineages drawn 
+from one or more populations at specified times in the past, providing a ground truth for coalescent rate-based 
+inference methods. To compute coalescence rates for a set of times $Ts=(t_0, t_1, \ldots)$ in the 
+past, we write
+\begin{lstlisting}[frame=single]
+dbg = demography.debug()
+rates = dbg.coalescence_rate_trajectory(Ts, num_samples)
+\end{lstlisting}
+where \texttt{num\_samples} is a list with length equal to the number of populations that
+specifies the number of sampled lineages from each population. For example, 
+to get pairwise rates within the same 
+population, this list would contain zeros for all populations except for the population index of interest, 
+which would have a $2$. To get the rate of coalescence for two lineages sampled from different 
+populations, we would set \texttt{num\_samples} to have ones in those population indices.
+
+More simply, \texttt{msprime} can also compute the probability of a lineage that was in a given
+population at some time to be found in other populations (through continuous or mass migration events). 
+For a given sample time $t\geq0$ and any list of times $Ts$ in the past, we write
+\begin{lstlisting}[frame=single]
+prob = dbg.lineage_probabilities(Ts, sample_time=t)
+\end{lstlisting}
+This function is also used to determine the time intervals that lineages can be found in each population,
+given a list of contemporary and/or ancient samples. For a given sampling configuration, by writing
+\begin{lstlisting}[frame=single]
+lineage_dict = dbg.possible_lineage_locations(samples=samples)
+\end{lstlisting}
+we get a dictionary with epoch intervals as keys whose values are a list that indicates True or False for 
+whether each population could contain ancestral material to our samples.
+If no sampling configuration is provided, we assume we sample lineages from each population at time zero.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-05-21 msprimepaper.tex
latex source file

\documentclass{article}
\usepackage[round]{natbib}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
\usepackage{showkeys}
%%
% local definitions
\newcommand{\msprime}[0]{{\texttt{msprime} }}
\newcommand{\tskit}[0]{{\texttt{tskit} }}
\newcommand{\ms}[0]{{\texttt{ms} }}
\newcommand{\scrm}[0]{{\texttt{scrm} }}
\newcommand{\stdpopsim}[0]{{\texttt{stdpopsim} }}
 \newcommand{\be}{\begin{equation}}
 \newcommand{\ee}{\end{equation}}
 \newcommand{\bd}{\begin{displaymath}}
 \newcommand{\ed}{\end{displaymath}}
\newcommand{\IN}{\ensuremath{\mathds{N}}}%
\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%

\NewEnviron{esplit}[1]{%
\begin{equation}
\label{#1}
\begin{split}
  \BODY
\end{split}\end{equation}
}

\begin{document}%

\linenumbers

\subsection*{Multiple merger coalescents}%
\label{mmc}%

Kingman's coalescent assumes that at most two ancestral lineages can
merge at each merger event.  This follows from the fact that the
Kingman coalescent can be shown to be the process describing the
random ancestral relations of gene copies sampled from a population
evolving according to the Wright-Fisher model (or similar models) of
genetic evolution.  Multiple-merger coalescent models can also be
obtained from appropriate extensions of the Moran
model \citep{EW06,HM12}, i.e.\ where a single randomly picked individual
(e.g.\ in a haploid population), or a pair of diploid parent
individuals \citep{BBE13}, produces offspring in each reproduction
event.  Convergence to the Kingman coalescent follows from certain
assumptions on the offspring distribution \citep{S99,MS01,SW08}.  These
assumptions may be violated in certain highly fecund
organisms \citep{hedgecock_94,B94,HP11,A04,irwin16}.  We take `high
fecundity' to mean the ability of individuals to produce numbers of
offspring on the order of the population size.  A key issue regarding
high fecundity involves the concept of `sweepstakes
reproduction' \citep{hedgecock_94}, where a few individuals may produce
the bulk of the offspring in any given generation.  Sweepstakes
reproduction is not captured by the Wrigth-Fisher model (or similar
models).  Population models, which do capture high fecundity and
sweepstakes reproduction, are in the domain of attraction of
`multiple-merger' coalescents \citep{DK99,P99,S99,S00}, in which a
random number of ancestral lineages may merge at any given time.  The
term `multiple merger' refers to a merger of at least three ancestral
lineages, or the simultaneous \citep{S00,MS01} merger of at least four
lineages in at least two distinct groups, and each group involving at
least two lineages.  Multiple-merger coalescent processes have also
been shown to be relevant for modeling the effects of selection on
gene genealogies \citep{Gillespie909,DS04}.

Multiple-merger coalescents, in which a random number of ancestral
lineages may merge at a given time in one group, and only one such
group of lineages can merge at any given time (asynchronous multiple
mergers), are referred to as $\Lambda$-coalescents; the rate at which  a given group of $k$ lineages out of a total of  $b$ lineages merges  is given by ($a\ge  0$ constant)
\begin{equation}\label{lambdabk}
\lambda_{b, k} =  \int_0^1  x^{k-2}(1-x)^{b-k}\Lambda(dx) + a\one{k=2}, \quad 2 \le k \le b, 
\end{equation}
where $\Lambda$ is a finite measure on the  (Borel subsets of)  unit interval without an atom at zero \citep{DK99,P99,S99}.    The  total rate  is given by
\be\label{lambdab}
 \lambda_{b} = \int_0^1 \left(1 - (1-x)^{b} - bx(1-x)^{b-1} \right)x^{-2}\Lambda(dx) + \binom{b}{2},
\ee
\citep{S99}  which can be useful in applications, in particular  provided the  antiderivative  can be explicitly identified.

A larger class of multiple-merger coalescents involving simultaneous
multiple mergers of distinct groups of ancestral lineages also exists
\citep{S00}. These are commonly referred to as $\Xi$-coalescents, and
can be shown to be the limits of ancestral processes derived from
population models incorporating diploidy (or more general polyploidy)
\citep{BBE13,Blath2016}, and certain models of selection \citep{DS04}.
To describe a general $\Xi$-coalescent let $\Delta$ denote the
infinite simplex \be\label{Delta} \Delta := \{ (x_1, \ldots ): x_1 \ge
x_2 \ge \cdots \ge 0, \sum_{j}x_j \le 1\}; \ee for any
$r \in \{1,2, \ldots\}$ let $k_1, \cdots, k_r \ge 2$, and
$b = s + k_1 + \cdots + k_r$ be the total number of blocks in a given
partition ($s = b - k_1 - \cdots - k_r$ is the number of blocks not
participating in the mergers at the given time).  The existence of
simultaneous multiple-merger coalescents was proved by \cite{S00}.
There exists a finite measure $\Xi$ on $\Delta$, with
$\Xi = \Xi_0 + a\delta_0$, the measure $\Xi_0$ has no atom at zero,
$a >0$ is fixed, and the rate at which one sees a given (simultaneous)
merger of ancestral lineages with merger sizes $k_1, \ldots, k_r$,
$s = n - k_1 - \cdots - k_r$, is given by
\begin{esplit}{xi}
  \lambda_{n; k_1, \ldots, k_r; s}  & = \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
  & +  a\one{r=1, k_1 = 2}.
\end{esplit}%
The number of such $(k_1, \ldots, k_r)$ mergers is
\be\label{N}
    \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
\ee
\citep{S00},   in particular  $\mathcal{N}(b;2) = b(b-1)/2$, and one can compute the total rate of a  $(k_1, \ldots, k_r)$ merger as
\be\label{lambdabkall}
      \lambda(n; k_1, \ldots, k_r)         =    \mathcal{N}(b; k_1, \ldots, k_r ) \lambda_{n; k_1, \ldots, k_r; s}.
\ee

The total rate is given by, with 
$n\ge 2$ denoting the total  number of ancestral lineages,  
\be
  \label{Xilambdab}
  \lambda_{n} = \int_\Delta \left(1 - \sum_{\ell = 0}^n \sum_{i_1 \neq
  \cdots \neq i_\ell } \binom{n}{\ell} x_{i_1}\cdots x_{i_\ell}\left(1
  - \sum_{i=1}^\infty x_i\right)^{n-\ell} \right)
  \frac{1}{\sum_{j=1}^\infty x_j^2}\Xi_0(dx) + a\binom{n}{2} \ee
  \citep{S00}.  Viewing coalescent processes strictly as mathematical
  objects, it is clear that the class of $\Xi$-coalescents contains
  $\Lambda$-coalescents as a specific example (i.e.\ allowing at most
  one group of lineages to merge each time), and the class of
  $\Lambda$-coalescents contain the Kingman-coalescent as a special
  case.  However, viewed as limits of ancestral processes derived from
  specific population models they are not nested, since one would
  obtain $\Lambda$-coalescents when deriving coalescent processes from
  haploid population models incorporating sweepstakes reproduction and
  high fecundity, and $\Xi$-coalescents for diploid populations.  One
  should therefore apply the models as appropriate, i.e.\
  $\Lambda$-coalescents to data (e.g.\ mtDNA data) inherited in a
  haploid fashion, and $\Xi$-coalescents to e.g.\ autosomal data
  inherited in diploid or polyploid fashion \citep{Blath2016}.

In \msprime we have incorporated two examples of multiple-merger
coalescents.  One is a diploid extension \citep{BBE13} of the haploid
model of sweepstakes reproduction considered by \cite{EW06}, which is
a haploid Moran model adapted to sweepstakes reproduction.  Let $N$
denote population size, and take $\psi \in (0,1]$ to be fixed.  In
every generation, with probability $1-\varepsilon_N$ a single
individual (picked uniformly at random) perishes, and one of the
surviving individuals (sampled uniformly at random) produces one
offspring; with probability $\varepsilon_N$ a total of
$\lfloor \psi N \rfloor$ individuals perish, and of the remaining
individuals a single individual produces $\lfloor \psi N \rfloor -1 $
offspring.  Taking $\varepsilon_N = 1/N^\gamma$ for some $\gamma > 0$,
\cite{EW06} obtain specific examples of $\Lambda$-coalescents, where
the $\Lambda$ measure in Eq \eqref{lambdabk} is a point mass at
$\psi$.  The simplicity of this model does allow one to obtain some
explicit mathematical results (see e.g.\
\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model
considered by \cite{EW06} has also been applied in algorithms for
simulating gene genealogies within phylogenies \citep{Zhu2015}. The
 specific model incorporated into \msprime is the diploid version
\citep{BBE13} of the model studied by \cite{EW06}, which would be
necessary in order to incorporate recombination.  In the model
considered by \cite{BBE13}, a single pair of diploid individuals
contribute offspring in each generation, selfing is excluded, and each
offspring is assigned one chromosome from each parent. There are,
therefore, four parent chromosomes involved in each reproduction
event, which can lead to up to four simultaneous mergers.  Let
\begin{esplit}{Cconst}
C_{b; k_1, \ldots, k_r;s } &=  \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell} (4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots + k_r + \ell},  \\
\end{esplit}
where  $C_{b; k_1, \ldots, k_r;s }$ corresponds to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq. \eqref{xi}) of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$.
The total merger rate (see Eq.\ \eqref{lambdabkall}) is then 
\be\label{xidirlambdabk}
      \lambda(b;k_1, \ldots, k_r) =    \mathcal{N}(b; k_1, \ldots, k_r ) \left( \frac{c\psi^2/4}{1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s } +     \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right),
\ee
and the total coalescence rate (see Eq.\ \eqref{xilambdab}) becomes
    \begin{esplit}{xilambdab}
\lambda_b & =  \frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0 }^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 - \psi)^{b-\ell } \right) +    \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.  \\
\end{esplit}
The interpretation of Eq.\ \eqref{xilambdab} is that with probability  $1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 + c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair produce  $\lfloor \psi N \rfloor$ offspring.

The other multiple-merger coalescent model incorporated in \msprime is
derived from an adaptation of the haploid population model considered
by \cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the
haploid version, in each generation individuals independently produce
random numbers of juveniles, where the random  number of juveniles $(X)$ produced by
a given individual  has the stable law \be\label{jX} \lim_{k\to
\infty} C k^\alpha \prb{X \ge k} = 1 \ee with index $\alpha > 0$, and
$C > 0$ is a normalising constant.   One can interpret  Eq.\ \eqref{jX} as  stating the form of the  probability distribution  for number of juveniles at least on the order of the population size.     If the random  total number of juveniles $(S_N)$ produced in this way is at least the population size $(N)$, then one samples  $N$ juveniles uniformly at random without replacement to form the next generation of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the same  set of individuals.   However,   assuming  $\EE{X} > 1$, one can show that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.         If $\alpha \ge 2$ the ancestral
process converges to the Kingman-coalescent; if $1 \le \alpha < 2$ the
ancestral process converges to a specific case of a
$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.\
\eqref{lambdabk} is associated with the Beta$(2-\alpha, \alpha)$
probability distribution, i.e.\
\be\label{Fbeta}
    \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)} x^{1 - \alpha}(1-x)^{\alpha - 1}  dx,
\ee
where $B(2-\alpha,\alpha)$ is the beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$ \citep{schweinsberg03}.    This model has been adapted to diploid populations  by                      \cite{BLS15}, where  the resulting coalescent process  is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$, where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.  The merger rate (see Eq.\ \eqref{xi}) is then ($1 \le r \le 4$)
\be\label{xibeta}
   \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b - k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)}
\ee
\citep{Blath2016,BLS15}. The interpretation of Eq.\ \eqref{xibeta} is  that the random   number of diploid  juveniles each  diploid pair of parents  produces  is  governed by  the law in Eq.\ \eqref{jX},   and   each diploid juvenile  is assigned  one chromosome  from each parent (selfing is excluded).

The model in Eq.\ \eqref{jX} assumes that individuals can produce
arbitrarily large numbers of juveniles. Considering diploid juveniles,
this assumption is probably rather strong, since diploid juveniles are
at least fertilised eggs, and so it is reasonable to suppose that the
number of juveniles surviving to the  particular life stage we are modeling  cannot be
arbitrarily large.  With this in mind, we also consider an adaptation
of the Schweinsberg model, where the random number of juveniles $(X)$
produced by a   given parent pair  is distributed according to
\be\label{jtr}
  \prb{X=k} =   \one{1 \le k \le \phi(N)} \frac{\phi(N+1)^\alpha }{ \phi(N+1)^\alpha - 1 }  \left( \frac{1}{k^\alpha} - \frac{1}{(k+1)^\alpha}  \right) ,
\ee
where $\phi(N)$ is a deterministic strict upper bound on the number of juveniles produced by  any given parent pair (see also \citep{Eldon2018}).   One can follow the calculations in  \citep{schweinsberg03} or \citep{BLS15}  to show  that  if $1 < \alpha < 2$   then  $(k = k_1 + \cdots + k_r)$ then the merger rate (see Eq.\ \eqref{xibeta}) is
\be
   \lambda_{b;k_1, \ldots, k_r} =  \sum_{\ell = 0}^{ (b - k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(M; k+\ell - \alpha, b-k-\ell + \alpha ) }{B(M;2-\alpha,\alpha)}
\ee
with $B(z;a,b) := \int_0^z t^{a-1}(1-t)^{b-1}dt$ for  $a,b>0$ and $0< z\le 1$, and
\be
M :=  \frac{K}{K+m} \one{\phi(N) = KN} + \one{\phi(N)/N \to \infty }
\ee
where $K > 0$ is a constant and  $m := \lim_{N\to \infty} \EE{X} = 1 + 2^{1-\alpha}/(\alpha - 1)$ \citep{CDEE2020,AEKKZ2020}.    In other words,  the measure $(\Lambda)$ driving the multiple mergers is of the same form as in Eq.\ \eqref{Fbeta}  with $0 < x \le M$ in the case $1 < \alpha < 2$ and $\phi(N) \ge KN$.   If $\alpha \ge 2$ or $\phi(N)/N \to 0$ then  the  ancestral process converges (in the sense of convergence of  finite-dimensional distributions) to  the Kingman-coalescent \citep{CDEE2020,AEKKZ2020}.

\cite{Becheler2020}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.

%assumption can be
%violated in some situations. Lambda coalescents
%%~\citep{donnelly1999particle,schweinsberg2000coalescents}
%%HybridLambda~\citep{zhu2015hybrid}

\begin{comment}%
A coalescent process is a continuous-time Markov process taking values among the
partitions of $\IN := \{1,2, \ldots \}$, such that the restriction to
any finite $n \in \IN $ takes values among partitions of
$[n] := \{1, 2, \ldots, n\}$.  Write $\one{A} = 1$ if $A$ holds, and zero otherwise.   Let $\cP_n$ denote the set of
partitions of $[n]$.  In the classical Kingman-coalescent, the only
possible transitions are the mergers of pairs of blocks (elements of a
partition $\pi \in \cP_n$), one pair at a time.  The $n$ leaves
(corresponding to the sampled DNA sequences) are arbitrarily labelled
from 1 to $n$, and  the blocks of a partition represent the common ancestors
of the labels of each block.    The initial state is  (usually) taken as 
$\{ \{1\}, \ldots, \{n\}\}$, and the final state, i.e.\ when the most 
recent common ancester is reached, as $\{ [n]\}$. A block in a partition of
$[n]$ represents an ancestor of the leaves in the block, i.e.\ the block
$\{i_1, \ldots, i_k\}$ in a given partition of $[n]$ is an ancestor of the
$k$ leaves $i_1, \ldots, i_k \in [n]$, and the leaves  correspond to
arbitrarily labelled DNA sequences in the sample.  
\end{comment}

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}%

### 2020-05-23 paper.tex
Update paper.tex

@@
 \usepackage{listings}
+\usepackage[english]{babel}%
+\usepackage[T1]{fontenc}%
+\usepackage[utf8]{inputenc}%
+\usepackage{amsmath,amssymb,amsfonts}%
+\usepackage{geometry}%
+\usepackage{dsfont}
+\usepackage{verbatim}%
+\usepackage{environ}%
+\usepackage[right]{lineno}%
+\usepackage{showkeys}
 
@@
 \newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
+
+%% local definitions for section multiple merger coalescents
+\newcommand{\msprime}[0]{{\texttt{msprime} }}
+\newcommand{\tskit}[0]{{\texttt{tskit} }}
+\newcommand{\ms}[0]{{\texttt{ms} }}
+\newcommand{\scrm}[0]{{\texttt{scrm} }}
+\newcommand{\stdpopsim}[0]{{\texttt{stdpopsim} }}
+ \newcommand{\be}{\begin{equation}}
+ \newcommand{\ee}{\end{equation}}
+ \newcommand{\bd}{\begin{displaymath}}
+ \newcommand{\ed}{\end{displaymath}}
+\newcommand{\IN}{\ensuremath{\mathds{N}}}%
+\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
+\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
+\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%
+
+\NewEnviron{esplit}[1]{%
+\begin{equation}
+\label{#1}
+\begin{split}
+  \BODY
+\end{split}\end{equation}
+}
 
@@
 \subsection*{Multiple merger coalescents}
-Kingman's coalescent assumes that all mergers are binary. This
-assumption can be violated in some situations. Lambda coalescents
-~\citep{donnelly1999particle,schweinsberg2000coalescents}
-HybridLambda~\citep{zhu2015hybrid}
-We support X Y Z.
+%%Kingman's coalescent assumes that all mergers are binary. This
+%%assumption can be violated in some situations. Lambda coalescents
+%%~\citep{donnelly1999particle,schweinsberg2000coalescents}
+%%HybridLambda~\citep{zhu2015hybrid}
+%%We support X Y Z.
+
+\label{mmc}%
+
+Kingman's coalescent assumes that at most two ancestral lineages can
+merge at each merger event.  This follows from the fact that the
+Kingman coalescent can be shown to be the process describing the
+random ancestral relations of gene copies sampled from a population
+evolving according to the Wright-Fisher model (or similar models) of
+genetic evolution.  Multiple-merger coalescent models can also be
+obtained from appropriate extensions of the Moran
+model \citep{EW06,HM12}, i.e.\ where a single randomly picked individual
+(e.g.\ in a haploid population), or a pair of diploid parent
+individuals \citep{BBE13}, produces offspring in each reproduction
+event.  Convergence to the Kingman coalescent follows from certain
+assumptions on the offspring distribution \citep{S99,MS01,SW08}.  These
+assumptions may be violated in certain highly fecund
+organisms \citep{hedgecock_94,B94,HP11,A04,irwin16}.  We take `high
+fecundity' to mean the ability of individuals to produce numbers of
+offspring on the order of the population size.  A key issue regarding
+high fecundity involves the concept of `sweepstakes
+reproduction' \citep{hedgecock_94}, where a few individuals may produce
+the bulk of the offspring in any given generation.  Sweepstakes
+reproduction is not captured by the Wrigth-Fisher model (or similar
+models).  Population models, which do capture high fecundity and
+sweepstakes reproduction, are in the domain of attraction of
+`multiple-merger' coalescents \citep{DK99,P99,S99,S00}, in which a
+random number of ancestral lineages may merge at any given time.  The
+term `multiple merger' refers to a merger of at least three ancestral
+lineages, or the simultaneous \citep{S00,MS01} merger of at least four
+lineages in at least two distinct groups, and each group involving at
+least two lineages.  Multiple-merger coalescent processes have also
+been shown to be relevant for modeling the effects of selection on
+gene genealogies \citep{Gillespie909,DS04}.
+
+Multiple-merger coalescents, in which a random number of ancestral
+lineages may merge at a given time in one group, and only one such
+group of lineages can merge at any given time (asynchronous multiple
+mergers), are referred to as $\Lambda$-coalescents; the rate at which  a given group of $k$ lineages out of a total of  $b$ lineages merges  is given by ($a\ge  0$ constant)
+\begin{equation}\label{lambdabk}
+\lambda_{b, k} =  \int_0^1  x^{k-2}(1-x)^{b-k}\Lambda(dx) + a\one{k=2}, \quad 2 \le k \le b, 
+\end{equation}
+where $\Lambda$ is a finite measure on the  (Borel subsets of)  unit interval without an atom at zero \citep{DK99,P99,S99}.    The  total rate  is given by
+\be\label{lambdab}
+ \lambda_{b} = \int_0^1 \left(1 - (1-x)^{b} - bx(1-x)^{b-1} \right)x^{-2}\Lambda(dx) + \binom{b}{2},
+\ee
+\citep{S99}  which can be useful in applications, in particular  provided the  antiderivative  can be explicitly identified.
+
+A larger class of multiple-merger coalescents involving simultaneous
+multiple mergers of distinct groups of ancestral lineages also exists
+\citep{S00}. These are commonly referred to as $\Xi$-coalescents, and
+can be shown to be the limits of ancestral processes derived from
+population models incorporating diploidy (or more general polyploidy)
+\citep{BBE13,Blath2016}, and certain models of selection \citep{DS04}.
+To describe a general $\Xi$-coalescent let $\Delta$ denote the
+infinite simplex \be\label{Delta} \Delta := \{ (x_1, \ldots ): x_1 \ge
+x_2 \ge \cdots \ge 0, \sum_{j}x_j \le 1\}; \ee for any
+$r \in \{1,2, \ldots\}$ let $k_1, \cdots, k_r \ge 2$, and
+$b = s + k_1 + \cdots + k_r$ be the total number of blocks in a given
+partition ($s = b - k_1 - \cdots - k_r$ is the number of blocks not
+participating in the mergers at the given time).  The existence of
+simultaneous multiple-merger coalescents was proved by \cite{S00}.
+There exists a finite measure $\Xi$ on $\Delta$, with
+$\Xi = \Xi_0 + a\delta_0$, the measure $\Xi_0$ has no atom at zero,
+$a >0$ is fixed, and the rate at which one sees a given (simultaneous)
+merger of ancestral lineages with merger sizes $k_1, \ldots, k_r$,
+$s = n - k_1 - \cdots - k_r$, is given by
+\begin{esplit}{xi}
+  \lambda_{n; k_1, \ldots, k_r; s}  & = \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
+_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
+  & +  a\one{r=1, k_1 = 2}.
+\end{esplit}%
+The number of such $(k_1, \ldots, k_r)$ mergers is
+\be\label{N}
+    \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
+\ee
+\citep{S00},   in particular  $\mathcal{N}(b;2) = b(b-1)/2$, and one can compute the total rate of a  $(k_1, \ldots, k_r)$ merger as
+\be\label{lambdabkall}
+      \lambda(n; k_1, \ldots, k_r)         =    \mathcal{N}(b; k_1, \ldots, k_r ) \lambda_{n; k_1, \ldots, k_r; s}.
+\ee
+
+The total rate is given by, with 
+$n\ge 2$ denoting the total  number of ancestral lineages,  
+\be
+  \label{Xilambdab}
+  \lambda_{n} = \int_\Delta \left(1 - \sum_{\ell = 0}^n \sum_{i_1 \neq
+  \cdots \neq i_\ell } \binom{n}{\ell} x_{i_1}\cdots x_{i_\ell}\left(1
+  - \sum_{i=1}^\infty x_i\right)^{n-\ell} \right)
+  \frac{1}{\sum_{j=1}^\infty x_j^2}\Xi_0(dx) + a\binom{n}{2} \ee
+  \citep{S00}.  Viewing coalescent processes strictly as mathematical
+  objects, it is clear that the class of $\Xi$-coalescents contains
+  $\Lambda$-coalescents as a specific example (i.e.\ allowing at most
+  one group of lineages to merge each time), and the class of
+  $\Lambda$-coalescents contain the Kingman-coalescent as a special
+  case.  However, viewed as limits of ancestral processes derived from
+  specific population models they are not nested, since one would
+  obtain $\Lambda$-coalescents when deriving coalescent processes from
+  haploid population models incorporating sweepstakes reproduction and
+  high fecundity, and $\Xi$-coalescents for diploid populations.  One
+  should therefore apply the models as appropriate, i.e.\
+  $\Lambda$-coalescents to data (e.g.\ mtDNA data) inherited in a
+  haploid fashion, and $\Xi$-coalescents to e.g.\ autosomal data
+  inherited in diploid or polyploid fashion \citep{Blath2016}.
+
+In \msprime we have incorporated two examples of multiple-merger
+coalescents.  One is a diploid extension \citep{BBE13} of the haploid
+model of sweepstakes reproduction considered by \cite{EW06}, which is
+a haploid Moran model adapted to sweepstakes reproduction.  Let $N$
+denote population size, and take $\psi \in (0,1]$ to be fixed.  In
+every generation, with probability $1-\varepsilon_N$ a single
+individual (picked uniformly at random) perishes, and one of the
+surviving individuals (sampled uniformly at random) produces one
+offspring; with probability $\varepsilon_N$ a total of
+$\lfloor \psi N \rfloor$ individuals perish, and of the remaining
+individuals a single individual produces $\lfloor \psi N \rfloor -1 $
+offspring.  Taking $\varepsilon_N = 1/N^\gamma$ for some $\gamma > 0$,
+\cite{EW06} obtain specific examples of $\Lambda$-coalescents, where
+the $\Lambda$ measure in Eq \eqref{lambdabk} is a point mass at
+$\psi$.  The simplicity of this model does allow one to obtain some
+explicit mathematical results (see e.g.\
+\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model
+considered by \cite{EW06} has also been applied in algorithms for
+simulating gene genealogies within phylogenies \citep{Zhu2015}. The
+ specific model incorporated into \msprime is the diploid version
+\citep{BBE13} of the model studied by \cite{EW06}, which would be
+necessary in order to incorporate recombination.  In the model
+considered by \cite{BBE13}, a single pair of diploid individuals
+contribute offspring in each generation, selfing is excluded, and each
+offspring is assigned one chromosome from each parent. There are,
+therefore, four parent chromosomes involved in each reproduction
+event, which can lead to up to four simultaneous mergers.  Let
+\begin{esplit}{Cconst}
+C_{b; k_1, \ldots, k_r;s } &=  \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell} (4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots + k_r + \ell},  \\
+\end{esplit}
+where  $C_{b; k_1, \ldots, k_r;s }$ corresponds to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq. \eqref{xi}) of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$.
+The total merger rate (see Eq.\ \eqref{lambdabkall}) is then 
+\be\label{xidirlambdabk}
+      \lambda(b;k_1, \ldots, k_r) =    \mathcal{N}(b; k_1, \ldots, k_r ) \left( \frac{c\psi^2/4}{1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s } +     \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right),
+\ee
+and the total coalescence rate (see Eq.\ \eqref{xilambdab}) becomes
+    \begin{esplit}{xilambdab}
+\lambda_b & =  \frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0 }^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 - \psi)^{b-\ell } \right) +    \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.
+\\
+\end{esplit}
+The interpretation of Eq.\ \eqref{xilambdab} is that with probability  $1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 + c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair produce  $\lfloor \psi N \rfloor$ offspring.
+
+The other multiple-merger coalescent model incorporated in \msprime is
+derived from an adaptation of the haploid population model considered
+by \cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the
+haploid version, in each generation individuals independently produce
+random numbers of juveniles, where the random  number of juveniles $(X)$ produced by
+a given individual  has the stable law \be\label{jX} \lim_{k\to
+\infty} C k^\alpha \prb{X \ge k} = 1 \ee with index $\alpha > 0$, and
+$C > 0$ is a normalising constant.
+One can interpret  Eq.\ \eqref{jX} as  stating the form of the  probability distribution  for number of juveniles at least on the order of the population size.
+If the random  total number of juveniles $(S_N)$ produced in this way is at least the population size $(N)$, then one samples  $N$ juveniles uniformly at random without replacement to form the next generation of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the same  set of individuals.
+However,   assuming  $\EE{X} > 1$, one can show that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.
+If $\alpha \ge 2$ the ancestral
+process converges to the Kingman-coalescent; if $1 \le \alpha < 2$ the
+ancestral process converges to a specific case of a
+$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.\
+\eqref{lambdabk} is associated with the Beta$(2-\alpha, \alpha)$
+probability distribution, i.e.\
+\be\label{Fbeta}
+    \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)} x^{1 - \alpha}(1-x)^{\alpha - 1}  dx,
+\ee
+where $B(2-\alpha,\alpha)$ is the beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$ \citep{schweinsberg03}.
+This model has been adapted to diploid populations  by                      \cite{BLS15}, where  the resulting coalescent process  is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$, where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.
+The merger rate (see Eq.\ \eqref{xi}) is then ($1 \le r \le 4$)
+\be\label{xibeta}
+   \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b - k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)}
+\ee
+\citep{Blath2016,BLS15}.
+The interpretation of Eq.\ \eqref{xibeta} is  that the random   number of diploid  juveniles each  diploid pair of parents  produces  is  governed by  the law in Eq.\ \eqref{jX},   and   each diploid juvenile  is assigned  one chromosome  from each parent (selfing is excluded).
+
+The model in Eq.\ \eqref{jX} assumes that individuals can produce
+arbitrarily large numbers of juveniles. Considering diploid juveniles,
+this assumption is probably rather strong, since diploid juveniles are
+at least fertilised eggs, and so it is reasonable to suppose that the
+number of juveniles surviving to the  particular life stage we are modeling  cannot be
+arbitrarily large.  With this in mind, we also consider an adaptation
+of the Schweinsberg model, where the random number of juveniles $(X)$
+produced by a   given parent pair  is distributed according to
+\be\label{jtr}
+  \prb{X=k} =   \one{1 \le k \le \phi(N)} \frac{\phi(N+1)^\alpha }{ \phi(N+1)^\alpha - 1 }  \left( \frac{1}{k^\alpha} - \frac{1}{(k+1)^\alpha}  \right) ,
+\ee
+where $\phi(N)$ is a deterministic strict upper bound on the number of juveniles produced by  any given parent pair (see also \citep{Eldon2018}).
+One can follow the calculations in  \citep{schweinsberg03} or \citep{BLS15}  to show  that  if $1 < \alpha < 2$   then  $(k = k_1 + \cdots + k_r)$ then the merger rate (see Eq.\ \eqref{xibeta}) is
+\be
+   \lambda_{b;k_1, \ldots, k_r} =  \sum_{\ell = 0}^{ (b - k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(M; k+\ell - \alpha, b-k-\ell + \alpha ) }{B(M;2-\alpha,\alpha)}
+\ee
+with $B(z;a,b) := \int_0^z t^{a-1}(1-t)^{b-1}dt$ for  $a,b>0$ and $0< z\le 1$, and
+\be
+M :=  \frac{K}{K+m} \one{\phi(N) = KN} + \one{\phi(N)/N \to \infty }
+\ee
+where $K > 0$ is a constant and  $m := \lim_{N\to \infty} \EE{X} = 1 + 2^{1-\alpha}/(\alpha - 1)$ \citep{CDEE2020,AEKKZ2020}.
+In other words,  the measure $(\Lambda)$ driving the multiple mergers is of the same form as in Eq.\ \eqref{Fbeta}  with $0 < x \le M$ in the case $1 < \alpha < 2$ and $\phi(N) \ge KN$.
+If $\alpha \ge 2$ or $\phi(N)/N \to 0$ then  the  ancestral process converges (in the sense of convergence of  finite-dimensional distributions) to  the Kingman-coalescent \citep{CDEE2020,AEKKZ2020}.
+
+\cite{Becheler2020}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.
+
+%assumption can be
+%violated in some situations. Lambda coalescents
+%%~\citep{donnelly1999particle,schweinsberg2000coalescents}
+%%HybridLambda~\citep{zhu2015hybrid}
+
+\begin{comment}%
+A coalescent process is a continuous-time Markov process taking values among the
+partitions of $\IN := \{1,2, \ldots \}$, such that the restriction to
+any finite $n \in \IN $ takes values among partitions of
+$[n] := \{1, 2, \ldots, n\}$.  Write $\one{A} = 1$ if $A$ holds, and zero otherwise.   Let $\cP_n$ denote the set of
+partitions of $[n]$.  In the classical Kingman-coalescent, the only
+possible transitions are the mergers of pairs of blocks (elements of a
+partition $\pi \in \cP_n$), one pair at a time.  The $n$ leaves
+(corresponding to the sampled DNA sequences) are arbitrarily labelled
+from 1 to $n$, and  the blocks of a partition represent the common ancestors
+of the labels of each block.    The initial state is  (usually) taken as 
+$\{ \{1\}, \ldots, \{n\}\}$, and the final state, i.e.\ when the most 
+recent common ancester is reached, as $\{ [n]\}$. A block in a partition of
+$[n]$ represents an ancestor of the leaves in the block, i.e.\ the block
+$\{i_1, \ldots, i_k\}$ in a given partition of $[n]$ is an ancestor of the
+$k$ leaves $i_1, \ldots, i_k \in [n]$, and the leaves  correspond to
+arbitrarily labelled DNA sequences in the sample.  
+\end{comment}
 

### 2020-06-12 paper.tex
Update paper.tex
Added a short feature section on instantaneous bottlenecks

@@
 
-\subsection*{Simultaneous bottlenecks}
-Overview of what simulataneous bottlenecks literature. What we
-do [Konrad].
+\subsection*{Instantaneous bottlenecks}
+A common approach to modelling the effect of demographic history on genealogies is to assume that effective population size ($N_e$) changes in discrete steps which define a series of epochs~\citep{griffiths1994sampling, marth2004allele, keightley2007joint, li2011inference}.
+In this setting of piece-wise constant $N_e$, capturing a population bottleneck requires three epochs: $N_e$ is reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and recovers to its initial value at time $T_{end}$~\citep{marth2004allele}.
+If bottlenecks are short both on the timescale of coalescence and mutations, one may expect little information about the duration of a bottleneck (T_end-T_end) in sequence data.
+Thus a simpler, alternative model is to assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$) and generate a sudden burst of coalescence events (a multiple merger event) in the genealogy.
+The strength of the bottleneck B can be thought of as an (imaginary) time period during which coalescence events are collapsed, i.e.
+there is no growth in genealogical branches during B and the probability that a single pair of lineages entering the bottleneck coalesce during the bottleneck is $1-e^{-B}$.
+Although this simple two parameter model of bottlenecks is attractive and both analytic results and empirical inference~\citep{griffiths1994sampling, galtier2000detecting, bunnefeld2015inferring} have been developed under this model, it has not been straightforward to simulate data under instantaneous bottleneck histories with \ms\.
+Instantaneous bottlenecks are implemented as a demographic event in \msprime\.
 

### 2020-09-27 paper.tex

@@
 \usepackage[right]{lineno}%
-\usepackage{showkeys}
+%\usepackage{showkeys}
 
@@
 %% local definitions for section multiple merger coalescents
-\newcommand{\msprime}[0]{{\texttt{msprime} }}
-\newcommand{\tskit}[0]{{\texttt{tskit} }}
-\newcommand{\ms}[0]{{\texttt{ms} }}
-\newcommand{\scrm}[0]{{\texttt{scrm} }}
-\newcommand{\stdpopsim}[0]{{\texttt{stdpopsim} }}
  \newcommand{\be}{\begin{equation}}
@@
 \subsection*{Ancestral recombination graphs}
+
+The \msprime~simulator generates tree sequences by simulating realisations of the
+Ancestral Recombination Graph (ARG).
+By default, only the marginal genealogy at each site is recorded in the tree sequence
+output.
+Marginal genealogies are sufficient for e.g.~simulating and storing sequence data,
+and involve fewer edges than nodes than storing a realisation of the full ARG.
+However, the mapping from ARG realisations to marginal tree sequences is not lossless.
+For example, times of recombination events can be recovered from the former
+but not the latter.
+The same is true of the number of extant ancestors at each time in the past.
+In addition to the default behaviour, we support the retention of the full
+ARG realisation using the \texttt{record\_full\_arg} option.
+This option augments the default tree sequence data structure with three
+new kinds of nodes:
+\begin{itemize}
+\item Each recombination event is marked by two nodes with identical birth times
+given by the time of the recombination event, and marked with the
+\texttt{NODE\_IS\_RE\_EVENT} flag. 
+Each node represents an ancestor of the recombinant child node, and edges connect
+the relevant segments of genome from the child node to these two ancestors.
+\item We store common ancestor nodes even when the corresponding merger event
+does not result in coalescence of ancestral material.
+The resulting nodes are marked with the \texttt{NODE\_IS\_CA\_EVENT} flag.
+\item When multiple populations are simulated, migration events are depicted by nodes
+with the \texttt{NODE\_IS\_MIG\_EVENT} flag, and with birth times equal to the time of
+the migration.
+The population flag of the migration node identifies the population to which the lineage
+belongs after the reverse time migration event.
+\end{itemize}
+
+The likelihood, i.e.~the joint sampling probability of the ancestry and mutations,
+is an example of a function which is intractable when the ancestry is stored as a
+marginal tree sequence, but tractable when the ancestry is given by a full
+ARG realisation.
+The likelihood is a central quantity in evaluating the plausibility of a putative
+ancestry as an explanation of DNA sequence data, both directly through
+e.g.~approaches based on maximum likelihood, and as an ingredient of
+methods such as Metropolis-Hastings
+\citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian}.
+
+We provide two functions to facilitate likelihood-based inference.
+Both are implemented only for the simplest case of the standard ARG with a 
+constant population size, and require tree sequences compatible with the 
+\texttt{record\_full\_arg} option as their arguments.
+\begin{itemize}
+\item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of 
+the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
+per-generation recombination probability \texttt{r} and population size \texttt{N}.
+Specifically, the function returns the logarithm of
+\begin{equation*}
+\Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
+	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 } 
+		+ r l_i \Big] t_i  \Bigg),
+\end{equation*}
+where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
+$k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
+in that interval that would split ancestral material should they recombine,
+$q$ is the total number of events in the tree sequence $\texttt{ts}$,
+$q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices 
+of time intervals which end in a recombination, and $g_i$ is the corresponding
+\emph{gap}:  the length of material in which the recombination at the end of the 
+$i$th period could have taken place without affecting the resulting genealogy.
+A recombination in ancestral material will always have $g_i = 1$, while a recombination
+in trapped non-ancestral material can result in other values of $g_i$.
+For a continuous model of the genome and a recombination in ancestral material,
+we set $g_i = 1$ and interpret the result as a density.
+\item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
+natural logarithm of the probability of the mutations recorded in the tree sequence
+\texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
+a normalising constant which depends on the pattern of mutations,
+but not on the tree sequence or the per-site, per-generation mutation 
+probability \texttt{m}.
+Specifically, the function returns the logarithm of
+\begin{equation*}
+e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
+\prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
+\end{equation*}
+where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
+in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
+total branch length on which $\gamma$ could have arisen while appearing on all
+of the leaves of \texttt{ts} it does, and on no others.
+Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
+mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
+edges.
+\end{itemize}
+
+\begin{comment}
+Previous draft of the ARG feature retained here as a comment.
+
 The definition of ARGs is vague and confusing.
@@
 under the~\cite{kuhner2000maximum} model.
+\end{comment}
 
@@
 In this setting of piece-wise constant $N_e$, capturing a population bottleneck requires three epochs: $N_e$ is reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and recovers to its initial value at time $T_{end}$~\citep{marth2004allele}.
-If bottlenecks are short both on the timescale of coalescence and mutations, one may expect little information about the duration of a bottleneck (T_end-T_end) in sequence data.
+If bottlenecks are short both on the timescale of coalescence and mutations, one may expect little information about the duration of a bottleneck ($T_{end}-T_{start}$) in sequence data.
 Thus a simpler, alternative model is to assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$) and generate a sudden burst of coalescence events (a multiple merger event) in the genealogy.
@@
 there is no growth in genealogical branches during B and the probability that a single pair of lineages entering the bottleneck coalesce during the bottleneck is $1-e^{-B}$.
-Although this simple two parameter model of bottlenecks is attractive and both analytic results and empirical inference~\citep{griffiths1994sampling, galtier2000detecting, bunnefeld2015inferring} have been developed under this model, it has not been straightforward to simulate data under instantaneous bottleneck histories with \ms\.
-Instantaneous bottlenecks are implemented as a demographic event in \msprime\.
+Although this simple two parameter model of bottlenecks is attractive and both analytic results and empirical inference~\citep{griffiths1994sampling, galtier2000detecting, bunnefeld2015inferring} have been developed under this model, it has not been straightforward to simulate data under instantaneous bottleneck histories with \ms.
+Instantaneous bottlenecks are implemented as a demographic event in \msprime.
 
@@
 \section*{Acknowledgments}
-JK is supported by the Robertson Foundation. [Others fill in their ACKs please]
+JK is supported by the Robertson Foundation. 
+Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
+[Others fill in their ACKs please]
 

### 2021-06-03 paper.tex

@@
 considered by \cite{EW06} has also been applied in algorithms for
-simulating gene genealogies within phylogenies \citep{Zhu2015}. The
+simulating gene genealogies within phylogenies \citep{zhu2015hybrid}. The
  specific model incorporated into \msprime is the diploid version
@@
 
-\cite{Becheler2020}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.
+\cite{becheler2020occupancy}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.
 

### 2021-06-07 paper.tex
Update mutation models.

@@
 Coalescent simulation is a key tool in population genetics and
-and has been integral to coalescent theory since its earliest days.
+has been integral to coalescent theory since its beginnings.
 Because of the ease at which the basic model can be simulated,
@@
 recombination must be fully accounted for. While the
-single-locus coalescent can be simulated in constant
+single-locus coalescent for $n$ samples can be simulated in $O(n)$
 time~\citep{hudson1990gene}, the coalescent with recombination is
-far more challenging to simulate.
+far more challenging.
 % Sentence is too long...
@@
 can only simulate short segments under the influence of recombination,
-and, until recently, we had to resort to
-approximate models of recombination~\citep{mcvean2005approximating,staab2014scrm}
-if we wished to simulate whole chromosomes.
+and, until recently, approximate models of
+recombination~\citep{mcvean2005approximating,staab2014scrm}
+were necessary to simulate whole chromosomes.
 
@@
 thousands of genomes, and there is every reason to believe that
-datasets of millions will soon be commonplace~\citep{stephens2015big}.
+researchers will be soon be routinely working with
+datasets containing millions of samples~\citep{stephens2015big}.
 Classical simulators such as \ms\ and even fast approximate simulators
@@
 sequence'' data
-structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring},
+structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
+wohns2021unified},
 introduced as part of \msprime\, makes it possible to store simulations
@@
 of magnitude smaller than the standard
-Newick~\citep{felsenstein1989phylip} formats. This data structure has
+Newick~\citep{felsenstein1989phylip} based formats. This data structure has
 also lead to major advances in forwards-time
 simulation~\citep{kelleher2018efficient,haller2018tree},
-ancestry inference~\citep{kelleher2019inferring}
+ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
 and calculation of population genetic statistics~\citep{ralph2019efficiently}.
@@
 \texttt{msHOT}~\citep{hellenthal2007mshot} and also empirical
-recombination maps like those supported by X Y Z.
+recombination maps like those supported \texttt{cosi2}
+\citep{shlyakhter2014cosi2}
 (What about ~\cite{wang2014new}?)
@@
 \subsection*{SMC approximations}
+
+\textbf{TODO:} It's important that we acknowledge that (a) the coalescent
+with recombination does become quite slow for large $N_e$; (b) our
+implementation of the SMC is naive; and (c) for sufficiently large
+$N_e$ programs like \texttt{scrm} will be faster.
 
@@
 
-\subsection*{Seedbank coalescent}
-Seedbank coalescent. [If we implement this fill in please Thibaud]
+\subsection*{Finite sites mutations}
+Version 0.x of \texttt{msprime} provided a very simple mutation model,
+supporting only the infinite sites model.
+With version 1.0, we now support a large number of mutation models,
+including
+JC69~\citep{jukes1969evolution},
+F84~\citep{felsenstein1996hidden},
+and
+GTR~\citep{tavare1986some} nucleotide models
+and
+the BLOSUM62~\citep{henikoff1992amino}
+and PAM~\citep{dayhoff1978} amino acid models.
+The general model accepts an arbitrary transition matrix between
+any number of alleles (which can be arbitrary unicode strings),
+and so other models such as the Kimura two and three
+parameter models~\citep{kimura1980simple,kimura1981estimation}
+parameters models can be easily
+and efficiently defined in user code.
+Full support for the original infinite sites model is also maintained.
+
+\textbf{TODO} Compare performance and features with
+Seq-Gen~\citep{rambaut1997seq},
+Dawg~\citep{cartwright2005dna} and Pyvolve~\citep{spielman2015pyvolve}
 
@@
 Thus, development of \msprime\ is not completed and much remains
-to be done. For example, one key aim of the 1.0 series is to
-improve the mutation models supported which are limited to
-simple infinite-sites currently. The ability to simulate
-sophisticated mutation models like Seq-Gen~\citep{rambaut1997seq},
-Dawg~\citep{cartwright2005dna} and Pyvolve~\citep{spielman2015pyvolve} at the megasample scale would be
-useful in many applications. Another key focus of development
-will be [WHAT?]
+to be done.
+
+% We need to acknowledge that for long genomes and for large population
+% sizes msprime's performance does drop off and SMC simulators like
+% scrm become more efficient.
+Msprime's performance for large sample sizes is excellent, but
+simulating species with large population sizes is still challenging.
+An important avenue for future development is to improve performance
+in these scenarios by incorporating data structures such as skip lists
+and implementing more efficient coalescent approximations, as done
+in \texttt{cosi2}~\citep{shlyakhter2014cosi2}.
 

### 2021-06-07 paper.tex

@@
 as input and complete the simulation (``recapitation'').
+
+\subsection*{Finite sites mutations}
+Version 0.x of \texttt{msprime} provided a very simple mutation model,
+supporting only the infinite sites model.
+With version 1.0, we now support a large number of mutation models,
+including
+JC69~\citep{jukes1969evolution},
+F84~\citep{felsenstein1996hidden},
+and
+GTR~\citep{tavare1986some} nucleotide models
+and
+the BLOSUM62~\citep{henikoff1992amino}
+and PAM~\citep{dayhoff1978} amino acid models.
+The general model accepts an arbitrary transition matrix between
+any number of alleles (which can be arbitrary unicode strings),
+and so other models such as the Kimura two and three
+parameter models~\citep{kimura1980simple,kimura1981estimation}
+parameters models can be easily
+and efficiently defined in user code.
+Full support for the original infinite sites model is also maintained.
+The results of mutation simulations across a range of models are
+extensively validated
+against both theoretical expectations and output from
+Seq-Gen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}, and are run before every
+release.
+
+Comparing the performance of mutation generation methods is problematic.
+Because there is no standard format for representing trees and
+the accompanying mutations, all existing methods are focused on
+writing out the resulting sequences. Methods are also not modular,
+and so it is very difficult to disentangle the time required to
+read in the tree file and output the sequences from the actual
+time required to perform the simulation.
+
+The mutation generation machinery in \msprime\ is extremely
+efficient. For example, we simulated mutations under the JC69
+model for a tree with $10^6$ leaves over $10^4$ sites (resulting
+in $\sim 280000$ mutations) in less that 1.5 seconds, including
+the time required for file input and output. While a systematic
+comparison of run-times is not feasible here, this is orders
+of magnitude faster than methods such as
+Seq-Gen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}
+and at least competitive with the recent method from
+\cite{demaio2021phastsim}.
+
+\textbf{TODO: would love to do a like-for-like comparison, but quite hard}
+
+Another substantial advantage of our approach is that the simulated
+mutations are stored directly in the output data format with
+a defined API. Although
+Pyvolve and Phastsim are implemented in Python, neither provide
+programatic access to the simulated output, outputting instead
+to a series of files which must subsequently be parsed.
+See the Data Model section for more details and discussion.
 
@@
 
-\subsection*{Finite sites mutations}
-Version 0.x of \texttt{msprime} provided a very simple mutation model,
-supporting only the infinite sites model.
-With version 1.0, we now support a large number of mutation models,
-including
-JC69~\citep{jukes1969evolution},
-F84~\citep{felsenstein1996hidden},
-and
-GTR~\citep{tavare1986some} nucleotide models
-and
-the BLOSUM62~\citep{henikoff1992amino}
-and PAM~\citep{dayhoff1978} amino acid models.
-The general model accepts an arbitrary transition matrix between
-any number of alleles (which can be arbitrary unicode strings),
-and so other models such as the Kimura two and three
-parameter models~\citep{kimura1980simple,kimura1981estimation}
-parameters models can be easily
-and efficiently defined in user code.
-Full support for the original infinite sites model is also maintained.
-
-\textbf{TODO} Compare performance and features with
-Seq-Gen~\citep{rambaut1997seq},
-Dawg~\citep{cartwright2005dna} and Pyvolve~\citep{spielman2015pyvolve}
-
 \subsection*{Simulation interface}
@@
 
-\subsection*{Output}
+\subsection*{Data model}
 The output of \msprime\ is a ``succinct tree sequence`` (or tree sequence,
@@
 of other simulators, and the perf advantages of tree sequences.]
+
+[Review the data model in Phastsim~\cite{demaio2021phastsim} and
+Usher~\citep{turakhia2021ultrafast}.]
 
@@
 
+In terms of mutation simulation, future work will need to do things like
+Dawg~\citep{cartwright2005dna} and to implement
+codon models and indels~\citep{fletcher2009indelible}.
+
 % Finish up with a paragraph saying that there loads of exciting

### 2021-07-05 paper.tex
Moved multiple merger derivations to the appendix.

@@
 \subsection*{Multiple merger coalescents}
-%%Kingman's coalescent assumes that all mergers are binary. This
-%%assumption can be violated in some situations. Lambda coalescents
-%%~\citep{donnelly1999particle,schweinsberg2000coalescents}
-%%HybridLambda~\citep{zhu2015hybrid}
-%%We support X Y Z.
-
-\label{mmc}%
 
@@
 
-Multiple-merger coalescents, in which a random number of ancestral
-lineages may merge at a given time in one group, and only one such
-group of lineages can merge at any given time (asynchronous multiple
-mergers), are referred to as $\Lambda$-coalescents; the rate at which  a given group of $k$ lineages out of a total of  $b$ lineages merges  is given by ($a\ge  0$ constant)
-\begin{equation}\label{lambdabk}
-\lambda_{b, k} =  \int_0^1  x^{k-2}(1-x)^{b-k}\Lambda(dx) + a\one{k=2}, \quad 2 \le k \le b,
-\end{equation}
-where $\Lambda$ is a finite measure on the  (Borel subsets of)  unit interval without an atom at zero \citep{DK99,P99,S99}.    The  total rate  is given by
-\be\label{lambdab}
- \lambda_{b} = \int_0^1 \left(1 - (1-x)^{b} - bx(1-x)^{b-1} \right)x^{-2}\Lambda(dx) + \binom{b}{2},
-\ee
-\citep{S99}  which can be useful in applications, in particular  provided the  antiderivative  can be explicitly identified.
-
-A larger class of multiple-merger coalescents involving simultaneous
-multiple mergers of distinct groups of ancestral lineages also exists
-\citep{S00}. These are commonly referred to as $\Xi$-coalescents, and
-can be shown to be the limits of ancestral processes derived from
-population models incorporating diploidy (or more general polyploidy)
-\citep{BBE13,Blath2016}, and certain models of selection \citep{DS04}.
-To describe a general $\Xi$-coalescent let $\Delta$ denote the
-infinite simplex \be\label{Delta} \Delta := \{ (x_1, \ldots ): x_1 \ge
-x_2 \ge \cdots \ge 0, \sum_{j}x_j \le 1\}; \ee for any
-$r \in \{1,2, \ldots\}$ let $k_1, \cdots, k_r \ge 2$, and
-$b = s + k_1 + \cdots + k_r$ be the total number of blocks in a given
-partition ($s = b - k_1 - \cdots - k_r$ is the number of blocks not
-participating in the mergers at the given time).  The existence of
-simultaneous multiple-merger coalescents was proved by \cite{S00}.
-There exists a finite measure $\Xi$ on $\Delta$, with
-$\Xi = \Xi_0 + a\delta_0$, the measure $\Xi_0$ has no atom at zero,
-$a >0$ is fixed, and the rate at which one sees a given (simultaneous)
-merger of ancestral lineages with merger sizes $k_1, \ldots, k_r$,
-$s = n - k_1 - \cdots - k_r$, is given by
-\begin{esplit}{xi}
-  \lambda_{n; k_1, \ldots, k_r; s}  & = \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
-_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
-  & +  a\one{r=1, k_1 = 2}.
-\end{esplit}%
-The number of such $(k_1, \ldots, k_r)$ mergers is
-\be\label{N}
-    \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
-\ee
-\citep{S00},   in particular  $\mathcal{N}(b;2) = b(b-1)/2$, and one can compute the total rate of a  $(k_1, \ldots, k_r)$ merger as
-\be\label{lambdabkall}
-      \lambda(n; k_1, \ldots, k_r)         =    \mathcal{N}(b; k_1, \ldots, k_r ) \lambda_{n; k_1, \ldots, k_r; s}.
-\ee
-
-The total rate is given by, with
-$n\ge 2$ denoting the total  number of ancestral lineages,
-\be
-  \label{Xilambdab}
-  \lambda_{n} = \int_\Delta \left(1 - \sum_{\ell = 0}^n \sum_{i_1 \neq
-  \cdots \neq i_\ell } \binom{n}{\ell} x_{i_1}\cdots x_{i_\ell}\left(1
-  - \sum_{i=1}^\infty x_i\right)^{n-\ell} \right)
-  \frac{1}{\sum_{j=1}^\infty x_j^2}\Xi_0(dx) + a\binom{n}{2} \ee
-  \citep{S00}.  Viewing coalescent processes strictly as mathematical
-  objects, it is clear that the class of $\Xi$-coalescents contains
-  $\Lambda$-coalescents as a specific example (i.e.\ allowing at most
-  one group of lineages to merge each time), and the class of
-  $\Lambda$-coalescents contain the Kingman-coalescent as a special
-  case.  However, viewed as limits of ancestral processes derived from
-  specific population models they are not nested, since one would
-  obtain $\Lambda$-coalescents when deriving coalescent processes from
-  haploid population models incorporating sweepstakes reproduction and
-  high fecundity, and $\Xi$-coalescents for diploid populations.  One
-  should therefore apply the models as appropriate, i.e.\
-  $\Lambda$-coalescents to data (e.g.\ mtDNA data) inherited in a
-  haploid fashion, and $\Xi$-coalescents to e.g.\ autosomal data
-  inherited in diploid or polyploid fashion \citep{Blath2016}.
-
-In \msprime we have incorporated two examples of multiple-merger
-coalescents.  One is a diploid extension \citep{BBE13} of the haploid
-model of sweepstakes reproduction considered by \cite{EW06}, which is
-a haploid Moran model adapted to sweepstakes reproduction.  Let $N$
-denote population size, and take $\psi \in (0,1]$ to be fixed.  In
-every generation, with probability $1-\varepsilon_N$ a single
-individual (picked uniformly at random) perishes, and one of the
-surviving individuals (sampled uniformly at random) produces one
-offspring; with probability $\varepsilon_N$ a total of
-$\lfloor \psi N \rfloor$ individuals perish, and of the remaining
-individuals a single individual produces $\lfloor \psi N \rfloor -1 $
-offspring.  Taking $\varepsilon_N = 1/N^\gamma$ for some $\gamma > 0$,
-\cite{EW06} obtain specific examples of $\Lambda$-coalescents, where
-the $\Lambda$ measure in Eq \eqref{lambdabk} is a point mass at
-$\psi$.  The simplicity of this model does allow one to obtain some
-explicit mathematical results (see e.g.\
-\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model
-considered by \cite{EW06} has also been applied in algorithms for
-simulating gene genealogies within phylogenies \citep{zhu2015hybrid}. The
- specific model incorporated into \msprime is the diploid version
-\citep{BBE13} of the model studied by \cite{EW06}, which would be
-necessary in order to incorporate recombination.  In the model
-considered by \cite{BBE13}, a single pair of diploid individuals
-contribute offspring in each generation, selfing is excluded, and each
-offspring is assigned one chromosome from each parent. There are,
-therefore, four parent chromosomes involved in each reproduction
-event, which can lead to up to four simultaneous mergers.  Let
-\begin{esplit}{Cconst}
-C_{b; k_1, \ldots, k_r;s } &=  \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell} (4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots + k_r + \ell},  \\
-\end{esplit}
-where  $C_{b; k_1, \ldots, k_r;s }$ corresponds to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq. \eqref{xi}) of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$.
-The total merger rate (see Eq.\ \eqref{lambdabkall}) is then
-\be\label{xidirlambdabk}
-      \lambda(b;k_1, \ldots, k_r) =    \mathcal{N}(b; k_1, \ldots, k_r ) \left( \frac{c\psi^2/4}{1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s } +     \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right),
-\ee
-and the total coalescence rate (see Eq.\ \eqref{xilambdab}) becomes
-    \begin{esplit}{xilambdab}
-\lambda_b & =  \frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0 }^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 - \psi)^{b-\ell } \right) +    \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.
-\\
-\end{esplit}
-The interpretation of Eq.\ \eqref{xilambdab} is that with probability  $1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 + c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair produce  $\lfloor \psi N \rfloor$ offspring.
-
-The other multiple-merger coalescent model incorporated in \msprime is
-derived from an adaptation of the haploid population model considered
-by \cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the
-haploid version, in each generation individuals independently produce
-random numbers of juveniles, where the random  number of juveniles $(X)$ produced by
-a given individual  has the stable law \be\label{jX} \lim_{k\to
-\infty} C k^\alpha \prb{X \ge k} = 1 \ee with index $\alpha > 0$, and
-$C > 0$ is a normalising constant.
-One can interpret  Eq.\ \eqref{jX} as  stating the form of the  probability distribution  for number of juveniles at least on the order of the population size.
-If the random  total number of juveniles $(S_N)$ produced in this way is at least the population size $(N)$, then one samples  $N$ juveniles uniformly at random without replacement to form the next generation of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the same  set of individuals.
-However,   assuming  $\EE{X} > 1$, one can show that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.
-If $\alpha \ge 2$ the ancestral
-process converges to the Kingman-coalescent; if $1 \le \alpha < 2$ the
-ancestral process converges to a specific case of a
-$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.\
-\eqref{lambdabk} is associated with the Beta$(2-\alpha, \alpha)$
-probability distribution, i.e.\
-\be\label{Fbeta}
-    \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)} x^{1 - \alpha}(1-x)^{\alpha - 1}  dx,
-\ee
-where $B(2-\alpha,\alpha)$ is the beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$ \citep{schweinsberg03}.
-This model has been adapted to diploid populations  by                      \cite{BLS15}, where  the resulting coalescent process  is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$, where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.
-The merger rate (see Eq.\ \eqref{xi}) is then ($1 \le r \le 4$)
-\be\label{xibeta}
-   \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b - k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)}
-\ee
-\citep{Blath2016,BLS15}.
-The interpretation of Eq.\ \eqref{xibeta} is  that the random   number of diploid  juveniles each  diploid pair of parents  produces  is  governed by  the law in Eq.\ \eqref{jX},   and   each diploid juvenile  is assigned  one chromosome  from each parent (selfing is excluded).
-
-The model in Eq.\ \eqref{jX} assumes that individuals can produce
-arbitrarily large numbers of juveniles. Considering diploid juveniles,
-this assumption is probably rather strong, since diploid juveniles are
-at least fertilised eggs, and so it is reasonable to suppose that the
-number of juveniles surviving to the  particular life stage we are modeling  cannot be
-arbitrarily large.  With this in mind, we also consider an adaptation
-of the Schweinsberg model, where the random number of juveniles $(X)$
-produced by a   given parent pair  is distributed according to
-\be\label{jtr}
-  \prb{X=k} =   \one{1 \le k \le \phi(N)} \frac{\phi(N+1)^\alpha }{ \phi(N+1)^\alpha - 1 }  \left( \frac{1}{k^\alpha} - \frac{1}{(k+1)^\alpha}  \right) ,
-\ee
-where $\phi(N)$ is a deterministic strict upper bound on the number of juveniles produced by  any given parent pair (see also \citep{Eldon2018}).
-One can follow the calculations in  \citep{schweinsberg03} or \citep{BLS15}  to show  that  if $1 < \alpha < 2$   then  $(k = k_1 + \cdots + k_r)$ then the merger rate (see Eq.\ \eqref{xibeta}) is
-\be
-   \lambda_{b;k_1, \ldots, k_r} =  \sum_{\ell = 0}^{ (b - k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(M; k+\ell - \alpha, b-k-\ell + \alpha ) }{B(M;2-\alpha,\alpha)}
-\ee
-with $B(z;a,b) := \int_0^z t^{a-1}(1-t)^{b-1}dt$ for  $a,b>0$ and $0< z\le 1$, and
-\be
-M :=  \frac{K}{K+m} \one{\phi(N) = KN} + \one{\phi(N)/N \to \infty }
-\ee
-where $K > 0$ is a constant and  $m := \lim_{N\to \infty} \EE{X} = 1 + 2^{1-\alpha}/(\alpha - 1)$ \citep{CDEE2020,AEKKZ2020}.
-In other words,  the measure $(\Lambda)$ driving the multiple mergers is of the same form as in Eq.\ \eqref{Fbeta}  with $0 < x \le M$ in the case $1 < \alpha < 2$ and $\phi(N) \ge KN$.
-If $\alpha \ge 2$ or $\phi(N)/N \to 0$ then  the  ancestral process converges (in the sense of convergence of  finite-dimensional distributions) to  the Kingman-coalescent \citep{CDEE2020,AEKKZ2020}.
-
 \cite{becheler2020occupancy}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.
@@
 %%HybridLambda~\citep{zhu2015hybrid}
-
-\begin{comment}%
-A coalescent process is a continuous-time Markov process taking values among the
-partitions of $\IN := \{1,2, \ldots \}$, such that the restriction to
-any finite $n \in \IN $ takes values among partitions of
-$[n] := \{1, 2, \ldots, n\}$.  Write $\one{A} = 1$ if $A$ holds, and zero otherwise.   Let $\cP_n$ denote the set of
-partitions of $[n]$.  In the classical Kingman-coalescent, the only
-possible transitions are the mergers of pairs of blocks (elements of a
-partition $\pi \in \cP_n$), one pair at a time.  The $n$ leaves
-(corresponding to the sampled DNA sequences) are arbitrarily labelled
-from 1 to $n$, and  the blocks of a partition represent the common ancestors
-of the labels of each block.    The initial state is  (usually) taken as
-$\{ \{1\}, \ldots, \{n\}\}$, and the final state, i.e.\ when the most
-recent common ancester is reached, as $\{ [n]\}$. A block in a partition of
-$[n]$ represents an ancestor of the leaves in the block, i.e.\ the block
-$\{i_1, \ldots, i_k\}$ in a given partition of $[n]$ is an ancestor of the
-$k$ leaves $i_1, \ldots, i_k \in [n]$, and the leaves  correspond to
-arbitrarily labelled DNA sequences in the sample.
-\end{comment}
-
-Mention~\citep{becheler2020occupancy} somewhere.
 
@@
 
+\appendix
+\label{app-multiple-mergers}
+\section{Multiple merger coalescent model}
+
+Multiple-merger coalescents, in which a random number of ancestral
+lineages may merge at a given time in one group, and only one such
+group of lineages can merge at any given time (asynchronous multiple
+mergers), are referred to as $\Lambda$-coalescents; the rate at which  a given group of $k$ lineages out of a total of  $b$ lineages merges  is given by ($a\ge  0$ constant)
+\begin{equation}\label{lambdabk}
+\lambda_{b, k} =  \int_0^1  x^{k-2}(1-x)^{b-k}\Lambda(dx) + a\one{k=2}, \quad 2 \le k \le b,
+\end{equation}
+where $\Lambda$ is a finite measure on the  (Borel subsets of)  unit interval without an atom at zero \citep{DK99,P99,S99}.    The  total rate  is given by
+\be\label{lambdab}
+ \lambda_{b} = \int_0^1 \left(1 - (1-x)^{b} - bx(1-x)^{b-1} \right)x^{-2}\Lambda(dx) + \binom{b}{2},
+\ee
+\citep{S99}  which can be useful in applications, in particular  provided the  antiderivative  can be explicitly identified.
+
+A larger class of multiple-merger coalescents involving simultaneous
+multiple mergers of distinct groups of ancestral lineages also exists
+\citep{S00}. These are commonly referred to as $\Xi$-coalescents, and
+can be shown to be the limits of ancestral processes derived from
+population models incorporating diploidy (or more general polyploidy)
+\citep{BBE13,Blath2016}, and certain models of selection \citep{DS04}.
+To describe a general $\Xi$-coalescent let $\Delta$ denote the
+infinite simplex \be\label{Delta} \Delta := \{ (x_1, \ldots ): x_1 \ge
+x_2 \ge \cdots \ge 0, \sum_{j}x_j \le 1\}; \ee for any
+$r \in \{1,2, \ldots\}$ let $k_1, \cdots, k_r \ge 2$, and
+$b = s + k_1 + \cdots + k_r$ be the total number of blocks in a given
+partition ($s = b - k_1 - \cdots - k_r$ is the number of blocks not
+participating in the mergers at the given time).  The existence of
+simultaneous multiple-merger coalescents was proved by \cite{S00}.
+There exists a finite measure $\Xi$ on $\Delta$, with
+$\Xi = \Xi_0 + a\delta_0$, the measure $\Xi_0$ has no atom at zero,
+$a >0$ is fixed, and the rate at which one sees a given (simultaneous)
+merger of ancestral lineages with merger sizes $k_1, \ldots, k_r$,
+$s = n - k_1 - \cdots - k_r$, is given by
+\begin{esplit}{xi}
+  \lambda_{n; k_1, \ldots, k_r; s}  & = \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
+_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
+  & +  a\one{r=1, k_1 = 2}.
+\end{esplit}%
+The number of such $(k_1, \ldots, k_r)$ mergers is
+\be\label{N}
+    \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
+\ee
+\citep{S00},   in particular  $\mathcal{N}(b;2) = b(b-1)/2$, and one can compute the total rate of a  $(k_1, \ldots, k_r)$ merger as
+\be\label{lambdabkall}
+      \lambda(n; k_1, \ldots, k_r)         =    \mathcal{N}(b; k_1, \ldots, k_r ) \lambda_{n; k_1, \ldots, k_r; s}.
+\ee
+
+The total rate is given by, with
+$n\ge 2$ denoting the total  number of ancestral lineages,
+\be
+\label{Xilambdab}
+\lambda_{n} = \int_\Delta \left(1 - \sum_{\ell = 0}^n \sum_{i_1 \neq
+\cdots \neq i_\ell } \binom{n}{\ell} x_{i_1}\cdots x_{i_\ell}\left(1
+- \sum_{i=1}^\infty x_i\right)^{n-\ell} \right)
+\frac{1}{\sum_{j=1}^\infty x_j^2}\Xi_0(dx) + a\binom{n}{2} \ee
+\citep{S00}.  Viewing coalescent processes strictly as mathematical
+objects, it is clear that the class of $\Xi$-coalescents contains
+$\Lambda$-coalescents as a specific example (i.e.\ allowing at most
+one group of lineages to merge each time), and the class of
+$\Lambda$-coalescents contain the Kingman-coalescent as a special
+case.  However, viewed as limits of ancestral processes derived from
+specific population models they are not nested, since one would
+obtain $\Lambda$-coalescents when deriving coalescent processes from
+haploid population models incorporating sweepstakes reproduction and
+high fecundity, and $\Xi$-coalescents for diploid populations.  One
+should therefore apply the models as appropriate, i.e.\
+$\Lambda$-coalescents to data (e.g.\ mtDNA data) inherited in a
+haploid fashion, and $\Xi$-coalescents to e.g.\ autosomal data
+inherited in diploid or polyploid fashion \citep{Blath2016}.
+
+In \msprime\ we have incorporated two examples of multiple-merger coalescents.
+One is a diploid extension \citep{BBE13} of the haploid model of sweepstakes
+reproduction considered by \cite{EW06}, which is a haploid Moran model adapted
+to sweepstakes reproduction.  Let $N$ denote population size, and take $\psi
+\in (0,1]$ to be fixed.  In every generation, with probability
+$1-\varepsilon_N$ a single individual (picked uniformly at random) perishes,
+and one of the surviving individuals (sampled uniformly at random) produces one
+offspring; with probability $\varepsilon_N$ a total of $\lfloor \psi N \rfloor$
+individuals perish, and of the remaining individuals a single individual
+produces $\lfloor \psi N \rfloor -1 $ offspring.  Taking $\varepsilon_N =
+1/N^\gamma$ for some $\gamma > 0$, \cite{EW06} obtain specific examples of
+$\Lambda$-coalescents, where the $\Lambda$ measure in Eq.~\eqref{lambdabk} is a
+point mass at $\psi$.  The simplicity of this model does allow one to obtain
+some explicit mathematical results (see e.g.\
+\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model considered by
+\cite{EW06} has also been applied in algorithms for simulating gene genealogies
+within phylogenies \citep{zhu2015hybrid}. The specific model incorporated into
+\msprime\ is the diploid version \citep{BBE13} of the model studied by
+\cite{EW06}, which would be necessary in order to incorporate recombination.
+In the model considered by \cite{BBE13}, a single pair of diploid individuals
+contribute offspring in each generation, selfing is excluded, and each
+offspring is assigned one chromosome from each parent. There are, therefore,
+four parent chromosomes involved in each reproduction event, which can lead to
+up to four simultaneous mergers.  Let \begin{esplit}{Cconst} C_{b; k_1, \ldots,
+k_r;s } &= \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell}
+(4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots +
+k_r + \ell},  \\ \end{esplit} where  $C_{b; k_1, \ldots, k_r;s }$ corresponds
+to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq~\eqref{xi})
+of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$. The
+total merger rate (see Eq~\eqref{lambdabkall}) is then
+\be\label{xidirlambdabk} \lambda(b;k_1, \ldots, k_r) = \mathcal{N}(b; k_1,
+\ldots, k_r ) \left( \frac{c\psi^2/4}{ 1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s
+} + \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right), \ee and the total
+coalescence rate becomes \begin{esplit}{xilambdab} \lambda_b & =
+\frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0
+}^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 -
+\psi)^{b-\ell } \right) + \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.  \\
+\end{esplit} The interpretation of \eqref{xilambdab} is that with probability
+$1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent
+pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 +
+c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair
+produce  $\lfloor \psi N \rfloor$ offspring.
+
+The other multiple-merger coalescent model incorporated in \msprime\ is derived
+from an adaptation of the haploid population model considered by
+\cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the haploid
+version, in each generation individuals independently produce random numbers of
+juveniles, where the random  number of juveniles $(X)$ produced by a given
+individual  has the stable law \be\label{jX} \lim_{k\to \infty} C k^\alpha
+\prb{X \ge k} = 1 \ee with index $\alpha > 0$, and $C > 0$ is a normalising
+constant.   One can interpret  Eq.~\eqref{jX} as  stating the form of the
+probability distribution  for number of juveniles at least on the order of the
+population size.     If the random  total number of juveniles $(S_N)$ produced
+in this way is at least the population size $(N)$, then one samples  $N$
+juveniles uniformly at random without replacement to form the next generation
+of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the
+same  set of individuals.   However,   assuming  $\EE{X} > 1$, one can show
+that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.
+If $\alpha \ge 2$ the ancestral process converges to the Kingman-coalescent; if
+$1 \le \alpha < 2$ the ancestral process converges to a specific case of a
+$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.~\eqref{lambdabk} is
+associated with the Beta$(2-\alpha, \alpha)$ probability distribution, i.e.\
+\be\label{Fbeta} \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)}
+x^{1 - \alpha}(1-x)^{\alpha - 1}  dx, \ee where $B(2-\alpha,\alpha)$ is the
+beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$
+\citep{schweinsberg03}.
+This model has been adapted to diploid populations
+by \cite{BLS15}, where  the resulting coalescent process
+is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$,
+where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.
+The merger rate (see Eq.~\eqref{xi}) is then ($1 \le r \le 4$)
+\be\label{xibeta} \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b -
+k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}}
+\frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)} \ee
+\citep{Blath2016,BLS15}. The interpretation of Eq.~\eqref{xibeta} is  that the
+random   number of diploid  juveniles each  diploid pair of parents  produces
+is  governed by  the law in Eq.~\eqref{jX},   and   each diploid juvenile  is
+assigned  one chromosome  from each parent (selfing is excluded).
+
+The model in Eq~\eqref{jX} assumes that individuals can produce arbitrarily
+large numbers of juveniles. Considering diploid juveniles, this assumption is
+probably rather strong, since diploid juveniles are at least fertilised eggs,
+and so it is reasonable to suppose that the number of juveniles surviving to
+the  particular life stage we are modeling  cannot be arbitrarily large.  With
+this in mind, we also consider an adaptation of the Schweinsberg model, where
+the random number of juveniles $(X)$ produced by a   given parent pair  is
+distributed according to \be\label{jtr} \prb{X=k} =   \one{1 \le k \le \phi(N)}
+\frac{\phi(N+1)^\alpha }{ \phi(N+1)^\alpha - 1 }  \left( \frac{1}{k^\alpha} -
+\frac{1}{(k+1)^\alpha}  \right) , \ee where $\phi(N)$ is a deterministic strict
+upper bound on the number of juveniles produced by  any given parent pair (see
+also \citep{Eldon2018}).   One can follow the calculations in
+\citep{schweinsberg03} or \citep{BLS15}  to show  that  if $1 < \alpha < 2$
+then  $(k = k_1 + \cdots + k_r)$ then the merger rate (see Eq.~\eqref{xibeta})
+is
+ \be \lambda_{b;k_1, \ldots, k_r} =  \sum_{\ell = 0}^{ (b - k)\wedge (4-r) }
+\binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(M; k+\ell - \alpha,
+b-k-\ell + \alpha ) }{B(M;2-\alpha,\alpha)}
+\ee
+with $B(z;a,b) := \int_0^z
+t^{a-1}(1-t)^{b-1}dt$ for  $a,b>0$ and $0< z\le 1$, and \be M :=  \frac{K}{K+m}
+\one{\phi(N) = KN} + \one{\phi(N)/N \to \infty } \ee where $K > 0$ is a
+constant and  $m := \lim_{N\to \infty} \EE{X} = 1 + 2^{1-\alpha}/(\alpha - 1)$
+\citep{CDEE2020,AEKKZ2020}.    In other words,  the measure $(\Lambda)$ driving
+the multiple mergers is of the same form as in Eq.~\eqref{Fbeta}  with $0 < x
+\le M$ in the case $1 < \alpha < 2$ and $\phi(N) \ge KN$.   If $\alpha \ge 2$
+or $\phi(N)/N \to 0$ then  the  ancestral process converges (in the sense of
+convergence of  finite-dimensional distributions) to  the Kingman-coalescent
+\citep{CDEE2020,AEKKZ2020}.
+
+\begin{comment}%
+A coalescent process is a continuous-time Markov process taking values among
+the partitions of $\IN := \{1,2, \ldots \}$, such that the restriction to any
+finite $n \in \IN $ takes values among partitions of $[n] := \{1, 2, \ldots,
+n\}$.  Write $\one{A} = 1$ if $A$ holds, and zero otherwise.   Let $\cP_n$
+denote the set of partitions of $[n]$.  In the classical Kingman-coalescent,
+the only possible transitions are the mergers of pairs of blocks (elements of a
+partition $\pi \in \cP_n$), one pair at a time.  The $n$ leaves (corresponding
+to the sampled DNA sequences) are arbitrarily labelled from 1 to $n$, and  the
+blocks of a partition represent the common ancestors of the labels of each
+block.    The initial state is  (usually) taken as $\{ \{1\}, \ldots, \{n\}\}$,
+and the final state, i.e.\ when the most recent common ancester is reached, as
+$\{ [n]\}$. A block in a partition of $[n]$ represents an ancestor of the
+leaves in the block, i.e.\ the block $\{i_1, \ldots, i_k\}$ in a given
+partition of $[n]$ is an ancestor of the $k$ leaves $i_1, \ldots, i_k \in [n]$,
+and the leaves  correspond to arbitrarily labelled DNA sequences in the sample.
+\end{comment}
+
 \end{document}

### 2021-07-05 paper.tex
Update main section text.

@@
 
-Kingman's coalescent assumes that at most two ancestral lineages can
-merge at each merger event.  This follows from the fact that the
-Kingman coalescent can be shown to be the process describing the
-random ancestral relations of gene copies sampled from a population
-evolving according to the Wright-Fisher model (or similar models) of
-genetic evolution.  Multiple-merger coalescent models can also be
-obtained from appropriate extensions of the Moran
-model \citep{EW06,HM12}, i.e.\ where a single randomly picked individual
-(e.g.\ in a haploid population), or a pair of diploid parent
-individuals \citep{BBE13}, produces offspring in each reproduction
-event.  Convergence to the Kingman coalescent follows from certain
-assumptions on the offspring distribution \citep{S99,MS01,SW08}.  These
-assumptions may be violated in certain highly fecund
-organisms \citep{hedgecock_94,B94,HP11,A04,irwin16}.  We take `high
-fecundity' to mean the ability of individuals to produce numbers of
-offspring on the order of the population size.  A key issue regarding
-high fecundity involves the concept of `sweepstakes
-reproduction' \citep{hedgecock_94}, where a few individuals may produce
-the bulk of the offspring in any given generation.  Sweepstakes
-reproduction is not captured by the Wrigth-Fisher model (or similar
-models).  Population models, which do capture high fecundity and
-sweepstakes reproduction, are in the domain of attraction of
-`multiple-merger' coalescents \citep{DK99,P99,S99,S00}, in which a
-random number of ancestral lineages may merge at any given time.  The
-term `multiple merger' refers to a merger of at least three ancestral
-lineages, or the simultaneous \citep{S00,MS01} merger of at least four
-lineages in at least two distinct groups, and each group involving at
-least two lineages.  Multiple-merger coalescent processes have also
-been shown to be relevant for modeling the effects of selection on
+Kingman's coalescent assumes that at most two ancestral lineages can merge at
+each merger event. Although this is generally a reasonable approximation there
+are certain situations in which the underlying mathematical assumptions are
+violated. For example in certain highly fecund organisms
+\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
+to produce numbers of offspring on the order of the population size and
+therefore a few individuals may produce the bulk of the offspring in any given
+generation\citep{hedgecock_94}. These population dynamics violate basic
+assumptions of the Kingman coalescent, and are better modelled by
+`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
+two lineages can merge in a given event. Multiple-merger coalescent processes
+have also been shown to be relevant for modeling the effects of selection on
 gene genealogies \citep{Gillespie909,DS04}.
 
-\cite{becheler2020occupancy}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.
+Although multiple merger coalescents have been of significant theoretical
+interest for around two decades, there has been little practical software
+available to simulate these models.
+\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
+simulate a related spatial continuum model~\citep{barton2010new},
+and \cite{zhu2015hybrid} simulate genealogies within a species tree
+based on a multiple-merger model.
+More recently \cite{becheler2020occupancy}
+provide a general method for simulating multiple merger processes
+as part of the Quetzal framework~\citep{becheler2019quetzal}.
+None of these methods work with large genomes, and very little work
+has been performed on simulating multiple merger processes with recombination.
 
-%assumption can be
-%violated in some situations. Lambda coalescents
-%%~\citep{donnelly1999particle,schweinsberg2000coalescents}
-%%HybridLambda~\citep{zhu2015hybrid}
+We have added two multiple merger coalescent models in \msprime\ 1.0, the
+$\beta$-coalescent and ``dirac''-coalescent, allowing us to simulate
+such models with recombination efficiently for the first time.
+Please see the Appendix for full details and derivations.
 

### 2021-07-07 paper.tex
Response to review.

@@
 
-Kingman's coalescent assumes that at most two ancestral lineages can merge at
+Kingman's coalescent assumes that only two ancestral lineages can merge at
 each merger event. Although this is generally a reasonable approximation there
@@
 simulate a related spatial continuum model~\citep{barton2010new},
-and \cite{zhu2015hybrid} simulate genealogies within a species tree
-based on a multiple-merger model.
-More recently \cite{becheler2020occupancy}
-provide a general method for simulating multiple merger processes
+\cite{zhu2015hybrid} simulate genealogies within a species tree
+based on a multiple-merger model, and
+\cite{becheler2020occupancy} provide a general method for simulating
+multiple merger processes
 as part of the Quetzal framework~\citep{becheler2019quetzal}.
+The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
+includes a number of extensions to the Beta-coalescent, although
+it only supports unlinked loci.
 None of these methods work with large genomes, and very little work
@@
 We have added two multiple merger coalescent models in \msprime\ 1.0, the
-$\beta$-coalescent and ``dirac''-coalescent, allowing us to simulate
+Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
 such models with recombination efficiently for the first time.
-Please see the Appendix for full details and derivations.
+These simulation models have been extensively validated against
+analytical results from the site frequency
+spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
+ as well as more general properties of coalescent processes.
+Please see the Appendix for more details and model derivations.
 
@@
 population models incorporating diploidy (or more general polyploidy)
-\citep{BBE13,Blath2016}, and certain models of selection \citep{DS04}.
+\citep{BBE13,blath2016site}, and certain models of selection \citep{DS04}.
 To describe a general $\Xi$-coalescent let $\Delta$ denote the
@@
 haploid fashion, and $\Xi$-coalescents to e.g.\ autosomal data
-inherited in diploid or polyploid fashion \citep{Blath2016}.
+inherited in diploid or polyploid fashion \citep{blath2016site}.
 
@@
 \frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)} \ee
-\citep{Blath2016,BLS15}. The interpretation of Eq.~\eqref{xibeta} is  that the
+\citep{blath2016site,BLS15}. The interpretation of Eq.~\eqref{xibeta} is  that the
 random   number of diploid  juveniles each  diploid pair of parents  produces

### 2021-07-07 paper.tex

@@
 The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
-includes a number of extensions to the Beta-coalescent, although
-it only supports unlinked loci.
+includes a number of extensions to the Beta-coalescent.
 None of these methods work with large genomes, and very little work

### 2021-07-08 paper.tex
Update bottleneck section.

@@
 F84~\citep{felsenstein1996hidden},
-and
-GTR~\citep{tavare1986some} nucleotide models
-and
-the BLOSUM62~\citep{henikoff1992amino}
+and GTR~\citep{tavare1986some} nucleotide models
+and the BLOSUM62~\citep{henikoff1992amino}
 and PAM~\citep{dayhoff1978} amino acid models.
@@
 Seq-Gen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}, and are run before every
-release.
+Pyvolve~\citep{spielman2015pyvolve}.
 
@@
 therefore a few individuals may produce the bulk of the offspring in any given
-generation\citep{hedgecock_94}. These population dynamics violate basic
+generation~\citep{hedgecock_94}. These population dynamics violate basic
 assumptions of the Kingman coalescent, and are better modelled by
@@
 \subsection*{Instantaneous bottlenecks}
-A common approach to modelling the effect of demographic history on genealogies is to assume that effective population size ($N_e$) changes in discrete steps which define a series of epochs~\citep{griffiths1994sampling, marth2004allele, keightley2007joint, li2011inference}.
-In this setting of piece-wise constant $N_e$, capturing a population bottleneck requires three epochs: $N_e$ is reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and recovers to its initial value at time $T_{end}$~\citep{marth2004allele}.
-If bottlenecks are short both on the timescale of coalescence and mutations, one may expect little information about the duration of a bottleneck ($T_{end}-T_{start}$) in sequence data.
-Thus a simpler, alternative model is to assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$) and generate a sudden burst of coalescence events (a multiple merger event) in the genealogy.
-The strength of the bottleneck B can be thought of as an (imaginary) time period during which coalescence events are collapsed, i.e.
-there is no growth in genealogical branches during B and the probability that a single pair of lineages entering the bottleneck coalesce during the bottleneck is $1-e^{-B}$.
-Although this simple two parameter model of bottlenecks is attractive and both analytic results and empirical inference~\citep{griffiths1994sampling, galtier2000detecting, bunnefeld2015inferring} have been developed under this model, it has not been straightforward to simulate data under instantaneous bottleneck histories with \ms.
-Instantaneous bottlenecks are implemented as a demographic event in \msprime.
+
+A common approach to modelling the effect of demographic history on genealogies
+is to assume that effective population size ($N_e$) changes in discrete steps
+which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
+keightley2007joint,li2011inference}. In this setting of piecewise constant
+$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
+reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
+recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
+bottlenecks are short both on the timescale of coalescence and mutations,
+there may be little information about the duration of a bottleneck
+($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
+assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
+and generate a sudden burst of coalescence events (a multiple merger event) in
+the genealogy. The strength of the bottleneck $B$ can be thought of as an
+(imaginary) time period during which coalescence events are collapsed,
+i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
+single pair of lineages entering the bottleneck coalesce during the bottleneck
+is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
+attractive and both analytic results and empirical
+inference~\citep{griffiths1994sampling, galtier2000detecting,
+bunnefeld2015inferring} have been developed under this model, there has
+been no software available to simulate data under instantaneous
+bottleneck histories.
+
+We have implemented instantaneous bottlenecks in \msprime~1.0
+using a variant of Hudson's linear time single-locus coalescent
+algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
+by adding events to the Demography object (see the Demography section)
+and can be used in combination with any other demographic modelling
+features. We have validated the results of these simulations by comparing
+against analytical expectations for coalescence times and the
+site frequency spectrum~\citep{bunnefeld2015inferring}.
 

### 2021-07-08 paper.tex
Updated discussion of recombination.

@@
 \usepackage{geometry}%
+\usepackage{color}
 \usepackage{dsfont}
@@
 \newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
+
+\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
 
@@
 The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
-has profoundly changed
-what is possible. Through the use of efficient data structures, it is
-now possible to simulate millions of whole chromosomes, without resorting
-to approximations. The ``succinct tree
-sequence'' data
+has greatly increased the scope of coalescent simulations.
+Through the use of efficient data structures, it is
+now straightforward to simulate millions of whole chromosomes.
+The ``succinct tree sequence'' data
 structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
@@
 of magnitude smaller than the standard
-Newick~\citep{felsenstein1989phylip} based formats. This data structure has
+Newick~\citep{felsenstein1989phylip} or
+VCF~\citep{danecek2011variant} based formats formats.
+This data structure has
 also lead to major advances in forwards-time
@@
 
-\subsection*{Recombination and gene conversion}
-Recombination is implemented in \msprime\ via
-Hudson's algorithm, which
+\subsection*{Recombination}
+
+Recombination is implemented in \msprime\ using Hudson's algorithm, which
 works backwards in time tracking the
-through the effects of common ancestor and recombination events
-~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}. Common
-ancestor events combine the ancestral material of two lineages, and may
+effects of common ancestor and recombination
+events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
+Common ancestor events combine the ancestral material of two lineages, and may
 result in coalescences in the marginal trees. Recombination events
@@
 or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
+In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
+us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
+and empirical recombination maps similar to other
+simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
 
-We also support variable recombination rates along a chromosome.
-This allows us to model both recombination hotspots like
-\texttt{msHOT}~\citep{hellenthal2007mshot} and also empirical
-recombination maps like those supported \texttt{cosi2}
-\citep{shlyakhter2014cosi2}
-(What about ~\cite{wang2014new}?)
-We have also implemented gene
-conversion following the model defined by~\cite{wiuf2000coalescent}
-and used in \ms.
-% % Uncomment this sentence, when the corresponding issue in the msprime github repo is resolved.
-% Furthermore, msprime is now able to simulate gene conversion events that are
-% initiated proportional to the local recombination rate specified by the recombination map.
-
-Gene conversion is particularly useful to model homologous recombination in bacterial evolution.
-We compare the performance of \msprime\ with gene conversion to \ms\ and
-the whole bacterial genome simulator SimBac~\citep{brown2016simbac}.
-Further features are planned to model bacterial evolution. In particular, we
-aim to include bacterial gene transfer and utilize the succinct tree sequence structure
-to represent the resulting ancestral gene transfer graph~\citep{baumdicker2014AGTG}.
-
-\subsection*{SMC approximations}
-
-\textbf{TODO:} It's important that we acknowledge that (a) the coalescent
-with recombination does become quite slow for large $N_e$; (b) our
-implementation of the SMC is naive; and (c) for sufficiently large
-$N_e$ programs like \texttt{scrm} will be faster.
-
-The SMC~\citep{mcvean2005approximating} and SMC'~\citep{marjoram2006fast}
-models are supported.
-The performance of msprime has made the SMC
-approximation largely unnecessary, at least when simulating human-like
-recombination rates. The SMC is an important analytical approximation
-forming the basis for several important inference methods
+The Sequentially Markovian Coalescent (SMC) is an approximation of the
+coalescent with recombination
+model~\citep{mcvean2005approximating,marjoram2006fast},
+and was originally motivated to a significant degree by the need to
+simulate longer genomes than was possible using tools like \ms.
+The SMC is a good approximation to the
+coalescent with recombination when we have four or fewer sampled
+genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
+effects of the approximation are less well understood for larger
+sample sizes and several approaches have been proposed to
+allow simulations more closely approximate the coalescent
+with recombination~\citep{chen2009fast,wang2014new,staab2014scrm}.
+The SMC is an important analytical approximation
+and has allowed the development of numerous inference
 methods~\citep{
@@
 zheng2014bayesian,
-terhorst2017robust},
-so simulations are still useful to examine  how good the approximation
-is.
+terhorst2017robust}.
 
-The SMC has been shown to be a good approximation to the
-coalescent with recombination for samples of less than
-four~\citep{hobolth2014markovian,wilton2015smc}
-and in simulations, produces patterns of LD indistinguishable from the
-full coalescent~\citep{mcvean2005approximating,marjoram2006fast}. However,
-LD is a crude approximation of the correlation structure of trees
-and it is not clear how good the SMC is for larger sample sizes.
+\begin{figure}
+\caption{\label{fig-recombination-perf}Approximate run time for \msprime\
+simulations as a function of scaled recombination rate.}
+\end{figure}
+
+In human-like parameter regimes, \msprime's implementation of the
+exact coalescent with recombination comprehensively outperforms
+all other simulators, including those based on SMC
+approximations~\citep{kelleher2016efficient}. However, it is important
+to note that although the implementation of Hudson's algorithm is
+very efficient, it is still quadratic in the scaled recombination rate
+$\rho$, and therefore for sufficiently large $N_e$, recombination rate
+or long genome, efficient SMC implementations will be faster
+that \msprime.
+
+\jkcomment{We want to round out this discussion here by giving some
+sort of rough argument for why Hudson's algorithm is quadratic, and discussing
+to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
+to simulate in how long. See issue 28.}
+
+The SMC~\citep{mcvean2005approximating} and
+SMC'~\citep{marjoram2006fast} models are available
+in \msprime\ 1.0. However, they are currently implemented using a
+naive rejection sampling approach, and are somewhat slower
+to simulate than the exact coalescent with recombination. These
+models are therefore only appropriate for studying the SMC approximations
+themselves; an important avenue for future work is to implement
+efficient SMC-like models, perhaps based on the approaches used
+in \texttt{cosi2}~\citep{shlyakhter2014cosi2}.
+
+\subsection*{Gene conversion}
+
+We have also implemented gene
+conversion following the model defined by~\cite{wiuf2000coalescent}
+and used in \ms.
+
+Gene conversion is particularly useful to model homologous recombination in
+bacterial evolution. We compare the performance of \msprime\ with gene
+conversion to \ms\ and the whole bacterial genome simulator
+SimBac~\citep{brown2016simbac}. Further features are planned to model bacterial
+evolution. In particular, we aim to include bacterial gene transfer and utilize
+the succinct tree sequence structure to represent the resulting ancestral gene
+transfer graph~\citep{baumdicker2014AGTG}.
 

### 2021-07-08 paper.tex

@@
 simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
+The implementation of recombination in \msprime\ is extensively validated
+against analytical results~\citep{hudson1983properties,kaplan1985use}
+and results from \ms.
 
@@
 efficient SMC-like models, perhaps based on the approaches used
-in \texttt{cosi2}~\citep{shlyakhter2014cosi2}.
+in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
+simulation of organisms such as \textit{Drosophila melagaster}.
 
@@
 
-We have also implemented gene
-conversion following the model defined by~\cite{wiuf2000coalescent}
-and used in \ms.
+Gene conversion is \jkcomment{A couple of sentences describing gene
+conversion, and what the model is. What other simulators simulate
+gc? }
+We have implemented this model of gene
+conversion in \msprime\ 1.0, and validated the output against
+\ms\ and analytical results~\citep{wiuf2000coalescent}.
+
+\begin{figure}
+\caption{\label{fig-gc-perf}Comparison of the running time
+for E-coli simulations using \msprime\ and SimBac for varying
+sample sizes. The parameters used are [] based on [references].}
+\end{figure}
 
 Gene conversion is particularly useful to model homologous recombination in
-bacterial evolution. We compare the performance of \msprime\ with gene
-conversion to \ms\ and the whole bacterial genome simulator
-SimBac~\citep{brown2016simbac}. Further features are planned to model bacterial
+bacterial evolution.
+We compare the performance of \msprime\ with gene conversion to
+SimBac~\citep{brown2016simbac}, a simulator developed specifically to
+simulate bacterial evolution. Although \msprime\ is missing some features
+relevant to this case, it is far more efficient at simulating the basic
+processes. An important avenue for future work
+will be to add further features to model bacterial
 evolution. In particular, we aim to include bacterial gene transfer and utilize
@@
 \section*{Results and Discussion}
-\subsection*{Performance}
-We compare \msprime\ with some existing simulators that have a subset of its
-features.
-
-\subsection*{Examples}
-We show a few examples of running \msprime.
 
@@
 way is vital with today's vast numbers of whole genome samples.
+
+A significant advantage to simulating the coalescent with recombination
+backwards in time using Hudson's algorithm rather than
+left-to-right
+approaches~\citep{wiuf1999recombination,chen2009fast,staab2014scrm}
+is the ease with which additional processes can be integrated.
+We have seen that processes such as gene conversion, complex
+demography, instantaneous bottlenecks, multiple merger caolescents
+and selective sweeps have been integrated seamlessly into
+the simulations. Incorporating any of these extensions into
+either the exact or approximate left-to-right processes would require
+complex derivations. Because we are going backwards
+in time, processes can be modelled as competing exponentials
+using Gillespie's algorithm \jkcomment{I think?},
+and therefore in principle we can
+add as many extra processes as we like.
+\jkcomment{so, backwards in time
+sims in msprime is extensible for the future.}
 

### 2021-07-08 paper.tex
Partial draft of data interchange section

@@
 
-\subsection*{Integration with forwards-time simulations}
-Forward simulations can also use tree
-sequences~\citep{kelleher2018efficient,haller2018tree}. In particular,
-SLiM 3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output tree sequences.
-Msprime can take tree sequences simulated by these forward simulators
-as input and complete the simulation (``recapitation'').
-
 \subsection*{Finite sites mutations}
@@
 
-\subsection*{Data model}
-The output of \msprime\ is a ``succinct tree sequence`` (or tree sequence,
-for short) using the \tskit\ library. [Quick overview of the output formats
-of other simulators, and the perf advantages of tree sequences.]
+\subsection*{Data interchange and interoperability}
+The point of simulations is to generate data and the efficiency with
+which simulated data can be incorporated into an overall application
+is an important---if often overlooked---issue.
+For example, Approximate Bayesian Computation
+(ABC)~\citep{beaumont2002approximate,csillery2010approximate}
+and more recent machine learning
+methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
+perform population genetic inference by using large numbers of
+simulations as training data. Clearly, simulation efficiency is
+crucial since the size and number of simulations that can be performed
+largely determines the scope of what can be inferred. Equally important,
+however, is the efficiency with which the simulation results can be
+transformed into the specific input required by the inference method.
+In the case of ABC, this is usually a set of summary statistics of the
+sequence data. General approaches such as
+ABCtoolbox~\citep{wegmann2010abctoolbox} connect to
+simulation and summary statistic computation programs via text streams
+and are quite flexible, but the overhead involved in processing
+the text is often a computational bottleneck.
+Other methods such as DIYABC~\citep{cornuet2008inferring}
+or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
+developing their own specialised simulators, tightly integrating them
+with the inference method. While this is certainly more efficient,
+there is a great deal of duplication of effort and significant
+room for error.
 
-[Review the data model in Phastsim~\cite{demaio2021phastsim} and
-Usher~\citep{turakhia2021ultrafast}.]
+\jkcomment{Stuff to cover}
+\begin{itemize}
+
+\item The core problem is that the field lacks an efficient mechanism for
+    interchanging simulated data, and so developers are forced to choose
+    between efficiency and duplication of effort.
+
+\item What are the output formats of simulators? Newick, or sequence data
+    basically. What's the problem with this? We don't have mutations.
+    Much more efficient to have the tree and mutations than the
+    sequence data. Fundamentally, the output of coalescent simulations
+    should be trees, otherwise we are throwing a lot of information
+    away.
+
+\item Side note somewhere: with efficient access to branch length
+    statisics~\citep{ralph2019efficiently}, maybe we don't need to
+    simulate mutations at all.
+
+\item Msprime uses the \tskit\ library~\citep{tskit2021tskit}, which provides
+    extremely efficient interchange and data processing. A quick example
+    of how efficient.
+
+\item Tskit has facilited unprecedented interoperability between simulators:
+    Forward simulations can also use tree
+    sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
+    3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
+    tree sequences. Msprime can take tree sequences simulated by these forward
+    simulators as input and complete the simulation (``recapitation'').
+    See \cite{haller2018tree} for more details.
+
+\item Recently, Phastsim/USHER~\cite{demaio2021phastsim,turakhia2021ultrafast}
+    have used this idea of storing the
+    tree plus mutations, but approach is crude compared to tskit.
+
+\end{itemize}
 

### 2021-07-09 paper.tex
Changes to data interchange section. Added some ABC references.

@@
 simulations as training data. Clearly, simulation efficiency is
-crucial since the size and number of simulations that can be performed
-largely determines the scope of what can be inferred. Equally important,
+crucial since the size and number of simulations that can be performed determines
+the depth to which the model and parameter space can be sampled from.
+Equally important,
 however, is the efficiency with which the simulation results can be
 transformed into the specific input required by the inference method.
-In the case of ABC, this is usually a set of summary statistics of the
-sequence data. General approaches such as
+In the case of ABC, this is usually a set of summary statistics of the sequence data.
+Modern approaches such as
+ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
+ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large numbers of statistics
+that are individually weakly informative, but jointly close to sufficient for the 
+desired model or parameter.
+However, the computational resources needed to calculate these statistics can be even 
+greater than those used to generate the simulated datasets.
+[An example would be good here? Hmmm...]
+
+General approaches such as
 ABCtoolbox~\citep{wegmann2010abctoolbox} connect to
@@
 \item Side note somewhere: with efficient access to branch length
-    statisics~\citep{ralph2019efficiently}, maybe we don't need to
+    statistics~\citep{ralph2019efficiently}, maybe we don't need to
     simulate mutations at all.
+[GT: I'm not so sure about this -- the desired statistics/summaries 
+must be calculated on the observed dataset as well as on the many
+simulated datasets.
+Unless the observed dataset is stored as a tree sequence,
+(because it has been through tsinfer, for example),
+it's not clear how branch or node stats could be calculated 
+from it.
+]
 

### 2021-07-09 paper.tex
Removed abctoolbox sentence, added a sentence about the all-in-one ABC softwares

@@
 For example, Approximate Bayesian Computation
-(ABC)~\citep{beaumont2002approximate,csillery2010approximate}
+(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
 and more recent machine learning
@@
 ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
-ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large numbers of statistics
-that are individually weakly informative, but jointly close to sufficient for the 
-desired model or parameter.
+ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large numbers of user-specified statistics that are individually weakly informative,
+but jointly close to sufficient for the desired model or parameter.
 However, the computational resources needed to calculate these statistics can be even 
 greater than those used to generate the simulated datasets.
-[An example would be good here? Hmmm...]
-
-General approaches such as
-ABCtoolbox~\citep{wegmann2010abctoolbox} connect to
-simulation and summary statistic computation programs via text streams
-and are quite flexible, but the overhead involved in processing
-the text is often a computational bottleneck.
-Other methods such as DIYABC~\citep{cornuet2008inferring}
+Older software such as DIYABC~\citep{cornuet2008inferring}
 or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
@@
 room for error.
+As ABC inference has become useful in an ever-widening range of applications,
+these all-in-one methods risk falling behind if they do not allow users to explore
+the models and statistics of most interest.
 

### 2021-07-09 paper.tex

@@
 
-\subsection*{Population structure}
-We support all the stuff that ms does.
-Demographic events where we change
-the properties of the population.
+\subsection*{Demography}
+% TODO should we put in some citations here for the models? Hardly seems
+% necessary.
+One of the key applications of population genetic simulations is to generate
+data for complex demographies. Beyond idealised cases such as stepping-stone or
+island models, or specialised cases such as isolation-with-migration models,
+analytical results are rarely possible. Simulation is therefore integral to the
+development and evaluation of methods for demographic inference. The demography
+model in \msprime\ is directly derived from the approach used in \ms, and
+supports an arbitrary number of randomly mating populations exchanging migrants
+at specified rates. A range of demographic events are provided, which allow for
+varying population parameters and migration rates over time, as well as
+population splits, admixtures and pulse migrations. The location of lineages
+over time can be tracked in as much detail as required: each tree node is
+automatically associated with the population in which it arose, the location of
+lineages can be recorded at any given time via census events, or every lineage
+migration can be recorded.
 
-We provide a debugger interface which allows the user to see the
-various population sizes and growth rates over time. This includes
-a way to compute coalescence rate trajectory, which provides a useful
-ground truth when comparing with methods that estimate population
-size over time~\citep{adrion2019community}.
+A major change for \msprime\ 1.0 is the introduction of the new Demography API,
+designed to address a serious flaw in the \msprime\ 0.x interface which lead to
+a number of avoidable errors in downstream
+simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
+separate parameters be provided to the \texttt{simulate} function to describe a
+demographic model, making it easy to accidentally omit information. The 1.0 API
+resolves this issue by creating a new \texttt{Demography} class, which
+encapsulates all information about the demographic model, and fully decouples
+the definition from other simulation details. An instance of this class is then
+provided as a parameter to the new \texttt{sim\_ancestry} function,
+substantially reducing the potential for error. Another improvement over the
+0.X APIs is the introduction of explicit population split and admixture events,
+and a population state machine that ensures that lineages cannot migrate into
+(or be sampled from) inactive populations. This demography model is compatible
+with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
+class supports importing Demes models. We also support importing demographic
+models in the form of Newick species tree descriptions, as well as those
+produced by programs such as *BEAST~\citep{heled2009bayesian}.
 
-The interface
-is very flexible and general. We also reading in population tree
-descriptions as produced by programs such as
-StarBeast~\citep{heled2009bayesian}.
-
-We support tracking lineages through time by associating a population
-with each node, and having migration records. We also have
-census events, which will record the location of every lineage
-at a particular time.
+Many popular approaches in population genetics use the distribution of
+coalescence rates between pairs of lineages in one or more populations to infer
+effective population sizes over
+time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
+split times and subsequent migration rates between
+populations~\citep{wang2020tracking}. The \texttt{DemographyDebugger} provides
+methods to compute a suite of numerical predictions about the location of
+lineages over time, for example the coalescence rates for two or more lineages
+drawn from populations at specified times in the past. This provides a valuable
+ground-truth when evaluating such inference methods, as illustrated by
+\cite{adrion2020community}.
 
@@
 that we can grow the interface over time. Another is that
-downstream applications such as \texttt{stdpopsim}~\citep{adrion2019community}
+downstream applications such as \texttt{stdpopsim}~\citep{adrion2020community}
 can be developed.
+
+The \texttt{msp simulate} command also supports demographic models
+specified in Demes format~\citep{gower2021demes}, allowing users with
+no Python experience to run complex demographic simulations.
 
@@
 results or existing simulators.
-
-\subsection*{Computing coalescence rates and lineage location probabilities}
-Many popular approaches in population genetics use the distribution of coalescence rates between
-pairs of lineages in one or more populations to infer effective population sizes over time
-\cite{li2011inference,sheehan2013estimating,schiffels2014inferring}
-or split times
-and subsequent migration rates between populations
-\cite{wang2020tracking}.
-Given a demographic model, the
-\texttt{msprime} DemographyDebugger is able to compute coalescence rates for two or more lineages drawn
-from one or more populations at specified times in the past, providing a ground truth for coalescent rate-based
-inference methods. To compute coalescence rates for a set of times $Ts=(t_0, t_1, \ldots)$ in the
-past, we write
-\begin{lstlisting}[frame=single]
-dbg = demography.debug()
-rates = dbg.coalescence_rate_trajectory(Ts, num_samples)
-\end{lstlisting}
-where \texttt{num\_samples} is a list with length equal to the number of populations that
-specifies the number of sampled lineages from each population. For example,
-to get pairwise rates within the same
-population, this list would contain zeros for all populations except for the population index of interest,
-which would have a $2$. To get the rate of coalescence for two lineages sampled from different
-populations, we would set \texttt{num\_samples} to have ones in those population indices.
-
-More simply, \texttt{msprime} can also compute the probability of a lineage that was in a given
-population at some time to be found in other populations (through continuous or mass migration events).
-For a given sample time $t\geq0$ and any list of times $Ts$ in the past, we write
-\begin{lstlisting}[frame=single]
-prob = dbg.lineage_probabilities(Ts, sample_time=t)
-\end{lstlisting}
-This function is also used to determine the time intervals that lineages can be found in each population,
-given a list of contemporary and/or ancient samples. For a given sampling configuration, by writing
-\begin{lstlisting}[frame=single]
-lineage_dict = dbg.possible_lineage_locations(samples=samples)
-\end{lstlisting}
-we get a dictionary with epoch intervals as keys whose values are a list that indicates True or False for
-whether each population could contain ancestral material to our samples.
-If no sampling configuration is provided, we assume we sample lineages from each population at time zero.
 
@@
 opens the possibility for new applications of coalescent simulations.
-Also \stdpopsim\ is cool~\citep{adrion2019community}
+Also \stdpopsim\ is cool~\citep{adrion2020community}
 

### 2021-07-10 paper.tex
Added blurb about trapped recombinations.

@@
 very efficient, it is still quadratic in the scaled recombination rate
-$\rho$, and therefore for sufficiently large $N_e$, recombination rate
-or long genome, efficient SMC implementations will be faster
-that \msprime.
+$\rho. This is because Hudson's algorithms tracks recombinations not only 
+in segments ancestral to the sample, but also between ancestral segments.     
+Estimates of the number of such "trapped" recombinations grow quadratically 
+with genome length (\cite{Hein, Schierup, Wiuf}, Eq. 5.10). 
+Because they do not track such recombinations, efficient SMC implementations 
+will be faster that \msprime for sufficiently high recombination rate, 
+genome length, or $N_e$. %Until we decide to approximate the recombination process
+%between distant ancestral segment. Hint hint.
 

### 2021-07-11 paper.tex
Minor tweaks.

@@
 very efficient, it is still quadratic in the scaled recombination rate
-$\rho. This is because Hudson's algorithms tracks recombinations not only 
+$\rho$. This is because Hudson's algorithms tracks recombinations not only
 in segments ancestral to the sample, but also between ancestral segments.
-Estimates of the number of such "trapped" recombinations grow quadratically 
-with genome length (\cite{Hein, Schierup, Wiuf}, Eq. 5.10). 
+Estimates of the number of such ``trapped'' recombinations grow quadratically
+with genome length \citep[Eq.~5.10]{hein2004gene}.
 Because they do not track such recombinations, efficient SMC implementations
@@
 See the Data Model section for more details and discussion.
+
+Things we can do: future work will need to do things like
+Dawg~\citep{cartwright2005dna} and to implement models and
+indels~\citep{fletcher2009indelible}.
 
@@
 functional and highly efficient simulator as a C source code listing in an
-appendix to his seminal review paper~\citep{hudson1990gene}. Because of
+appendix to the seminal review paper~\citep{hudson1990gene}. Because of
 this ease, the dominant approach to producing simulators for extensions
@@
 possible to implement the model efficiently, and only through the use
-of sophisticated data structures and an intricate and subtle
-algorithm~\citep{kelleher2016efficient}.
-This efficiency, and the ability to handle recombination in a principled
-way is vital with today's vast numbers of whole genome samples.
+of sophisticated data structures and
+intricate algorithms~\citep{kelleher2016efficient}.
+
+% Making lots of simulators has given us a fragmented and poor
+% quality ecosystem.
+The approach of developing ``in-house'' simulators
+has lead to a huge proliferation of different programs, with
+slightly different feature sets and often incompatible interfaces.
+It is a fragmented ecosystem with significant issues with software
+quality (see, e.g.,~\cite{yang2014critical}),
+and few simulators are actively maintained after publication.
+We have seen that \msprime\ (for larger samples at least)
+has at worst comparable, but usually far superior
+performance to more specialised methods.
+
+% JK: I want to make the point that the established idea that
+% developing your own simulation software is a good learning experience
+% is pretty questionable these days, and tends to produce a lot of
+% crappy unmaintained software. Not sure if the following paragraph
+% does the job that well, though.
+
+There seems to be little point in continuing the huge duplication of effort we
+have seen. It is not possible to achieve the performance levels of \msprime\
+without investing substantially in software engineering. As demonstrated here,
+\msprime\ is extensible and we would argue that developing this simulator as an
+open-source community asset is of more benefit to all. Implementing a
+coalescent simulator may once have been a good way for (e.g.) a student to
+learn about programming and simulation, but it is no longer reasonable to
+expect this. There is simply too much engineering required. For example, as of
+the 1.0.0 release \msprime\ consisted of around 13K lines of C and 11K
+lines of Python, with test suites of 122 C test cases (7K lines of code) and
+1350 Python test cases (22K lines of code), along with a suite of 217
+statistical validation tests (6K lines of code). It requires a sustained effort
+over years to produce well engineered, high-quality software. It may reasonably
+be asked if students would learn more from contributing to an existing project
+with established high-quality engineering standards, or from producing their own
+software in a self-taught manner.
 
@@
 
-% Making lots of simulators has given us a fragmented and poort
-% quality ecosystem.
-The approach of developing ``in-house'' simulators
-has lead to a huge proliferation of different programs, with
-slightly different feature sets and often incompatible interfaces.
-It is a fragmented ecosystem with significant issues with software
-quality (see, e.g.,~\cite{yang2014critical});
-few simulators are actively maintained after publication.
-There seems to be little point
-in continuing the huge duplication of effort we have seen.
-It is not possible to achieve the performance levels of
-\msprime\ without investing substantially in software engineering.
-As demonstrated here, \msprime\ is extensible and we would argue
-that developing this simulator as an open-source community asset
-is of more benefit to all.
-
 There is also the welcome development of libraries such as
@@
 to be done.
-
-% We need to acknowledge that for long genomes and for large population
-% sizes msprime's performance does drop off and SMC simulators like
-% scrm become more efficient.
-Msprime's performance for large sample sizes is excellent, but
-simulating species with large population sizes is still challenging.
-An important avenue for future development is to improve performance
-in these scenarios by incorporating data structures such as skip lists
-and implementing more efficient coalescent approximations, as done
-in \texttt{cosi2}~\citep{shlyakhter2014cosi2}.
-
-In terms of mutation simulation, future work will need to do things like
-Dawg~\citep{cartwright2005dna} and to implement
-codon models and indels~\citep{fletcher2009indelible}.
 

### 2021-07-12 paper.tex
Move arg section next to recombination.
Add ARG comparison plot.

@@
 \usepackage{color}
+\usepackage{graphicx}
 \usepackage{dsfont}
@@
 \newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
-
-%% local definitions for section multiple merger coalescents
- \newcommand{\be}{\begin{equation}}
- \newcommand{\ee}{\end{equation}}
- \newcommand{\bd}{\begin{displaymath}}
- \newcommand{\ed}{\end{displaymath}}
-\newcommand{\IN}{\ensuremath{\mathds{N}}}%
-\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
-\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
-\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%
-
-\NewEnviron{esplit}[1]{%
-\begin{equation}
-\label{#1}
-\begin{split}
-  \BODY
-\end{split}\end{equation}
-}
 
@@
 \begin{figure}
-\caption{\label{fig-recombination-perf}Approximate run time for \msprime\
-simulations as a function of scaled recombination rate.}
+\includegraphics[width=\textwidth]{figures/recombination}
+\caption{\label{fig-recombination-perf}(A) Approximate run time for \msprime\
+simulations as a function of scaled recombination rate.
+(B) Comparison of the ARG simulations vs minimal tree sequence representation.}
 \end{figure}
@@
 simulation of organisms such as \textit{Drosophila melagaster}.
-
-\subsection*{Gene conversion}
-
-Gene conversion is \jkcomment{A couple of sentences describing gene
-conversion, and what the model is. What other simulators simulate
-gc? }
-We have implemented this model of gene
-conversion in \msprime\ 1.0, and validated the output against
-\ms\ and analytical results~\citep{wiuf2000coalescent}.
-
-\begin{figure}
-\caption{\label{fig-gc-perf}Comparison of the running time
-for E-coli simulations using \msprime\ and SimBac for varying
-sample sizes. The parameters used are [] based on [references].}
-\end{figure}
-
-Gene conversion is particularly useful to model homologous recombination in
-bacterial evolution.
-We compare the performance of \msprime\ with gene conversion to
-SimBac~\citep{brown2016simbac}, a simulator developed specifically to
-simulate bacterial evolution. Although \msprime\ is missing some features
-relevant to this case, it is far more efficient at simulating the basic
-processes. An important avenue for future work
-will be to add further features to model bacterial
-evolution. In particular, we aim to include bacterial gene transfer and utilize
-the succinct tree sequence structure to represent the resulting ancestral gene
-transfer graph~\citep{baumdicker2014AGTG}.
-
-\subsection*{Demography}
-% TODO should we put in some citations here for the models? Hardly seems
-% necessary.
-One of the key applications of population genetic simulations is to generate
-data for complex demographies. Beyond idealised cases such as stepping-stone or
-island models, or specialised cases such as isolation-with-migration models,
-analytical results are rarely possible. Simulation is therefore integral to the
-development and evaluation of methods for demographic inference. The demography
-model in \msprime\ is directly derived from the approach used in \ms, and
-supports an arbitrary number of randomly mating populations exchanging migrants
-at specified rates. A range of demographic events are provided, which allow for
-varying population parameters and migration rates over time, as well as
-population splits, admixtures and pulse migrations. The location of lineages
-over time can be tracked in as much detail as required: each tree node is
-automatically associated with the population in which it arose, the location of
-lineages can be recorded at any given time via census events, or every lineage
-migration can be recorded.
-
-A major change for \msprime\ 1.0 is the introduction of the new Demography API,
-designed to address a serious flaw in the \msprime\ 0.x interface which lead to
-a number of avoidable errors in downstream
-simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
-separate parameters be provided to the \texttt{simulate} function to describe a
-demographic model, making it easy to accidentally omit information. The 1.0 API
-resolves this issue by creating a new \texttt{Demography} class, which
-encapsulates all information about the demographic model, and fully decouples
-the definition from other simulation details. An instance of this class is then
-provided as a parameter to the new \texttt{sim\_ancestry} function,
-substantially reducing the potential for error. Another improvement over the
-0.X APIs is the introduction of explicit population split and admixture events,
-and a population state machine that ensures that lineages cannot migrate into
-(or be sampled from) inactive populations. This demography model is compatible
-with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
-class supports importing Demes models. We also support importing demographic
-models in the form of Newick species tree descriptions, as well as those
-produced by programs such as *BEAST~\citep{heled2009bayesian}.
-
-Many popular approaches in population genetics use the distribution of
-coalescence rates between pairs of lineages in one or more populations to infer
-effective population sizes over
-time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
-split times and subsequent migration rates between
-populations~\citep{wang2020tracking}. The \texttt{DemographyDebugger} provides
-methods to compute a suite of numerical predictions about the location of
-lineages over time, for example the coalescence rates for two or more lineages
-drawn from populations at specified times in the past. This provides a valuable
-ground-truth when evaluating such inference methods, as illustrated by
-\cite{adrion2020community}.
-
-\subsection*{Selection}
-Selection was added to the coalescent ... Lots of simulators exist:
-SelSim~\citep{spencer2004selsim}
-mbs~\citep{teshima2009mbs},
-msms~\citep{ewing2010msms},
-% What about Cosi, did it simulate selection?
-Cosi2~\citep{shlyakhter2014cosi2},
-and discoal~\citep{kern2016discoal}.
-
-We have implemented the basic infrastructure exists for the structured
-coalescent, which makes adding support for, e.g.,
-inversions straightforward~\citep{peischl2013sequential}
-
-\subsection*{Discrete time Wright-Fisher}
-The coalescent is a good approximation for ancient history, but
-is bad for the recent past and in particular when we have large
-sample
-sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
-Describe basic
-approach [DOM, can you fill in please].
-See ~\cite{nelson2020accounting} for more details.
-We compare the performance of this simulation in Fig X
-with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
-
-We support changing the simulation model, so that we can easily
-run hybrid simulations like described by~\cite{bhaskar2014distortion}
-This ia a powerful approach because it lets us combine any
-number of different simulation models.
-
-\subsection*{Finite sites mutations}
-Version 0.x of \texttt{msprime} provided a very simple mutation model,
-supporting only the infinite sites model.
-With version 1.0, we now support a large number of mutation models,
-including
-JC69~\citep{jukes1969evolution},
-F84~\citep{felsenstein1996hidden},
-and GTR~\citep{tavare1986some} nucleotide models
-and the BLOSUM62~\citep{henikoff1992amino}
-and PAM~\citep{dayhoff1978} amino acid models.
-The general model accepts an arbitrary transition matrix between
-any number of alleles (which can be arbitrary unicode strings),
-and so other models such as the Kimura two and three
-parameter models~\citep{kimura1980simple,kimura1981estimation}
-parameters models can be easily
-and efficiently defined in user code.
-Full support for the original infinite sites model is also maintained.
-The results of mutation simulations across a range of models are
-extensively validated
-against both theoretical expectations and output from
-Seq-Gen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}.
-
-Comparing the performance of mutation generation methods is problematic.
-Because there is no standard format for representing trees and
-the accompanying mutations, all existing methods are focused on
-writing out the resulting sequences. Methods are also not modular,
-and so it is very difficult to disentangle the time required to
-read in the tree file and output the sequences from the actual
-time required to perform the simulation.
-
-The mutation generation machinery in \msprime\ is extremely
-efficient. For example, we simulated mutations under the JC69
-model for a tree with $10^6$ leaves over $10^4$ sites (resulting
-in $\sim 280000$ mutations) in less that 1.5 seconds, including
-the time required for file input and output. While a systematic
-comparison of run-times is not feasible here, this is orders
-of magnitude faster than methods such as
-Seq-Gen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}
-and at least competitive with the recent method from
-\cite{demaio2021phastsim}.
-
-\textbf{TODO: would love to do a like-for-like comparison, but quite hard}
-
-Another substantial advantage of our approach is that the simulated
-mutations are stored directly in the output data format with
-a defined API. Although
-Pyvolve and Phastsim are implemented in Python, neither provide
-programatic access to the simulated output, outputting instead
-to a series of files which must subsequently be parsed.
-See the Data Model section for more details and discussion.
-
-Things we can do: future work will need to do things like
-Dawg~\citep{cartwright2005dna} and to implement models and
-indels~\citep{fletcher2009indelible}.
-
-\subsection*{Multiple merger coalescents}
-
-Kingman's coalescent assumes that only two ancestral lineages can merge at
-each merger event. Although this is generally a reasonable approximation there
-are certain situations in which the underlying mathematical assumptions are
-violated. For example in certain highly fecund organisms
-\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
-to produce numbers of offspring on the order of the population size and
-therefore a few individuals may produce the bulk of the offspring in any given
-generation~\citep{hedgecock_94}. These population dynamics violate basic
-assumptions of the Kingman coalescent, and are better modelled by
-`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
-two lineages can merge in a given event. Multiple-merger coalescent processes
-have also been shown to be relevant for modeling the effects of selection on
-gene genealogies \citep{Gillespie909,DS04}.
-
-Although multiple merger coalescents have been of significant theoretical
-interest for around two decades, there has been little practical software
-available to simulate these models.
-\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
-simulate a related spatial continuum model~\citep{barton2010new},
-\cite{zhu2015hybrid} simulate genealogies within a species tree
-based on a multiple-merger model, and
-\cite{becheler2020occupancy} provide a general method for simulating
-multiple merger processes
-as part of the Quetzal framework~\citep{becheler2019quetzal}.
-The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
-includes a number of extensions to the Beta-coalescent.
-None of these methods work with large genomes, and very little work
-has been performed on simulating multiple merger processes with recombination.
-
-We have added two multiple merger coalescent models in \msprime\ 1.0, the
-Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
-such models with recombination efficiently for the first time.
-These simulation models have been extensively validated against
-analytical results from the site frequency
-spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
- as well as more general properties of coalescent processes.
-Please see the Appendix for more details and model derivations.
 
@@
 
+\subsection*{Gene conversion}
+
+Gene conversion is \jkcomment{A couple of sentences describing gene
+conversion, and what the model is. What other simulators simulate
+gc? }
+We have implemented this model of gene
+conversion in \msprime\ 1.0, and validated the output against
+\ms\ and analytical results~\citep{wiuf2000coalescent}.
+
+\begin{figure}
+\caption{\label{fig-gc-perf}Comparison of the running time
+for E-coli simulations using \msprime\ and SimBac for varying
+sample sizes. The parameters used are [] based on [references].}
+\end{figure}
+
+Gene conversion is particularly useful to model homologous recombination in
+bacterial evolution.
+We compare the performance of \msprime\ with gene conversion to
+SimBac~\citep{brown2016simbac}, a simulator developed specifically to
+simulate bacterial evolution. Although \msprime\ is missing some features
+relevant to this case, it is far more efficient at simulating the basic
+processes. An important avenue for future work
+will be to add further features to model bacterial
+evolution. In particular, we aim to include bacterial gene transfer and utilize
+the succinct tree sequence structure to represent the resulting ancestral gene
+transfer graph~\citep{baumdicker2014AGTG}.
+
+\subsection*{Demography}
+% TODO should we put in some citations here for the models? Hardly seems
+% necessary.
+One of the key applications of population genetic simulations is to generate
+data for complex demographies. Beyond idealised cases such as stepping-stone or
+island models, or specialised cases such as isolation-with-migration models,
+analytical results are rarely possible. Simulation is therefore integral to the
+development and evaluation of methods for demographic inference. The demography
+model in \msprime\ is directly derived from the approach used in \ms, and
+supports an arbitrary number of randomly mating populations exchanging migrants
+at specified rates. A range of demographic events are provided, which allow for
+varying population parameters and migration rates over time, as well as
+population splits, admixtures and pulse migrations. The location of lineages
+over time can be tracked in as much detail as required: each tree node is
+automatically associated with the population in which it arose, the location of
+lineages can be recorded at any given time via census events, or every lineage
+migration can be recorded.
+
+A major change for \msprime\ 1.0 is the introduction of the new Demography API,
+designed to address a serious flaw in the \msprime\ 0.x interface which lead to
+a number of avoidable errors in downstream
+simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
+separate parameters be provided to the \texttt{simulate} function to describe a
+demographic model, making it easy to accidentally omit information. The 1.0 API
+resolves this issue by creating a new \texttt{Demography} class, which
+encapsulates all information about the demographic model, and fully decouples
+the definition from other simulation details. An instance of this class is then
+provided as a parameter to the new \texttt{sim\_ancestry} function,
+substantially reducing the potential for error. Another improvement over the
+0.X APIs is the introduction of explicit population split and admixture events,
+and a population state machine that ensures that lineages cannot migrate into
+(or be sampled from) inactive populations. This demography model is compatible
+with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
+class supports importing Demes models. We also support importing demographic
+models in the form of Newick species tree descriptions, as well as those
+produced by programs such as *BEAST~\citep{heled2009bayesian}.
+
+Many popular approaches in population genetics use the distribution of
+coalescence rates between pairs of lineages in one or more populations to infer
+effective population sizes over
+time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
+split times and subsequent migration rates between
+populations~\citep{wang2020tracking}. The \texttt{DemographyDebugger} provides
+methods to compute a suite of numerical predictions about the location of
+lineages over time, for example the coalescence rates for two or more lineages
+drawn from populations at specified times in the past. This provides a valuable
+ground-truth when evaluating such inference methods, as illustrated by
+\cite{adrion2020community}.
+
+\subsection*{Selection}
+Selection was added to the coalescent ... Lots of simulators exist:
+SelSim~\citep{spencer2004selsim}
+mbs~\citep{teshima2009mbs},
+msms~\citep{ewing2010msms},
+% What about Cosi, did it simulate selection?
+Cosi2~\citep{shlyakhter2014cosi2},
+and discoal~\citep{kern2016discoal}.
+
+We have implemented the basic infrastructure exists for the structured
+coalescent, which makes adding support for, e.g.,
+inversions straightforward~\citep{peischl2013sequential}
+
+\subsection*{Discrete time Wright-Fisher}
+The coalescent is a good approximation for ancient history, but
+is bad for the recent past and in particular when we have large
+sample
+sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
+Describe basic
+approach [DOM, can you fill in please].
+See ~\cite{nelson2020accounting} for more details.
+We compare the performance of this simulation in Fig X
+with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
+
+We support changing the simulation model, so that we can easily
+run hybrid simulations like described by~\cite{bhaskar2014distortion}
+This ia a powerful approach because it lets us combine any
+number of different simulation models.
+
+\subsection*{Finite sites mutations}
+Version 0.x of \texttt{msprime} provided a very simple mutation model,
+supporting only the infinite sites model.
+With version 1.0, we now support a large number of mutation models,
+including
+JC69~\citep{jukes1969evolution},
+F84~\citep{felsenstein1996hidden},
+and GTR~\citep{tavare1986some} nucleotide models
+and the BLOSUM62~\citep{henikoff1992amino}
+and PAM~\citep{dayhoff1978} amino acid models.
+The general model accepts an arbitrary transition matrix between
+any number of alleles (which can be arbitrary unicode strings),
+and so other models such as the Kimura two and three
+parameter models~\citep{kimura1980simple,kimura1981estimation}
+parameters models can be easily
+and efficiently defined in user code.
+Full support for the original infinite sites model is also maintained.
+The results of mutation simulations across a range of models are
+extensively validated
+against both theoretical expectations and output from
+Seq-Gen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}.
+
+Comparing the performance of mutation generation methods is problematic.
+Because there is no standard format for representing trees and
+the accompanying mutations, all existing methods are focused on
+writing out the resulting sequences. Methods are also not modular,
+and so it is very difficult to disentangle the time required to
+read in the tree file and output the sequences from the actual
+time required to perform the simulation.
+
+The mutation generation machinery in \msprime\ is extremely
+efficient. For example, we simulated mutations under the JC69
+model for a tree with $10^6$ leaves over $10^4$ sites (resulting
+in $\sim 280000$ mutations) in less that 1.5 seconds, including
+the time required for file input and output. While a systematic
+comparison of run-times is not feasible here, this is orders
+of magnitude faster than methods such as
+Seq-Gen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}
+and at least competitive with the recent method from
+\cite{demaio2021phastsim}.
+
+\textbf{TODO: would love to do a like-for-like comparison, but quite hard}
+
+Another substantial advantage of our approach is that the simulated
+mutations are stored directly in the output data format with
+a defined API. Although
+Pyvolve and Phastsim are implemented in Python, neither provide
+programatic access to the simulated output, outputting instead
+to a series of files which must subsequently be parsed.
+See the Data Model section for more details and discussion.
+
+Things we can do: future work will need to do things like
+Dawg~\citep{cartwright2005dna} and to implement models and
+indels~\citep{fletcher2009indelible}.
+
+\subsection*{Multiple merger coalescents}
+
+Kingman's coalescent assumes that only two ancestral lineages can merge at
+each merger event. Although this is generally a reasonable approximation there
+are certain situations in which the underlying mathematical assumptions are
+violated. For example in certain highly fecund organisms
+\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
+to produce numbers of offspring on the order of the population size and
+therefore a few individuals may produce the bulk of the offspring in any given
+generation~\citep{hedgecock_94}. These population dynamics violate basic
+assumptions of the Kingman coalescent, and are better modelled by
+`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
+two lineages can merge in a given event. Multiple-merger coalescent processes
+have also been shown to be relevant for modeling the effects of selection on
+gene genealogies \citep{Gillespie909,DS04}.
+
+Although multiple merger coalescents have been of significant theoretical
+interest for around two decades, there has been little practical software
+available to simulate these models.
+\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
+simulate a related spatial continuum model~\citep{barton2010new},
+\cite{zhu2015hybrid} simulate genealogies within a species tree
+based on a multiple-merger model, and
+\cite{becheler2020occupancy} provide a general method for simulating
+multiple merger processes
+as part of the Quetzal framework~\citep{becheler2019quetzal}.
+The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
+includes a number of extensions to the Beta-coalescent.
+None of these methods work with large genomes, and very little work
+has been performed on simulating multiple merger processes with recombination.
+
+We have added two multiple merger coalescent models in \msprime\ 1.0, the
+Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
+such models with recombination efficiently for the first time.
+These simulation models have been extensively validated against
+analytical results from the site frequency
+spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
+ as well as more general properties of coalescent processes.
+Please see the Appendix for more details and model derivations.
+
 \subsection*{Instantaneous bottlenecks}
@@
 \bibliography{paper}
+
+%% local definitions for section multiple merger coalescents
+ \newcommand{\be}{\begin{equation}}
+ \newcommand{\ee}{\end{equation}}
+ \newcommand{\bd}{\begin{displaymath}}
+ \newcommand{\ed}{\end{displaymath}}
+\newcommand{\IN}{\ensuremath{\mathds{N}}}%
+\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
+\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
+\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%
+
+\NewEnviron{esplit}[1]{%
+\begin{equation}
+\label{#1}
+\begin{split}
+  \BODY
+\end{split}\end{equation}
+}
 

### 2021-07-12 evaluation/recombination.py

@@ -1,0 +1,62 @@
+"""
+Generate data for the ARG and recombination performance figures.
+"""
+import numpy as np
+import msprime
+import matplotlib.pyplot as plt
+import pandas as pd
+
+
+def run_arg_sim():
+
+
+    L_col = []
+    size_col = []
+    arg_col = []
+    for megabases in np.linspace(0.1, 5, 20):
+        L = int(megabases * 1_000_000)
+        arg_ts = msprime.sim_ancestry(
+            100,
+            population_size=10_000,
+            sequence_length=L,
+            recombination_rate=1e-8,
+            random_seed=42,
+            record_full_arg=True,
+        )
+        flags = arg_ts.tables.nodes.flags
+        # Samples have flags == 1 and ordinary coalescent nodes have
+        # flags == 0. So, anything > 1 is an ARG node.
+        arg_nodes = flags > 1
+        L_col.append(L)
+        arg_fraction = np.sum(arg_nodes) / arg_ts.num_nodes
+        ts = arg_ts.simplify()
+        size_ratio = ts.tables.nbytes / arg_ts.nbytes
+        size_col.append(size_ratio)
+        arg_col.append(arg_fraction)
+        print(L, arg_fraction, size_ratio)
+    data = {"L": L_col, "size_ratio": size_col, "arg_nodes": arg_col}
+
+    df = pd.DataFrame(data)
+    print(df)
+    df.to_csv("data/arg.csv")
+
+
+def plot_arg_data():
+    df = pd.read_csv("data/arg.csv")
+    print(df)
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
+    ax1.set_title("(A)")
+    ax1.set_xlabel("Scaled recombination rate")
+
+    ax2.set_title("(B)")
+    ax2.plot(df.L, df.arg_nodes, label="Fraction of ARG nodes")
+    ax2.plot(df.L, df.size_ratio, label="size(tree sequence) / size(ARG)")
+    ax2.set_xlabel("Sequence length")
+    ax2.legend()
+
+    plt.savefig("figures/recombination.pdf")
+
+
+if __name__ == "__main__":
+    # run_arg_sim()
+    plot_arg_data()

### 2021-07-12 paper.tex
Draft of arg stuff.

@@
 
-The \msprime~simulator generates tree sequences by simulating realisations of the
-Ancestral Recombination Graph (ARG).
-By default, only the marginal genealogy at each site is recorded in the tree sequence
-output.
-Marginal genealogies are sufficient for e.g.~simulating and storing sequence data,
-and involve fewer edges than nodes than storing a realisation of the full ARG.
-However, the mapping from ARG realisations to marginal tree sequences is not lossless.
-For example, times of recombination events can be recovered from the former
-but not the latter.
-The same is true of the number of extant ancestors at each time in the past.
-In addition to the default behaviour, we support the retention of the full
-ARG realisation using the \texttt{record\_full\_arg} option.
-This option augments the default tree sequence data structure with three
-new kinds of nodes:
-\begin{itemize}
-\item Each recombination event is marked by two nodes with identical birth times
-given by the time of the recombination event, and marked with the
-\texttt{NODE\_IS\_RE\_EVENT} flag.
-Each node represents an ancestor of the recombinant child node, and edges connect
-the relevant segments of genome from the child node to these two ancestors.
-\item We store common ancestor nodes even when the corresponding merger event
-does not result in coalescence of ancestral material.
-The resulting nodes are marked with the \texttt{NODE\_IS\_CA\_EVENT} flag.
-\item When multiple populations are simulated, migration events are depicted by nodes
-with the \texttt{NODE\_IS\_MIG\_EVENT} flag, and with birth times equal to the time of
-the migration.
-The population flag of the migration node identifies the population to which the lineage
-belongs after the reverse time migration event.
-\end{itemize}
-
-The likelihood, i.e.~the joint sampling probability of the ancestry and mutations,
-is an example of a function which is intractable when the ancestry is stored as a
-marginal tree sequence, but tractable when the ancestry is given by a full
-ARG realisation.
+The term Ancestral Recombination Graph (ARG) was introduced by
+Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
+representation of the coalescent with recombination stochastic process. This
+formulation was complementary to Hudson's earlier
+work~\citep{hudson1983properties}, and substantially increased our theoretical
+understanding of the recombination process. In Griffiths' ARG formulation, we
+regard a realisation of the coalescent with recombination as a graph in which
+nodes represent common ancestor or recombination events, and edges represent
+lineages through which ancestral material flows. We may consider the ``big''
+ARG, in which we wait an exponentially large time until all sampled lineages
+coalescence into a single ancestor~\citep{ethier1990two}, or the ``little'' ARG
+in which we wait until all marginal trees have fully coalescenced (like
+Hudson's algorithm).
+
+% Seems somewhat unfair to single out Iain and Alywyn here, but a systematic
+% review would be long and tedious, and I'm sure they won't be too upset about it.
+However, this original definition of the ARG in terms of a stochastic process
+is rarely what is meant when the term is currently used. The term has come to
+essentially mean a record of genetic ancestry as a graph; for example,
+\cite{mathieson2020ancestry} informally define ARGs as the subset of an
+individual's pedigree from which they have inherited ancestral material. Such
+imprecision is unhelpful, since there are any number of graphical encodings of
+genetic ancestry, and referring to the original Griffiths papers does little
+to clarify. As we shall see, there are situations in which the precise form of the
+graph is important, and conflating high-level representation of genetic
+ancestry with this precise mathematical object is not helpful.
+One may wonder if a more generic term such as ``genetic pedigree''
+would be a more useful description, in such cases where the precise form
+of the encoding is not important.
+
+% ~\cite[.e.g.][]{minichiello2007mapping}
+% \cite{minichiello2006mapping} are somewhat more explicit
+% "An ARG provides a record of all coalescence and recombination events since the
+% divergence of the sequences under study and specifies a complete genealogy at
+% each genomic position
+% \citep{rasmussen2014genome}
+For our purposes here, an ARG is a realisation of the coalescent with
+recombination, in the Griffiths sense.
+As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
+essentially works by dynamically generating and traversing a little ARG.
+The graph is not explicitly represented in memory, but is partially
+present through the extant lineages and the ancestral material they carry
+over time. Since the ARG is a large and unwieldy object
+(Fig.~\ref{fig-recombination-perf}B) we do not use it as output, but
+rather store the information required to recover the genealogical
+history as nodes and edges in a tree sequence (see section XX),
+or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
+This approach is one of the key reasons for \msprime's efficiency.
+
+Although the tree sequences output by \msprime\ contain full information
+about the genealogical trees, their correlation structure along the genome,
+and the ancestral haplotypes on which coalescences occured, information
+is lost in this mapping down from ARG space. In particular, we lose
+information about individuals who were common ancestors but where
+no coalescences occured and also information about the precise time
+and genomic location of recombination events. In most cases, such
+information is of little relevance as it is in principle unknowable,
+but there are occasions in which is useful. We therefore provide
+the \texttt{record\_full\_arg} option in \msprime\
+to store a representation of the compete ARG traversed during simulation.
+This is done by storing extra nodes (marked with specific flags, so they
+can be easily identified later) in the tree sequence.
+Fig.~\ref{fig-recombination-perf}B shows that these extra ARG nodes
+and edges quickly come to dominate the output as we increase the sequence length,
+and the resulting tree sequence files are many times larger
+than the standard minimal tree sequence representation.
+
+One situation in which a record of the full ARG is necessary is when we
+wish to compute likelihoods during inference.
 The likelihood is a central quantity in evaluating the plausibility of a putative
@@
 e.g.~approaches based on maximum likelihood, and as an ingredient of
-methods such as Metropolis-Hastings
-\citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian}.
-
-We provide two functions to facilitate likelihood-based inference.
-Both are implemented only for the simplest case of the standard ARG with a
-constant population size, and require tree sequences compatible with the
-\texttt{record\_full\_arg} option as their arguments.
-\begin{itemize}
-\item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
-the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
-per-generation recombination probability \texttt{r} and population size \texttt{N}.
-Specifically, the function returns the logarithm of
-\begin{equation*}
-\Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
-	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
-		+ r l_i \Big] t_i  \Bigg),
-\end{equation*}
-where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
-$k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
-in that interval that would split ancestral material should they recombine,
-$q$ is the total number of events in the tree sequence $\texttt{ts}$,
-$q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
-of time intervals which end in a recombination, and $g_i$ is the corresponding
-\emph{gap}:  the length of material in which the recombination at the end of the
-$i$th period could have taken place without affecting the resulting genealogy.
-A recombination in ancestral material will always have $g_i = 1$, while a recombination
-in trapped non-ancestral material can result in other values of $g_i$.
-For a continuous model of the genome and a recombination in ancestral material,
-we set $g_i = 1$ and interpret the result as a density.
-\item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
-natural logarithm of the probability of the mutations recorded in the tree sequence
-\texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
-a normalising constant which depends on the pattern of mutations,
-but not on the tree sequence or the per-site, per-generation mutation
-probability \texttt{m}.
-Specifically, the function returns the logarithm of
-\begin{equation*}
-e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
-\prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
-\end{equation*}
-where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
-in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
-total branch length on which $\gamma$ could have arisen while appearing on all
-of the leaves of \texttt{ts} it does, and on no others.
-Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
-mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
-edges.
-\end{itemize}
-
-\begin{comment}
-Previous draft of the ARG feature retained here as a comment.
-
-The definition of ARGs is vague and confusing.
-\cite{minichiello2006mapping} define it as the data structure
-rather than the original stochastic
-process~\citep{griffiths1991two,griffiths1997ancestral}.
-\cite{rasmussen2014genome} have a slightly different approach.
-Sometimes the ARG is useful.
-The tree sequence output for \msprime\ can be turned into
-an ARG using the \texttt{record\_full\_arg} option, where we store
-recombination nodes.
-
-We also provide an interface to compute the likelihood of a given ARG
-under the~\cite{kuhner2000maximum} model.
-\end{comment}
+methods such as
+Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
+We provide functions to compute the likelihood of ARG realisations
+(encoded using the representation of the \texttt{record\_full\_arg} option)
+and mutational patterns under
+the standard coalescent and infinite sites mutation model.
+
+% We provide two functions to facilitate likelihood-based inference.
+% Both are implemented only for the simplest case of the standard ARG with a
+% constant population size, and require tree sequences compatible with the
+% \texttt{record\_full\_arg} option as their arguments.
+% \begin{itemize}
+% \item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
+% the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
+% per-generation recombination probability \texttt{r} and population size \texttt{N}.
+% Specifically, the function returns the logarithm of
+% \begin{equation*}
+% \Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
+% 	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
+% 		+ r l_i \Big] t_i  \Bigg),
+% \end{equation*}
+% where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
+% $k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
+% in that interval that would split ancestral material should they recombine,
+% $q$ is the total number of events in the tree sequence $\texttt{ts}$,
+% $q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
+% of time intervals which end in a recombination, and $g_i$ is the corresponding
+% \emph{gap}:  the length of material in which the recombination at the end of the
+% $i$th period could have taken place without affecting the resulting genealogy.
+% A recombination in ancestral material will always have $g_i = 1$, while a recombination
+% in trapped non-ancestral material can result in other values of $g_i$.
+% For a continuous model of the genome and a recombination in ancestral material,
+% we set $g_i = 1$ and interpret the result as a density.
+% \item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
+% natural logarithm of the probability of the mutations recorded in the tree sequence
+% \texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
+% a normalising constant which depends on the pattern of mutations,
+% but not on the tree sequence or the per-site, per-generation mutation
+% probability \texttt{m}.
+% Specifically, the function returns the logarithm of
+% \begin{equation*}
+% e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
+% \prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
+% \end{equation*}
+% where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
+% in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
+% total branch length on which $\gamma$ could have arisen while appearing on all
+% of the leaves of \texttt{ts} it does, and on no others.
+% Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
+% mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
+% edges.
+% \end{itemize}
+
+% \begin{comment}
+% Previous draft of the ARG feature retained here as a comment.
+
+% The definition of ARGs is vague and confusing.
+% \cite{minichiello2006mapping} define it as the data structure
+% rather than the original stochastic
+% process~\citep{griffiths1991two,griffiths1997ancestral}.
+% \cite{rasmussen2014genome} have a slightly different approach.
+% Sometimes the ARG is useful.
+% The tree sequence output for \msprime\ can be turned into
+% an ARG using the \texttt{record\_full\_arg} option, where we store
+% recombination nodes.
+
+% We also provide an interface to compute the likelihood of a given ARG
+% under the~\cite{kuhner2000maximum} model.
+% \end{comment}
 

### 2021-07-12 paper.tex
Revisions.

@@
 
-The term Ancestral Recombination Graph (ARG) was introduced by
+The Ancestral Recombination Graph (ARG) was introduced by
 Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
@@
 work~\citep{hudson1983properties}, and substantially increased our theoretical
-understanding of the recombination process. In Griffiths' ARG formulation, we
+understanding of the recombination. In Griffiths' ARG formulation, we
 regard a realisation of the coalescent with recombination as a graph in which
@@
 
-% Seems somewhat unfair to single out Iain and Alywyn here, but a systematic
-% review would be long and tedious, and I'm sure they won't be too upset about it.
+% Seems somewhat unfair to single out Iain and Aylwyn here, but a systematic
+% review would be long and tedious, and I'm sure they won't be too upset about
+% it.
 However, this original definition of the ARG in terms of a stochastic process
 is rarely what is meant when the term is currently used. The term has come to
-essentially mean a record of genetic ancestry as a graph; for example,
-\cite{mathieson2020ancestry} informally define ARGs as the subset of an
+essentially mean a representation of genetic ancestry as a graph; for example,
+\cite{mathieson2020ancestry} informally define an ARG as the subset of an
 individual's pedigree from which they have inherited ancestral material. Such
-imprecision is unhelpful, since there are any number of graphical encodings of
-genetic ancestry, and referring to the original Griffiths papers does little
-to clarify. As we shall see, there are situations in which the precise form of the
-graph is important, and conflating high-level representation of genetic
-ancestry with this precise mathematical object is not helpful.
-One may wonder if a more generic term such as ``genetic pedigree''
-would be a more useful description, in such cases where the precise form
-of the encoding is not important.
-
-% ~\cite[.e.g.][]{minichiello2007mapping}
-% \cite{minichiello2006mapping} are somewhat more explicit
-% "An ARG provides a record of all coalescence and recombination events since the
-% divergence of the sequences under study and specifies a complete genealogy at
-% each genomic position
-% \citep{rasmussen2014genome}
+imprecision is unfortunate, since there are any number of graphical encodings of
+genetic ancestry, and referring to the original Griffiths papers does little to
+clarify. As we shall see, there are situations in which the precise information
+stored of the graph is important, and conflating a high-level
+discussion of genetic ancestry with this mathematical object is not
+helpful. One may wonder if a more generic term such as ``genetic pedigree''
+would be a more useful description, in such cases where the precise form of the
+encoding is not important.
+
 For our purposes here, an ARG is a realisation of the coalescent with
@@
 As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
-essentially works by dynamically generating and traversing a little ARG.
+works by dynamically generating and traversing a little ARG.
 The graph is not explicitly represented in memory, but is partially
@@
 or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
-This approach is one of the key reasons for \msprime's efficiency.
 
@@
 is lost in this mapping down from ARG space. In particular, we lose
-information about individuals who were common ancestors but where
-no coalescences occured and also information about the precise time
+information about ancestral haplotypes that were common ancestors but
+in which no coalescences occured, and also information about the precise time
 and genomic location of recombination events. In most cases, such
 information is of little relevance as it is in principle unknowable,
-but there are occasions in which is useful. We therefore provide
+but there are occasions such as visualisation or computing likelihoods (see
+below) in which is useful. We therefore provide
 the \texttt{record\_full\_arg} option in \msprime\
@@
 This is done by storing extra nodes (marked with specific flags, so they
-can be easily identified later) in the tree sequence.
-Fig.~\ref{fig-recombination-perf}B shows that these extra ARG nodes
-and edges quickly come to dominate the output as we increase the sequence length,
+can be easily identified later) and edges in the tree sequence.
+Fig.~\ref{fig-recombination-perf}B shows that this extra ARG information
+quickly comes to dominate the output as we increase the sequence length,
 and the resulting tree sequence files are many times larger
@@
 
-% \begin{comment}
-% Previous draft of the ARG feature retained here as a comment.
-
-% The definition of ARGs is vague and confusing.
-% \cite{minichiello2006mapping} define it as the data structure
-% rather than the original stochastic
-% process~\citep{griffiths1991two,griffiths1997ancestral}.
-% \cite{rasmussen2014genome} have a slightly different approach.
-% Sometimes the ARG is useful.
-% The tree sequence output for \msprime\ can be turned into
-% an ARG using the \texttt{record\_full\_arg} option, where we store
-% recombination nodes.
-
-% We also provide an interface to compute the likelihood of a given ARG
-% under the~\cite{kuhner2000maximum} model.
-% \end{comment}
-
 \subsection*{Gene conversion}

### 2021-07-12 evaluation/recombination.py

@@ -55,6 +55,7 @@ def plot_arg_data():
     ax2.legend()
 
     plt.savefig("figures/recombination.pdf")
+    plt.savefig("figures/recombination.png")
 
 
 if __name__ == "__main__":

### 2021-07-12 paper.tex
fixups

@@
 work~\citep{hudson1983properties}, and substantially increased our theoretical
-understanding of the recombination. In Griffiths' ARG formulation, we
+understanding of recombination. In Griffiths' ARG formulation, we
 regard a realisation of the coalescent with recombination as a graph in which
-nodes represent common ancestor or recombination events, and edges represent
+vertices represent common ancestor or recombination events, and edges represent
 lineages through which ancestral material flows. We may consider the ``big''
@@
 clarify. As we shall see, there are situations in which the precise information
-stored of the graph is important, and conflating a high-level
+stored in the graph is important, and conflating a high-level
 discussion of genetic ancestry with this mathematical object is not
@@
 the \texttt{record\_full\_arg} option in \msprime\
-to store a representation of the compete ARG traversed during simulation.
+to store a representation of the complete ARG traversed during simulation.
 This is done by storing extra nodes (marked with specific flags, so they

### 2021-07-12 paper.tex
Updates for review.

@@
 vertices represent common ancestor or recombination events, and edges represent
-lineages through which ancestral material flows. We may consider the ``big''
-ARG, in which we wait an exponentially large time until all sampled lineages
-coalescence into a single ancestor~\citep{ethier1990two}, or the ``little'' ARG
-in which we wait until all marginal trees have fully coalescenced (like
-Hudson's algorithm).
+lineages through which ancestral material flows.
+We may consider the big ARG, in which we track lineages arising out of
+recombinations regardless of whether they carry ancestral material and
+consequently wait an exponentially large time until all sampled lineages
+coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
+which we only track genetic ancestors, and wait until all marginal trees
+have fully coalesced (like Hudson's algorithm).
 
@@
 
-For our purposes here, an ARG is a realisation of the coalescent with
-recombination, in the Griffiths sense.
+For our purposes, an ARG is a realisation of the coalescent with
+recombination, in the Griffiths (little ARG) sense.
 As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
@@
 the standard coalescent and infinite sites mutation model.
-
-% We provide two functions to facilitate likelihood-based inference.
-% Both are implemented only for the simplest case of the standard ARG with a
-% constant population size, and require tree sequences compatible with the
-% \texttt{record\_full\_arg} option as their arguments.
-% \begin{itemize}
-% \item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
-% the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
-% per-generation recombination probability \texttt{r} and population size \texttt{N}.
-% Specifically, the function returns the logarithm of
-% \begin{equation*}
-% \Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
-% 	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
-% 		+ r l_i \Big] t_i  \Bigg),
-% \end{equation*}
-% where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
-% $k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
-% in that interval that would split ancestral material should they recombine,
-% $q$ is the total number of events in the tree sequence $\texttt{ts}$,
-% $q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
-% of time intervals which end in a recombination, and $g_i$ is the corresponding
-% \emph{gap}:  the length of material in which the recombination at the end of the
-% $i$th period could have taken place without affecting the resulting genealogy.
-% A recombination in ancestral material will always have $g_i = 1$, while a recombination
-% in trapped non-ancestral material can result in other values of $g_i$.
-% For a continuous model of the genome and a recombination in ancestral material,
-% we set $g_i = 1$ and interpret the result as a density.
-% \item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
-% natural logarithm of the probability of the mutations recorded in the tree sequence
-% \texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
-% a normalising constant which depends on the pattern of mutations,
-% but not on the tree sequence or the per-site, per-generation mutation
-% probability \texttt{m}.
-% Specifically, the function returns the logarithm of
-% \begin{equation*}
-% e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
-% \prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
-% \end{equation*}
-% where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
-% in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
-% total branch length on which $\gamma$ could have arisen while appearing on all
-% of the leaves of \texttt{ts} it does, and on no others.
-% Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
-% mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
-% edges.
-% \end{itemize}
+See the Appendix for details.
 
@@
 \appendix
+
+\label{app-likelihoods}
+\section{Likelihood calculations}
+
+We provide two functions to facilitate likelihood-based inference.
+Both are implemented only for the simplest case of the standard ARG with a
+constant population size, and require tree sequences compatible with the
+\texttt{record\_full\_arg} option as their arguments.
+\begin{itemize}
+\item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
+the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
+per-generation recombination probability \texttt{r} and population size \texttt{N}.
+Specifically, the function returns the logarithm of
+\begin{equation*}
+\Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
+	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
+		+ r l_i \Big] t_i  \Bigg),
+\end{equation*}
+where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
+$k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
+in that interval that would split ancestral material should they recombine,
+$q$ is the total number of events in the tree sequence $\texttt{ts}$,
+$q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
+of time intervals which end in a recombination, and $g_i$ is the corresponding
+\emph{gap}:  the length of material in which the recombination at the end of the
+$i$th period could have taken place without affecting the resulting genealogy.
+A recombination in ancestral material will always have $g_i = 1$, while a recombination
+in trapped non-ancestral material can result in other values of $g_i$.
+For a continuous model of the genome and a recombination in ancestral material,
+we set $g_i = 1$ and interpret the result as a density.
+\item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
+natural logarithm of the probability of the mutations recorded in the tree sequence
+\texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
+a normalising constant which depends on the pattern of mutations,
+but not on the tree sequence or the per-site, per-generation mutation
+probability \texttt{m}.
+Specifically, the function returns the logarithm of
+\begin{equation*}
+e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
+\prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
+\end{equation*}
+where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
+in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
+total branch length on which $\gamma$ could have arisen while appearing on all
+of the leaves of \texttt{ts} it does, and on no others.
+Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
+mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
+edges.
+\end{itemize}
+
 \label{app-multiple-mergers}

### 2021-07-13 paper.tex

@@
 genetic ancestry, and referring to the original Griffiths papers does little to
-clarify. As we shall see, there are situations in which the precise information
+clarify. As we see below, there are situations in which the precise information
 stored in the graph is important, and conflating a high-level
@@
 would be a more useful description, in such cases where the precise form of the
-encoding is not important.
+encoding is not important. \jkcomment{Genetic pedigree is probably
+just as confusing, on reflection. We can either come up with a different
+term, or delete this sentence.}
 
@@
 about the genealogical trees, their correlation structure along the genome,
-and the ancestral haplotypes on which coalescences occured, information
-is lost in this mapping down from ARG space. In particular, we lose
+and the ancestral haplotypes on which coalescences occured, some information
+is lost in this mapping down from ARG space to the minimal tree sequence
+form. In particular, we lose
 information about ancestral haplotypes that were common ancestors but

### 2021-07-13 paper.tex

@@
 and, until recently, approximate models of
-recombination~\citep{mcvean2005approximating,staab2014scrm}
+recombination~\citep{mcvean2005approximating,staab2015scrm}
 were necessary to simulate whole chromosomes.
@@
 allow simulations more closely approximate the coalescent
-with recombination~\citep{chen2009fast,wang2014new,staab2014scrm}.
+with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
 The SMC is an important analytical approximation
@@
 \subsection*{Simulation interface}
-The primary interface for \msprime\ is through a Python application
-programming interface (API). This has many advantages. One is
-that we can grow the interface over time. Another is that
-downstream applications such as \texttt{stdpopsim}~\citep{adrion2020community}
-can be developed.
-
-The \texttt{msp simulate} command also supports demographic models
-specified in Demes format~\citep{gower2021demes}, allowing users with
-no Python experience to run complex demographic simulations.
+The majority of simulation packages are controlled either through
+a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
+a text-based input file
+format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
+or a mixture of both.
+Command line interfaces such as \ms\ make it very easy to run simple
+simulations, but as model complexity and the number of parameters specified increases
+they become difficult to understand and
+error-prone~\citep{ragsdale2020lessons,gower2021demes}.
+Specifying parameters through a text file alleviates this problem to a degree,
+but lacks flexibility, for example, when running simulations with parameters
+drawn from some distribution. In practice, for any significant simulation
+project (or at least, any reproducible one), users will write a script
+to generate the required command lines or input parameter files,
+invoke the simulation engine, and process the results in some way.
+This process is cumbersome and labour intensive, and packages
+such \texttt{coala}~\citep{staab2016coala} have been developed
+to allow simulations be run directly in a high-level
+scripting language.
+
+The more recent trend has been to move away from this file and command-line
+driven approach and to instead provide direct interfaces to the simulation
+engines via an Application Programming Interface (API)~\citep[e.g.][]{
+thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}. This
+approach greatly increases the flexibility of simulations, and allows all
+processing for many simulation analyses to be performed in a single script. The
+primary interface for \msprime\ is through a well documented and stable Python
+API, which has allowed a number of downstream tools to be built on
+% JK: must be more, right?
+top of it~\citep{adrion2020community,rivera2021simulation}
+as well as numerous easily reproducible simulation
+% TODO: more papers with exemplorary reproducible simulation pipelines based
+% on msprime!
+pipelines~\citep[e.g.][]{martin2017human}.
+We also provide an \ms\ compatible
+command line interface to support existing workflows, as well as the
+\texttt{msp simulate} command that supports specifying demographic models in
+Demes format~\citep{gower2021demes}.
+
+A major change for the 1.0 release is the introduction of a new set of APIs,
+designed in part to avoid sources of error (see the Demography section) but
+also to provide more appropriate defaults while keeping compatibility with
+existing code. In the new APIs, ancestry and mutation simulation are fully
+separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
+functions replacing the legacy \texttt{simulate} function. Among other changes,
+the new APIs default to discrete genome coordinates and finite sites mutations,
+resolving a major source of confusion and error. The 0.x APIs are fully
+supported and tested, and will be maintained for the foreseeable future.
 
@@
 left-to-right
-approaches~\citep{wiuf1999recombination,chen2009fast,staab2014scrm}
+approaches~\citep{wiuf1999recombination,chen2009fast,staab2015scrm}
 is the ease with which additional processes can be integrated.

### 2021-07-13 paper.tex
Add dependent packages from GitHub

@@
 API, which has allowed a number of downstream tools to be built on
-% JK: must be more, right?
-top of it~\citep{adrion2020community,rivera2021simulation}
+top of
+% Found these through the Dependent packages on GitHub
+it~\citep{kamm2020efficiently,adrion2020predicting,adrion2020community,
+rivera2021simulation,de2021geonomics}
 as well as numerous easily reproducible simulation

### 2021-07-14 paper.tex

@@
 engines via an Application Programming Interface (API)~\citep[e.g.][]{
-thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}. This
-approach greatly increases the flexibility of simulations, and allows all
-processing for many simulation analyses to be performed in a single script. The
-primary interface for \msprime\ is through a well documented and stable Python
-API, which has allowed a number of downstream tools to be built on
-top of
+thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
+The primary interface for \msprime\ is through a thoroughly documented and
+stable Python
+API, which has allowed a number of downstream tools to be built on top of
 % Found these through the Dependent packages on GitHub
-it~\citep{kamm2020efficiently,adrion2020predicting,adrion2020community,
-rivera2021simulation,de2021geonomics}
-as well as numerous easily reproducible simulation
-% TODO: more papers with exemplorary reproducible simulation pipelines based
-% on msprime!
-pipelines~\citep[e.g.][]{martin2017human}.
-We also provide an \ms\ compatible
-command line interface to support existing workflows, as well as the
-\texttt{msp simulate} command that supports specifying demographic models in
-Demes format~\citep{gower2021demes}.
-
-A major change for the 1.0 release is the introduction of a new set of APIs,
+it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
+adrion2020community,adrion2020predicting, kamm2020efficiently,
+mckenzie2020ipcoal, montinaro2020revisiting,
+de2021geonomics,rivera2021simulation}.
+As well as providing a stable and efficient platform for building
+downstream applications, \msprime's Python API makes it much easier to
+build reproducible simulation pipelines, as the entire workflow can
+be encapsulated in a single script, and package and version
+dependencies can be explicitly stated using the \texttt{pip}
+or \texttt{conda} package managers.
+For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
+in the influential simulation analysis of
+\cite{martin2017human} were only detected because the pipeline
+could be easily run and reanalysed.
+
+A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
 designed in part to avoid sources of error (see the Demography section) but
@@
 supported and tested, and will be maintained for the foreseeable future.
+We also provide an \ms\ compatible
+command line interface to support existing workflows, and the
+\texttt{msp} program that supports specifying demographic models in
+Demes format~\citep{gower2021demes} from the command line.
 

### 2021-07-14 paper.tex
adds gene conversion section

@@
 \usepackage[right]{lineno}%
+\usepackage{microtype}
 %\usepackage{showkeys}
@@
 \begin{figure}
+% \begin{center}
+% \begin{tabular}{cc}
+% % \includegraphics[width=5cm]{msprime-scaling.png} &
+% % \includegraphics[width=5cm]{figures/recombination} \\
+% \end{tabular}
+% \end{center}
 \includegraphics[width=\textwidth]{figures/recombination}
@@
 Because they do not track such recombinations, efficient SMC implementations
-will be faster that \msprime for sufficiently high recombination rate,
+will be faster than \msprime\ for sufficiently high recombination rate,
 genome length, or $N_e$. %Until we decide to approximate the recombination process
@@
 
-% Seems somewhat unfair to single out Iain and Aylwyn here, but a systematic
-% review would be long and tedious, and I'm sure they won't be too upset about
-% it.
+\jkcomment{If we're going to have this paragraph of ARG bashing then
+I guess we'll have to do more than just single out Iain and Aylwyn.}
 However, this original definition of the ARG in terms of a stochastic process
 is rarely what is meant when the term is currently used. The term has come to
-essentially mean a representation of genetic ancestry as a graph; for example,
+essentially mean any representation of genetic ancestry in the presence
+of recombination as a graph-like structure; for example,
 \cite{mathieson2020ancestry} informally define an ARG as the subset of an
@@
 discussion of genetic ancestry with this mathematical object is not
-helpful. One may wonder if a more generic term such as ``genetic pedigree''
-would be a more useful description, in such cases where the precise form of the
-encoding is not important. \jkcomment{Genetic pedigree is probably
-just as confusing, on reflection. We can either come up with a different
-term, or delete this sentence.}
+helpful. The point of technical terms is to provide clarity and
+precision on what is being discussed, and it is now essentially
+impossible to know what is meant by the term ``ARG'';
+we therefore use the term ``genetic ancestry'' in high-level
+discussions.
 
@@
 
-Gene conversion is \jkcomment{A couple of sentences describing gene
-conversion, and what the model is. What other simulators simulate
-gc? }
+%introduction
+Gene conversion is a unidirectional transfer of a short segment of genetic material,
+for example between homologous chromosomes~\citep{Chen2007}.
+%description of ancestral process
+Just as recombination, gene conversion events split the ancestral material
+for some lineage, creating two independent lineages.
+In contrast to the previously described recombination process, where the ancestral
+material to the left and right of a recombination breakpoint follows two
+different ancestral lineages, gene conversion events distribute the genetic material
+of a short segment to one ancestral lineage and the part to the left and right of this
+segment to another lineage.
+%Franz Baumdicker: Not sure if this next sentence is necessary,
+% but it might be of interest for some people?
+The resulting distribution of ancestral material after a gene conversion event
+equals the distribution after two recombination events at the left and right position
+of the gene conversion tract, followed by a common ancestor event.
+%Impact on LD
+However, gene conversions impact much shorter segments than crossover recombinations,
+typically below 1kb, and thus affects linkage pattern differently than
+recombination~\citep{Korunes2017}.
+
+% other simulators
+Hudson's \ms\ optionally includes gene conversion and so does
+MaCS~\citep{chen2009fast}, while scrm~\citep{staab2014scrm} does not
+support gene conversions.
+As bacterial recombination can be modeled analogously to eukaryotic allelic
+gene conversion, simulators for microbial evolution often include gene conversion
+instead of recombination.
+This includes the backward simulator SimBac~\citep{brown2016simbac},
+the SMC based approach used in FastSimBac~\citep{DeMaio2017},
+and forward in time tools within the SLiM framework~\citep{Cury2020}.
+
+In \msprime\ 1.0 we implemented the coalescent with gene conversion as defined by
+\cite{wiuf2000coalescent}, where gene conversion events are initiated between any
+two sites in the sequence at a constant (scaled) rate $\gamma$ and
+the length of the transferred segment is given by a geometric distributed
+random variable with mean length $L$.
+% Above I used the discrete case, but we could easily switch to a description for the
+% continuous case.
 We have implemented this model of gene
@@
 \begin{figure}
+\includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
+% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
 \caption{\label{fig-gc-perf}Comparison of the running time
 for E-coli simulations using \msprime\ and SimBac for varying
-sample sizes. The parameters used are [] based on [references].}
+sample sizes. The parameters used are chosen to mimic the simulation
+of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
+gene conversion rate was set to $N_e g = \gamma = 0.015$ and
+the mean gene conversion tract length was set to $L = 500$, which is close to
+estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
+$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
+with mean tract length $L = 542$~\citep{Lapierre2016}.}
 \end{figure}
@@
 simulate bacterial evolution. Although \msprime\ is missing some features
-relevant to this case, it is far more efficient at simulating the basic
-processes. An important avenue for future work
+relevant to this case, e.g.\ the simulation of circular genomes, it is far
+more efficient at simulating the basic processes.
+Furthermore, gene conversion can be considered in combination
+with other current and future features of \msprime\ 1.0,
+including varying recombination rates and complex demographies.
+
+An important avenue for future work
 will be to add further features to model bacterial
@@
 
+\begin{figure}
+\Large{A FIGURE SHOWING DISCOAL AND MSPRIME}
+% \includegraphics[width=\textwidth]{figures/recombination}
+\caption{\label{fig-selection-perf}Comparison of the runtime for
+\msprime\ and \texttt{discoal} for simulating selective sweeps.
+}
+\end{figure}
+
 We have implemented the basic infrastructure exists for the structured
@@
 See ~\cite{nelson2020accounting} for more details.
-We compare the performance of this simulation in Fig X
+We compare the performance of this simulation in Fig.~\ref{fig-dtwf-perf}
 with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
+
+\begin{figure}
+\begin{center}
+% \includegraphics[width=6cm]{msprime-argon-runtimes.png}
+\end{center}
+\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
+simulations with ARGON. To ensure the comparison with ARGON is
+as fair as possible we run it with a mutation rate of 0 and with
+the minimum output options (so that it writes out only an empty VCF file)}
+\end{figure}
 

### 2021-07-14 evaluation/generate_gc_performance_data.py

@@ -1,0 +1,113 @@
+import subprocess
+import sys
+import time
+
+
+#import sys
+#sys.path.insert(0, "/home/franz/syncedprojects/msprime_performance/msprime")
+
+import msprime
+
+REPLICATES = 1
+
+Ne = 1
+mu = 1e-9
+r = 0
+
+
+def run_simbac():
+    L = mb(4.5)
+    #L = 10000
+    #sample_sizes = list(range(10, 100 + 1, 10)) + list(range(200, 900 + 1, 100)) + list(range(1000, 10000 + 1, 1000))
+    sample_sizes = list(range(10,10+1,1)) + list(range(50,500 + 1, 50))
+    #sample_sizes = range(10, 100 + 1, 10)
+    #sample_sizes = range(200, 900 + 1, 100)
+
+    def simbac(sample_size):
+        return simbac_run(sample_size, L)
+    def msp(sample_size):
+        return msp_run(samples=sample_size, sequence_length=L, population_size=1, recombination_rate=0, 
+gene_conversion_rate=0.015/2.0, gene_conversion_tract_length=500)
+
+    create_data(sample_sizes, [msp, simbac], ["msprime", "simbac"], "simbac")
+
+
+def ms_run(sample_size, L):
+    args = ms_style_args(sample_size, L)
+    exec_cli("./sims/ms", args)
+
+
+def simbac_run(sample_size, L):
+    ms_args = f"-N {sample_size} -B {int(L)}"
+    # using R=2*0.015 as gene conversion/recombination rate as SimBac uses R/2
+    simbac_args = "-R 0.03"
+    args = ms_args + " " + simbac_args
+    for _ in range(REPLICATES):
+        exec_cli("./sims/SimBac", args)
+
+
+def msp_run(**kwargs):
+    if "population_size" not in kwargs:
+        kwargs["population_size"] = Ne
+    if "recombination_rate" not in kwargs:
+        kwargs["recombination_rate"] = r
+    kwargs["num_replicates"] = REPLICATES
+
+    replicates = msprime.sim_ancestry(**kwargs)
+    for ts in replicates:
+        pass
+
+
+def exec_cli(command, args_str):
+    arg_list = [command] + args_str.split()
+    print(" ".join(arg_list))
+    subprocess.call(arg_list, stdout=subprocess.DEVNULL)
+
+
+def exec_jar(jar_path, sim_args_str):
[…]

### 2021-07-14 paper.tex
Updates for paper read-through.

--- 2021-07-14 paper.tex
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/recombination}
+\begin{center}
+\begin{tabular}{cc}
+\includegraphics[width=5cm]{msprime-scaling.png} &
+\includegraphics[width=5cm]{figures/recombination} \\
+\end{tabular}
+\end{center}
+% \includegraphics[width=\textwidth]{figures/recombination}
 \caption{\label{fig-recombination-perf}(A) Approximate run time for \msprime\
@@
 \begin{figure}
+\begin{center}
+\includegraphics[width=6cm]{simbac.png}
+\end{center}
 \caption{\label{fig-gc-perf}Comparison of the running time
@@
 See ~\cite{nelson2020accounting} for more details.
-We compare the performance of this simulation in Fig X
+We compare the performance of this simulation in Fig.~\ref{fig-dtwf-perf}
 with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=6cm]{msprime-argon-runtimes.png}
+\end{center}
+\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
+simulations with ARGON. To ensure the comparison with ARGON is
+as fair as possible we run it with a mutation rate of 0 and with
+the minimum output options (so that it writes out only an empty VCF file)}
+\end{figure}
 

### 2021-07-14 evaluation/recombination.py

@@ -44,9 +44,10 @@ def run_arg_sim():
 def plot_arg_data():
     df = pd.read_csv("data/arg.csv")
     print(df)
-    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
-    ax1.set_title("(A)")
-    ax1.set_xlabel("Scaled recombination rate")
+    # fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
+    fig, ax2 = plt.subplots(1, 1, figsize=(6, 6))
+    # ax1.set_title("(A)")
+    # ax1.set_xlabel("Scaled recombination rate")
 
     ax2.set_title("(B)")
     ax2.plot(df.L, df.arg_nodes, label="Fraction of ARG nodes")

### 2021-07-14 evaluation/recombination.py
Comment out png includes

@@ -44,10 +44,10 @@ def run_arg_sim():
 def plot_arg_data():
     df = pd.read_csv("data/arg.csv")
     print(df)
-    # fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
-    fig, ax2 = plt.subplots(1, 1, figsize=(6, 6))
-    # ax1.set_title("(A)")
-    # ax1.set_xlabel("Scaled recombination rate")
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
+    # fig, ax2 = plt.subplots(1, 1, figsize=(6, 6))
+    ax1.set_title("(A)")
+    ax1.set_xlabel("Scaled recombination rate")
 
     ax2.set_title("(B)")
     ax2.plot(df.L, df.arg_nodes, label="Fraction of ARG nodes")

### 2021-07-15 paper.tex
Comment out png includes
Point out the ancestry/mutation distinction
And outline the subsections below

@@
 
-In this section we describe the main features of \msprime\ 1.0.
+In this section we describe the main features of \msprime\ 1.0. Methodologically,
+there exists a logical separation between simulating coalescent genealogies and
+adding neutral mutations onto these genealogies. Simulating genetic ancestry
+requires methods that model the underlying reproductive process (e.g. Moran,
+Wright-Fisher, etc.) as well as accounting for recombination of various sorts
+(including gene conversion), demography, and selection. These methods are
+discussed in the first N subsections below. One of the strengths of the
+approach taken in msprime is that neutral mutations can be retrospectively
+added to simulated genealogies. Methods to generate mutations, and thereby
+generate genetic variation, are therefore discussed in a separate and subsequent
+subsection. Finally, we describe the \msprime\ API, which is designed to enforce
+the distinction between simulating genealogies and mutations; discuss the
+interoperability provided by the use of \tskit\, the tree sequence toolkit;
+and highlight the open source development process which allows for \msprime\
+to be extended by addition of new features while ensuring robustness and
+maintainability of the core codebase.
 
@@
 \begin{figure}
-\begin{center}
-\begin{tabular}{cc}
-\includegraphics[width=5cm]{msprime-scaling.png} &
-\includegraphics[width=5cm]{figures/recombination} \\
-\end{tabular}
-\end{center}
-% \includegraphics[width=\textwidth]{figures/recombination}
+% \begin{center}
+% \begin{tabular}{cc}
+% % \includegraphics[width=5cm]{msprime-scaling.png} &
+% % \includegraphics[width=5cm]{figures/recombination} \\
+% \end{tabular}
+% \end{center}
+\includegraphics[width=\textwidth]{figures/recombination}
 \caption{\label{fig-recombination-perf}(A) Approximate run time for \msprime\
@@
 \begin{center}
-\includegraphics[width=6cm]{simbac.png}
+% \includegraphics[width=6cm]{simbac.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=6cm]{msprime-argon-runtimes.png}
+% \includegraphics[width=6cm]{msprime-argon-runtimes.png}
 \end{center}

### 2021-07-15 paper.tex
Place bottlenecks+demography+multiple-mergers together
I think this is the subsection rearrangement that @JereKoskela suggested. Makes sense to me.

--- 2021-07-14 paper.tex
@@
 
+\subsection*{Instantaneous bottlenecks}
+
+A common approach to modelling the effect of demographic history on genealogies
+is to assume that effective population size ($N_e$) changes in discrete steps
+which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
+keightley2007joint,li2011inference}. In this setting of piecewise constant
+$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
+reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
+recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
+bottlenecks are short both on the timescale of coalescence and mutations,
+there may be little information about the duration of a bottleneck
+($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
+assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
+and generate a sudden burst of coalescence events (a multiple merger event) in
+the genealogy. The strength of the bottleneck $B$ can be thought of as an
+(imaginary) time period during which coalescence events are collapsed,
+i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
+single pair of lineages entering the bottleneck coalesce during the bottleneck
+is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
+attractive and both analytic results and empirical
+inference~\citep{griffiths1994sampling, galtier2000detecting,
+bunnefeld2015inferring} have been developed under this model, there has
+been no software available to simulate data under instantaneous
+bottleneck histories.
+
+We have implemented instantaneous bottlenecks in \msprime~1.0
+using a variant of Hudson's linear time single-locus coalescent
+algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
+by adding events to the Demography object (see the Demography section)
+and can be used in combination with any other demographic modelling
+features. We have validated the results of these simulations by comparing
+against analytical expectations for coalescence times and the
+site frequency spectrum~\citep{bunnefeld2015inferring}.
+
+\subsection*{Multiple merger coalescents}
+
+Kingman's coalescent assumes that only two ancestral lineages can merge at
+each merger event. Although this is generally a reasonable approximation there
+are certain situations in which the underlying mathematical assumptions are
+violated. For example in certain highly fecund organisms
+\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
+to produce numbers of offspring on the order of the population size and
+therefore a few individuals may produce the bulk of the offspring in any given
+generation~\citep{hedgecock_94}. These population dynamics violate basic
+assumptions of the Kingman coalescent, and are better modelled by
+`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
+two lineages can merge in a given event. Multiple-merger coalescent processes
+have also been shown to be relevant for modeling the effects of selection on
+gene genealogies \citep{Gillespie909,DS04}.
+
+Although multiple merger coalescents have been of significant theoretical
+interest for around two decades, there has been little practical software
+available to simulate these models.
+\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
+simulate a related spatial continuum model~\citep{barton2010new},
+\cite{zhu2015hybrid} simulate genealogies within a species tree
+based on a multiple-merger model, and
+\cite{becheler2020occupancy} provide a general method for simulating
+multiple merger processes
+as part of the Quetzal framework~\citep{becheler2019quetzal}.
+The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
+includes a number of extensions to the Beta-coalescent.
+None of these methods work with large genomes, and very little work
+has been performed on simulating multiple merger processes with recombination.
+
+We have added two multiple merger coalescent models in \msprime\ 1.0, the
+Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
+such models with recombination efficiently for the first time.
+These simulation models have been extensively validated against
+analytical results from the site frequency
+spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
+ as well as more general properties of coalescent processes.
+Please see the Appendix for more details and model derivations.
+
 \subsection*{Selection}
@@
 
-\subsection*{Multiple merger coalescents}
-
-Kingman's coalescent assumes that only two ancestral lineages can merge at
-each merger event. Although this is generally a reasonable approximation there
-are certain situations in which the underlying mathematical assumptions are
-violated. For example in certain highly fecund organisms
-\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
-to produce numbers of offspring on the order of the population size and
-therefore a few individuals may produce the bulk of the offspring in any given
-generation~\citep{hedgecock_94}. These population dynamics violate basic
-assumptions of the Kingman coalescent, and are better modelled by
-`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
-two lineages can merge in a given event. Multiple-merger coalescent processes
-have also been shown to be relevant for modeling the effects of selection on
-gene genealogies \citep{Gillespie909,DS04}.
-
-Although multiple merger coalescents have been of significant theoretical
-interest for around two decades, there has been little practical software
-available to simulate these models.
-\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
-simulate a related spatial continuum model~\citep{barton2010new},
-\cite{zhu2015hybrid} simulate genealogies within a species tree
-based on a multiple-merger model, and
-\cite{becheler2020occupancy} provide a general method for simulating
-multiple merger processes
-as part of the Quetzal framework~\citep{becheler2019quetzal}.
-The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
-includes a number of extensions to the Beta-coalescent.
-None of these methods work with large genomes, and very little work
-has been performed on simulating multiple merger processes with recombination.
-
-We have added two multiple merger coalescent models in \msprime\ 1.0, the
-Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
-such models with recombination efficiently for the first time.
-These simulation models have been extensively validated against
-analytical results from the site frequency
-spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
- as well as more general properties of coalescent processes.
-Please see the Appendix for more details and model derivations.
-
-\subsection*{Instantaneous bottlenecks}
-
-A common approach to modelling the effect of demographic history on genealogies
-is to assume that effective population size ($N_e$) changes in discrete steps
-which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
-keightley2007joint,li2011inference}. In this setting of piecewise constant
-$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
-reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
-recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
-bottlenecks are short both on the timescale of coalescence and mutations,
-there may be little information about the duration of a bottleneck
-($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
-assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
-and generate a sudden burst of coalescence events (a multiple merger event) in
-the genealogy. The strength of the bottleneck $B$ can be thought of as an
-(imaginary) time period during which coalescence events are collapsed,
-i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
-single pair of lineages entering the bottleneck coalesce during the bottleneck
-is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
-attractive and both analytic results and empirical
-inference~\citep{griffiths1994sampling, galtier2000detecting,
-bunnefeld2015inferring} have been developed under this model, there has
-been no software available to simulate data under instantaneous
-bottleneck histories.
-
-We have implemented instantaneous bottlenecks in \msprime~1.0
-using a variant of Hudson's linear time single-locus coalescent
-algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
-by adding events to the Demography object (see the Demography section)
-and can be used in combination with any other demographic modelling
-features. We have validated the results of these simulations by comparing
-against analytical expectations for coalescence times and the
-site frequency spectrum~\citep{bunnefeld2015inferring}.
-
 \subsection*{Simulation interface}
+
 The majority of simulation packages are controlled either through

### 2021-07-15 paper.tex
make refs appear

@@
 
-\title{msprime 1.0: an efficient and extensible coalescent simulation framework}
+\title{Msprime 1.0: an efficient and extensible coalescent simulation framework}
 \author{Author list to be filled in
@@
 Hudson's \ms\ optionally includes gene conversion and so does
-MaCS~\citep{chen2009fast}, while scrm~\citep{staab2014scrm} does not
+MaCS~\citep{chen2009fast}, while scrm~\citep{staab2015scrm} does not
 support gene conversions.

### 2021-07-15 paper.tex

--- 2021-07-15 paper.tex
@@
 
+\usepackage[]{hyperref}
+\hypersetup{
+    pdftitle={Msprime 1.0: an efficient and extensible coalescent simulation framework},
+    pdfauthor={Msprime developers},
+    pdfsubject={Coalescent simulation},
+    pdfkeywords={coalescent, mutation, simulation, genetic ancestry, genealogy, evolution},
+    bookmarksnumbered=true,     
+    bookmarksopen=true,
+    bookmarksopenlevel=3,
+    colorlinks=true,
+    pdfstartview=Fit,
+    pdfpagemode=UseOutlines,
+    pdfpagelayout=TwoPageRight
+}
+
 % local definitions
@@
 
+\setcounter{secnumdepth}{0} %Do not print out section numbers
+
 \begin{document}
 
-\title{msprime 1.0: an efficient and extensible coalescent simulation framework}
+\title{Msprime 1.0: an efficient and extensible coalescent simulation framework}
 \author{Author list to be filled in
@@
 
-\section*{Introduction}
+\section{Introduction}
 
@@
 
-\section*{Methods}
+\section{Methods}
 
@@
 
-\subsection*{Recombination}
+\subsection{Recombination}
 
@@
 
-\subsection*{Ancestral recombination graphs}
+\subsection{Ancestral recombination graphs}
 
@@
 
-\subsection*{Gene conversion}
+\subsection{Gene conversion}
 
@@
 
-\subsection*{Demography}
+\subsection{Demography}
 % TODO should we put in some citations here for the models? Hardly seems
@@
 
-\subsection*{Instantaneous bottlenecks}
+\subsection{Instantaneous bottlenecks}
 
@@
 
-\subsection*{Multiple merger coalescents}
+\subsection{Multiple merger coalescents}
 
@@
 
-\subsection*{Selection}
+\subsection{Selection}
 Selection was added to the coalescent ... Lots of simulators exist:
@@
 
-\subsection*{Discrete time Wright-Fisher}
+\subsection{Discrete time Wright-Fisher}
 The coalescent is a good approximation for ancient history, but
@@
 
-\subsection*{Finite sites mutations}
+\subsection{Finite sites mutations}
 Version 0.x of \texttt{msprime} provided a very simple mutation model,
@@
 
-\subsection*{Simulation interface}
+\subsection{Simulation interface}
 
@@
 
-\subsection*{Data interchange and interoperability}
+\subsection{Data interchange and interoperability}
 The point of simulations is to generate data and the efficiency with
@@
 
-\subsection*{Development model}
+\subsection{Development model}
 The community development process for \msprime\ is enabled by following an
@@
 
-\section*{Results and Discussion}
-
-\subsection*{Discussion}
+\section{Results and Discussion}
+
+\subsection{Discussion}
 % Simulation is useful, and it's really easy for a single locus.
@@
 
-\section*{Acknowledgments}
+\section{Acknowledgments}
 JK is supported by the Robertson Foundation.
@@
 }
+
+\setcounter{secnumdepth}{2} % Print out appendix section numbers
 

### 2021-07-15 figures/gene_conversion/plot_performance_gene_conversion.R
include gc performance figure into Makefile
Remove simbac

@@ -3,12 +3,12 @@
 # Rscript plot_performance_gene_conversion.R
 
 
-df = read.csv(file = "../../data/simbac.csv")
+df = read.csv(file = "./data/simbac.csv")
 
 library(ggplot2)
 
 
-pdf(file = "gc_perf.pdf", width = 6, height = 3)
+pdf(file = "./figures/gene_conversion/gc_perf.pdf", width = 6, height = 3)
 ggplot(data = df, aes(x = sample_size, y = time, color = tool)) +
   geom_point() + 
   labs(x = "sample size") +
@@ -22,8 +22,8 @@ ggplot(data = df, aes(x = sample_size, y = time, color = tool)) +
 dev.off()
 
 
-pdf(file = "gc_perf_log.pdf", width = 6, height = 3)
-ggplot(data = df_long, aes(x = sample_size, y = time, color = tool)) +
+pdf(file = "./figures/gene_conversion/gc_perf_log.pdf", width = 6, height = 3)
+ggplot(data = df, aes(x = sample_size, y = time, color = tool)) +
   geom_point() +
   labs(x = "sample size") +
   xlim(0,500) +

### 2021-07-15 paper.tex
Update
Remove hyperref
It turns the page into a horrible looking mess.

--- 2021-07-15 paper.tex
@@
 
-% \section{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}
+% \section*{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}
 
 \maketitle
-
-% Abstract: ~ 150 words.  Concise summary of the results.
-
-% Key points in abstract.
-% 1) Coalescent simulation is a vital tool
-% 2) Coalescent with recombination is hard
-% 3) Large sample sizes are crucial today
-% 4) Msprime has lots of features and uses a community development model
 
@@
 
-% Keywords: For software, include the relevant models, algorithms and language;
 \textbf{Keywords:} Coalescent simulation, Python
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Introduction: Limit of 500 words. The introduction should describe the
-% significance of the software or resource presented, discuss novelty, provide an
-% overview of the software or resource, and describe how researchers can access it.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Methods: For software, the authors should describe key elements of the algorithms
-% implemented, and are expected to provide both the source code as well as
-% examples illustrating the use of the software.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Methods}
+\section*{Results}
 
@@
 Gene conversion is a unidirectional transfer of a short segment of genetic material,
-for example between homologous chromosomes~\citep{Chen2007}.
+for example between homologous chromosomes~\citep{chen2007gene}.
 %description of ancestral process
@@
 typically below 1kb, and thus affects linkage pattern differently than
-recombination~\citep{Korunes2017}.
+recombination~\citep{korunes2017gene}.
 
@@
 This includes the backward simulator SimBac~\citep{brown2016simbac},
-the SMC based approach used in FastSimBac~\citep{DeMaio2017},
-and forward in time tools within the SLiM framework~\citep{Cury2020}.
+the SMC based approach used in FastSimBac~\citep{demaio2017the},
+and forward in time tools within the SLiM framework~\citep{cury2020simulation}.
 
@@
 \begin{figure}
-\includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
+% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
 % \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
@@
 $N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
-with mean tract length $L = 542$~\citep{Lapierre2016}.}
+with mean tract length $L = 542$~\citep{lapierre2016the}.}
 \end{figure}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Results and Discussion: For software, the authors are expected to present a
-% few examples that demonstrate the use of the software and benchmarks. Ideally,
-% the examples should be simple. We encourage the use of boxes with scripts
-% supported by a narrative. Benchmarks carried under more realistic situations
-% than the one presented in the examples are also encouraged.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Results and Discussion}
-
-\subsection*{Discussion}
+\section*{Discussion}
 % Simulation is useful, and it's really easy for a single locus.
@@
 
+\setcounter{secnumdepth}{2} % Print out appendix section numbers
+
 \appendix
@@
 \label{app-likelihoods}
-\section{Likelihood calculations}
+\section*{Likelihood calculations}
 
@@
 \label{app-multiple-mergers}
-\section{Multiple merger coalescent model}
+\section*{Multiple merger coalescent model}
 

### 2021-07-15 evaluation/mutations.py
Add mutations benchmark.

@@ -1,0 +1,176 @@
+import time
+import pathlib
+
+import matplotlib.pyplot as plt
+import pandas as pd
+import numpy as np
+import click
+import pyvolve
+import tskit
+import msprime
+import cpuinfo
+
+
+@click.group()
+def cli():
+    pass
+
+
+@click.command()
+@click.argument("tree_file", type=click.Path())
+def benchmark_pyvolve(tree_file):
+    click.echo("Benchmarking pyvolve")
+
+    ts = tskit.load(tree_file)
+    tree = ts.first()
+    # node_labels = {u: str(u) for u in ts.samples()}
+    newick = tree.newick()  # node_labels=node_labels)
+    # Not entirely sure this is the correct rate, but pyvolve won't tell
+    # me how many mutations it has generated, so it's not simple.
+    pyvolve_tree = pyvolve.read_tree(tree=newick, scale_tree=1)
+    pyvolve_model = pyvolve.Partition(
+        models=pyvolve.Model("nucleotide"), size=int(ts.sequence_length)
+    )
+    print("starting sim")
+    sim = pyvolve.Evolver(tree=pyvolve_tree, partitions=pyvolve_model)
+    before = time.perf_counter()
+    sim(ratefile=None, infofile=None, seqfile=None)
+    duration = time.perf_counter() - before
+    print("Ran in duration ", duration)
+    # print(sim)
+    # seqs = sim.get_sequences(anc=True)  # seq-dict is sorted in pre-order
+
+
+@click.command()
+@click.argument("tree_file", type=click.Path())
+def benchmark_msprime(tree_file):
+    click.echo("Benchmarking msprime")
+    ts = tskit.load(tree_file)
+    before = time.perf_counter()
+    ts = msprime.sim_mutations(ts, rate=1, random_seed=42)
+    duration = time.perf_counter() - before
+    print("Ran in duration ", duration)
+    print("simulated ", ts.num_mutations, "mutations at ", ts.num_sites, "sites")
+    ts.dump("tmp.trees")
+
+
+@click.command()
+@click.argument("tree_file", type=click.Path())
+def convert_newick(tree_file):
+    ts = tskit.load(tree_file)
+    tree = ts.first()
+    newick = tree.newick()
+    print(newick)
+
+
+def generate_tree(n, L, random_seed=42):
+    print(f"Simulate = n {n} L = {L}Mb")
+    ts = msprime.sim_ancestry(
+        n,
+        population_size=10 ** 4,
+        ploidy=1,
[…]

### 2021-07-15 paper.tex
Add figure and a little text.

@@
 
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/mutation_perf}
+\caption{\label{fig-mutation-perf} Time required to run
+\texttt{sim\_mutations} on tree sequences generated
+by \texttt{sim\_ancestry} (with a population size of $10^4$
+and recombination rate of $10^{-8}$) for varying (haploid) sample
+size and sequence length. We run 10 replicate mutation simulations
+each for three different mutation rates, and report the average
+CPU time required (Intel Core i7-9700).
+(A) Holding sequence length fixed at 10 megabases and varying the
+number of samples (tree tips) from 10 to 100,000.
+(B) Holding number of samples fixed at 1000, and varying the sequence
+length from 1 to 100 megabases.}
+\end{figure}
+
 Comparing the performance of mutation generation methods is problematic.
@@
 
-The mutation generation machinery in \msprime\ is extremely
-efficient. For example, we simulated mutations under the JC69
+Fig.~\ref{fig-mutation-perf} shows that the mutation generation
+machinery in \msprime\ is extremely efficient, requiring
+fractions of a second to generate mutations for most simulations.
+For example, the longest running simulation in
+Fig.~\ref{fig-mutation-perf}B required less than 2 seconds to
+generate an average of 1.5 million mutations over 137,081 trees
+in a tree sequence with 508,125 edges.
+Generating mutations for a tree sequence is particularly efficient
+because we can work edge-by-edge rather than having to
+consider each tree independently.
+
+\jkcomment{We probably want to keep a version of this narrative too.}
+For example, we simulated mutations under the JC69
 model for a tree with $10^6$ leaves over $10^4$ sites (resulting

### 2021-07-15 evaluation/mutations.py

@@ -151,14 +151,16 @@ def plot():
         dfr = dfL[dfL.rate == rate].sort_values("n")
         ax1.plot(dfr.n, dfr.time, label=f"rate={rate}")
 
+
     ax2.set_title("(B)")
     ax2.set_xlabel("Sequence length (Megabases)")
     ax1.set_ylabel("Time (seconds)")
 
-    dfn = df[df.n == 1000]
+    dfn = df[df.n == 1000].sort_values("L")
     for rate in rates:
-        dfr = dfn[dfn.rate == rate].sort_values("L")
+        dfr = dfn[dfn.rate == rate]
         ax2.plot(dfr.L, dfr.time, label=f"rate={rate}")
+    print(dfn[dfn.L == 100])
 
     ax1.legend()
     plt.savefig("figures/mutation_perf.png")

### 2021-07-15 paper.tex
Update

@@
 
-\jkcomment{If we're going to have this paragraph of ARG bashing then
-I guess we'll have to do more than just single out Iain and Aylwyn.}
 However, this original definition of the ARG in terms of a stochastic process
@@
 \cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material. Such
+individual's pedigree from which they have inherited ancestral material.
+Others state that an ARG is equivalent to the sequence of marginal
+% Also SARGE paper
+trees along the genome~\citep{mirzaei2016rent}
+
+Such
 imprecision is unfortunate, since there are any number of graphical encodings of
@@
 \begin{figure}
-% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
-% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
+\begin{center}
+\includegraphics[width=7cm]{figures/gc-perf}
+\end{center}
 \caption{\label{fig-gc-perf}Comparison of the running time
@@
 \subsection*{Discrete time Wright-Fisher}
-The coalescent is a good approximation for ancient history, but
-is bad for the recent past and in particular when we have large
-sample
+
+The assumptions of the classical coalescent model are that both sample size and
+the simulated sequence length are small. As modern datasets often include
+whole-genome sequences for tens or hundreds of thousands of samples, simulating
+such data requires breaking these assumptions. While the coalescent remains a
+good approximation to the Wright-Fisher model in the distant past, it can
+exhibit large biases in the recent history. Identity-by-descent (IBD),
+long-range linkage disequilibrium (LD), and ancestry patterns deviate from
+Wright-Fisher expectations, and this bias grows considerably when simulating
+larger sample
 sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
-Describe basic
-approach [DOM, can you fill in please].
-See ~\cite{nelson2020accounting} for more details.
-We compare the performance of this simulation in Fig.~\ref{fig-dtwf-perf}
-with ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
+
+A backwards-in-time Wright-Fisher model is supported in \msprime, which is
+shown to properly capture the expected distributions of IBD and long-range LD,
+as well as ancestry patterns in admixed
+populations~\citep{nelson2020accounting}. In short, the classical coalescent
+draws recombination events from exponential waiting times, which are continuous
+in time, and each recombination then splits a segment into two distinct
+ancestors. For large sequence lengths, this can dramatically overestimate the
+number of genealogical ancestors of a given sample. Instead, the Wright-Fisher
+model iterates through discrete generations so that each gamete has a unique
+diploid parent, and multiple recombinations within a generation results in
+crossover events between the same two parental haploid copies.  This method is
+described in more detail in~\cite{nelson2020accounting}.
 
@@
 
-We support changing the simulation model, so that we can easily
-run hybrid simulations like described by~\cite{bhaskar2014distortion}
-This ia a powerful approach because it lets us combine any
-number of different simulation models.
+We support changing the simulation model at specified times so that we can
+easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}. We
+can combine any number of different simulation models, allowing for the
+flexible choice of simulation scenarios without severely damaging efficiency.
+As the discrete time Wright-Fisher model improves accuracy of genealogical
+patterns in the recent past, we can simulate the recent history using this
+model and then switch to the standard coalescent to more efficiently simulate
+the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the perfomance of
+this approach with ARGON~\citep{palamara2016argon}, a specialised DTWF
+simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
+simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
+hybrid approach provides large efficiency gains.
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/mutation_perf}
+\includegraphics[width=\textwidth]{figures/mutation-perf}
 \caption{\label{fig-mutation-perf} Time required to run
@@
 speidel2019method,tang2019genealogy}
+% also SARGE
 opens the possibility for new applications of coalescent simulations.
+With lots of applications, e.g., [cite Osmond and Coop].
 Also \stdpopsim\ is cool~\citep{adrion2020community}

### 2021-07-15 paper.tex

@@
 
+\jkcomment{If we're going to have this paragraph of ARG bashing then
+I guess we'll have to do more than just single out Iain and Aylwyn.}
 However, this original definition of the ARG in terms of a stochastic process
@@
 \cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material.
-Others state that an ARG is equivalent to the sequence of marginal
-% Also SARGE paper
-trees along the genome~\citep{mirzaei2016rent}
-
-Such
+individual's pedigree from which they have inherited ancestral material. Such
 imprecision is unfortunate, since there are any number of graphical encodings of
@@
 \begin{figure}
-\begin{center}
-\includegraphics[width=7cm]{figures/gc-perf}
-\end{center}
+% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
+% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
 \caption{\label{fig-gc-perf}Comparison of the running time
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{figures/mutation-perf}
-\caption{\label{fig-mutation-perf} Time required to run
-\texttt{sim\_mutations} on tree sequences generated
-by \texttt{sim\_ancestry} (with a population size of $10^4$
-and recombination rate of $10^{-8}$) for varying (haploid) sample
-size and sequence length. We run 10 replicate mutation simulations
-each for three different mutation rates, and report the average
-CPU time required (Intel Core i7-9700).
-(A) Holding sequence length fixed at 10 megabases and varying the
-number of samples (tree tips) from 10 to 100,000.
-(B) Holding number of samples fixed at 1000, and varying the sequence
-length from 1 to 100 megabases.}
-\end{figure}
-
 Comparing the performance of mutation generation methods is problematic.
@@
 
-Fig.~\ref{fig-mutation-perf} shows that the mutation generation
-machinery in \msprime\ is extremely efficient, requiring
-fractions of a second to generate mutations for most simulations.
-For example, the longest running simulation in
-Fig.~\ref{fig-mutation-perf}B required less than 2 seconds to
-generate an average of 1.5 million mutations over 137,081 trees
-in a tree sequence with 508,125 edges.
-Generating mutations for a tree sequence is particularly efficient
-because we can work edge-by-edge rather than having to
-consider each tree independently.
-
-\jkcomment{We probably want to keep a version of this narrative too.}
-For example, we simulated mutations under the JC69
+The mutation generation machinery in \msprime\ is extremely
+efficient. For example, we simulated mutations under the JC69
 model for a tree with $10^6$ leaves over $10^4$ sites (resulting
@@
 speidel2019method,tang2019genealogy}
-% also SARGE
 opens the possibility for new applications of coalescent simulations.
-With lots of applications, e.g., [cite Osmond and Coop].
 Also \stdpopsim\ is cool~\citep{adrion2020community}

### 2021-07-16 paper.tex

--- 2021-07-15 paper.tex
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/mutation-perf}
+\includegraphics[width=\textwidth]{figures/mutation_perf}
 \caption{\label{fig-mutation-perf} Time required to run

### 2021-07-16 evaluation/plot.py
Make new plot script and remove R code for gc

@@ -1,0 +1,75 @@
+import time
+import pathlib
+
+import matplotlib.pyplot as plt
+import pandas as pd
+import numpy as np
+import click
+import pyvolve
+import tskit
+import msprime
+import cpuinfo
+
+
+@click.group()
+def cli():
+    pass
+
+
+@click.command()
+def mutation_perf():
+    """
+    Plot the mutations benchmark.
+    """
+    df = pd.read_csv("data/mutations.csv")
+    df.L /= 1e6
+    rates = np.unique(df.rate)
+
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6), sharey=True)
+    ax1.set_title("(A)")
+    ax1.set_xlabel("Sample size")
+    dfL = df[df.L == 10]
+    for rate in rates:
+        dfr = dfL[dfL.rate == rate].sort_values("n")
+        ax1.plot(dfr.n, dfr.time, label=f"rate={rate}")
+
+    ax2.set_title("(B)")
+    ax2.set_xlabel("Sequence length (Megabases)")
+    ax1.set_ylabel("Time (seconds)")
+
+    dfn = df[df.n == 1000].sort_values("L")
+    for rate in rates:
+        dfr = dfn[dfn.rate == rate]
+        ax2.plot(dfr.L, dfr.time, label=f"rate={rate}")
+    print(dfn[dfn.L == 100])
+
+    ax1.legend()
+    plt.savefig("figures/mutation-perf.png")
+    plt.savefig("figures/mutation-perf.pdf")
+
+
+@click.command()
+def gc_perf():
+    """
+    Plot the gene-conversion benchmark.
+    """
+    df = pd.read_csv("data/gc.csv")
+
+    fig, ax1 = plt.subplots(1, 1)
+    for tool in set(df.tool):
+        dft = df[df.tool == tool]
+        ax1.plot(dft.sample_size, dft.time, label=tool)
+
+    ax1.set_xlabel("Sample size")
+    ax1.set_ylabel("Time (seconds)")
+
+    ax1.legend()
+    plt.savefig("figures/gc-perf.png")
+    plt.savefig("figures/gc-perf.pdf")
+
+
+cli.add_command(mutation_perf)
+cli.add_command(gc_perf)
+
+if __name__ == "__main__":
+    cli()

### 2021-07-16 paper.tex
More systematic treatment of data files.

--- 2021-07-15 paper.tex
@@
 % \end{center}
-\includegraphics[width=\textwidth]{figures/recombination}
-\caption{\label{fig-recombination-perf}(A) Approximate run time for \msprime\
+% \includegraphics[width=\textwidth]{figures/recombination}
+\Large{Plot of msprime run time against $\rho$}
+\caption{\label{fig-recombination-perf} Approximate run time for \msprime\
 simulations as a function of scaled recombination rate.
-(B) Comparison of the ARG simulations vs minimal tree sequence representation.}
+}
 \end{figure}
@@
 simulation of organisms such as \textit{Drosophila melagaster}.
-
-\subsection*{Ancestral recombination graphs}
-
-The Ancestral Recombination Graph (ARG) was introduced by
-Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
-representation of the coalescent with recombination stochastic process. This
-formulation was complementary to Hudson's earlier
-work~\citep{hudson1983properties}, and substantially increased our theoretical
-understanding of recombination. In Griffiths' ARG formulation, we
-regard a realisation of the coalescent with recombination as a graph in which
-vertices represent common ancestor or recombination events, and edges represent
-lineages through which ancestral material flows.
-We may consider the big ARG, in which we track lineages arising out of
-recombinations regardless of whether they carry ancestral material and
-consequently wait an exponentially large time until all sampled lineages
-coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
-which we only track genetic ancestors, and wait until all marginal trees
-have fully coalesced (like Hudson's algorithm).
-
-However, this original definition of the ARG in terms of a stochastic process
-is rarely what is meant when the term is currently used. The term has come to
-essentially mean any representation of genetic ancestry in the presence
-of recombination as a graph-like structure; for example,
-\cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material.
-Others state that an ARG is equivalent to the sequence of marginal
-% Also SARGE paper
-trees along the genome~\citep{mirzaei2016rent}
-
-Such
-imprecision is unfortunate, since there are any number of graphical encodings of
-genetic ancestry, and referring to the original Griffiths papers does little to
-clarify. As we see below, there are situations in which the precise information
-stored in the graph is important, and conflating a high-level
-discussion of genetic ancestry with this mathematical object is not
-helpful. The point of technical terms is to provide clarity and
-precision on what is being discussed, and it is now essentially
-impossible to know what is meant by the term ``ARG'';
-we therefore use the term ``genetic ancestry'' in high-level
-discussions.
-
-For our purposes, an ARG is a realisation of the coalescent with
-recombination, in the Griffiths (little ARG) sense.
-As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
-works by dynamically generating and traversing a little ARG.
-The graph is not explicitly represented in memory, but is partially
-present through the extant lineages and the ancestral material they carry
-over time. Since the ARG is a large and unwieldy object
-(Fig.~\ref{fig-recombination-perf}B) we do not use it as output, but
-rather store the information required to recover the genealogical
-history as nodes and edges in a tree sequence (see section XX),
-or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
-
-Although the tree sequences output by \msprime\ contain full information
-about the genealogical trees, their correlation structure along the genome,
-and the ancestral haplotypes on which coalescences occured, some information
-is lost in this mapping down from ARG space to the minimal tree sequence
-form. In particular, we lose
-information about ancestral haplotypes that were common ancestors but
-in which no coalescences occured, and also information about the precise time
-and genomic location of recombination events. In most cases, such
-information is of little relevance as it is in principle unknowable,
-but there are occasions such as visualisation or computing likelihoods (see
-below) in which is useful. We therefore provide
-the \texttt{record\_full\_arg} option in \msprime\
-to store a representation of the complete ARG traversed during simulation.
-This is done by storing extra nodes (marked with specific flags, so they
-can be easily identified later) and edges in the tree sequence.
-Fig.~\ref{fig-recombination-perf}B shows that this extra ARG information
-quickly comes to dominate the output as we increase the sequence length,
-and the resulting tree sequence files are many times larger
-than the standard minimal tree sequence representation.
-
-One situation in which a record of the full ARG is necessary is when we
-wish to compute likelihoods during inference.
-The likelihood is a central quantity in evaluating the plausibility of a putative
-ancestry as an explanation of DNA sequence data, both directly through
-e.g.~approaches based on maximum likelihood, and as an ingredient of
-methods such as
-Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
-We provide functions to compute the likelihood of ARG realisations
-(encoded using the representation of the \texttt{record\_full\_arg} option)
-and mutational patterns under
-the standard coalescent and infinite sites mutation model.
-See the Appendix for details.
 
@@
 Please see the Appendix for more details and model derivations.
+
+\subsection*{Ancestral recombination graphs}
+
+The Ancestral Recombination Graph (ARG) was introduced by
+Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
+representation of the coalescent with recombination stochastic process. This
+formulation was complementary to Hudson's earlier
+work~\citep{hudson1983properties}, and substantially increased our theoretical
+understanding of recombination. In Griffiths' ARG formulation, we
+regard a realisation of the coalescent with recombination as a graph in which
+vertices represent common ancestor or recombination events, and edges represent
+lineages through which ancestral material flows.
+We may consider the big ARG, in which we track lineages arising out of
+recombinations regardless of whether they carry ancestral material and
+consequently wait an exponentially large time until all sampled lineages
+coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
+which we only track genetic ancestors, and wait until all marginal trees
+have fully coalesced (like Hudson's algorithm).
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=7cm]{figures/arg}
+\end{center}
+\caption{\label{fig-arg} A figure illustrating the relative size of
+the full ARG representation compared the minimal tree sequence.
+\jkcomment{This figure is quite preliminary and we need to improve it.}}
+\end{figure}
+
+However, this original definition of the ARG in terms of a stochastic process
+is rarely what is meant when the term is currently used. The term has come to
+essentially mean any representation of genetic ancestry in the presence
+of recombination as a graph-like structure; for example,
+\cite{mathieson2020ancestry} informally define an ARG as the subset of an
+individual's pedigree from which they have inherited ancestral material.
+Others state that an ARG is equivalent to the sequence of marginal
+% Also SARGE paper
+trees along the genome~\citep{mirzaei2016rent}
+
+Such
+imprecision is unfortunate, since there are any number of graphical encodings of
+genetic ancestry, and referring to the original Griffiths papers does little to
+clarify. As we see below, there are situations in which the precise information
+stored in the graph is important, and conflating a high-level
+discussion of genetic ancestry with this mathematical object is not
+helpful. The point of technical terms is to provide clarity and
+precision on what is being discussed, and it is now essentially
+impossible to know what is meant by the term ``ARG'';
+we therefore use the term ``genetic ancestry'' in high-level
+discussions.
+
+For our purposes, an ARG is a realisation of the coalescent with
+recombination, in the Griffiths (little ARG) sense.
+As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
+works by dynamically generating and traversing a little ARG.
+The graph is not explicitly represented in memory, but is partially
+present through the extant lineages and the ancestral material they carry
+over time. Since the ARG is a large and unwieldy object
+(Fig.~\ref{fig-arg}) we do not use it as output, but
+rather store the information required to recover the genealogical
+history as nodes and edges in a tree sequence (see section XX),
+or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
+
+Although the tree sequences output by \msprime\ contain full information
+about the genealogical trees, their correlation structure along the genome,
+and the ancestral haplotypes on which coalescences occured, some information
+is lost in this mapping down from ARG space to the minimal tree sequence
+form. In particular, we lose
+information about ancestral haplotypes that were common ancestors but
+in which no coalescences occured, and also information about the precise time
+and genomic location of recombination events. In most cases, such
+information is of little relevance as it is in principle unknowable,
+but there are occasions such as visualisation or computing likelihoods (see
+below) in which is useful. We therefore provide
+the \texttt{record\_full\_arg} option in \msprime\
+to store a representation of the complete ARG traversed during simulation.
+This is done by storing extra nodes (marked with specific flags, so they
+can be easily identified later) and edges in the tree sequence.
+Fig.~\ref{fig-arg} shows that this extra ARG information
+quickly comes to dominate the output as we increase the sequence length,
+and the resulting tree sequence files are many times larger
+than the standard minimal tree sequence representation.
+
+One situation in which a record of the full ARG is necessary is when we
+wish to compute likelihoods during inference.
+The likelihood is a central quantity in evaluating the plausibility of a putative
+ancestry as an explanation of DNA sequence data, both directly through
+e.g.~approaches based on maximum likelihood, and as an ingredient of
+methods such as
+Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
+We provide functions to compute the likelihood of ARG realisations
+(encoded using the representation of the \texttt{record\_full\_arg} option)
+and mutational patterns under
+the standard coalescent and infinite sites mutation model.
+See the Appendix for details.
 

### 2021-07-16 evaluation/plot.py

@@ -15,6 +15,9 @@ import cpuinfo
 def cli():
     pass
 
+def save(name):
+    plt.savefig(f"figures/{name}.png")
+    plt.savefig(f"figures/{name}.pdf")
 
 @click.command()
 def mutation_perf():
@@ -44,8 +47,7 @@ def mutation_perf():
     print(dfn[dfn.L == 100])
 
     ax1.legend()
-    plt.savefig("figures/mutation-perf.png")
-    plt.savefig("figures/mutation-perf.pdf")
+    save("mutation-perf")
 
 
 @click.command()
@@ -64,12 +66,27 @@ def gc_perf():
     ax1.set_ylabel("Time (seconds)")
 
     ax1.legend()
-    plt.savefig("figures/gc-perf.png")
-    plt.savefig("figures/gc-perf.pdf")
+    save("gc-perf")
+
+
+@click.command()
+def arg():
+    """
+    Plot the ARG size illustration figure.
+    """
+    df = pd.read_csv("data/arg.csv")
+
+    fig, ax1 = plt.subplots(1, 1)
+    ax1.plot(df.L, df.arg_nodes, label="Fraction of ARG nodes")
+    ax1.plot(df.L, df.size_ratio, label="size(tree sequence) / size(ARG)")
+    ax1.set_xlabel("Sequence length")
+    ax1.legend()
+    save("arg")
 
 
 cli.add_command(mutation_perf)
 cli.add_command(gc_perf)
+cli.add_command(arg)
 
 if __name__ == "__main__":
     cli()

### 2021-07-16 paper.tex
Systematise data/plots

@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/mutation-perf}
-\caption{\label{fig-mutation-perf} Time required to run
+\includegraphics[width=\textwidth]{figures/mutations-perf}
+\caption{\label{fig-mutations-perf} Time required to run
 \texttt{sim\_mutations} on tree sequences generated
@@
 
-Fig.~\ref{fig-mutation-perf} shows that the mutation generation
+Fig.~\ref{fig-mutations-perf} shows that the mutation generation
 machinery in \msprime\ is extremely efficient, requiring
@@
 For example, the longest running simulation in
-Fig.~\ref{fig-mutation-perf}B required less than 2 seconds to
+Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
 generate an average of 1.5 million mutations over 137,081 trees

### 2021-07-16 evaluation/plot.py

@@ -20,11 +20,11 @@ def save(name):
     plt.savefig(f"figures/{name}.pdf")
 
 @click.command()
-def mutation_perf():
+def mutations_perf():
     """
     Plot the mutations benchmark.
     """
-    df = pd.read_csv("data/mutations.csv")
+    df = pd.read_csv("data/mutations_perf.csv")
     df.L /= 1e6
     rates = np.unique(df.rate)
 
@@ -47,7 +47,7 @@ def mutation_perf():
     print(dfn[dfn.L == 100])
 
     ax1.legend()
-    save("mutation-perf")
+    save("mutations-perf")
 
 
 @click.command()
@@ -84,7 +84,7 @@ def arg():
     save("arg")
 
 
-cli.add_command(mutation_perf)
+cli.add_command(mutations_perf)
 cli.add_command(gc_perf)
 cli.add_command(arg)
 

### 2021-07-16 evaluation/generate_mutations_perf_data.py
Remove unused files.

@@ -1,7 +1,6 @@
 import time
 import pathlib
 
-import matplotlib.pyplot as plt
 import pandas as pd
 import numpy as np
 import click

### 2021-07-17 evaluation/generate_gc_perf_data.py
Update the GC code to validate the parameters.

@@ -1,114 +1,149 @@
 import subprocess
 import sys
+import tempfile
 import time
 
-
-#import sys
-#sys.path.insert(0, "/home/franz/syncedprojects/msprime_performance/msprime")
-
+import pandas as pd
+import numpy as np
+import click
 import msprime
-
-REPLICATES = 1
-
-Ne = 1
-mu = 1e-9
-r = 0
+import matplotlib.pyplot as plt
+import statsmodels.api as sm
 
 
-def run_simbac():
-    L = mb(4.5)
-    #L = 10000
-    #sample_sizes = list(range(10, 100 + 1, 10)) + list(range(200, 900 + 1, 100)) + list(range(1000, 10000 + 1, 1000))
-    sample_sizes = list(range(10,10+1,1)) + list(range(50,500 + 1, 50))
-    #sample_sizes = range(10, 100 + 1, 10)
-    #sample_sizes = range(200, 900 + 1, 100)
+def run_simbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=False):
 
-    def simbac(sample_size):
-        return simbac_run(sample_size, L)
-    def msp(sample_size):
-        return msp_run(samples=sample_size, sequence_length=L, population_size=1,
-                       ploidy = 1, recombination_rate=0,
-                       gene_conversion_rate=0.015, gene_conversion_tract_length=500)
+    # using R=2*gc_rate as gene conversion/recombination rate as SimBac uses R/2
+    R = gc_rate * 2
+    # Set theta to 0 to remove mutations (defaults to 0.01)
+    args = f"-N {sample_size} -B {int(L)} -R {R} -D {gc_tract_length} -T 0"
+    if count_trees:
+        # This is cleaned up automatically when the variable goes out of scope
+        tempdir = tempfile.TemporaryDirectory()
+        outfile = tempdir.name + "/out.newick"
+        args += " -l " + outfile
 
-    create_data(sample_sizes, [msp, simbac], ["msprime", "simbac"], "simbac")
+    # print("running", args)
+    subprocess.check_call(
+        "./tools/simbac " + args, shell=True, stdout=subprocess.DEVNULL
+    )
+    num_trees = 0
+    if count_trees:
+        with open(outfile) as f:
+            for line in f:
+                num_trees += 1
 
-
-def ms_run(sample_size, L):
-    args = ms_style_args(sample_size, L)
-    exec_cli("./sims/ms", args)
-
-
[…]

### 2021-07-17 evaluation/plot.py

@@ -55,7 +55,7 @@ def gc_perf():
     """
     Plot the gene-conversion benchmark.
     """
-    df = pd.read_csv("data/gc.csv")
+    df = pd.read_csv("data/gc-perf.csv")
 
     fig, ax1 = plt.subplots(1, 1)
     for tool in set(df.tool):

### 2021-07-17 evaluation/generate_gc_perf_data.py
Add validation for fastsimbac and runner code

@@ -24,13 +24,36 @@ def run_simbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=False):
         args += " -l " + outfile
 
     # print("running", args)
-    subprocess.check_call(
-        "./tools/simbac " + args, shell=True, stdout=subprocess.DEVNULL
+    subprocess.run(
+        "./tools/simbac " + args, shell=True, stdout=subprocess.DEVNULL, check=True
     )
     num_trees = 0
     if count_trees:
         with open(outfile) as f:
             for line in f:
+                num_trees += 1
+
+    return num_trees
+
+
+def run_fastsimbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=False):
+
+    # using R=2*gc_rate as gene conversion/recombination rate as SimBac uses R/2
+    R = gc_rate * 2
+    # Set theta to 0 to remove mutations (defaults to 0.01)
+    args = f"{sample_size} {int(L)} -r {R} {gc_tract_length} -t 0"
+    if count_trees:
+        args += " -T "
+
+    # print("running", args)
+    output = subprocess.run(
+        "./tools/fastSimBac " + args, shell=True, check=True, capture_output=True
+    )
+    # print(output.stdout)
+    num_trees = 0
+    if count_trees:
+        for line in output.stdout.splitlines():
+            if line.startswith(b"["):
                 num_trees += 1
 
     return num_trees
@@ -51,10 +74,9 @@ def run_msprime(*, sample_size, L, gc_rate, gc_tract_length):
 
 
 @click.command()
-@click.option("--output", type=click.Path(), default="tmp/gc_validation.png")
 @click.option("--replicates", type=int, default=1000)
 @click.option("--sample-size", type=int, default=10)
-def validate(output, replicates, sample_size):
+def validate(replicates, sample_size):
     """
     Validate that we are simulating the same things in the simulators
     by running some replicates and plotting the distributions of the
@@ -65,11 +87,19 @@ def validate(output, replicates, sample_size):
     gc_tract_length = 10
 
     nt_simbac = np.zeros(replicates)
+    nt_fastsimbac = np.zeros(replicates)
     nt_msprime = np.zeros(replicates)
 
     with click.progressbar(range(replicates)) as bar:
         for j in bar:
             nt_simbac[j] = run_simbac(
+                sample_size=sample_size,
+                L=L,
[…]

### 2021-07-17 evaluation/generate_sweeps_perf_data.py
Update sweeps perf code.

@@ -1,218 +1,128 @@
-import matplotlib
 import numpy as np
 import pandas as pd
 import msprime
-import attr
 import time
-import sys
-import functools
 import subprocess
-import matplotlib
-from matplotlib import pyplot as plt
-
-"""
-this outputs a csv file of the performance comparison (time_comp.csv)
-along with a rough figure to visualize stuff (time_comp.pdf)
-"""
-
-_mspms_executable = [sys.executable, "mspms_dev.py"]
-_discoal_executable = ["../tools/discoal"]
 
 
-class Timer:
-    def __init__(self):
-        self._start_time = None
+def _discoal_str_to_msprime(args):
+    # takes discoal command line as input and returns an iterator over the
+    # msprime tree sequences.
 
-    def start(self):
-        if self._start_time is not None:
-            raise Exception(f"Timer is running. Use .stop() to stop it")
-
-        self._start_time = time.perf_counter()
-
-    def stop(self):
-        if self._start_time is None:
-            raise Exception(f"Timer is not running. Use .start() to start it")
-
-        elapsed_time = time.perf_counter() - self._start_time
-        self._start_time = None
-        return(elapsed_time)
-
-@attr.s
-class Run:
-    """
-    The superclass of all tests. The only attribute defined is the output
-    directory for the test, which is guaranteed to exist when the
-    test method is called.
-    """
-
-    output_dir = attr.ib(type=str, default=None)
-
-    def _run_simulation(self, args):
-        p1 = subprocess.Popen(args, stdout=subprocess.PIPE)
-        output = p1.communicate()[0]
-        p1.wait()
-        if p1.returncode != 0:
-            raise ValueError("Error occured in subprocess: ", p1.returncode)
-
-    def get_ms_seeds(self):
-        max_seed = 2 ** 16
-        seeds = [np.random.randint(1, max_seed) for j in range(3)]
-        return ["-seed"] + list(map(str, seeds))
-
-    def _run_msprime(self, args):
-        return self._run_simulation(
-            _mspms_executable + args.split() + self.get_ms_seeds()
+    tokens = args.split(" ")
+    # positional args
+    sample_size = int(tokens[0])
+    nreps = int(tokens[1])
+    seq_length = int(tokens[2])
+    # parse discoal command line for params
+    # init ones we definitely need for comparison
+    theta = rho = alpha = sweep_site = sweep_mod_time = None
+    refsize = 1e6
+    for i in range(3, len(tokens)):
[…]

### 2021-07-17 evaluation/plot.py

@@ -68,6 +68,24 @@ def gc_perf():
     ax1.legend()
     save("gc-perf")
 
+@click.command()
+def sweeps_perf():
+    """
+    Plot the sweeps benchmark.
+    """
+    df = pd.read_csv("data/sweeps_perf.csv")
+
+    fig, ax1 = plt.subplots(1, 1)
+    for tool in set(df.tool):
+        dft = df[df.tool == tool]
+        ax1.plot(dft.L, dft.time, label=tool)
+
+    ax1.set_xlabel("Sequence length")
+    ax1.set_ylabel("Time (seconds)")
+
+    ax1.legend()
+    save("sweeps-perf")
+
 
 @click.command()
 def arg():
@@ -86,6 +104,7 @@ def arg():
 
 cli.add_command(mutations_perf)
 cli.add_command(gc_perf)
+cli.add_command(sweeps_perf)
 cli.add_command(arg)
 
 if __name__ == "__main__":

### 2021-07-18 paper.tex
Update sweeps perf code.
Some restructuring

@@
 
+\jkcomment{We should trim this down quite a bit - all we want to do is some basic
+roadmapping, like, we discuss the underlying tree sequence object,
+then mutations (because it's quite separate from the rest)
+and then a bunch of new features for ancestry simulation. There's no
+point in listing it out section by section.}
 In this section we describe the main features of \msprime\ 1.0. Methodologically,
@@
 maintainability of the core codebase.
+
+\subsection*{Tree sequences}
+One of the key reasons for \msprime's significiant performance advantage
+over other simulators is its usage of the ``succinct tree sequence''
+(usually, tree sequence) data stucture and the supporting \tskit\
+library~\citep{tskit2021tskit}.
+While a full discussion of tree sequences and the capabilities
+of \tskit\ is beyond the scope of this article, we briefly summarise
+here.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=6cm]{figures/example_tree_sequence}
+\end{center}
+\caption{\label{fig-ts-example} Example tree sequence.
+An example tree sequence describing genealogies and genotypes
+for four samples at ten sites.
+}
+\end{figure}
+
+\jkcomment{Paragraph describing what a tree sequence is, referring to
+Fig.~\ref{fig-ts-example}.}
+
+The point of simulations is to generate data and the efficiency with
+which simulated data can be incorporated into an overall application
+is an important issue.
+For example, Approximate Bayesian Computation
+(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
+and more recent machine learning
+methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
+perform population genetic inference by using large numbers of
+simulations as training data. Clearly, simulation efficiency is
+crucial since the size and number of simulations that can be performed determines
+the depth to which the model and parameter space can be sampled from.
+Equally important,
+however, is the efficiency with which the simulation results can be
+transformed into the specific input required by the inference method.
+In the case of ABC, this is usually a set of summary statistics of the sequence data.
+Modern approaches such as
+ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
+ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large
+numbers of user-specified statistics that are individually weakly informative,
+but jointly close to sufficient for the desired model or parameter.
+However, the computational resources needed to calculate these statistics can be even
+greater than those used to generate the simulated datasets.
+Older software such as DIYABC~\citep{cornuet2008inferring}
+or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
+developing their own specialised simulators, tightly integrating them
+with the inference method. While this is certainly more efficient,
+there is a great deal of duplication of effort and significant
+room for error.
+As ABC inference has become useful in an ever-widening range of applications,
+these all-in-one methods risk falling behind if they do not allow users to explore
+the models and statistics of most interest.
+
+\jkcomment{Stuff to cover}
+\begin{itemize}
+
+\item The core problem is that the field lacked an efficient mechanism for
+    interchanging simulated data, and so developers were forced to choose
+    between efficiency and duplication of effort.
+
+\item What are the output formats of simulators? Newick, or sequence data
+    basically. What's the problem with this? We don't have mutations.
+    Much more efficient to have the tree and mutations than the
+    sequence data. Fundamentally, the output of coalescent simulations
+    should be trees, otherwise we are throwing a lot of information
+    away.
+
+\item Lots of examples of papers where people have been forced to
+    create a modified fork of some simulator so that they can get
+    access to some internal details, not exposed by the simulators
+    interface. Tree sequences provide a complete record of the
+    ancestral process, so we can do things like compute IBD
+    segments and introgressed tracts directly on the output. See
+    also ARG section for recording more information, though.
+
+\item Side note somewhere: with efficient access to branch length
+    statistics~\citep{ralph2019efficiently}, maybe we don't need to
+    simulate mutations at all. (not in the ABC context, but elsewhere)
+
+\item Msprime uses the \tskit\ library~\citep{tskit2021tskit}, which provides
+    extremely efficient interchange and data processing. A quick example
+    of how efficient.
+
+\item Tskit has facilited unprecedented interoperability between simulators:
+    Forward simulations can also use tree
+    sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
+    3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
+    tree sequences. Msprime can take tree sequences simulated by these forward
+    simulators as input and complete the simulation (``recapitation'').
+    See \cite{haller2018tree} for more details.
+    (Although, this could be a discussion point?)
+
+\item Recently, Phastsim/USHER~\cite{demaio2021phastsim,turakhia2021ultrafast}
+    have used this idea of storing the
+    tree plus mutations, but approach is crude compared to tskit.
+
+\end{itemize}
+
+\subsection*{Finite sites mutations}
+Version 0.x of \texttt{msprime} provided a very simple mutation model,
+supporting only the infinite sites model.
+With version 1.0, we now support a large number of mutation models,
+including
+JC69~\citep{jukes1969evolution},
+F84~\citep{felsenstein1996hidden},
+and GTR~\citep{tavare1986some} nucleotide models
+and the BLOSUM62~\citep{henikoff1992amino}
+and PAM~\citep{dayhoff1978} amino acid models.
+The general model accepts an arbitrary transition matrix between
+any number of alleles (which can be arbitrary unicode strings),
+and so other models such as the Kimura two and three
+parameter models~\citep{kimura1980simple,kimura1981estimation}
+parameters models can be easily
+and efficiently defined in user code.
+Full support for the original infinite sites model is also maintained.
+The results of mutation simulations across a range of models are
+extensively validated
+against both theoretical expectations and output from
+Seq-Gen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/mutations-perf}
+\caption{\label{fig-mutations-perf} Time required to run
+\texttt{sim\_mutations} on tree sequences generated
+by \texttt{sim\_ancestry} (with a population size of $10^4$
+and recombination rate of $10^{-8}$) for varying (haploid) sample
+size and sequence length. We run 10 replicate mutation simulations
+each for three different mutation rates, and report the average
+CPU time required (Intel Core i7-9700).
+(A) Holding sequence length fixed at 10 megabases and varying the
+number of samples (tree tips) from 10 to 100,000.
+(B) Holding number of samples fixed at 1000, and varying the sequence
+length from 1 to 100 megabases.}
+\end{figure}
+
+Comparing the performance of mutation generation methods is problematic.
+Because there is no standard format for representing trees and
+the accompanying mutations, all existing methods are focused on
+writing out the resulting sequences. Methods are also not modular,
+and so it is very difficult to disentangle the time required to
+read in the tree file and output the sequences from the actual
+time required to perform the simulation.
+
+Fig.~\ref{fig-mutations-perf} shows that the mutation generation
+machinery in \msprime\ is extremely efficient, requiring
+fractions of a second to generate mutations for most simulations.
+For example, the longest running simulation in
+Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
+generate an average of 1.5 million mutations over 137,081 trees
+in a tree sequence with 508,125 edges.
+Generating mutations for a tree sequence is particularly efficient
+because we can work edge-by-edge rather than having to
+consider each tree independently.
+
+\jkcomment{We probably want to keep a version of this narrative too.}
+For example, we simulated mutations under the JC69
+model for a tree with $10^6$ leaves over $10^4$ sites (resulting
+in $\sim 280000$ mutations) in less that 1.5 seconds, including
+the time required for file input and output. While a systematic
+comparison of run-times is not feasible here, this is orders
+of magnitude faster than methods such as
+Seq-Gen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}
+and at least competitive with the recent method from
+\cite{demaio2021phastsim}.
+
+\textbf{TODO: would love to do a like-for-like comparison, but quite hard}
+
+Another substantial advantage of our approach is that the simulated
+mutations are stored directly in the output data format with
+a defined API. Although
+Pyvolve and Phastsim are implemented in Python, neither provide
+programatic access to the simulated output, outputting instead
+to a series of files which must subsequently be parsed.
+See the Data Model section for more details and discussion.
+
+Things we can do: future work will need to do things like
+Dawg~\citep{cartwright2005dna} and to implement models and
+indels~\citep{fletcher2009indelible}.
 
@@
 \begin{figure}
-\Large{A FIGURE SHOWING DISCOAL AND MSPRIME}
-% \includegraphics[width=\textwidth]{figures/recombination}
+\includegraphics[width=6cm]{figures/sweeps-perf}
 \caption{\label{fig-selection-perf}Comparison of the runtime for
@@
 hybrid approach provides large efficiency gains.
-
-\subsection*{Finite sites mutations}
-Version 0.x of \texttt{msprime} provided a very simple mutation model,
-supporting only the infinite sites model.
-With version 1.0, we now support a large number of mutation models,
-including
-JC69~\citep{jukes1969evolution},
-F84~\citep{felsenstein1996hidden},
-and GTR~\citep{tavare1986some} nucleotide models
-and the BLOSUM62~\citep{henikoff1992amino}
-and PAM~\citep{dayhoff1978} amino acid models.
-The general model accepts an arbitrary transition matrix between
-any number of alleles (which can be arbitrary unicode strings),
-and so other models such as the Kimura two and three
-parameter models~\citep{kimura1980simple,kimura1981estimation}
-parameters models can be easily
-and efficiently defined in user code.
-Full support for the original infinite sites model is also maintained.
-The results of mutation simulations across a range of models are
-extensively validated
-against both theoretical expectations and output from
-Seq-Gen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{figures/mutations-perf}
-\caption{\label{fig-mutations-perf} Time required to run
-\texttt{sim\_mutations} on tree sequences generated
-by \texttt{sim\_ancestry} (with a population size of $10^4$
-and recombination rate of $10^{-8}$) for varying (haploid) sample
-size and sequence length. We run 10 replicate mutation simulations
-each for three different mutation rates, and report the average
-CPU time required (Intel Core i7-9700).
-(A) Holding sequence length fixed at 10 megabases and varying the
-number of samples (tree tips) from 10 to 100,000.
-(B) Holding number of samples fixed at 1000, and varying the sequence
-length from 1 to 100 megabases.}
-\end{figure}
-
-Comparing the performance of mutation generation methods is problematic.
-Because there is no standard format for representing trees and
-the accompanying mutations, all existing methods are focused on
-writing out the resulting sequences. Methods are also not modular,
-and so it is very difficult to disentangle the time required to
-read in the tree file and output the sequences from the actual
-time required to perform the simulation.
-
-Fig.~\ref{fig-mutations-perf} shows that the mutation generation
-machinery in \msprime\ is extremely efficient, requiring
-fractions of a second to generate mutations for most simulations.
-For example, the longest running simulation in
-Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
-generate an average of 1.5 million mutations over 137,081 trees
-in a tree sequence with 508,125 edges.
-Generating mutations for a tree sequence is particularly efficient
-because we can work edge-by-edge rather than having to
-consider each tree independently.
-
-\jkcomment{We probably want to keep a version of this narrative too.}
-For example, we simulated mutations under the JC69
-model for a tree with $10^6$ leaves over $10^4$ sites (resulting
-in $\sim 280000$ mutations) in less that 1.5 seconds, including
-the time required for file input and output. While a systematic
-comparison of run-times is not feasible here, this is orders
-of magnitude faster than methods such as
-Seq-Gen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}
-and at least competitive with the recent method from
-\cite{demaio2021phastsim}.
-
-\textbf{TODO: would love to do a like-for-like comparison, but quite hard}
-
-Another substantial advantage of our approach is that the simulated
-mutations are stored directly in the output data format with
-a defined API. Although
-Pyvolve and Phastsim are implemented in Python, neither provide
-programatic access to the simulated output, outputting instead
-to a series of files which must subsequently be parsed.
-See the Data Model section for more details and discussion.
-
-Things we can do: future work will need to do things like
-Dawg~\citep{cartwright2005dna} and to implement models and
-indels~\citep{fletcher2009indelible}.
 
@@
 Demes format~\citep{gower2021demes} from the command line.
-
-\subsection*{Data interchange and interoperability}
-The point of simulations is to generate data and the efficiency with
-which simulated data can be incorporated into an overall application
-is an important---if often overlooked---issue.
-For example, Approximate Bayesian Computation
-(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
-and more recent machine learning
-methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
-perform population genetic inference by using large numbers of
-simulations as training data. Clearly, simulation efficiency is
-crucial since the size and number of simulations that can be performed determines
-the depth to which the model and parameter space can be sampled from.
-Equally important,
-however, is the efficiency with which the simulation results can be
-transformed into the specific input required by the inference method.
-In the case of ABC, this is usually a set of summary statistics of the sequence data.
-Modern approaches such as
-ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
-ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large numbers of user-specified statistics that are individually weakly informative,
-but jointly close to sufficient for the desired model or parameter.
-However, the computational resources needed to calculate these statistics can be even
-greater than those used to generate the simulated datasets.
-Older software such as DIYABC~\citep{cornuet2008inferring}
-or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
-developing their own specialised simulators, tightly integrating them
-with the inference method. While this is certainly more efficient,
-there is a great deal of duplication of effort and significant
-room for error.
-As ABC inference has become useful in an ever-widening range of applications,
-these all-in-one methods risk falling behind if they do not allow users to explore
-the models and statistics of most interest.
-
-\jkcomment{Stuff to cover}
-\begin{itemize}
-
-\item The core problem is that the field lacks an efficient mechanism for
-    interchanging simulated data, and so developers are forced to choose
-    between efficiency and duplication of effort.
-
-\item What are the output formats of simulators? Newick, or sequence data
-    basically. What's the problem with this? We don't have mutations.
-    Much more efficient to have the tree and mutations than the
-    sequence data. Fundamentally, the output of coalescent simulations
-    should be trees, otherwise we are throwing a lot of information
-    away.
-
-\item Side note somewhere: with efficient access to branch length
-    statistics~\citep{ralph2019efficiently}, maybe we don't need to
-    simulate mutations at all.
-[GT: I'm not so sure about this -- the desired statistics/summaries
-must be calculated on the observed dataset as well as on the many
-simulated datasets.
-Unless the observed dataset is stored as a tree sequence,
-(because it has been through tsinfer, for example),
-it's not clear how branch or node stats could be calculated
-from it.
-]
-
-\item Msprime uses the \tskit\ library~\citep{tskit2021tskit}, which provides
-    extremely efficient interchange and data processing. A quick example
-    of how efficient.
-
-\item Tskit has facilited unprecedented interoperability between simulators:
-    Forward simulations can also use tree
-    sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
-    3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
-    tree sequences. Msprime can take tree sequences simulated by these forward
-    simulators as input and complete the simulation (``recapitation'').
-    See \cite{haller2018tree} for more details.
-
-\item Recently, Phastsim/USHER~\cite{demaio2021phastsim,turakhia2021ultrafast}
-    have used this idea of storing the
-    tree plus mutations, but approach is crude compared to tskit.
-
-\end{itemize}
 

### 2021-07-18 paper.tex
Comment out out-of-date roadmap text.

@@
 
-\jkcomment{We should trim this down quite a bit - all we want to do is some basic
-roadmapping, like, we discuss the underlying tree sequence object,
+\jkcomment{Give a quick roadmap,
+like, we discuss the underlying tree sequence object,
 then mutations (because it's quite separate from the rest)
@@
 point in listing it out section by section.}
-In this section we describe the main features of \msprime\ 1.0. Methodologically,
-there exists a logical separation between simulating coalescent genealogies and
-adding neutral mutations onto these genealogies. Simulating genetic ancestry
-requires methods that model the underlying reproductive process (e.g. Moran,
-Wright-Fisher, etc.) as well as accounting for recombination of various sorts
-(including gene conversion), demography, and selection. These methods are
-discussed in the first N subsections below. One of the strengths of the
-approach taken in msprime is that neutral mutations can be retrospectively
-added to simulated genealogies. Methods to generate mutations, and thereby
-generate genetic variation, are therefore discussed in a separate and subsequent
-subsection. Finally, we describe the \msprime\ API, which is designed to enforce
-the distinction between simulating genealogies and mutations; discuss the
-interoperability provided by the use of \tskit\, the tree sequence toolkit;
-and highlight the open source development process which allows for \msprime\
-to be extended by addition of new features while ensuring robustness and
-maintainability of the core codebase.
+
+% JK: commented out this text as it's misleading and would confuse
+% people. We should adapt it when writing the roadmap when the structure
+% has settled down.
+
+% In this section we describe the main features of \msprime\ 1.0. Methodologically,
+% there exists a logical separation between simulating coalescent genealogies and
+% adding neutral mutations onto these genealogies. Simulating genetic ancestry
+% requires methods that model the underlying reproductive process (e.g. Moran,
+% Wright-Fisher, etc.) as well as accounting for recombination of various sorts
+% (including gene conversion), demography, and selection. These methods are
+% discussed in the first N subsections below. One of the strengths of the
+% approach taken in msprime is that neutral mutations can be retrospectively
+% added to simulated genealogies. Methods to generate mutations, and thereby
+% generate genetic variation, are therefore discussed in a separate and subsequent
+% subsection. Finally, we describe the \msprime\ API, which is designed to enforce
+% the distinction between simulating genealogies and mutations; discuss the
+% interoperability provided by the use of \tskit\, the tree sequence toolkit;
+% and highlight the open source development process which allows for \msprime\
+% to be extended by addition of new features while ensuring robustness and
+% maintainability of the core codebase.
 

### 2021-07-19 evaluation/generate_gc_perf_data.py
Apply suggestions from code review
Co-authored-by: fbaumdicker <franz.baumdicker@gmx.de>

@@ -36,7 +36,7 @@ def run_simbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=False):
     return num_trees
 
 
-def run_fastsimbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=False):
+def run_fastsimbac(*, sample_size, L, gc_rate, gc_tract_length, set_seed=0, count_trees=False):
 
     # using R=2*gc_rate as gene conversion/recombination rate as SimBac uses R/2
     R = gc_rate * 2
@@ -44,6 +44,8 @@ def run_fastsimbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=Fals
     args = f"{sample_size} {int(L)} -r {R} {gc_tract_length} -t 0"
     if count_trees:
         args += " -T "
+    if set_seed > 0:
+        args += " -s " + str(set_seed)
 
     # print("running", args)
     output = subprocess.run(
@@ -104,6 +106,7 @@ def validate(replicates, sample_size):
                 L=L,
                 gc_rate=gc_rate,
                 gc_tract_length=gc_tract_length,
+                set_seed = j,
                 count_trees=True,
             )
             nt_msprime[j] = run_msprime(

### 2021-07-19 evaluation/generate_gc_perf_data.py
Add fastSimBac and skip larger sims.

@@ -3,6 +3,7 @@ import sys
 import tempfile
 import time
 
+import cpuinfo
 import pandas as pd
 import numpy as np
 import click
@@ -150,6 +151,7 @@ def validate(replicates, sample_size):
 
     plt.close("all")
 
+
     sm.graphics.qqplot(nt_fastsimbac)
     sm.qqplot_2samples(nt_fastsimbac, nt_msprime, line="45")
     plt.xlabel("fastsimbac")
@@ -163,10 +165,13 @@ def benchmark_ecoli(replicates):
     """
     Runs the benchmarks for E-coli simulations.
     """
+
+    cpu = cpuinfo.get_cpu_info()
+    with open("data/gc_perf_cpu.txt", "w") as f:
+        for k, v in cpu.items():
+            print(k, "\t", v, file=f)
+
     L = 4_500_000
-    # Make this run in a reasonable amount of time, change this when
-    # we're doing final runs.
-    L /= 100
     gc_rate = 0.015
     gc_tract_length = 500
 
@@ -174,10 +179,15 @@ def benchmark_ecoli(replicates):
     tool_map = {
         "msprime": run_msprime,
         "SimBac": run_simbac,
+        "fastSimBac": run_fastsimbac,
     }
     data = []
-    for sample_size in sample_sizes:
+    for j, sample_size in enumerate(sample_sizes):
         for name, func in tool_map.items():
+            if j > 1 and name == "SimBac":
+                continue
+            if j > 3 and name == "fastSimBac":
+                continue
             print(name, "n =", sample_size)
             before = time.perf_counter()
             for _ in range(replicates):

### 2021-07-19 paper.tex
Add tree sequence paragraph.

@@
 \usepackage[right]{lineno}%
-\usepackage{microtype}
+\usepackage{nameref}
 %\usepackage{showkeys}
@@
 
+\label{sec-ts}
 \subsection*{Tree sequences}
@@
 
-\jkcomment{Paragraph describing what a tree sequence is, referring to
-Fig.~\ref{fig-ts-example}.}
+The succinct tree sequence is a data structure for concisely encoding
+genetic ancestry. Given a set of sample chromosomes, the topology
+of their genetic ancestry is described by a set of \emph{nodes}
+and \emph{edges}. A node in tree sequence terminology corresponds
+to a single chromosome strand, and is associated with its birth time,
+birth population and other metadata. A node may optionally be
+associated with an \emph{individual}, so that, for example, a diploid
+individual will be associated with two separate nodes (we omit
+the individual table in Fig.~\ref{fig-ts-example} as it is optional).
+Nodes define haplotypes that exist at a specific time within a
+specific individual, and edges define genetic inheritence
+relationships between these nodes. An edge consists of a
+parent node, a child node and the left and right coordinates
+of the contiguous genomic interval over which
+the child inherited genetic material from the parent.
+Critically, edges can span multiple marginal trees: for
+example, in Fig.~\ref{fig-ts-example} the branch from node
+0 to 4 is present in both marginal trees, and represented
+by a single edge (the first row in the example edge table).
+This simple device, of explicitly associating tree nodes
+with specific ancestral chromosomes and recording the
+genomic intervals over which parent/child relationships exist,
+is a generalisation from the original concept  of
+``coalescence records''~\citep{kelleher2016efficient}, and
+is the key to the efficiency of tree sequences.
+\jkcomment{Summarise and add references?}
+As discussed in the \ref{sec-mutations} section, tree sequences
+can also store genetic variation data very efficiently.
+
+The term Ancestral Recombination Graph (ARG) is often understood
+to mean the complete record of a sample's genetic
+ancestry~\citep[e.g.][]{mathieson2020ancestry}, and in this sense a tree
+sequence \emph{is} an ARG. More technically, an ARG is a particular
+graphical representation of genetic ancestry based on the
+flow of ancestral material through recombination and common
+ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
+In this sense, tree sequences can be used to \emph{encode}
+ARGs without loss of information---although we do not store
+the additional ARG information by default
+in \msprime\ because the minimal tree sequence encoding is
+much more concise (see the \nameref{sec-arg} section for more details).
 
@@
 
+\label{sec-mutations}
 \subsection*{Finite sites mutations}
@@
 
-\subsection*{Ancestral recombination graphs}
+\subsection*{Ancestral Recombination Graphs}
+\label{sec-arg}
 

### 2021-07-19 paper.tex
Discuss mutations.

@@
 ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
-and calculation of population genetic statistics~\citep{ralph2019efficiently}.
+and calculation of population genetic statistics~\citep{ralph2020efficiently}.
 Through a rigorous open-source community development process,
@@
 
+\subsection*{Tree sequences}
 \label{sec-ts}
-\subsection*{Tree sequences}
 One of the key reasons for \msprime's significiant performance advantage
@@
 here.
+\jkcomment{We need to mention that tree sequences and tskit are
+derived from the 2016 paper and have been extracted from msprime
+into a separate C/Python library.}
 
@@
 \jkcomment{Summarise and add references?}
-As discussed in the \ref{sec-mutations} section, tree sequences
-can also store genetic variation data very efficiently.
 
@@
 much more concise (see the \nameref{sec-arg} section for more details).
-
+It should be noted that, despite efforts to
+standardise~\citep{cardona2008extended,mcgill2013graphml}, there is
+no shared format for interchanging ARG data, and no supporting software
+libraries.
+
+Tree sequences encode a complete and concise description of a set of sampled
+chromosome's genetic ancestry, and they can also efficiently represent
+the genetic variation (extant and ancestral).
+As illustrated in Fig.~\ref{fig-ts-example}, the \tskit\ tree
+sequence encoding includes \emph{site} and \emph{mutation} tables. Each
+row in the site table represents some position along the genome, and
+each row in the mutation table represents a change of allelic state
+at a particular site and node. Since the number of mutations is typically
+small, we can usually store the genetic variation for a given variant
+site in constant space. This approach of storing the genetic variation
+via the genetic ancestry is both very concise and extremely efficient
+to process---for example, computing Tajima's $D$ using \tskit\
+is several orders of magnitude faster for large sample sizes than
+efficient libraries working with the variant matrix
+representation~\citep{ralph2020efficiently}.
+\jkcomment{Mention Phastsim/USHER~\citep{demaio2021phastsim,turakhia2021ultrafast}
+here. Same basic idea, but much cruder.}
+
+\jkcomment{Make some sort of link here between this paper, which is
+about msprime simulations, but that we can't really disentangle
+simulations from their downstream processing.}
 The point of simulations is to generate data and the efficiency with
@@
 
-\jkcomment{Stuff to cover}
+\jkcomment{Other potential stuff to cover}
 \begin{itemize}
-
-\item The core problem is that the field lacked an efficient mechanism for
-    interchanging simulated data, and so developers were forced to choose
-    between efficiency and duplication of effort.
 
@@
 \item Side note somewhere: with efficient access to branch length
-    statistics~\citep{ralph2019efficiently}, maybe we don't need to
+    statistics~\citep{ralph2020efficiently}, maybe we don't need to
     simulate mutations at all. (not in the ABC context, but elsewhere)
-
-\item Msprime uses the \tskit\ library~\citep{tskit2021tskit}, which provides
-    extremely efficient interchange and data processing. A quick example
-    of how efficient.
 
@@
 
-\item Recently, Phastsim/USHER~\cite{demaio2021phastsim,turakhia2021ultrafast}
-    have used this idea of storing the
-    tree plus mutations, but approach is crude compared to tskit.
-
 \end{itemize}
 
+\subsection*{Finite sites mutations}
 \label{sec-mutations}
-\subsection*{Finite sites mutations}
 Version 0.x of \texttt{msprime} provided a very simple mutation model,

### 2021-07-19 paper.tex
small changes in the introduction

--- 2021-07-18 paper.tex
@@
 process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
-models the ancestry of set of sampled genomes. Given some sampled
+models the ancestry of a set of sampled genomes. Given some sampled
 genomes, the coalescent provides a mathematical description of the
@@
 arenas2012simulation,yuan2012overview,hoban2012computer}.
-Two developments of recent years, however, have made the majority
+%% CQ: Reading this sentence, I thought we were still talking abour simulators, that's what I proposed these changes
+Two technological developments of recent years, however, have made the majority
 of these simulators of little relevance to modern datasets.
 % Some citations here? Darwin tree of life maybe?
-Firstly, whole genome sequencing technology---and in particular fourth-generation
+Firstly, whole genome sequencing ---and in particular fourth-generation
 sequencing technologies---have made chromosome-level assemblies
@@
 
-\jkcomment{Give a quick roadmap,
-like, we discuss the underlying tree sequence object,
-then mutations (because it's quite separate from the rest)
-and then a bunch of new features for ancestry simulation. There's no
-point in listing it out section by section.}
-
-% JK: commented out this text as it's misleading and would confuse
-% people. We should adapt it when writing the roadmap when the structure
-% has settled down.
-
-% In this section we describe the main features of \msprime\ 1.0. Methodologically,
-% there exists a logical separation between simulating coalescent genealogies and
-% adding neutral mutations onto these genealogies. Simulating genetic ancestry
-% requires methods that model the underlying reproductive process (e.g. Moran,
-% Wright-Fisher, etc.) as well as accounting for recombination of various sorts
-% (including gene conversion), demography, and selection. These methods are
-% discussed in the first N subsections below. One of the strengths of the
-% approach taken in msprime is that neutral mutations can be retrospectively
-% added to simulated genealogies. Methods to generate mutations, and thereby
-% generate genetic variation, are therefore discussed in a separate and subsequent
-% subsection. Finally, we describe the \msprime\ API, which is designed to enforce
-% the distinction between simulating genealogies and mutations; discuss the
-% interoperability provided by the use of \tskit\, the tree sequence toolkit;
-% and highlight the open source development process which allows for \msprime\
-% to be extended by addition of new features while ensuring robustness and
-% maintainability of the core codebase.
-
-\subsection*{Tree sequences}
-One of the key reasons for \msprime's significiant performance advantage
-over other simulators is its usage of the ``succinct tree sequence''
-(usually, tree sequence) data stucture and the supporting \tskit\
-library~\citep{tskit2021tskit}.
-While a full discussion of tree sequences and the capabilities
-of \tskit\ is beyond the scope of this article, we briefly summarise
-here.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=6cm]{figures/example_tree_sequence}
-\end{center}
-\caption{\label{fig-ts-example} Example tree sequence.
-An example tree sequence describing genealogies and genotypes
-for four samples at ten sites.
-}
-\end{figure}
-
-\jkcomment{Paragraph describing what a tree sequence is, referring to
-Fig.~\ref{fig-ts-example}.}
-
-The point of simulations is to generate data and the efficiency with
-which simulated data can be incorporated into an overall application
-is an important issue.
-For example, Approximate Bayesian Computation
-(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
-and more recent machine learning
-methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
-perform population genetic inference by using large numbers of
-simulations as training data. Clearly, simulation efficiency is
-crucial since the size and number of simulations that can be performed determines
-the depth to which the model and parameter space can be sampled from.
-Equally important,
-however, is the efficiency with which the simulation results can be
-transformed into the specific input required by the inference method.
-In the case of ABC, this is usually a set of summary statistics of the sequence data.
-Modern approaches such as
-ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
-ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large
-numbers of user-specified statistics that are individually weakly informative,
-but jointly close to sufficient for the desired model or parameter.
-However, the computational resources needed to calculate these statistics can be even
-greater than those used to generate the simulated datasets.
-Older software such as DIYABC~\citep{cornuet2008inferring}
-or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
-developing their own specialised simulators, tightly integrating them
-with the inference method. While this is certainly more efficient,
-there is a great deal of duplication of effort and significant
-room for error.
-As ABC inference has become useful in an ever-widening range of applications,
-these all-in-one methods risk falling behind if they do not allow users to explore
-the models and statistics of most interest.
-
-\jkcomment{Stuff to cover}
-\begin{itemize}
-
-\item The core problem is that the field lacked an efficient mechanism for
-    interchanging simulated data, and so developers were forced to choose
-    between efficiency and duplication of effort.
-
-\item What are the output formats of simulators? Newick, or sequence data
-    basically. What's the problem with this? We don't have mutations.
-    Much more efficient to have the tree and mutations than the
-    sequence data. Fundamentally, the output of coalescent simulations
-    should be trees, otherwise we are throwing a lot of information
-    away.
-
-\item Lots of examples of papers where people have been forced to
-    create a modified fork of some simulator so that they can get
-    access to some internal details, not exposed by the simulators
-    interface. Tree sequences provide a complete record of the
-    ancestral process, so we can do things like compute IBD
-    segments and introgressed tracts directly on the output. See
-    also ARG section for recording more information, though.
-
-\item Side note somewhere: with efficient access to branch length
-    statistics~\citep{ralph2019efficiently}, maybe we don't need to
-    simulate mutations at all. (not in the ABC context, but elsewhere)
-
-\item Msprime uses the \tskit\ library~\citep{tskit2021tskit}, which provides
-    extremely efficient interchange and data processing. A quick example
-    of how efficient.
-
-\item Tskit has facilited unprecedented interoperability between simulators:
-    Forward simulations can also use tree
-    sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
-    3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
-    tree sequences. Msprime can take tree sequences simulated by these forward
-    simulators as input and complete the simulation (``recapitation'').
-    See \cite{haller2018tree} for more details.
-    (Although, this could be a discussion point?)
-
-\item Recently, Phastsim/USHER~\cite{demaio2021phastsim,turakhia2021ultrafast}
-    have used this idea of storing the
-    tree plus mutations, but approach is crude compared to tskit.
-
-\end{itemize}
-
-\subsection*{Finite sites mutations}
-Version 0.x of \texttt{msprime} provided a very simple mutation model,
-supporting only the infinite sites model.
-With version 1.0, we now support a large number of mutation models,
-including
-JC69~\citep{jukes1969evolution},
-F84~\citep{felsenstein1996hidden},
-and GTR~\citep{tavare1986some} nucleotide models
-and the BLOSUM62~\citep{henikoff1992amino}
-and PAM~\citep{dayhoff1978} amino acid models.
-The general model accepts an arbitrary transition matrix between
-any number of alleles (which can be arbitrary unicode strings),
-and so other models such as the Kimura two and three
-parameter models~\citep{kimura1980simple,kimura1981estimation}
-parameters models can be easily
-and efficiently defined in user code.
-Full support for the original infinite sites model is also maintained.
-The results of mutation simulations across a range of models are
-extensively validated
-against both theoretical expectations and output from
-Seq-Gen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{figures/mutations-perf}
-\caption{\label{fig-mutations-perf} Time required to run
-\texttt{sim\_mutations} on tree sequences generated
-by \texttt{sim\_ancestry} (with a population size of $10^4$
-and recombination rate of $10^{-8}$) for varying (haploid) sample
-size and sequence length. We run 10 replicate mutation simulations
-each for three different mutation rates, and report the average
-CPU time required (Intel Core i7-9700).
-(A) Holding sequence length fixed at 10 megabases and varying the
-number of samples (tree tips) from 10 to 100,000.
-(B) Holding number of samples fixed at 1000, and varying the sequence
-length from 1 to 100 megabases.}
-\end{figure}
-
-Comparing the performance of mutation generation methods is problematic.
-Because there is no standard format for representing trees and
-the accompanying mutations, all existing methods are focused on
-writing out the resulting sequences. Methods are also not modular,
-and so it is very difficult to disentangle the time required to
-read in the tree file and output the sequences from the actual
-time required to perform the simulation.
-
-Fig.~\ref{fig-mutations-perf} shows that the mutation generation
-machinery in \msprime\ is extremely efficient, requiring
-fractions of a second to generate mutations for most simulations.
-For example, the longest running simulation in
-Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
-generate an average of 1.5 million mutations over 137,081 trees
-in a tree sequence with 508,125 edges.
-Generating mutations for a tree sequence is particularly efficient
-because we can work edge-by-edge rather than having to
-consider each tree independently.
-
-\jkcomment{We probably want to keep a version of this narrative too.}
-For example, we simulated mutations under the JC69
-model for a tree with $10^6$ leaves over $10^4$ sites (resulting
-in $\sim 280000$ mutations) in less that 1.5 seconds, including
-the time required for file input and output. While a systematic
-comparison of run-times is not feasible here, this is orders
-of magnitude faster than methods such as
-Seq-Gen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}
-and at least competitive with the recent method from
-\cite{demaio2021phastsim}.
-
-\textbf{TODO: would love to do a like-for-like comparison, but quite hard}
-
-Another substantial advantage of our approach is that the simulated
-mutations are stored directly in the output data format with
-a defined API. Although
-Pyvolve and Phastsim are implemented in Python, neither provide
-programatic access to the simulated output, outputting instead
-to a series of files which must subsequently be parsed.
-See the Data Model section for more details and discussion.
-
-Things we can do: future work will need to do things like
-Dawg~\citep{cartwright2005dna} and to implement models and
-indels~\citep{fletcher2009indelible}.
+In this section we describe the main features of \msprime\ 1.0. Methodologically,
+there exists a logical separation between simulating coalescent genealogies and
+adding neutral mutations onto these genealogies. Simulating genetic ancestry
+requires methods that model the underlying reproductive process (e.g. Moran,
+Wright-Fisher, etc.) as well as accounting for recombination of various sorts
+(including gene conversion), demography, and selection. These methods are
+discussed in the first N subsections below. One of the strengths of the
+approach taken in msprime is that neutral mutations can be retrospectively
+added to simulated genealogies. Methods to generate mutations, and thereby
+generate genetic variation, are therefore discussed in a separate and subsequent
+subsection. Finally, we describe the \msprime\ API, which is designed to enforce
+the distinction between simulating genealogies and mutations; discuss the
+interoperability provided by the use of \tskit\, the tree sequence toolkit;
+and highlight the open source development process which allows for \msprime\
+to be extended by addition of new features while ensuring robustness and
+maintainability of the core codebase.
 
@@
 
+\subsection*{Finite sites mutations}
+Version 0.x of \texttt{msprime} provided a very simple mutation model,
+supporting only the infinite sites model.
+With version 1.0, we now support a large number of mutation models,
+including
+JC69~\citep{jukes1969evolution},
+F84~\citep{felsenstein1996hidden},
+and GTR~\citep{tavare1986some} nucleotide models
+and the BLOSUM62~\citep{henikoff1992amino}
+and PAM~\citep{dayhoff1978} amino acid models.
+The general model accepts an arbitrary transition matrix between
+any number of alleles (which can be arbitrary unicode strings),
+and so other models such as the Kimura two and three
+parameter models~\citep{kimura1980simple,kimura1981estimation}
+parameters models can be easily
+and efficiently defined in user code.
+Full support for the original infinite sites model is also maintained.
+The results of mutation simulations across a range of models are
+extensively validated
+against both theoretical expectations and output from
+Seq-Gen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/mutations-perf}
+\caption{\label{fig-mutations-perf} Time required to run
+\texttt{sim\_mutations} on tree sequences generated
+by \texttt{sim\_ancestry} (with a population size of $10^4$
+and recombination rate of $10^{-8}$) for varying (haploid) sample
+size and sequence length. We run 10 replicate mutation simulations
+each for three different mutation rates, and report the average
+CPU time required (Intel Core i7-9700).
+(A) Holding sequence length fixed at 10 megabases and varying the
+number of samples (tree tips) from 10 to 100,000.
+(B) Holding number of samples fixed at 1000, and varying the sequence
+length from 1 to 100 megabases.}
+\end{figure}
+
+Comparing the performance of mutation generation methods is problematic.
+Because there is no standard format for representing trees and
+the accompanying mutations, all existing methods are focused on
+writing out the resulting sequences. Methods are also not modular,
+and so it is very difficult to disentangle the time required to
+read in the tree file and output the sequences from the actual
+time required to perform the simulation.
+
+Fig.~\ref{fig-mutations-perf} shows that the mutation generation
+machinery in \msprime\ is extremely efficient, requiring
+fractions of a second to generate mutations for most simulations.
+For example, the longest running simulation in
+Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
+generate an average of 1.5 million mutations over 137,081 trees
+in a tree sequence with 508,125 edges.
+Generating mutations for a tree sequence is particularly efficient
+because we can work edge-by-edge rather than having to
+consider each tree independently.
+
+\jkcomment{We probably want to keep a version of this narrative too.}
+For example, we simulated mutations under the JC69
+model for a tree with $10^6$ leaves over $10^4$ sites (resulting
+in $\sim 280000$ mutations) in less that 1.5 seconds, including
+the time required for file input and output. While a systematic
+comparison of run-times is not feasible here, this is orders
+of magnitude faster than methods such as
+Seq-Gen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}
+and at least competitive with the recent method from
+\cite{demaio2021phastsim}.
+
+\textbf{TODO: would love to do a like-for-like comparison, but quite hard}
+
+Another substantial advantage of our approach is that the simulated
+mutations are stored directly in the output data format with
+a defined API. Although
+Pyvolve and Phastsim are implemented in Python, neither provide
+programatic access to the simulated output, outputting instead
+to a series of files which must subsequently be parsed.
+See the Data Model section for more details and discussion.
+
+Things we can do: future work will need to do things like
+Dawg~\citep{cartwright2005dna} and to implement models and
+indels~\citep{fletcher2009indelible}.
+
 \subsection*{Simulation interface}
@@
 Demes format~\citep{gower2021demes} from the command line.
+
+\subsection*{Data interchange and interoperability}
+The point of simulations is to generate data and the efficiency with
+which simulated data can be incorporated into an overall application
+is an important---if often overlooked---issue.
+For example, Approximate Bayesian Computation
+(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
+and more recent machine learning
+methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
+perform population genetic inference by using large numbers of
+simulations as training data. Clearly, simulation efficiency is
+crucial since the size and number of simulations that can be performed determines
+the depth to which the model and parameter space can be sampled from.
+Equally important,
+however, is the efficiency with which the simulation results can be
+transformed into the specific input required by the inference method.
+In the case of ABC, this is usually a set of summary statistics of the sequence data.
+Modern approaches such as
+ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
+ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large numbers of user-specified statistics that are individually weakly informative,
+but jointly close to sufficient for the desired model or parameter.
+However, the computational resources needed to calculate these statistics can be even
+greater than those used to generate the simulated datasets.
+Older software such as DIYABC~\citep{cornuet2008inferring}
+or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
+developing their own specialised simulators, tightly integrating them
+with the inference method. While this is certainly more efficient,
+there is a great deal of duplication of effort and significant
+room for error.
+As ABC inference has become useful in an ever-widening range of applications,
+these all-in-one methods risk falling behind if they do not allow users to explore
+the models and statistics of most interest.
+
+\jkcomment{Stuff to cover}
+\begin{itemize}
+
+\item The core problem is that the field lacks an efficient mechanism for
+    interchanging simulated data, and so developers are forced to choose
+    between efficiency and duplication of effort.
+
+\item What are the output formats of simulators? Newick, or sequence data
+    basically. What's the problem with this? We don't have mutations.
+    Much more efficient to have the tree and mutations than the
+    sequence data. Fundamentally, the output of coalescent simulations
+    should be trees, otherwise we are throwing a lot of information
+    away.
+
+\item Side note somewhere: with efficient access to branch length
+    statistics~\citep{ralph2019efficiently}, maybe we don't need to
+    simulate mutations at all.
+[GT: I'm not so sure about this -- the desired statistics/summaries
+must be calculated on the observed dataset as well as on the many
+simulated datasets.
+Unless the observed dataset is stored as a tree sequence,
+(because it has been through tsinfer, for example),
+it's not clear how branch or node stats could be calculated
+from it.
+]
+
+\item Msprime uses the \tskit\ library~\citep{tskit2021tskit}, which provides
+    extremely efficient interchange and data processing. A quick example
+    of how efficient.
+
+\item Tskit has facilited unprecedented interoperability between simulators:
+    Forward simulations can also use tree
+    sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
+    3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
+    tree sequences. Msprime can take tree sequences simulated by these forward
+    simulators as input and complete the simulation (``recapitation'').
+    See \cite{haller2018tree} for more details.
+
+\item Recently, Phastsim/USHER~\cite{demaio2021phastsim,turakhia2021ultrafast}
+    have used this idea of storing the
+    tree plus mutations, but approach is crude compared to tskit.
+
+\end{itemize}
 

### 2021-07-19 paper.tex
More work on processing simulations.

--- 2021-07-19 paper.tex
@@
 
-\jkcomment{Make some sort of link here between this paper, which is
-about msprime simulations, but that we can't really disentangle
-simulations from their downstream processing.}
-The point of simulations is to generate data and the efficiency with
-which simulated data can be incorporated into an overall application
-is an important issue.
-For example, Approximate Bayesian Computation
+\jkcomment{We need a link paragraph here between what this paper is
+about (what we can simulate) and what the rest of this section is about
+(efficiently using the \emph{results} of simulations). This is where
+we should review the classical approaches. Then lead into the
+the discussion of why we need simulations to be efficiently processable,
+most importantly, in inference methods like ABC and ML.}
+
+The abilily to efficiently process simulation results is
+particularly important in simulation-based inference methods
+such as Approximate Bayesian Computation
 (ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
-and more recent machine learning
-methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
-perform population genetic inference by using large numbers of
-simulations as training data. Clearly, simulation efficiency is
+and machine learning based
+approaches~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable,
+sanchez2020deep}. Clearly, simulation efficiency is
 crucial since the size and number of simulations that can be performed determines
@@
 transformed into the specific input required by the inference method.
-In the case of ABC, this is usually a set of summary statistics of the sequence data.
-Modern approaches such as
+In the case of ABC, this is usually a set of summary statistics of the sequence
+data, and methods avoid the bottleneck of parsing
+text-based file formats to compute these statistics
+by either developing their own
+simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
+or creating forked versions of existing
+simulators~\cite[e.g.][]{pavlidis2010msabc,quinto2018modeling},
+tightly integrated with the inference method.
+While this is certainly more efficient,
+there is a great deal of duplication of effort and significant
+room for error (and propagation of bugs).
+Modern approaches to ABC such as
 ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
-ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large
-numbers of user-specified statistics that are individually weakly informative,
-but jointly close to sufficient for the desired model or parameter.
-However, the computational resources needed to calculate these statistics can be even
-greater than those used to generate the simulated datasets.
-Older software such as DIYABC~\citep{cornuet2008inferring}
-or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
-developing their own specialised simulators, tightly integrating them
-with the inference method. While this is certainly more efficient,
-there is a great deal of duplication of effort and significant
-room for error.
-As ABC inference has become useful in an ever-widening range of applications,
-these all-in-one methods risk falling behind if they do not allow users to explore
-the models and statistics of most interest.
-
-\jkcomment{Other potential stuff to cover}
-\begin{itemize}
-
-\item What are the output formats of simulators? Newick, or sequence data
-    basically. What's the problem with this? We don't have mutations.
-    Much more efficient to have the tree and mutations than the
-    sequence data. Fundamentally, the output of coalescent simulations
-    should be trees, otherwise we are throwing a lot of information
-    away.
-
-\item Lots of examples of papers where people have been forced to
-    create a modified fork of some simulator so that they can get
-    access to some internal details, not exposed by the simulators
-    interface. Tree sequences provide a complete record of the
-    ancestral process, so we can do things like compute IBD
-    segments and introgressed tracts directly on the output. See
-    also ARG section for recording more information, though.
-
-\item Side note somewhere: with efficient access to branch length
-    statistics~\citep{ralph2020efficiently}, maybe we don't need to
-    simulate mutations at all. (not in the ABC context, but elsewhere)
-
-\item Tskit has facilited unprecedented interoperability between simulators:
+ABC-NN~\citep{csillery2012abc,blum2010abc} use large
+numbers of weakly informative statistics,
+making the need to efficiently compute statistics from simulation
+results even more acute.
+
+Classical text based output formats like \ms\ are inefficient to process,
+but also lacking a great deal of important information.
+The tree-by-tree topology information output by simulators in Newick
+format lacks any concept of node (in the tree sequence sense) identity,
+and means that we cannot reliably infer information about ancestors
+from the output \jkcomment{Also there's the inherent limitation of
+newick using branch lenghts; I think there's some citations for
+the issues this causes. Probably in the original msprime paper.}.
+This has been worked around by using approximate definitions of
+concepts like identity by descent (cite carmi paper) \jkcomment{
+any other computation methods developed just to work around the
+crappiness of ms output?}
+Numerous forks
+of simulators have been created to access information not provided
+in the output. For example, \ms\  has been forked to
+output information about migrating segments~\citep{rosenzweig2016powerful},
+information about ancestral lineages~\citep{chen2013asymptotic},
+% Any more?
+and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
+has in turn been forked to output information on local
+ancestry~\citep{racimo2017archaic}.
+All of this information
+is either directly available by default in \msprime, or can be optionally
+stored via options such as \texttt{record\_migrations} or
+\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
+can be efficiently and conveniently processed via \tskit\ APIs.
+
+\jkcomment{Fit this in somewhere? Or maybe it should be a discussion point?}
+Tskit has facilited unprecedented interoperability between simulators:
 Forward simulations can also use tree
@@
 See \cite{haller2018tree} for more details.
-    (Although, this could be a discussion point?)
-
-\end{itemize}
 

### 2021-07-19 paper.tex

@@
 for four samples at ten sites.
+Figure based on~\citet{kelleher2018efficient}.
 }

### 2021-07-20 paper.tex

@@
 invoke the simulation engine, and process the results in some way.
+%% Maybe we can cite simprily below (https://www.sciencedirect.com/science/article/pii/S2352711018301213). Ariella worked on putting
+%% together a pipeline to facilite running simulations for ABC. 
 This process is cumbersome and labour intensive, and packages
-such \texttt{coala}~\citep{staab2016coala} have been developed
+such \texttt{coala}~\citep{staab2016coala} and \texttt{gladstein2018simprily} have been developed
 to allow simulations be run directly in a high-level

### 2021-07-20 paper.tex
More simulation frontend citations.

--- 2021-07-19 paper.tex
@@
 invoke the simulation engine, and process the results in some way.
-This process is cumbersome and labour intensive, and packages
-such \texttt{coala}~\citep{staab2016coala} have been developed
+This process is cumbersome and labour intensive, and
+a number of packages have been developed
 to allow simulations be run directly in a high-level
-scripting language.
+scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
 

### 2021-07-20 paper.tex
More citations for forks of ms

@@
 or creating forked versions of existing
-simulators~\cite[e.g.][]{pavlidis2010msabc,quinto2018modeling},
+simulators~\cite[e.g.][]{thornton2006approximate,
+hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
 tightly integrated with the inference method.
@@
 information about ancestral lineages~\citep{chen2013asymptotic},
+an array of population genetic statistics~\citep{ramos2007mlcoalsim}
 % Any more?

### 2021-07-20 paper.tex
Updates to the mutations section narrative.

@@
 % \address{
+
+%%Affiliations:
+% Franz Baumdicker:
+% Cluster of Excellence "Controlling Microbes to Fight Infections", Mathematical and Computational Population Genetics, University of Tübingen, Germany
 
@@
 
-\subsection*{Finite sites mutations}
+\subsection*{Simulating mutations}
 \label{sec-mutations}
+% Explain the background: what have people done classically?
+Because we are usually simulating neutral evolution in coalescent simulations,
+the problem of simulating generating variation can be decomposed into
+two independent steps:
+first simulating the genetic ancestry (the trees), and then simulating
+variation by superimposing neutral mutation processes on those trees.
+The classical \texttt{Seq-Gen}~\citep{rambaut1997seq} program allows a range
+of different models of sequence evolution to be imposed on a given
+tree in Newick format. Population genetic simulators have tended to
+include their own implementations of mutation simulation, with
+most just supporting the infinite sites
+model~\citep[e.g.][]{hudson2002generating}
+but in several cases including a range of different
+models of sequence evolution
+built-in~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal}.
+The ultimate output of the simulations of trees and mutations is
+then some representation of the resulting nucleotide sequences,
+and information about the simulated mutations is lost.
+
+\jkcomment{stuff to cover somewhere in this section}
+\begin{itemize}
+\item Tskit's approach of storing the mutations is a very simple idea,
+but has profound computational consequences. We can compute statistics
+about the sequences \emph{without ever seeing those sequences}.
+\item Mutation simulations is only included in msprime for historical reasons.
+It would be better to see msprime's mutation simulation as a separate
+package. The mutation and ancestry code in msprime have no priveledged
+access to each other, and mutation simulation would be no less efficient
+if done in a different package.
+\item Really emphasise this point: you can use msprime's mutation simulation
+on \empH{any} tskit tree sequence, produced by SLiM, fwdpp, etc or
+even tsinfer. It's completely modular.
+\end{itemize}
+
 Version 0.x of \texttt{msprime} provided a very simple mutation model,
@@
 \subsection*{Selection}
-Selection was added to the coalescent ... Lots of simulators exist:
-SelSim~\citep{spencer2004selsim}
-mbs~\citep{teshima2009mbs},
-msms~\citep{ewing2010msms},
-% What about Cosi, did it simulate selection?
-Cosi2~\citep{shlyakhter2014cosi2},
-and discoal~\citep{kern2016discoal}.
+Another ellaboration of the standard neutral coalescent with recombination
+is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
+In the context of the coalescent this amounts to creating a structured population
+during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
+in which lineages may transit between favored and unfavored backgrounds through
+recombination. This allows for many selective sweep scenarios, including recurrent, partial,
+and soft selective sweeps to be simulated efficiently, however in general this
+efficiency comes at the cost of flexibility in comparison to forwards in
+time simulation.
+
+Currently there are a small number of other simulators that will generate
+coalescent samples with sweeps including, SelSim~\citep{spencer2004selsim}
+mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
+Cosi2~\citep{shlyakhter2014cosi2}, and discoal~\citep{kern2016discoal}.
+Each of these packages has their own benefits and possible drawbacks,
+however none of them at present will output tree sequence formatted data.
 
@@
 
-We have implemented the basic infrastructure exists for the structured
-coalescent, which makes adding support for, e.g.,
-inversions straightforward~\citep{peischl2013sequential}
+We have implemented selective sweeps in \msprime\ similarly to
+the \texttt{discoal} implementation-- sweep trajectories are
+generated from a jump process approximation to the conditional diffusion
+of an allele bound for fixation~\citep{coop2004ancestral}.
+This implementation allows us to further generalize sweep models to
+changing population sizes as well as selection from standing variation,
+both of which are features that are planned for the near future.
+Moreover because this implements the basic infrastructure for the structured
+coalescent, adding support for, e.g.,
+inversions is straightforward~\citep{peischl2013sequential}.
 
@@
 \begin{center}
-% \includegraphics[width=6cm]{msprime-argon-runtimes.png}
+\includegraphics[width=6cm]{figures/dtwf-perf}
 \end{center}
@@
 Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
-[Others fill in their ACKs please]
+ADK was supported by NIH awards R01GM117241 and R01HG010774.
+Franz Baumdicker is funded by the Deutsche Forschungsgemeinschaft
+EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
 

### 2021-07-20 paper.tex
first draft of sweep section

--- 2021-07-20 paper.tex
@@
 \subsection*{Selection}
-Selection was added to the coalescent ... Lots of simulators exist:
-SelSim~\citep{spencer2004selsim}
-mbs~\citep{teshima2009mbs},
-msms~\citep{ewing2010msms},
-% What about Cosi, did it simulate selection?
-Cosi2~\citep{shlyakhter2014cosi2},
-and discoal~\citep{kern2016discoal}.
+Another ellaboration of the standard neutral coalescent with recombination
+is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
+In the context of the coalescent this amounts to creating a structured population
+during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
+in which lineages may transit between favored and unfavored backgrounds through 
+recombination. This allows for many selective sweep scenarios, including recurrent, partial, 
+and soft selective sweeps to be simulated efficiently, however in general this
+efficiency comes at the cost of flexibility in comparison to forwards in
+time simulation.
+
+Currently there are a small number of other simulators that will generate
+coalescent samples with sweeps including, SelSim~\citep{spencer2004selsim}
+mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
+Cosi2~\citep{shlyakhter2014cosi2}, and discoal~\citep{kern2016discoal}. 
+Each of these packages has their own benefits and possible drawbacks,
+however none of them at present will output tree sequence formatted data.
 
@@
 
-We have implemented the basic infrastructure exists for the structured
-coalescent, which makes adding support for, e.g.,
-inversions straightforward~\citep{peischl2013sequential}
+We have implemented selective sweeps in \msprime\ similarly to 
+the \texttt{discoal} implementation-- sweep trajectories are 
+generated from a jump process approximation to the conditional diffusion
+of an allele bound for fixation~\citep{coop2004ancestral}.
+This implementation allows us to further generalize sweep models to
+changing population sizes as well as selection from standing variation,
+both of which are features that are planned for the near future.
+Moreover because this implements the basic infrastructure for the structured
+coalescent, adding support for, e.g.,
+inversions is straightforward~\citep{peischl2013sequential}.
 

### 2021-07-21 evaluation/simulate-argon-genomelengths.py
Added scripts that simulate and plot ARGON against msprime.

@@ -1,0 +1,140 @@
+#!/usr/bin/env python3
+
+import tskit, msprime
+import subprocess
+import pandas as pd
+import time
+
+# Methods for simulating with ARGON and msprime.
+
+def sim_argon(sample_size, chrom_length, min_length=0, no_rec=False,
+    classpath="/Users/gtsambos/Software/ARGON/", ibd=True, trees=True):
+    # Defaults: mutation rate 1.65e-8, rec rate 1e-8
+    diploid_size = 2*sample_size
+    arguments = ['java',
+            '-jar', classpath + 'ARGON.0.1.jar',
+            '-N', '10000',
+            '-pop', '1', '%i' % diploid_size,
+            '-size', '%i' % int(chrom_length/1000000),
+            '-mut', '0',
+            '-quiet'
+            ]
+    if trees:
+        arguments.extend(['-trees'])
+    if ibd:
+        arguments.extend(['-IBD', '%i' % min_length])
+    if no_rec:
+        arguments.extend(['-rec', '0'])
+
+    process = subprocess.run(arguments)
+    stdout, stderr = process.communicate()
+    return (stdout, stderr)
+
+# Methods for simulating with msprime.
+
+def sim_msprime(sample_size, chrom_length, min_length=0, max_time=None, ibd=True,
+    no_rec=False):
+    if no_rec:
+        rho = 0
+    else:
+        rho = 1e-8
+    ts = msprime.sim_ancestry(
+        samples=sample_size, sequence_length=chrom_length,
+        discrete_genome=True,
+        recombination_rate=rho, population_size=5000,
+        model='dtwf')
+    # ts = msprime.sim_mutations(ts, rate=1e-20)
+    if not ibd:
+        return ts
+    else:
+        samples = list(itertools.combinations(range(sample_size), 2))
+        ibd = ts.tables.find_ibd(
+            min_length=min_length, samples=samples, max_time=max_time)
+        # print("Number of simulated trees:", ts.num_trees)
+        return ibd
+
+def sim_msprime_hybrid(sample_size, chrom_length, min_length=0, max_time=None, ibd=True,
+    no_rec=False):
+    if no_rec:
+        rho = 0
+    else:
+        rho = 1e-8
+    ts = msprime.sim_ancestry(
+        samples=sample_size, sequence_length=chrom_length,
+        discrete_genome=True,
+        recombination_rate=rho, population_size=5000,
+        model = [msprime.DiscreteTimeWrightFisher(duration=100),
[…]

### 2021-07-21 evaluation/generate_gc_perf_data.py
Update gc code to run in parallel.

@@ -1,3 +1,4 @@
+import concurrent.futures
 import subprocess
 import sys
 import tempfile
@@ -37,7 +38,9 @@ def run_simbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=False):
     return num_trees
 
 
-def run_fastsimbac(*, sample_size, L, gc_rate, gc_tract_length, set_seed=0, count_trees=False):
+def run_fastsimbac(
+    *, sample_size, L, gc_rate, gc_tract_length, set_seed=0, count_trees=False
+):
 
     # using R=2*gc_rate as gene conversion/recombination rate as SimBac uses R/2
     R = gc_rate * 2
@@ -89,6 +92,26 @@ def run_msprime(*, sample_size, L, gc_rate, gc_tract_length, ret_breakpoints = T
     return treenumber
 
 
+def run_benchmark(work):
+    tool, sample_size = work
+
+    L = 4_500_000
+    gc_rate = 0.015
+    gc_tract_length = 500
+
+    func = tool_map[tool]
+    before = time.perf_counter()
+    func(
+        sample_size=sample_size, L=L, gc_rate=gc_rate, gc_tract_length=gc_tract_length,
+    )
+    duration = time.perf_counter() - before
+    return {
+        "sample_size": sample_size,
+        "tool": name,
+        "time": duration / replicates,
+    }
+
+
 @click.command()
 @click.option("--replicates", type=int, default=1000)
 @click.option("--sample-size", type=int, default=10)
@@ -121,7 +144,7 @@ def validate(replicates, sample_size):
                 L=L,
                 gc_rate=gc_rate,
                 gc_tract_length=gc_tract_length,
-                set_seed = j,
+                set_seed=j,
                 count_trees=True,
             )
             nt_msprime[j], nb_msprime[j] = run_msprime(
@@ -151,7 +174,6 @@ def validate(replicates, sample_size):
 
     plt.close("all")
 
-
     sm.graphics.qqplot(nt_fastsimbac)
     sm.qqplot_2samples(nt_fastsimbac, nt_msprime, line="45")
     plt.xlabel("fastsimbac")
@@ -159,9 +181,17 @@ def validate(replicates, sample_size):
     plt.savefig("figures/verify_fastsimbac_v_msprime.png")
 
 
+tool_map = {
+    "msprime": run_msprime,
+    "SimBac": run_simbac,
+    "fastSimBac": run_fastsimbac,
+}
+
+
 @click.command()
[…]

### 2021-07-21 evaluation/generate_gc_perf_data.py
Fix errors and change to concurrent.futures.

@@ -1,6 +1,5 @@
 import concurrent.futures
 import subprocess
-import sys
 import tempfile
 import time
 
@@ -42,7 +41,7 @@ def run_fastsimbac(
     *, sample_size, L, gc_rate, gc_tract_length, set_seed=0, count_trees=False
 ):
 
-    # using R=2*gc_rate as gene conversion/recombination rate as SimBac uses R/2
+    # using R=2*gc_rate as gene conversion/recombination rate
     R = gc_rate * 2
     # Set theta to 0 to remove mutations (defaults to 0.01)
     args = f"{sample_size} {int(L)} -r {R} {gc_tract_length} -t 0"
@@ -107,8 +106,8 @@ def run_benchmark(work):
     duration = time.perf_counter() - before
     return {
         "sample_size": sample_size,
-        "tool": name,
-        "time": duration / replicates,
+        "tool": tool,
+        "time": duration,
     }
 
 
@@ -212,11 +211,14 @@ def benchmark_ecoli(replicates, processes):
             work.extend([(name, sample_size)] * replicates)
     data = []
     with concurrent.futures.ProcessPoolExecutor(max_workers=processes) as executor:
-        for datum in executor.map(run_benchmark, work):
-            print(datum)
-            data.append(datum)
+        futures = [executor.submit(run_benchmark, item) for item in work]
+        for future in concurrent.futures.as_completed(futures):
+            data.append(future.result())
+            print(data[-1])
             df = pd.DataFrame(data)
             df.to_csv("data/gc-perf.csv")
+
+
 
 @click.group()
 def cli():

### 2021-07-21 evaluation/plot.py
Update plotting code

@@ -58,13 +58,16 @@ def gc_perf():
     Plot the gene-conversion benchmark.
     """
     df = pd.read_csv("data/gc-perf.csv")
+    # Get the mean among replicates
+    dfg = df.groupby(["sample_size", "tool"]).mean().reset_index()
 
     fig, ax1 = plt.subplots(1, 1)
     for tool in set(df.tool):
-        dft = df[df.tool == tool]
+        dft = dfg[dfg["tool"] == tool]
         ax1.plot(dft.sample_size, dft.time, label=tool)
 
     ax1.set_xlabel("Sample size")
+    ax1.set_ylim(0, 3600)
     ax1.set_ylabel("Time (seconds)")
 
     ax1.legend()

### 2021-07-21 evaluation/plot.py
Refactor dtwf code to fit patterns.

@@ -15,9 +15,11 @@ import cpuinfo
 def cli():
     pass
 
+
 def save(name):
     plt.savefig(f"figures/{name}.png")
     plt.savefig(f"figures/{name}.pdf")
+
 
 @click.command()
 def mutations_perf():
@@ -68,6 +70,7 @@ def gc_perf():
     ax1.legend()
     save("gc-perf")
 
+
 @click.command()
 def sweeps_perf():
     """
@@ -88,6 +91,65 @@ def sweeps_perf():
 
 
 @click.command()
+def dtwf_perf():
+    """
+    Plot the DTWF benchark.
+    """
+
+    df = pd.read_csv(
+        "data/dtwf-perf.csv", sep="\t", usecols=["length", "program", "runtime"]
+    )
+
+    # Make a new dataframe with just the runtime means.
+    means = df.groupby(["length", "program"]).mean().reset_index()
+    sample_sizes = [0, 1, 2, 3, 4]
+    sample_size_labs = [
+        r"$10^{-2}$",
+        r"$10^{-1}$",
+        r"$10^{0}$",
+        r"$10^{1}$",
+        r"$10^{2}$",
+    ]
+
+    # print(means)
+
+    # # Plot the means
+    plt.plot(
+        sample_sizes,
+        np.log10(means[means["program"] == "msprime-DTWF"]["runtime"]),
+        label="msprime-DTWF",
+        marker="o",
+        color="C1",
+    )
+    plt.plot(
+        sample_sizes,
+        np.log10(means[means["program"] == "msprime-hybrid"]["runtime"]),
+        label="msprime-hybrid",
+        marker="o",
+        color="C1",
+        linestyle="dashed",
+    )
+    plt.plot(
+        sample_sizes,
+        np.log10(means[means["program"] == "argon"]["runtime"]),
+        label="argon",
+        marker="o",
+        color="C0",
+    )
+    plt.legend(loc="upper left")
+    plt.xlabel("Chromosome lengths (Mb)")
+    plt.ylabel("Runtimes (secs)")
+    plt.title("Simulation runtimes for different genome lengths")
+    plt.xticks(sample_sizes, labels=sample_size_labs)
+    plt.yticks(
+        [-2, -1, 0, 1, 2],
+        labels=[r"$10^{-2}$", r"$10^{1}$", r"$10^{0}$", r"$10^{1}$", r"$10^{2}$"],
+    )
[…]

### 2021-07-21 evaluation/generate_dtwf_perf_data.py
Update script style and point to local ARGON

@@ -1,140 +1,171 @@
-#!/usr/bin/env python3
-
-import tskit, msprime
-import subprocess
-import pandas as pd
 import time
+import subprocess
+
+import tskit
+import msprime
+import pandas as pd
 
 # Methods for simulating with ARGON and msprime.
 
-def sim_argon(sample_size, chrom_length, min_length=0, no_rec=False,
-    classpath="/Users/gtsambos/Software/ARGON/", ibd=True, trees=True):
-    # Defaults: mutation rate 1.65e-8, rec rate 1e-8
-    diploid_size = 2*sample_size
-    arguments = ['java',
-            '-jar', classpath + 'ARGON.0.1.jar',
-            '-N', '10000',
-            '-pop', '1', '%i' % diploid_size,
-            '-size', '%i' % int(chrom_length/1000000),
-            '-mut', '0',
-            '-quiet'
-            ]
-    if trees:
-        arguments.extend(['-trees'])
-    if ibd:
-        arguments.extend(['-IBD', '%i' % min_length])
-    if no_rec:
-        arguments.extend(['-rec', '0'])
 
-    process = subprocess.run(arguments)
-    stdout, stderr = process.communicate()
-    return (stdout, stderr)
+def sim_argon(
+    sample_size,
+    chrom_length,
+    min_length=0,
+    no_rec=False,
+    ibd=True,
+    trees=True,
+):
+    # Defaults: mutation rate 1.65e-8, rec rate 1e-8
+    diploid_size = 2 * sample_size
+    arguments = [
+        "java",
+        "-jar",
+        "tools/ARGON.jar",
+        "-N",
+        "10000",
+        "-pop",
+        "1",
+        "%i" % diploid_size,
+        "-size",
+        "%i" % int(chrom_length / 1_000_000),
+        "-mut",
+        "0",
+        "-quiet",
+    ]
+    if trees:
+        arguments.extend(["-trees"])
+    if ibd:
+        arguments.extend(["-IBD", "%i" % min_length])
+    if no_rec:
+        arguments.extend(["-rec", "0"])
+
+    output = subprocess.run(arguments, check=True, capture_output=True)
+    return (output.stdout, output.stderr)
+
 
 # Methods for simulating with msprime.
 
-def sim_msprime(sample_size, chrom_length, min_length=0, max_time=None, ibd=True,
-    no_rec=False):
+
+def sim_msprime(
[…]

### 2021-07-21 paper.tex
added ack
Refactor dtwf code to fit patterns.
Update

@@
 
-Tree sequences encode a complete and concise description of a set of sampled
-chromosome's genetic ancestry, and they can also efficiently represent
-the genetic variation (extant and ancestral).
-As illustrated in Fig.~\ref{fig-ts-example}, the \tskit\ tree
-sequence encoding includes \emph{site} and \emph{mutation} tables. Each
-row in the site table represents some position along the genome, and
-each row in the mutation table represents a change of allelic state
-at a particular site and node. Since the number of mutations is typically
-small, we can usually store the genetic variation for a given variant
-site in constant space. This approach of storing the genetic variation
-via the genetic ancestry is both very concise and extremely efficient
-to process---for example, computing Tajima's $D$ using \tskit\
-is several orders of magnitude faster for large sample sizes than
-efficient libraries working with the variant matrix
-representation~\citep{ralph2020efficiently}.
-\jkcomment{Mention Phastsim/USHER~\citep{demaio2021phastsim,turakhia2021ultrafast}
-here. Same basic idea, but much cruder.}
+% JK: commenting this out for now as we're covering it in the Mutations
+% section.
+
+% Tree sequences encode a complete and concise description of a set of sampled
+% chromosome's genetic ancestry, and they can also efficiently represent
+% the genetic variation (extant and ancestral).
+% As illustrated in Fig.~\ref{fig-ts-example}, the \tskit\ tree
+% sequence encoding includes \emph{site} and \emph{mutation} tables. Each
+% row in the site table represents some position along the genome, and
+% each row in the mutation table represents a change of allelic state
+% at a particular site and node. Since the number of mutations is typically
+% small, we can usually store the genetic variation for a given variant
+% site in constant space. This approach of storing the genetic variation
+% via the genetic ancestry is both very concise and extremely efficient
+% to process---for example, computing Tajima's $D$ using \tskit\
+% is several orders of magnitude faster for large sample sizes than
+% efficient libraries working with the variant matrix
+% representation~\citep{ralph2020efficiently}.
+% \jkcomment{Mention Phastsim/USHER~\citep{demaio2021phastsim,turakhia2021ultrafast}
+% here. Same basic idea, but much cruder.}
 
@@
 
+\jkcomment{A point to make somewhere in here: the tskit data model is
+integrated, with all of the information that you need in one file. Not
+splattered across a number of different files, that you have to keep track
+of.}
+
 \jkcomment{Fit this in somewhere? Or maybe it should be a discussion point?}
@@
 
-\subsection*{Finite sites mutations}
+\subsection*{Simulating mutations}
 \label{sec-mutations}
+% Explain the background: what have people done classically?
+Because coalescent simulations are usually concerned with
+neutral evolution (see the \nameref{sec-selection} section, however)
+the problem of generating synthetic genetic variation can be decomposed into
+two independent steps:
+simulating genetic ancestry (the trees), and then simulating
+variation by superimposing mutation processes on those trees.
+The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
+allows a range of different models of sequence evolution to be imposed on a given
+tree in Newick format. Population genetic simulators have tended to
+include their own implementations of mutation simulation, with
+most just supporting the infinite sites
+model~\citep[e.g.][]{hudson2002generating}
+but with several supporting a range of different models
+of sequence
+evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
+virgoulay2021gspace}.
+% Mutations are not treated as first-class data --- the sequences
+% were really all people were after, classically.
+The output of the combined simulations of ancestry and mutations is
+then some representation of the resulting nucleotide sequences
+(or perhaps some derived statistics of those sequences);
+information about the mutations themselves is not typically
+provided to the user.
+
+The approach taken in \msprime\ and \tskit\ is quite different.
+As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
+integral part of the tree sequence data model, and genetic
+variation is encoded by sites and the mutations at those sites.
+This may seem at first inconvenient, but is extremely powerful.
+Firstly, the storage space required for simulations is dramatically
+reduced. For a simulation of $n$ samples with $m$ variant sites,
+we required $O(nm)$ space to store the sequence data as a variant
+matrix. However, if this simulation was of a recombining genome
+with $t$ trees, then the \tskit\ tree sequence encoding
+requires $O(n + t + m)$ space~\citep{kelleher2016efficient}. For
+large sample sizes, this difference is profound, making it
+conceivable, for example, to store the genetic ancestry
+and variation data for the entire human population on a
+laptop~\citep{kelleher2019inferring}. Besides the huge
+difference in storage efficiency, it is also often far more efficient
+to compute statistics of the sequence data from the
+trees and mutations than it is to work with the sequences themselves.
+For example, computing Tajima's $D$ from simulated data using \tskit\
+is several orders of magnitude faster for large
+sample sizes~\citep{ralph2020efficiently}.
+
+Motivated by the large numbers of SARS-CoV-2 genomes currently
+available, \cite{turakhia2021ultrafast} propose a similar approach
+to encoding variation data for non-recombining organisms,
+where a Newick tree is annotated by mutations~\citep{mcbroome2021daily}.
+Similarly, the Phastsim~\citep{demaio2021phastsim} mutation simulator,
+developed primarily to support simulation of SARS-CoV-2 trees
+with hundreds of thousands of leaves, can optionally output a
+Newick tree, where nodes are annotated with the mutations.
+However, the dependence on the Newick format for encoding
+topological information means that it suffers from all the
+problems of newick [what?], and there is no software library support
+[confirm].
+
+\jkcomment{stuff to cover somewhere in this section}
+\begin{itemize}
+\item Mutation simulations is only included in msprime for historical reasons.
+It would be better to see msprime's mutation simulation as a separate
+package. The mutation and ancestry code in msprime have no priveledged
+access to each other, and mutation simulation would be no less efficient
+if done in a different package.
+\item Really emphasise this point: you can use msprime's mutation simulation
+on \emph{any} tskit tree sequence, produced by SLiM, fwdpp, etc or
+even tsinfer. It's completely modular.
+\end{itemize}
+
 Version 0.x of \texttt{msprime} provided a very simple mutation model,
@@
 \subsection*{Selection}
+\label{sec-selection}
 Another ellaboration of the standard neutral coalescent with recombination

### 2021-07-21 evaluation/generate_gc_perf_data.py
update fastSimBac evaluation and Affiliation/Acknowledgements

@@ -43,7 +43,7 @@ def run_fastsimbac(*, sample_size, L, gc_rate, gc_tract_length, set_seed=0, coun
     # Set theta to 0 to remove mutations (defaults to 0.01)
     args = f"{sample_size} {int(L)} -r {R} {gc_tract_length} -t 0"
     if count_trees:
-        args += " -T "
+        args += " -T -b 500"
     if set_seed > 0:
         args += " -s " + str(set_seed)
 
@@ -61,18 +61,31 @@ def run_fastsimbac(*, sample_size, L, gc_rate, gc_tract_length, set_seed=0, coun
     return num_trees
 
 
-def run_msprime(*, sample_size, L, gc_rate, gc_tract_length):
-    # We use an internal msprime API here because we want to get at the
-    # number of breakpoints, not the distinct trees.
-    sim = msprime.ancestry._parse_sim_ancestry(
+def run_msprime(*, sample_size, L, gc_rate, gc_tract_length, ret_breakpoints = True):
+    sim = msprime.sim_ancestry(
         samples=sample_size,
         sequence_length=L,
         ploidy=1,
         gene_conversion_rate=gc_rate,
         gene_conversion_tract_length=gc_tract_length,
     )
-    sim.run()
-    return sim.num_breakpoints
+    treenumber = sim.num_trees
+        
+    # We use an internal msprime API here because we want to get at the
+    # number of breakpoints, not the distinct trees.
+    if ret_breakpoints:
+        sim = msprime.ancestry._parse_sim_ancestry(
+            samples=sample_size,
+            sequence_length=L,
+            ploidy=1,
+            gene_conversion_rate=gc_rate,
+            gene_conversion_tract_length=gc_tract_length,
+        )
+        sim.run()
+        breakpointnumber = sim.num_breakpoints
+        return treenumber, breakpointnumber
+        
+    return treenumber
 
 
 @click.command()
@@ -91,6 +104,7 @@ def validate(replicates, sample_size):
     nt_simbac = np.zeros(replicates)
     nt_fastsimbac = np.zeros(replicates)
     nt_msprime = np.zeros(replicates)
+    nb_msprime = np.zeros(replicates)
 
     with click.progressbar(range(replicates)) as bar:
         for j in bar:
@@ -109,11 +123,12 @@ def validate(replicates, sample_size):
                 set_seed = j,
                 count_trees=True,
             )
[…]

### 2021-07-21 evaluation/generate_gc_perf_data.py
Run fastSimBac for all values.

@@ -64,7 +64,7 @@ def run_fastsimbac(
     return num_trees
 
 
-def run_msprime(*, sample_size, L, gc_rate, gc_tract_length, ret_breakpoints = True):
+def run_msprime(*, sample_size, L, gc_rate, gc_tract_length, ret_breakpoints=True):
     sim = msprime.sim_ancestry(
         samples=sample_size,
         sequence_length=L,
@@ -73,7 +73,7 @@ def run_msprime(*, sample_size, L, gc_rate, gc_tract_length, ret_breakpoints = T
         gene_conversion_tract_length=gc_tract_length,
     )
     treenumber = sim.num_trees
-        
+
     # We use an internal msprime API here because we want to get at the
     # number of breakpoints, not the distinct trees.
     if ret_breakpoints:
@@ -87,7 +87,7 @@ def run_msprime(*, sample_size, L, gc_rate, gc_tract_length, ret_breakpoints = T
         sim.run()
         breakpointnumber = sim.num_breakpoints
         return treenumber, breakpointnumber
-        
+
     return treenumber
 
 
@@ -206,8 +206,6 @@ def benchmark_ecoli(replicates, processes):
         for name in tool_map.keys():
             if sample_size > 50 and name == "SimBac":
                 continue
-            if sample_size > 400 and name == "fastSimBac":
-                continue
             work.extend([(name, sample_size)] * replicates)
     data = []
     with concurrent.futures.ProcessPoolExecutor(max_workers=processes) as executor:
@@ -217,7 +215,6 @@ def benchmark_ecoli(replicates, processes):
             print(data[-1])
             df = pd.DataFrame(data)
             df.to_csv("data/gc-perf.csv")
-
 
 
 @click.group()

### 2021-07-21 paper.tex
Updates on MAT etc.

@@
 
-Motivated by the large numbers of SARS-CoV-2 genomes currently
-available, \cite{turakhia2021ultrafast} propose a similar approach
-to encoding variation data for non-recombining organisms,
-where a Newick tree is annotated by mutations~\citep{mcbroome2021daily}.
-Similarly, the Phastsim~\citep{demaio2021phastsim} mutation simulator,
-developed primarily to support simulation of SARS-CoV-2 trees
-with hundreds of thousands of leaves, can optionally output a
-Newick tree, where nodes are annotated with the mutations.
-However, the dependence on the Newick format for encoding
-topological information means that it suffers from all the
-problems of newick [what?], and there is no software library support
-[confirm].
+The advantages of storing mutations rather than sequences have
+recently become clear in the context of vast genomic
+datasets produced during the SARS-CoV-2 pandemic.
+\cite{turakhia2021ultrafast} propose the
+Mutation Annotated Tree (MAT)
+format~\citep{mcbroome2021daily}---consisting of a Newick tree
+and associated mutations in a binary format---as an efficient way to store
+and process large viral datasets,
+and achieve excellent compression and processing performance.
+Similarly, \texttt{phastsim}~\citep{demaio2021phastsim}
+was developed to simulate sequence evolution on such large SARS-CoV-2
+phylogenies, and also outputs a tree annotated with mutations (not
+in MAT format, however) to avoid the bottleneck of generating
+and storing the simulated sequences.
+While these methods illustate the advantages of the general
+approach of storing ancestry and mutations over sequences,
+they do not generalise beyond their immediate settings, and
+no software library support is available.
+
+% on the other hand, tskit is awesome.
+Sites and mutations are fully integrated into the tree sequence data
+model and the \tskit\ library.
+As illustrated in Fig~\ref{fig-ts-example},
+each site corresponds to some position
+along the genome, and each mutation occurs at a given site, above
+a particular node, with a given derived state (along with other
+information, such as the time the mutation occured).
+% Now talk about why tskit allows us to easily generate mutations
+% on any input tree sequence topology.
 

### 2021-07-22 evaluation/plot.py
Finalise plotting

@@ -1,14 +1,8 @@
-import time
-import pathlib
 
 import matplotlib.pyplot as plt
 import pandas as pd
 import numpy as np
 import click
-import pyvolve
-import tskit
-import msprime
-import cpuinfo
 
 
 @click.group()
@@ -59,18 +53,34 @@ def gc_perf():
     """
     df = pd.read_csv("data/gc-perf.csv")
     # Get the mean among replicates
+    df.time /= 3600
     dfg = df.groupby(["sample_size", "tool"]).mean().reset_index()
 
     fig, ax1 = plt.subplots(1, 1)
-    for tool in set(df.tool):
+    lines = {}
+    for tool in ["SimBac", "fastSimBac", "msprime"]:
         dft = dfg[dfg["tool"] == tool]
-        ax1.plot(dft.sample_size, dft.time, label=tool)
+        (line,) = ax1.plot(dft.sample_size, dft.time, label=tool)
+        lines[tool] = line
 
     ax1.set_xlabel("Sample size")
-    ax1.set_ylim(0, 3600)
-    ax1.set_ylabel("Time (seconds)")
+    ax1.set_ylabel("Time (hours)")
+    # We could set a log-scale on the y-axis here but it really
+    # doesn't make much difference
 
+    dfmsp = dfg[dfg["tool"] == "msprime"]
+    largest_n = np.array(dfmsp.sample_size)[-1]
+    largest_value = np.array(dfmsp.time)[-1]
+    ax1.plot([largest_n], [largest_value], "o", color=lines["msprime"].get_color())
+    ax1.annotate(
+        f"{round(largest_value * 60)} mins",
+        textcoords="offset points",
+        xytext=(-30, 15),
+        xy=(largest_n, largest_value),
+        xycoords="data",
+    )
     ax1.legend()
+    plt.tight_layout()
     save("gc-perf")
 
 

### 2021-07-22 evaluation/generate_gc_perf_data.py
Notes on burnin

@@ -44,9 +44,11 @@ def run_fastsimbac(
     # using R=2*gc_rate as gene conversion/recombination rate
     R = gc_rate * 2
     # Set theta to 0 to remove mutations (defaults to 0.01)
-    args = f"{sample_size} {int(L)} -r {R} {gc_tract_length} -t 0"
+    # We need the -b 500 to set the burn-in so we get the correct distribution
+    # of trees.
+    args = f"{sample_size} {int(L)} -r {R} {gc_tract_length} -t 0 -b 500"
     if count_trees:
-        args += " -T -b 500"
+        args += " -T "
     if set_seed > 0:
         args += " -s " + str(set_seed)
 

### 2021-07-22 paper.tex
Update narrative for mutation generation

--- 2021-07-21 paper.tex
@@
 \newcommand{\scrm}[0]{\texttt{scrm}}
+\newcommand{\SLiM}[0]{\texttt{SLiM}}
+\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
 \newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
+\newcommand{\SeqGen}[0]{\texttt{Seq-Gen}}
+\newcommand{\Pyvolve}[0]{\texttt{Pyvolve}}
+\newcommand{\phastsim}[0]{\texttt{phastsim}}
 
@@
 then some representation of the resulting nucleotide sequences
-(or perhaps some derived statistics of those sequences);
-information about the mutations themselves is not typically
-provided to the user.
-
-The approach taken in \msprime\ and \tskit\ is quite different.
+(or some derived statistics of those sequences);
+information about the mutations that have been simulated
+is usually not retained or made available for subsequent analysis.
+
+The approach taken in \msprime\ is quite different.
 As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
-integral part of the tree sequence data model, and genetic
-variation is encoded by sites and the mutations at those sites.
-This may seem at first inconvenient, but is extremely powerful.
+integral part of \tskit's tree sequence data model, and genetic
+variation is encoded by sites and the mutations at those sites;
+the sequences themselves are never stored.
+This may seem inconvenient at first, but it is extremely powerful.
 Firstly, the storage space required for simulations is dramatically
@@
 with $t$ trees, then the \tskit\ tree sequence encoding
-requires $O(n + t + m)$ space~\citep{kelleher2016efficient}. For
+requires $O(n + t + m)$ space~\citep{kelleher2016efficient},
+assuming we have $O(1)$ mutations at each site. For
 large sample sizes, this difference is profound, making it
@@
 For example, computing Tajima's $D$ from simulated data using \tskit\
-is several orders of magnitude faster for large
-sample sizes~\citep{ralph2020efficiently}.
-
-Motivated by the large numbers of SARS-CoV-2 genomes currently
-available, \cite{turakhia2021ultrafast} propose a similar approach
-to encoding variation data for non-recombining organisms,
-where a Newick tree is annotated by mutations~\citep{mcbroome2021daily}.
-Similarly, the Phastsim~\citep{demaio2021phastsim} mutation simulator,
-developed primarily to support simulation of SARS-CoV-2 trees
-with hundreds of thousands of leaves, can optionally output a
-Newick tree, where nodes are annotated with the mutations.
-However, the dependence on the Newick format for encoding
-topological information means that it suffers from all the
-problems of newick [what?], and there is no software library support
-[confirm].
-
-\jkcomment{stuff to cover somewhere in this section}
-\begin{itemize}
-\item Mutation simulations is only included in msprime for historical reasons.
-It would be better to see msprime's mutation simulation as a separate
-package. The mutation and ancestry code in msprime have no priveledged
-access to each other, and mutation simulation would be no less efficient
-if done in a different package.
-\item Really emphasise this point: you can use msprime's mutation simulation
-on \emph{any} tskit tree sequence, produced by SLiM, fwdpp, etc or
-even tsinfer. It's completely modular.
-\end{itemize}
-
-Version 0.x of \texttt{msprime} provided a very simple mutation model,
-supporting only the infinite sites model.
-With version 1.0, we now support a large number of mutation models,
-including
+is several orders of magnitude faster than efficient variant matrix
+libraries for large sample sizes~\citep{ralph2020efficiently}.
+
+The advantages of storing mutations rather than sequences have
+recently become clear in the context of the vast genomic
+datasets produced during the SARS-CoV-2 pandemic.
+\cite{turakhia2021ultrafast} propose the
+Mutation Annotated Tree (MAT)
+format---consisting of a Newick tree
+and associated mutations in a binary format---and
+associated \texttt{matUtils} program as an efficient way to store
+and process large viral datasets~\citep{mcbroome2021daily},
+achieving excellent compression and processing performance.
+Similarly, \phastsim~\citep{demaio2021phastsim}
+was developed to simulate sequence evolution on such large SARS-CoV-2
+phylogenies, and also outputs a Newick tree annotated with mutations (not
+in MAT format, however) to avoid the bottleneck of generating
+and storing the simulated sequences.
+While these methods illustate the advantages of the general
+approach of storing ancestry and mutations over sequences,
+they do not generalise beyond their immediate settings, and
+no software library support is available.
+
+% on the other hand, tskit is awesome.
+The \tskit\ library provides first-class support for site and mutation
+data, fully integrated with the topological information encoded in the
+node and edge tables (Fig~\ref{fig-ts-example}). The mature and
+efficient C and Python interfaces makes it straightforward to add
+further information to an existing file, and to output the results.
+As such, it is essentially a historical accident
+that \msprime\ (whose main focus is on ancestry simulation)
+contains code to simulate mutations: the process could (and arguably should)
+be performed in a separate package, and there would be no performance
+penalty for doing so. It is important to note that \msprime's
+mutation generator can work with \emph{any} \tskit\ tree sequence,
+be it simulated using \SLiM~\citep{haller2019slim} or
+\fwdpy~\citep{thornton2014cpp}, or estimated from real data
+using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
+or \texttt{Relate}~\citep{speidel2019method}. It is a modular
+component intended to fit into a larger software ecosystem, and
+is in no way dependent on \msprime's ancestry simulator.
+
+For version 1.0 we have greatly extended the sophistication of
+\msprime's mutation generation engine, achieving near feature-parity
+with \SeqGen.
+We  support a large number of mutation models, including
 JC69~\citep{jukes1969evolution},
@@
 and efficiently defined in user code.
-Full support for the original infinite sites model is also maintained.
-The results of mutation simulations across a range of models are
-extensively validated
+Mutation rates can vary along the genome, and multiple mutation
+models can be imposed on a tree sequence by overlaying mutations
+in multiple passes.
+The new \texttt{sim\_mutations} function
+(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
+mutation model, but full support for the
+original infinite sites model is also maintained.
+We have extensively validated the results of mutation simulations
 against both theoretical expectations and output from
-Seq-Gen~\citep{rambaut1997seq} and
+\SeqGen~\citep{rambaut1997seq} and
 Pyvolve~\citep{spielman2015pyvolve}.
@@
 
+Generating simulations in \msprime\ is extremely efficient.
+Fig.~\ref{fig-mutations-perf} shows the time requried to generate
+mutations (using the default JC69 model) on
+simulated tree sequences for a variety of mutation
+rates as we vary the number of samples
+(Fig.~\ref{fig-mutations-perf}A) and the sequence
+length (Fig.~\ref{fig-mutations-perf}B).
+For example, the longest running simulation in
+Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
+generate an average of 1.5 million mutations over 137,081 trees
+in a tree sequence with $508,125$ edges.
+This efficiency for large numbers of trees is possible because
+the tree sequence encoding allows us to generate mutations
+on an edge-by-edge (see Fig.\ref{fig-ts-example}) basis, rather than
+tree-by-tree and branch-by-branch as would otherwise be required.
+In the above example from Fig.~\ref{fig-mutations-perf}B,
+we would have to consider $273,887,838$ branches
+since there are $137,081$ trees and $1,998$ branches in each
+tree ($\sim500$ times more work).
+
+Even when we consider a single tree, \msprime\ is far more
+efficient than methods such as \SeqGen\ and \Pyvolve,
+and has similar performance to \phastsim.
+
 Comparing the performance of mutation generation methods is problematic.
@@
 time required to perform the simulation.
-
-Fig.~\ref{fig-mutations-perf} shows that the mutation generation
-machinery in \msprime\ is extremely efficient, requiring
-fractions of a second to generate mutations for most simulations.
-For example, the longest running simulation in
-Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
-generate an average of 1.5 million mutations over 137,081 trees
-in a tree sequence with 508,125 edges.
-Generating mutations for a tree sequence is particularly efficient
-because we can work edge-by-edge rather than having to
-consider each tree independently.
 
@@
 \textbf{TODO: would love to do a like-for-like comparison, but quite hard}
-
-Another substantial advantage of our approach is that the simulated
-mutations are stored directly in the output data format with
-a defined API. Although
-Pyvolve and Phastsim are implemented in Python, neither provide
-programatic access to the simulated output, outputting instead
-to a series of files which must subsequently be parsed.
-See the Data Model section for more details and discussion.
 
@@
 \subsection*{Simulation interface}
+\label{sec-sim-interface}
 

### 2021-07-22 paper.tex
Finalise msprime benchmark.

@@
 
-Even when we consider a single tree, \msprime\ is far more
-efficient than methods such as \SeqGen\ and \Pyvolve,
-and has similar performance to \phastsim.
-
-Comparing the performance of mutation generation methods is problematic.
-Because there is no standard format for representing trees and
-the accompanying mutations, all existing methods are focused on
-writing out the resulting sequences. Methods are also not modular,
-and so it is very difficult to disentangle the time required to
-read in the tree file and output the sequences from the actual
-time required to perform the simulation.
-
-\jkcomment{We probably want to keep a version of this narrative too.}
-For example, we simulated mutations under the JC69
-model for a tree with $10^6$ leaves over $10^4$ sites (resulting
-in $\sim 280000$ mutations) in less that 1.5 seconds, including
-the time required for file input and output. While a systematic
-comparison of run-times is not feasible here, this is orders
-of magnitude faster than methods such as
-Seq-Gen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}
-and at least competitive with the recent method from
-\cite{demaio2021phastsim}.
-
-\textbf{TODO: would love to do a like-for-like comparison, but quite hard}
+% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
+Even if we consider a single tree, \msprime\ generates mutations
+extremely efficiently.
+For example, we simulated mutations under the BLOSUM62 amino
+acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
+in $\sim 260,000$ mutations) in about $0.8$ seconds, including
+the time required for file input and output.
+We do not attempt a systematic benchmarking of \msprime's
+mutation generation code against other methods because it is
+difficult to disentangle the effects of inefficient input and
+output formats from the mutation generation algorithms
+at this scale. It seems quite unlikely, however,
+that generating mutations with \msprime\ would be a bottleneck in any
+realistic analysis.
 

### 2021-07-22 paper.tex
Good draft of mutations section

@@
 allows a range of different models of sequence evolution to be imposed on a given
-tree in Newick format. Population genetic simulators have tended to
+tree in Newick format, and there are various
+extensions to the basic nucleotide
+models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
+Population genetic simulators have tended to
 include their own implementations of mutation simulation, with
@@
 Mutation Annotated Tree (MAT)
-format---consisting of a Newick tree
-and associated mutations in a binary format---and
-associated \texttt{matUtils} program as an efficient way to store
+format (consisting of a Newick tree
+and associated mutations in a binary format) and
+the \texttt{matUtils} program as an efficient way to store
 and process large viral datasets~\citep{mcbroome2021daily},
@@
 phylogenies, and also outputs a Newick tree annotated with mutations (not
-in MAT format, however) to avoid the bottleneck of generating
+in MAT format) to avoid the bottleneck of generating
 and storing the simulated sequences.
-While these methods illustate the advantages of the general
-approach of storing ancestry and mutations over sequences,
+While these methods illustrate the advantages of the general
+approach of storing ancestry and mutations rather than sequences,
 they do not generalise beyond their immediate settings, and
@@
 parameter models~\citep{kimura1980simple,kimura1981estimation}
-parameters models can be easily
-and efficiently defined in user code.
+can be easily and efficiently defined in user code.
 Mutation rates can vary along the genome, and multiple mutation
@@
 Generating simulations in \msprime\ is extremely efficient.
-Fig.~\ref{fig-mutations-perf} shows the time requried to generate
+Fig.~\ref{fig-mutations-perf} shows the time required to generate
 mutations (using the default JC69 model) on
@@
 generate an average of 1.5 million mutations over 137,081 trees
-in a tree sequence with $508,125$ edges.
+in a tree sequence with 508,125 edges.
 This efficiency for large numbers of trees is possible because
 the tree sequence encoding allows us to generate mutations
-on an edge-by-edge (see Fig.\ref{fig-ts-example}) basis, rather than
+on an edge-by-edge basis (see Fig.~\ref{fig-ts-example}), rather than
 tree-by-tree and branch-by-branch as would otherwise be required.
 In the above example from Fig.~\ref{fig-mutations-perf}B,
-we would have to consider $273,887,838$ branches
-since there are $137,081$ trees and $1,998$ branches in each
+we would have to consider 273,887,838 branches
+since there are 137,081 trees and 1,998 branches in each
 tree ($\sim500$ times more work).
@@
 acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
-in $\sim 260,000$ mutations) in about $0.8$ seconds, including
+in $\sim$260,000 mutations) in about $0.8$ seconds, including
 the time required for file input and output.
@@
 
-Things we can do: future work will need to do things like
-Dawg~\citep{cartwright2005dna} and to implement models and
-indels~\citep{fletcher2009indelible}.
+There are many ways in which the mutation generation code
+in \msprime\ could be extended by adapting existing methods.
+For example, we hope to add support for
+microsatellites~\citep{mailund2005coasim},
+codon models~\citep{arenas2007recodon}
+and indels~\citep{cartwright2005dna,fletcher2009indelible}.
 

### 2021-07-22 paper.tex
Update tree sequences section

@@
 Newick~\citep{felsenstein1989phylip} or
-VCF~\citep{danecek2011variant} based formats formats.
+VCF~\citep{danecek2011variant} based formats.
 This data structure has
@@
 One of the key reasons for \msprime's significiant performance advantage
-over other simulators is its usage of the ``succinct tree sequence''
-(usually, tree sequence) data stucture and the supporting \tskit\
-library~\citep{tskit2021tskit}.
+over other simulators~\citep{kelleher2016efficient}
+is its usage of the ``succinct tree sequence''
+(usually, tree sequence) data stucture
+The basic concepts were introduced by \cite{kelleher2016efficient}
+and originally developed as part of \msprime. We have subsequently
+extracted the tree sequence code from \msprime\ into
+\tskit~\citep{tskit2021tskit}, a standalone Python and C library.
 While a full discussion of tree sequences and the capabilities
-of \tskit\ is beyond the scope of this article, we briefly summarise
-here.
-\jkcomment{We need to mention that tree sequences and tskit are
-derived from the 2016 paper and have been extracted from msprime
-into a separate C/Python library.}
+of \tskit\ is beyond the scope of this article, we summarise
+the aspects that are important for simulation here.
 
@@
 for four samples at ten sites.
-Figure based on~\citet{kelleher2018efficient}.
 }
@@
 The succinct tree sequence is a data structure for concisely encoding
-genetic ancestry. Given a set of sample chromosomes, the topology
+genetic ancestry and variation. Given a set of sample chromosomes, the topology
 of their genetic ancestry is described by a set of \emph{nodes}
-and \emph{edges}. A node in tree sequence terminology corresponds
+and \emph{edges} (Fig.~\ref{fig-ts-example}).
+A node in tree sequence terminology corresponds
 to a single chromosome strand, and is associated with its birth time,
-birth population and other metadata. A node may optionally be
+population and other metadata. A node may optionally be
 associated with an \emph{individual}, so that, for example, a diploid
@@
 Nodes define haplotypes that exist at a specific time within a
-specific individual, and edges define genetic inheritence
+some individual, and edges define genetic inheritence
 relationships between these nodes. An edge consists of a
@@
 ``coalescence records''~\citep{kelleher2016efficient}, and
-is the key to the efficiency of tree sequences.
-\jkcomment{Summarise and add references?}
-
+is the key to the efficiency of tree
+sequences~\citep{
+kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
 The term Ancestral Recombination Graph (ARG) is often understood
-to mean the complete record of a sample's genetic
-ancestry~\citep[e.g.][]{mathieson2020ancestry}, and in this sense a tree
+to mean the complete record of the genetic
+ancestry of a set of sampled
+genomes~\citep[e.g.][]{mathieson2020ancestry},
+and in this sense a tree
 sequence \emph{is} an ARG. More technically, an ARG is a particular
@@
 In this sense, tree sequences can be used to \emph{encode}
-ARGs without loss of information---although we do not store
+ARGs without loss of information (although we do not store
 the additional ARG information by default
 in \msprime\ because the minimal tree sequence encoding is
-much more concise (see the \nameref{sec-arg} section for more details).
+much more concise; see the \nameref{sec-arg} section for more details).
 It should be noted that, despite efforts to
@@
 
-% JK: commenting this out for now as we're covering it in the Mutations
-% section.
-
-% Tree sequences encode a complete and concise description of a set of sampled
-% chromosome's genetic ancestry, and they can also efficiently represent
-% the genetic variation (extant and ancestral).
-% As illustrated in Fig.~\ref{fig-ts-example}, the \tskit\ tree
-% sequence encoding includes \emph{site} and \emph{mutation} tables. Each
-% row in the site table represents some position along the genome, and
-% each row in the mutation table represents a change of allelic state
-% at a particular site and node. Since the number of mutations is typically
-% small, we can usually store the genetic variation for a given variant
-% site in constant space. This approach of storing the genetic variation
-% via the genetic ancestry is both very concise and extremely efficient
-% to process---for example, computing Tajima's $D$ using \tskit\
-% is several orders of magnitude faster for large sample sizes than
-% efficient libraries working with the variant matrix
-% representation~\citep{ralph2020efficiently}.
-% \jkcomment{Mention Phastsim/USHER~\citep{demaio2021phastsim,turakhia2021ultrafast}
-% here. Same basic idea, but much cruder.}
-
-\jkcomment{We need a link paragraph here between what this paper is
-about (what we can simulate) and what the rest of this section is about
-(efficiently using the \emph{results} of simulations). This is where
-we should review the classical approaches. Then lead into the
-the discussion of why we need simulations to be efficiently processable,
-most importantly, in inference methods like ABC and ML.}
-
-The abilily to efficiently process simulation results is
+The point of running simulations is to generate data, and there
+are two important ways in which the tree sequences output by
+\msprime\ are more useful than classical approaches: (1) tree
+sequences are more efficient; and (2) tree sequences contain
+more information. We will consider these points in turn.
+
+The classical approach to generating simulation data is
+to render the results in some text format. For example,
+\ms~\citep{hudson2002generating} outputs a set of sequences
+and can also optionally output the marginal trees along
+the genome in Newick format,
+and variants of this approach are used by many simulators.
+Text files have many advantages, but are slow to process
+at scale. With sample sizes becoming ever larger, processing
+text output can become a significant bottleneck in many analyses.
+The ability to efficiently process simulation results is
 particularly important in simulation-based inference methods
@@
 and means that we cannot reliably infer information about ancestors
-from the output \jkcomment{Also there's the inherent limitation of
-newick using branch lenghts; I think there's some citations for
-the issues this causes. Probably in the original msprime paper.}.
-This has been worked around by using approximate definitions of
-concepts like identity by descent (cite carmi paper) \jkcomment{
-any other computation methods developed just to work around the
-crappiness of ms output?}
+from the output. Because Newick stores branch lengths rather
+than node times, numerical precision issues also arise when summing
+over many short branches~\citep{mcgill2013graphml}.
+% JK: not sure this is worth leaving in tbh
+This lack of reliable information about ancestors in simulation data
+has lead, for example, to the definition of segments of
+identity by descent as a set of samples having the same time
+to their most recent common ancestor along some section of the
+genome~\citep{yang2016rapidly}.
 Numerous forks
@@
 
-\jkcomment{A point to make somewhere in here: the tskit data model is
-integrated, with all of the information that you need in one file. Not
-splattered across a number of different files, that you have to keep track
-of.}
-
-\jkcomment{Fit this in somewhere? Or maybe it should be a discussion point?}
-Tskit has facilited unprecedented interoperability between simulators:
-Forward simulations can also use tree
-sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
-3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
-tree sequences. Msprime can take tree sequences simulated by these forward
-simulators as input and complete the simulation (``recapitation'').
-See \cite{haller2018tree} for more details.
+\jkcomment{Summary paragraph: tskit gives us complete information
+about the genetic ancestry. and is very efficient}
+
+% \jkcomment{A point to make somewhere in here: the tskit data model is
+% integrated, with all of the information that you need in one file. Not
+% splattered across a number of different files, that you have to keep track
+% of.}
 
@@
 % Also SARGE paper
-trees along the genome~\citep{mirzaei2016rent}
+trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}
 
@@
 to be done.
+
+Tskit has facilited unprecedented interoperability between simulators:
+Forward simulations can also use tree
+sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
+3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
+tree sequences. Msprime can take tree sequences simulated by these forward
+simulators as input and complete the simulation (``recapitation'').
+See \cite{haller2018tree} for more details.
 

### 2021-07-22 paper.tex
added appendix for sweeps
incorporated kevin and peter's comments

@@
 \usepackage{nameref}
+\usepackage{tikz}
 %\usepackage{showkeys}
@@
 \newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
+\newcommand{\plrcomment}[1]{\textcolor{cyan}{#1}}
 
@@
 
+% The data model
 \subsection*{Tree sequences}
 \label{sec-ts}
-One of the key reasons for \msprime's significiant performance advantage
-over other simulators~\citep{kelleher2016efficient}
-is its usage of the ``succinct tree sequence''
-(usually, tree sequence) data stucture
-The basic concepts were introduced by \cite{kelleher2016efficient}
-and originally developed as part of \msprime. We have subsequently
-extracted the tree sequence code from \msprime\ into
-\tskit~\citep{tskit2021tskit}, a standalone Python and C library.
-While a full discussion of tree sequences and the capabilities
-of \tskit\ is beyond the scope of this article, we summarise
-the aspects that are important for simulation here.
 
@@
 \begin{center}
-\includegraphics[width=6cm]{figures/example_tree_sequence}
+\includegraphics[width=6cm]{illustrations/example_tree_sequence}
 \end{center}
@@
 
-The succinct tree sequence is a data structure for concisely encoding
-genetic ancestry and variation. Given a set of sample chromosomes, the topology
+One of the key reasons for \msprime's substantial performance advantage
+over other simulators~\citep{kelleher2016efficient}
+is its usage of the ``succinct tree sequence''
+ data stucture to represent simulation results.
+The succinct tree sequence (usually abbreviated to ``tree sequence'')
+is a data structure for concisely encoding
+genetic ancestry and sequence variation.
+The basic concepts were introduced by \cite{kelleher2016efficient}
+and originally developed as part of \msprime. We subsequently
+extracted the tree sequence code that was independent of simulations
+from \msprime\ into \tskit,
+a standalone Python and C library~\citep{tskit2021tskit}.
+While a full discussion of tree sequences and the capabilities
+of \tskit\ is beyond the scope of this article, we summarise
+some aspects that are important for simulation here.
+
+% Encoding ancestry
+Given a set of sample chromosomes, the topology
 of their genetic ancestry is described by a set of \emph{nodes}
@@
 A node in tree sequence terminology corresponds
-to a single chromosome strand, and is associated with its birth time,
-population and other metadata. A node may optionally be
+to a single chromosome strand, associated with its birth time
+and other optional metadata. A node may be
 associated with an \emph{individual}, so that, for example, a diploid
-individual will be associated with two separate nodes (we omit
+individual is associated with two separate nodes (we omit
 the individual table in Fig.~\ref{fig-ts-example} as it is optional).
-Nodes define haplotypes that exist at a specific time within a
+Nodes define haplotypes that exist at a specific time within
 some individual, and edges define genetic inheritence
-relationships between these nodes. An edge consists of a
+relationships between these nodes. An \emph{edge} consists of a
 parent node, a child node and the left and right coordinates
@@
 0 to 4 is present in both marginal trees, and represented
-by a single edge (the first row in the example edge table).
+by a single edge (the first row in the edge table).
 This simple device, of explicitly associating tree nodes
@@
 genomic intervals over which parent/child relationships exist,
-is a generalisation from the original concept  of
-``coalescence records''~\citep{kelleher2016efficient}, and
-is the key to the efficiency of tree
+generalises the original
+``coalescence records'' concept~\citep{kelleher2016efficient},
+and is the key to the efficiency of tree
 sequences~\citep{
 kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
+
+% Quick digression: relationship to ARGs
 The term Ancestral Recombination Graph (ARG) is often understood
@@
 genomes~\citep[e.g.][]{mathieson2020ancestry},
-and in this sense a tree
+and in this sense a succinct tree
 sequence \emph{is} an ARG. More technically, an ARG is a particular
@@
 In this sense, tree sequences can be used to \emph{encode}
-ARGs without loss of information (although we do not store
-the additional ARG information by default
-in \msprime\ because the minimal tree sequence encoding is
-much more concise; see the \nameref{sec-arg} section for more details).
+ARGs without loss of information (although
+the default minimal tree sequence encoding is
+much more concise; see the \nameref{sec-arg} section).
 It should be noted that, despite efforts to
@@
 
+% OK, back to tree sequences: what about sequence variation?
+The final output of most population genetic simulations is some representation
+of sequence variation among the specified samples. For coalescent
+simulations, we usually have three steps: (1) simulate
+the genetic ancestry, and optionally output the resulting marginal trees;
+(2) simulate sequence evolution conditioned on this ancestry by generating
+mutations (see the \nameref{sec-mutations} section); and (3) output the
+resulting sequences by percolating the effects of the mutations through
+the trees. Information about the mutations
+themselves---e.g., where they have occurred on the
+trees---is usually not retained or made available for subsequent analysis.
+In \msprime, however, we skip step (3), instead using \tskit's
+data model of ancestry and mutations
+to represent the simulated sequences.
+As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
+fully integrated part of \tskit's tree sequence data model, and genetic
+variation is encoded by recording sites at which mutations
+have occurred, and where each mutation at those sites
+has occurred on the marginal tree.
+Crucially, the sequences themselves are never stored, or indeed directly
+represented in memory (unless explicitly decoded using \tskit\ APIs).
+It may at first seem inconvenient to have only this indirect representation
+of the genome sequences, but it is extremely powerful.
+Firstly, the storage space required for simulations is dramatically
+reduced. For a simulation of $n$ samples with $m$ variant sites,
+we would require $O(nm)$ space to store the sequence data as a variant
+matrix. However, if this simulation was of a recombining genome
+with $t$ trees, then the \tskit\ tree sequence encoding
+requires $O(n + t + m)$ space,
+assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
+For large sample sizes, this difference is profound, making it
+conceivable, for example, to store the genetic ancestry
+and variation data for the entire human population on a
+laptop~\citep{kelleher2019inferring}. Besides the huge
+difference in storage efficiency, it is also often far more efficient
+to compute statistics of the sequence data from the
+trees and mutations than it is to work with the sequences themselves.
+For example, computing Tajima's $D$ from simulated data using \tskit\
+is several orders of magnitude faster than efficient variant matrix
+libraries for large sample sizes~\citep{ralph2020efficiently}.
+
+The advantages of storing genetic variation data using the underlying trees
+have recently become clear in the context of the vast genomic
+datasets produced during the SARS-CoV-2 pandemic.
+\cite{turakhia2021ultrafast} propose the
+Mutation Annotated Tree (MAT)
+format (consisting of a Newick tree
+and associated mutations in a binary format) and
+the \texttt{matUtils} program as an efficient way to store
+and process large viral datasets~\citep{mcbroome2021daily},
+achieving excellent compression and processing performance.
+Similarly, \phastsim~\citep{demaio2021phastsim}
+was developed to simulate sequence evolution on such large SARS-CoV-2
+phylogenies, and also outputs a Newick tree annotated with mutations (not
+in MAT format) to avoid the bottleneck of generating
+and storing the simulated sequences.
+While these methods illustrate the advantages of the general
+approach of storing ancestry and mutations rather than sequences,
+they do not generalise beyond their immediate settings, and
+no software library support is available.
+% \msprime, on the other hand, provides a fast, powerful, and well-tested
+% generic interface for generating and storing mutations on trees.
+
+% The input - how do we run simulations?
+\subsection*{User interface}
+\label{sec-sim-interface}
+
+The majority of simulation packages are controlled either through
+a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
+a text-based input file
+format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
+or a mixture of both.
+Command line interfaces such as \ms\ make it very easy to run simple
+simulations, but as model complexity and the number of parameters specified increases
+they become difficult to understand and
+error-prone~\citep{ragsdale2020lessons,gower2021demes}.
+Specifying parameters through a text file alleviates this problem to a degree,
+but lacks flexibility, for example, when running simulations with parameters
+drawn from some distribution. In practice, for any significant simulation
+project (or at least, any reproducible one), users will write a script
+to generate the required command lines or input parameter files,
+invoke the simulation engine, and process the results in some way.
+This process is cumbersome and labour intensive, and
+a number of packages have been developed
+to allow simulations be run directly in a high-level
+scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
+
+The more recent trend has been to move away from this file and command-line
+driven approach and to instead provide direct interfaces to the simulation
+engines via an Application Programming Interface (API)~\citep[e.g.][]{
+thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
+The primary interface for \msprime\ is through a thoroughly documented and
+stable Python
+API, which has allowed a number of downstream tools to be built on top of
+% Found these through the Dependent packages on GitHub
+it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
+adrion2020community,adrion2020predicting, kamm2020efficiently,
+mckenzie2020ipcoal, montinaro2020revisiting,
+de2021geonomics,rivera2021simulation}.
+As well as providing a stable and efficient platform for building
+downstream applications, \msprime's Python API makes it much easier to
+build reproducible simulation pipelines, as the entire workflow can
+be encapsulated in a single script, and package and version
+dependencies can be explicitly stated using the \texttt{pip}
+or \texttt{conda} package managers.
+For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
+in the influential simulation analysis of
+\cite{martin2017human} were only detected because the pipeline
+could be easily run and reanalysed.
+
+A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
+designed in part to avoid sources of error (see the Demography section) but
+also to provide more appropriate defaults while keeping compatibility with
+existing code. In the new APIs, ancestry and mutation simulation are fully
+separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
+functions replacing the legacy \texttt{simulate} function. Among other changes,
+the new APIs default to discrete genome coordinates and finite sites mutations,
+resolving a major source of confusion and error. The 0.x APIs are fully
+supported and tested, and will be maintained for the foreseeable future.
+We also provide an \ms\ compatible
+command line interface to support existing workflows, and the
+\texttt{msp} program that supports specifying demographic models in
+Demes format~\citep{gower2021demes} from the command line.
+
+% The output - what do we get out?
+\subsection*{Data analysis}
+\label{sec-data-analysis}
 The point of running simulations is to generate data, and there
-are two important ways in which the tree sequences output by
-\msprime\ are more useful than classical approaches: (1) tree
+are two important ways in which the \tskit\ tree sequences output by
+\msprime\ are more useful than established methods: (1) tree
 sequences are more efficient; and (2) tree sequences contain
@@
 
-The classical approach to generating simulation data is
-to render the results in some text format. For example,
+The standard way of representing simulation data is
+to render the results in a text format. For example,
 \ms~\citep{hudson2002generating} outputs a set of sequences
@@
 Text files have many advantages, but are slow to process
-at scale. With sample sizes becoming ever larger, processing
-text output can become a significant bottleneck in many analyses.
+at scale.
 The ability to efficiently process simulation results is
@@
 and machine learning based
-approaches~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable,
-sanchez2020deep}. Clearly, simulation efficiency is
+approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
+flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
 crucial since the size and number of simulations that can be performed determines
-the depth to which the model and parameter space can be sampled from.
+the depth to which one can sample from the model and parameter space.
 Equally important,
@@
 tightly integrated with the inference method.
-While this is certainly more efficient,
-there is a great deal of duplication of effort and significant
-room for error (and propagation of bugs).
 Modern approaches to ABC such as
@@
 making the need to efficiently compute statistics from simulation
-results even more acute.
+results all the more acute.
 
 Classical text based output formats like \ms\ are inefficient to process,
-but also lacking a great deal of important information.
+but also lack a great deal of important information.
 The tree-by-tree topology information output by simulators in Newick
@@
 
-\jkcomment{Summary paragraph: tskit gives us complete information
-about the genetic ancestry. and is very efficient}
-
-% \jkcomment{A point to make somewhere in here: the tskit data model is
-% integrated, with all of the information that you need in one file. Not
-% splattered across a number of different files, that you have to keep track
-% of.}
-
+These problems, of inefficient and limited access to the simulated
+data are comprehensively solved by \msprime\ and \tskit. Complete
+information about ancestry and mutation simulations is immediately
+available to the user in a single integrated data structure
+(rather than multiple independent files), with no intermediate
+processing required.
+Users can subsequently compute precisely the information that they
+require using the powerful suite of population genetic statistics
+and other methods available in \tskit.
+
+% Stuff we simulate: mutations
 \subsection*{Simulating mutations}
 \label{sec-mutations}
+
 % Explain the background: what have people done classically?
@@
 simulating genetic ancestry (the trees), and then simulating
-variation by superimposing mutation processes on those trees.
-The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
-allows a range of different models of sequence evolution to be imposed on a given
-tree in Newick format, and there are various
-extensions to the basic nucleotide
-models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
-Population genetic simulators have tended to
+variation by superimposing mutation processes on those trees
+(see Fig.~\ref{fig-mutated-trees}).
+A number of programs exist to place mutations on trees: for instance,
+the classical \SeqGen\ program~\citep{rambaut1997seq}
+supports a range of different models of sequence evolution,
+and various extensions to the basic
+models have been proposed~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
+Partly for efficiency and partly in the interest of
+simplicity for users (i.e., to avoid intermediate text format conversions),
+population genetic simulators have tended to
 include their own implementations of mutation simulation, with
-most just supporting the infinite sites
+most supporting the infinite sites
 model~\citep[e.g.][]{hudson2002generating}
-but with several supporting a range of different models
-of sequence
+but with several supporting a wide range of different models of sequence
 evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
-virgoulay2021gspace}.
-% Mutations are not treated as first-class data --- the sequences
-% were really all people were after, classically.
-The output of the combined simulations of ancestry and mutations is
-then some representation of the resulting nucleotide sequences
-(or some derived statistics of those sequences);
-information about the mutations that have been simulated
-is usually not retained or made available for subsequent analysis.
-
-The approach taken in \msprime\ is quite different.
-As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
-integral part of \tskit's tree sequence data model, and genetic
-variation is encoded by sites and the mutations at those sites;
-the sequences themselves are never stored.
-This may seem inconvenient at first, but it is extremely powerful.
-Firstly, the storage space required for simulations is dramatically
-reduced. For a simulation of $n$ samples with $m$ variant sites,
-we required $O(nm)$ space to store the sequence data as a variant
-matrix. However, if this simulation was of a recombining genome
-with $t$ trees, then the \tskit\ tree sequence encoding
-requires $O(n + t + m)$ space~\citep{kelleher2016efficient},
-assuming we have $O(1)$ mutations at each site. For
-large sample sizes, this difference is profound, making it
-conceivable, for example, to store the genetic ancestry
-and variation data for the entire human population on a
-laptop~\citep{kelleher2019inferring}. Besides the huge
-difference in storage efficiency, it is also often far more efficient
-to compute statistics of the sequence data from the
-trees and mutations than it is to work with the sequences themselves.
-For example, computing Tajima's $D$ from simulated data using \tskit\
-is several orders of magnitude faster than efficient variant matrix
-libraries for large sample sizes~\citep{ralph2020efficiently}.
-
-The advantages of storing mutations rather than sequences have
-recently become clear in the context of the vast genomic
-datasets produced during the SARS-CoV-2 pandemic.
-\cite{turakhia2021ultrafast} propose the
-Mutation Annotated Tree (MAT)
-format (consisting of a Newick tree
-and associated mutations in a binary format) and
-the \texttt{matUtils} program as an efficient way to store
-and process large viral datasets~\citep{mcbroome2021daily},
-achieving excellent compression and processing performance.
-Similarly, \phastsim~\citep{demaio2021phastsim}
-was developed to simulate sequence evolution on such large SARS-CoV-2
-phylogenies, and also outputs a Newick tree annotated with mutations (not
-in MAT format) to avoid the bottleneck of generating
-and storing the simulated sequences.
-While these methods illustrate the advantages of the general
-approach of storing ancestry and mutations rather than sequences,
-they do not generalise beyond their immediate settings, and
-no software library support is available.
-
-% on the other hand, tskit is awesome.
-The \tskit\ library provides first-class support for site and mutation
-data, fully integrated with the topological information encoded in the
-node and edge tables (Fig~\ref{fig-ts-example}). The mature and
-efficient C and Python interfaces makes it straightforward to add
-further information to an existing file, and to output the results.
-As such, it is essentially a historical accident
-that \msprime\ (whose main focus is on ancestry simulation)
-contains code to simulate mutations: the process could (and arguably should)
-be performed in a separate package, and there would be no performance
-penalty for doing so. It is important to note that \msprime's
+virgoulay2021gspace}. Thus, despite the logical separation between
+the tasks of simulating ancestry and neutral sequence evolution,
+the two have been conflated in practice.
+
+\begin{figure}
+    \includegraphics{illustrations/unmutated_tree}
+    \includegraphics{illustrations/mutated_tree}
+    \caption{
+        \label{fig-mutated-trees}
+        Visualization of the separation between ancestry and mutation simulation:
+        \texttt{sim\_ancestry} produces ancestry
+        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
+        relating 200 samples),
+        and \texttt{sim\_mutations} adds mutations
+        (\textbf{right;} fourteen nucleotide mutations have been added).
+        Graphics produced with \tskit's \texttt{draw\_svg} method.
+        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
+        \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
+    }
+\end{figure}
+
+Part of the reason for this poor record of software reuse and modularity is the
+lack of standardised file formats, and in particular, the lack of common
+library infrastructure to abstract the details of interchanging simulation
+data. On the surface, it would appear that \msprime\ is also guilty of this
+lack of modularity since it supports simulating both ancestry and mutations.
+However, this is essentially a historical accident and because of the \tskit\
+library, the mutation process in \msprime\ could just as easily
+be performed in a separate package.
+The mature and efficient C and Python interfaces for
+\tskit\ makes it straightforward to add further information to
+an existing file, and because of its efficient data interchange mechanisms,
+there is no performance penalty for performing additional operations
+in a different Python package.
+Thanks to this interoperability, \msprime's
 mutation generator can work with \emph{any} \tskit\ tree sequence,
@@
 
-For version 1.0 we have greatly extended the sophistication of
-\msprime's mutation generation engine, achieving near feature-parity
-with \SeqGen.
-We  support a large number of mutation models, including
+As well as providing a new API (see the \nameref{sec-sim-interface} section)
+which emphasises this logical split between ancestry and mutation simulation,
+we have greatly extended the sophistication of
+\msprime's mutation generation engine for version 1.0,
+achieving near feature-parity with \SeqGen.
+We  support a large number of mutation models, including the
 JC69~\citep{jukes1969evolution},
@@
 and PAM~\citep{dayhoff1978} amino acid models.
-The general model accepts an arbitrary transition matrix between
+A general model accepts an arbitrary transition matrix between
 any number of alleles (which can be arbitrary unicode strings),
@@
 in multiple passes.
-The new \texttt{sim\_mutations} function
-(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
-mutation model, but full support for the
-original infinite sites model is also maintained.
 We have extensively validated the results of mutation simulations
@@
 
-Generating simulations in \msprime\ is extremely efficient.
+Simulating mutations in \msprime\ is efficient.
 Fig.~\ref{fig-mutations-perf} shows the time required to generate
@@
 the tree sequence encoding allows us to generate mutations
-on an edge-by-edge basis (see Fig.~\ref{fig-ts-example}), rather than
-tree-by-tree and branch-by-branch as would otherwise be required.
+on an edge-by-edge basis
+(see Fig.~\ref{fig-ts-example} and Appendix~\ref{app-mutation-algorithm}),
+rather than tree-by-tree and branch-by-branch as would otherwise be required.
 In the above example from Fig.~\ref{fig-mutations-perf}B,
@@
 tree ($\sim500$ times more work).
-
 % python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
-Even if we consider a single tree, \msprime\ generates mutations
-extremely efficiently.
+Even if we have a tree sequence consisting of a single tree
+(so that there is no advantage to working edge-by-edge),
+\msprime's mutation generator is still very efficient.
 For example, we simulated mutations under the BLOSUM62 amino
@@
 We do not attempt a systematic benchmarking of \msprime's
-mutation generation code against other methods because it is
+mutation generation code against other methods because at this scale it is
 difficult to disentangle the effects of inefficient input and
-output formats from the mutation generation algorithms
-at this scale. It seems quite unlikely, however,
+output formats from the mutation generation algorithms.
+It seems quite unlikely, however,
 that generating mutations with \msprime\ would be a bottleneck in any
@@
 There are many ways in which the mutation generation code
-in \msprime\ could be extended by adapting existing methods.
-For example, we hope to add support for
+in \msprime\ could be extended. For example, we hope to add support for
 microsatellites~\citep{mailund2005coasim},
 codon models~\citep{arenas2007recodon}
-and indels~\citep{cartwright2005dna,fletcher2009indelible}.
+and indels~\citep{cartwright2005dna,fletcher2009indelible},
+although some may also require upstream additions to \tskit's data model
+which is currently based on the assumption of independent sites.
 
@@
 
-Recombination is implemented in \msprime\ using Hudson's algorithm, which
+Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
 works backwards in time tracking the
@@
 sample sizes and several approaches have been proposed to
-allow simulations more closely approximate the coalescent
+allow simulations that more closely approximate the coalescent
 with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
@@
 \begin{figure}
-% \begin{center}
-% \begin{tabular}{cc}
-% % \includegraphics[width=5cm]{msprime-scaling.png} &
-% % \includegraphics[width=5cm]{figures/recombination} \\
-% \end{tabular}
-% \end{center}
-% \includegraphics[width=\textwidth]{figures/recombination}
-\Large{Plot of msprime run time against $\rho$}
-\caption{\label{fig-recombination-perf} Approximate run time for \msprime\
-simulations as a function of scaled recombination rate.
+\begin{center}
+    \includegraphics{figures/ancestry-perf}
+\end{center}
+\caption{\label{fig-ancestry-perf}
+    Approximate run time for \msprime\ simulations,
+    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
+    Each point is the runtime of one simulation,
+    for various values of effective population size ($N$),
+    genome length ($L$), and number of samples.
+    Runtime scales quadratically with the product of $N$ and $L$;
+    since recombination rate in these simulations was $10^{-8}$,
+    the scaled recombination rate here is $\rho = NL/10^8$,
+    shown on the horizontal axis.
+    The black line is a quadratic fit separately in each panel.
 }
@@
 very efficient, it is still quadratic in the scaled recombination rate
-$\rho$. This is because Hudson's algorithms tracks recombinations not only
+$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
+This is because Hudson's algorithms tracks recombinations not only
 in segments ancestral to the sample, but also between ancestral segments.
@@
 
-\jkcomment{We want to round out this discussion here by giving some
-sort of rough argument for why Hudson's algorithm is quadratic, and discussing
-to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
-to simulate in how long. See issue 28.}
+% \jkcomment{We want to round out this discussion here by giving some
+% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
+% to Fig.~\ref{fig-ancestry-perf} to show roughly what we can expect
+% to simulate in how long. See issue 28.}
+
+Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
+a given simulation of a constant size population,
+and verifies that runtime scales quadratically with $\rho$,
+(\citet{hein2004gene} gives an equation
+including the effects of sample size --
+see Appendix~\ref{app-theoretical-runtime}.)
+and depends much less strongly on the number of samples.
+Taking a typical chromosome to be 1 Morgan in length,
+these plot show, roughly, that simulating
+samples from a population of thousands of individuals takes seconds,
+while samples from a population of tens of thousands takes minutes.
+
+However, what about a simulation with \emph{changing} size?
+To understand how runtime depends on demography
+it helps to consider \emph{why} runtime is quadratic in $\rho$.
+At any point in time, \msprime\ must keep track of some number of lineages,
+each of which contains some number of chunks of genetic material.
+Common ancestor events reduce the number of lineages,
+and recombination events increase their number.
+However, with long genomes
+only a small fraction of the common ancestor events
+will involve overlapping segments of ancestry,
+and lead to coalescence in the marginal trees.
+Such disjoint segments are often far apart (on average, about distance $L/2$),
+and so recombine apart again immediately;
+it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
+\plrcomment{(this is discussed also in the ARG section; check for overlap)}
+The maximum number of lineages occurs when
+the increase and decrease in numbers of lineages due to
+common ancestor and recombination events balance out;
+to get an idea of runtime we can estimate when this occurs.
+Suppose that the maximal number of lineages is $M$;
+at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
+and the total rate of recombination is $M \ell$,
+where $\ell$ is the mean length of genome carried by each lineage
+(including ``trapped'' non-ancestral material).
+It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
+and at this maximum, recombination and coalescence rates are equal,
+so many (roughly half) of these lineages have at least two distant segments of ancestral material
+separated by a long segment of genome, implying that $\ell \approx L / 4$.
+This implies that $M$ is roughly $L N_e$.
+Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
+Since the total rate of events when there are the maximum number of lineages
+happens at rate roughly $L^2 N_e / 4$,
+then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
+as desired.
+What does this tell us about time-varying population sizes?
+The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
+in other words,
+more events happen in the time period from 0 to $N_e/2$ generations
+than from $N_e/2$ to $N_e$ generations,
+but the runtime is not dominated but the more recent time period.
+
+Putting this together:
+suppose that population size today is $N_1$,
+while longer ago than $T$ generations ago it was $N_2$:
+does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
+The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
+then runtime will be similar to a population of size $N_1$;
+while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
+For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
+and the runtime will depend critically on $T$ -- in other words,
+simulation will be quick in a species with a strong domestication bottleneck,
+and slow otherwise.
+In practice, we recommend that users estimate runtime by
+measuring runtime in a series of simulations with short genome lengths,
+possibly varying other parameters to check for interactions,
+and then predict runtime by fitting a quadratic curve to genome length.
+There is only a weak dependency on sample size
+-- notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
+only increases runtime by a factor of 2 or so.
 
@@
 in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
-simulation of organisms such as \textit{Drosophila melagaster}.
+simulation of organisms such as \textit{Drosophila melanogaster}.
 
@@
 However, gene conversions impact much shorter segments than crossover recombinations,
-typically below 1kb, and thus affects linkage pattern differently than
-recombination~\citep{korunes2017gene}.
+typically below 1kb, and thus affect linkage patterns differently than
+recombinations~\citep{korunes2017gene}.
 
@@
 Gene conversion is particularly useful to model homologous recombination in
-bacterial evolution.
-We compare the performance of \msprime\ with gene conversion to
+bacterial evolution,
+so we compare the performance of \msprime\ with gene conversion to
 SimBac~\citep{brown2016simbac}, a simulator developed specifically to
@@
 supports an arbitrary number of randomly mating populations exchanging migrants
-at specified rates. A range of demographic events are provided, which allow for
-varying population parameters and migration rates over time, as well as
-population splits, admixtures and pulse migrations. The location of lineages
-over time can be tracked in as much detail as required: each tree node is
+at specified rates. A range of demographic events are supported, which allow for
+varying population sizes and growth rates, changing migration rates over time,
+as well as population splits, admixtures and pulse migrations.
+The location of sampled lineages
+can be tracked through time in as much detail as required: each tree node is
 automatically associated with the population in which it arose, the location of
@@
 provided as a parameter to the new \texttt{sim\_ancestry} function,
-substantially reducing the potential for error. Another improvement over the
+substantially reducing the potential for error.
+Another improvement over the
 0.X APIs is the introduction of explicit population split and admixture events,
@@
 with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
-class supports importing Demes models. We also support importing demographic
+class supports importing and exporting Demes models.
+Models previously constructed using the 0.x API can be seamlessly imported into
+the \texttt{Demography} class, and we also support importing demographic
 models in the form of Newick species tree descriptions, as well as those
@@
 
+The \texttt{DemographyDebugger} can output detailed information for debugging
+demographic models, but also provides
+methods to compute a suite of numerical predictions about the location of
+lineages over time, for example the coalescence rates for two or more lineages
+drawn from populations at specified times in the past.
 Many popular approaches in population genetics use the distribution of
@@
 split times and subsequent migration rates between
-populations~\citep{wang2020tracking}. The \texttt{DemographyDebugger} provides
-methods to compute a suite of numerical predictions about the location of
-lineages over time, for example the coalescence rates for two or more lineages
-drawn from populations at specified times in the past. This provides a valuable
+populations~\citep{wang2020tracking}.
+The \texttt{DemographyDebugger} provides a valuable
 ground-truth when evaluating such inference methods, as illustrated by
@@
 \begin{figure}
+
 \begin{center}
-\includegraphics[width=7cm]{figures/arg}
+    \begin{tikzpicture}[scale=0.45]
+
+    \draw (-2,0) -- (-2,3) -- (0, 3);
+    \draw (1,0) -- (1,1);
+    \draw (0,1) -- (2,1) -- (2,2) -- (4,2) -- (4,0);
+    \draw (3,2) -- (3,5);
+    \draw (0,1) -- (0,3);
+    \draw (-1, 3) -- (-1, 5) -- (3,5);
+
+    \draw[fill] (-2,0) circle [radius=0.05];
+    \draw[fill] (-1,3) circle [radius=0.05];
+    \draw[fill] (1,0) circle [radius=0.05];
+    \draw[fill] (4,0) circle [radius=0.05];
+    \draw[fill] (1,1) circle [radius=0.05];
+    \draw[fill] (3,2) circle [radius=0.05];
+    \draw[fill] (1,5) circle [radius=0.05];
+
+    \node [below] at (-2,0) {0};
+    \node [below] at (1,0) {1};
+    \node [below] at (4,0) {2};
+    \node [above] at (1,1) {3};
+    \node [below right] at (1,1) {$x = 0.3$};
+    \node [above right] at (3,2) {4};
+    \node [above left] at (-1,3) {5};
+    \node [above] at (1,5) {6};
+
+    \end{tikzpicture}
+% \begin{center}
+% $\equiv$
+% \end{center}
+% \end{column}
+% \begin{column}{5cm}
+\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
+\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
+\includegraphics[width=3.8cm]{figures/arg.pdf}
+
 \end{center}
-\caption{\label{fig-arg} A figure illustrating the relative size of
-the full ARG representation compared the minimal tree sequence.
-\jkcomment{This figure is quite preliminary and we need to improve it.}}
+\caption{\label{fig-arg} (A) An simple ARG in which a recombination
+occurs at position 0.3; (B) the equivalent topology depicted as a tree
+sequence, including the recombination node; (C) the same tree sequence
+topology ``simplified'' down to its minimal tree sequence representation;
+(D) the fraction of nodes in a tree sequence that are concerned with
+extra ARG information in a tree sequences returned using \msprime's
+\texttt{record\_full\_arg} mode}
 \end{figure}
@@
 \label{sec-selection}
-Another ellaboration of the standard neutral coalescent with recombination
+Another elaboration of the standard neutral coalescent with recombination
 is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
@@
 during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
-in which lineages may transit between favored and unfavored backgrounds through
+in which lineages may transit between favoured and unfavoured backgrounds through
 recombination. This allows for many selective sweep scenarios, including recurrent, partial,
@@
 \begin{figure}
+\begin{center}
 \includegraphics[width=6cm]{figures/sweeps-perf}
+\end{center}
 \caption{\label{fig-selection-perf}Comparison of the runtime for
@@
 of an allele bound for fixation~\citep{coop2004ancestral}.
+Details of our implementation are available in the Appendix.
 This implementation allows us to further generalize sweep models to
@@
 
-\subsection*{Simulation interface}
-\label{sec-sim-interface}
-
-The majority of simulation packages are controlled either through
-a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
-a text-based input file
-format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
-or a mixture of both.
-Command line interfaces such as \ms\ make it very easy to run simple
-simulations, but as model complexity and the number of parameters specified increases
-they become difficult to understand and
-error-prone~\citep{ragsdale2020lessons,gower2021demes}.
-Specifying parameters through a text file alleviates this problem to a degree,
-but lacks flexibility, for example, when running simulations with parameters
-drawn from some distribution. In practice, for any significant simulation
-project (or at least, any reproducible one), users will write a script
-to generate the required command lines or input parameter files,
-invoke the simulation engine, and process the results in some way.
-This process is cumbersome and labour intensive, and
-a number of packages have been developed
-to allow simulations be run directly in a high-level
-scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
-
-The more recent trend has been to move away from this file and command-line
-driven approach and to instead provide direct interfaces to the simulation
-engines via an Application Programming Interface (API)~\citep[e.g.][]{
-thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
-The primary interface for \msprime\ is through a thoroughly documented and
-stable Python
-API, which has allowed a number of downstream tools to be built on top of
-% Found these through the Dependent packages on GitHub
-it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
-adrion2020community,adrion2020predicting, kamm2020efficiently,
-mckenzie2020ipcoal, montinaro2020revisiting,
-de2021geonomics,rivera2021simulation}.
-As well as providing a stable and efficient platform for building
-downstream applications, \msprime's Python API makes it much easier to
-build reproducible simulation pipelines, as the entire workflow can
-be encapsulated in a single script, and package and version
-dependencies can be explicitly stated using the \texttt{pip}
-or \texttt{conda} package managers.
-For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
-in the influential simulation analysis of
-\cite{martin2017human} were only detected because the pipeline
-could be easily run and reanalysed.
-
-A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
-designed in part to avoid sources of error (see the Demography section) but
-also to provide more appropriate defaults while keeping compatibility with
-existing code. In the new APIs, ancestry and mutation simulation are fully
-separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
-functions replacing the legacy \texttt{simulate} function. Among other changes,
-the new APIs default to discrete genome coordinates and finite sites mutations,
-resolving a major source of confusion and error. The 0.x APIs are fully
-supported and tested, and will be maintained for the foreseeable future.
-We also provide an \ms\ compatible
-command line interface to support existing workflows, and the
-\texttt{msp} program that supports specifying demographic models in
-Demes format~\citep{gower2021demes} from the command line.
-
 \subsection*{Development model}
@@
 and its use in understanding observed patterns of genetic variation.
-The basic model can be simulated for $n$ samples in $O(n)$ time
+The Kingman coalescent can be simulated for $n$ samples in $O(n)$ time
 and is straightforward to implement---for example, Hudson provided a fully
-functional and highly efficient simulator as a C source code listing in an
-appendix to the seminal review paper~\citep{hudson1990gene}. Because of
+functional and efficient simulator as a C source code listing
+in a paper appendix~\citep{hudson1990gene}. Because of
 this ease, the dominant approach to producing simulators for extensions
@@
 possible to implement the model efficiently, and only through the use
-of sophisticated data structures and
-intricate algorithms~\citep{kelleher2016efficient}.
+of sophisticated algorithms and data structures~\citep{kelleher2016efficient}.
 
@@
 
-Thus, development of \msprime\ is not completed and much remains
-to be done.
-
 Tskit has facilited unprecedented interoperability between simulators:
@@
 
+\section*{Theoretical runtime}
+    \label{app-theoretical-runtime}
+
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/ancestry-perf-expected}
+\end{center}
+\caption{\label{fig-expected-perf}
+    Data from Figure~\ref{fig-ancestry-perf}, compared with
+    equation~\eqref{eqn-hsw-bound} (``expected'').
+    Left panel shows observed times plotted against expected number
+    of operations; right panel shows the ratio of the two, plotted against $\rho$.
+}
+\end{figure}
+
+The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
+on the number of events involving trapped ancestral material in Hudson's algorithm.
+For $n$ samples from
+a diploid population of effective size $N_e$,
+and a genome of genetic length $r$,
+with $\rho = 4 N_e r$ the scaled recombination rate,
+and $R^T$ the number of recombination recombination events in trapped material,
+this upper bound is
+\begin{equation} \label{eqn-hsw-bound}
+    \mathbb{E}[R^T]
+    \le
+    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
+\end{equation}
+Note that the sum on the right is well-approximated by $\log(n)$.
+Although \citet{hein2004gene} refer to this bound as ``crude'',
+it turns out to predict the scaling of runtime quite well for large genomes.
+Figure~\ref{fig-expected-perf} shows the observed runtime
+(from the same simulations as Figure~\ref{fig-ancestry-perf})
+plotted against equation~\eqref{eqn-hsw-bound},
+showing that for values of $\rho$ above $4 \times 10^4$ or so
+(i.e., chromosome-length simulations in populations above $10^4$),
+the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
+This implies the runtime for a large sample
+can be predicted from the runtime for a small sample.
+For smaller values of $\rho$, however, the prediction
+severely underestimates runtime.
+
+\section*{Mutation generation}
+\label{app-mutation-algorithm}
+
+The algorithm that \msprime\ uses to simulate mutations on a tree sequence
+proceeds in two steps:
+first, mutations are ``placed'' on the tree sequence
+(i.e., sampling their locations in time, along the genome, and on the marginal tree),
+and then the ancestral and derived alleles of each mutation are generated.
+All mutation models share the code to place mutations,
+but choose alleles in different ways.
+
+First, mutations are placed on the tree sequence under an inhomgeneous Poisson model
+by applying them independently to each edge.
+If an edge spans a region $[a, b)$ of the genome
+and connected parent and child nodes with times $s < t$,
+and the mutation rate locally is $\mu$,
+then the number of mutations on the edge is Poisson with mean $\mu (t-s) (b-a)$,
+and each mutation is placed independently at a position chosen uniformly in $[a,b)$
+and a time uniformly in $[s,t)$.
+In a discrete genome, all positions are integers and so more than one mutation
+may occur at the same position on the same edge.
+Otherwise (i.e., for an infinite-sites model),
+positions are rejection sampled to obtain a unique floating-point number.
+If an edge spans a region of the genome with more than one mutation rate,
+this is done separately for each distinct region.
+The algorithm scales with the number of edges in the tree sequence.
+Since the number of edges is proportional to the number of marginal trees,
+an algorithm that applies mutations separately to each marginal tree
+would take at least as many steps,
+and would require additional bookkeeping necessary to place mutations uniformly on a tree.
+
+Next, alleles are chosen for each mutation.
+If the site was not previously mutated, then a new ancestral allele is chosen for the site,
+according to an input distribution of ancestral state allele probabilities.
+Then, each mutation on the tree is considered in turn,
+in order so that older mutations are considered first,
+and a derived allele is chosen based on the parental allele
+(which may be the ancestral allele or the derived allele of a previous mutation)
+and some randomness.
+Finally, information about the mutations are recorded in the site and mutation tables
+of the tree sequence.
+
+A mutation model must, therefore, provide two things:
+a way of choosing an ancestral allele for each new variant site,
+and a way of choosing a derived allele given the parental allele at each mutation.
+Perhaps the simplest mutation model implemented in \msprime\ is the
+\texttt{InfiniteAlleles} mutation model,
+which keeps an internal counter so that the requested alleles are 
+assigned subsequent (and therefore unique) integers.
+
+Since the algorithm above applies mutations at a single rate
+independent of ancestral state,
+a model in which different alleles mutate at different rates
+must necessarily produce some silent mutations,
+i.e., mutations in which the derived allele is equal to the parental allele.
+To illustrate this, consider a mutation model on the binary alphabet
+$\{X, Y\}$ in which $X$ mutates to $Y$ at rate $\alpha$
+and $Y$ mutates to $X$ at rate $\beta$, with $\beta < \alpha$.
+To implement this, first place mutations at the largest total rate, which is $\alpha$.
+Then, at each site, choose an ancestral allele from the ``root distribution''
+(a natural choice would be to choose from the stationary distribution,
+$X$ with probability $\beta / (\alpha + \beta)$),
+and for each mutation, choose a derived allele as follows:
+if the parental allele is $X$, then choose $Y$;
+if the parental allele is $Y$, then choose $X$ with probability $\beta/(\alpha + \beta)$,
+and choose $Y$ otherwise.
+This produces the correct distribution by Poisson thinning:
+a Poisson process with rate $\alpha$ in which each point is discarded independently
+with probability $\beta / (\alpha + \beta)$ is equivalent to a Poisson
+process with rate $\beta$.
+All finite-state models (implented under the generic \texttt{MatrixMutationModel} class)
+work in this way: mutations are placed at the maximum mutation rate,
+and then some silent mutations will result.
+
+In previous versions of \msprime, silent mutations were disallowed,
+and we could have removed them from the output entirely.
+However, we have chosen to leave them in, so that for instance
+simulating with the \texttt{HKY} mutation model will result in silent mutations
+if not all equilibrium frequencies are the same.
+Although the presence of silent mutations may at first be surprising
+(although they don't affect calculations of statistics in any way),
+there is a good reason to leave them in:
+to allow layering of different mutation models.
+Suppose that we wanted to model the mutation process as a mixture of more than one model,
+e.g., Jukes-Cantor mutations at rate $\mu_1$, and
+HKY mutations occur at rate $\mu_2$.
+Layering multiple calls to \texttt{sim_mutations} is allowed,
+so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
+and then add more with the \texttt{HKY} model at rate $\mu_2$.
+However, there is a small statistical problem:
+suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
+but then the HKY mutations inserts another mutation in the middle,
+resulting in $A \to C \to C$.
+If neither mutation model allows silent transitions,
+then this is clearly not correct,
+i.e., it is not equivalent to a model that simultaneously applies the two models.
+(The impact is small, however, as it only affects sites with more than one mutation.)
+The solution is to make the Jukes-Cantor model \emph{state-independent}
+(also called ``parent-independent''),
+by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
+\emph{independently} of the parent (so that 1/4 of mutations will be silent).
+If so -- and, more generally, if the first mutational process put down is state-independent --
+then the result of sequentially applying the two mutation models
+is equivalent to the simultaneous model.
+To facilitate this, many mutation models have a \texttt{state\_independent} option
+that increases the number of silent mutations
+and makes the model closer to state-independent.
+
+\section*{Likelihood calculations}
 \label{app-likelihoods}
-\section*{Likelihood calculations}
 
@@
 
+\label{app-sweeps}
+\section*{Selective sweep trajectories}
+We have implemented the coalescent with selective sweeps as a structured 
+coalescent process (as in~\citep{braverman1995hitchhiking}). During the
+``sweep phase'' of the simulation the rate of coalescence and recombination
+is tracked separately in lineages linked to the beneficial allele or not,
+with those rates depending on the frequency of the beneficial allele in 
+the population at a given time. The allele frequency trajectory of the
+sweeping allele is modeled using a jump process that approximates
+the conditional diffusion of the allele, conditioning on its eventual 
+fixation~\citep{coop2004ancestral}. 
+The jump process moves back in time following the beneficial
+allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
+the origination of the allele at $p=1/2N$, and we track time in small
+increments $\delta t$. This works such that the allele frequency at
+the next time step back in time, call it $p^\prime$ is
+
+\begin{eqnarray*}
+	p'=\left\{ \begin{array}{ll}
+	    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
+	    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
+	\end{array} \right.
+\end{eqnarray*}
+where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes.
+\mu(p)$ is
+\begin{eqnarray*}
+	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p)}
+\end{eqnarray*}
+
+While our current implementation only deals with genic selection,
+we plan in the future to add sweeps with arbitrary dominance. Further planned
+additions include trajectories of purely neutral alleles, as well as trajectories
+that model soft selective sweeps.
+
 \end{document}

### 2021-07-23 paper.tex
rearranging mutation section

@@
 
-\subsection*{Simulating mutations}
-\label{sec-mutations}
-% Explain the background: what have people done classically?
-Because coalescent simulations are usually concerned with
-neutral evolution (see the \nameref{sec-selection} section, however)
-the problem of generating synthetic genetic variation can be decomposed into
-two independent steps:
-simulating genetic ancestry (the trees), and then simulating
-variation by superimposing mutation processes on those trees.
-The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
-allows a range of different models of sequence evolution to be imposed on a given
-tree in Newick format, and there are various
-extensions to the basic nucleotide
-models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
-Population genetic simulators have tended to
-include their own implementations of mutation simulation, with
-most just supporting the infinite sites
-model~\citep[e.g.][]{hudson2002generating}
-but with several supporting a range of different models
-of sequence
-evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
-virgoulay2021gspace}.
+\subsection*{Genetic variation}
+\label{sec-genetic-variation}
+
+The output of population genetics simulators, quite naturally,
+tends to be some representation of the resulting nucleotide sequences
+(or some derived statistics of those sequences),
+although the underlying genealogies are also of interest.
+For instance,
+the classic simulator \ms\ will output trees or genome sequences,
+but not both.
 % Mutations are not treated as first-class data --- the sequences
 % were really all people were after, classically.
-The output of the combined simulations of ancestry and mutations is
-then some representation of the resulting nucleotide sequences
-(or some derived statistics of those sequences);
-information about the mutations that have been simulated
+However, joint information about the mutations that have been simulated
+-- e.g., where they have occurred on the trees -- 
 is usually not retained or made available for subsequent analysis.
@@
 integral part of \tskit's tree sequence data model, and genetic
-variation is encoded by sites and the mutations at those sites;
+variation is encoded by recording sites at which mutations
+have occurred, and where each mutation at those sites
+has occurred on the marginal tree;
 the sequences themselves are never stored.
-This may seem inconvenient at first, but it is extremely powerful.
+It may at first seem inconvenient to have only this indirect representation
+of the genome sequences, but it is extremely powerful.
 Firstly, the storage space required for simulations is dramatically
 reduced. For a simulation of $n$ samples with $m$ variant sites,
-we required $O(nm)$ space to store the sequence data as a variant
+we would require $O(nm)$ space to store the sequence data as a variant
 matrix. However, if this simulation was of a recombining genome
@@
 
-The advantages of storing mutations rather than sequences have
-recently become clear in the context of the vast genomic
+The advantages of storing genetic variation data using the underlying trees
+have recently become clear in the context of the vast genomic
 datasets produced during the SARS-CoV-2 pandemic.
@@
 no software library support is available.
+\msprime, on the other hand, provides a fast, powerful, and well-tested
+generic interface for generating and storing mutations on trees.
+
+\subsection*{Simulating mutations}
+\label{sec-mutations}
+
+% Explain the background: what have people done classically?
+Because coalescent simulations are usually concerned with
+neutral evolution (see the \nameref{sec-selection} section, however)
+the problem of generating synthetic genetic variation can be decomposed into
+two independent steps:
+simulating genetic ancestry (the trees), and then simulating
+variation by superimposing mutation processes on those trees.
+Indeed, to call a mutation ``neutral'' means that its presence or absence
+does not affect demography.
+A number of programs exist to place mutations on trees:
+in addition to \phastsim~\citep{demaio2021phastsim},
+the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
+allows a range of different models of sequence evolution to be imposed on a given
+tree in Newick format, and there are various
+extensions to the basic nucleotide
+models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
+Population genetic simulators have tended to
+include their own implementations of mutation simulation, with
+most just supporting the infinite sites
+model~\citep[e.g.][]{hudson2002generating}
+but with several supporting a range of different models
+of sequence
+evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
+virgoulay2021gspace}.
 
 % on the other hand, tskit is awesome.
-The \tskit\ library provides first-class support for site and mutation
-data, fully integrated with the topological information encoded in the
-node and edge tables (Fig~\ref{fig-ts-example}). The mature and
+Clean separation of the ancestry and mutation generation steps
+is made possible by the \tskit\ library, whose mature and
 efficient C and Python interfaces makes it straightforward to add
@@
 that \msprime\ (whose main focus is on ancestry simulation)
-contains code to simulate mutations: the process could (and arguably should)
-be performed in a separate package, and there would be no performance
-penalty for doing so. It is important to note that \msprime's
+contains code to simulate mutations: the mutation process could
+be performed by a separate package, and there would be no performance
+penalty for doing so.
+Thanks to this interoperability, \msprime's
 mutation generator can work with \emph{any} \tskit\ tree sequence,

### 2021-07-23 paper.tex
added diagram of mutated trees

@@
 \newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
+\newcommand{\plrcomment}[1]{\textcolor{purple}{#1}}
 
@@
 simulating genetic ancestry (the trees), and then simulating
-variation by superimposing mutation processes on those trees.
+variation by superimposing mutation processes on those trees
+(see figure~\ref{fig-mutated-trees}).
 Indeed, to call a mutation ``neutral'' means that its presence or absence
@@
 virgoulay2021gspace}.
+
+\begin{figure}
+    \includegraphics{figures/unmutated_tree}
+    \includegraphics{figures/mutated_tree}
+    \caption{
+        \label{fig-mutated-trees}
+        Visualization of the separation between ancestry and mutation simulation:
+        \texttt{sim\_ancestry} produces ancestry
+        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
+        relating 200 samples),
+        and \texttt{sim\_mutations} adds mutations
+        (\textbf{right;} fourteen nucleotide mutations have been added).
+        Graphics produced with \tskit's \texttt{draw\_svg} method.
+        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
+    }
+\end{figure}
 

### 2021-07-23 pretty_pictures.py

@@ -1,0 +1,40 @@
+#!/usr/bin/env python
+
+import msprime
+
+
+## Adding mutations
+# 
+# The goal here is to make a figure with (a) a tree and (b) some mutations added to it.
+
+
+ts = msprime.sim_ancestry(
+        50,
+        population_size=1e4,
+        recombination_rate=0.5e-8,
+        sequence_length=1000,
+        random_seed=14
+)
+
+model = msprime.F84(kappa=2)
+mts = msprime.sim_mutations(
+        ts,
+        rate=5e-8,
+        model=model,
+        random_seed=5
+)
+
+
+def do_svg(ts, **kwargs):
+    ts.draw_svg(
+            size=(300,200),
+            node_labels={},
+            mutation_labels={m.id: m.derived_state for m in ts.mutations()},
+            symbol_size=5,
+            force_root_branch=True,
+            **kwargs
+    )
+
+do_svg(ts, path="figures/unmutated_tree.svg")
+do_svg(mts, path="figures/mutated_tree.svg")
+

### 2021-07-23 paper.tex

@@
 although the underlying genealogies are also of interest.
-For instance,
-the classic simulator \ms\ will output trees or genome sequences,
-but not both.
 % Mutations are not treated as first-class data --- the sequences
@@
 (see figure~\ref{fig-mutated-trees}).
-Indeed, to call a mutation ``neutral'' means that its presence or absence
-does not affect demography.
 A number of programs exist to place mutations on trees:
-in addition to \phastsim~\citep{demaio2021phastsim},
+for instance,
 the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}

### 2021-07-23 paper.tex
paragraphs about runtime

@@
 \newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
-\newcommand{\plrcomment}[1]{\textcolor{purple}{#1}}
+% @plr - purple doesn't exist in my Latex
+\newcommand{\plrcomment}[1]{\textcolor{blue}{#1}}
 
@@
 \begin{figure}
-    \includegraphics{figures/unmutated_tree}
-    \includegraphics{figures/mutated_tree}
+    \includegraphics{illustrations/unmutated_tree}
+    \includegraphics{illustrations/mutated_tree}
     \caption{
@@
 
-\jkcomment{We want to round out this discussion here by giving some
-sort of rough argument for why Hudson's algorithm is quadratic, and discussing
-to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
-to simulate in how long. See issue 28.}
+% \jkcomment{We want to round out this discussion here by giving some
+% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
+% to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
+% to simulate in how long. See issue 28.}
+
+Figure~\ref{fig-recombination-perf} gives a rough idea of the runtime of
+a given simulation of a constant size population.
+However, what about a simulation with \emph{changing} size?
+To understand how runtime depends on demography
+it helps to consider \emph{why} runtime is quadratic in $\rho$.
+At any point in time, \msprime\ must keep track of some number of lineages,
+each of which contains some number of chunks of genetic material.
+Common ancestor events reduce the number of lineages,
+and recombination events increase their number.
+However, with long genomes
+only a small fraction of the common ancestor events
+will involve overlapping segments of ancesty,
+and lead to coalescence in the marginal trees.
+Such disjoint segments are often far apart (on average, about distance $L/2$),
+and so recombine apart again immediately;
+it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
+\plrcomment{(this is discussed also in the ARG section; check for overlap)}
+The maximum number of lineages occurs when
+the increase and decrease in numbers of lineages due to
+common ancestor and recombination events balance out;
+to get an idea of runtime we can estimate when this occurs.
+Suppose that the maximal number of lineages is $M$;
+at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
+and the total rate of recombination is $M \ell$,
+where $\ell$ is the mean length of genome carried by each lineage
+(including ``trapped'' non-ancestral material).
+It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
+and at this maximum, recombination and coalescence rates are equal,
+so many (roughly half) of these lineages have at least two distant segments of ancestral material
+separated by a long segment of genome, implying that $\ell \approx L / 4$.
+This implies that $M$ is roughly $L N_e$.
+Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
+Since the total rate of events when there are the maximum number of lineages
+happens at rate roughly $L^2 N_e / 4$,
+then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
+as desired.
+What does this tell us about time-varying population sizes?
+The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
+in other words,
+more events happen in the time period from 0 to $N_e/2$ generations
+than from $N_e/2$ to $N_e$ generations,
+but the runtime is not dominated but the more recent time period.
+
+Putting this together:
+suppose that population size today is $N_1$,
+while longer ago than $T$ generations ago it was $N_2$:
+does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
+The answer depends on how $T$ commpares to $N_1$: if $T/N_1$ is large,
+then runtime will be similar to a population of size $N_1$;
+while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
+For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
+and the runtime will depend critically on $T$ -- in other words,
+simulation will be quick in a species with a strong domestication bottleneck,
+and slow otherwise.
 

### 2021-07-24 pretty_pictures.py
Fixup makefile and add ARG pictures

@@ -1,40 +1,90 @@
-#!/usr/bin/env python
-
+import click
+import tskit
 import msprime
 
 
-## Adding mutations
-# 
-# The goal here is to make a figure with (a) a tree and (b) some mutations added to it.
-
-
-ts = msprime.sim_ancestry(
+@click.command()
+def mutated_tree():
+    """
+    Make a figure with (a) a tree and (b) some mutations added to it.
+    """
+    ts = msprime.sim_ancestry(
         50,
         population_size=1e4,
         recombination_rate=0.5e-8,
         sequence_length=1000,
-        random_seed=14
-)
+        random_seed=14,
+    )
 
-model = msprime.F84(kappa=2)
-mts = msprime.sim_mutations(
-        ts,
-        rate=5e-8,
-        model=model,
-        random_seed=5
-)
+    model = msprime.F84(kappa=2)
+    mts = msprime.sim_mutations(ts, rate=5e-8, model=model, random_seed=5)
 
 
-def do_svg(ts, **kwargs):
-    ts.draw_svg(
-            size=(300,200),
+    def do_svg(ts, **kwargs):
+        ts.draw_svg(
+            size=(300, 200),
             node_labels={},
             mutation_labels={m.id: m.derived_state for m in ts.mutations()},
             symbol_size=5,
             force_root_branch=True,
             **kwargs
-    )
+        )
 
-do_svg(ts, path="figures/unmutated_tree.svg")
-do_svg(mts, path="figures/mutated_tree.svg")
 
+    do_svg(ts, path="illustrations/unmutated_tree.svg")
+    do_svg(mts, path="illustrations/mutated_tree.svg")
+
+
+@click.command()
+def arg_ts():
+    tables = tskit.TableCollection(1.0)
+    tables.nodes.add_row(flags=tskit.NODE_IS_SAMPLE, time=0)
+    tables.nodes.add_row(flags=tskit.NODE_IS_SAMPLE, time=0)
+    tables.nodes.add_row(flags=tskit.NODE_IS_SAMPLE, time=0)
+    tables.nodes.add_row(flags=msprime.NODE_IS_RE_EVENT, time=0.5)
+    # tables.nodes.add_row(flags=msprime.NODE_IS_RE_EVENT, time=0.5)
+    tables.nodes.add_row(flags=0, time=1.0)
+    tables.nodes.add_row(flags=0, time=1.5)
[…]

### 2021-07-24 pretty_pictures.py
Fixup mutations illustration

@@ -21,12 +21,21 @@ def mutated_tree():
 
 
     def do_svg(ts, **kwargs):
+
+        # Sizes are probably all wrong here
+        style = """\
+            @media print {
+              @page { margin: 0; size: 3in 2in}
+              body { margin: 1.6cm; }
+            }
+        """
         ts.draw_svg(
             size=(300, 200),
             node_labels={},
             mutation_labels={m.id: m.derived_state for m in ts.mutations()},
             symbol_size=5,
             force_root_branch=True,
+            style=style,
             **kwargs
         )
 

### 2021-07-25 paper.tex
runtime figures

--- 2021-07-23 paper.tex
@@
 \newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
-% @plr - purple doesn't exist in my Latex
-\newcommand{\plrcomment}[1]{\textcolor{blue}{#1}}
+\newcommand{\plrcomment}[1]{\textcolor{cyan}{#1}}
 
@@
 \begin{figure}
-% \begin{center}
-% \begin{tabular}{cc}
-% % \includegraphics[width=5cm]{msprime-scaling.png} &
-% % \includegraphics[width=5cm]{figures/recombination} \\
-% \end{tabular}
-% \end{center}
-% \includegraphics[width=\textwidth]{figures/recombination}
-\Large{Plot of msprime run time against $\rho$}
-\caption{\label{fig-recombination-perf} Approximate run time for \msprime\
-simulations as a function of scaled recombination rate.
+\begin{center}
+    \includegraphics{figures/ancestry-perf}
+\end{center}
+\caption{\label{fig-ancestry-perf}
+    Approximate run time for \msprime\ simulations,
+    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
+    Each point is the runtime of one simulation,
+    for various values of effective population size ($N$),
+    genome length ($L$), and number of samples.
+    Runtime scales quadratically with the product of $N$ and $L$;
+    since recombination rate in these simulations was $10^{-8}$,
+    the scaled recombination rate here is $\rho = NL/10^8$,
+    shown on the horizontal axis.
+    The black line is a quadratic fit separately in each panel.
+    \plrcomment{TODO: run out more simulations to get more triangles}
 }
@@
 % sort of rough argument for why Hudson's algorithm is quadratic, and discussing
-% to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
+% to Fig.~\ref{fig-ancestry-perf} to show roughly what we can expect
 % to simulate in how long. See issue 28.}
 
-Figure~\ref{fig-recombination-perf} gives a rough idea of the runtime of
-a given simulation of a constant size population.
+Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
+a given simulation of a constant size population,
+and verifies that runtime scales quadratically with $\rho$,
+and depends much less strongly on the number of samples.
+Taking a typical chromosome to be 1 Morgan in length,
+these plot show, roughly, that simulating
+samples from a population of thousands of individuals takes seconds,
+while samples from a population of tens of thousands takes minutes.
+
 However, what about a simulation with \emph{changing} size?
@@
 and slow otherwise.
+In practice, we recommend that users estimate runtime by
+measuring runtime in a series of simulations with short genome lengths,
+possibly varying other parameters to check for interactions,
+and then predict runtime by fitting a quadratic curve to genome length.
 

### 2021-07-25 evaluation/generate_ancestry_perf_data.py

@@ -1,0 +1,42 @@
+import time
+
+import msprime
+import numpy as np
+
+def run(L, N, num_samples):
+    start = time.process_time()
+    ts = msprime.sim_ancestry(
+            samples=num_samples,
+            population_size=N,
+            sequence_length=L,
+            recombination_rate=1e-8)
+    end = time.process_time()
+    return (max(ts.tables.nodes.time), ts.num_trees, end - start)
+
+
+def csv(x):
+    return ",".join(map(str, x)) + "\n"
+
+
+def run_sims():
+    outfile = "data/ancestry-perf.csv"
+    num_reps = 3
+    num_samples = [1000, 100000]
+    L_VALS = [1e6, 5e6, 1e7, 5e7, 1e8]
+    N_VALS = [1000, 5000, 10000, 50000, 100000, 200000, 300000]
+    ln_pairs = [(L, N) for L in L_VALS for N in N_VALS]
+    ln_pairs.sort(key=lambda x: x[0] * x[1])
+
+    with open(outfile, "w") as f:
+        f.write(csv(["N", "L", "num_samples", "height", "num_trees", "time"]))
+        for L, N in ln_pairs:
+            for ns in num_samples:
+                if L * N * np.log10(ns) < 3e12:
+                    for _ in range(num_reps):
+                        print(f"... L={L}, N={N}, num_samples={ns}, L*N={int(L*N)}.")
+                        h, n, t = run(L, N, ns)
+                        f.write(csv([N, L, ns, h, n, t]))
+                        f.flush()
+
+if __name__ == "__main__":
+    run_sims()

### 2021-07-25 evaluation/plot.py

@@ -1,4 +1,5 @@
 
+import matplotlib
 import matplotlib.pyplot as plt
 import pandas as pd
 import numpy as np
@@ -11,6 +12,7 @@ def cli():
 
 
 def save(name):
+    plt.tight_layout()
     plt.savefig(f"figures/{name}.png")
     plt.savefig(f"figures/{name}.pdf")
 
@@ -163,6 +165,115 @@ def dtwf_perf():
 
 
 @click.command()
+def ancestry_perf():
+    """
+    Plot the ancestry benchark.
+    """
+
+    df = pd.read_csv(
+        "data/ancestry-perf.csv", sep=",",
+    )
+
+    # Plot the times
+    fig, axes = plt.subplots(1, 2, figsize=(6, 3))
+    rgb = matplotlib.cm.get_cmap("Set1")(np.linspace(0.0, 1.0, len(set(df["N"]))))
+    time_lims = [5, 1e12]
+    for ax, tl  in zip(axes, time_lims):
+        legend_adds = []
+        for ns, m in zip((1000, 100000), ("o", "v")):
+            legend_adds.append(
+                    matplotlib.lines.Line2D(
+                        [], [], color='grey', marker=m,
+                        linestyle="none",
+                        label=f"num samples={ns}"
+                    )
+            )
+            ut = np.logical_and(
+                    df["time"] <= tl,
+                    df["num_samples"] == ns
+            )
+            # fit a qudratic with no intercept:
+            # in R this would be lm(time ~ 0 + L : N + I((N*L)^2))
+            rho = 4 * df["N"] * df["L"] / 1e8
+            X = np.empty((sum(ut), 3))
+            X[:,0] = rho[ut]
+            X[:,1] = ( rho ** 2 )[ut]
+            X[:,2] = 1
+            b, _, _, _ = np.linalg.lstsq(X, df["time"][ut], rcond=None)
+
+            ax.set_xlabel("rho (4NL/1e8)")
+            ax.set_ylabel("time (seconds)")
+            Nvals = sorted(list(set(df["N"])))
+            for k, Nval in enumerate(Nvals):
+                utN = np.logical_and(ut, df["N"] == Nval)
+                if np.sum(utN) > 0:
+                    sargs = {}
+                    if m == "o":
+                        sargs["label"] = f"N={Nval}"
+                    ax.scatter(
+                        rho[utN],
[…]

### 2021-07-26 paper.tex
Move the "Genetic variation" section into tree sequences.

@@
 \label{sec-ts}
-One of the key reasons for \msprime's significiant performance advantage
-over other simulators~\citep{kelleher2016efficient}
-is its usage of the ``succinct tree sequence''
-(usually, tree sequence) data stucture
-The basic concepts were introduced by \cite{kelleher2016efficient}
-and originally developed as part of \msprime. We have subsequently
-extracted the tree sequence code from \msprime\ into
-\tskit~\citep{tskit2021tskit}, a standalone Python and C library.
-While a full discussion of tree sequences and the capabilities
-of \tskit\ is beyond the scope of this article, we summarise
-the aspects that are important for simulation here.
 
@@
 
-The succinct tree sequence is a data structure for concisely encoding
-genetic ancestry and variation. Given a set of sample chromosomes, the topology
+One of the key reasons for \msprime's substantial performance advantage
+over other simulators~\citep{kelleher2016efficient}
+is its usage of the ``succinct tree sequence''
+ data stucture to represent simulation results.
+The succinct tree sequence (usually abbreviated to tree sequence)
+is a data structure for concisely encoding
+genetic ancestry and sequence variation.
+The basic concepts were introduced by \cite{kelleher2016efficient}
+and originally developed as part of \msprime. We subsequently
+extracted the tree sequence code that was independent of simulations
+from \msprime\ into \tskit,
+a standalone Python and C library~\citep{tskit2021tskit}.
+While a full discussion of tree sequences and the capabilities
+of \tskit\ is beyond the scope of this article, we summarise
+some aspects that are important for simulation here.
+
+% Encoding ancestry
+Given a set of sample chromosomes, the topology
 of their genetic ancestry is described by a set of \emph{nodes}
@@
 to a single chromosome strand, and is associated with its birth time,
-population and other metadata. A node may optionally be
+and other optional metadata. A node may optionally be
 associated with an \emph{individual}, so that, for example, a diploid
@@
 the individual table in Fig.~\ref{fig-ts-example} as it is optional).
-Nodes define haplotypes that exist at a specific time within a
+Nodes define haplotypes that exist at a specific time within
 some individual, and edges define genetic inheritence
-relationships between these nodes. An edge consists of a
+relationships between these nodes. An \emph{edge} consists of a
 parent node, a child node and the left and right coordinates
@@
 genomes~\citep[e.g.][]{mathieson2020ancestry},
-and in this sense a tree
+and in this sense a succint tree
 sequence \emph{is} an ARG. More technically, an ARG is a particular
@@
 libraries.
+
+% OK, back to tree sequences: what about sequence variation?
+The final output of most population simulations is some representation
+of the sequence variation among the specified samples. For coalescent
+simulations, we usually have three steps (1) simulate
+the genetic ancestry, and optionally output the resulting marginal trees;
+(2) simulate sequence evolution conditioned on this ancestry by generating
+mutations (see the \nameref{sec-mutations} section); and (3) output the
+resulting sequences by percolating the effects of the mutations through
+the trees. Information about the mutations
+themselves---e.g., where they have occurred on the
+trees---is usually not retained or made available for subsequent analysis.
+In \msprime, however, we skip step (3), instead using \tskit's
+data model of ancestry and mutations
+to represent the simulated sequences.
+As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
+fully integrated part of \tskit's tree sequence data model, and genetic
+variation is encoded by recording sites at which mutations
+have occurred, and where each mutation at those sites
+has occurred on the marginal tree.
+Crucially, the sequences themselves are never stored, or indeed directly
+represented in memory (unless explicitly decoded using \tskit\ APIs).
+It may at first seem inconvenient to have only this indirect representation
+of the genome sequences, but it is extremely powerful.
+Firstly, the storage space required for simulations is dramatically
+reduced. For a simulation of $n$ samples with $m$ variant sites,
+we would require $O(nm)$ space to store the sequence data as a variant
+matrix. However, if this simulation was of a recombining genome
+with $t$ trees, then the \tskit\ tree sequence encoding
+requires $O(n + t + m)$ space,
+assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
+For large sample sizes, this difference is profound, making it
+conceivable, for example, to store the genetic ancestry
+and variation data for the entire human population on a
+laptop~\citep{kelleher2019inferring}. Besides the huge
+difference in storage efficiency, it is also often far more efficient
+to compute statistics of the sequence data from the
+trees and mutations than it is to work with the sequences themselves.
+For example, computing Tajima's $D$ from simulated data using \tskit\
+is several orders of magnitude faster than efficient variant matrix
+libraries for large sample sizes~\citep{ralph2020efficiently}.
+
+The advantages of storing genetic variation data using the underlying trees
+have recently become clear in the context of the vast genomic
+datasets produced during the SARS-CoV-2 pandemic.
+\cite{turakhia2021ultrafast} propose the
+Mutation Annotated Tree (MAT)
+format (consisting of a Newick tree
+and associated mutations in a binary format) and
+the \texttt{matUtils} program as an efficient way to store
+and process large viral datasets~\citep{mcbroome2021daily},
+achieving excellent compression and processing performance.
+Similarly, \phastsim~\citep{demaio2021phastsim}
+was developed to simulate sequence evolution on such large SARS-CoV-2
+phylogenies, and also outputs a Newick tree annotated with mutations (not
+in MAT format) to avoid the bottleneck of generating
+and storing the simulated sequences.
+While these methods illustrate the advantages of the general
+approach of storing ancestry and mutations rather than sequences,
+they do not generalise beyond their immediate settings, and
+no software library support is available.
+% \msprime, on the other hand, provides a fast, powerful, and well-tested
+% generic interface for generating and storing mutations on trees.
+
+\subsection*{Simulating mutations}
+\label{sec-mutations}
+
+% Explain the background: what have people done classically?
+Because coalescent simulations are usually concerned with
+neutral evolution (see the \nameref{sec-selection} section, however)
+the problem of generating synthetic genetic variation can be decomposed into
+two independent steps:
+simulating genetic ancestry (the trees), and then simulating
+variation by superimposing mutation processes on those trees
+(see Fig.~\ref{fig-mutated-trees}).
+A number of programs exist to place mutations on trees:
+for instance,
+the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
+allows a range of different models of sequence evolution to be imposed on a given
+tree in Newick format, and there are various
+extensions to the basic
+models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
+Population genetic simulators have tended to
+include their own implementations of mutation simulation, with
+most supporting the infinite sites
+model~\citep[e.g.][]{hudson2002generating}
+but with several supporting a range of different models
+of sequence
+evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
+virgoulay2021gspace}.
+
+\begin{figure}
+    \includegraphics{illustrations/unmutated_tree}
+    \includegraphics{illustrations/mutated_tree}
+    \caption{
+        \label{fig-mutated-trees}
+        Visualization of the separation between ancestry and mutation simulation:
+        \texttt{sim\_ancestry} produces ancestry
+        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
+        relating 200 samples),
+        and \texttt{sim\_mutations} adds mutations
+        (\textbf{right;} fourteen nucleotide mutations have been added).
+        Graphics produced with \tskit's \texttt{draw\_svg} method.
+        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
+    }
+\end{figure}
+
+% on the other hand, tskit is awesome.
+Clean separation of the ancestry and mutation generation steps
+is made possible by the \tskit\ library, whose mature and
+efficient C and Python interfaces makes it straightforward to add
+further information to an existing file, and to output the results.
+As such, it is essentially a historical accident
+that \msprime\ (whose main focus is on ancestry simulation)
+contains code to simulate mutations: the mutation process could
+be performed by a separate package, and there would be no performance
+penalty for doing so.
+Thanks to this interoperability, \msprime's
+mutation generator can work with \emph{any} \tskit\ tree sequence,
+be it simulated using \SLiM~\citep{haller2019slim} or
+\fwdpy~\citep{thornton2014cpp}, or estimated from real data
+using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
+or \texttt{Relate}~\citep{speidel2019method}. It is a modular
+component intended to fit into a larger software ecosystem, and
+is in no way dependent on \msprime's ancestry simulator.
+
+For version 1.0 we have greatly extended the sophistication of
+\msprime's mutation generation engine, achieving near feature-parity
+with \SeqGen.
+We  support a large number of mutation models, including
+JC69~\citep{jukes1969evolution},
+F84~\citep{felsenstein1996hidden},
+and GTR~\citep{tavare1986some} nucleotide models
+and the BLOSUM62~\citep{henikoff1992amino}
+and PAM~\citep{dayhoff1978} amino acid models.
+A general model accepts an arbitrary transition matrix between
+any number of alleles (which can be arbitrary unicode strings),
+and so other models such as the Kimura two and three
+parameter models~\citep{kimura1980simple,kimura1981estimation}
+can be easily and efficiently defined in user code.
+Mutation rates can vary along the genome, and multiple mutation
+models can be imposed on a tree sequence by overlaying mutations
+in multiple passes.
+The new \texttt{sim\_mutations} function
+(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
+mutation model, but full support for the
+original infinite sites model is also maintained.
+We have extensively validated the results of mutation simulations
+against both theoretical expectations and output from
+\SeqGen~\citep{rambaut1997seq} and
+Pyvolve~\citep{spielman2015pyvolve}.
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{figures/mutations-perf}
+\caption{\label{fig-mutations-perf} Time required to run
+\texttt{sim\_mutations} on tree sequences generated
+by \texttt{sim\_ancestry} (with a population size of $10^4$
+and recombination rate of $10^{-8}$) for varying (haploid) sample
+size and sequence length. We run 10 replicate mutation simulations
+each for three different mutation rates, and report the average
+CPU time required (Intel Core i7-9700).
+(A) Holding sequence length fixed at 10 megabases and varying the
+number of samples (tree tips) from 10 to 100,000.
+(B) Holding number of samples fixed at 1000, and varying the sequence
+length from 1 to 100 megabases.}
+\end{figure}
+
+Generating simulations in \msprime\ is extremely efficient.
+Fig.~\ref{fig-mutations-perf} shows the time required to generate
+mutations (using the default JC69 model) on
+simulated tree sequences for a variety of mutation
+rates as we vary the number of samples
+(Fig.~\ref{fig-mutations-perf}A) and the sequence
+length (Fig.~\ref{fig-mutations-perf}B).
+For example, the longest running simulation in
+Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
+generate an average of 1.5 million mutations over 137,081 trees
+in a tree sequence with 508,125 edges.
+This efficiency for large numbers of trees is possible because
+the tree sequence encoding allows us to generate mutations
+on an edge-by-edge basis (see Fig.~\ref{fig-ts-example}), rather than
+tree-by-tree and branch-by-branch as would otherwise be required.
+In the above example from Fig.~\ref{fig-mutations-perf}B,
+we would have to consider 273,887,838 branches
+since there are 137,081 trees and 1,998 branches in each
+tree ($\sim500$ times more work).
+
+% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
+Even if we consider a single tree, \msprime\ generates mutations
+extremely efficiently.
+For example, we simulated mutations under the BLOSUM62 amino
+acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
+in $\sim$260,000 mutations) in about $0.8$ seconds, including
+the time required for file input and output.
+We do not attempt a systematic benchmarking of \msprime's
+mutation generation code against other methods because at this scale it is
+difficult to disentangle the effects of inefficient input and
+output formats from the mutation generation algorithms.
+It seems quite unlikely, however,
+that generating mutations with \msprime\ would be a bottleneck in any
+realistic analysis.
+
+There are many ways in which the mutation generation code
+in \msprime\ could be extended by adapting existing methods.
+For example, we hope to add support for
+microsatellites~\citep{mailund2005coasim},
+codon models~\citep{arenas2007recodon}
+and indels~\citep{cartwright2005dna,fletcher2009indelible}.
+
+\subsection*{Recombination}
+
+Recombination is implemented in \msprime\ using Hudson's algorithm, which
+works backwards in time tracking the
+effects of common ancestor and recombination
+events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
+Common ancestor events combine the ancestral material of two lineages, and may
+result in coalescences in the marginal trees. Recombination events
+split the ancestral material for some lineage, creating two independent
+lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
+this process is much more efficient to simulate than either
+methods based on the Ancestral Recombination Graph
+formulation~\citep{griffiths1991two,griffiths1997ancestral}
+or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
+In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
+us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
+and empirical recombination maps similar to other
+simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
+The implementation of recombination in \msprime\ is extensively validated
+against analytical results~\citep{hudson1983properties,kaplan1985use}
+and results from \ms.
+
+The Sequentially Markovian Coalescent (SMC) is an approximation of the
+coalescent with recombination
+model~\citep{mcvean2005approximating,marjoram2006fast},
+and was originally motivated to a significant degree by the need to
+simulate longer genomes than was possible using tools like \ms.
+The SMC is a good approximation to the
+coalescent with recombination when we have four or fewer sampled
+genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
+effects of the approximation are less well understood for larger
+sample sizes and several approaches have been proposed to
+allow simulations that more closely approximate the coalescent
+with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
+The SMC is an important analytical approximation
+and has allowed the development of numerous inference
+methods~\citep{
+li2011inference,
+harris2013inferring,
+sheehan2013estimating,
+schiffels2014inferring,
+carmi2014renewal,
+rasmussen2014genome,
+zheng2014bayesian,
+terhorst2017robust}.
+
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/ancestry-perf}
+\end{center}
+\caption{\label{fig-ancestry-perf}
+    Approximate run time for \msprime\ simulations,
+    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
+    Each point is the runtime of one simulation,
+    for various values of effective population size ($N$),
+    genome length ($L$), and number of samples.
+    Runtime scales quadratically with the product of $N$ and $L$;
+    since recombination rate in these simulations was $10^{-8}$,
+    the scaled recombination rate here is $\rho = NL/10^8$,
+    shown on the horizontal axis.
+    The black line is a quadratic fit separately in each panel.
+}
+\end{figure}
+
+In human-like parameter regimes, \msprime's implementation of the
+exact coalescent with recombination comprehensively outperforms
+all other simulators, including those based on SMC
+approximations~\citep{kelleher2016efficient}. However, it is important
+to note that although the implementation of Hudson's algorithm is
+very efficient, it is still quadratic in the scaled recombination rate
+$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
+This is because Hudson's algorithms tracks recombinations not only
+in segments ancestral to the sample, but also between ancestral segments.
+Estimates of the number of such ``trapped'' recombinations grow quadratically
+with genome length \citep[Eq.~5.10]{hein2004gene}.
+Because they do not track such recombinations, efficient SMC implementations
+will be faster than \msprime\ for sufficiently high recombination rate,
+genome length, or $N_e$. %Until we decide to approximate the recombination process
+%between distant ancestral segment. Hint hint.
+
+% \jkcomment{We want to round out this discussion here by giving some
+% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
+% to Fig.~\ref{fig-ancestry-perf} to show roughly what we can expect
+% to simulate in how long. See issue 28.}
+
+Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
+a given simulation of a constant size population,
+and verifies that runtime scales quadratically with $\rho$,
+(\citet{hein2004gene} gives an equation
+including the effects of sample size --
+see Appendix~\ref{app-theoretical-runtime}.)
+and depends much less strongly on the number of samples.
+Taking a typical chromosome to be 1 Morgan in length,
+these plot show, roughly, that simulating
+samples from a population of thousands of individuals takes seconds,
+while samples from a population of tens of thousands takes minutes.
+
+However, what about a simulation with \emph{changing} size?
+To understand how runtime depends on demography
+it helps to consider \emph{why} runtime is quadratic in $\rho$.
+At any point in time, \msprime\ must keep track of some number of lineages,
+each of which contains some number of chunks of genetic material.
+Common ancestor events reduce the number of lineages,
+and recombination events increase their number.
+However, with long genomes
+only a small fraction of the common ancestor events
+will involve overlapping segments of ancestry,
+and lead to coalescence in the marginal trees.
+Such disjoint segments are often far apart (on average, about distance $L/2$),
+and so recombine apart again immediately;
+it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
+\plrcomment{(this is discussed also in the ARG section; check for overlap)}
+The maximum number of lineages occurs when
+the increase and decrease in numbers of lineages due to
+common ancestor and recombination events balance out;
+to get an idea of runtime we can estimate when this occurs.
+Suppose that the maximal number of lineages is $M$;
+at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
+and the total rate of recombination is $M \ell$,
+where $\ell$ is the mean length of genome carried by each lineage
+(including ``trapped'' non-ancestral material).
+It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
+and at this maximum, recombination and coalescence rates are equal,
+so many (roughly half) of these lineages have at least two distant segments of ancestral material
+separated by a long segment of genome, implying that $\ell \approx L / 4$.
+This implies that $M$ is roughly $L N_e$.
+Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
+Since the total rate of events when there are the maximum number of lineages
+happens at rate roughly $L^2 N_e / 4$,
+then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
+as desired.
+What does this tell us about time-varying population sizes?
+The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
+in other words,
+more events happen in the time period from 0 to $N_e/2$ generations
+than from $N_e/2$ to $N_e$ generations,
+but the runtime is not dominated but the more recent time period.
+
+Putting this together:
+suppose that population size today is $N_1$,
+while longer ago than $T$ generations ago it was $N_2$:
+does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
+The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
+then runtime will be similar to a population of size $N_1$;
+while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
+For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
+and the runtime will depend critically on $T$ -- in other words,
+simulation will be quick in a species with a strong domestication bottleneck,
+and slow otherwise.
+In practice, we recommend that users estimate runtime by
+measuring runtime in a series of simulations with short genome lengths,
+possibly varying other parameters to check for interactions,
+and then predict runtime by fitting a quadratic curve to genome length.
+There is only a weak dependency on sample size
+-- notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
+only increases runtime by a factor of 2 or so.
+
+The SMC~\citep{mcvean2005approximating} and
+SMC'~\citep{marjoram2006fast} models are available
+in \msprime\ 1.0. However, they are currently implemented using a
+naive rejection sampling approach, and are somewhat slower
+to simulate than the exact coalescent with recombination. These
+models are therefore only appropriate for studying the SMC approximations
+themselves; an important avenue for future work is to implement
+efficient SMC-like models, perhaps based on the approaches used
+in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
+simulation of organisms such as \textit{Drosophila melanogaster}.
+
+\subsection*{Gene conversion}
+
+%introduction
+Gene conversion is a unidirectional transfer of a short segment of genetic material,
+for example between homologous chromosomes~\citep{chen2007gene}.
+%description of ancestral process
+Just as recombination, gene conversion events split the ancestral material
+for some lineage, creating two independent lineages.
+In contrast to the previously described recombination process, where the ancestral
+material to the left and right of a recombination breakpoint follows two
+different ancestral lineages, gene conversion events distribute the genetic material
+of a short segment to one ancestral lineage and the part to the left and right of this
+segment to another lineage.
+%Franz Baumdicker: Not sure if this next sentence is necessary,
+% but it might be of interest for some people?
+The resulting distribution of ancestral material after a gene conversion event
+equals the distribution after two recombination events at the left and right position
+of the gene conversion tract, followed by a common ancestor event.
+%Impact on LD
+However, gene conversions impact much shorter segments than crossover recombinations,
+typically below 1kb, and thus affect linkage patterns differently than
+recombinations~\citep{korunes2017gene}.
+
+% other simulators
+Hudson's \ms\ optionally includes gene conversion and so does
+MaCS~\citep{chen2009fast}, while scrm~\citep{staab2015scrm} does not
+support gene conversions.
+As bacterial recombination can be modeled analogously to eukaryotic allelic
+gene conversion, simulators for microbial evolution often include gene conversion
+instead of recombination.
+This includes the backward simulator SimBac~\citep{brown2016simbac},
+the SMC based approach used in FastSimBac~\citep{demaio2017the},
+and forward in time tools within the SLiM framework~\citep{cury2020simulation}.
+
+In \msprime\ 1.0 we implemented the coalescent with gene conversion as defined by
+\cite{wiuf2000coalescent}, where gene conversion events are initiated between any
+two sites in the sequence at a constant (scaled) rate $\gamma$ and
+the length of the transferred segment is given by a geometric distributed
+random variable with mean length $L$.
+% Above I used the discrete case, but we could easily switch to a description for the
+% continuous case.
+We have implemented this model of gene
+conversion in \msprime\ 1.0, and validated the output against
+\ms\ and analytical results~\citep{wiuf2000coalescent}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=7cm]{figures/gc-perf}
+\end{center}
+\caption{\label{fig-gc-perf}Comparison of the running time
+for E-coli simulations using \msprime\ and SimBac for varying
+sample sizes. The parameters used are chosen to mimic the simulation
+of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
+gene conversion rate was set to $N_e g = \gamma = 0.015$ and
+the mean gene conversion tract length was set to $L = 500$, which is close to
+estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
+$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
+with mean tract length $L = 542$~\citep{lapierre2016the}.}
+\end{figure}
+
+Gene conversion is particularly useful to model homologous recombination in
+bacterial evolution,
+so we compare the performance of \msprime\ with gene conversion to
+SimBac~\citep{brown2016simbac}, a simulator developed specifically to
+simulate bacterial evolution. Although \msprime\ is missing some features
+relevant to this case, e.g.\ the simulation of circular genomes, it is far
+more efficient at simulating the basic processes.
+Furthermore, gene conversion can be considered in combination
+with other current and future features of \msprime\ 1.0,
+including varying recombination rates and complex demographies.
+
+An important avenue for future work
+will be to add further features to model bacterial
+evolution. In particular, we aim to include bacterial gene transfer and utilize
+the succinct tree sequence structure to represent the resulting ancestral gene
+transfer graph~\citep{baumdicker2014AGTG}.
+
+\subsection*{Demography}
+% TODO should we put in some citations here for the models? Hardly seems
+% necessary.
+One of the key applications of population genetic simulations is to generate
+data for complex demographies. Beyond idealised cases such as stepping-stone or
+island models, or specialised cases such as isolation-with-migration models,
+analytical results are rarely possible. Simulation is therefore integral to the
+development and evaluation of methods for demographic inference. The demography
+model in \msprime\ is directly derived from the approach used in \ms, and
+supports an arbitrary number of randomly mating populations exchanging migrants
+at specified rates. A range of demographic events are supported, which allow for
+varying population sizes and growth rates, changing migration rates over time,
+as well as population splits, admixtures and pulse migrations.
+The location of sampled lineages
+can be tracked through time in as much detail as required: each tree node is
+automatically associated with the population in which it arose, the location of
+lineages can be recorded at any given time via census events, or every lineage
+migration can be recorded.
+
+A major change for \msprime\ 1.0 is the introduction of the new Demography API,
+designed to address a serious flaw in the \msprime\ 0.x interface which lead to
+a number of avoidable errors in downstream
+simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
+separate parameters be provided to the \texttt{simulate} function to describe a
+demographic model, making it easy to accidentally omit information. The 1.0 API
+resolves this issue by creating a new \texttt{Demography} class, which
+encapsulates all information about the demographic model, and fully decouples
+the definition from other simulation details. An instance of this class is then
+provided as a parameter to the new \texttt{sim\_ancestry} function,
+substantially reducing the potential for error.
+Another improvement over the
+0.X APIs is the introduction of explicit population split and admixture events,
+and a population state machine that ensures that lineages cannot migrate into
+(or be sampled from) inactive populations. This demography model is compatible
+with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
+class supports importing and exporting Demes models.
+Models previously constructed using the 0.x API can be seamlessly imported into
+the \texttt{Demography} class, and we also support importing demographic
+models in the form of Newick species tree descriptions, as well as those
+produced by programs such as *BEAST~\citep{heled2009bayesian}.
+
+The \texttt{DemographyDebugger} can output detailed information for debugging
+demographic models, but also provides
+methods to compute a suite of numerical predictions about the location of
+lineages over time, for example the coalescence rates for two or more lineages
+drawn from populations at specified times in the past.
+Many popular approaches in population genetics use the distribution of
+coalescence rates between pairs of lineages in one or more populations to infer
+effective population sizes over
+time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
+split times and subsequent migration rates between
+populations~\citep{wang2020tracking}.
+The \texttt{DemographyDebugger} provides a valuable
+ground-truth when evaluating such inference methods, as illustrated by
+\cite{adrion2020community}.
+
+\subsection*{Instantaneous bottlenecks}
+
+A common approach to modelling the effect of demographic history on genealogies
+is to assume that effective population size ($N_e$) changes in discrete steps
+which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
+keightley2007joint,li2011inference}. In this setting of piecewise constant
+$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
+reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
+recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
+bottlenecks are short both on the timescale of coalescence and mutations,
+there may be little information about the duration of a bottleneck
+($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
+assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
+and generate a sudden burst of coalescence events (a multiple merger event) in
+the genealogy. The strength of the bottleneck $B$ can be thought of as an
+(imaginary) time period during which coalescence events are collapsed,
+i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
+single pair of lineages entering the bottleneck coalesce during the bottleneck
+is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
+attractive and both analytic results and empirical
+inference~\citep{griffiths1994sampling, galtier2000detecting,
+bunnefeld2015inferring} have been developed under this model, there has
+been no software available to simulate data under instantaneous
+bottleneck histories.
+
+We have implemented instantaneous bottlenecks in \msprime~1.0
+using a variant of Hudson's linear time single-locus coalescent
+algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
+by adding events to the Demography object (see the Demography section)
+and can be used in combination with any other demographic modelling
+features. We have validated the results of these simulations by comparing
+against analytical expectations for coalescence times and the
+site frequency spectrum~\citep{bunnefeld2015inferring}.
+
+\subsection*{Multiple merger coalescents}
+
+Kingman's coalescent assumes that only two ancestral lineages can merge at
+each merger event. Although this is generally a reasonable approximation there
+are certain situations in which the underlying mathematical assumptions are
+violated. For example in certain highly fecund organisms
+\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
+to produce numbers of offspring on the order of the population size and
+therefore a few individuals may produce the bulk of the offspring in any given
+generation~\citep{hedgecock_94}. These population dynamics violate basic
+assumptions of the Kingman coalescent, and are better modelled by
+`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
+two lineages can merge in a given event. Multiple-merger coalescent processes
+have also been shown to be relevant for modeling the effects of selection on
+gene genealogies \citep{Gillespie909,DS04}.
+
+Although multiple merger coalescents have been of significant theoretical
+interest for around two decades, there has been little practical software
+available to simulate these models.
+\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
+simulate a related spatial continuum model~\citep{barton2010new},
+\cite{zhu2015hybrid} simulate genealogies within a species tree
+based on a multiple-merger model, and
+\cite{becheler2020occupancy} provide a general method for simulating
+multiple merger processes
+as part of the Quetzal framework~\citep{becheler2019quetzal}.
+The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
+includes a number of extensions to the Beta-coalescent.
+None of these methods work with large genomes, and very little work
+has been performed on simulating multiple merger processes with recombination.
+
+We have added two multiple merger coalescent models in \msprime\ 1.0, the
+Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
+such models with recombination efficiently for the first time.
+These simulation models have been extensively validated against
+analytical results from the site frequency
+spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
+ as well as more general properties of coalescent processes.
+Please see the Appendix for more details and model derivations.
+
+\subsection*{Ancestral Recombination Graphs}
+\label{sec-arg}
+
+The Ancestral Recombination Graph (ARG) was introduced by
+Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
+representation of the coalescent with recombination stochastic process. This
+formulation was complementary to Hudson's earlier
+work~\citep{hudson1983properties}, and substantially increased our theoretical
+understanding of recombination. In Griffiths' ARG formulation, we
+regard a realisation of the coalescent with recombination as a graph in which
+vertices represent common ancestor or recombination events, and edges represent
+lineages through which ancestral material flows.
+We may consider the big ARG, in which we track lineages arising out of
+recombinations regardless of whether they carry ancestral material and
+consequently wait an exponentially large time until all sampled lineages
+coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
+which we only track genetic ancestors, and wait until all marginal trees
+have fully coalesced (like Hudson's algorithm).
+
+\begin{figure}
+
+\begin{center}
+    \begin{tikzpicture}[scale=0.45]
+
+    \draw (-2,0) -- (-2,3) -- (0, 3);
+    \draw (1,0) -- (1,1);
+    \draw (0,1) -- (2,1) -- (2,2) -- (4,2) -- (4,0);
+    \draw (3,2) -- (3,5);
+    \draw (0,1) -- (0,3);
+    \draw (-1, 3) -- (-1, 5) -- (3,5);
+
+    \draw[fill] (-2,0) circle [radius=0.05];
+    \draw[fill] (-1,3) circle [radius=0.05];
+    \draw[fill] (1,0) circle [radius=0.05];
+    \draw[fill] (4,0) circle [radius=0.05];
+    \draw[fill] (1,1) circle [radius=0.05];
+    \draw[fill] (3,2) circle [radius=0.05];
+    \draw[fill] (1,5) circle [radius=0.05];
+
+    \node [below] at (-2,0) {0};
+    \node [below] at (1,0) {1};
+    \node [below] at (4,0) {2};
+    \node [above] at (1,1) {3};
+    \node [below right] at (1,1) {$x = 0.3$};
+    \node [above right] at (3,2) {4};
+    \node [above left] at (-1,3) {5};
+    \node [above] at (1,5) {6};
+
+    \end{tikzpicture}
+% \begin{center}
+% $\equiv$
+% \end{center}
+% \end{column}
+% \begin{column}{5cm}
+\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
+\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
+\includegraphics[width=3.8cm]{figures/arg.pdf}
+
+\end{center}
+\caption{\label{fig-arg} (A) An simple ARG in which a recombination
+occurs at position 0.3; (B) the equivalent topology depicted as a tree
+sequence, including the recombination node; (C) the same tree sequence
+topology ``simplified'' down to its minimal tree sequence representation;
+(D) the fraction of nodes in a tree sequence that are concerned with
+extra ARG information in a tree sequences returned using \msprime's
+\texttt{record\_full\_arg} mode}
+\end{figure}
+
+However, this original definition of the ARG in terms of a stochastic process
+is rarely what is meant when the term is currently used. The term has come to
+essentially mean any representation of genetic ancestry in the presence
+of recombination as a graph-like structure; for example,
+\cite{mathieson2020ancestry} informally define an ARG as the subset of an
+individual's pedigree from which they have inherited ancestral material.
+Others state that an ARG is equivalent to the sequence of marginal
+% Also SARGE paper
+trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}
+
+Such
+imprecision is unfortunate, since there are any number of graphical encodings of
+genetic ancestry, and referring to the original Griffiths papers does little to
+clarify. As we see below, there are situations in which the precise information
+stored in the graph is important, and conflating a high-level
+discussion of genetic ancestry with this mathematical object is not
+helpful. The point of technical terms is to provide clarity and
+precision on what is being discussed, and it is now essentially
+impossible to know what is meant by the term ``ARG'';
+we therefore use the term ``genetic ancestry'' in high-level
+discussions.
+
+For our purposes, an ARG is a realisation of the coalescent with
+recombination, in the Griffiths (little ARG) sense.
+As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
+works by dynamically generating and traversing a little ARG.
+The graph is not explicitly represented in memory, but is partially
+present through the extant lineages and the ancestral material they carry
+over time. Since the ARG is a large and unwieldy object
+(Fig.~\ref{fig-arg}) we do not use it as output, but
+rather store the information required to recover the genealogical
+history as nodes and edges in a tree sequence (see section XX),
+or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
+
+Although the tree sequences output by \msprime\ contain full information
+about the genealogical trees, their correlation structure along the genome,
+and the ancestral haplotypes on which coalescences occured, some information
+is lost in this mapping down from ARG space to the minimal tree sequence
+form. In particular, we lose
+information about ancestral haplotypes that were common ancestors but
+in which no coalescences occured, and also information about the precise time
+and genomic location of recombination events. In most cases, such
+information is of little relevance as it is in principle unknowable,
+but there are occasions such as visualisation or computing likelihoods (see
+below) in which is useful. We therefore provide
+the \texttt{record\_full\_arg} option in \msprime\
+to store a representation of the complete ARG traversed during simulation.
+This is done by storing extra nodes (marked with specific flags, so they
+can be easily identified later) and edges in the tree sequence.
+Fig.~\ref{fig-arg} shows that this extra ARG information
+quickly comes to dominate the output as we increase the sequence length,
+and the resulting tree sequence files are many times larger
+than the standard minimal tree sequence representation.
+
+One situation in which a record of the full ARG is necessary is when we
+wish to compute likelihoods during inference.
+The likelihood is a central quantity in evaluating the plausibility of a putative
+ancestry as an explanation of DNA sequence data, both directly through
+e.g.~approaches based on maximum likelihood, and as an ingredient of
+methods such as
+Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
+We provide functions to compute the likelihood of ARG realisations
+(encoded using the representation of the \texttt{record\_full\_arg} option)
+and mutational patterns under
+the standard coalescent and infinite sites mutation model.
+See the Appendix for details.
+
+\subsection*{Selection}
+\label{sec-selection}
+Another ellaboration of the standard neutral coalescent with recombination
+is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
+In the context of the coalescent this amounts to creating a structured population
+during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
+in which lineages may transit between favored and unfavored backgrounds through
+recombination. This allows for many selective sweep scenarios, including recurrent, partial,
+and soft selective sweeps to be simulated efficiently, however in general this
+efficiency comes at the cost of flexibility in comparison to forwards in
+time simulation.
+
+Currently there are a small number of other simulators that will generate
+coalescent samples with sweeps including, SelSim~\citep{spencer2004selsim}
+mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
+Cosi2~\citep{shlyakhter2014cosi2}, and discoal~\citep{kern2016discoal}.
+Each of these packages has their own benefits and possible drawbacks,
+however none of them at present will output tree sequence formatted data.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=6cm]{figures/sweeps-perf}
+\end{center}
+\caption{\label{fig-selection-perf}Comparison of the runtime for
+\msprime\ and \texttt{discoal} for simulating selective sweeps.
+}
+\end{figure}
+
+We have implemented selective sweeps in \msprime\ similarly to
+the \texttt{discoal} implementation-- sweep trajectories are
+generated from a jump process approximation to the conditional diffusion
+of an allele bound for fixation~\citep{coop2004ancestral}.
+This implementation allows us to further generalize sweep models to
+changing population sizes as well as selection from standing variation,
+both of which are features that are planned for the near future.
+Moreover because this implements the basic infrastructure for the structured
+coalescent, adding support for, e.g.,
+inversions is straightforward~\citep{peischl2013sequential}.
+
+\subsection*{Discrete time Wright-Fisher}
+
+The assumptions of the classical coalescent model are that both sample size and
+the simulated sequence length are small. As modern datasets often include
+whole-genome sequences for tens or hundreds of thousands of samples, simulating
+such data requires breaking these assumptions. While the coalescent remains a
+good approximation to the Wright-Fisher model in the distant past, it can
+exhibit large biases in the recent history. Identity-by-descent (IBD),
+long-range linkage disequilibrium (LD), and ancestry patterns deviate from
+Wright-Fisher expectations, and this bias grows considerably when simulating
+larger sample
+sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
+
+A backwards-in-time Wright-Fisher model is supported in \msprime, which is
+shown to properly capture the expected distributions of IBD and long-range LD,
+as well as ancestry patterns in admixed
+populations~\citep{nelson2020accounting}. In short, the classical coalescent
+draws recombination events from exponential waiting times, which are continuous
+in time, and each recombination then splits a segment into two distinct
+ancestors. For large sequence lengths, this can dramatically overestimate the
+number of genealogical ancestors of a given sample. Instead, the Wright-Fisher
+model iterates through discrete generations so that each gamete has a unique
+diploid parent, and multiple recombinations within a generation results in
+crossover events between the same two parental haploid copies.  This method is
+described in more detail in~\cite{nelson2020accounting}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=6cm]{figures/dtwf-perf}
+\end{center}
+\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
+simulations with ARGON. To ensure the comparison with ARGON is
+as fair as possible we run it with a mutation rate of 0 and with
+the minimum output options (so that it writes out only an empty VCF file)}
+\end{figure}
+
+We support changing the simulation model at specified times so that we can
+easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}. We
+can combine any number of different simulation models, allowing for the
+flexible choice of simulation scenarios without severely damaging efficiency.
+As the discrete time Wright-Fisher model improves accuracy of genealogical
+patterns in the recent past, we can simulate the recent history using this
+model and then switch to the standard coalescent to more efficiently simulate
+the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the perfomance of
+this approach with ARGON~\citep{palamara2016argon}, a specialised DTWF
+simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
+simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
+hybrid approach provides large efficiency gains.
+
+\subsection*{Simulation interface}
+\label{sec-sim-interface}
+
+The majority of simulation packages are controlled either through
+a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
+a text-based input file
+format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
+or a mixture of both.
+Command line interfaces such as \ms\ make it very easy to run simple
+simulations, but as model complexity and the number of parameters specified increases
+they become difficult to understand and
+error-prone~\citep{ragsdale2020lessons,gower2021demes}.
+Specifying parameters through a text file alleviates this problem to a degree,
+but lacks flexibility, for example, when running simulations with parameters
+drawn from some distribution. In practice, for any significant simulation
+project (or at least, any reproducible one), users will write a script
+to generate the required command lines or input parameter files,
+invoke the simulation engine, and process the results in some way.
+This process is cumbersome and labour intensive, and
+a number of packages have been developed
+to allow simulations be run directly in a high-level
+scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
+
+The more recent trend has been to move away from this file and command-line
+driven approach and to instead provide direct interfaces to the simulation
+engines via an Application Programming Interface (API)~\citep[e.g.][]{
+thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
+The primary interface for \msprime\ is through a thoroughly documented and
+stable Python
+API, which has allowed a number of downstream tools to be built on top of
+% Found these through the Dependent packages on GitHub
+it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
+adrion2020community,adrion2020predicting, kamm2020efficiently,
+mckenzie2020ipcoal, montinaro2020revisiting,
+de2021geonomics,rivera2021simulation}.
+As well as providing a stable and efficient platform for building
+downstream applications, \msprime's Python API makes it much easier to
+build reproducible simulation pipelines, as the entire workflow can
+be encapsulated in a single script, and package and version
+dependencies can be explicitly stated using the \texttt{pip}
+or \texttt{conda} package managers.
+For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
+in the influential simulation analysis of
+\cite{martin2017human} were only detected because the pipeline
+could be easily run and reanalysed.
+
+A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
+designed in part to avoid sources of error (see the Demography section) but
+also to provide more appropriate defaults while keeping compatibility with
+existing code. In the new APIs, ancestry and mutation simulation are fully
+separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
+functions replacing the legacy \texttt{simulate} function. Among other changes,
+the new APIs default to discrete genome coordinates and finite sites mutations,
+resolving a major source of confusion and error. The 0.x APIs are fully
+supported and tested, and will be maintained for the foreseeable future.
+We also provide an \ms\ compatible
+command line interface to support existing workflows, and the
+\texttt{msp} program that supports specifying demographic models in
+Demes format~\citep{gower2021demes} from the command line.
+
+\subsection*{Analysis of simulation results}
+\label{sec-data-analysis}
+
+\jkcomment{This is probably part of the discussion, but putting here out of
+the way for now in its own subsection}
 
@@
 
-\subsection*{Genetic variation}
-\label{sec-genetic-variation}
-
-The output of population genetics simulators, quite naturally,
-tends to be some representation of the resulting nucleotide sequences
-(or some derived statistics of those sequences),
-although the underlying genealogies are also of interest.
-% Mutations are not treated as first-class data --- the sequences
-% were really all people were after, classically.
-However, joint information about the mutations that have been simulated
--- e.g., where they have occurred on the trees --
-is usually not retained or made available for subsequent analysis.
-
-The approach taken in \msprime\ is quite different.
-As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
-integral part of \tskit's tree sequence data model, and genetic
-variation is encoded by recording sites at which mutations
-have occurred, and where each mutation at those sites
-has occurred on the marginal tree;
-the sequences themselves are never stored.
-It may at first seem inconvenient to have only this indirect representation
-of the genome sequences, but it is extremely powerful.
-Firstly, the storage space required for simulations is dramatically
-reduced. For a simulation of $n$ samples with $m$ variant sites,
-we would require $O(nm)$ space to store the sequence data as a variant
-matrix. However, if this simulation was of a recombining genome
-with $t$ trees, then the \tskit\ tree sequence encoding
-requires $O(n + t + m)$ space~\citep{kelleher2016efficient},
-assuming we have $O(1)$ mutations at each site. For
-large sample sizes, this difference is profound, making it
-conceivable, for example, to store the genetic ancestry
-and variation data for the entire human population on a
-laptop~\citep{kelleher2019inferring}. Besides the huge
-difference in storage efficiency, it is also often far more efficient
-to compute statistics of the sequence data from the
-trees and mutations than it is to work with the sequences themselves.
-For example, computing Tajima's $D$ from simulated data using \tskit\
-is several orders of magnitude faster than efficient variant matrix
-libraries for large sample sizes~\citep{ralph2020efficiently}.
-
-The advantages of storing genetic variation data using the underlying trees
-have recently become clear in the context of the vast genomic
-datasets produced during the SARS-CoV-2 pandemic.
-\cite{turakhia2021ultrafast} propose the
-Mutation Annotated Tree (MAT)
-format (consisting of a Newick tree
-and associated mutations in a binary format) and
-the \texttt{matUtils} program as an efficient way to store
-and process large viral datasets~\citep{mcbroome2021daily},
-achieving excellent compression and processing performance.
-Similarly, \phastsim~\citep{demaio2021phastsim}
-was developed to simulate sequence evolution on such large SARS-CoV-2
-phylogenies, and also outputs a Newick tree annotated with mutations (not
-in MAT format) to avoid the bottleneck of generating
-and storing the simulated sequences.
-While these methods illustrate the advantages of the general
-approach of storing ancestry and mutations rather than sequences,
-they do not generalise beyond their immediate settings, and
-no software library support is available.
-\msprime, on the other hand, provides a fast, powerful, and well-tested
-generic interface for generating and storing mutations on trees.
-
-\subsection*{Simulating mutations}
-\label{sec-mutations}
-
-% Explain the background: what have people done classically?
-Because coalescent simulations are usually concerned with
-neutral evolution (see the \nameref{sec-selection} section, however)
-the problem of generating synthetic genetic variation can be decomposed into
-two independent steps:
-simulating genetic ancestry (the trees), and then simulating
-variation by superimposing mutation processes on those trees
-(see figure~\ref{fig-mutated-trees}).
-A number of programs exist to place mutations on trees:
-for instance,
-the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
-allows a range of different models of sequence evolution to be imposed on a given
-tree in Newick format, and there are various
-extensions to the basic nucleotide
-models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
-Population genetic simulators have tended to
-include their own implementations of mutation simulation, with
-most just supporting the infinite sites
-model~\citep[e.g.][]{hudson2002generating}
-but with several supporting a range of different models
-of sequence
-evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
-virgoulay2021gspace}.
-
-\begin{figure}
-    \includegraphics{illustrations/unmutated_tree}
-    \includegraphics{illustrations/mutated_tree}
-    \caption{
-        \label{fig-mutated-trees}
-        Visualization of the separation between ancestry and mutation simulation:
-        \texttt{sim\_ancestry} produces ancestry
-        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
-        relating 200 samples),
-        and \texttt{sim\_mutations} adds mutations
-        (\textbf{right;} fourteen nucleotide mutations have been added).
-        Graphics produced with \tskit's \texttt{draw\_svg} method.
-        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
-    }
-\end{figure}
-
-% on the other hand, tskit is awesome.
-Clean separation of the ancestry and mutation generation steps
-is made possible by the \tskit\ library, whose mature and
-efficient C and Python interfaces makes it straightforward to add
-further information to an existing file, and to output the results.
-As such, it is essentially a historical accident
-that \msprime\ (whose main focus is on ancestry simulation)
-contains code to simulate mutations: the mutation process could
-be performed by a separate package, and there would be no performance
-penalty for doing so.
-Thanks to this interoperability, \msprime's
-mutation generator can work with \emph{any} \tskit\ tree sequence,
-be it simulated using \SLiM~\citep{haller2019slim} or
-\fwdpy~\citep{thornton2014cpp}, or estimated from real data
-using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
-or \texttt{Relate}~\citep{speidel2019method}. It is a modular
-component intended to fit into a larger software ecosystem, and
-is in no way dependent on \msprime's ancestry simulator.
-
-For version 1.0 we have greatly extended the sophistication of
-\msprime's mutation generation engine, achieving near feature-parity
-with \SeqGen.
-We  support a large number of mutation models, including
-JC69~\citep{jukes1969evolution},
-F84~\citep{felsenstein1996hidden},
-and GTR~\citep{tavare1986some} nucleotide models
-and the BLOSUM62~\citep{henikoff1992amino}
-and PAM~\citep{dayhoff1978} amino acid models.
-A general model accepts an arbitrary transition matrix between
-any number of alleles (which can be arbitrary unicode strings),
-and so other models such as the Kimura two and three
-parameter models~\citep{kimura1980simple,kimura1981estimation}
-can be easily and efficiently defined in user code.
-Mutation rates can vary along the genome, and multiple mutation
-models can be imposed on a tree sequence by overlaying mutations
-in multiple passes.
-The new \texttt{sim\_mutations} function
-(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
-mutation model, but full support for the
-original infinite sites model is also maintained.
-We have extensively validated the results of mutation simulations
-against both theoretical expectations and output from
-\SeqGen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}.
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{figures/mutations-perf}
-\caption{\label{fig-mutations-perf} Time required to run
-\texttt{sim\_mutations} on tree sequences generated
-by \texttt{sim\_ancestry} (with a population size of $10^4$
-and recombination rate of $10^{-8}$) for varying (haploid) sample
-size and sequence length. We run 10 replicate mutation simulations
-each for three different mutation rates, and report the average
-CPU time required (Intel Core i7-9700).
-(A) Holding sequence length fixed at 10 megabases and varying the
-number of samples (tree tips) from 10 to 100,000.
-(B) Holding number of samples fixed at 1000, and varying the sequence
-length from 1 to 100 megabases.}
-\end{figure}
-
-Generating simulations in \msprime\ is extremely efficient.
-Fig.~\ref{fig-mutations-perf} shows the time required to generate
-mutations (using the default JC69 model) on
-simulated tree sequences for a variety of mutation
-rates as we vary the number of samples
-(Fig.~\ref{fig-mutations-perf}A) and the sequence
-length (Fig.~\ref{fig-mutations-perf}B).
-For example, the longest running simulation in
-Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
-generate an average of 1.5 million mutations over 137,081 trees
-in a tree sequence with 508,125 edges.
-This efficiency for large numbers of trees is possible because
-the tree sequence encoding allows us to generate mutations
-on an edge-by-edge basis (see Fig.~\ref{fig-ts-example}), rather than
-tree-by-tree and branch-by-branch as would otherwise be required.
-In the above example from Fig.~\ref{fig-mutations-perf}B,
-we would have to consider 273,887,838 branches
-since there are 137,081 trees and 1,998 branches in each
-tree ($\sim500$ times more work).
-
-% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
-Even if we consider a single tree, \msprime\ generates mutations
-extremely efficiently.
-For example, we simulated mutations under the BLOSUM62 amino
-acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
-in $\sim$260,000 mutations) in about $0.8$ seconds, including
-the time required for file input and output.
-We do not attempt a systematic benchmarking of \msprime's
-mutation generation code against other methods because at this scale it is
-difficult to disentangle the effects of inefficient input and
-output formats from the mutation generation algorithms.
-It seems quite unlikely, however,
-that generating mutations with \msprime\ would be a bottleneck in any
-realistic analysis.
-
-There are many ways in which the mutation generation code
-in \msprime\ could be extended by adapting existing methods.
-For example, we hope to add support for
-microsatellites~\citep{mailund2005coasim},
-codon models~\citep{arenas2007recodon}
-and indels~\citep{cartwright2005dna,fletcher2009indelible}.
-
-\subsection*{Recombination}
-
-Recombination is implemented in \msprime\ using Hudson's algorithm, which
-works backwards in time tracking the
-effects of common ancestor and recombination
-events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
-Common ancestor events combine the ancestral material of two lineages, and may
-result in coalescences in the marginal trees. Recombination events
-split the ancestral material for some lineage, creating two independent
-lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
-this process is much more efficient to simulate than either
-methods based on the Ancestral Recombination Graph
-formulation~\citep{griffiths1991two,griffiths1997ancestral}
-or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
-In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
-us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
-and empirical recombination maps similar to other
-simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
-The implementation of recombination in \msprime\ is extensively validated
-against analytical results~\citep{hudson1983properties,kaplan1985use}
-and results from \ms.
-
-The Sequentially Markovian Coalescent (SMC) is an approximation of the
-coalescent with recombination
-model~\citep{mcvean2005approximating,marjoram2006fast},
-and was originally motivated to a significant degree by the need to
-simulate longer genomes than was possible using tools like \ms.
-The SMC is a good approximation to the
-coalescent with recombination when we have four or fewer sampled
-genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
-effects of the approximation are less well understood for larger
-sample sizes and several approaches have been proposed to
-allow simulations that more closely approximate the coalescent
-with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
-The SMC is an important analytical approximation
-and has allowed the development of numerous inference
-methods~\citep{
-li2011inference,
-harris2013inferring,
-sheehan2013estimating,
-schiffels2014inferring,
-carmi2014renewal,
-rasmussen2014genome,
-zheng2014bayesian,
-terhorst2017robust}.
-
-\begin{figure}
-\begin{center}
-    \includegraphics{figures/ancestry-perf}
-\end{center}
-\caption{\label{fig-ancestry-perf}
-    Approximate run time for \msprime\ simulations,
-    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
-    Each point is the runtime of one simulation,
-    for various values of effective population size ($N$),
-    genome length ($L$), and number of samples.
-    Runtime scales quadratically with the product of $N$ and $L$;
-    since recombination rate in these simulations was $10^{-8}$,
-    the scaled recombination rate here is $\rho = NL/10^8$,
-    shown on the horizontal axis.
-    The black line is a quadratic fit separately in each panel.
-    \plrcomment{TODO: run out more simulations to get more triangles}
-}
-\end{figure}
-
-In human-like parameter regimes, \msprime's implementation of the
-exact coalescent with recombination comprehensively outperforms
-all other simulators, including those based on SMC
-approximations~\citep{kelleher2016efficient}. However, it is important
-to note that although the implementation of Hudson's algorithm is
-very efficient, it is still quadratic in the scaled recombination rate
-$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
-This is because Hudson's algorithms tracks recombinations not only
-in segments ancestral to the sample, but also between ancestral segments.
-Estimates of the number of such ``trapped'' recombinations grow quadratically
-with genome length \citep[Eq.~5.10]{hein2004gene}.
-Because they do not track such recombinations, efficient SMC implementations
-will be faster than \msprime\ for sufficiently high recombination rate,
-genome length, or $N_e$. %Until we decide to approximate the recombination process
-%between distant ancestral segment. Hint hint.
-
-% \jkcomment{We want to round out this discussion here by giving some
-% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
-% to Fig.~\ref{fig-ancestry-perf} to show roughly what we can expect
-% to simulate in how long. See issue 28.}
-
-Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
-a given simulation of a constant size population,
-and verifies that runtime scales quadratically with $\rho$,
-and depends much less strongly on the number of samples.
-Taking a typical chromosome to be 1 Morgan in length,
-these plot show, roughly, that simulating
-samples from a population of thousands of individuals takes seconds,
-while samples from a population of tens of thousands takes minutes.
-
-However, what about a simulation with \emph{changing} size?
-To understand how runtime depends on demography
-it helps to consider \emph{why} runtime is quadratic in $\rho$.
-At any point in time, \msprime\ must keep track of some number of lineages,
-each of which contains some number of chunks of genetic material.
-Common ancestor events reduce the number of lineages,
-and recombination events increase their number.
-However, with long genomes
-only a small fraction of the common ancestor events
-will involve overlapping segments of ancesty,
-and lead to coalescence in the marginal trees.
-Such disjoint segments are often far apart (on average, about distance $L/2$),
-and so recombine apart again immediately;
-it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
-\plrcomment{(this is discussed also in the ARG section; check for overlap)}
-The maximum number of lineages occurs when
-the increase and decrease in numbers of lineages due to
-common ancestor and recombination events balance out;
-to get an idea of runtime we can estimate when this occurs.
-Suppose that the maximal number of lineages is $M$;
-at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
-and the total rate of recombination is $M \ell$,
-where $\ell$ is the mean length of genome carried by each lineage
-(including ``trapped'' non-ancestral material).
-It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
-and at this maximum, recombination and coalescence rates are equal,
-so many (roughly half) of these lineages have at least two distant segments of ancestral material
-separated by a long segment of genome, implying that $\ell \approx L / 4$.
-This implies that $M$ is roughly $L N_e$.
-Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
-Since the total rate of events when there are the maximum number of lineages
-happens at rate roughly $L^2 N_e / 4$,
-then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
-as desired.
-What does this tell us about time-varying population sizes?
-The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
-in other words,
-more events happen in the time period from 0 to $N_e/2$ generations
-than from $N_e/2$ to $N_e$ generations,
-but the runtime is not dominated but the more recent time period.
-
-Putting this together:
-suppose that population size today is $N_1$,
-while longer ago than $T$ generations ago it was $N_2$:
-does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
-The answer depends on how $T$ commpares to $N_1$: if $T/N_1$ is large,
-then runtime will be similar to a population of size $N_1$;
-while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
-For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
-and the runtime will depend critically on $T$ -- in other words,
-simulation will be quick in a species with a strong domestication bottleneck,
-and slow otherwise.
-In practice, we recommend that users estimate runtime by
-measuring runtime in a series of simulations with short genome lengths,
-possibly varying other parameters to check for interactions,
-and then predict runtime by fitting a quadratic curve to genome length.
-
-The SMC~\citep{mcvean2005approximating} and
-SMC'~\citep{marjoram2006fast} models are available
-in \msprime\ 1.0. However, they are currently implemented using a
-naive rejection sampling approach, and are somewhat slower
-to simulate than the exact coalescent with recombination. These
-models are therefore only appropriate for studying the SMC approximations
-themselves; an important avenue for future work is to implement
-efficient SMC-like models, perhaps based on the approaches used
-in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
-simulation of organisms such as \textit{Drosophila melanogaster}.
-
-\subsection*{Gene conversion}
-
-%introduction
-Gene conversion is a unidirectional transfer of a short segment of genetic material,
-for example between homologous chromosomes~\citep{chen2007gene}.
-%description of ancestral process
-Just as recombination, gene conversion events split the ancestral material
-for some lineage, creating two independent lineages.
-In contrast to the previously described recombination process, where the ancestral
-material to the left and right of a recombination breakpoint follows two
-different ancestral lineages, gene conversion events distribute the genetic material
-of a short segment to one ancestral lineage and the part to the left and right of this
-segment to another lineage.
-%Franz Baumdicker: Not sure if this next sentence is necessary,
-% but it might be of interest for some people?
-The resulting distribution of ancestral material after a gene conversion event
-equals the distribution after two recombination events at the left and right position
-of the gene conversion tract, followed by a common ancestor event.
-%Impact on LD
-However, gene conversions impact much shorter segments than crossover recombinations,
-typically below 1kb, and thus affect linkage patterns differently than
-recombinations~\citep{korunes2017gene}.
-
-% other simulators
-Hudson's \ms\ optionally includes gene conversion and so does
-MaCS~\citep{chen2009fast}, while scrm~\citep{staab2015scrm} does not
-support gene conversions.
-As bacterial recombination can be modeled analogously to eukaryotic allelic
-gene conversion, simulators for microbial evolution often include gene conversion
-instead of recombination.
-This includes the backward simulator SimBac~\citep{brown2016simbac},
-the SMC based approach used in FastSimBac~\citep{demaio2017the},
-and forward in time tools within the SLiM framework~\citep{cury2020simulation}.
-
-In \msprime\ 1.0 we implemented the coalescent with gene conversion as defined by
-\cite{wiuf2000coalescent}, where gene conversion events are initiated between any
-two sites in the sequence at a constant (scaled) rate $\gamma$ and
-the length of the transferred segment is given by a geometric distributed
-random variable with mean length $L$.
-% Above I used the discrete case, but we could easily switch to a description for the
-% continuous case.
-We have implemented this model of gene
-conversion in \msprime\ 1.0, and validated the output against
-\ms\ and analytical results~\citep{wiuf2000coalescent}.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=7cm]{figures/gc-perf}
-\end{center}
-\caption{\label{fig-gc-perf}Comparison of the running time
-for E-coli simulations using \msprime\ and SimBac for varying
-sample sizes. The parameters used are chosen to mimic the simulation
-of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
-gene conversion rate was set to $N_e g = \gamma = 0.015$ and
-the mean gene conversion tract length was set to $L = 500$, which is close to
-estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
-$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
-with mean tract length $L = 542$~\citep{lapierre2016the}.}
-\end{figure}
-
-Gene conversion is particularly useful to model homologous recombination in
-bacterial evolution,
-so we compare the performance of \msprime\ with gene conversion to
-SimBac~\citep{brown2016simbac}, a simulator developed specifically to
-simulate bacterial evolution. Although \msprime\ is missing some features
-relevant to this case, e.g.\ the simulation of circular genomes, it is far
-more efficient at simulating the basic processes.
-Furthermore, gene conversion can be considered in combination
-with other current and future features of \msprime\ 1.0,
-including varying recombination rates and complex demographies.
-
-An important avenue for future work
-will be to add further features to model bacterial
-evolution. In particular, we aim to include bacterial gene transfer and utilize
-the succinct tree sequence structure to represent the resulting ancestral gene
-transfer graph~\citep{baumdicker2014AGTG}.
-
-\subsection*{Demography}
-% TODO should we put in some citations here for the models? Hardly seems
-% necessary.
-One of the key applications of population genetic simulations is to generate
-data for complex demographies. Beyond idealised cases such as stepping-stone or
-island models, or specialised cases such as isolation-with-migration models,
-analytical results are rarely possible. Simulation is therefore integral to the
-development and evaluation of methods for demographic inference. The demography
-model in \msprime\ is directly derived from the approach used in \ms, and
-supports an arbitrary number of randomly mating populations exchanging migrants
-at specified rates. A range of demographic events are supported, which allow for
-varying population sizes and growth rates, changing migration rates over time,
-as well as population splits, admixtures and pulse migrations.
-The location of sampled lineages
-can be tracked through time in as much detail as required: each tree node is
-automatically associated with the population in which it arose, the location of
-lineages can be recorded at any given time via census events, or every lineage
-migration can be recorded.
-
-A major change for \msprime\ 1.0 is the introduction of the new Demography API,
-designed to address a serious flaw in the \msprime\ 0.x interface which lead to
-a number of avoidable errors in downstream
-simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
-separate parameters be provided to the \texttt{simulate} function to describe a
-demographic model, making it easy to accidentally omit information. The 1.0 API
-resolves this issue by creating a new \texttt{Demography} class, which
-encapsulates all information about the demographic model, and fully decouples
-the definition from other simulation details. An instance of this class is then
-provided as a parameter to the new \texttt{sim\_ancestry} function,
-substantially reducing the potential for error.
-Another improvement over the
-0.X APIs is the introduction of explicit population split and admixture events,
-and a population state machine that ensures that lineages cannot migrate into
-(or be sampled from) inactive populations. This demography model is compatible
-with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
-class supports importing and exporting Demes models.
-Models previously constructed using the 0.x API can be seamlessly imported into
-the \texttt{Demography} class, and we also support importing demographic
-models in the form of Newick species tree descriptions, as well as those
-produced by programs such as *BEAST~\citep{heled2009bayesian}.
-
-The \texttt{DemographyDebugger} can output detailed information for debugging
-demographic models, but also provides
-methods to compute a suite of numerical predictions about the location of
-lineages over time, for example the coalescence rates for two or more lineages
-drawn from populations at specified times in the past.
-Many popular approaches in population genetics use the distribution of
-coalescence rates between pairs of lineages in one or more populations to infer
-effective population sizes over
-time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
-split times and subsequent migration rates between
-populations~\citep{wang2020tracking}.
-The \texttt{DemographyDebugger} provides a valuable
-ground-truth when evaluating such inference methods, as illustrated by
-\cite{adrion2020community}.
-
-\subsection*{Instantaneous bottlenecks}
-
-A common approach to modelling the effect of demographic history on genealogies
-is to assume that effective population size ($N_e$) changes in discrete steps
-which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
-keightley2007joint,li2011inference}. In this setting of piecewise constant
-$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
-reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
-recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
-bottlenecks are short both on the timescale of coalescence and mutations,
-there may be little information about the duration of a bottleneck
-($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
-assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
-and generate a sudden burst of coalescence events (a multiple merger event) in
-the genealogy. The strength of the bottleneck $B$ can be thought of as an
-(imaginary) time period during which coalescence events are collapsed,
-i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
-single pair of lineages entering the bottleneck coalesce during the bottleneck
-is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
-attractive and both analytic results and empirical
-inference~\citep{griffiths1994sampling, galtier2000detecting,
-bunnefeld2015inferring} have been developed under this model, there has
-been no software available to simulate data under instantaneous
-bottleneck histories.
-
-We have implemented instantaneous bottlenecks in \msprime~1.0
-using a variant of Hudson's linear time single-locus coalescent
-algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
-by adding events to the Demography object (see the Demography section)
-and can be used in combination with any other demographic modelling
-features. We have validated the results of these simulations by comparing
-against analytical expectations for coalescence times and the
-site frequency spectrum~\citep{bunnefeld2015inferring}.
-
-\subsection*{Multiple merger coalescents}
-
-Kingman's coalescent assumes that only two ancestral lineages can merge at
-each merger event. Although this is generally a reasonable approximation there
-are certain situations in which the underlying mathematical assumptions are
-violated. For example in certain highly fecund organisms
-\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
-to produce numbers of offspring on the order of the population size and
-therefore a few individuals may produce the bulk of the offspring in any given
-generation~\citep{hedgecock_94}. These population dynamics violate basic
-assumptions of the Kingman coalescent, and are better modelled by
-`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
-two lineages can merge in a given event. Multiple-merger coalescent processes
-have also been shown to be relevant for modeling the effects of selection on
-gene genealogies \citep{Gillespie909,DS04}.
-
-Although multiple merger coalescents have been of significant theoretical
-interest for around two decades, there has been little practical software
-available to simulate these models.
-\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
-simulate a related spatial continuum model~\citep{barton2010new},
-\cite{zhu2015hybrid} simulate genealogies within a species tree
-based on a multiple-merger model, and
-\cite{becheler2020occupancy} provide a general method for simulating
-multiple merger processes
-as part of the Quetzal framework~\citep{becheler2019quetzal}.
-The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
-includes a number of extensions to the Beta-coalescent.
-None of these methods work with large genomes, and very little work
-has been performed on simulating multiple merger processes with recombination.
-
-We have added two multiple merger coalescent models in \msprime\ 1.0, the
-Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
-such models with recombination efficiently for the first time.
-These simulation models have been extensively validated against
-analytical results from the site frequency
-spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
- as well as more general properties of coalescent processes.
-Please see the Appendix for more details and model derivations.
-
-\subsection*{Ancestral Recombination Graphs}
-\label{sec-arg}
-
-The Ancestral Recombination Graph (ARG) was introduced by
-Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
-representation of the coalescent with recombination stochastic process. This
-formulation was complementary to Hudson's earlier
-work~\citep{hudson1983properties}, and substantially increased our theoretical
-understanding of recombination. In Griffiths' ARG formulation, we
-regard a realisation of the coalescent with recombination as a graph in which
-vertices represent common ancestor or recombination events, and edges represent
-lineages through which ancestral material flows.
-We may consider the big ARG, in which we track lineages arising out of
-recombinations regardless of whether they carry ancestral material and
-consequently wait an exponentially large time until all sampled lineages
-coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
-which we only track genetic ancestors, and wait until all marginal trees
-have fully coalesced (like Hudson's algorithm).
-
-\begin{figure}
-
-\begin{center}
-    \begin{tikzpicture}[scale=0.45]
-
-    \draw (-2,0) -- (-2,3) -- (0, 3);
-    \draw (1,0) -- (1,1);
-    \draw (0,1) -- (2,1) -- (2,2) -- (4,2) -- (4,0);
-    \draw (3,2) -- (3,5);
-    \draw (0,1) -- (0,3);
-    \draw (-1, 3) -- (-1, 5) -- (3,5);
-
-    \draw[fill] (-2,0) circle [radius=0.05];
-    \draw[fill] (-1,3) circle [radius=0.05];
-    \draw[fill] (1,0) circle [radius=0.05];
-    \draw[fill] (4,0) circle [radius=0.05];
-    \draw[fill] (1,1) circle [radius=0.05];
-    \draw[fill] (3,2) circle [radius=0.05];
-    \draw[fill] (1,5) circle [radius=0.05];
-
-    \node [below] at (-2,0) {0};
-    \node [below] at (1,0) {1};
-    \node [below] at (4,0) {2};
-    \node [above] at (1,1) {3};
-    \node [below right] at (1,1) {$x = 0.3$};
-    \node [above right] at (3,2) {4};
-    \node [above left] at (-1,3) {5};
-    \node [above] at (1,5) {6};
-
-    \end{tikzpicture}
-% \begin{center}
-% $\equiv$
-% \end{center}
-% \end{column}
-% \begin{column}{5cm}
-\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
-\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
-\includegraphics[width=3.8cm]{figures/arg.pdf}
-
-\end{center}
-\caption{\label{fig-arg} (A) An simple ARG in which a recombination
-occurs at position 0.3; (B) the equivalent topology depicted as a tree
-sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation;
-(D) the fraction of nodes in a tree sequence that are concerned with
-extra ARG information in a tree sequences returned using \msprime's
-\texttt{record\_full\_arg} mode}
-\end{figure}
-
-However, this original definition of the ARG in terms of a stochastic process
-is rarely what is meant when the term is currently used. The term has come to
-essentially mean any representation of genetic ancestry in the presence
-of recombination as a graph-like structure; for example,
-\cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material.
-Others state that an ARG is equivalent to the sequence of marginal
-% Also SARGE paper
-trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}
-
-Such
-imprecision is unfortunate, since there are any number of graphical encodings of
-genetic ancestry, and referring to the original Griffiths papers does little to
-clarify. As we see below, there are situations in which the precise information
-stored in the graph is important, and conflating a high-level
-discussion of genetic ancestry with this mathematical object is not
-helpful. The point of technical terms is to provide clarity and
-precision on what is being discussed, and it is now essentially
-impossible to know what is meant by the term ``ARG'';
-we therefore use the term ``genetic ancestry'' in high-level
-discussions.
-
-For our purposes, an ARG is a realisation of the coalescent with
-recombination, in the Griffiths (little ARG) sense.
-As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
-works by dynamically generating and traversing a little ARG.
-The graph is not explicitly represented in memory, but is partially
-present through the extant lineages and the ancestral material they carry
-over time. Since the ARG is a large and unwieldy object
-(Fig.~\ref{fig-arg}) we do not use it as output, but
-rather store the information required to recover the genealogical
-history as nodes and edges in a tree sequence (see section XX),
-or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
-
-Although the tree sequences output by \msprime\ contain full information
-about the genealogical trees, their correlation structure along the genome,
-and the ancestral haplotypes on which coalescences occured, some information
-is lost in this mapping down from ARG space to the minimal tree sequence
-form. In particular, we lose
-information about ancestral haplotypes that were common ancestors but
-in which no coalescences occured, and also information about the precise time
-and genomic location of recombination events. In most cases, such
-information is of little relevance as it is in principle unknowable,
-but there are occasions such as visualisation or computing likelihoods (see
-below) in which is useful. We therefore provide
-the \texttt{record\_full\_arg} option in \msprime\
-to store a representation of the complete ARG traversed during simulation.
-This is done by storing extra nodes (marked with specific flags, so they
-can be easily identified later) and edges in the tree sequence.
-Fig.~\ref{fig-arg} shows that this extra ARG information
-quickly comes to dominate the output as we increase the sequence length,
-and the resulting tree sequence files are many times larger
-than the standard minimal tree sequence representation.
-
-One situation in which a record of the full ARG is necessary is when we
-wish to compute likelihoods during inference.
-The likelihood is a central quantity in evaluating the plausibility of a putative
-ancestry as an explanation of DNA sequence data, both directly through
-e.g.~approaches based on maximum likelihood, and as an ingredient of
-methods such as
-Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
-We provide functions to compute the likelihood of ARG realisations
-(encoded using the representation of the \texttt{record\_full\_arg} option)
-and mutational patterns under
-the standard coalescent and infinite sites mutation model.
-See the Appendix for details.
-
-\subsection*{Selection}
-\label{sec-selection}
-Another ellaboration of the standard neutral coalescent with recombination
-is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
-In the context of the coalescent this amounts to creating a structured population
-during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
-in which lineages may transit between favored and unfavored backgrounds through
-recombination. This allows for many selective sweep scenarios, including recurrent, partial,
-and soft selective sweeps to be simulated efficiently, however in general this
-efficiency comes at the cost of flexibility in comparison to forwards in
-time simulation.
-
-Currently there are a small number of other simulators that will generate
-coalescent samples with sweeps including, SelSim~\citep{spencer2004selsim}
-mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
-Cosi2~\citep{shlyakhter2014cosi2}, and discoal~\citep{kern2016discoal}.
-Each of these packages has their own benefits and possible drawbacks,
-however none of them at present will output tree sequence formatted data.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=6cm]{figures/sweeps-perf}
-\end{center}
-\caption{\label{fig-selection-perf}Comparison of the runtime for
-\msprime\ and \texttt{discoal} for simulating selective sweeps.
-}
-\end{figure}
-
-We have implemented selective sweeps in \msprime\ similarly to
-the \texttt{discoal} implementation-- sweep trajectories are
-generated from a jump process approximation to the conditional diffusion
-of an allele bound for fixation~\citep{coop2004ancestral}.
-This implementation allows us to further generalize sweep models to
-changing population sizes as well as selection from standing variation,
-both of which are features that are planned for the near future.
-Moreover because this implements the basic infrastructure for the structured
-coalescent, adding support for, e.g.,
-inversions is straightforward~\citep{peischl2013sequential}.
-
-\subsection*{Discrete time Wright-Fisher}
-
-The assumptions of the classical coalescent model are that both sample size and
-the simulated sequence length are small. As modern datasets often include
-whole-genome sequences for tens or hundreds of thousands of samples, simulating
-such data requires breaking these assumptions. While the coalescent remains a
-good approximation to the Wright-Fisher model in the distant past, it can
-exhibit large biases in the recent history. Identity-by-descent (IBD),
-long-range linkage disequilibrium (LD), and ancestry patterns deviate from
-Wright-Fisher expectations, and this bias grows considerably when simulating
-larger sample
-sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
-
-A backwards-in-time Wright-Fisher model is supported in \msprime, which is
-shown to properly capture the expected distributions of IBD and long-range LD,
-as well as ancestry patterns in admixed
-populations~\citep{nelson2020accounting}. In short, the classical coalescent
-draws recombination events from exponential waiting times, which are continuous
-in time, and each recombination then splits a segment into two distinct
-ancestors. For large sequence lengths, this can dramatically overestimate the
-number of genealogical ancestors of a given sample. Instead, the Wright-Fisher
-model iterates through discrete generations so that each gamete has a unique
-diploid parent, and multiple recombinations within a generation results in
-crossover events between the same two parental haploid copies.  This method is
-described in more detail in~\cite{nelson2020accounting}.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=6cm]{figures/dtwf-perf}
-\end{center}
-\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
-simulations with ARGON. To ensure the comparison with ARGON is
-as fair as possible we run it with a mutation rate of 0 and with
-the minimum output options (so that it writes out only an empty VCF file)}
-\end{figure}
-
-We support changing the simulation model at specified times so that we can
-easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}. We
-can combine any number of different simulation models, allowing for the
-flexible choice of simulation scenarios without severely damaging efficiency.
-As the discrete time Wright-Fisher model improves accuracy of genealogical
-patterns in the recent past, we can simulate the recent history using this
-model and then switch to the standard coalescent to more efficiently simulate
-the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the perfomance of
-this approach with ARGON~\citep{palamara2016argon}, a specialised DTWF
-simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
-simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
-hybrid approach provides large efficiency gains.
-
-\subsection*{Simulation interface}
-\label{sec-sim-interface}
-
-The majority of simulation packages are controlled either through
-a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
-a text-based input file
-format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
-or a mixture of both.
-Command line interfaces such as \ms\ make it very easy to run simple
-simulations, but as model complexity and the number of parameters specified increases
-they become difficult to understand and
-error-prone~\citep{ragsdale2020lessons,gower2021demes}.
-Specifying parameters through a text file alleviates this problem to a degree,
-but lacks flexibility, for example, when running simulations with parameters
-drawn from some distribution. In practice, for any significant simulation
-project (or at least, any reproducible one), users will write a script
-to generate the required command lines or input parameter files,
-invoke the simulation engine, and process the results in some way.
-This process is cumbersome and labour intensive, and
-a number of packages have been developed
-to allow simulations be run directly in a high-level
-scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
-
-The more recent trend has been to move away from this file and command-line
-driven approach and to instead provide direct interfaces to the simulation
-engines via an Application Programming Interface (API)~\citep[e.g.][]{
-thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
-The primary interface for \msprime\ is through a thoroughly documented and
-stable Python
-API, which has allowed a number of downstream tools to be built on top of
-% Found these through the Dependent packages on GitHub
-it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
-adrion2020community,adrion2020predicting, kamm2020efficiently,
-mckenzie2020ipcoal, montinaro2020revisiting,
-de2021geonomics,rivera2021simulation}.
-As well as providing a stable and efficient platform for building
-downstream applications, \msprime's Python API makes it much easier to
-build reproducible simulation pipelines, as the entire workflow can
-be encapsulated in a single script, and package and version
-dependencies can be explicitly stated using the \texttt{pip}
-or \texttt{conda} package managers.
-For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
-in the influential simulation analysis of
-\cite{martin2017human} were only detected because the pipeline
-could be easily run and reanalysed.
-
-A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
-designed in part to avoid sources of error (see the Demography section) but
-also to provide more appropriate defaults while keeping compatibility with
-existing code. In the new APIs, ancestry and mutation simulation are fully
-separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
-functions replacing the legacy \texttt{simulate} function. Among other changes,
-the new APIs default to discrete genome coordinates and finite sites mutations,
-resolving a major source of confusion and error. The 0.x APIs are fully
-supported and tested, and will be maintained for the foreseeable future.
-We also provide an \ms\ compatible
-command line interface to support existing workflows, and the
-\texttt{msp} program that supports specifying demographic models in
-Demes format~\citep{gower2021demes} from the command line.
-
 \subsection*{Development model}
@@
 
+\section*{Theoretical runtime}
+    \label{app-theoretical-runtime}
+
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/ancestry-perf-expected}
+\end{center}
+\caption{\label{fig-expected-perf}
+    Data from Figure~\ref{fig-ancestry-perf}, compared with
+    equation~\eqref{eqn-hsw-bound} (``expected'').
+    Left panel shows observed times plotted against expected number
+    of operations; right panel shows the ratio of the two, plotted against $\rho$.
+}
+\end{figure}
+
+The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
+on the number of events involving trapped ancestral material in Hudson's algorithm.
+For $n$ samples from 
+a diploid population of effective size $N_e$,
+and a genome of genetic length $r$,
+with $\rho = 4 N_e r$ the scaled recombination rate,
+and $R^T$ the number of recombination recombination events in trapped material,
+this upper bound is
+\begin{equation} \label{eqn-hsw-bound}
+    \mathbb{E}[R^T]
+    \le
+    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
+\end{equation}
+Note that the sum on the right is well-approximated by $\log(n)$.
+Although \citet{hein2004gene} refer to this bound as ``crude'',
+it turns out to predict the scaling of runtime quite well for large genomes.
+Figure~\ref{fig-expected-perf} shows the observed runtime
+(from the same simulations as Figure~\ref{fig-ancestry-perf})
+plotted against equation~\eqref{eqn-hsw-bound},
+showing that for values of $\rho$ above $4 \times 10^4$ or so
+(i.e., chromosome-length simulations in populations above $10^4$),
+the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
+This implies the runtime for a large sample
+can be predicted from the runtime for a small sample.
+For smaller values of $\rho$, however, the prediction
+severely underestimates runtime.
+
+\section*{Likelihood calculations}
 \label{app-likelihoods}
-\section*{Likelihood calculations}
 

### 2021-07-26 paper.tex
appendix

--- 2021-07-25 paper.tex
@@
     The black line is a quadratic fit separately in each panel.
-    \plrcomment{TODO: run out more simulations to get more triangles}
 }
@@
 and verifies that runtime scales quadratically with $\rho$,
+(\citet{hein2004gene} gives an equation
+including the effects of sample size --
+see Appendix~\ref{app-theoretical-runtime}.)
 and depends much less strongly on the number of samples.
@@
 only a small fraction of the common ancestor events
-will involve overlapping segments of ancesty,
+will involve overlapping segments of ancestry,
 and lead to coalescence in the marginal trees.
@@
 does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
-The answer depends on how $T$ commpares to $N_1$: if $T/N_1$ is large,
+The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
 then runtime will be similar to a population of size $N_1$;
@@
 and then predict runtime by fitting a quadratic curve to genome length.
+There is only a weak dependency on sample size
+-- notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
+only increases runtime by a factor of 2 or so.
 

### 2021-07-26 evaluation/generate_ancestry_perf_data.py

@@ -1,5 +1,6 @@
 import time
 
+import cpuinfo
 import msprime
 import numpy as np
 
@@ -26,6 +27,11 @@ def run_sims():
     N_VALS = [1000, 5000, 10000, 50000, 100000, 200000, 300000]
     ln_pairs = [(L, N) for L in L_VALS for N in N_VALS]
     ln_pairs.sort(key=lambda x: x[0] * x[1])
+
+    cpu = cpuinfo.get_cpu_info()
+    with open("data/ancestry_perf_cpu.txt", "w") as f:
+        for k, v in cpu.items():
+            print(k, "\t", v, file=f)
 
     with open(outfile, "w") as f:
         f.write(csv(["N", "L", "num_samples", "height", "num_trees", "time"]))

### 2021-07-26 evaluation/plot.py

@@ -231,15 +231,25 @@ def ancestry_perf():
     axes[1].legend(prop={'size': 6})
     save("ancestry-perf")
 
-    # Compare to Hein, Schierup & Wiuf:
-    # recombination rate is 1e-8, so rho = N * L * 1e-8
+
+@click.command()
+def ancestry_perf_expected():
+    """
+    Compare the ancestry benchark to Hein, Schierup & Wiuf:
+    recombination rate is 1e-8, so rho = N * L * 1e-8
+    """
+
+    df = pd.read_csv(
+        "data/ancestry-perf.csv", sep=",",
+    )
+
     h = np.cumsum(1/np.arange(1, np.max(df["num_samples"]) + 1))
     rho = df["L"] * df["N"] * 1e-8
     expected = rho * (rho + 1) * (h[df["num_samples"].astype("int") - 1] ** 2)
 
     fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(6, 3))
 
-    ax0.set_xlabel("expected time (rho * (rho+1) * log(num_samples))")
+    ax0.set_xlabel("expected time (theory)")
     ax0.set_ylabel("observed time (seconds)")
     for ns, m in zip((1000, 100000), ("o", "v")):
         ut = (df["num_samples"] == ns)
@@ -258,17 +268,7 @@ def ancestry_perf():
                 (df["time"] / expected)[ut],
                 marker=m,
         )
-    ax1.set_xscale("log")
     ax1.set_yscale("log")
-
-    # ax1.set_xlabel("number of samples")
-    # ax1.set_ylabel("observed / expected time")
-    # sc = ax1.scatter(
-    #         df["num_samples"], df["time"] / expected,
-    #         c=np.log10(df["N"] * df["L"] / 1e8), cmap="cool",
-    # )
-    # ax1.set_yscale("log")
-    # plt.colorbar(sc, label="log10(rho)")
 
     save("ancestry-perf-expected")
 
@@ -293,6 +293,7 @@ cli.add_command(gc_perf)
 cli.add_command(sweeps_perf)
 cli.add_command(dtwf_perf)
 cli.add_command(ancestry_perf)
+cli.add_command(ancestry_perf_expected)
 cli.add_command(arg)
 
 if __name__ == "__main__":

### 2021-07-26 paper.tex
Mutation stuff.

--- 2021-07-26 paper.tex
@@
 genomes~\citep[e.g.][]{mathieson2020ancestry},
-and in this sense a succint tree
+and in this sense a succinct tree
 sequence \emph{is} an ARG. More technically, an ARG is a particular
@@
 (see Fig.~\ref{fig-mutated-trees}).
-A number of programs exist to place mutations on trees:
-for instance,
-the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
-allows a range of different models of sequence evolution to be imposed on a given
-tree in Newick format, and there are various
-extensions to the basic
+A number of programs exist to place mutations on trees: for instance,
+the classical \SeqGen\ program~\citep{rambaut1997seq}
+supports a range of different models of sequence evolution,
+and there are various extensions to the basic
 models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
-Population genetic simulators have tended to
+Partly for efficiency and partly in the interest of
+simplicity for users (i.e., to avoid intermediate text format conversions),
+population genetic simulators have tended to
 include their own implementations of mutation simulation, with
@@
 
-Generating simulations in \msprime\ is extremely efficient.
+Generating simulations in \msprime\ is efficient.
 Fig.~\ref{fig-mutations-perf} shows the time required to generate
@@
 tree ($\sim500$ times more work).
-
 % python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
-Even if we consider a single tree, \msprime\ generates mutations
-extremely efficiently.
+Even if we have a tree sequence consisting of a single tree
+(so that there is no advantage to working edge-by-edge),
+\msprime's mutation generator is quite efficient.
 For example, we simulated mutations under the BLOSUM62 amino
@@
 There are many ways in which the mutation generation code
-in \msprime\ could be extended by adapting existing methods.
-For example, we hope to add support for
+in \msprime\ could be extended. For example, we hope to add support for
 microsatellites~\citep{mailund2005coasim},

### 2021-07-26 paper.tex
Pass through mutations section

@@
  data stucture to represent simulation results.
-The succinct tree sequence (usually abbreviated to tree sequence)
+The succinct tree sequence (usually abbreviated to ``tree sequence'')
 is a data structure for concisely encoding
@@
 A node in tree sequence terminology corresponds
-to a single chromosome strand, and is associated with its birth time,
-and other optional metadata. A node may optionally be
+to a single chromosome strand, associated with its birth time
+and other optional metadata. A node may be
 associated with an \emph{individual}, so that, for example, a diploid
-individual will be associated with two separate nodes (we omit
+individual is associated with two separate nodes (we omit
 the individual table in Fig.~\ref{fig-ts-example} as it is optional).
@@
 % OK, back to tree sequences: what about sequence variation?
-The final output of most population simulations is some representation
-of the sequence variation among the specified samples. For coalescent
-simulations, we usually have three steps (1) simulate
+The final output of most population genetic simulations is some representation
+of sequence variation among the specified samples. For coalescent
+simulations, we usually have three steps: (1) simulate
 the genetic ancestry, and optionally output the resulting marginal trees;
@@
 supports a range of different models of sequence evolution,
-and there are various extensions to the basic
-models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
+and various extensions to the basic
+models have been proposed~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
 Partly for efficiency and partly in the interest of
@@
 model~\citep[e.g.][]{hudson2002generating}
-but with several supporting a range of different models
-of sequence
+but with several supporting a wide range of different models of sequence
 evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
-virgoulay2021gspace}.
+virgoulay2021gspace}. Thus, despite the logical separation between
+the tasks of simulating ancestry and neutral sequence evolution,
+the two have been conflated in practice.
 
@@
         \plrcomment{TODO: fix up the axis labels, if we like this figure?}
+        \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
     }
@@
 
-% on the other hand, tskit is awesome.
-Clean separation of the ancestry and mutation generation steps
-is made possible by the \tskit\ library, whose mature and
-efficient C and Python interfaces makes it straightforward to add
-further information to an existing file, and to output the results.
-As such, it is essentially a historical accident
-that \msprime\ (whose main focus is on ancestry simulation)
-contains code to simulate mutations: the mutation process could
-be performed by a separate package, and there would be no performance
-penalty for doing so.
+Part of the reason for this poor record of software reuse and modularity is the
+lack of standardised file formats, and in particular, the lack of common
+library infrastructure to abstract the details of interchanging simulation
+data. On the surface, it would appear that \msprime\ is also guilty of this
+lack of modularity since it supports simulating both ancestry and mutations.
+However, this is essentially a historical accident and because of the \tskit\
+library, the mutation process in \msprime\ could just as easily
+be performed in a separate package.
+The mature and efficient C and Python interfaces for
+\tskit\ makes it straightforward to add further information to
+an existing file, and because of its efficient data interchange mechanisms,
+there is no performance penalty for performing additional operations
+in a different Python package.
 Thanks to this interoperability, \msprime's
@@
 
-Generating simulations in \msprime\ is efficient.
+Simulating mutations in \msprime\ is efficient.
 Fig.~\ref{fig-mutations-perf} shows the time required to generate
@@
 (so that there is no advantage to working edge-by-edge),
-\msprime's mutation generator is quite efficient.
+\msprime's mutation generator is still very efficient.
 For example, we simulated mutations under the BLOSUM62 amino
@@
 codon models~\citep{arenas2007recodon}
-and indels~\citep{cartwright2005dna,fletcher2009indelible}.
+and indels~\citep{cartwright2005dna,fletcher2009indelible},
+although some may also require upstream additions to \tskit's data model
+which is currently based on the assumption of independent sites.
 

### 2021-07-26 paper.tex
Move the Simulation Interface up to the top.

@@
 
+\subsection*{Simulation interface}
+\label{sec-sim-interface}
+
+The majority of simulation packages are controlled either through
+a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
+a text-based input file
+format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
+or a mixture of both.
+Command line interfaces such as \ms\ make it very easy to run simple
+simulations, but as model complexity and the number of parameters specified increases
+they become difficult to understand and
+error-prone~\citep{ragsdale2020lessons,gower2021demes}.
+Specifying parameters through a text file alleviates this problem to a degree,
+but lacks flexibility, for example, when running simulations with parameters
+drawn from some distribution. In practice, for any significant simulation
+project (or at least, any reproducible one), users will write a script
+to generate the required command lines or input parameter files,
+invoke the simulation engine, and process the results in some way.
+This process is cumbersome and labour intensive, and
+a number of packages have been developed
+to allow simulations be run directly in a high-level
+scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
+
+The more recent trend has been to move away from this file and command-line
+driven approach and to instead provide direct interfaces to the simulation
+engines via an Application Programming Interface (API)~\citep[e.g.][]{
+thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
+The primary interface for \msprime\ is through a thoroughly documented and
+stable Python
+API, which has allowed a number of downstream tools to be built on top of
+% Found these through the Dependent packages on GitHub
+it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
+adrion2020community,adrion2020predicting, kamm2020efficiently,
+mckenzie2020ipcoal, montinaro2020revisiting,
+de2021geonomics,rivera2021simulation}.
+As well as providing a stable and efficient platform for building
+downstream applications, \msprime's Python API makes it much easier to
+build reproducible simulation pipelines, as the entire workflow can
+be encapsulated in a single script, and package and version
+dependencies can be explicitly stated using the \texttt{pip}
+or \texttt{conda} package managers.
+For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
+in the influential simulation analysis of
+\cite{martin2017human} were only detected because the pipeline
+could be easily run and reanalysed.
+
+A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
+designed in part to avoid sources of error (see the Demography section) but
+also to provide more appropriate defaults while keeping compatibility with
+existing code. In the new APIs, ancestry and mutation simulation are fully
+separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
+functions replacing the legacy \texttt{simulate} function. Among other changes,
+the new APIs default to discrete genome coordinates and finite sites mutations,
+resolving a major source of confusion and error. The 0.x APIs are fully
+supported and tested, and will be maintained for the foreseeable future.
+We also provide an \ms\ compatible
+command line interface to support existing workflows, and the
+\texttt{msp} program that supports specifying demographic models in
+Demes format~\citep{gower2021demes} from the command line.
+
 \subsection*{Simulating mutations}
@@
 
-\subsection*{Simulation interface}
-\label{sec-sim-interface}
-
-The majority of simulation packages are controlled either through
-a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
-a text-based input file
-format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
-or a mixture of both.
-Command line interfaces such as \ms\ make it very easy to run simple
-simulations, but as model complexity and the number of parameters specified increases
-they become difficult to understand and
-error-prone~\citep{ragsdale2020lessons,gower2021demes}.
-Specifying parameters through a text file alleviates this problem to a degree,
-but lacks flexibility, for example, when running simulations with parameters
-drawn from some distribution. In practice, for any significant simulation
-project (or at least, any reproducible one), users will write a script
-to generate the required command lines or input parameter files,
-invoke the simulation engine, and process the results in some way.
-This process is cumbersome and labour intensive, and
-a number of packages have been developed
-to allow simulations be run directly in a high-level
-scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
-
-The more recent trend has been to move away from this file and command-line
-driven approach and to instead provide direct interfaces to the simulation
-engines via an Application Programming Interface (API)~\citep[e.g.][]{
-thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
-The primary interface for \msprime\ is through a thoroughly documented and
-stable Python
-API, which has allowed a number of downstream tools to be built on top of
-% Found these through the Dependent packages on GitHub
-it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
-adrion2020community,adrion2020predicting, kamm2020efficiently,
-mckenzie2020ipcoal, montinaro2020revisiting,
-de2021geonomics,rivera2021simulation}.
-As well as providing a stable and efficient platform for building
-downstream applications, \msprime's Python API makes it much easier to
-build reproducible simulation pipelines, as the entire workflow can
-be encapsulated in a single script, and package and version
-dependencies can be explicitly stated using the \texttt{pip}
-or \texttt{conda} package managers.
-For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
-in the influential simulation analysis of
-\cite{martin2017human} were only detected because the pipeline
-could be easily run and reanalysed.
-
-A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
-designed in part to avoid sources of error (see the Demography section) but
-also to provide more appropriate defaults while keeping compatibility with
-existing code. In the new APIs, ancestry and mutation simulation are fully
-separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
-functions replacing the legacy \texttt{simulate} function. Among other changes,
-the new APIs default to discrete genome coordinates and finite sites mutations,
-resolving a major source of confusion and error. The 0.x APIs are fully
-supported and tested, and will be maintained for the foreseeable future.
-We also provide an \ms\ compatible
-command line interface to support existing workflows, and the
-\texttt{msp} program that supports specifying demographic models in
-Demes format~\citep{gower2021demes} from the command line.
-
 \subsection*{Analysis of simulation results}
@@
 which break this pattern of monolithic simulation software.
-
-Thus, development of \msprime\ is not completed and much remains
-to be done.
 

### 2021-07-26 paper.tex
Move more stuff up to the top

@@
 
+% The data model
 \subsection*{Tree sequences}
@@
 
-\subsection*{Simulation interface}
+% The input - how do we run simulations?
+\subsection*{User interface}
 \label{sec-sim-interface}
@@
 
+% The output - what do we get out?
+\subsection*{Data analysis}
+\label{sec-data-analysis}
+The point of running simulations is to generate data, and there
+are two important ways in which the \tskit\ tree sequences output by
+\msprime\ are more useful than established methods: (1) tree
+sequences are more efficient; and (2) tree sequences contain
+more information. We will consider these points in turn.
+
+The standard way of representing simulation data is
+to render the results in a text format. For example,
+\ms~\citep{hudson2002generating} outputs a set of sequences
+and can also optionally output the marginal trees along
+the genome in Newick format,
+and variants of this approach are used by many simulators.
+Text files have many advantages, but are slow to process
+at scale.
+The ability to efficiently process simulation results is
+particularly important in simulation-based inference methods
+such as Approximate Bayesian Computation
+(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
+and machine learning based
+approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
+flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
+crucial since the size and number of simulations that can be performed determines
+the depth to which one can sample from the model and parameter space.
+Equally important,
+however, is the efficiency with which the simulation results can be
+transformed into the specific input required by the inference method.
+In the case of ABC, this is usually a set of summary statistics of the sequence
+data, and methods avoid the bottleneck of parsing
+text-based file formats to compute these statistics
+by either developing their own
+simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
+or creating forked versions of existing
+simulators~\cite[e.g.][]{thornton2006approximate,
+hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
+tightly integrated with the inference method.
+Modern approaches to ABC such as
+ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
+ABC-NN~\citep{csillery2012abc,blum2010abc} use large
+numbers of weakly informative statistics,
+making the need to efficiently compute statistics from simulation
+results all the more acute.
+
+Classical text based output formats like \ms\ are inefficient to process,
+but also lack a great deal of important information.
+The tree-by-tree topology information output by simulators in Newick
+format lacks any concept of node (in the tree sequence sense) identity,
+and means that we cannot reliably infer information about ancestors
+from the output. Because Newick stores branch lengths rather
+than node times, numerical precision issues also arise when summing
+over many short branches~\citep{mcgill2013graphml}.
+% JK: not sure this is worth leaving in tbh
+This lack of reliable information about ancestors in simulation data
+has lead, for example, to the definition of segments of
+identity by descent as a set of samples having the same time
+to their most recent common ancestor along some section of the
+genome~\citep{yang2016rapidly}.
+Numerous forks
+of simulators have been created to access information not provided
+in the output. For example, \ms\  has been forked to
+output information about migrating segments~\citep{rosenzweig2016powerful},
+information about ancestral lineages~\citep{chen2013asymptotic},
+an array of population genetic statistics~\citep{ramos2007mlcoalsim}
+% Any more?
+and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
+has in turn been forked to output information on local
+ancestry~\citep{racimo2017archaic}.
+All of this information
+is either directly available by default in \msprime, or can be optionally
+stored via options such as \texttt{record\_migrations} or
+\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
+can be efficiently and conveniently processed via \tskit\ APIs.
+
+These problems, of inefficient and limited access to the simulated
+data are comprehensively solved by \msprime\ and \tskit. Complete
+information about ancestry and mutation simulations is immediately
+available to the user in a single integrated data structure
+(rather than multiple independent files), with no intermediate
+processing required.
+Users can subsequently compute precisely the information that they
+require using the powerful suite of population genetic statistics
+and other methods available in \tskit.
+
+% Stuff we simulate: mutations
 \subsection*{Simulating mutations}
@@
 hybrid approach provides large efficiency gains.
-
-\subsection*{Analysis of simulation results}
-\label{sec-data-analysis}
-
-\jkcomment{This is probably part of the discussion, but putting here out of
-the way for now in its own subsection}
-
-The point of running simulations is to generate data, and there
-are two important ways in which the tree sequences output by
-\msprime\ are more useful than classical approaches: (1) tree
-sequences are more efficient; and (2) tree sequences contain
-more information. We will consider these points in turn.
-
-The classical approach to generating simulation data is
-to render the results in some text format. For example,
-\ms~\citep{hudson2002generating} outputs a set of sequences
-and can also optionally output the marginal trees along
-the genome in Newick format,
-and variants of this approach are used by many simulators.
-Text files have many advantages, but are slow to process
-at scale. With sample sizes becoming ever larger, processing
-text output can become a significant bottleneck in many analyses.
-The ability to efficiently process simulation results is
-particularly important in simulation-based inference methods
-such as Approximate Bayesian Computation
-(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
-and machine learning based
-approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
-flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
-crucial since the size and number of simulations that can be performed determines
-the depth to which one can sample from the model and parameter space.
-Equally important,
-however, is the efficiency with which the simulation results can be
-transformed into the specific input required by the inference method.
-In the case of ABC, this is usually a set of summary statistics of the sequence
-data, and methods avoid the bottleneck of parsing
-text-based file formats to compute these statistics
-by either developing their own
-simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
-or creating forked versions of existing
-simulators~\cite[e.g.][]{thornton2006approximate,
-hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
-tightly integrated with the inference method.
-While this is certainly more efficient,
-there is a great deal of duplication of effort and significant
-room for error (and propagation of bugs).
-Modern approaches to ABC such as
-ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
-ABC-NN~\citep{csillery2012abc,blum2010abc} use large
-numbers of weakly informative statistics,
-making the need to efficiently compute statistics from simulation
-results even more acute.
-
-Classical text based output formats like \ms\ are inefficient to process,
-but also lacking a great deal of important information.
-The tree-by-tree topology information output by simulators in Newick
-format lacks any concept of node (in the tree sequence sense) identity,
-and means that we cannot reliably infer information about ancestors
-from the output. Because Newick stores branch lengths rather
-than node times, numerical precision issues also arise when summing
-over many short branches~\citep{mcgill2013graphml}.
-% JK: not sure this is worth leaving in tbh
-This lack of reliable information about ancestors in simulation data
-has lead, for example, to the definition of segments of
-identity by descent as a set of samples having the same time
-to their most recent common ancestor along some section of the
-genome~\citep{yang2016rapidly}.
-Numerous forks
-of simulators have been created to access information not provided
-in the output. For example, \ms\  has been forked to
-output information about migrating segments~\citep{rosenzweig2016powerful},
-information about ancestral lineages~\citep{chen2013asymptotic},
-an array of population genetic statistics~\citep{ramos2007mlcoalsim}
-% Any more?
-and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
-has in turn been forked to output information on local
-ancestry~\citep{racimo2017archaic}.
-All of this information
-is either directly available by default in \msprime, or can be optionally
-stored via options such as \texttt{record\_migrations} or
-\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
-can be efficiently and conveniently processed via \tskit\ APIs.
-
-\jkcomment{Summary paragraph: tskit gives us complete information
-about the genetic ancestry. and is very efficient}
-
-% \jkcomment{A point to make somewhere in here: the tskit data model is
-% integrated, with all of the information that you need in one file. Not
-% splattered across a number of different files, that you have to keep track
-% of.}
 

### 2021-07-26 paper.tex
first draft of mutation appendix

--- 2021-07-26 paper.tex
@@
 \label{sec-ts}
+One of the key reasons for \msprime's significiant performance advantage
+over other simulators~\citep{kelleher2016efficient}
+is its usage of the ``succinct tree sequence''
+(usually, tree sequence) data stucture
+The basic concepts were introduced by \cite{kelleher2016efficient}
+and originally developed as part of \msprime. We have subsequently
+extracted the tree sequence code from \msprime\ into
+\tskit~\citep{tskit2021tskit}, a standalone Python and C library.
+While a full discussion of tree sequences and the capabilities
+of \tskit\ is beyond the scope of this article, we summarise
+the aspects that are important for simulation here.
 
@@
 
-One of the key reasons for \msprime's substantial performance advantage
-over other simulators~\citep{kelleher2016efficient}
-is its usage of the ``succinct tree sequence''
- data stucture to represent simulation results.
-The succinct tree sequence (usually abbreviated to ``tree sequence'')
-is a data structure for concisely encoding
-genetic ancestry and sequence variation.
-The basic concepts were introduced by \cite{kelleher2016efficient}
-and originally developed as part of \msprime. We subsequently
-extracted the tree sequence code that was independent of simulations
-from \msprime\ into \tskit,
-a standalone Python and C library~\citep{tskit2021tskit}.
-While a full discussion of tree sequences and the capabilities
-of \tskit\ is beyond the scope of this article, we summarise
-some aspects that are important for simulation here.
-
-% Encoding ancestry
-Given a set of sample chromosomes, the topology
+The succinct tree sequence is a data structure for concisely encoding
+genetic ancestry and variation. Given a set of sample chromosomes, the topology
 of their genetic ancestry is described by a set of \emph{nodes}
@@
 A node in tree sequence terminology corresponds
-to a single chromosome strand, associated with its birth time
-and other optional metadata. A node may be
+to a single chromosome strand, and is associated with its birth time,
+population and other metadata. A node may optionally be
 associated with an \emph{individual}, so that, for example, a diploid
-individual is associated with two separate nodes (we omit
+individual will be associated with two separate nodes (we omit
 the individual table in Fig.~\ref{fig-ts-example} as it is optional).
-Nodes define haplotypes that exist at a specific time within
+Nodes define haplotypes that exist at a specific time within a
 some individual, and edges define genetic inheritence
-relationships between these nodes. An \emph{edge} consists of a
+relationships between these nodes. An edge consists of a
 parent node, a child node and the left and right coordinates
@@
 0 to 4 is present in both marginal trees, and represented
-by a single edge (the first row in the edge table).
+by a single edge (the first row in the example edge table).
 This simple device, of explicitly associating tree nodes
@@
 genomic intervals over which parent/child relationships exist,
-generalises the original
-``coalescence records'' concept~\citep{kelleher2016efficient},
-and is the key to the efficiency of tree
+is a generalisation from the original concept  of
+``coalescence records''~\citep{kelleher2016efficient}, and
+is the key to the efficiency of tree
 sequences~\citep{
 kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
-
-% Quick digression: relationship to ARGs
 The term Ancestral Recombination Graph (ARG) is often understood
@@
 genomes~\citep[e.g.][]{mathieson2020ancestry},
-and in this sense a succinct tree
+and in this sense a tree
 sequence \emph{is} an ARG. More technically, an ARG is a particular
@@
 In this sense, tree sequences can be used to \emph{encode}
-ARGs without loss of information (although
-the default minimal tree sequence encoding is
-much more concise; see the \nameref{sec-arg} section).
+ARGs without loss of information (although we do not store
+the additional ARG information by default
+in \msprime\ because the minimal tree sequence encoding is
+much more concise; see the \nameref{sec-arg} section for more details).
 It should be noted that, despite efforts to
@@
 
-% OK, back to tree sequences: what about sequence variation?
-The final output of most population genetic simulations is some representation
-of sequence variation among the specified samples. For coalescent
-simulations, we usually have three steps: (1) simulate
-the genetic ancestry, and optionally output the resulting marginal trees;
-(2) simulate sequence evolution conditioned on this ancestry by generating
-mutations (see the \nameref{sec-mutations} section); and (3) output the
-resulting sequences by percolating the effects of the mutations through
-the trees. Information about the mutations
-themselves---e.g., where they have occurred on the
-trees---is usually not retained or made available for subsequent analysis.
-In \msprime, however, we skip step (3), instead using \tskit's
-data model of ancestry and mutations
-to represent the simulated sequences.
-As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
-fully integrated part of \tskit's tree sequence data model, and genetic
+The point of running simulations is to generate data, and there
+are two important ways in which the tree sequences output by
+\msprime\ are more useful than classical approaches: (1) tree
+sequences are more efficient; and (2) tree sequences contain
+more information. We will consider these points in turn.
+
+The classical approach to generating simulation data is
+to render the results in some text format. For example,
+\ms~\citep{hudson2002generating} outputs a set of sequences
+and can also optionally output the marginal trees along
+the genome in Newick format,
+and variants of this approach are used by many simulators.
+Text files have many advantages, but are slow to process
+at scale. With sample sizes becoming ever larger, processing
+text output can become a significant bottleneck in many analyses.
+The ability to efficiently process simulation results is
+particularly important in simulation-based inference methods
+such as Approximate Bayesian Computation
+(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
+and machine learning based
+approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
+flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
+crucial since the size and number of simulations that can be performed determines
+the depth to which one can sample from the model and parameter space.
+Equally important,
+however, is the efficiency with which the simulation results can be
+transformed into the specific input required by the inference method.
+In the case of ABC, this is usually a set of summary statistics of the sequence
+data, and methods avoid the bottleneck of parsing
+text-based file formats to compute these statistics
+by either developing their own
+simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
+or creating forked versions of existing
+simulators~\cite[e.g.][]{thornton2006approximate,
+hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
+tightly integrated with the inference method.
+While this is certainly more efficient,
+there is a great deal of duplication of effort and significant
+room for error (and propagation of bugs).
+Modern approaches to ABC such as
+ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
+ABC-NN~\citep{csillery2012abc,blum2010abc} use large
+numbers of weakly informative statistics,
+making the need to efficiently compute statistics from simulation
+results even more acute.
+
+Classical text based output formats like \ms\ are inefficient to process,
+but also lacking a great deal of important information.
+The tree-by-tree topology information output by simulators in Newick
+format lacks any concept of node (in the tree sequence sense) identity,
+and means that we cannot reliably infer information about ancestors
+from the output. Because Newick stores branch lengths rather
+than node times, numerical precision issues also arise when summing
+over many short branches~\citep{mcgill2013graphml}.
+% JK: not sure this is worth leaving in tbh
+This lack of reliable information about ancestors in simulation data
+has lead, for example, to the definition of segments of
+identity by descent as a set of samples having the same time
+to their most recent common ancestor along some section of the
+genome~\citep{yang2016rapidly}.
+Numerous forks
+of simulators have been created to access information not provided
+in the output. For example, \ms\  has been forked to
+output information about migrating segments~\citep{rosenzweig2016powerful},
+information about ancestral lineages~\citep{chen2013asymptotic},
+an array of population genetic statistics~\citep{ramos2007mlcoalsim}
+% Any more?
+and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
+has in turn been forked to output information on local
+ancestry~\citep{racimo2017archaic}.
+All of this information
+is either directly available by default in \msprime, or can be optionally
+stored via options such as \texttt{record\_migrations} or
+\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
+can be efficiently and conveniently processed via \tskit\ APIs.
+
+\jkcomment{Summary paragraph: tskit gives us complete information
+about the genetic ancestry. and is very efficient}
+
+% \jkcomment{A point to make somewhere in here: the tskit data model is
+% integrated, with all of the information that you need in one file. Not
+% splattered across a number of different files, that you have to keep track
+% of.}
+
+\subsection*{Genetic variation}
+\label{sec-genetic-variation}
+
+The output of population genetics simulators, quite naturally,
+tends to be some representation of the resulting nucleotide sequences
+(or some derived statistics of those sequences),
+although the underlying genealogies are also of interest.
+% Mutations are not treated as first-class data --- the sequences
+% were really all people were after, classically.
+However, joint information about the mutations that have been simulated
+-- e.g., where they have occurred on the trees --
+is usually not retained or made available for subsequent analysis.
+
+The approach taken in \msprime\ is quite different.
+As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
+integral part of \tskit's tree sequence data model, and genetic
 variation is encoded by recording sites at which mutations
 have occurred, and where each mutation at those sites
-has occurred on the marginal tree.
-Crucially, the sequences themselves are never stored, or indeed directly
-represented in memory (unless explicitly decoded using \tskit\ APIs).
+has occurred on the marginal tree;
+the sequences themselves are never stored.
 It may at first seem inconvenient to have only this indirect representation
@@
 with $t$ trees, then the \tskit\ tree sequence encoding
-requires $O(n + t + m)$ space,
-assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
-For large sample sizes, this difference is profound, making it
+requires $O(n + t + m)$ space~\citep{kelleher2016efficient},
+assuming we have $O(1)$ mutations at each site. For
+large sample sizes, this difference is profound, making it
 conceivable, for example, to store the genetic ancestry
@@
 no software library support is available.
-% \msprime, on the other hand, provides a fast, powerful, and well-tested
-% generic interface for generating and storing mutations on trees.
+\msprime, on the other hand, provides a fast, powerful, and well-tested
+generic interface for generating and storing mutations on trees.
 
@@
 variation by superimposing mutation processes on those trees
-(see Fig.~\ref{fig-mutated-trees}).
-A number of programs exist to place mutations on trees: for instance,
-the classical \SeqGen\ program~\citep{rambaut1997seq}
-supports a range of different models of sequence evolution,
-and various extensions to the basic
-models have been proposed~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
-Partly for efficiency and partly in the interest of
-simplicity for users (i.e., to avoid intermediate text format conversions),
-population genetic simulators have tended to
+(see figure~\ref{fig-mutated-trees}).
+A number of programs exist to place mutations on trees:
+for instance,
+the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
+allows a range of different models of sequence evolution to be imposed on a given
+tree in Newick format, and there are various
+extensions to the basic nucleotide
+models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
+Population genetic simulators have tended to
 include their own implementations of mutation simulation, with
-most supporting the infinite sites
+most just supporting the infinite sites
 model~\citep[e.g.][]{hudson2002generating}
-but with several supporting a wide range of different models of sequence
+but with several supporting a range of different models
+of sequence
 evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
-virgoulay2021gspace}. Thus, despite the logical separation between
-the tasks of simulating ancestry and neutral sequence evolution,
-the two have been conflated in practice.
+virgoulay2021gspace}.
 
@@
         \plrcomment{TODO: fix up the axis labels, if we like this figure?}
-        \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
     }
@@
 
-Part of the reason for this poor record of software reuse and modularity is the
-lack of standardised file formats, and in particular, the lack of common
-library infrastructure to abstract the details of interchanging simulation
-data. On the surface, it would appear that \msprime\ is also guilty of this
-lack of modularity since it supports simulating both ancestry and mutations.
-However, this is essentially a historical accident and because of the \tskit\
-library, the mutation process in \msprime\ could just as easily
-be performed in a separate package.
-The mature and efficient C and Python interfaces for
-\tskit\ makes it straightforward to add further information to
-an existing file, and because of its efficient data interchange mechanisms,
-there is no performance penalty for performing additional operations
-in a different Python package.
+% on the other hand, tskit is awesome.
+Clean separation of the ancestry and mutation generation steps
+is made possible by the \tskit\ library, whose mature and
+efficient C and Python interfaces makes it straightforward to add
+further information to an existing file, and to output the results.
+As such, it is essentially a historical accident
+that \msprime\ (whose main focus is on ancestry simulation)
+contains code to simulate mutations: the mutation process could
+be performed by a separate package, and there would be no performance
+penalty for doing so.
 Thanks to this interoperability, \msprime's
@@
 
-As well as providing a new API (see the \nameref{sec-sim-interface} section)
-which emphasises this logical split between ancestry and mutation simulation,
-we have greatly extended the sophistication of
-\msprime's mutation generation engine for version 1.0,
-achieving near feature-parity with \SeqGen.
-We  support a large number of mutation models, including the
+For version 1.0 we have greatly extended the sophistication of
+\msprime's mutation generation engine, achieving near feature-parity
+with \SeqGen.
+We  support a large number of mutation models, including
 JC69~\citep{jukes1969evolution},
@@
 
-Simulating mutations in \msprime\ is efficient.
+Generating simulations in \msprime\ is extremely efficient.
 Fig.~\ref{fig-mutations-perf} shows the time required to generate
@@
 tree ($\sim500$ times more work).
+
 % python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
-Even if we have a tree sequence consisting of a single tree
-(so that there is no advantage to working edge-by-edge),
-\msprime's mutation generator is still very efficient.
+Even if we consider a single tree, \msprime\ generates mutations
+extremely efficiently.
 For example, we simulated mutations under the BLOSUM62 amino
@@
 There are many ways in which the mutation generation code
-in \msprime\ could be extended. For example, we hope to add support for
+in \msprime\ could be extended by adapting existing methods.
+For example, we hope to add support for
 microsatellites~\citep{mailund2005coasim},
 codon models~\citep{arenas2007recodon}
-and indels~\citep{cartwright2005dna,fletcher2009indelible},
-although some may also require upstream additions to \tskit's data model
-which is currently based on the assumption of independent sites.
+and indels~\citep{cartwright2005dna,fletcher2009indelible}.
 
@@
 
-Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
+Recombination is implemented in \msprime\ using Hudson's algorithm, which
 works backwards in time tracking the
@@
 
-\subsection*{Analysis of simulation results}
-\label{sec-data-analysis}
-
-\jkcomment{This is probably part of the discussion, but putting here out of
-the way for now in its own subsection}
-
-The point of running simulations is to generate data, and there
-are two important ways in which the tree sequences output by
-\msprime\ are more useful than classical approaches: (1) tree
-sequences are more efficient; and (2) tree sequences contain
-more information. We will consider these points in turn.
-
-The classical approach to generating simulation data is
-to render the results in some text format. For example,
-\ms~\citep{hudson2002generating} outputs a set of sequences
-and can also optionally output the marginal trees along
-the genome in Newick format,
-and variants of this approach are used by many simulators.
-Text files have many advantages, but are slow to process
-at scale. With sample sizes becoming ever larger, processing
-text output can become a significant bottleneck in many analyses.
-The ability to efficiently process simulation results is
-particularly important in simulation-based inference methods
-such as Approximate Bayesian Computation
-(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
-and machine learning based
-approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
-flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
-crucial since the size and number of simulations that can be performed determines
-the depth to which one can sample from the model and parameter space.
-Equally important,
-however, is the efficiency with which the simulation results can be
-transformed into the specific input required by the inference method.
-In the case of ABC, this is usually a set of summary statistics of the sequence
-data, and methods avoid the bottleneck of parsing
-text-based file formats to compute these statistics
-by either developing their own
-simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
-or creating forked versions of existing
-simulators~\cite[e.g.][]{thornton2006approximate,
-hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
-tightly integrated with the inference method.
-While this is certainly more efficient,
-there is a great deal of duplication of effort and significant
-room for error (and propagation of bugs).
-Modern approaches to ABC such as
-ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
-ABC-NN~\citep{csillery2012abc,blum2010abc} use large
-numbers of weakly informative statistics,
-making the need to efficiently compute statistics from simulation
-results even more acute.
-
-Classical text based output formats like \ms\ are inefficient to process,
-but also lacking a great deal of important information.
-The tree-by-tree topology information output by simulators in Newick
-format lacks any concept of node (in the tree sequence sense) identity,
-and means that we cannot reliably infer information about ancestors
-from the output. Because Newick stores branch lengths rather
-than node times, numerical precision issues also arise when summing
-over many short branches~\citep{mcgill2013graphml}.
-% JK: not sure this is worth leaving in tbh
-This lack of reliable information about ancestors in simulation data
-has lead, for example, to the definition of segments of
-identity by descent as a set of samples having the same time
-to their most recent common ancestor along some section of the
-genome~\citep{yang2016rapidly}.
-Numerous forks
-of simulators have been created to access information not provided
-in the output. For example, \ms\  has been forked to
-output information about migrating segments~\citep{rosenzweig2016powerful},
-information about ancestral lineages~\citep{chen2013asymptotic},
-an array of population genetic statistics~\citep{ramos2007mlcoalsim}
-% Any more?
-and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
-has in turn been forked to output information on local
-ancestry~\citep{racimo2017archaic}.
-All of this information
-is either directly available by default in \msprime, or can be optionally
-stored via options such as \texttt{record\_migrations} or
-\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
-can be efficiently and conveniently processed via \tskit\ APIs.
-
-\jkcomment{Summary paragraph: tskit gives us complete information
-about the genetic ancestry. and is very efficient}
-
-% \jkcomment{A point to make somewhere in here: the tskit data model is
-% integrated, with all of the information that you need in one file. Not
-% splattered across a number of different files, that you have to keep track
-% of.}
-
 \subsection*{Development model}
@@
 severely underestimates runtime.
+
+\section*{Mutation generation}
+\label{app-mutation-algorithm}
+
+The algorithm that \msprime\ uses to simulate mutations on a tree sequence
+proceeds in two steps:
+first, mutations are ``placed'' on the tree sequence
+(i.e., sampling their locations in time, along the genome, and on the marginal tree),
+and then the ancestral and derived alleles of each mutation are generated.
+All mutation models share the code to place mutations,
+but choose alleles in different ways.
+
+First, mutations are placed on the tree sequence under an inhomgeneous Poisson model
+by applying them independently to each edge.
+If an edge spans a region $[a, b)$ of the genome
+and connected parent and child nodes with times $s < t$,
+and the mutation rate locally is $\mu$,
+then the number of mutations on the edge is Poisson with mean $\mu (t-s) (b-a)$,
+and each mutation is placed independently at a position chosen uniformly in $[a,b)$
+and a time uniformly in $[s,t)$.
+In a discrete genome, all positions are integers and so more than one mutation
+may occur at the same position on the same edge.
+Otherwise (i.e., for an infinite-sites model),
+positions are rejection sampled to obtain a unique floating-point number.
+If an edge spans a region of the genome with more than one mutation rate,
+this is done separately for each distinct region.
+The algorithm scales with the number of edges in the tree sequence.
+Since the number of edges is proportional to the number of marginal trees,
+this is equivalent to an algorithm that applies mutations separately
+to each marginal tree,
+but avoids the bookkeeping necessary to place mutations uniformly on a tree
+(as opposed to drawing locations uniformly on the box $[a,b) \times [s, t)$
+as we do here).
+
+Next, alleles are chosen for each mutation.
+If the site was not previously mutated, then a new ancestral allele is chosen for the site.
+Then, each mutation on the tree is considered in turn,
+in decreasing order of time ago,
+and a derived allele is chosen based on the parental allele
+(which may be the ancestral allele or the derived allele of a previous mutation)
+and some randomness.
+Finally, information about the mutations are recorded in the site and mutation tables
+of the tree sequence.
+
+A mutation model must, therefore, provide two things:
+a way of choosing an ancestral allele for each new variant site,
+and a way of choosing a derived allele given the parental allele at each mutation.
+Perhaps the simplest mutation model implemented in \msprime\ is the
+\texttt{InfiniteAlleles} mutation model,
+which keeps an internal counter so that the requested alleles are 
+assigned subsequent (and therefore unique) integers.
+
+Since the algorithm above applies mutations at a single rate
+independent of ancestral state,
+a model in which different alleles mutate at different rates
+must necessarily produce some silent mutations.
+To illustrate this, consider a mutation model on the binary alphabet
+$\{X, Y\}$ in which $X$ mutates to $Y$ at rate $\alpha$
+and $Y$ mutates to $X$ at rate $\beta$, with $\beta < \alpha$.
+To implement this, first place mutations at the largest total rate, which is $\alpha$.
+Then, at each site, choose an ancestral allele from the ``root distribution''
+(a natural choice would be to choose from the stationary distribution,
+$X$ with probability $\beta / (\alpha + \beta)$),
+and for each mutation, choose a derived allele as follows:
+if the parental allele is $X$, then choose $Y$;
+if the parental allele is $Y$, then choose $X$ with probability $\beta/(\alpha + \beta)$,
+and choose $Y$ otherwise.
+This produces the correct distribution by Poisson thinning:
+a Poisson process with rate $\alpha$ in which each point is discarded independently
+with probability $\beta / (\alpha + \beta)$ is equivalent to a Poisson
+process with rate $\beta$.
+All finite-state models (implented under the generic \texttt{MatrixMutationModel} class)
+work in this way: mutations are placed at the maximum mutation rate,
+and then some silent mutations will result.
+
+In previous versions of \msprime, silent mutations were disallowed,
+and we could have removed them from the output entirely.
+However, we have chosen to leave them in, so that for instance
+simulating with the \texttt{HKY} mutation model will result in silent mutations
+if not all equilibrium frequencies are the same.
+Although the presence of silent mutations may at first be surprising
+(although they don't affect calculations of statistics in any way),
+there is a good reason to leave them in:
+to allow layering of different mutation models.
+Suppose that we wanted to model the mutation process as a mixture of more than one model,
+e.g., Jukes-Cantor mutations at rate $\mu_1$, and
+HKY mutations occur at rate $\mu_2$.
+Layering multiple calls to \texttt{sim_mutations} is allowed,
+so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
+and then add more with the \texttt{HKY} model at rate $\mu_2$.
+However, there is a small statistical problem:
+suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
+but then the HKY mutations inserts another mutation in the middle,
+resulting in $A \to C \to C$.
+If neither mutation model allows silent transitions,
+then this is clearly not correct,
+i.e., it is not equivalent to a model that simultaneously applies the two models.
+(The impact is small, however, as it only affects sites with more than one mutation.)
+The solution is to make the Jukes-Cantor model \emph{state-independent}
+(also called ``parent-independent''),
+by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
+\emph{independently} of the parent (so that 1/4 of mutations will be silent).
+If so -- and, more generally, if the first mutational process put down is state-independent --
+then the result of sequentially applying the two mutation models
+is equivalent to the simultaneous model.
+To facilitate this, many mutation models have a \texttt{state\_independent} option
+that increases the number of silent mutations
+and makes the model closer to state-independent.
 

### 2021-07-26 paper.tex
few more edits from peter

--- 2021-07-26 paper.tex
@@
 severely underestimates runtime.
+
+\section*{Mutation generation}
+\label{app-mutation-algorithm}
+
+The algorithm that \msprime\ uses to simulate mutations on a tree sequence
+proceeds in two steps:
+first, mutations are ``placed'' on the tree sequence
+(i.e., sampling their locations in time, along the genome, and on the marginal tree),
+and then the ancestral and derived alleles of each mutation are generated.
+All mutation models share the code to place mutations,
+but choose alleles in different ways.
+
+First, mutations are placed on the tree sequence under an inhomgeneous Poisson model
+by applying them independently to each edge.
+If an edge spans a region $[a, b)$ of the genome
+and connected parent and child nodes with times $s < t$,
+and the mutation rate locally is $\mu$,
+then the number of mutations on the edge is Poisson with mean $\mu (t-s) (b-a)$,
+and each mutation is placed independently at a position chosen uniformly in $[a,b)$
+and a time uniformly in $[s,t)$.
+In a discrete genome, all positions are integers and so more than one mutation
+may occur at the same position on the same edge.
+Otherwise (i.e., for an infinite-sites model),
+positions are rejection sampled to obtain a unique floating-point number.
+If an edge spans a region of the genome with more than one mutation rate,
+this is done separately for each distinct region.
+The algorithm scales with the number of edges in the tree sequence.
+Since the number of edges is proportional to the number of marginal trees,
+an algorithm that applies mutations separately to each marginal tree
+would take at least as many steps,
+and would require additional bookkeeping necessary to place mutations uniformly on a tree.
+
+Next, alleles are chosen for each mutation.
+If the site was not previously mutated, then a new ancestral allele is chosen for the site,
+according to an input distribution of ancestral state allele probabilities.
+Then, each mutation on the tree is considered in turn,
+in order so that older mutations are considered first,
+and a derived allele is chosen based on the parental allele
+(which may be the ancestral allele or the derived allele of a previous mutation)
+and some randomness.
+Finally, information about the mutations are recorded in the site and mutation tables
+of the tree sequence.
+
+A mutation model must, therefore, provide two things:
+a way of choosing an ancestral allele for each new variant site,
+and a way of choosing a derived allele given the parental allele at each mutation.
+Perhaps the simplest mutation model implemented in \msprime\ is the
+\texttt{InfiniteAlleles} mutation model,
+which keeps an internal counter so that the requested alleles are 
+assigned subsequent (and therefore unique) integers.
+
+Since the algorithm above applies mutations at a single rate
+independent of ancestral state,
+a model in which different alleles mutate at different rates
+must necessarily produce some silent mutations,
+i.e., mutations in which the derived allele is equal to the parental allele.
+To illustrate this, consider a mutation model on the binary alphabet
+$\{X, Y\}$ in which $X$ mutates to $Y$ at rate $\alpha$
+and $Y$ mutates to $X$ at rate $\beta$, with $\beta < \alpha$.
+To implement this, first place mutations at the largest total rate, which is $\alpha$.
+Then, at each site, choose an ancestral allele from the ``root distribution''
+(a natural choice would be to choose from the stationary distribution,
+$X$ with probability $\beta / (\alpha + \beta)$),
+and for each mutation, choose a derived allele as follows:
+if the parental allele is $X$, then choose $Y$;
+if the parental allele is $Y$, then choose $X$ with probability $\beta/(\alpha + \beta)$,
+and choose $Y$ otherwise.
+This produces the correct distribution by Poisson thinning:
+a Poisson process with rate $\alpha$ in which each point is discarded independently
+with probability $\beta / (\alpha + \beta)$ is equivalent to a Poisson
+process with rate $\beta$.
+All finite-state models (implented under the generic \texttt{MatrixMutationModel} class)
+work in this way: mutations are placed at the maximum mutation rate,
+and then some silent mutations will result.
+
+In previous versions of \msprime, silent mutations were disallowed,
+and we could have removed them from the output entirely.
+However, we have chosen to leave them in, so that for instance
+simulating with the \texttt{HKY} mutation model will result in silent mutations
+if not all equilibrium frequencies are the same.
+Although the presence of silent mutations may at first be surprising
+(although they don't affect calculations of statistics in any way),
+there is a good reason to leave them in:
+to allow layering of different mutation models.
+Suppose that we wanted to model the mutation process as a mixture of more than one model,
+e.g., Jukes-Cantor mutations at rate $\mu_1$, and
+HKY mutations occur at rate $\mu_2$.
+Layering multiple calls to \texttt{sim_mutations} is allowed,
+so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
+and then add more with the \texttt{HKY} model at rate $\mu_2$.
+However, there is a small statistical problem:
+suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
+but then the HKY mutations inserts another mutation in the middle,
+resulting in $A \to C \to C$.
+If neither mutation model allows silent transitions,
+then this is clearly not correct,
+i.e., it is not equivalent to a model that simultaneously applies the two models.
+(The impact is small, however, as it only affects sites with more than one mutation.)
+The solution is to make the Jukes-Cantor model \emph{state-independent}
+(also called ``parent-independent''),
+by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
+\emph{independently} of the parent (so that 1/4 of mutations will be silent).
+If so -- and, more generally, if the first mutational process put down is state-independent --
+then the result of sequentially applying the two mutation models
+is equivalent to the simultaneous model.
+To facilitate this, many mutation models have a \texttt{state\_independent} option
+that increases the number of silent mutations
+and makes the model closer to state-independent.
 
@@
 
+\label{app-sweeps}
+\section*{Selective sweep trajectories}
+We have implemented the coalescent with selective sweeps as a structured 
+coalescent process (as in~\citep{braverman1995hitchhiking}). During the
+``sweep phase'' of the simulation the rate of coalescence and recombination
+is tracked separately in lineages linked to the beneficial allele or not,
+with those rates depending on the frequency of the beneficial allele in 
+the population at a given time. The allele frequency trajectory of the
+sweeping allele is modeled using a jump process that approximates
+the conditional diffusion of the allele, conditioning on its eventual 
+fixation~\citep{coop2004ancestral}. 
+The jump process moves back in time following the beneficial
+allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
+the origination of the allele at $p=1/2N$, and we track time in small
+increments $\delta t$. This works such that the allele frequency at
+the next time step back in time, call it $p^\prime$ is
+
+\begin{eqnarray*}
+	p'=\left\{ \begin{array}{ll}
+	    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
+	    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
+	\end{array} \right.
+\end{eqnarray*}
+where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$. Further
+$\mu(p)$ is
+\begin{eqnarray*}
+	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
+\end{eqnarray*}
+
+While our current implementation only deals with genic selection,
+we plan in the future to add sweeps with arbitrary dominance. Further planned
+additions include trajectories of purely neutral alleles, as well as trajectories
+that model soft selective sweeps.
+
 \end{document}

### 2021-07-27 paper.tex
minor edits

--- 2021-07-26 paper.tex
@@
 Since the number of edges is proportional to the number of marginal trees,
-this is equivalent to an algorithm that applies mutations separately
-to each marginal tree,
-but avoids the bookkeeping necessary to place mutations uniformly on a tree
-(as opposed to drawing locations uniformly on the box $[a,b) \times [s, t)$
-as we do here).
+an algorithm that applies mutations separately to each marginal tree
+would take at least as many steps,
+and would require additional bookkeeping necessary to place mutations uniformly on a tree.
 
 Next, alleles are chosen for each mutation.
-If the site was not previously mutated, then a new ancestral allele is chosen for the site.
+If the site was not previously mutated, then a new ancestral allele is chosen for the site,
+according to an input distribution of ancestral state allele probabilities.
 Then, each mutation on the tree is considered in turn,
-in decreasing order of time ago,
+in order so that older mutations are considered first,
 and a derived allele is chosen based on the parental allele
@@
 a model in which different alleles mutate at different rates
-must necessarily produce some silent mutations.
+must necessarily produce some silent mutations,
+i.e., mutations in which the derived allele is equal to the parental allele.
 To illustrate this, consider a mutation model on the binary alphabet

### 2021-07-27 paper.tex

--- 2021-07-26 paper.tex
@@
 of an allele bound for fixation~\citep{coop2004ancestral}.
-Details of our implementation are available in the Appendix.
 This implementation allows us to further generalize sweep models to
@@
 
-\label{app-sweeps}
-\section*{Selective sweep trajectories}
-We have implemented the coalescent with selective sweeps as a structured 
-coalescent process (as in~\citep{braverman1995hitchhiking}). During the
-``sweep phase'' of the simulation the rate of coalescence and recombination
-is tracked separately in lineages linked to the beneficial allele or not,
-with those rates depending on the frequency of the beneficial allele in 
-the population at a given time. The allele frequency trajectory of the
-sweeping allele is modeled using a jump process that approximates
-the conditional diffusion of the allele, conditioning on its eventual 
-fixation~\citep{coop2004ancestral}. 
-The jump process moves back in time following the beneficial
-allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
-the origination of the allele at $p=1/2N$, and we track time in small
-increments $\delta t$. This works such that the allele frequency at
-the next time step back in time, call it $p^\prime$ is
-
-\begin{eqnarray*}
-	p'=\left\{ \begin{array}{ll}
-	    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
-	    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
-	\end{array} \right.
-\end{eqnarray*}
-where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$. Further
-$\mu(p)$ is
-\begin{eqnarray*}
-	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
-\end{eqnarray*}
-
-While our current implementation only deals with genic selection,
-we plan in the future to add sweeps with arbitrary dominance. Further planned
-additions include trajectories of purely neutral alleles, as well as trajectories
-that model soft selective sweeps.
-
 \end{document}

### 2021-07-27 paper.tex

--- 2021-07-26 paper.tex
@@
 HKY mutations occur at rate $\mu_2$.
-Layering multiple calls to \texttt{sim_mutations} is allowed,
+Layering multiple calls to \texttt{sim\_mutations} is allowed,
 so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
@@
 where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$. Further
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$). Further
 $\mu(p)$ is

### 2021-07-28 paper.tex
Some updates to first few sections.

@@
 extracted the tree sequence code that was independent of simulations
-from \msprime\ into \tskit,
-a standalone Python and C library~\citep{tskit2021tskit}.
+from \msprime\ into
+a standalone Python and C library, called \tskit~\citep{tskit2021tskit}.
 While a full discussion of tree sequences and the capabilities
@@
 to a single chromosome strand, associated with its birth time
-and other optional metadata. A node may be
+and other optional information. A node may be
 associated with an \emph{individual}, so that, for example, a diploid
@@
 In \msprime, however, we skip step (3), instead using \tskit's
-data model of ancestry and mutations
+combined data model of ancestry and mutations
 to represent the simulated sequences.
@@
 
+The software ecosystem built around \tskit\ is stable and mature,
+with simulators such as
+\fwdpy~\citep{thornton2014cpp},
+\SLiM~\citep{haller2019slim},
+\stdpopsim~\citep{adrion2020community},
+\texttt{Geonomics}~\citep{de2021geonomics}
+and
+\texttt{GSpace}~\citep{virgoulay2021gspace},
+and inference methods such as
+\texttt{tsinfer}~\citep{kelleher2019inferring},
+\texttt{tsdate}~\citep{,wohns2021unified}
+and \texttt{Relate}~\citep{speidel2019method}
+using either the Python or C
+APIs to support outputting results in tree sequence format.
+Tree sequences are stored in an efficient binary file format, and
+are fully portable across operating systems and processor
+architectures. The \tskit\ library ensures interoperability
+between programs by having strict definitions of how the
+information in each of the tables is interpreted, and
+stringent checks for the internal consistency of the data model.
+
 % The input - how do we run simulations?
@@
 or \texttt{conda} package managers.
-For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
+For example, the errors made
 in the influential simulation analysis of
 \cite{martin2017human} were only detected because the pipeline
-could be easily run and reanalysed.
+could be easily run and reanalysed~\citep{ragsdale2020lessons,martin2020erratum}.
 
@@
 \texttt{msp} program that supports specifying demographic models in
-Demes format~\citep{gower2021demes} from the command line.
+Demes format~\citep{gower2021demes} from the command line, as
+well as simulating mutations on a given tree sequence file.
 
@@
 are two important ways in which the \tskit\ tree sequences output by
-\msprime\ are more useful than established methods: (1) tree
-sequences are more efficient; and (2) tree sequences contain
-more information. We will consider these points in turn.
+\msprime\ are more useful than established methods:
+they are more efficient and they contain more information.
+We will consider these points in turn.
 
@@
 The tree-by-tree topology information output by simulators in Newick
-format lacks any concept of node (in the tree sequence sense) identity,
+format lacks any concept of node identity,
 and means that we cannot reliably infer information about ancestors
@@
 has lead, for example, to the definition of segments of
-identity by descent as a set of samples having the same time
+identity by descent as a set of samples having approximately the same time
 to their most recent common ancestor along some section of the
 genome~\citep{yang2016rapidly}.
-Numerous forks
-of simulators have been created to access information not provided
+Numerous forks (i.e., independent copies) of simulators have been created to access information not provided
 in the output. For example, \ms\  has been forked to
@@
 be it simulated using \SLiM~\citep{haller2019slim} or
-\fwdpy~\citep{thornton2014cpp}, or estimated from real data
-using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
-or \texttt{Relate}~\citep{speidel2019method}. It is a modular
+\fwdpy~\citep{thornton2014cpp}, or estimated from real
+data~\citep{kelleher2019inferring,speidel2019method,wohns2021unified}.
+It is a modular
 component intended to fit into a larger software ecosystem, and
@@
 As well as providing a new API (see the \nameref{sec-sim-interface} section)
-which emphasises this logical split between ancestry and mutation simulation,
+which emphasises the logical split between ancestry and mutation simulation,
 we have greatly extended the sophistication of
@@
 on an edge-by-edge basis
-(see Fig.~\ref{fig-ts-example} and Appendix~\ref{app-mutation-algorithm}),
+(see Fig.~\ref{fig-ts-example} and the~\nameref{app-mutation-algorithm}
+appendix),
 rather than tree-by-tree and branch-by-branch as would otherwise be required.

### 2021-07-28 paper.tex
Update development model.

@@
 \newcommand{\phastsim}[0]{\texttt{phastsim}}
+\newcommand{\discoal}[0]{\texttt{discoal}}
 
@@
 mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
-Cosi2~\citep{shlyakhter2014cosi2}, and discoal~\citep{kern2016discoal}.
+Cosi2~\citep{shlyakhter2014cosi2}, and \discoal~\citep{kern2016discoal}.
 Each of these packages has their own benefits and possible drawbacks,
@@
 \caption{\label{fig-selection-perf}Comparison of the runtime for
-\msprime\ and \texttt{discoal} for simulating selective sweeps.
+\msprime\ and \discoal\ for simulating selective sweeps.
 }
@@
 We have implemented selective sweeps in \msprime\ similarly to
-the \texttt{discoal} implementation-- sweep trajectories are
+the \discoal\ implementation-- sweep trajectories are
 generated from a jump process approximation to the conditional diffusion
@@
 model and then switch to the standard coalescent to more efficiently simulate
-the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the perfomance of
+the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the performance of
 this approach with ARGON~\citep{palamara2016argon}, a specialised DTWF
@@
 \subsection*{Development model}
-The community development process for \msprime\ is enabled by following an
-open source development practise with a strong emphasis on code quality
-and correctness. To ensure a consistent style, we require that all code
-follows the PEP8 guidelines. Unit tests for all additions are required, and
-are automatically run on each pull request on a variety of continuous
-integration testing platforms. For new simulation features, statistical tests
-must also be added, comparing the distribution of results to either analytical
-results or existing simulators.
+\texttt{Msprime} has a large number of features, encompassing
+the functionality of several more specialised simulators
+while maintaining excellent performance.
+It is developed by a geographically distributed team of volunteers under an
+open source community development model, with a strong emphasis
+on code quality and correctness. As in any large code base,
+unit tests play a key role in ensuring that new additions behave
+as expected and \msprime\ has an extensive suite.
+As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
+lines of Python, with suites of 122 C tests (7K lines of code) and
+1350 Python tests (22K lines of code). These tests are run
+automatically on different operating systems on each pull request
+(where a contributor updates the codebase), using standard Continuous
+Integration (CI) methodology. Other CI services check code formatting
+and for common errors, as well as producing reports on the level of
+test coverage, e.g., pinpointing areas where additional unit tests may
+be required.
+
+Unit tests are vital for ensuring software quality and correctness, but
+they are usually of little value in assessing the statistical properties
+of simulations. To validate the correctness of simulation output
+we maintain a suite of statistical tests (as of 1.0.0,
+217 validation tests, in 6K lines of code). These consist of running many
+replicate simulations, and checking the properties of the output
+against either analytical results of the output of other simulators
+where such results are not available.
+For example, simulations of complex demography are validated against \ms,
+selective sweeps against \discoal\, and Wright-Fisher simulations
+against forwards-time simulations in \SLiM.
+This suite of tests is run before every release, to ensure
+that statistical errors have not been introduced.
 

### 2021-07-28 paper.tex
Add forward simulations section.

@@
 
+\subsection*{Integration with forward simulators}
+A unique feature of \msprime\ is its ability to simulate
+genetic ancestries conditioned on an existing partial
+genetic ancestry. Given a tree sequence that
+is complete up until time $t$ as input
+(where marginal trees may or may not have fully coalesced),
+\msprime\ can efficiently obtain the segments of ancestral material
+present at this time, and then run the simulation backwards in time from there.
+This allows a simulated ancestry to be produced by any
+number of different processes across disjoint time slices.
+In practise this feature is used to ``complete''
+forwards-time ancestry simulations~\citep{kelleher2018efficient}
+that may have not fully coalesced. This process
+(``recapitation'') can be orders of magnitude faster than
+the standard approach of neutral burn-in; see
+\cite{haller2018tree} for more details and examples.
+This interoperability between simulators, where a partial ancestry
+simulation produced by \SLiM~\citep{haller2019slim}
+or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
+by another simulator, with complete information
+retained---at scale---is unprecedented.
+
 \subsection*{Development model}
@@
 which break this pattern of monolithic simulation software.
-
-Tskit has facilited unprecedented interoperability between simulators:
-Forward simulations can also use tree
-sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
-3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
-tree sequences. Msprime can take tree sequences simulated by these forward
-simulators as input and complete the simulation (``recapitation'').
-See \cite{haller2018tree} for more details.
 

### 2021-07-28 paper.tex
Paragraph on inference methods.

@@
 
-There seems to be little point in continuing the huge duplication of effort we
-have seen. It is not possible to achieve the performance levels of \msprime\
-without investing substantially in software engineering. As demonstrated here,
-\msprime\ is extensible and we would argue that developing this simulator as an
-open-source community asset is of more benefit to all. Implementing a
-coalescent simulator may once have been a good way for (e.g.) a student to
-learn about programming and simulation, but it is no longer reasonable to
-expect this. There is simply too much engineering required. For example, as of
-the 1.0.0 release \msprime\ consisted of around 13K lines of C and 11K
-lines of Python, with test suites of 122 C test cases (7K lines of code) and
-1350 Python test cases (22K lines of code), along with a suite of 217
-statistical validation tests (6K lines of code). It requires a sustained effort
-over years to produce well engineered, high-quality software. It may reasonably
-be asked if students would learn more from contributing to an existing project
-with established high-quality engineering standards, or from producing their own
-software in a self-taught manner.
-
-A significant advantage to simulating the coalescent with recombination
-backwards in time using Hudson's algorithm rather than
-left-to-right
-approaches~\citep{wiuf1999recombination,chen2009fast,staab2015scrm}
-is the ease with which additional processes can be integrated.
-We have seen that processes such as gene conversion, complex
-demography, instantaneous bottlenecks, multiple merger caolescents
-and selective sweeps have been integrated seamlessly into
-the simulations. Incorporating any of these extensions into
-either the exact or approximate left-to-right processes would require
-complex derivations. Because we are going backwards
-in time, processes can be modelled as competing exponentials
-using Gillespie's algorithm \jkcomment{I think?},
-and therefore in principle we can
-add as many extra processes as we like.
-\jkcomment{so, backwards in time
-sims in msprime is extensible for the future.}
-
-There is also the welcome development of libraries such as
-Quetzal~\citep{becheler2019quetzal} and libsequence~\citep{thornton2014cpp}
-which break this pattern of monolithic simulation software.
-
-% Finish up with a paragraph saying that there loads of exciting
-% new things that simulation can do. We need MORE sims!!!
-The ability to infer trees at scale for the first time
-in the presence of
-recombination~\citep{harris2019database,kelleher2019inferring,
-speidel2019method,tang2019genealogy}
-% also SARGE
-opens the possibility for new applications of coalescent simulations.
-With lots of applications, e.g., [cite Osmond and Coop].
-Also \stdpopsim\ is cool~\citep{adrion2020community}
+% There seems to be little point in continuing the huge duplication of effort we
+% have seen. It is not possible to achieve the performance levels of \msprime\
+% without investing substantially in software engineering. As demonstrated here,
+% \msprime\ is extensible and we would argue that developing this simulator as an
+% open-source community asset is of more benefit to all. Implementing a
+% coalescent simulator may once have been a good way for (e.g.) a student to
+% learn about programming and simulation, but it is no longer reasonable to
+% expect this. There is simply too much engineering required. For example, as of
+% the 1.0.0 release \msprime\ consisted of around 13K lines of C and 11K
+% lines of Python, with test suites of 122 C test cases (7K lines of code) and
+% 1350 Python test cases (22K lines of code), along with a suite of 217
+% statistical validation tests (6K lines of code). It requires a sustained effort
+% over years to produce well engineered, high-quality software. It may reasonably
+% be asked if students would learn more from contributing to an existing project
+% with established high-quality engineering standards, or from producing their own
+% software in a self-taught manner.
+
+% A significant advantage to simulating the coalescent with recombination
+% backwards in time using Hudson's algorithm rather than
+% left-to-right
+% approaches~\citep{wiuf1999recombination,chen2009fast,staab2015scrm}
+% is the ease with which additional processes can be integrated.
+% We have seen that processes such as gene conversion, complex
+% demography, instantaneous bottlenecks, multiple merger caolescents
+% and selective sweeps have been integrated seamlessly into
+% the simulations. Incorporating any of these extensions into
+% either the exact or approximate left-to-right processes would require
+% complex derivations. Because we are going backwards
+% in time, processes can be modelled as competing exponentials
+% using Gillespie's algorithm \jkcomment{I think?},
+% and therefore in principle we can
+% add as many extra processes as we like.
+% \jkcomment{so, backwards in time
+% sims in msprime is extensible for the future.}
+
+Recent breakthroughs in methods to infer genetic ancestry in recombining
+organisms~\citep{kelleher2019inferring,
+speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
+have made it possible to estimate such ancestry at scale for
+the first time~\citep{harris2019database,tang2019genealogy}.
+Given such inferred ancestry, many exciting applications
+become possible. For example, \cite{osmond2021estimating}
+developed a method to estimate the location of genetic ancestors
+based on inferred trees, and other uses are sure to follow.
+Since the inferred genetic ancestry
+becomes the input for other downstream inferences, it is vitally
+important that these primary inferences are thoroughly validated,
+with the detailed properties of the inferred ancestries
+catalogued and understood.
+As we have seen, \msprime\---through the succinct tree sequence
+data structure and \tskit\ library---can provide simulated
+genetic ancestry at any scale or degree of detail required, and should
+provide a valuable source of ground-truth for evaluating
+these methods.
 

### 2021-07-28 paper.tex
Another stab at the discussion

@@
 developed over the decades~\citep{carvajal2008simulation,liu2008survey,
-arenas2012simulation,yuan2012overview,hoban2012computer}.
-%% CQ: Reading this sentence, I thought we were still talking abour simulators, that's what I proposed these changes
+arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
+peng2015genetic}.
 Two technological developments of recent years, however, have made the majority
@@
 For example, simulations of complex demography are validated against \ms,
-selective sweeps against \discoal\, and Wright-Fisher simulations
+selective sweeps against \discoal, and Wright-Fisher simulations
 against forwards-time simulations in \SLiM.
@@
 \section*{Discussion}
-% Simulation is useful, and it's really easy for a single locus.
-% This has lead to a culture of developing lots of simulators
-% rather than focusing on one
-Simulation has played a key role in the development of coalescent theory
-and its use in understanding observed patterns of genetic variation.
-The Kingman coalescent can be simulated for $n$ samples in $O(n)$ time
-and is straightforward to implement---for example, Hudson provided a fully
-functional and efficient simulator as a C source code listing
-in a paper appendix~\citep{hudson1990gene}. Because of
-this ease, the dominant approach to producing simulators for extensions
-to this basic model has been to develop them independently. This is
-often seen as a learning opportunity, since those developing extensions
-to the coalescent must have an excellent understanding of the basic
-model in order to develop extensions to the coalescent.
-The coalescent with recombination, however, is \emph{not} straightforward
-to implement, and numerous different approaches have been
-proposed over the
-years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
-mcvean2005approximating}. It has only recently become clear that it is
-possible to implement the model efficiently, and only through the use
-of sophisticated algorithms and data structures~\citep{kelleher2016efficient}.
-
-% Making lots of simulators has given us a fragmented and poor
-% quality ecosystem.
-The approach of developing ``in-house'' simulators
-has lead to a huge proliferation of different programs, with
-slightly different feature sets and often incompatible interfaces.
-It is a fragmented ecosystem with significant issues with software
-quality (see, e.g.,~\cite{yang2014critical}),
-and few simulators are actively maintained after publication.
-We have seen that \msprime\ (for larger samples at least)
-has at worst comparable, but usually far superior
-performance to more specialised methods.
-
-% JK: I want to make the point that the established idea that
-% developing your own simulation software is a good learning experience
-% is pretty questionable these days, and tends to produce a lot of
-% crappy unmaintained software. Not sure if the following paragraph
-% does the job that well, though.
-
-% There seems to be little point in continuing the huge duplication of effort we
-% have seen. It is not possible to achieve the performance levels of \msprime\
-% without investing substantially in software engineering. As demonstrated here,
-% \msprime\ is extensible and we would argue that developing this simulator as an
-% open-source community asset is of more benefit to all. Implementing a
-% coalescent simulator may once have been a good way for (e.g.) a student to
-% learn about programming and simulation, but it is no longer reasonable to
-% expect this. There is simply too much engineering required. For example, as of
-% the 1.0.0 release \msprime\ consisted of around 13K lines of C and 11K
-% lines of Python, with test suites of 122 C test cases (7K lines of code) and
-% 1350 Python test cases (22K lines of code), along with a suite of 217
-% statistical validation tests (6K lines of code). It requires a sustained effort
-% over years to produce well engineered, high-quality software. It may reasonably
-% be asked if students would learn more from contributing to an existing project
-% with established high-quality engineering standards, or from producing their own
-% software in a self-taught manner.
-
-% A significant advantage to simulating the coalescent with recombination
-% backwards in time using Hudson's algorithm rather than
-% left-to-right
-% approaches~\citep{wiuf1999recombination,chen2009fast,staab2015scrm}
-% is the ease with which additional processes can be integrated.
-% We have seen that processes such as gene conversion, complex
-% demography, instantaneous bottlenecks, multiple merger caolescents
-% and selective sweeps have been integrated seamlessly into
-% the simulations. Incorporating any of these extensions into
-% either the exact or approximate left-to-right processes would require
-% complex derivations. Because we are going backwards
-% in time, processes can be modelled as competing exponentials
-% using Gillespie's algorithm \jkcomment{I think?},
-% and therefore in principle we can
-% add as many extra processes as we like.
-% \jkcomment{so, backwards in time
-% sims in msprime is extensible for the future.}
+
+% Simulation software is historically a total shitshow
+
+Simulation is a vital tool in population genetics, but the approach to
+developing simulation software has historically been dysfunctional. A huge
+proliferation of tools have been published (the references here are not
+exhaustive) and only a small minority of these are actively developed and
+maintained today. The ecosystem is highly fragmented, with numerous different
+ways of specifying parameters and representing results, and there are
+significant software quality issues at all stages. This is unsurprising, since
+the majority of simulation software development is performed by students, who
+are themselves learning to code, often without formal training in software
+development. Although developing a simulator is certainly a learning
+opportunity, it is also an opportunity to learn poor development practices. A
+large number of unmaintained tools with mostly overlapping features represents
+a profound waste of effort, and it is worth considering what could have been
+achieved if even a small fraction of this effort had been invested in shared
+community infrastructure.
+
+% THERE IS A BETTER WAY!!!
+Scientific software is vital apparatus, and must be engineered
+to a high quality if we are to trust its results. There is a growing
+realisation across the sciences [citations] that investing in
+shared community infrastructure produces better results than a
+proliferation of individually maintained tools, allowing scientists
+to focus on their specific questions rather than software engineering.
+\texttt{Msprime} 1.0 is the result of such a community process,
+with features added by motivated users, taking advantage of the
+established development practices and infrastructure.
+% I'm going out on a limb here - do folks agree?
+Most contributors would agree that this is also a learning experience,
+allowing them to improve their software development skills in a
+welcoming community environment and being exposed to
+state-of-the-art infrastructure. We hope that users who find that
+features they require are missing will continue to contribute to
+\msprime, leading to a community project that is both high-quality
+and sustainable in the long term.
+
+\jkcomment{TODO either need another paragraph linking these up, or we just
+ drop this last one?}
 

### 2021-07-29 paper.tex
Move recombination equation into main text.

@@
 In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
-us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
-and empirical recombination maps similar to other
-simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
+us to simulate recombination hotspots and patterns
+of recombination from empirical maps.
 The implementation of recombination in \msprime\ is extensively validated
 against analytical results~\citep{hudson1983properties,kaplan1985use}
-and results from \ms.
+and \ms.
 
@@
 model~\citep{mcvean2005approximating,marjoram2006fast},
-and was originally motivated to a significant degree by the need to
+and was originally largely motivated by the need to
 simulate longer genomes than was possible using tools like \ms.
@@
 effects of the approximation are less well understood for larger
-sample sizes and several approaches have been proposed to
-allow simulations that more closely approximate the coalescent
+sample sizes and several approaches have been proposed
+to allow simulations more closely approximate the coalescent
 with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
-The SMC is an important analytical approximation
-and has allowed the development of numerous inference
-methods~\citep{
-li2011inference,
-harris2013inferring,
-sheehan2013estimating,
-schiffels2014inferring,
-carmi2014renewal,
-rasmussen2014genome,
-zheng2014bayesian,
-terhorst2017robust}.
+The SMC and SMC' models are available
+in \msprime\ 1.0. However, they are currently implemented using a
+naive rejection sampling approach, and are somewhat slower
+to simulate than the exact coalescent with recombination. These
+models are therefore currently only appropriate for studying the
+SMC approximations themselves, although we hope to
+implement them more efficiently in future versions.
 
@@
     \includegraphics{figures/ancestry-perf}
+    \includegraphics{figures/ancestry-perf-expected}
 \end{center}
@@
 
-In human-like parameter regimes, \msprime's implementation of the
+% \begin{figure}
+% \begin{center}
+% \end{center}
+% \caption{\label{fig-expected-perf}
+%     Data from Figure~\ref{fig-ancestry-perf}, compared with
+%     equation~\eqref{eqn-hsw-bound} (``expected'').
+%     Left panel shows observed times plotted against expected number
+%     of operations; right panel shows the ratio of the two, plotted against $\rho$.
+% }
+% \end{figure}
+
+In human-like parameter regimes and for large sample sizes,
+\msprime's implementation of the
 exact coalescent with recombination comprehensively outperforms
@@
 This is because Hudson's algorithms tracks recombinations not only
-in segments ancestral to the sample, but also between ancestral segments.
-Estimates of the number of such ``trapped'' recombinations grow quadratically
-with genome length \citep[Eq.~5.10]{hein2004gene}.
-Because they do not track such recombinations, efficient SMC implementations
-will be faster than \msprime\ for sufficiently high recombination rate,
-genome length, or $N_e$. %Until we decide to approximate the recombination process
-%between distant ancestral segment. Hint hint.
-
-% \jkcomment{We want to round out this discussion here by giving some
-% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
-% to Fig.~\ref{fig-ancestry-perf} to show roughly what we can expect
-% to simulate in how long. See issue 28.}
+in segments ancestral to the sample, but also between ancestral segments
+and for large $\rho$, these events will dominate.
+\citet[Eq.~5.10]{hein2004gene} derived an upper bound
+on $R_T$, the number of recombination events within trapped ancestral material
+in Hudson's algorithm,
+\begin{equation} \label{eqn-hsw-bound}
+    \mathbb{E}[R_T]
+    \le
+    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
+\end{equation}
+Therefore, since the sum on the right is well approximated by $\log{n}$,
+we can very roughly say that the time
+complexity of Hudson's algorithm is $O(\rho^2 \log^2 n)$.
 
@@
 and verifies that runtime scales quadratically with $\rho$,
-(\citet{hein2004gene} gives an equation
-including the effects of sample size --
-see Appendix~\ref{app-theoretical-runtime}.)
 and depends much less strongly on the number of samples.
@@
 
-However, what about a simulation with \emph{changing} size?
+Although \citet{hein2004gene} refer to the bound in \eqref{eqn-hsw-bound}
+as ``crude'', it predicts the scaling of run time quite well for large genomes
+Figure~\ref{fig-expected-perf} shows the observed runtime
+(from the same simulations as Figure~\ref{fig-ancestry-perf})
+plotted against equation~\eqref{eqn-hsw-bound},
+showing that for values of $\rho$ above $4 \times 10^4$ or so
+(i.e., chromosome-length simulations in populations above $10^4$),
+the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
+This implies the runtime for a large sample
+can be predicted from the runtime for a small sample.
+For smaller values of $\rho$, however, the prediction
+severely underestimates runtime.
+
+However, what about a simulation with changing population size?
 To understand how runtime depends on demography
-it helps to consider \emph{why} runtime is quadratic in $\rho$.
+it helps to consider why runtime is quadratic in $\rho$.
 At any point in time, \msprime\ must keep track of some number of lineages,
@@
 
-The SMC~\citep{mcvean2005approximating} and
-SMC'~\citep{marjoram2006fast} models are available
-in \msprime\ 1.0. However, they are currently implemented using a
-naive rejection sampling approach, and are somewhat slower
-to simulate than the exact coalescent with recombination. These
-models are therefore only appropriate for studying the SMC approximations
-themselves; an important avenue for future work is to implement
+an important avenue for future work is to implement
 efficient SMC-like models, perhaps based on the approaches used
@@
 sample sizes. The parameters used are chosen to mimic the simulation
-of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
+of \textit{E. coli} genomes. The genome length was set to 4.5 Mb, the scaled per site
 gene conversion rate was set to $N_e g = \gamma = 0.015$ and
@@
 
-\section*{Theoretical runtime}
-    \label{app-theoretical-runtime}
-
-\begin{figure}
-\begin{center}
-    \includegraphics{figures/ancestry-perf-expected}
-\end{center}
-\caption{\label{fig-expected-perf}
-    Data from Figure~\ref{fig-ancestry-perf}, compared with
-    equation~\eqref{eqn-hsw-bound} (``expected'').
-    Left panel shows observed times plotted against expected number
-    of operations; right panel shows the ratio of the two, plotted against $\rho$.
-}
-\end{figure}
-
-The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
-on the number of events involving trapped ancestral material in Hudson's algorithm.
-For $n$ samples from
-a diploid population of effective size $N_e$,
-and a genome of genetic length $r$,
-with $\rho = 4 N_e r$ the scaled recombination rate,
-and $R^T$ the number of recombination recombination events in trapped material,
-this upper bound is
-\begin{equation} \label{eqn-hsw-bound}
-    \mathbb{E}[R^T]
-    \le
-    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
-\end{equation}
-Note that the sum on the right is well-approximated by $\log(n)$.
-Although \citet{hein2004gene} refer to this bound as ``crude'',
-it turns out to predict the scaling of runtime quite well for large genomes.
-Figure~\ref{fig-expected-perf} shows the observed runtime
-(from the same simulations as Figure~\ref{fig-ancestry-perf})
-plotted against equation~\eqref{eqn-hsw-bound},
-showing that for values of $\rho$ above $4 \times 10^4$ or so
-(i.e., chromosome-length simulations in populations above $10^4$),
-the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
-This implies the runtime for a large sample
-can be predicted from the runtime for a small sample.
-For smaller values of $\rho$, however, the prediction
-severely underestimates runtime.
+% \section*{Theoretical runtime}
+%     \label{app-theoretical-runtime}
+% \begin{figure}
+% \begin{center}
+%     \includegraphics{figures/ancestry-perf-expected}
+% \end{center}
+% \caption{\label{fig-expected-perf}
+%     Data from Figure~\ref{fig-ancestry-perf}, compared with
+%     equation~\eqref{eqn-hsw-bound} (``expected'').
+%     Left panel shows observed times plotted against expected number
+%     of operations; right panel shows the ratio of the two, plotted against $\rho$.
+% }
+% \end{figure}
+
+% The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
+% on the number of events involving trapped ancestral material in Hudson's algorithm.
+% For $n$ samples from
+% a diploid population of effective size $N_e$,
+% and a genome of genetic length $r$,
+% with $\rho = 4 N_e r$ the scaled recombination rate,
+% and $R^T$ the number of recombination recombination events in trapped material,
+% this upper bound is
+% \begin{equation} \label{eqn-hsw-bound}
+%     \mathbb{E}[R^T]
+%     \le
+%     \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
+% \end{equation}
+% Note that the sum on the right is well-approximated by $\log(n)$.
+% Although \citet{hein2004gene} refer to this bound as ``crude'',
+% it turns out to predict the scaling of runtime quite well for large genomes.
+% Figure~\ref{fig-expected-perf} shows the observed runtime
+% (from the same simulations as Figure~\ref{fig-ancestry-perf})
+% plotted against equation~\eqref{eqn-hsw-bound},
+% showing that for values of $\rho$ above $4 \times 10^4$ or so
+% (i.e., chromosome-length simulations in populations above $10^4$),
+% the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
+% This implies the runtime for a large sample
+% can be predicted from the runtime for a small sample.
+% For smaller values of $\rho$, however, the prediction
+% severely underestimates runtime.
 

### 2021-07-29 evaluation/plot.py
Fit a single function to data.

@@ -3,6 +3,7 @@ import matplotlib
 import matplotlib.pyplot as plt
 import pandas as pd
 import numpy as np
+import scipy.optimize
 import click
 
 
@@ -170,9 +171,20 @@ def ancestry_perf():
     Plot the ancestry benchark.
     """
 
-    df = pd.read_csv(
-        "data/ancestry-perf.csv", sep=",",
-    )
+    df = pd.read_csv("data/ancestry-perf.csv", sep=",")
+
+    def objective(X, a, b):
+        r, n = X
+        return a * (r ** 2) * (np.log(n)**2) + b
+
+    # Constants aren't going to matter here since we're fitting, but not
+    # to worry.
+    rho = np.array(4 * df["N"] * df["L"] / 1e8)
+    n = np.array(df["num_samples"])
+    T = np.array(df["time"])
+    popt, _ = scipy.optimize.curve_fit(objective, [rho, n], T, [1, 0])
+
+    fig, axes = plt.subplots(1, 2, figsize=(6, 3))
 
     # Plot the times
     fig, axes = plt.subplots(1, 2, figsize=(6, 3))
@@ -219,11 +231,14 @@ def ancestry_perf():
                     )
 
             xx = np.linspace(0, 1.05 * max(X[:, 0]), 51)
+
+            ax.plot(xx, [objective((x,ns), *popt) for x in xx], color="red")
+
             # Note the two quadratic curves are not the same!
-            pargs = {}
-            if m == "o":
-                pargs["label"] = "quadratic"
-            ax.plot(xx, b[2] + b[0] * xx + b[1] * (xx ** 2), color="black", **pargs)
+            # pargs = {}
+            # if m == "o":
+            #     pargs["label"] = "quadratic"
+            # ax.plot(xx, b[2] + b[0] * xx + b[1] * (xx ** 2), color="black", **pargs)
             print(f"Times less than {tl}: "
                   f"{b[2]:.2f} + {b[0]} * rho + {b[1]} * rho^2")
 

### 2021-07-29 paper.tex
Text tweaks

@@
 and so recombine apart again immediately;
-it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
+it is these large numbers of rapid yet inconsequential events that
+lead to the quadratic runtime.
 \plrcomment{(this is discussed also in the ARG section; check for overlap)}
@@
 (including ``trapped'' non-ancestral material).
-It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
-and at this maximum, recombination and coalescence rates are equal,
-so many (roughly half) of these lineages have at least two distant segments of ancestral material
+It is known that the maximum number of lineages is reached
+very quickly~\citep{nelson2020accounting},
+and at this maximum, recombination and coalescence rates are equal.
+Threfore, many (roughly half) of these lineages have at
+least two distant segments of ancestral material
 separated by a long segment of genome, implying that $\ell \approx L / 4$.
 This implies that $M$ is roughly $L N_e$.
-Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
+The number of linages decreases gradually,
+on the coalescent time scale---so, over roughly $2 N_e$ generations.
 Since the total rate of events when there are the maximum number of lineages
-happens at rate roughly $L^2 N_e / 4$,
-then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
-as desired.
+happens is roughly $L^2 N_e / 4$,
+then the total number of events is proportional to $(L N_e)^2$---i.e.,
+proportional to $\rho^2$, as desired.
+
 What does this tell us about time-varying population sizes?
-The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
-in other words,
+The argument above implies that the work is spread out relatively
+evenly on the coalescent time scale: in other words,
 more events happen in the time period from 0 to $N_e/2$ generations
@@
 but the runtime is not dominated but the more recent time period.
-
-Putting this together:
-suppose that population size today is $N_1$,
+Suppose that population size today is $N_1$,
 while longer ago than $T$ generations ago it was $N_2$:
@@
 For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
-and the runtime will depend critically on $T$ -- in other words,
+and the runtime will depend critically on $T$---in other words,
 simulation will be quick in a species with a strong domestication bottleneck,
@@
 There is only a weak dependency on sample size
--- notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
+---notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
 only increases runtime by a factor of 2 or so.
-
-an important avenue for future work is to implement
-efficient SMC-like models, perhaps based on the approaches used
-in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
-simulation of organisms such as \textit{Drosophila melanogaster}.
+\jkcomment{as expected, since it's $O(\log^2{n})$?}
 

### 2021-07-30 paper.tex
Updates from Overleaf

--- 2021-07-28 paper.tex
@@
 with $\rho = 4 N_e r$ the scaled recombination rate,
-and $R^T$ the number of recombination recombination events in trapped material,
+and $R^T$ the number of recombination events in trapped material,
 this upper bound is

### 2021-07-30 paper.tex
Intermediate version for discussion.

--- 2021-07-29 paper.tex
@@
 against analytical results~\citep{hudson1983properties,kaplan1985use}
-and \ms.
+and simulations by \ms.
 
@@
 sample sizes and several approaches have been proposed
-to allow simulations more closely approximate the coalescent
+that allow simulations to more closely approximate the coalescent
 with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
@@
 \caption{\label{fig-ancestry-perf}
+    \jkcomment{This is a temporary version of the figure pending the outcome
+    of discussion in \#96}
+    \textbf{TOP}
     Approximate run time for \msprime\ simulations,
@@
     The black line is a quadratic fit separately in each panel.
+    \textbf{BOTTOM}
+    Time data from compared with
+    equation~\eqref{eqn-hsw-bound} (``expected'').
+    Left panel shows observed times plotted against expected number
+    of operations; right panel shows the ratio of the two, plotted against $\rho$.
 }
 \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% \end{center}
-% \caption{\label{fig-expected-perf}
-%     Data from Figure~\ref{fig-ancestry-perf}, compared with
-%     equation~\eqref{eqn-hsw-bound} (``expected'').
-%     Left panel shows observed times plotted against expected number
-%     of operations; right panel shows the ratio of the two, plotted against $\rho$.
-% }
-% \end{figure}
 
@@
 in segments ancestral to the sample, but also between ancestral segments
-and for large $\rho$, these events will dominate.
+(``trapped'' ancestral material). For large $\rho$, recombination events
+in trapped material will dominate.
 \citet[Eq.~5.10]{hein2004gene} derived an upper bound
@@
 Therefore, since the sum on the right is well approximated by $\log{n}$,
-we can very roughly say that the time
-complexity of Hudson's algorithm is $O(\rho^2 \log^2 n)$.
+we can conjecture that the time complexity of Hudson's
+algorithm is $O(\rho^2 \log^2 n)$.
 
@@
 
-Although \citet{hein2004gene} refer to the bound in \eqref{eqn-hsw-bound}
-as ``crude'', it predicts the scaling of run time quite well for large genomes
-Figure~\ref{fig-expected-perf} shows the observed runtime
-(from the same simulations as Figure~\ref{fig-ancestry-perf})
-plotted against equation~\eqref{eqn-hsw-bound},
-showing that for values of $\rho$ above $4 \times 10^4$ or so
-(i.e., chromosome-length simulations in populations above $10^4$),
-the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
-This implies the runtime for a large sample
-can be predicted from the runtime for a small sample.
-For smaller values of $\rho$, however, the prediction
-severely underestimates runtime.
+% Although \citet{hein2004gene} refer to the bound in \eqref{eqn-hsw-bound}
+% as ``crude'', it predicts the scaling of run time quite well for large genomes
+% Figure~\ref{fig-expected-perf} shows the observed runtime
+% (from the same simulations as Figure~\ref{fig-ancestry-perf})
+% plotted against equation~\eqref{eqn-hsw-bound},
+% showing that for values of $\rho$ above $4 \times 10^4$ or so
+% (i.e., chromosome-length simulations in populations above $10^4$),
+% the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
+% This implies the runtime for a large sample
+% can be predicted from the runtime for a small sample.
+% For smaller values of $\rho$, however, the prediction
+% severely underestimates runtime.
 
@@
 then the total number of events is proportional to $(L N_e)^2$---i.e.,
-proportional to $\rho^2$, as desired.
+proportional to $\rho^2$, in agreement with Eq.~\eqref{eqn-hsw-bound}.
 
@@
 \appendix
-
-% \section*{Theoretical runtime}
-%     \label{app-theoretical-runtime}
-% \begin{figure}
-% \begin{center}
-%     \includegraphics{figures/ancestry-perf-expected}
-% \end{center}
-% \caption{\label{fig-expected-perf}
-%     Data from Figure~\ref{fig-ancestry-perf}, compared with
-%     equation~\eqref{eqn-hsw-bound} (``expected'').
-%     Left panel shows observed times plotted against expected number
-%     of operations; right panel shows the ratio of the two, plotted against $\rho$.
-% }
-% \end{figure}
-
-% The book \citet{hein2004gene} (Equation 5.10) gives a theoretical upper bound
-% on the number of events involving trapped ancestral material in Hudson's algorithm.
-% For $n$ samples from
-% a diploid population of effective size $N_e$,
-% and a genome of genetic length $r$,
-% with $\rho = 4 N_e r$ the scaled recombination rate,
-% and $R^T$ the number of recombination recombination events in trapped material,
-% this upper bound is
-% \begin{equation} \label{eqn-hsw-bound}
-%     \mathbb{E}[R^T]
-%     \le
-%     \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
-% \end{equation}
-% Note that the sum on the right is well-approximated by $\log(n)$.
-% Although \citet{hein2004gene} refer to this bound as ``crude'',
-% it turns out to predict the scaling of runtime quite well for large genomes.
-% Figure~\ref{fig-expected-perf} shows the observed runtime
-% (from the same simulations as Figure~\ref{fig-ancestry-perf})
-% plotted against equation~\eqref{eqn-hsw-bound},
-% showing that for values of $\rho$ above $4 \times 10^4$ or so
-% (i.e., chromosome-length simulations in populations above $10^4$),
-% the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
-% This implies the runtime for a large sample
-% can be predicted from the runtime for a small sample.
-% For smaller values of $\rho$, however, the prediction
-% severely underestimates runtime.
 

### 2021-07-30 evaluation/plot.py

@@ -178,11 +178,11 @@ def ancestry_perf():
         return a * (r ** 2) * (np.log(n)**2) + b
 
     # Constants aren't going to matter here since we're fitting, but not
-    # to worry.
-    rho = np.array(4 * df["N"] * df["L"] / 1e8)
-    n = np.array(df["num_samples"])
-    T = np.array(df["time"])
-    popt, _ = scipy.optimize.curve_fit(objective, [rho, n], T, [1, 0])
+    # # to worry.
+    # rho = np.array(4 * df["N"] * df["L"] / 1e8)
+    # n = np.array(df["num_samples"])
+    # T = np.array(df["time"])
+    # popt, _ = scipy.optimize.curve_fit(objective, [rho, n], T, [1, 0])
 
     fig, axes = plt.subplots(1, 2, figsize=(6, 3))
 
@@ -232,13 +232,13 @@ def ancestry_perf():
 
             xx = np.linspace(0, 1.05 * max(X[:, 0]), 51)
 
-            ax.plot(xx, [objective((x,ns), *popt) for x in xx], color="red")
+            # ax.plot(xx, [objective((x,ns), *popt) for x in xx], color="red")
 
             # Note the two quadratic curves are not the same!
-            # pargs = {}
-            # if m == "o":
-            #     pargs["label"] = "quadratic"
-            # ax.plot(xx, b[2] + b[0] * xx + b[1] * (xx ** 2), color="black", **pargs)
+            pargs = {}
+            if m == "o":
+                pargs["label"] = "quadratic"
+            ax.plot(xx, b[2] + b[0] * xx + b[1] * (xx ** 2), color="black", **pargs)
             print(f"Times less than {tl}: "
                   f"{b[2]:.2f} + {b[0]} * rho + {b[1]} * rho^2")
 

### 2021-07-30 paper.tex

@@
 genomes, the coalescent provides a mathematical description of the
-genealogical tree relating these to each other. Because of its
-elegance and simplicity, the coalescent is now central to much of
-population genetics and
-genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
+genealogical tree relating the samples to each other.
+The coalescent has proved to be a powerful model,
+and it is now central to
+population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
 Simulation has been a vital
@@
 peng2015genetic}.
-Two technological developments of recent years, however, have made the majority
-of these simulators of little relevance to modern datasets.
+Two technological developments of recent years, however,
+pose major challenges to the majority of these methods.
 % Some citations here? Darwin tree of life maybe?
-Firstly, whole genome sequencing ---and in particular fourth-generation
+Firstly, whole genome sequencing---and in particular fourth-generation
 sequencing technologies---have made chromosome-level assemblies
@@
 were necessary to simulate whole chromosomes.
-
-The second development that has made the majority of
-present day simulators inapplicable to modern data sets is the
+The second development reducing the utility of many simulators is the
 exponential increase in sample size. Human datasets such
 UK Biobank~\citep{bycroft2018genome} and
-gnomAD~\citep{karczewski2019variation} now contain hundreds of
-thousands of genomes, and there is every reason to believe that
-researchers will be soon be routinely working with
-datasets containing millions of samples~\citep{stephens2015big}.
-Classical simulators such as \ms\ and even fast approximate simulators
-such as \scrm\ simply cannot cope with this scale. Even if we
-spend months of CPU time on running simulations with millions of samples,
-the output would require terabytes of storage space to store and
-days of CPU time to parse~\citep{kelleher2016efficient}.
+gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
+thousands of genomes and many other datasets on a similar scale
+are becoming available~\citep{tanjo2021practical}.
+Classical simulators such as \ms\ and even fast approximate methods
+such as \scrm\ simply cannot cope with such a large number of samples.
 
 The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
-has greatly increased the scope of coalescent simulations.
-Through the use of efficient data structures, it is
-now straightforward to simulate millions of whole chromosomes.
+has greatly increased the scope of coalescent simulations,
+and it now straightforward to simulate millions of whole chromosomes.
 The ``succinct tree sequence'' data
@@
 wohns2021unified},
-introduced as part of \msprime\, makes it possible to store simulations
-of millions of whole chromosomes in a few gigabytes, several orders
-of magnitude smaller than the standard
-Newick~\citep{felsenstein1989phylip} or
-VCF~\citep{danecek2011variant} based formats.
-This data structure has
-also lead to major advances in forwards-time
+introduced as part of \msprime\, makes it possible to store
+such huge simulations in a few gigabytes, several orders
+of magnitude smaller than commonly used formats.
+The succinct tree sequence has also lead to major advances in forwards-time
 simulation~\citep{kelleher2018efficient,haller2018tree},
@@
 \msprime\ has gained a large number of features since its introduction,
-making a highly efficient and flexible platform for population
+making it a highly efficient and flexible platform for population
 genetic simulation.
@@
 This paper marks the release of \msprime\ 1.0.
-
-% JK: Earlier notes on this para
-% - Msprime 1.0 combines the features of many different simulators and is
-%   far more efficient than all of them. The combination of a Python API,
-%   an extremely efficient storage format, and a well-engineered and
-%   extensible codebase changes the game. Stdpopsim builds on this base
-%   to provide easy access to standard models of population genetics.
-%
 

### 2021-07-30 paper.tex
took a pass through the abstract and intro

@@
 Coalescent simulation is a key tool in population genetics and
-has been integral to coalescent theory since its beginnings.
-Because of the ease at which the basic model can be simulated,
+has been integral to theoretical developments since its beginnings.
+Because of the ease with which the basic coalescent model can be simulated,
 a large number of different simulators have been developed. However,
@@
 implementing recombination via the succinct tree sequence data
-structure. We advocate a community oriented open-source development
-process as a way to reduce duplication of effort and increase
-the quality of simulation software.
+structure. Moreover, \msprime\ is a community oriented open-source development
+project, aimed as a way to reduce duplication of effort and increase
+the quality of key simulation software.
 \end{abstract}
@@
 process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
-models the ancestry of a set of sampled genomes. Given some sampled
-genomes, the coalescent provides a mathematical description of the
-genealogical tree relating the samples to each other.
+models the ancestry of a set of sampled genomes. Specifically,
+the coalescent provides a mathematical description of the
+genealogical tree that relates sampled genomes to one another.
 The coalescent has proved to be a powerful model,
@@
 population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
-Simulation has been a vital
-tool in coalescent theory since its earliest days~\citep{hudson1983testing},
-and dozens of different simulation tools have been
+Among the most impactful contributions of the coalescent since its origins
+~\citep{hudson1983testing} is that it provides a path towards more efficient
+simulation of population genetic samples than is possible using traditional,
+forwards in time simulation.
+As a result of this great utility, dozens of different simulation tools have been
 developed over the decades~\citep{carvajal2008simulation,liu2008survey,
@@
 sequencing technologies---have made chromosome-level assemblies
-commonplace. Thus, regarding input data as a series of unlinked loci
+commonplace. Thus, modeling genetic variation data as a series of unlinked loci
 (a previously reasonable assumption), is no longer defensible and
@@
 were necessary to simulate whole chromosomes.
-The second development reducing the utility of many simulators is the
-exponential increase in sample size. Human datasets such
+
+%adk-- splitting this to new paragraph
+The second major challenge to existing coalescent simulators is the
+exponential increase in sample size of genetic studies. Human datasets such
 UK Biobank~\citep{bycroft2018genome} and
@@
 % This is a bit flat --- anything else to say at the end here?
-This paper marks the release of \msprime\ 1.0.
+This paper marks the release of \msprime\ 1.0, the current state-of-the-art
+coalescent simulator both with respect to efficiency as well as feature-richness.
+Here we provide an overview of the extensive abilities of \msprime\ 1.0, 
+demonstrate through numerous performance comparisons its advantages over
+current, alternative software packages, and finally lay out future plans for
+further open-source development of what we hope will become an industry
+standard piece of computational infrastructure for the genetics community.
 

### 2021-07-31 paper.tex

@@
 % This is a bit flat --- anything else to say at the end here?
-This paper marks the release of \msprime\ 1.0, the current state-of-the-art
+This paper marks the release of \msprime\ 1.0, a state-of-the-art
 coalescent simulator both with respect to efficiency as well as feature-richness.

### 2021-08-01 paper.tex
Updates from overleaf

--- 2021-07-30 paper.tex
@@
 ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
-and calculation of population genetic statistics~\citep{ralph2020efficiently}.
+and calculation of population genetic 
+statistics~\citep{kelleher2016efficient,ralph2020efficiently}.
 Through a rigorous open-source community development process,
@@
 or a mixture of both.
-Command line interfaces such as \ms\ make it very easy to run simple
+Command line interfaces make it very easy to run simple
 simulations, but as model complexity and the number of parameters specified increases
@@
 the new APIs default to discrete genome coordinates and finite sites mutations,
-resolving a major source of confusion and error. The 0.x APIs are fully
+resolving a major source of confusion and error. The previous APIs (like \texttt{simulate}) are fully
 supported and tested, and will be maintained for the foreseeable future.
 We also provide an \ms\ compatible
-command line interface to support existing workflows, and the
-\texttt{msp} program that supports specifying demographic models in
-Demes format~\citep{gower2021demes} from the command line, as
-well as simulating mutations on a given tree sequence file.
+command line interface to support existing workflows. The
+\texttt{msp} program has been extended to include new commands 
+for simulating ancestry and mutations separately. A particularly useful 
+feature is the ability to specify demographic models in
+Demes format~\citep{gower2021demes} from the command line,
+making simulating complex demography models straightforward.
 
@@
 are two important ways in which the \tskit\ tree sequences output by
-\msprime\ are more useful than established methods:
+\msprime\ are more useful than classical methods:
 they are more efficient and they contain more information.
@@
 simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
-or creating forked versions of existing
+or creating forked versions (i.e., independent copies) of existing
 simulators~\cite[e.g.][]{thornton2006approximate,
@@
 genome~\citep{yang2016rapidly}.
-Numerous forks (i.e., independent copies) of simulators have been created to access information not provided
+Numerous forks  of simulators have been created to access information not provided
 in the output. For example, \ms\  has been forked to
@@
 
-These problems, of inefficient and limited access to the simulated
+These problems of inefficient and limited access to the simulated
 data are comprehensively solved by \msprime\ and \tskit. Complete
-information about ancestry and mutation simulations is immediately
+information about ancestry and mutation in simulations is immediately
 available to the user in a single integrated data structure
@@
 Taking a typical chromosome to be 1 Morgan in length,
-these plot show, roughly, that simulating
+these plot show, roughly, that simulating chromosome-length
 samples from a population of thousands of individuals takes seconds,
@@
 and at this maximum, recombination and coalescence rates are equal.
-Threfore, many (roughly half) of these lineages have at
+Therefore, many (roughly half) of these lineages have at
 least two distant segments of ancestral material
@@
 ---notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
-only increases runtime by a factor of 2 or so.
-\jkcomment{as expected, since it's $O(\log^2{n})$?}
+only increases runtime by a factor of 2 or so,
+as predicted from the $\log^2{n}$? factor
+of equation \eqref{eqn-hsw-bound}.
 
@@
 %description of ancestral process
-Just as recombination, gene conversion events split the ancestral material
-for some lineage, creating two independent lineages.
+Just as recombination, gene conversion events split how genetic material
+is inherited along some lineage, creating two independent lineages.
 In contrast to the previously described recombination process, where the ancestral
@@
 lineages over time, for example the coalescence rates for two or more lineages
-drawn from populations at specified times in the past.
+drawn from populations at specified times in the past,
+which can be inverted to obtain the ``inverse instanteous coalescence rate'',
+or IICR of \citet{chikhi2018inverse}.
 Many popular approaches in population genetics use the distribution of
@@
 Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
-representation of the coalescent with recombination stochastic process. This
+representation of the stochastic process that describes the coalescent with recombination. This
 formulation was complementary to Hudson's earlier

### 2021-08-01 paper.tex

@@
 % This is a bit flat --- anything else to say at the end here?
-This paper marks the release of \msprime\ 1.0.
+This paper marks the release of \msprime\ 1.0, a state-of-the-art
+coalescent simulator both with respect to efficiency as well as feature-richness.
+Here we provide an overview of the extensive abilities of \msprime\ 1.0, 
+demonstrate through numerous performance comparisons its advantages over
+current, alternative software packages, and finally lay out future plans for
+further open-source development of what we hope will become an industry
+standard piece of computational infrastructure for the genetics community.
 

### 2021-08-02 paper.tex
Update from overleaf

@@
 sequencing technologies---have made chromosome-level assemblies
-commonplace. Thus, modeling genetic variation data as a series of unlinked loci
-(a previously reasonable assumption), is no longer defensible and
-recombination must be fully accounted for. While the
-single-locus coalescent for $n$ samples can be simulated in $O(n)$
-time~\citep{hudson1990gene}, the coalescent with recombination is
+commonplace. Thus, modelling genetic variation data as a series of unlinked loci,
+a previously reasonable assumption, is no longer defensible and
+we must fully account for recombination. 
+While a genealogical tree relating $n$ samples in the single-locus coalescent 
+can be simulated in $O(n)$ time~\citep{hudson1990gene}, 
+the coalescent with recombination, which generates multiple correlated trees, is
 far more challenging.
@@
 Nodes define haplotypes that exist at a specific time within
-some individual, and edges define genetic inheritence
+some individual, and edges define genetic inheritance
 relationships between these nodes. An \emph{edge} consists of a
@@
 
-% Quick digression: relationship to ARGs
-The term Ancestral Recombination Graph (ARG) is often understood
-to mean the complete record of the genetic
-ancestry of a set of sampled
-genomes~\citep[e.g.][]{mathieson2020ancestry},
-and in this sense a succinct tree
-sequence \emph{is} an ARG. More technically, an ARG is a particular
-graphical representation of genetic ancestry based on the
+The tree sequence encoding of ancestry as set of nodes and edges 
+as just described, is equivalent to a graph in which edges 
+are annotated with a set of disjoint genomic intervals.
+This is closely related to the Ancestral Recombination Graph (ARG),
+a term that is often loosely understood
+to mean some representation of genetic
+ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
+In this sense a succinct tree sequence \emph{is} an ARG. 
+More technically, an ARG is a particular
+representation of genetic ancestry as a graph based on the
 flow of ancestral material through recombination and common
@@
 much more concise; see the \nameref{sec-arg} section).
-It should be noted that, despite efforts to
-standardise~\citep{cardona2008extended,mcgill2013graphml}, there is
+It should be noted that, despite standardisation 
+efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
 no shared format for interchanging ARG data, and no supporting software
@@
 has occurred on the marginal tree.
-Crucially, the sequences themselves are never stored, or indeed directly
+Crucially, the genetic sequences themselves are never stored, or indeed directly
 represented in memory (unless explicitly decoded using \tskit\ APIs).
@@
 and variation data for the entire human population on a
-laptop~\citep{kelleher2019inferring}. Besides the huge
-difference in storage efficiency, it is also often far more efficient
+laptop~\citep{kelleher2019inferring}. As well as the huge
+difference in storage efficiency, it is often far more efficient
 to compute statistics of the sequence data from the
 trees and mutations than it is to work with the sequences themselves.
-For example, computing Tajima's $D$ from simulated data using \tskit\
+For example, computing Tajima's $D$ from simulated data stored in the \tskit\ format
 is several orders of magnitude faster than efficient variant matrix
@@
 
-The advantages of storing genetic variation data using the underlying trees
-have recently become clear in the context of the vast genomic
-datasets produced during the SARS-CoV-2 pandemic.
+The vast genomic datasets produced during the SARS-CoV-2 pandemic
+have highlighed the advantages of storing genetic variation data using the underlying trees.
 \cite{turakhia2021ultrafast} propose the
@@
 
-The software ecosystem built around \tskit\ is stable and mature,
-with simulators such as
+The software ecosystem built around \tskit\ is stable and mature;
+simulators such as
 \fwdpy~\citep{thornton2014cpp},
@@
 and \texttt{Relate}~\citep{speidel2019method}
-using either the Python or C
+use either the Python or C
 APIs to support outputting results in tree sequence format.
@@
 Command line interfaces make it very easy to run simple
-simulations, but as model complexity and the number of parameters specified increases
+simulations, but as model complexity and the number of parameters specified increase,
 they become difficult to understand and
@@
 but lacks flexibility, for example, when running simulations with parameters
-drawn from some distribution. In practice, for any significant simulation
-project (or at least, any reproducible one), users will write a script
+drawn from some distribution. In practice, for any reproducible simulation
+project users will write a script
 to generate the required command lines or input parameter files,
@@
 stable Python
-API, which has allowed a number of downstream tools to be built on top of
+API, which has encouraged the development of an ecosystem of downstream 
 % Found these through the Dependent packages on GitHub
-it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
+tools~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
 adrion2020community,adrion2020predicting, kamm2020efficiently,
@@
 be encapsulated in a single script, and package and version
-dependencies can be explicitly stated using the \texttt{pip}
+dependencies explicitly stated using the \texttt{pip}
 or \texttt{conda} package managers.
@@
 the new APIs default to discrete genome coordinates and finite sites mutations,
-resolving a major source of confusion and error. The previous APIs (like \texttt{simulate}) are fully
+making the default settings more realistic and resolving a major source of confusion and error. 
+The previous APIs are fully
 supported and tested, and will be maintained for the foreseeable future.
@@
 This lack of reliable information about ancestors in simulation data
-has lead, for example, to the definition of segments of
+has led, for example, to the definition of segments of
 identity by descent as a set of samples having approximately the same time
@@
 Because coalescent simulations are usually concerned with
-neutral evolution (see the \nameref{sec-selection} section, however)
+neutral evolution (however, see the \nameref{sec-selection} section)
 the problem of generating synthetic genetic variation can be decomposed into
 two independent steps:
-simulating genetic ancestry (the trees), and then simulating
+firstly, simulating genetic ancestry (the trees), then subsequently simulating
 variation by superimposing mutation processes on those trees
@@
 Part of the reason for this poor record of software reuse and modularity is the
-lack of standardised file formats, and in particular, the lack of common
+lack of standardised file formats, and in particular, the absence of common
 library infrastructure to abstract the details of interchanging simulation
@@
 any number of alleles (which can be arbitrary unicode strings),
-and so other models such as the Kimura two and three
-parameter models~\citep{kimura1980simple,kimura1981estimation}
+so that other models, such as the Kimura two and three
+parameter models~\citep{kimura1980simple,kimura1981estimation},
 can be easily and efficiently defined in user code.
@@
 In the above example from Fig.~\ref{fig-mutations-perf}B,
-we would have to consider 273,887,838 branches
-since there are 137,081 trees and 1,998 branches in each
-tree ($\sim500$ times more work).
+if we generated mutations tree-by-tree, we would have to iterate over 273,887,838 branches
+(since there are 137,081 trees and 1,998 branches in each
+tree) rather than 508,125 edges, resulting in $\sim500$ times more work.
 % python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
 Even if we have a tree sequence consisting of a single tree
-(so that there is no advantage to working edge-by-edge),
+(negating the advantage of working edge-by-edge),
 \msprime's mutation generator is still very efficient.
@@
 We do not attempt a systematic benchmarking of \msprime's
-mutation generation code against other methods because at this scale it is
+mutation generation code against other methods, because at this scale it is
 difficult to disentangle the effects of inefficient input and
 output formats from the mutation generation algorithms.
-It seems quite unlikely, however,
+Given these timings, it seems quite unlikely
 that generating mutations with \msprime\ would be a bottleneck in any
@@
 There are many ways in which the mutation generation code
-in \msprime\ could be extended. For example, we hope to add support for
+in \msprime\ could be extended. For example, we intend to add support for
 microsatellites~\citep{mailund2005coasim},
@@
 Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
-works backwards in time tracking the
+works backwards in time, tracking the
 effects of common ancestor and recombination
@@
 The Sequentially Markovian Coalescent (SMC) is an approximation of the
-coalescent with recombination
-model~\citep{mcvean2005approximating,marjoram2006fast},
-and was originally largely motivated by the need to
+coalescent with recombination~\citep{mcvean2005approximating,marjoram2006fast},
+and was primarily motivated by the need to
 simulate longer genomes than was possible using tools like \ms.
@@
 effects of the approximation are less well understood for larger
-sample sizes and several approaches have been proposed
+sample sizes, and several approaches have been proposed
 that allow simulations to more closely approximate the coalescent
@@
 models are therefore currently only appropriate for studying the
-SMC approximations themselves, although we hope to
+SMC approximations themselves, although we intend to
 implement them more efficiently in future versions.
@@
 in segments ancestral to the sample, but also between ancestral segments
-(``trapped'' ancestral material). For large $\rho$, recombination events
+(i.e. in ``trapped'' ancestral material). For large $\rho$, recombination events
 in trapped material will dominate.
@@
 
-However, what about a simulation with changing population size?
+What about simulations with changing population size?
 To understand how runtime depends on demography
@@
 than from $N_e/2$ to $N_e$ generations,
-but the runtime is not dominated but the more recent time period.
+but the runtime is not dominated by the more recent time period.
 Suppose that population size today is $N_1$,
@@
 %introduction
-Gene conversion is a unidirectional transfer of a short segment of genetic material,
+Gene conversion is a form of recombination that results in unidirectional transfer
+of a short segment of genetic material,
 for example between homologous chromosomes~\citep{chen2007gene}.
 %description of ancestral process
-Just as recombination, gene conversion events split how genetic material
+Just as in crossover recombination, gene conversion events split how genetic material
 is inherited along some lineage, creating two independent lineages.
@@
 However, gene conversions impact much shorter segments than crossover recombinations,
-typically below 1kb, and thus affect linkage patterns differently than
+typically below 1kb, and thus affect linkage patterns differently to
 recombinations~\citep{korunes2017gene}.
@@
 the SMC based approach used in FastSimBac~\citep{demaio2017the},
-and forward in time tools within the SLiM framework~\citep{cury2020simulation}.
+and forward-in-time tools within the SLiM framework~\citep{cury2020simulation}.
 
@@
 \cite{wiuf2000coalescent}, where gene conversion events are initiated between any
-two sites in the sequence at a constant (scaled) rate $\gamma$ and
+two sites in the sequence at a constant rate $\gamma$ and
 the length of the transferred segment is given by a geometric distributed
@@
 \ms\ and analytical results~\citep{wiuf2000coalescent}.
+
+% In terms of file size, as expected from the discussion above, we observe that gene
+% conversion is equivalent to double the rate of
+% crossover recombination, so that a tree sequence produced with a `recombination_rate` of $r$
+% and a `gene_conversion_rate` of $g$ is roughly the same size as a tree sequence with no gene 
+% conversion and a `recombination_rate` of $r+2g$. However, because gene conversion tends to
+% affect only short regions of the genome, each gene conversion event usually creates
+% less trapped material, hence simulations using mammalian-like gene conversion parameters
+% run faster than simulations in which an equivalent amount of crossover recombination is
+% imposed.
 
@@
 A major change for \msprime\ 1.0 is the introduction of the new Demography API,
-designed to address a serious flaw in the \msprime\ 0.x interface which lead to
+designed to address a design flaw in the \msprime\ 0.x interface which lead to
 a number of avoidable errors in downstream
@@
 drawn from populations at specified times in the past,
-which can be inverted to obtain the ``inverse instanteous coalescence rate'',
+which can be inverted to obtain the ``inverse instantaneous coalescence rate'',
 or IICR of \citet{chikhi2018inverse}.
@@
 We have added two multiple merger coalescent models in \msprime\ 1.0, the
-Beta-coalescent and ``dirac''-coalescent, allowing us to simulate
+Beta-coalescent and ``Dirac''-coalescent, allowing us to simulate
 such models with recombination efficiently for the first time.
@@
 The Ancestral Recombination Graph (ARG) was introduced by
-Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
-representation of the stochastic process that describes the coalescent with recombination. This
-formulation was complementary to Hudson's earlier
+Griffiths~\citep{griffiths1991two,griffiths1997ancestral} 
+representation of the coalescent with recombination in terms of a graph. 
+This formulation was complementary to Hudson's earlier
 work~\citep{hudson1983properties}, and substantially increased our theoretical
-understanding of recombination. In Griffiths' ARG formulation, we
-regard a realisation of the coalescent with recombination as a graph in which
+understanding of recombination. In Griffiths' ARG formulation, 
+a realisation of the coalescent with recombination is a graph in which
 vertices represent common ancestor or recombination events, and edges represent
@@
 is rarely what is meant when the term is currently used. The term has come to
-essentially mean any representation of genetic ancestry in the presence
-of recombination as a graph-like structure; for example,
+essentially mean any graph-like representation of genetic ancestry in the presence
+of recombination; for example,
 \cite{mathieson2020ancestry} informally define an ARG as the subset of an
@@
 
-Although the tree sequences output by \msprime\ contain full information
+Although by default \msprime\ outputs tree sequences that contain full information
 about the genealogical trees, their correlation structure along the genome,
@@
 
-Currently there are a small number of other simulators that will generate
-coalescent samples with sweeps including, SelSim~\citep{spencer2004selsim}
+Currently a small number of other simulators generate
+coalescent samples with sweeps, including SelSim~\citep{spencer2004selsim},
 mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
@@
 Moreover because this implements the basic infrastructure for the structured
-coalescent, adding support for, e.g.,
-inversions is straightforward~\citep{peischl2013sequential}.
+coalescent, adding support for similar situations,
+such as inversions, is straightforward~\citep{peischl2013sequential}.
 
@@
 % I'm going out on a limb here - do folks agree?
-Most contributors would agree that this is also a learning experience,
+This is a learning experience for many users,
 allowing them to improve their software development skills in a

### 2021-08-02 paper.tex

@@
 which can be inverted to obtain the ``inverse instantaneous coalescence rate'',
-or IICR of \citet{chikhi2018inverse}.
+or IICR of \citet{chikhi2018iicr}.
 Many popular approaches in population genetics use the distribution of
@@
 Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
-representation of the coalescent with recombination in terms of a graph. 
+as a representation of the coalescent with recombination in terms of a graph.
 This formulation was complementary to Hudson's earlier

### 2021-08-03 paper.tex
Updates from overleaf

@@
 some individual, and edges define genetic inheritance
-relationships between these nodes. An \emph{edge} consists of a
+between these nodes. An \emph{edge} consists of a
 parent node, a child node and the left and right coordinates
@@
 library infrastructure to abstract the details of interchanging simulation
-data. On the surface, it would appear that \msprime\ is also guilty of this
-lack of modularity since it supports simulating both ancestry and mutations.
-However, this is essentially a historical accident and because of the \tskit\
-library, the mutation process in \msprime\ could just as easily
-be performed in a separate package.
+data. Although \msprime\ also supports simulating both ancestry and mutations,
+the two aspects are functionally independent within the software.
+Both ancestry and mutation simulators are present in \msprime
+for reasons of convenience and history,
+and could be split into separate packages.
 The mature and efficient C and Python interfaces for
@@
 open source community development model, with a strong emphasis
-on code quality and correctness. As in any large code base,
+on code quality, correctness, good documentation, and inclusive development.
+As in any large code base,
 unit tests play a key role in ensuring that new additions behave
@@
 that statistical errors have not been introduced.
+
+More visibly to the end user, we also have a high standard for documentation,
+with precise, comprehensive, and cross-linked documentation
+that is automatically built from the code base
+and served through the website \url{https://tskit.dev}.
+With the goal of lowering the entry barrier to new users,
+we have invested significant effort in writing examples and introductions,
+and making common tasks discoverable.
+We also view contributions to documentation as equally important to the project
+as writing code or designing methods---what use would it be
+to write reliable, stable software if no-one used it?
 

### 2021-08-03 paper.tex
Update the recombination plot.

@@
     \includegraphics{figures/ancestry-perf}
-    \includegraphics{figures/ancestry-perf-expected}
 \end{center}
 \caption{\label{fig-ancestry-perf}
-    \jkcomment{This is a temporary version of the figure pending the outcome
-    of discussion in \#96}
-    \textbf{TOP}
     Approximate run time for \msprime\ simulations,
-    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
+    for (A) small and (B) larger simulations.
     Each point is the runtime of one simulation,
@@
     shown on the horizontal axis.
-    The black line is a quadratic fit separately in each panel.
-    \textbf{BOTTOM}
-    Time data from compared with
-    equation~\eqref{eqn-hsw-bound} (``expected'').
-    Left panel shows observed times plotted against expected number
-    of operations; right panel shows the ratio of the two, plotted against $\rho$.
+    The black lines are quadratic fits separately in each panel
+    and sample size.
 }
@@
 that is automatically built from the code base
-and served through the website \url{https://tskit.dev}.
+and served through the website \texttt{https://tskit.dev}.
 With the goal of lowering the entry barrier to new users,
@@
 
-% Simulation software is historically a total shitshow
-
-Simulation is a vital tool in population genetics, but the approach to
-developing simulation software has historically been dysfunctional. A huge
-proliferation of tools have been published (the references here are not
-exhaustive) and only a small minority of these are actively developed and
+% We can simulate so much more now
+
+This release makes \msprime\ able to produce simulations
+with a great deal more biological realism, without a loss in speed and efficiency.
+This ability will allow researchers to perform more reliable % different word here
+power analyses, more robustly test new methods,
+and carry out more reliable inferences,
+without worrying that gene conversion or backmutation might be sabatoging
+their applications to real data.
+These advances have only been possible thanks to a distributed,
+collaborative model of software development,
+and the work of many people.
+We hope the tool continues to be useful to many.
+
+% Simulation software is historically a beautiful shitshow
+
+Even though simulation has long been a vital tool in population genetics,
+such collaborative software development has historically been uncommon.
+A huge proliferation of tools have been published
+(the references here are not exhaustive)
+and only a small minority of these are actively developed and
 maintained today. The ecosystem is highly fragmented, with numerous different
@@
 are themselves learning to code, often without formal training in software
-development. Although developing a simulator is certainly a learning
-opportunity, it is also an opportunity to learn poor development practices. A
-large number of unmaintained tools with mostly overlapping features represents
-a profound waste of effort, and it is worth considering what could have been
-achieved if even a small fraction of this effort had been invested in shared
-community infrastructure.
+development.
+Although this ``Haldane's sieve'' of software has produced many good tools
+and enabled decades of research,
+it also represents a missed opportunity to invest as a community
+in shared infrastructure and mentorship in good software development practice.
 
@@
 to a high quality if we are to trust its results. There is a growing
-realisation across the sciences [citations] that investing in
+realisation across the sciences \citet[OTHER CITATIONS]{siepel2019challenges}
+that investing in
 shared community infrastructure produces better results than a
@@
 % I'm going out on a limb here - do folks agree?
-This is a learning experience for many users,
-allowing them to improve their software development skills in a
-welcoming community environment and being exposed to
-state-of-the-art infrastructure. We hope that users who find that
+% PLR- personally I have learned SO MUCH, and I'm passing it on to others
+Software development in a welcoming community,
+with mentorship by experienced developers,
+is a hugely useul experience for many users,
+The skills that contributors learn
+can lead to greatly increased productivity in subsequent
+work (e.g., through more reliable code and better debugging skills).
+We hope that users who find that
 features they require are missing will continue to contribute to
@@
 
-\jkcomment{TODO either need another paragraph linking these up, or we just
- drop this last one?}
-
+The tree sequence developed for \msprime\ allows users a view of not only
+genetic variation but also the genetic ancestry that produced that variation.
 Recent breakthroughs in methods to infer genetic ancestry in recombining
@@
 speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
-have made it possible to estimate such ancestry at scale for
+have made it possible to estimate such ancestry from real data at scale for
 the first time~\citep{harris2019database,tang2019genealogy}.
@@
 catalogued and understood.
-As we have seen, \msprime\---through the succinct tree sequence
-data structure and \tskit\ library---can provide simulated
-genetic ancestry at any scale or degree of detail required, and should
-provide a valuable source of ground-truth for evaluating
-these methods.
+\texttt{Msprime} will be an important tool for these inferences and validations,
+and in this context the ability to interoperate
+with other methods---particularly, forwards simulators---through
+the succinct tree sequence data structure and \tskit\ library
+will be vital.
 

### 2021-08-03 evaluation/plot.py

@@ -1,4 +1,3 @@
-
 import matplotlib
 import matplotlib.pyplot as plt
 import pandas as pd
@@ -173,48 +172,35 @@ def ancestry_perf():
 
     df = pd.read_csv("data/ancestry-perf.csv", sep=",")
 
-    def objective(X, a, b):
-        r, n = X
-        return a * (r ** 2) * (np.log(n)**2) + b
-
-    # Constants aren't going to matter here since we're fitting, but not
-    # # to worry.
-    # rho = np.array(4 * df["N"] * df["L"] / 1e8)
-    # n = np.array(df["num_samples"])
-    # T = np.array(df["time"])
-    # popt, _ = scipy.optimize.curve_fit(objective, [rho, n], T, [1, 0])
-
-    fig, axes = plt.subplots(1, 2, figsize=(6, 3))
-
     # Plot the times
-    fig, axes = plt.subplots(1, 2, figsize=(6, 3))
+    fig, axes = plt.subplots(1, 2, figsize=(12, 6))
     rgb = matplotlib.cm.get_cmap("Set1")(np.linspace(0.0, 1.0, len(set(df["N"]))))
     time_lims = [5, 1e12]
-    for ax, tl  in zip(axes, time_lims):
+    for ax, tl in zip(axes, time_lims):
         legend_adds = []
         for ns, m in zip((1000, 100000), ("o", "v")):
             legend_adds.append(
-                    matplotlib.lines.Line2D(
-                        [], [], color='grey', marker=m,
-                        linestyle="none",
-                        label=f"num samples={ns}"
-                    )
+                matplotlib.lines.Line2D(
+                    [],
+                    [],
+                    color="grey",
+                    marker=m,
+                    linestyle="none",
+                    label=f"num samples={ns}",
+                )
             )
-            ut = np.logical_and(
-                    df["time"] <= tl,
-                    df["num_samples"] == ns
-            )
+            ut = np.logical_and(df["time"] <= tl, df["num_samples"] == ns)
             # fit a qudratic with no intercept:
             # in R this would be lm(time ~ 0 + L : N + I((N*L)^2))
             rho = 4 * df["N"] * df["L"] / 1e8
             X = np.empty((sum(ut), 3))
-            X[:,0] = rho[ut]
-            X[:,1] = ( rho ** 2 )[ut]
-            X[:,2] = 1
[…]

### 2021-08-03 paper.tex

@@
     \includegraphics{figures/ancestry-perf}
+    \includegraphics{figures/ancestry-perf-expected}
 \end{center}
 \caption{\label{fig-ancestry-perf}
+    \jkcomment{This is a temporary version of the figure pending the outcome
+    of discussion in \#96}
+    \textbf{TOP}
     Approximate run time for \msprime\ simulations,
-    for (A) small and (B) larger simulations.
+    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
     Each point is the runtime of one simulation,
@@
     shown on the horizontal axis.
-    The black lines are quadratic fits separately in each panel
-    and sample size.
+    The black line is a quadratic fit separately in each panel.
+    \textbf{BOTTOM}
+    Time data from compared with
+    equation~\eqref{eqn-hsw-bound} (``expected'').
+    Left panel shows observed times plotted against expected number
+    of operations; right panel shows the ratio of the two, plotted against $\rho$.
 }
@@
 that is automatically built from the code base
-and served through the website \texttt{https://tskit.dev}.
+and served through the website \url{https://tskit.dev}.
 With the goal of lowering the entry barrier to new users,

### 2021-08-03 paper.tex
rewrite of runtime section

--- 2021-08-03 paper.tex
@@
 This is because Hudson's algorithms tracks recombinations not only
-in segments ancestral to the sample, but also between ancestral segments
-(i.e. in ``trapped'' ancestral material). For large $\rho$, recombination events
-in trapped material will dominate.
-\citet[Eq.~5.10]{hein2004gene} derived an upper bound
-on $R_T$, the number of recombination events within trapped ancestral material
-in Hudson's algorithm,
-\begin{equation} \label{eqn-hsw-bound}
-    \mathbb{E}[R_T]
-    \le
-    \rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2 .
-\end{equation}
-Therefore, since the sum on the right is well approximated by $\log{n}$,
-we can conjecture that the time complexity of Hudson's
+in segments ancestral to the sample, but also between ancestral segments.
+In other words, lineages may ``coalesce'' even if the pieces of ancestral material they carry
+do not overlap---representing an ancestral chromosome
+that is a genetic ancestor to each of two linages,
+but not the most recent common genetic ancestor at any location along the genome.
+When this happens, the resulting merged lineage carries ``trapped'' genetic material
+that is not ancestral to any samples;
+recombinations in the trapped material will cause the two segments of ancestry
+to split apart into separate lineages again.
+For large $\rho$, recombination events
+in this trapped material will dominate.
+\citet[Eq.~5.10]{hein2004gene} gave 
+$\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2$
+as an upper bound
+on the number of recombination events within trapped ancestral material
+in Hudson's algorithm.
+Since the sum is well approximated by $\log{n}$,
+a natural conjecture is that the time complexity of Hudson's
 algorithm is $O(\rho^2 \log^2 n)$.
+This turns out to have the correct dependence on $\rho$---although
+caution is needed; see below---but the dependence on $n$ is less accurate.
 
@@
 while samples from a population of tens of thousands takes minutes.
+However, it is important to note that the quadratic curves in the two
+panels are different---predicting the runtimes of days-long simulations
+using the timing of minutes-long runs is unlikely to be very accurate.
+Further experiments show that adding a $\log^2 n$ factor
+to predict runtimes across different sample sizes does not do well
+except possibly at very large values of $\rho$.
 
@@
 then the total number of events is proportional to $(L N_e)^2$---i.e.,
-proportional to $\rho^2$, in agreement with Eq.~\eqref{eqn-hsw-bound}.
+proportional to $\rho^2$.
 
@@
 and slow otherwise.
+
 In practice, we recommend that users estimate runtime by
@@
 and then predict runtime by fitting a quadratic curve to genome length.
-There is only a weak dependency on sample size
+In these experiments, sample sizes should be roughly
+those desired---which is not a problem in practice because of
+the weak dependency on sample size
 ---notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
-only increases runtime by a factor of 2 or so,
-as predicted from the $\log^2{n}$? factor
-of equation \eqref{eqn-hsw-bound}.
+only increases runtime by a factor of 2 or so.
 

### 2021-08-04 paper.tex
Updates from Overleaf

--- 2021-08-03 paper.tex
@@
 far more challenging.
-% Sentence is too long...
+% JK will need to revise this sentence, in light of what we'll be saying 
+% down further about needing the SMC for very large rho.
 Programs such as Hudson's classical \ms~\citep{hudson2002generating}
@@
 
-%adk-- splitting this to new paragraph
 The second major challenge to existing coalescent simulators is the
@@
 genetic simulation.
-% This is a bit flat --- anything else to say at the end here?
 This paper marks the release of \msprime\ 1.0, a state-of-the-art
@@
 simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
-or creating forked versions (i.e., independent copies) of existing
+or creating forked versions (i.e., modified copies) of existing
 simulators~\cite[e.g.][]{thornton2006approximate,
@@
 over many short branches~\citep{mcgill2013graphml}.
-% JK: not sure this is worth leaving in tbh
-This lack of reliable information about ancestors in simulation data
-has led, for example, to the definition of segments of
-identity by descent as a set of samples having approximately the same time
-to their most recent common ancestor along some section of the
-genome~\citep{yang2016rapidly}.
 Numerous forks  of simulators have been created to access information not provided
@@
 with a great deal more biological realism, without a loss in speed and efficiency.
-This ability will allow researchers to perform more reliable % different word here
-power analyses, more robustly test new methods,
+This ability will allow researchers to perform more robust
+power analyses, more reliably test new methods,
 and carry out more reliable inferences,
-without worrying that gene conversion or backmutation might be sabatoging
+without worrying that gene conversion or back mutation might be sabotaging
 their applications to real data.

### 2021-08-04 evaluation/plot.py
figure adjustment

@@ -173,11 +173,19 @@ def ancestry_perf():
     df = pd.read_csv("data/ancestry-perf.csv", sep=",")
 
     # Plot the times
-    fig, axes = plt.subplots(1, 2, figsize=(12, 6))
+    fig, axes = plt.subplots(1, 2, figsize=(6, 3))
     rgb = matplotlib.cm.get_cmap("Set1")(np.linspace(0.0, 1.0, len(set(df["N"]))))
     time_lims = [5, 1e12]
     for ax, tl in zip(axes, time_lims):
-        legend_adds = []
+        legend_adds = [
+                matplotlib.lines.Line2D(
+                    [],
+                    [],
+                    color="black",
+                    linestyle="-",
+                    label=f"quadratic",
+                )
+        ]
         for ns, m in zip((1000, 100000), ("o", "v")):
             legend_adds.append(
                 matplotlib.lines.Line2D(
@@ -186,7 +194,7 @@ def ancestry_perf():
                     color="grey",
                     marker=m,
                     linestyle="none",
-                    label=f"num samples={ns}",
+                    label=f"n={ns}",
                 )
             )
             ut = np.logical_and(df["time"] <= tl, df["num_samples"] == ns)
@@ -199,7 +207,7 @@ def ancestry_perf():
             X[:, 2] = 1
             b, _, _, _ = np.linalg.lstsq(X, df["time"][ut], rcond=None)
 
-            ax.set_xlabel("Scaled recombination rate $\\rho= 4NL/ 10^{8}$")
+            ax.set_xlabel("Scaled recombination rate $\\rho/4 = N_e L$")
             ax.set_ylabel("Time (seconds)")
             Nvals = sorted(list(set(df["N"])))
             for k, Nval in enumerate(Nvals):
@@ -207,24 +215,21 @@ def ancestry_perf():
                 if np.sum(utN) > 0:
                     sargs = {}
                     if m == "o":
-                        sargs["label"] = f"N={Nval}"
+                        sargs["label"] = f"N_e={Nval}"
                     ax.scatter(
-                        rho[utN], df["time"][utN], color=rgb[k], marker=m, **sargs
+                        rho[utN]/4, df["time"][utN], color=rgb[k], marker=m, **sargs
                     )
 
             xx = np.linspace(0, 1.05 * max(X[:, 0]), 51)
 
             # Note the two quadratic curves are not the same!
[…]

### 2021-08-04 paper.tex
Add species annotations.

@@
 The SMC is a good approximation to the
-coalescent with recombination when we have four or fewer sampled
+coalescent with recombination when we have fewer than five sampled
 genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
@@
 \begin{center}
-    \includegraphics{figures/ancestry-perf}
+    \includegraphics[width=\textwidth]{figures/ancestry-perf}
 \end{center}
 \caption{\label{fig-ancestry-perf}
-    Approximate run time for \msprime\ simulations,
-    for (A) small and (B) larger simulations.
+    Running time for \msprime\
+    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
     Each point is the runtime of one simulation,
-    for various values of effective population size ($N$),
-    genome length ($L$), and number of samples.
-    Runtime scales quadratically with the product of $N$ and $L$;
-    since recombination rate in these simulations was $10^{-8}$,
-    the scaled recombination rate here is $\rho = NL/10^8$,
-    shown on the horizontal axis.
+    for various values of effective population size ($N_e$),
+    genome length in Morgans ($L$), and number of samples ($n$).
+    Runtime scales quadratically with the product of $N_e$ and $L$,
+    shown on the horizontal axis---for instance, figure A shows that
+    1,000 samples of 1 Morgan-length chromosomes
+    from a population of $N_e=2,000$ diploids
+    would take about 2 seconds, and that the same number of
+    0.01 Morgan segments in a population of size 200,000
+    would take about the same time.
+    Since recombination rate in these simulations was $10^{-8}$,
+    $L$ is the number of base pairs divided by $10^8$.
     The black lines are quadratic fits separately in each panel
-    and sample size.
+    and sample size. Also shown are the approximate values of
+    $\rho$ for chromosome 1 in three species, using values from
+    the \stdpopsim\ catalog~\citep{adrion2020community}.
 }
 \end{figure}
+
+% Output from  python3 evaluation/plot.py ancestry-perf
+% Predicted time for DroMel chr2L with n = 1000 = 107.85106995938845 hours
+% Predicted time for DroMel chr2L with n = 100000 = -511.74273443828514 hours
+% Predicted time for DroMel chr2L with n = 1000 = 177.40184421122058 hours
+% Predicted time for DroMel chr2L with n = 100000 = 361.90415260982536 hours
 
@@
 on the number of recombination events within trapped ancestral material
-in Hudson's algorithm.
+in Hudson's algorithm with $n$ samples.
 Since the sum is well approximated by $\log{n}$,
@@
 Taking a typical chromosome to be 1 Morgan in length,
-these plot show, roughly, that simulating chromosome-length
+these plots show, roughly, that simulating chromosome-length
 samples from a population of thousands of individuals takes seconds,

### 2021-08-04 evaluation/plot-argon-genomelengths.py

@@ -1,36 +1,0 @@
-#!/usr/bin/env python3
-
-import pandas as pd
-import matplotlib.pyplot as plt
-import numpy as np
-
-plt.rcParams['text.usetex'] = True
-
-df = pd.read_csv("output/x.msprime-paper-simulate-genomelengths.csv", sep="\t",
-    usecols=['length', 'program', 'runtime'])
-
-# print(df)
-
-# Make a new dataframe with just the runtime means.
-means = df.groupby(['length', 'program']).mean().reset_index()
-sample_sizes = [0, 1, 2, 3, 4]
-sample_size_labs = [r'$10^{-2}$', r'$10^{-1}$', r'$10^{0}$', r'$10^{1}$', r'$10^{2}$']
-
-# print(means)
-
-# # Plot the means
-plt.plot(sample_sizes, np.log10(means[means['program']=='msprime-DTWF']['runtime']),
-    label='msprime-DTWF', marker='o', color='C1')
-plt.plot(sample_sizes, np.log10(means[means['program']=='msprime-hybrid']['runtime']),
-    label='msprime-hybrid', marker='o', color='C1', linestyle='dashed')
-plt.plot(sample_sizes, np.log10(means[means['program']=='argon']['runtime']),
-    label='argon', marker='o', color='C0')
-plt.legend(loc='upper left')
-plt.xlabel("Chromosome lengths (Mb)")
-plt.ylabel("Runtimes (secs)")
-plt.title("Simulation runtimes for different genome lengths")
-plt.xticks(sample_sizes, labels=sample_size_labs)
-plt.yticks([-2, -1, 0, 1, 2],
-    labels=[r'$10^{-2}$', r'$10^{1}$', r'$10^{0}$', r'$10^{1}$', r'$10^{2}$'])
-plt.grid(b=True, linewidth=0.5, linestyle='dotted')
-plt.show()

### 2021-08-04 paper.tex

@@
 \begin{center}
-    \includegraphics[width=\textwidth]{figures/ancestry-perf}
+    \includegraphics{figures/ancestry-perf}
 \end{center}
 \caption{\label{fig-ancestry-perf}
-    Running time for \msprime\
-    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
+    Approximate run time for \msprime\ simulations,
+    for (A) small and (B) larger simulations.
     Each point is the runtime of one simulation,
@@
     The black lines are quadratic fits separately in each panel
-    and sample size. Also shown are the approximate values of
-    $\rho$ for chromosome 1 in three species, using values from
-    the \stdpopsim\ catalog~\citep{adrion2020community}.
+    and sample size.
 }
 \end{figure}
-
-% Output from  python3 evaluation/plot.py ancestry-perf
-% Predicted time for DroMel chr2L with n = 1000 = 107.85106995938845 hours
-% Predicted time for DroMel chr2L with n = 100000 = -511.74273443828514 hours
-% Predicted time for DroMel chr2L with n = 1000 = 177.40184421122058 hours
-% Predicted time for DroMel chr2L with n = 100000 = 361.90415260982536 hours
 

### 2021-08-04 paper.tex
Changed from overleaf

@@
 Such
-imprecision is unfortunate, since there are any number of graphical encodings of
-genetic ancestry, and referring to the original Griffiths papers does little to
+imprecision is unfortunate, since there are many ways to encode genetic
+ancestry as a graph, and referring to the original Griffiths papers does little to
 clarify. As we see below, there are situations in which the precise information
@@
 A unique feature of \msprime\ is its ability to simulate
-genetic ancestries conditioned on an existing partial
+genetic ancestries by extending an existing partial
 genetic ancestry. Given a tree sequence that

### 2021-08-05 paper.tex
Updates from overleaf

@@
 % Cluster of Excellence "Controlling Microbes to Fight Infections", Mathematical and Computational Population Genetics, University of Tübingen, Germany
+
+% Consuelo D. Quinto-Cortés 
+%National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico
 
@@
 The second major challenge to existing coalescent simulators is the
-exponential increase in sample size of genetic studies. Human datasets such
+exponential increase in sample size of genetic studies. Human datasets like the
 UK Biobank~\citep{bycroft2018genome} and
@@
 has greatly increased the scope of coalescent simulations,
-and it now straightforward to simulate millions of whole chromosomes.
+and it is now straightforward to simulate millions of whole chromosomes.
 The ``succinct tree sequence'' data
@@
 is its usage of the ``succinct tree sequence''
- data stucture to represent simulation results.
+ data structure to represent simulation results.
 The succinct tree sequence (usually abbreviated to ``tree sequence'')
@@
 The vast genomic datasets produced during the SARS-CoV-2 pandemic
-have highlighed the advantages of storing genetic variation data using the underlying trees.
+have highlighted the advantages of storing genetic variation data using the underlying trees.
 \cite{turakhia2021ultrafast} propose the
@@
 two lineages can merge in a given event. Multiple-merger coalescent processes
-have also been shown to be relevant for modeling the effects of selection on
+have also been shown to be relevant for modelling the effects of selection on
 gene genealogies \citep{Gillespie909,DS04}.
@@
 about the genealogical trees, their correlation structure along the genome,
-and the ancestral haplotypes on which coalescences occured, some information
+and the ancestral haplotypes on which coalescences occurred, some information
 is lost in this mapping down from ARG space to the minimal tree sequence
@@
 information about ancestral haplotypes that were common ancestors but
-in which no coalescences occured, and also information about the precise time
+in which no coalescences occurred, and also information about the precise time
 and genomic location of recombination events. In most cases, such

### 2021-08-05 paper.tex
Pass through recomb section

@@
 Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
-works backwards in time, tracking the
-effects of common ancestor and recombination
-events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
-Common ancestor events combine the ancestral material of two lineages, and may
-result in coalescences in the marginal trees. Recombination events
-split the ancestral material for some lineage, creating two independent
+works backwards in time, generating common ancestor and recombination
+events and tracking their effects on segments of ancestral
+material inherited from the
+sample~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
+Common ancestor events merge the ancestral material of two lineages,
+and result in coalescences in the marginal trees when ancestral
+segments overlap.
+Recombination events split the ancestral material for some lineage
+at a given breakpoint, creating two independent
 lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
@@
 against analytical results~\citep{hudson1983properties,kaplan1985use}
-and simulations by \ms.
+and simulations by \ms, \texttt{msHot} and \SLiM.
 
@@
     for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
-    Each point is the runtime of one simulation,
+    Each point is the run time of one simulation,
     for various values of effective population size ($N_e$),
     genome length in Morgans ($L$), and number of samples ($n$).
-    Runtime scales quadratically with the product of $N_e$ and $L$,
-    shown on the horizontal axis---for instance, figure A shows that
+    Run time scales quadratically with the product of $N_e$ and $L$,
+    shown on the horizontal axis. For example, (A) shows that
     1,000 samples of 1 Morgan-length chromosomes
     from a population of $N_e=2,000$ diploids
-    would take about 2 seconds, and that the same number of
-    0.01 Morgan segments in a population of size 200,000
-    would take about the same time.
+    would take about 2 seconds, and (equivalently) that the same number of
+    0.01 Morgan segments with $N_e= 200,000$ would take the same time.
     Since recombination rate in these simulations was $10^{-8}$,
@@
     and sample size. Also shown are the approximate values of
-    $\rho$ for chromosome 1 in three species, using values from
-    the \stdpopsim\ catalog~\citep{adrion2020community}.
+    $N_e L$ for chromosome 1 in three species, using values from
+    the \stdpopsim\ catalogue~\citep{adrion2020community}.
 }
@@
 to note that although the implementation of Hudson's algorithm is
-very efficient, it is still quadratic in the scaled recombination rate
+very efficient, it is still quadratic in the population scaled recombination rate
 $\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
-This is because Hudson's algorithms tracks recombinations not only
+This is because Hudson's algorithm tracks recombinations not only
 in segments ancestral to the sample, but also between ancestral segments.
-In other words, lineages may ``coalesce'' even if the pieces of ancestral material they carry
-do not overlap---representing an ancestral chromosome
-that is a genetic ancestor to each of two linages,
-but not the most recent common genetic ancestor at any location along the genome.
-When this happens, the resulting merged lineage carries ``trapped'' genetic material
-that is not ancestral to any samples;
-recombinations in the trapped material will cause the two segments of ancestry
-to split apart into separate lineages again.
-For large $\rho$, recombination events
-in this trapped material will dominate.
+As mentioned above, common ancestor events in which the ancestral material
+of two lineages is merged only result in coalescences in the marginal trees
+if their ancestral segments overlap. If there is no overlap, the merged
+segments represent an ancestral chromosome that is a genetic ancestor
+of the two linages, but not the most recent common genetic ancestor
+at any location along the genome.
+When this happens, the merged lineage carries ``trapped'' genetic material
+that is not ancestral to any samples, but where recombinations can still occur.
+The SMC approximations work by disallowing common ancestor events
+that generate trapped material, greatly simplifying the process.
+However, this also removes subtle long-range
+correlations in the trees since there are many ways in which ancestry
+segments can merge without overlapping.
+
+For large $\rho$, recombination events in trapped ancestral material will
+dominate, and so we can use this as a proxy for the overall number
+of events in Hudson's algorithm.
 \citet[Eq.~5.10]{hein2004gene} gave
-$\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2$
+\begin{equation}\label{eqn-hsw-bound}
+\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
+\end{equation}
 as an upper bound
 on the number of recombination events within trapped ancestral material
-in Hudson's algorithm with $n$ samples.
-Since the sum is well approximated by $\log{n}$,
-a natural conjecture is that the time complexity of Hudson's
-algorithm is $O(\rho^2 \log^2 n)$.
-This turns out to have the correct dependence on $\rho$---although
-caution is needed; see below---but the dependence on $n$ is less accurate.
-
-Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
-a given simulation of a constant size population,
-and verifies that runtime scales quadratically with $\rho$,
-and depends much less strongly on the number of samples.
-Taking a typical chromosome to be 1 Morgan in length,
+in Hudson's algorithm for $n$ samples.
+% Since the sum is well approximated by $\log{n}$,
+% a natural conjecture is that the time complexity of Hudson's
+% algorithm is $O(\rho^2 \log^2 n)$.
+Fig.~\ref{fig-ancestry-perf} shows the observed run time for
+simulations with a variety of population size, chromosome
+length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
+correctly predicts the quadratic dependence on $\rho$,
+as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
+We also see that the dependence on $n$ is quite weak,
+since increasing sample size 100-fold
+only increases run time by a factor of 2 or so. However, the
+$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
+(the sum is a Harmonic number, and approximated by $\log{n}$)
+is not well supported by the data except possibly at very
+large values of $\rho$, and so it seems a tighter bound
+than Eq.~\eqref{eqn-hsw-bound} would
+be required to accurately predict simulation time for a
+given $\rho$ and $n$.
+
+Fig.~\ref{fig-ancestry-perf} is a useful yardstick, allowing us
+to predict how long simulations should take for a wide range of
+species. Taking a typical chromosome to be 1 Morgan in length,
 these plots show, roughly, that simulating chromosome-length
@@
 while samples from a population of tens of thousands takes minutes.
-However, it is important to note that the quadratic curves in the two
-panels are different---predicting the runtimes of days-long simulations
+Simulating whole chromosomes for many species is very fast,
+with 1000 samples of chromosome 1 for
+\emph{Arabidopsis thaliana} taking less than a second, and a few
+minutes for dogs and humans. However, the dependence on $\rho$
+\emph{is} quadratic, and if $\rho$ is sufficiently large simulations
+may not be feasible. For example, for
+\emph{Drosophila melanogaster} chromosome 2L, with
+$\rho\approx 3.9\times10^6$, Fig.~\ref{fig-ancestry-perf}
+predicts that simulation would require around 177 hours. For such
+large values of $\rho$  we recommend
+users consider approximate simulations. Since \msprime\ does not currently
+have efficient implementations of approximate coalescent with recombination
+models, we recommend using SMC based methods such as \scrm\,
+particularly for moderate sample sizes.
+In practice, to predict the running time of a given
+simulation in \msprime, we recommend that users
+measure run time in a series of simulations with short genome lengths and
+the desired sample size,
+% possibly varying other parameters to check for interactions,
+% JK: why? ^
+% I'm worried people will take this as meaning "run every possible
+% combination of parameters". I think the genome length should be basically
+% independent of everything else, right?
+and then predict run time by fitting a quadratic curve to genome length
+as in Fig.~\ref{fig-ancestry-perf}.
+It is important to note that the quadratic curves in the two
+panels of Fig.~\ref{fig-ancestry-perf} are different, and
+predicting the run times of days-long simulations
 using the timing of minutes-long runs is unlikely to be very accurate.
-Further experiments show that adding a $\log^2 n$ factor
-to predict runtimes across different sample sizes does not do well
-except possibly at very large values of $\rho$.
-
-% Although \citet{hein2004gene} refer to the bound in \eqref{eqn-hsw-bound}
-% as ``crude'', it predicts the scaling of run time quite well for large genomes
-% Figure~\ref{fig-expected-perf} shows the observed runtime
-% (from the same simulations as Figure~\ref{fig-ancestry-perf})
-% plotted against equation~\eqref{eqn-hsw-bound},
-% showing that for values of $\rho$ above $4 \times 10^4$ or so
-% (i.e., chromosome-length simulations in populations above $10^4$),
-% the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
-% This implies the runtime for a large sample
-% can be predicted from the runtime for a small sample.
-% For smaller values of $\rho$, however, the prediction
-% severely underestimates runtime.
 
 What about simulations with changing population size?
-To understand how runtime depends on demography
-it helps to consider why runtime is quadratic in $\rho$.
+To understand how run time depends on demography
+it helps to consider why run time is quadratic in $\rho$.
 At any point in time, \msprime\ must keep track of some number of lineages,
@@
 and so recombine apart again immediately;
-it is these large numbers of rapid yet inconsequential events that
-lead to the quadratic runtime.
-\plrcomment{(this is discussed also in the ARG section; check for overlap)}
+it is these large numbers of rapid and inconsequential events that
+lead to the quadratic run time.
 The maximum number of lineages occurs when
@@
 common ancestor and recombination events balance out;
-to get an idea of runtime we can estimate when this occurs.
-Suppose that the maximal number of lineages is $M$;
+to get an idea of run time we can estimate when this occurs.
+Suppose that the maximum number of lineages is $M$;
 at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
@@
 (including ``trapped'' non-ancestral material).
-It is known that the maximum number of lineages is reached
+Since the maximum number of lineages is reached
 very quickly~\citep{nelson2020accounting},
-and at this maximum, recombination and coalescence rates are equal.
-Therefore, many (roughly half) of these lineages have at
+many (roughly half) of these lineages have at
 least two distant segments of ancestral material
 separated by a long segment of genome, implying that $\ell \approx L / 4$.
-This implies that $M$ is roughly $L N_e$.
-The number of linages decreases gradually,
-on the coalescent time scale---so, over roughly $2 N_e$ generations.
-Since the total rate of events when there are the maximum number of lineages
-happens is roughly $L^2 N_e / 4$,
+Then, since the rates of recombination and common ancestor events
+are equal the time the maximum number of lineages is present,
+we see that $M$ is roughly equal to $L N_e$.
+The number of lineages then decreases gradually from this maximum
+on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
+Since the total rate of events when the maximum number of lineages is
+present is roughly $L^2 N_e / 4$,
 then the total number of events is proportional to $(L N_e)^2$---i.e.,
@@
 The argument above implies that the work is spread out relatively
-evenly on the coalescent time scale: in other words,
-more events happen in the time period from 0 to $N_e/2$ generations
-than from $N_e/2$ to $N_e$ generations,
-but the runtime is not dominated by the more recent time period.
+evenly on the coalescent time scale.
+% JK: I just found this confusing tbh.
+% in other words,
+% more events happen in the time period from 0 to $N_e/2$ generations
+% than from $N_e/2$ to $N_e$ generations,
+% but the run time is not dominated by the more recent time period.
 Suppose that population size today is $N_1$,
 while longer ago than $T$ generations ago it was $N_2$:
-does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
+does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
 The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
-then runtime will be similar to a population of size $N_1$;
+then run time will be similar to a population of size $N_1$;
 while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
 For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
-and the runtime will depend critically on $T$---in other words,
+and the run time will depend critically on $T$---in other words,
 simulation will be quick in a species with a strong domestication bottleneck,
 and slow otherwise.
-
-In practice, we recommend that users estimate runtime by
-measuring runtime in a series of simulations with short genome lengths,
-possibly varying other parameters to check for interactions,
-and then predict runtime by fitting a quadratic curve to genome length.
-In these experiments, sample sizes should be roughly
-those desired---which is not a problem in practice because of
-the weak dependency on sample size
----notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
-only increases runtime by a factor of 2 or so.
 
@@
 \end{center}
-\caption{\label{fig-selection-perf}Comparison of the runtime for
+\caption{\label{fig-selection-perf}Comparison of the run time for
 \msprime\ and \discoal\ for simulating selective sweeps.
@@
 
-Multiple-merger coalescents, in which a random number of ancestral
-lineages may merge at a given time in one group, and only one such
-group of lineages can merge at any given time (asynchronous multiple
-mergers), are referred to as $\Lambda$-coalescents; the rate at which  a given group of $k$ lineages out of a total of  $b$ lineages merges  is given by ($a\ge  0$ constant)
+Multiple merger coalescents, in which a random number of ancestral
+lineages may merge into a common ancestor at a given time,
+are referred to as $\Lambda$-coalescents.
+The rate at which  a given group of $k$ out of a total of  $b$ lineages merges  is
 \begin{equation}\label{lambdabk}
@@
 \end{equation}
-where $\Lambda$ is a finite measure on the  (Borel subsets of)  unit interval without an atom at zero \citep{DK99,P99,S99}.    The  total rate  is given by
-\be\label{lambdab}
- \lambda_{b} = \int_0^1 \left(1 - (1-x)^{b} - bx(1-x)^{b-1} \right)x^{-2}\Lambda(dx) + \binom{b}{2},
-\ee
-\citep{S99}  which can be useful in applications, in particular  provided the  antiderivative  can be explicitly identified.
-
-A larger class of multiple-merger coalescents involving simultaneous
-multiple mergers of distinct groups of ancestral lineages also exists
-\citep{S00}. These are commonly referred to as $\Xi$-coalescents, and
-can be shown to be the limits of ancestral processes derived from
-population models incorporating diploidy (or more general polyploidy)
-\citep{BBE13,blath2016site}, and certain models of selection \citep{DS04}.
-To describe a general $\Xi$-coalescent let $\Delta$ denote the
-infinite simplex \be\label{Delta} \Delta := \{ (x_1, \ldots ): x_1 \ge
-x_2 \ge \cdots \ge 0, \sum_{j}x_j \le 1\}; \ee for any
-$r \in \{1,2, \ldots\}$ let $k_1, \cdots, k_r \ge 2$, and
-$b = s + k_1 + \cdots + k_r$ be the total number of blocks in a given
-partition ($s = b - k_1 - \cdots - k_r$ is the number of blocks not
-participating in the mergers at the given time).  The existence of
-simultaneous multiple-merger coalescents was proved by \cite{S00}.
-There exists a finite measure $\Xi$ on $\Delta$, with
-$\Xi = \Xi_0 + a\delta_0$, the measure $\Xi_0$ has no atom at zero,
-$a >0$ is fixed, and the rate at which one sees a given (simultaneous)
-merger of ancestral lineages with merger sizes $k_1, \ldots, k_r$,
-$s = n - k_1 - \cdots - k_r$, is given by
-\begin{esplit}{xi}
-  \lambda_{n; k_1, \ldots, k_r; s}  & = \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
+where $a \geq 0$ is a constant and $\Lambda$ is a finite measure on the unit interval without an atom at zero \citep{DK99,P99,S99}.
+There is also a larger class of multiple merger coalescents involving simultaneous
+multiple mergers of distinct groups of ancestral lineages \citep{S00}.
+These are commonly referred to as $\Xi$-coalescents, and
+often arise from population models incorporating diploidy or more general polyploidy
+\citep{BBE13,blath2016site}.
+To describe a general $\Xi$-coalescent, let $\Delta$ denote the
+infinite simplex
+\begin{equation*}
+ \Delta := \Bigg\{ (x_1,x_2,  \ldots ): x_1 \ge x_2 \ge \cdots \ge 0, \sum_{j = 1}^{ \infty} x_j \le 1\Bigg\}.
+ \end{equation*}
+The rate of mergers is determined by
+$\Xi = \Xi_0 + a\delta_0$, where $a \geq 0$ is a constant, $\delta_0$ is the Dirac delta
+measure, and $\Xi_0$ is a finite measure on $\Delta$ with no atom at (0, 0, \ldots).
+For an initial number of blocks $b \geq 2$ and
+$r \in \{1,2, \ldots, b - 1 \}$, let $k_1 \geq 2, \ldots, k_r \ge 2$ be the sizes of $r$ merger events
+and $s = b - k_1 - \cdots - k_r$ be the number of blocks not participating in any merger.
+The rate of each possible set of mergers with sizes $(k_1, \ldots, k_r)$ is
+\begin{align*}
+  \lambda_{n; k_1, \ldots, k_r; s}  = {}& \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
 _{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
-  & +  a\one{r=1, k_1 = 2}.
-\end{esplit}%
-The number of such $(k_1, \ldots, k_r)$ mergers is
-\be\label{N}
+  & +  a\one{r=1, k_1 = 2},
+\end{align*}
+and the number of such $(k_1, \ldots, k_r)$ mergers is
+\begin{equation*}
     \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
-\ee
-\citep{S00},   in particular  $\mathcal{N}(b;2) = b(b-1)/2$, and one can compute the total rate of a  $(k_1, \ldots, k_r)$ merger as
-\be\label{lambdabkall}
-      \lambda(n; k_1, \ldots, k_r)         =    \mathcal{N}(b; k_1, \ldots, k_r ) \lambda_{n; k_1, \ldots, k_r; s}.
-\ee
-
-The total rate is given by, with
-$n\ge 2$ denoting the total  number of ancestral lineages,
-\be
-\label{Xilambdab}
-\lambda_{n} = \int_\Delta \left(1 - \sum_{\ell = 0}^n \sum_{i_1 \neq
-\cdots \neq i_\ell } \binom{n}{\ell} x_{i_1}\cdots x_{i_\ell}\left(1
-- \sum_{i=1}^\infty x_i\right)^{n-\ell} \right)
-\frac{1}{\sum_{j=1}^\infty x_j^2}\Xi_0(dx) + a\binom{n}{2} \ee
-\citep{S00}.  Viewing coalescent processes strictly as mathematical
+\end{equation*}
+where $\ell_j := \#\{ i \in \{ 1, \ldots, r \} : k_i = j \}$ is the number of mergers of size $j \geq 2$ \citep{S00}.
+
+Viewing coalescent processes strictly as mathematical
 objects, it is clear that the class of $\Xi$-coalescents contains
-$\Lambda$-coalescents as a specific example (i.e.\ allowing at most
-one group of lineages to merge each time), and the class of
+$\Lambda$-coalescents as a specific example in which at most
+one group of lineages can merge at each time, and the class of
 $\Lambda$-coalescents contain the Kingman-coalescent as a special
 case.  However, viewed as limits of ancestral processes derived from
-specific population models they are not nested, since one would
-obtain $\Lambda$-coalescents when deriving coalescent processes from
-haploid population models incorporating sweepstakes reproduction and
-high fecundity, and $\Xi$-coalescents for diploid populations.  One
-should therefore apply the models as appropriate, i.e.\
-$\Lambda$-coalescents to data (e.g.\ mtDNA data) inherited in a
-haploid fashion, and $\Xi$-coalescents to e.g.\ autosomal data
-inherited in diploid or polyploid fashion \citep{blath2016site}.
+specific population models they are not nested. For example, one can
+obtain $\Lambda$-coalescents from haploid population models
+incorporating sweepstakes reproduction and high fecundity,
+and $\Xi$-coalescents for the same models for diploid populations \citep{BBE13}.
+One should therefore apply the models as appropriate, i.e.\
+$\Lambda$-coalescents to haploid (e.g.\ mtDNA) data,
+and $\Xi$-coalescents to diploid or polyploid (e.g.\ autosomal) data \citep{blath2016site}.
 
 In \msprime\ we have incorporated two examples of multiple-merger coalescents.
-One is a diploid extension \citep{BBE13} of the haploid model of sweepstakes
-reproduction considered by \cite{EW06}, which is a haploid Moran model adapted
-to sweepstakes reproduction.  Let $N$ denote population size, and take $\psi
-\in (0,1]$ to be fixed.  In every generation, with probability
-$1-\varepsilon_N$ a single individual (picked uniformly at random) perishes,
-and one of the surviving individuals (sampled uniformly at random) produces one
-offspring; with probability $\varepsilon_N$ a total of $\lfloor \psi N \rfloor$
-individuals perish, and of the remaining individuals a single individual
-produces $\lfloor \psi N \rfloor -1 $ offspring.  Taking $\varepsilon_N =
-1/N^\gamma$ for some $\gamma > 0$, \cite{EW06} obtain specific examples of
-$\Lambda$-coalescents, where the $\Lambda$ measure in Eq.~\eqref{lambdabk} is a
+One is a diploid extension \citep{BBE13} of the haploid Moran model adapted
+to sweepstakes reproduction considered by \cite{EW06}.
+Let $N$ denote the population size, and take $\psi \in (0,1]$ to be fixed.
+In every generation, with probability $1-\varepsilon_N$ a single individual
+(picked uniformly at random) perishes.
+With probability $\varepsilon_N$, $\lfloor \psi N \rfloor$ individuals picked uniformly
+without replacement perish instead.
+In either case, a surviving individual picked uniformly at random produces enough offspring
+to restore the population size back to $N$.  Taking $\varepsilon_N =
+1/N^\gamma$ for some $\gamma > 0$, \cite{EW06} obtain $\Lambda$-coalescents
+for which the $\Lambda$ measure in \eqref{lambdabk} is a
 point mass at $\psi$.  The simplicity of this model does allow one to obtain
 some explicit mathematical results (see e.g.\
-\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model considered by
-\cite{EW06} has also been applied in algorithms for simulating gene genealogies
-within phylogenies \citep{zhu2015hybrid}. The specific model incorporated into
-\msprime\ is the diploid version \citep{BBE13} of the model studied by
-\cite{EW06}, which would be necessary in order to incorporate recombination.
-In the model considered by \cite{BBE13}, a single pair of diploid individuals
-contribute offspring in each generation, selfing is excluded, and each
-offspring is assigned one chromosome from each parent. There are, therefore,
-four parent chromosomes involved in each reproduction event, which can lead to
-up to four simultaneous mergers.  Let \begin{esplit}{Cconst} C_{b; k_1, \ldots,
-k_r;s } &= \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell}
+\cite{Der2012, EF2018,Freund2020,Matuszewski2017}), and the model has also been
+used to simulate gene genealogies within phylogenies \citep{zhu2015hybrid}.
+As well as the haploid model of \cite{EW06}, \msprime\ provides the diploid version
+of \cite{BBE13}, in which individuals perish as above, but replacements are
+generated by sampling a single pair of diploid individuals as parents, with
+children sampling one chromosome from each parent.
+Hence, there are four parent chromosomes involved in each reproduction event, which can lead to
+up to four simultaneous mergers, giving rise to a $\Xi$-coalescent with merger rate
+\begin{esplit}{Cconst} \lambda^{\text{Dirac}}_{b; k_1, \ldots,
+k_r;s } &= \frac{c \psi^2 / 4}{1 + c \psi^2 / 4} \frac{4}{ \psi^2}\sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell}
 (4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots +
-k_r + \ell},  \\ \end{esplit} where  $C_{b; k_1, \ldots, k_r;s }$ corresponds
-to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq~\eqref{xi})
-of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$. The
-total merger rate (see Eq~\eqref{lambdabkall}) is then
-\be\label{xidirlambdabk} \lambda(b;k_1, \ldots, k_r) = \mathcal{N}(b; k_1,
-\ldots, k_r ) \left( \frac{c\psi^2/4}{ 1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s
-} + \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right), \ee and the total
-coalescence rate becomes \begin{esplit}{xilambdab} \lambda_b & =
-\frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0
-}^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 -
-\psi)^{b-\ell } \right) + \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.  \\
-\end{esplit} The interpretation of \eqref{xilambdab} is that with probability
-$1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent
-pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 +
-c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair
-produce  $\lfloor \psi N \rfloor$ offspring.
-
-The other multiple-merger coalescent model incorporated in \msprime\ is derived
-from an adaptation of the haploid population model considered by
-\cite{schweinsberg03} to diploid populations \citep{BLS15}.  In the haploid
-version, in each generation individuals independently produce random numbers of
-juveniles, where the random  number of juveniles $(X)$ produced by a given
-individual  has the stable law \be\label{jX} \lim_{k\to \infty} C k^\alpha
-\prb{X \ge k} = 1 \ee with index $\alpha > 0$, and $C > 0$ is a normalising
-constant.   One can interpret  Eq.~\eqref{jX} as  stating the form of the
-probability distribution  for number of juveniles at least on the order of the
-population size.     If the random  total number of juveniles $(S_N)$ produced
-in this way is at least the population size $(N)$, then one samples  $N$
-juveniles uniformly at random without replacement to form the next generation
-of  reproducing individuals; if  $S_N < N$ one simply  carries on with  the
-same  set of individuals.   However,   assuming  $\EE{X} > 1$, one can show
-that $\prb{S_N < N}$ decays exponentially fast in $N$ \citep{schweinsberg03}.
+k_r + \ell} + \frac{\one{r=1, k_1 = 2}}{1 + c \psi^2 / 4},  \\ \end{esplit}
+The interpretation of \eqref{Cconst} is that `small' reproduction events in which
+two lineages merge occur at rate $1/(1 + c\psi^2/4)$, while large reproduction events
+giving rise to simultaneous multiple mergers occur at rate $(c\psi^2/4) / (1 + c\psi^2/4)$.
+
+The other multiple merger coalescent model incorporated in \msprime\ is
+the haploid population model considered by \cite{schweinsberg03},
+as well as its diploid extension \citep{BLS15}.  In the haploid
+version, in each generation of fixed size $N$, individuals produce random numbers of
+juveniles $(X_1, \ldots, X_N)$ independently, each distributed according to a stable law satisfying
+\be\label{jX} \lim_{k\to \infty} C k^\alpha \prb{X \ge k} = 1 \ee
+with index $\alpha > 0$, and where $C > 0$ is a normalising constant.
+If the  total number of juveniles $S_N := X_1 + \ldots + X_n$ produced
+in this way is at least $N$, then $N$ juveniles are sampled uniformly at random
+without replacement to form the next generation.
+As long as $\EE{X_1} > 1$, one can show that $\{ S_N < N \}$ has exponentially small
+probability in $N$, and does not affect the resulting coalescent as $N \to \infty$ \citep{schweinsberg03}.
 If $\alpha \ge 2$ the ancestral process converges to the Kingman-coalescent; if
-$1 \le \alpha < 2$ the ancestral process converges to a specific case of a
-$\Lambda$-coalescent, where the $\Lambda$ measure in Eq.~\eqref{lambdabk} is
-associated with the Beta$(2-\alpha, \alpha)$ probability distribution, i.e.\
+$1 \le \alpha < 2$ the ancestral process converges to a $\Lambda$-coalescent
+with $\Lambda$ in \eqref{lambdabk} given by the Beta$(2-\alpha, \alpha)$ distribution, i.e.\
 \be\label{Fbeta} \Lambda(dx) = \one{0< x \le 1} \frac{1}{B(2-\alpha,\alpha)}
-x^{1 - \alpha}(1-x)^{\alpha - 1}  dx, \ee where $B(2-\alpha,\alpha)$ is the
-beta function $B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$, $a,b > 0$
-\citep{schweinsberg03}.
+x^{1 - \alpha}(1-x)^{\alpha - 1}  dx, \ee where
+$B(a,b) = \Gamma(a)\Gamma(b)/\Gamma(a+b)$ for $a,b > 0$ is the
+beta function \citep{schweinsberg03}.
 This model has been adapted to diploid populations
-by \cite{BLS15}, where  the resulting coalescent process
-is a  four-fold $\Xi$-coalescent on  $(x/4, x/4, x/4, x/4, 0, 0, \ldots)$,
-where $x$ is a random variate with  the Beta$(2-\alpha,\alpha)$ distribution.
-The merger rate (see Eq.~\eqref{xi}) is then ($1 \le r \le 4$)
-\be\label{xibeta} \lambda_{b;k_1, \ldots, k_r} = \sum_{\ell = 0}^{ (b -
-k)\wedge (4-r) } \binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}}
-\frac{B(k+\ell - \alpha, b-k-\ell + \alpha ) }{B(2-\alpha,\alpha)} \ee
-\citep{blath2016site,BLS15}. The interpretation of Eq.~\eqref{xibeta} is  that the
-random   number of diploid  juveniles each  diploid pair of parents  produces
-is  governed by  the law in Eq.~\eqref{jX},   and   each diploid juvenile  is
-assigned  one chromosome  from each parent (selfing is excluded).
-
-The model in Eq~\eqref{jX} assumes that individuals can produce arbitrarily
-large numbers of juveniles. Considering diploid juveniles, this assumption is
-probably rather strong, since diploid juveniles are at least fertilised eggs,
-and so it is reasonable to suppose that the number of juveniles surviving to
-the  particular life stage we are modeling  cannot be arbitrarily large.  With
-this in mind, we also consider an adaptation of the Schweinsberg model, where
-the random number of juveniles $(X)$ produced by a   given parent pair  is
-distributed according to \be\label{jtr} \prb{X=k} =   \one{1 \le k \le \phi(N)}
+by \cite{BLS15}, and the resulting coalescent is $\Xi$-coalescent with merger rate
+\be\label{xibeta} \lambda_{b;k_1, \ldots, k_r; s}^{\text{Beta}} = \sum_{\ell = 0}^{ s\wedge (4-r) } \binom{s}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}}
+\frac{B(k+\ell - \alpha, s-\ell + \alpha ) }{B(2-\alpha,\alpha)}, \ee
+where $k := k_1 + \ldots + k_r$ \citep{blath2016site,BLS15}. The interpretation of \eqref{xibeta} is  that the
+random number of lineages participating in a potential merger is governed by the
+$\Lambda$-coalescent with rate \eqref{Fbeta}, and all participating lineages are
+randomly allocated into one of four groups corresponding to the four parental
+chromosomes, giving rise to up to four simultaneous mergers.
+
+The stable law \eqref{jX} assumes that individuals can produce arbitrarily
+large numbers of juveniles. Since juveniles are at least fertilised eggs,
+it may be desirable suppose that the number of juveniles
+surviving to reproductive maturity cannot be arbitrarily large.  Hence we also consider
+an adaptation of the \cite{schweinsberg03} model, where
+the random number of juveniles has a deterministic upper bound $\phi(N)$,
+and the distribution of the number of juveniles produced by a given parent
+(or pair of parents in the diploid case) is
+\be\label{jtr} \prb{X=k} =   \one{1 \le k \le \phi(N)}
 \frac{\phi(N+1)^\alpha }{ \phi(N+1)^\alpha - 1 }  \left( \frac{1}{k^\alpha} -
-\frac{1}{(k+1)^\alpha}  \right) , \ee where $\phi(N)$ is a deterministic strict
-upper bound on the number of juveniles produced by  any given parent pair (see
-also \citep{Eldon2018}).   One can follow the calculations in
-\citep{schweinsberg03} or \citep{BLS15}  to show  that  if $1 < \alpha < 2$
-then  $(k = k_1 + \cdots + k_r)$ then the merger rate (see Eq.~\eqref{xibeta})
-is
- \be \lambda_{b;k_1, \ldots, k_r} =  \sum_{\ell = 0}^{ (b - k)\wedge (4-r) }
-\binom{b-k}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(M; k+\ell - \alpha,
-b-k-\ell + \alpha ) }{B(M;2-\alpha,\alpha)}
+\frac{1}{(k+1)^\alpha}  \right) . \ee
+See \citet{Eldon2018} for a related model.   One can follow the calculations of
+\cite{schweinsberg03} or \cite{BLS15}  to show  that  if $1 < \alpha < 2$
+then, recalling that $k = k_1 + \cdots + k_r$, the merger rate is
+ \be \lambda_{b;k_1, \ldots, k_r; s}^{\text{Beta}, M} =  \sum_{\ell = 0}^{ s \wedge (4-r) }
+\binom{s}{\ell} \frac{ (4)_{r+\ell} }{4^{k+\ell}} \frac{B(M; k+\ell - \alpha,
+s-\ell + \alpha ) }{B(M;2-\alpha,\alpha)}
 \ee
-with $B(z;a,b) := \int_0^z
-t^{a-1}(1-t)^{b-1}dt$ for  $a,b>0$ and $0< z\le 1$, and \be M :=  \frac{K}{K+m}
-\one{\phi(N) = KN} + \one{\phi(N)/N \to \infty } \ee where $K > 0$ is a
-constant and  $m := \lim_{N\to \infty} \EE{X} = 1 + 2^{1-\alpha}/(\alpha - 1)$
-\citep{CDEE2020,AEKKZ2020}.    In other words,  the measure $(\Lambda)$ driving
-the multiple mergers is of the same form as in Eq.~\eqref{Fbeta}  with $0 < x
-\le M$ in the case $1 < \alpha < 2$ and $\phi(N) \ge KN$.   If $\alpha \ge 2$
-or $\phi(N)/N \to 0$ then  the  ancestral process converges (in the sense of
-convergence of  finite-dimensional distributions) to  the Kingman-coalescent
-\citep{CDEE2020,AEKKZ2020}.
-
-\begin{comment}%
-A coalescent process is a continuous-time Markov process taking values among
-the partitions of $\IN := \{1,2, \ldots \}$, such that the restriction to any
-finite $n \in \IN $ takes values among partitions of $[n] := \{1, 2, \ldots,
-n\}$.  Write $\one{A} = 1$ if $A$ holds, and zero otherwise.   Let $\cP_n$
-denote the set of partitions of $[n]$.  In the classical Kingman-coalescent,
-the only possible transitions are the mergers of pairs of blocks (elements of a
-partition $\pi \in \cP_n$), one pair at a time.  The $n$ leaves (corresponding
-to the sampled DNA sequences) are arbitrarily labelled from 1 to $n$, and  the
-blocks of a partition represent the common ancestors of the labels of each
-block.    The initial state is  (usually) taken as $\{ \{1\}, \ldots, \{n\}\}$,
-and the final state, i.e.\ when the most recent common ancester is reached, as
-$\{ [n]\}$. A block in a partition of $[n]$ represents an ancestor of the
-leaves in the block, i.e.\ the block $\{i_1, \ldots, i_k\}$ in a given
-partition of $[n]$ is an ancestor of the $k$ leaves $i_1, \ldots, i_k \in [n]$,
-and the leaves  correspond to arbitrarily labelled DNA sequences in the sample.
-\end{comment}
+where $B(z;a,b) := \int_0^z
+t^{a-1}(1-t)^{b-1}dt$ for  $a,b>0$ and $0< z\le 1$ is the incomplete beta function, and
+\begin{equation*}
+M :=  \lim_{ N \to \infty } \frac{\phi(N) / N}{\phi(N) / N + \EE{X_1}} \in (0, 1]
+\end{equation*}
+\citep{CDEE2020}.    In other words,  the measure $\Lambda$ driving
+the multiple mergers is of the same form as in \eqref{Fbeta}  with $0 < x
+\le M$ in the case $1 < \alpha < 2$ and $\lim_{ N \to \infty } \phi(N) / N > 0$.   If $\alpha \ge 2$
+or $\phi(N)/N \to 0$ then  the  ancestral process converges to  the Kingman-coalescent
+\citep{CDEE2020}.
 

### 2021-08-05 paper.tex

@@
 Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
-works backwards in time, generating common ancestor and recombination
-events and tracking their effects on segments of ancestral
-material inherited from the
-sample~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
-Common ancestor events merge the ancestral material of two lineages,
-and result in coalescences in the marginal trees when ancestral
-segments overlap.
-Recombination events split the ancestral material for some lineage
-at a given breakpoint, creating two independent
+works backwards in time, tracking the
+effects of common ancestor and recombination
+events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
+Common ancestor events combine the ancestral material of two lineages, and may
+result in coalescences in the marginal trees. Recombination events
+split the ancestral material for some lineage, creating two independent
 lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
@@
 against analytical results~\citep{hudson1983properties,kaplan1985use}
-and simulations by \ms, \texttt{msHot} and \SLiM.
+and simulations by \ms.
 
@@
     for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
-    Each point is the run time of one simulation,
+    Each point is the runtime of one simulation,
     for various values of effective population size ($N_e$),
     genome length in Morgans ($L$), and number of samples ($n$).
-    Run time scales quadratically with the product of $N_e$ and $L$,
-    shown on the horizontal axis. For example, (A) shows that
+    Runtime scales quadratically with the product of $N_e$ and $L$,
+    shown on the horizontal axis---for instance, figure A shows that
     1,000 samples of 1 Morgan-length chromosomes
     from a population of $N_e=2,000$ diploids
-    would take about 2 seconds, and (equivalently) that the same number of
-    0.01 Morgan segments with $N_e= 200,000$ would take the same time.
+    would take about 2 seconds, and that the same number of
+    0.01 Morgan segments in a population of size 200,000
+    would take about the same time.
     Since recombination rate in these simulations was $10^{-8}$,
@@
     and sample size. Also shown are the approximate values of
-    $N_e L$ for chromosome 1 in three species, using values from
-    the \stdpopsim\ catalogue~\citep{adrion2020community}.
+    $\rho$ for chromosome 1 in three species, using values from
+    the \stdpopsim\ catalog~\citep{adrion2020community}.
 }
@@
 to note that although the implementation of Hudson's algorithm is
-very efficient, it is still quadratic in the population scaled recombination rate
+very efficient, it is still quadratic in the scaled recombination rate
 $\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
-This is because Hudson's algorithm tracks recombinations not only
+This is because Hudson's algorithms tracks recombinations not only
 in segments ancestral to the sample, but also between ancestral segments.
-As mentioned above, common ancestor events in which the ancestral material
-of two lineages is merged only result in coalescences in the marginal trees
-if their ancestral segments overlap. If there is no overlap, the merged
-segments represent an ancestral chromosome that is a genetic ancestor
-of the two linages, but not the most recent common genetic ancestor
-at any location along the genome.
-When this happens, the merged lineage carries ``trapped'' genetic material
-that is not ancestral to any samples, but where recombinations can still occur.
-The SMC approximations work by disallowing common ancestor events
-that generate trapped material, greatly simplifying the process.
-However, this also removes subtle long-range
-correlations in the trees since there are many ways in which ancestry
-segments can merge without overlapping.
-
-For large $\rho$, recombination events in trapped ancestral material will
-dominate, and so we can use this as a proxy for the overall number
-of events in Hudson's algorithm.
+In other words, lineages may ``coalesce'' even if the pieces of ancestral material they carry
+do not overlap---representing an ancestral chromosome
+that is a genetic ancestor to each of two linages,
+but not the most recent common genetic ancestor at any location along the genome.
+When this happens, the resulting merged lineage carries ``trapped'' genetic material
+that is not ancestral to any samples;
+recombinations in the trapped material will cause the two segments of ancestry
+to split apart into separate lineages again.
+For large $\rho$, recombination events
+in this trapped material will dominate.
 \citet[Eq.~5.10]{hein2004gene} gave
-\begin{equation}\label{eqn-hsw-bound}
-\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
-\end{equation}
+$\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2$
 as an upper bound
 on the number of recombination events within trapped ancestral material
-in Hudson's algorithm for $n$ samples.
-% Since the sum is well approximated by $\log{n}$,
-% a natural conjecture is that the time complexity of Hudson's
-% algorithm is $O(\rho^2 \log^2 n)$.
-Fig.~\ref{fig-ancestry-perf} shows the observed run time for
-simulations with a variety of population size, chromosome
-length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
-correctly predicts the quadratic dependence on $\rho$,
-as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
-We also see that the dependence on $n$ is quite weak,
-since increasing sample size 100-fold
-only increases run time by a factor of 2 or so. However, the
-$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
-(the sum is a Harmonic number, and approximated by $\log{n}$)
-is not well supported by the data except possibly at very
-large values of $\rho$, and so it seems a tighter bound
-than Eq.~\eqref{eqn-hsw-bound} would
-be required to accurately predict simulation time for a
-given $\rho$ and $n$.
-
-Fig.~\ref{fig-ancestry-perf} is a useful yardstick, allowing us
-to predict how long simulations should take for a wide range of
-species. Taking a typical chromosome to be 1 Morgan in length,
+in Hudson's algorithm with $n$ samples.
+Since the sum is well approximated by $\log{n}$,
+a natural conjecture is that the time complexity of Hudson's
+algorithm is $O(\rho^2 \log^2 n)$.
+This turns out to have the correct dependence on $\rho$---although
+caution is needed; see below---but the dependence on $n$ is less accurate.
+
+Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
+a given simulation of a constant size population,
+and verifies that runtime scales quadratically with $\rho$,
+and depends much less strongly on the number of samples.
+Taking a typical chromosome to be 1 Morgan in length,
 these plots show, roughly, that simulating chromosome-length
@@
 while samples from a population of tens of thousands takes minutes.
-Simulating whole chromosomes for many species is very fast,
-with 1000 samples of chromosome 1 for
-\emph{Arabidopsis thaliana} taking less than a second, and a few
-minutes for dogs and humans. However, the dependence on $\rho$
-\emph{is} quadratic, and if $\rho$ is sufficiently large simulations
-may not be feasible. For example, for
-\emph{Drosophila melanogaster} chromosome 2L, with
-$\rho\approx 3.9\times10^6$, Fig.~\ref{fig-ancestry-perf}
-predicts that simulation would require around 177 hours. For such
-large values of $\rho$  we recommend
-users consider approximate simulations. Since \msprime\ does not currently
-have efficient implementations of approximate coalescent with recombination
-models, we recommend using SMC based methods such as \scrm\,
-particularly for moderate sample sizes.
-In practice, to predict the running time of a given
-simulation in \msprime, we recommend that users
-measure run time in a series of simulations with short genome lengths and
-the desired sample size,
-% possibly varying other parameters to check for interactions,
-% JK: why? ^
-% I'm worried people will take this as meaning "run every possible
-% combination of parameters". I think the genome length should be basically
-% independent of everything else, right?
-and then predict run time by fitting a quadratic curve to genome length
-as in Fig.~\ref{fig-ancestry-perf}.
-It is important to note that the quadratic curves in the two
-panels of Fig.~\ref{fig-ancestry-perf} are different, and
-predicting the run times of days-long simulations
+However, it is important to note that the quadratic curves in the two
+panels are different---predicting the runtimes of days-long simulations
 using the timing of minutes-long runs is unlikely to be very accurate.
+Further experiments show that adding a $\log^2 n$ factor
+to predict runtimes across different sample sizes does not do well
+except possibly at very large values of $\rho$.
+
+% Although \citet{hein2004gene} refer to the bound in \eqref{eqn-hsw-bound}
+% as ``crude'', it predicts the scaling of run time quite well for large genomes
+% Figure~\ref{fig-expected-perf} shows the observed runtime
+% (from the same simulations as Figure~\ref{fig-ancestry-perf})
+% plotted against equation~\eqref{eqn-hsw-bound},
+% showing that for values of $\rho$ above $4 \times 10^4$ or so
+% (i.e., chromosome-length simulations in populations above $10^4$),
+% the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
+% This implies the runtime for a large sample
+% can be predicted from the runtime for a small sample.
+% For smaller values of $\rho$, however, the prediction
+% severely underestimates runtime.
 
 What about simulations with changing population size?
-To understand how run time depends on demography
-it helps to consider why run time is quadratic in $\rho$.
+To understand how runtime depends on demography
+it helps to consider why runtime is quadratic in $\rho$.
 At any point in time, \msprime\ must keep track of some number of lineages,
@@
 and so recombine apart again immediately;
-it is these large numbers of rapid and inconsequential events that
-lead to the quadratic run time.
+it is these large numbers of rapid yet inconsequential events that
+lead to the quadratic runtime.
+\plrcomment{(this is discussed also in the ARG section; check for overlap)}
 The maximum number of lineages occurs when
@@
 common ancestor and recombination events balance out;
-to get an idea of run time we can estimate when this occurs.
-Suppose that the maximum number of lineages is $M$;
+to get an idea of runtime we can estimate when this occurs.
+Suppose that the maximal number of lineages is $M$;
 at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
@@
 (including ``trapped'' non-ancestral material).
-Since the maximum number of lineages is reached
+It is known that the maximum number of lineages is reached
 very quickly~\citep{nelson2020accounting},
-many (roughly half) of these lineages have at
+and at this maximum, recombination and coalescence rates are equal.
+Therefore, many (roughly half) of these lineages have at
 least two distant segments of ancestral material
 separated by a long segment of genome, implying that $\ell \approx L / 4$.
-Then, since the rates of recombination and common ancestor events
-are equal the time the maximum number of lineages is present,
-we see that $M$ is roughly equal to $L N_e$.
-The number of lineages then decreases gradually from this maximum
-on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
-Since the total rate of events when the maximum number of lineages is
-present is roughly $L^2 N_e / 4$,
+This implies that $M$ is roughly $L N_e$.
+The number of linages decreases gradually,
+on the coalescent time scale---so, over roughly $2 N_e$ generations.
+Since the total rate of events when there are the maximum number of lineages
+happens is roughly $L^2 N_e / 4$,
 then the total number of events is proportional to $(L N_e)^2$---i.e.,
@@
 The argument above implies that the work is spread out relatively
-evenly on the coalescent time scale.
-% JK: I just found this confusing tbh.
-% in other words,
-% more events happen in the time period from 0 to $N_e/2$ generations
-% than from $N_e/2$ to $N_e$ generations,
-% but the run time is not dominated by the more recent time period.
+evenly on the coalescent time scale: in other words,
+more events happen in the time period from 0 to $N_e/2$ generations
+than from $N_e/2$ to $N_e$ generations,
+but the runtime is not dominated by the more recent time period.
 Suppose that population size today is $N_1$,
 while longer ago than $T$ generations ago it was $N_2$:
-does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
+does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
 The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
-then run time will be similar to a population of size $N_1$;
+then runtime will be similar to a population of size $N_1$;
 while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
 For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
-and the run time will depend critically on $T$---in other words,
+and the runtime will depend critically on $T$---in other words,
 simulation will be quick in a species with a strong domestication bottleneck,
 and slow otherwise.
+
+In practice, we recommend that users estimate runtime by
+measuring runtime in a series of simulations with short genome lengths,
+possibly varying other parameters to check for interactions,
+and then predict runtime by fitting a quadratic curve to genome length.
+In these experiments, sample sizes should be roughly
+those desired---which is not a problem in practice because of
+the weak dependency on sample size
+---notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
+only increases runtime by a factor of 2 or so.
 
@@
 \end{center}
-\caption{\label{fig-selection-perf}Comparison of the run time for
+\caption{\label{fig-selection-perf}Comparison of the runtime for
 \msprime\ and \discoal\ for simulating selective sweeps.

### 2021-08-05 evaluation/generate_dtwf_perf_data.py
Update DTWF benchmarks and data.

@@ -1,21 +1,21 @@
 import time
 import subprocess
+import concurrent.futures
+import tempfile
 
+import matplotlib.pyplot as plt
+import statsmodels.api as sm
+import numpy as np
+import click
 import tskit
 import msprime
 import pandas as pd
+import cpuinfo
 
-# Methods for simulating with ARGON and msprime.
 
-
-def sim_argon(
-    sample_size,
-    chrom_length,
-    min_length=0,
-    no_rec=False,
-    ibd=True,
-    trees=True,
-):
+def sim_argon(sample_size, chrom_length_mb, count_trees=False):
+    tmpdir = tempfile.TemporaryDirectory()
+    prefix = tmpdir.name + "/x"
     # Defaults: mutation rate 1.65e-8, rec rate 1e-8
     diploid_size = 2 * sample_size
     arguments = [
@@ -26,146 +26,155 @@ def sim_argon(
         "10000",
         "-pop",
         "1",
-        "%i" % diploid_size,
+        f"{diploid_size}",
         "-size",
-        "%i" % int(chrom_length / 1_000_000),
+        f"{chrom_length_mb}",
         "-mut",
         "0",
         "-quiet",
+        "-out",
+        prefix,
     ]
-    if trees:
+    if count_trees:
         arguments.extend(["-trees"])
-    if ibd:
-        arguments.extend(["-IBD", "%i" % min_length])
-    if no_rec:
-        arguments.extend(["-rec", "0"])
+    # print("  ".join(arguments))
+    subprocess.run(arguments, check=True, capture_output=True)
+    num_trees = 0
+    if count_trees:
+        treefile = prefix + ".trees"
+        # count the unique trees in the output. Technically we only want to
+        # count *adjacent* identical trees, though.
+        cmd = f"cut -f 4 {treefile} | uniq | wc -l"
+        output = subprocess.run(cmd, shell=True, check=True, capture_output=True)
+        num_trees = int(output.stdout)
+    return num_trees
 
-    output = subprocess.run(arguments, check=True, capture_output=True)
-    return (output.stdout, output.stderr)
 
-
-# Methods for simulating with msprime.
-
-
-def sim_msprime(
-    sample_size, chrom_length, min_length=0, max_time=None, ibd=True, no_rec=False
-):
-    if no_rec:
-        rho = 0
-    else:
-        rho = 1e-8
+def sim_msprime(sample_size, chrom_length_mb, model="dtwf"):
[…]

### 2021-08-05 evaluation/plot.py

@@ -111,56 +111,17 @@ def dtwf_perf():
     Plot the DTWF benchark.
     """
 
-    df = pd.read_csv(
-        "data/dtwf-perf.csv", sep="\t", usecols=["length", "program", "runtime"]
-    )
+    df = pd.read_csv("data/dtwf-perf.csv")
+    df = df.groupby(["L", "tool"]).mean().reset_index()
 
-    # Make a new dataframe with just the runtime means.
-    means = df.groupby(["length", "program"]).mean().reset_index()
-    sample_sizes = [0, 1, 2, 3, 4]
-    sample_size_labs = [
-        r"$10^{-2}$",
-        r"$10^{-1}$",
-        r"$10^{0}$",
-        r"$10^{1}$",
-        r"$10^{2}$",
-    ]
+    fig, ax1 = plt.subplots(1, 1)
+    for tool in set(df.tool):
+        dft = df[df.tool == tool]
+        ax1.plot(dft.L, dft.time, label=tool)
 
-    # print(means)
-
-    # # Plot the means
-    plt.plot(
-        sample_sizes,
-        np.log10(means[means["program"] == "msprime-DTWF"]["runtime"]),
-        label="msprime-DTWF",
-        marker="o",
-        color="C1",
-    )
-    plt.plot(
-        sample_sizes,
-        np.log10(means[means["program"] == "msprime-hybrid"]["runtime"]),
-        label="msprime-hybrid",
-        marker="o",
-        color="C1",
-        linestyle="dashed",
-    )
-    plt.plot(
-        sample_sizes,
-        np.log10(means[means["program"] == "argon"]["runtime"]),
-        label="argon",
-        marker="o",
-        color="C0",
-    )
-    plt.legend(loc="upper left")
-    plt.xlabel("Chromosome lengths (Mb)")
-    plt.ylabel("Runtimes (secs)")
-    plt.title("Simulation runtimes for different genome lengths")
-    plt.xticks(sample_sizes, labels=sample_size_labs)
-    plt.yticks(
-        [-2, -1, 0, 1, 2],
-        labels=[r"$10^{-2}$", r"$10^{1}$", r"$10^{0}$", r"$10^{1}$", r"$10^{2}$"],
-    )
-    plt.grid(b=True, linewidth=0.5, linestyle="dotted")
+    ax1.set_xlabel("Sequence length (Megabases)")
+    ax1.set_ylabel("Time (seconds)")
+    ax1.legend()
     save("dtwf-perf")
 
 

### 2021-08-05 paper.tex
Typo corrections in multiple merger appendix

@@
 where $a \geq 0$ is a constant and $\Lambda$ is a finite measure on the unit interval without an atom at zero \citep{DK99,P99,S99}.
-There is also a larger class of multiple merger coalescents involving simultaneous
-multiple mergers of distinct groups of ancestral lineages \citep{S00}.
+There is also a larger class of simultaneous multiple merger coalescents involving
+simultaneous mergers of distinct groups of lineages into several common ancestors \citep{S00}.
 These are commonly referred to as $\Xi$-coalescents, and
@@
 two lineages merge occur at rate $1/(1 + c\psi^2/4)$, while large reproduction events
-giving rise to simultaneous multiple mergers occur at rate $(c\psi^2/4) / (1 + c\psi^2/4)$.
+with the potential to result in simultaneous multiple mergers occur at rate $(c\psi^2/4) / (1 + c\psi^2/4)$.
 
@@
 with index $\alpha > 0$, and where $C > 0$ is a normalising constant.
-If the  total number of juveniles $S_N := X_1 + \ldots + X_n$ produced
+If the  total number of juveniles $S_N := X_1 + \ldots + X_N$ produced
 in this way is at least $N$, then $N$ juveniles are sampled uniformly at random
@@
 large numbers of juveniles. Since juveniles are at least fertilised eggs,
-it may be desirable suppose that the number of juveniles
+it may be desirable to suppose that the number of juveniles
 surviving to reproductive maturity cannot be arbitrarily large.  Hence we also consider

### 2021-08-05 paper.tex
Response to review

--- 2021-08-05 paper.tex
@@
 $\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
-(the sum is a Harmonic number, and approximated by $\log{n}$)
+(the sum is a harmonic number, and approximated by $\log{n}$)
 is not well supported by the data except possibly at very
-large values of $\rho$, and so it seems a tighter bound
-than Eq.~\eqref{eqn-hsw-bound} would
-be required to accurately predict simulation time for a
+large values of $\rho$, and so it seems a different dependence
+on $n$ would be required to accurately predict simulation time for a
 given $\rho$ and $n$.
@@
 the desired sample size,
-% possibly varying other parameters to check for interactions,
-% JK: why? ^
-% I'm worried people will take this as meaning "run every possible
-% combination of parameters". I think the genome length should be basically
-% independent of everything else, right?
 and then predict run time by fitting a quadratic curve to genome length
@@
 evenly on the coalescent time scale.
-% JK: I just found this confusing tbh.
-% in other words,
-% more events happen in the time period from 0 to $N_e/2$ generations
-% than from $N_e/2$ to $N_e$ generations,
-% but the run time is not dominated by the more recent time period.
 Suppose that population size today is $N_1$,

### 2021-08-05 paper.tex

@@
         \label{fig-mutated-trees}
-        Visualization of the separation between ancestry and mutation simulation:
+        Visualisation of the separation between ancestry and mutation simulation:
         \texttt{sim\_ancestry} produces ancestry
@@
 segments represent an ancestral chromosome that is a genetic ancestor
-of the two linages, but not the most recent common genetic ancestor
+of the two lineages, but not the most recent common genetic ancestor
 at any location along the genome.
@@
 support gene conversions.
-As bacterial recombination can be modeled analogously to eukaryotic allelic
+As bacterial recombination can be modelled analogously to eukaryotic allelic
 gene conversion, simulators for microbial evolution often include gene conversion
@@
 will be to add further features to model bacterial
-evolution. In particular, we aim to include bacterial gene transfer and utilize
+evolution. In particular, we aim to include bacterial gene transfer and utilise
 the succinct tree sequence structure to represent the resulting ancestral gene
@@
 Details of our implementation are available in the Appendix.
-This implementation allows us to further generalize sweep models to
+This implementation allows us to further generalise sweep models to
 changing population sizes as well as selection from standing variation,
@@
 with mentorship by experienced developers,
-is a hugely useul experience for many users,
+is a hugely useful experience for many users,
 The skills that contributors learn
@@
 
-\section*{Acknowledgments}
+\section*{Acknowledgements}
 JK is supported by the Robertson Foundation.
@@
 
-First, mutations are placed on the tree sequence under an inhomgeneous Poisson model
+First, mutations are placed on the tree sequence under an inhomogeneous Poisson model
 by applying them independently to each edge.
@@
 process with rate $\beta$.
-All finite-state models (implented under the generic \texttt{MatrixMutationModel} class)
+All finite-state models (implemented under the generic \texttt{MatrixMutationModel} class)
 work in this way: mutations are placed at the maximum mutation rate,
@@
 the population at a given time. The allele frequency trajectory of the
-sweeping allele is modeled using a jump process that approximates
+sweeping allele is modelled using a jump process that approximates
 the conditional diffusion of the allele, conditioning on its eventual

### 2021-08-05 evaluation/generate_dtwf_perf_data.py
Add support for tracking memory (and thread time for ARGON)

@@ -1,7 +1,10 @@
 import time
 import subprocess
 import concurrent.futures
+import multiprocessing
 import tempfile
+import os
+import resource
 
 import matplotlib.pyplot as plt
 import statsmodels.api as sm
@@ -11,6 +14,7 @@ import tskit
 import msprime
 import pandas as pd
 import cpuinfo
+import sys
 
 
 def sim_argon(sample_size, chrom_length_mb, count_trees=False):
@@ -73,19 +77,38 @@ def sim_msprime_hybrid(sample_size, chrom_length_mb):
     )
 
 
-def run_benchmark(work):
-    tool, L = work
-    sample_size = 500
+def process_resources():
+    utime = 0
+    stime = 0
+    mem = 0
+    for who in [resource.RUSAGE_CHILDREN, resource.RUSAGE_SELF]:
+        info = resource.getrusage(who)
+        utime += info.ru_utime
+        stime += info.ru_stime
+        mem += info.ru_maxrss
+    # Memory is returned in KiB, scale to bytes
+    return {"user_time": utime, "sys_time": stime, "memory": mem * 1024}
 
+
+def run_benchmark_process(tool, L, queue):
+    sample_size = 500
     func = tool_map[tool]
-    before = time.perf_counter()
     func(sample_size=sample_size, chrom_length_mb=L)
-    duration = time.perf_counter() - before
-    return {
-        "L": L,
-        "tool": tool,
-        "time": duration,
-    }
+    queue.put(process_resources())
+
+
+def run_benchmark(work):
+
+    tool, L = work
+    queue = multiprocessing.Queue()
+    p = multiprocessing.Process(target=run_benchmark_process, args=(tool, L, queue))
+    before = time.perf_counter()
+    p.start()
+    perf_metrics = queue.get()
+    p.join()
+    if p.exitcode < 0:
+        raise ValueError("Error occured", p.exitcode)
+    return {"L": L, "tool": tool, **perf_metrics}
 
 
 tool_map = {

### 2021-08-05 evaluation/plot.py
Update with full DWTF data

@@ -113,14 +113,22 @@ def dtwf_perf():
 
     df = pd.read_csv("data/dtwf-perf.csv")
     df = df.groupby(["L", "tool"]).mean().reset_index()
+    # print(df)
 
-    fig, ax1 = plt.subplots(1, 1)
-    for tool in set(df.tool):
+    df.memory /= 1024 ** 3
+    label_map = {"ARGON": "ARGON", "msprime": "DTWF",
+            "hybrid": "DTWF + Hudson"}
+
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
+    for tool in ["ARGON", "msprime", "hybrid"]:
         dft = df[df.tool == tool]
-        ax1.plot(dft.L, dft.time, label=tool)
+        ax1.plot(dft.L, dft.user_time, label=label_map[tool])
+        ax2.plot(dft.L, dft.memory, label=label_map[tool])
 
     ax1.set_xlabel("Sequence length (Megabases)")
     ax1.set_ylabel("Time (seconds)")
+    ax2.set_xlabel("Sequence length (Megabases)")
+    ax2.set_ylabel("Memory (GiB)")
     ax1.legend()
     save("dtwf-perf")
 

### 2021-08-05 paper.tex
Update paper for figure

--- 2021-08-05 paper.tex
@@
 to note that although the implementation of Hudson's algorithm is
-very efficient, it is still quadratic in the population scaled recombination rate
+very efficient, it is still quadratic in the scaled recombination rate
 $\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
-This is because Hudson's algorithm tracks recombinations not only
+This is because Hudson's algorithms tracks recombinations not only
 in segments ancestral to the sample, but also between ancestral segments.
-As mentioned above, common ancestor events in which the ancestral material
-of two lineages is merged only result in coalescences in the marginal trees
-if their ancestral segments overlap. If there is no overlap, the merged
-segments represent an ancestral chromosome that is a genetic ancestor
-of the two linages, but not the most recent common genetic ancestor
-at any location along the genome.
-When this happens, the merged lineage carries ``trapped'' genetic material
-that is not ancestral to any samples, but where recombinations can still occur.
-The SMC approximations work by disallowing common ancestor events
-that generate trapped material, greatly simplifying the process.
-However, this also removes subtle long-range
-correlations in the trees since there are many ways in which ancestry
-segments can merge without overlapping.
-
-For large $\rho$, recombination events in trapped ancestral material will
-dominate, and so we can use this as a proxy for the overall number
-of events in Hudson's algorithm.
+In other words, lineages may ``coalesce'' even if the pieces of ancestral material they carry
+do not overlap---representing an ancestral chromosome
+that is a genetic ancestor to each of two linages,
+but not the most recent common genetic ancestor at any location along the genome.
+When this happens, the resulting merged lineage carries ``trapped'' genetic material
+that is not ancestral to any samples;
+recombinations in the trapped material will cause the two segments of ancestry
+to split apart into separate lineages again.
+For large $\rho$, recombination events
+in this trapped material will dominate.
 \citet[Eq.~5.10]{hein2004gene} gave
-\begin{equation}\label{eqn-hsw-bound}
-\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
-\end{equation}
+$\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2$
 as an upper bound
 on the number of recombination events within trapped ancestral material
-in Hudson's algorithm for $n$ samples.
-% Since the sum is well approximated by $\log{n}$,
-% a natural conjecture is that the time complexity of Hudson's
-% algorithm is $O(\rho^2 \log^2 n)$.
-Fig.~\ref{fig-ancestry-perf} shows the observed run time for
-simulations with a variety of population size, chromosome
-length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
-correctly predicts the quadratic dependence on $\rho$,
-as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
-We also see that the dependence on $n$ is quite weak,
-since increasing sample size 100-fold
-only increases run time by a factor of 2 or so. However, the
-$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
-(the sum is a harmonic number, and approximated by $\log{n}$)
-is not well supported by the data except possibly at very
-large values of $\rho$, and so it seems a different dependence
-on $n$ would be required to accurately predict simulation time for a
-given $\rho$ and $n$.
-
-Fig.~\ref{fig-ancestry-perf} is a useful yardstick, allowing us
-to predict how long simulations should take for a wide range of
-species. Taking a typical chromosome to be 1 Morgan in length,
+in Hudson's algorithm with $n$ samples.
+Since the sum is well approximated by $\log{n}$,
+a natural conjecture is that the time complexity of Hudson's
+algorithm is $O(\rho^2 \log^2 n)$.
+This turns out to have the correct dependence on $\rho$---although
+caution is needed; see below---but the dependence on $n$ is less accurate.
+
+Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
+a given simulation of a constant size population,
+and verifies that runtime scales quadratically with $\rho$,
+and depends much less strongly on the number of samples.
+Taking a typical chromosome to be 1 Morgan in length,
 these plots show, roughly, that simulating chromosome-length
@@
 while samples from a population of tens of thousands takes minutes.
-Simulating whole chromosomes for many species is very fast,
-with 1000 samples of chromosome 1 for
-\emph{Arabidopsis thaliana} taking less than a second, and a few
-minutes for dogs and humans. However, the dependence on $\rho$
-\emph{is} quadratic, and if $\rho$ is sufficiently large simulations
-may not be feasible. For example, for
-\emph{Drosophila melanogaster} chromosome 2L, with
-$\rho\approx 3.9\times10^6$, Fig.~\ref{fig-ancestry-perf}
-predicts that simulation would require around 177 hours. For such
-large values of $\rho$  we recommend
-users consider approximate simulations. Since \msprime\ does not currently
-have efficient implementations of approximate coalescent with recombination
-models, we recommend using SMC based methods such as \scrm\,
-particularly for moderate sample sizes.
-In practice, to predict the running time of a given
-simulation in \msprime, we recommend that users
-measure run time in a series of simulations with short genome lengths and
-the desired sample size,
-and then predict run time by fitting a quadratic curve to genome length
-as in Fig.~\ref{fig-ancestry-perf}.
-It is important to note that the quadratic curves in the two
-panels of Fig.~\ref{fig-ancestry-perf} are different, and
-predicting the run times of days-long simulations
+However, it is important to note that the quadratic curves in the two
+panels are different---predicting the runtimes of days-long simulations
 using the timing of minutes-long runs is unlikely to be very accurate.
+Further experiments show that adding a $\log^2 n$ factor
+to predict runtimes across different sample sizes does not do well
+except possibly at very large values of $\rho$.
+
+% Although \citet{hein2004gene} refer to the bound in \eqref{eqn-hsw-bound}
+% as ``crude'', it predicts the scaling of run time quite well for large genomes
+% Figure~\ref{fig-expected-perf} shows the observed runtime
+% (from the same simulations as Figure~\ref{fig-ancestry-perf})
+% plotted against equation~\eqref{eqn-hsw-bound},
+% showing that for values of $\rho$ above $4 \times 10^4$ or so
+% (i.e., chromosome-length simulations in populations above $10^4$),
+% the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
+% This implies the runtime for a large sample
+% can be predicted from the runtime for a small sample.
+% For smaller values of $\rho$, however, the prediction
+% severely underestimates runtime.
 
 What about simulations with changing population size?
-To understand how run time depends on demography
-it helps to consider why run time is quadratic in $\rho$.
+To understand how runtime depends on demography
+it helps to consider why runtime is quadratic in $\rho$.
 At any point in time, \msprime\ must keep track of some number of lineages,
@@
 The argument above implies that the work is spread out relatively
-evenly on the coalescent time scale.
+evenly on the coalescent time scale: in other words,
+more events happen in the time period from 0 to $N_e/2$ generations
+than from $N_e/2$ to $N_e$ generations,
+but the runtime is not dominated by the more recent time period.
 Suppose that population size today is $N_1$,
 while longer ago than $T$ generations ago it was $N_2$:
-does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
+does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
 The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
-then run time will be similar to a population of size $N_1$;
+then runtime will be similar to a population of size $N_1$;
 while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
 For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
-and the run time will depend critically on $T$---in other words,
+and the runtime will depend critically on $T$---in other words,
 simulation will be quick in a species with a strong domestication bottleneck,
 and slow otherwise.
+
+In practice, we recommend that users estimate runtime by
+measuring runtime in a series of simulations with short genome lengths,
+possibly varying other parameters to check for interactions,
+and then predict runtime by fitting a quadratic curve to genome length.
+In these experiments, sample sizes should be roughly
+those desired---which is not a problem in practice because of
+the weak dependency on sample size
+---notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
+only increases runtime by a factor of 2 or so.
 
@@
 \begin{center}
-\includegraphics[width=6cm]{figures/dtwf-perf}
+\includegraphics[width=\textwidth]{figures/dtwf-perf}
 \end{center}

### 2021-08-05 paper.tex
handwaving

--- 2021-08-05 paper.tex
@@
 %National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico
+
+% Peter L. Ralph
+% Institute of Ecology and Evolution, Departments of Mathematics and Biology, University of Oregon, USA
 
@@
 \emph{is} quadratic, and if $\rho$ is sufficiently large simulations
-may not be feasible. For example, for
-\emph{Drosophila melanogaster} chromosome 2L, with
-$\rho\approx 3.9\times10^6$, Fig.~\ref{fig-ancestry-perf}
-predicts that simulation would require around 177 hours. For such
-large values of $\rho$  we recommend
+may not be feasible. For example,
+the \emph{Drosophila melanogaster} chromosome 2L
+is about 23.5Mb long and an average recombination rate of around $2.4 \times 10^{-8}$,
+so $L \approx 0.57$,
+and with $N_e = 1.7 \times 10^6$ \citep{li2006inferring},
+$N_e L \approx 10^6$, so extrapolating the curve in Fig.~\ref{fig-ancestry-perf}B
+predicts that simulation would require around 177 hours.
+For such large values of $\rho$  we recommend
 users consider approximate simulations. Since \msprime\ does not currently
@@
 (including ``trapped'' non-ancestral material).
+At the maximum, coalescence and recombination rates are equal,
+so a typical segment of ancestry will spend roughly half its time
+in a lineage with at least one other such segment---and,
+since such lineages carry at least two segments,
+at most one-third of the lineages carry long trapped segments of ancesry.
 Since the maximum number of lineages is reached
 very quickly~\citep{nelson2020accounting},
-many (roughly half) of these lineages have at
-least two distant segments of ancestral material
-separated by a long segment of genome, implying that $\ell \approx L / 4$.
-Then, since the rates of recombination and common ancestor events
-are equal the time the maximum number of lineages is present,
-we see that $M$ is roughly equal to $L N_e$.
+this implies that $\ell \approx L / 6$.
+Setting the rates of recombination and common ancestor events
+to be equal and solving for $M$,
+we find that $M$ is roughly equal to $L N_e$.
 The number of lineages then decreases gradually from this maximum
@@
 Since the total rate of events when the maximum number of lineages is
-present is roughly $L^2 N_e / 4$,
+present is roughly $L^2 N_e / 6$,
 then the total number of events is proportional to $(L N_e)^2$---i.e.,

### 2021-08-05 evaluation/plot.py

@@ -230,9 +230,6 @@ def ancestry_perf():
             X[:, 2] = 1
             b, _, _, _ = np.linalg.lstsq(X, df["time"][ut], rcond=None)
 
-            # JK: I don't feel strongly about this, but maybe it's less
-            # confusing.
-            # ax.set_xlabel("Scaled recombination rate $\\rho/4 = N_e L$")
             ax.set_xlabel("$N_e L$ (= scaled recombination rate $\\rho/4$)")
 
             def fitted_quadratic(x):

### 2021-08-06 evaluation/generate_dtwf_perf_data.py
Add note about apparent discrepency.

@@ -159,6 +159,10 @@ def validate(replicates):
     # at the qqplots, the distributions looks about the same, but there's
     # consistently more from ARGON. We've check the parameters as closely
     # as we can here, so I'm not sure there's much we can do.
+    # However, see the discussion here:
+    # https://github.com/tskit-dev/msprime-1.0-paper/pull/109
+    # When we export to a tree sequence and squash the trees down properly,
+    # we get the same distributions. So, this is fine.
     L = 1  # Megabases
     sample_size = 10
 

### 2021-08-06 evaluation/generate_dtwf_perf_data.py
Update GC benchmarks to capture memory usage
GC perf data

Final gc perf

Update

@@ -77,7 +77,7 @@ def sim_msprime_hybrid(sample_size, chrom_length_mb):
     )
 
 
-def process_resources():
+def get_process_resources():
     utime = 0
     stime = 0
     mem = 0
@@ -86,7 +86,7 @@ def process_resources():
         utime += info.ru_utime
         stime += info.ru_stime
         mem += info.ru_maxrss
-    # Memory is returned in KiB, scale to bytes
+    # Memory is returned in KiB on Linux, scale to bytes
     return {"user_time": utime, "sys_time": stime, "memory": mem * 1024}
 
 
@@ -94,7 +94,7 @@ def run_benchmark_process(tool, L, queue):
     sample_size = 500
     func = tool_map[tool]
     func(sample_size=sample_size, chrom_length_mb=L)
-    queue.put(process_resources())
+    queue.put(get_process_resources())
 
 
 def run_benchmark(work):

### 2021-08-06 evaluation/plot.py

@@ -54,25 +54,31 @@ def gc_perf():
     Plot the gene-conversion benchmark.
     """
     df = pd.read_csv("data/gc-perf.csv")
-    # Get the mean among replicates
-    df.time /= 3600
-    dfg = df.groupby(["sample_size", "tool"]).mean().reset_index()
+    df = df.groupby(["sample_size", "tool"]).mean().reset_index()
+    print(df)
 
-    fig, ax1 = plt.subplots(1, 1)
+    df.memory /= 1024 ** 3
+    df.user_time /= 3600
+
     lines = {}
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
     for tool in ["SimBac", "fastSimBac", "msprime"]:
-        dft = dfg[dfg["tool"] == tool]
-        (line,) = ax1.plot(dft.sample_size, dft.time, label=tool)
+        dft = df[df.tool == tool]
+        (line,) = ax1.plot(dft.sample_size, dft.user_time, label=tool)
+        ax2.plot(dft.sample_size, dft.memory, label=tool, color=line.get_color())
         lines[tool] = line
 
     ax1.set_xlabel("Sample size")
     ax1.set_ylabel("Time (hours)")
-    # We could set a log-scale on the y-axis here but it really
-    # doesn't make much difference
+    ax1.set_ylim(0, None)
+    ax2.set_ylim(0, None)
+    ax2.set_xlabel("Sample size")
+    ax2.set_ylabel("Memory (GiB)")
+    ax1.legend()
 
-    dfmsp = dfg[dfg["tool"] == "msprime"]
+    dfmsp = df[df["tool"] == "msprime"]
     largest_n = np.array(dfmsp.sample_size)[-1]
-    largest_value = np.array(dfmsp.time)[-1]
+    largest_value = np.array(dfmsp.user_time)[-1]
     ax1.plot([largest_n], [largest_value], "o", color=lines["msprime"].get_color())
     ax1.annotate(
         f"{round(largest_value * 60)} mins",
@@ -81,8 +87,8 @@ def gc_perf():
         xy=(largest_n, largest_value),
         xycoords="data",
     )
-    ax1.legend()
     plt.tight_layout()
+
     save("gc-perf")
 
 
@@ -116,8 +122,7 @@ def dtwf_perf():
     # print(df)
 
     df.memory /= 1024 ** 3
[…]

### 2021-08-06 paper.tex
Update DTWF section

@@
 \newcommand{\ms}[0]{\texttt{ms}}
+\newcommand{\msHOT}[0]{\texttt{msHOT}}
 \newcommand{\scrm}[0]{\texttt{scrm}}
@@
 \newcommand{\discoal}[0]{\texttt{discoal}}
+\newcommand{\ARGON}[0]{\texttt{ARGON}}
 
@@
 \newcommand{\plrcomment}[1]{\textcolor{cyan}{#1}}
+\newcommand{\grgcomment}[1]{\textcolor{yellow!60!red}{#1}}
 
@@
 % Institute of Ecology and Evolution, Departments of Mathematics and Biology, University of Oregon, USA
+
+% \author[n]{Graham Gower}
+% \affil[n]{Lundbeck GeoGenetics Centre, Globe Institute, Faculty of Health and Medical Sciences, University of Copenhagen, Denmark}
 
@@
 implementing recombination via the succinct tree sequence data
-structure. Moreover, \msprime\ is a community oriented open-source development
-project, aimed as a way to reduce duplication of effort and increase
-the quality of key simulation software.
+structure. Moreover, development of \msprime\ is collaborative
+and open source, which reduces duplication of effort and promotes
+software quality via community engagement.
+% "Bazaar" development model: https://en.wikipedia.org/wiki/The_Cathedral_and_the_Bazaar
 \end{abstract}
@@
 Here we provide an overview of the extensive abilities of \msprime\ 1.0,
-demonstrate through numerous performance comparisons its advantages over
-current, alternative software packages, and finally lay out future plans for
+demonstrate its performance advantages over alternative software,
+and finally lay out future plans for
 further open-source development of what we hope will become an industry
@@
 of \tskit\ is beyond the scope of this article, we summarise
-some aspects that are important for simulation here.
+some aspects that are important for simulation.
 
@@
 
-The tree sequence encoding of ancestry as set of nodes and edges
+The tree sequence encoding of ancestry as a set of nodes and edges
 as just described, is equivalent to a graph in which edges
@@
 a number of packages have been developed
-to allow simulations be run directly in a high-level
+to allow simulations to be run directly in a high-level
 scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
@@
 Demes format~\citep{gower2021demes} from the command line,
-making simulating complex demography models straightforward.
+making simulation of complex demographies straightforward.
 
@@
 % Any more?
-and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
+and \ms's fork \msHOT~\citep{hellenthal2007mshot},
 has in turn been forked to output information on local
@@
         \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
+        \grgcomment{Add \texttt{sim\_ancestry} and \texttt{sim\_mutations} as subfigure titles?}
     }
@@
 the two aspects are functionally independent within the software.
-Both ancestry and mutation simulators are present in \msprime
+Both ancestry and mutation simulators are present in \msprime\
 for reasons of convenience and history,
@@
 The mature and efficient C and Python interfaces for
-\tskit\ makes it straightforward to add further information to
+\tskit\ make it straightforward to add further information to
 an existing file, and because of its efficient data interchange mechanisms,
@@
 
-As well as providing a new API (see the \nameref{sec-sim-interface} section)
-which emphasises the logical split between ancestry and mutation simulation,
+As well as providing a new API
+that emphasises the logical split between ancestry and mutation simulation
+(see the \nameref{sec-sim-interface} section),
 we have greatly extended the sophistication of
@@
 and PAM~\citep{dayhoff1978} amino acid models.
-A general model accepts an arbitrary transition matrix between
-any number of alleles (which can be arbitrary unicode strings),
-so that other models, such as the Kimura two and three
+Other models, such as the Kimura two and three
 parameter models~\citep{kimura1980simple,kimura1981estimation},
-can be easily and efficiently defined in user code.
+can be defined easily and efficiently in
+user code by specifying an arbitrary transition matrix between
+any number of alleles (which can be arbitrary unicode strings).
 Mutation rates can vary along the genome, and multiple mutation
@@
 \SeqGen~\citep{rambaut1997seq} and
-Pyvolve~\citep{spielman2015pyvolve}.
+\Pyvolve~\citep{spielman2015pyvolve}.
 
@@
 and recombination rate of $10^{-8}$) for varying (haploid) sample
-size and sequence length. We run 10 replicate mutation simulations
+size and sequence length. We ran 10 replicate mutation simulations
 each for three different mutation rates, and report the average
@@
 (since there are 137,081 trees and 1,998 branches in each
-tree) rather than 508,125 edges, resulting in $\sim500$ times more work.
+tree) rather than 508,125 edges, resulting in $\sim$500 times more work.
 % python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
@@
 and indels~\citep{cartwright2005dna,fletcher2009indelible},
-although some may also require upstream additions to \tskit's data model
+although changes may be required to \tskit's data model
 which is currently based on the assumption of independent sites.
@@
 against analytical results~\citep{hudson1983properties,kaplan1985use}
-and simulations by \ms, \texttt{msHot} and \SLiM.
+and simulations by \ms, \msHOT\ and \SLiM.
 
@@
     The black lines are quadratic fits separately in each panel
-    and sample size. Also shown are the approximate values of
+    and sample size. Vertical grey lines show the approximate values of
     $N_e L$ for chromosome 1 in three species, using values from
@@
 since such lineages carry at least two segments,
-at most one-third of the lineages carry long trapped segments of ancesry.
+at most one-third of the lineages carry long trapped segments of ancestry.
 Since the maximum number of lineages is reached
@@
 \subsection*{Gene conversion}
+\grgcomment{TODO: discuss Fig~\ref{fig-gc-perf} in text of GC section.}
 
@@
 Kingman's coalescent assumes that only two ancestral lineages can merge at
-each merger event. Although this is generally a reasonable approximation there
+each merger event. Although this is generally a reasonable approximation, there
 are certain situations in which the underlying mathematical assumptions are
@@
 \end{center}
-\caption{\label{fig-arg} (A) An simple ARG in which a recombination
+\caption{\label{fig-arg} (A) A simple ARG in which a recombination
 occurs at position 0.3; (B) the equivalent topology depicted as a tree
@@
 (D) the fraction of nodes in a tree sequence that are concerned with
-extra ARG information in a tree sequences returned using \msprime's
+extra ARG information in a tree sequence returned using \msprime's
 \texttt{record\_full\_arg} mode}
@@
 rather store the information required to recover the genealogical
-history as nodes and edges in a tree sequence (see section XX),
+history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
 or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
@@
 in which lineages may transit between favoured and unfavoured backgrounds through
-recombination. This allows for many selective sweep scenarios, including recurrent, partial,
-and soft selective sweeps to be simulated efficiently, however in general this
+recombination. This allows for many selective sweep scenarios to be simulated
+efficiently, including recurrent, partial,
+and soft selective sweeps. However in general this
 efficiency comes at the cost of flexibility in comparison to forwards in
@@
 We have implemented selective sweeps in \msprime\ similarly to
-the \discoal\ implementation-- sweep trajectories are
+the \discoal\ implementation---sweep trajectories are
 generated from a jump process approximation to the conditional diffusion
@@
 Details of our implementation are available in the Appendix.
-This implementation allows us to further generalise sweep models to
+This implementation facilitates future changes to generalise sweep models to
 changing population sizes as well as selection from standing variation,
@@
 \subsection*{Discrete time Wright-Fisher}
-
-The assumptions of the classical coalescent model are that both sample size and
-the simulated sequence length are small. As modern datasets often include
-whole-genome sequences for tens or hundreds of thousands of samples, simulating
-such data requires breaking these assumptions. While the coalescent remains a
-good approximation to the Wright-Fisher model in the distant past, it can
-exhibit large biases in the recent history. Identity-by-descent (IBD),
-long-range linkage disequilibrium (LD), and ancestry patterns deviate from
-Wright-Fisher expectations, and this bias grows considerably when simulating
-larger sample
+The coalescent is an idealised model and makes many simplifying assumptions,
+but it is often surprisingly robust to violations of these
+assumptions~\citep{wakeley2012gene}. One situation in which the
+model does break down is the combination of large sample size
+and with long recombining genomes, where the large
+number of recombination events in the recent past results in
+more than the biologically possible $2^t$ ancestors in
+$t$ diploid generations~\citep{nelson2020accounting}.
+This pathological behaviour results in
+Identity-by-descent,
+long-range linkage disequilibrium (LD), and ancestry patterns deviating from
+Wright-Fisher expectations, and the bias grows with larger sample
 sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
-
-A backwards-in-time Wright-Fisher model is supported in \msprime, which is
-shown to properly capture the expected distributions of IBD and long-range LD,
-as well as ancestry patterns in admixed
-populations~\citep{nelson2020accounting}. In short, the classical coalescent
-draws recombination events from exponential waiting times, which are continuous
-in time, and each recombination then splits a segment into two distinct
-ancestors. For large sequence lengths, this can dramatically overestimate the
-number of genealogical ancestors of a given sample. Instead, the Wright-Fisher
-model iterates through discrete generations so that each gamete has a unique
+Precisely this problem occurs when simulating modern human datasets,
+and we have implemented a Discrete Time Wright-Fisher (DTWF) model
+in \msprime\ to address the issue. The DTWF simulates backwards in
+time generation-by-generation so that each gamete has a unique
 diploid parent, and multiple recombinations within a generation results in
-crossover events between the same two parental haploid copies.  This method is
-described in more detail in~\cite{nelson2020accounting}.
+crossover events between the same two parental haploid copies.
+The method is described in more detail by~\cite{nelson2020accounting}.
 
@@
 \begin{center}
-\includegraphics[width=6cm]{figures/dtwf-perf}
+\includegraphics[width=\textwidth]{figures/dtwf-perf}
 \end{center}
-\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
-simulations with ARGON. To ensure the comparison with ARGON is
-as fair as possible we run it with a mutation rate of 0 and with
-the minimum output options (so that it writes out only an empty VCF file)}
+\caption{\label{fig-dtwf-perf} Comparison of Discrete Time Wright-Fisher
+(DTWF) simulation performance in \msprime\ and \ARGON\ (Intel Xeon E5-2680 CPU).
+We simulated ancestry for a sample of 1000 haploids from a population of 10000,
+and report the (A) total CPU time and (B) maximum memory usage for varying
+sequence lengths, and a per-base recombination rate of $10^{-8}$. Each
+point is the average over $5$ replicate simulations.
+We show observations for \ARGON, \msprime's DTWF implementation (``DWTF'')
+and a hybrid simulation of 100 generations of the DTWF followed by
+the standard coalescent with recombination model (``DTWF + Hudson'').
+We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
+to ensure we are measuring only ancestry simulation time.}
 \end{figure}
 
-We support changing the simulation model at specified times so that we can
-easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}. We
-can combine any number of different simulation models, allowing for the
-flexible choice of simulation scenarios without severely damaging efficiency.
+Fig.~\ref{fig-dtwf-perf} shows that \msprime\ simulates the DTWF
+more quickly and requires substantially less memory than
+\ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
+However, the generation-by-generation approach of the DTWF is less
+efficient than the coalescent with recombination when the
+number of lineages is significantly less than the population size
+(the regime where the coalescent is an accurate approximat),
+which usually happens in the quite recent
+past~\cite{bhaskar2014distortion}.
+We therefore support changing the simulation model during a simulation
+so that we can run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}.
+Any number of different simulation models can be combined, allowing for the
+flexible choice of simulation scenarios.
 As the discrete time Wright-Fisher model improves accuracy of genealogical
@@
 model and then switch to the standard coalescent to more efficiently simulate
-the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the performance of
-this approach with ARGON~\citep{palamara2016argon}, a specialised DTWF
-simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
-simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
-hybrid approach provides large efficiency gains.
+the more ancient history.
 
@@
 number of different processes across disjoint time slices.
-In practise this feature is used to ``complete''
+In practice this feature is used to ``complete''
 forwards-time ancestry simulations~\citep{kelleher2018efficient}
@@
 automatically on different operating systems on each pull request
-(where a contributor updates the codebase), using standard Continuous
+(where a contributor proposes a code change), using standard Continuous
 Integration (CI) methodology. Other CI services check code formatting
@@
 217 validation tests, in 6K lines of code). These consist of running many
-replicate simulations, and checking the properties of the output
-against either analytical results of the output of other simulators
-where such results are not available.
+replicate simulations to check the properties of the output
+against other simulators, and where possible against analytical results.
 For example, simulations of complex demography are validated against \ms,
 selective sweeps against \discoal, and Wright-Fisher simulations
-against forwards-time simulations in \SLiM.
+against forwards in time simulations in \SLiM.
 This suite of tests is run before every release, to ensure
@@
 realisation across the sciences \citet[OTHER CITATIONS]{siepel2019challenges}
+% Maybe cite: https://www.biorxiv.org/content/10.1101/092205v3
 that investing in
@@
 features they require are missing will continue to contribute to
-\msprime, leading to a community project that is both high-quality
+\msprime, leading to a community project that is both high quality
 and sustainable in the long term.
@@
 catalogued and understood.
-\texttt{Msprime} will be an important tool for these inferences and validations,
+\texttt{Msprime} will continue to be an important tool for these inferences and validations,
 and in this context the ability to interoperate
-with other methods---particularly, forwards simulators---through
+with other methods---particularly forwards simulators---through
 the succinct tree sequence data structure and \tskit\ library
-will be vital.
+will be essential.
+% this final sentence is hard to parse
 
@@
 EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
+Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
 

### 2021-08-06 paper.tex

@@
 \subsection*{Discrete time Wright-Fisher}
-The coalescent is an idealised model and makes many simplifying assumptions,
-but it is often surprisingly robust to violations of these
-assumptions~\citep{wakeley2012gene}. One situation in which the
-model does break down is the combination of large sample size
-and with long recombining genomes, where the large
-number of recombination events in the recent past results in
-more than the biologically possible $2^t$ ancestors in
-$t$ diploid generations~\citep{nelson2020accounting}.
-This pathological behaviour results in
-Identity-by-descent,
-long-range linkage disequilibrium (LD), and ancestry patterns deviating from
-Wright-Fisher expectations, and the bias grows with larger sample
+
+The assumptions of the classical coalescent model are that both sample size and
+the simulated sequence length are small. As modern datasets often include
+whole-genome sequences for tens or hundreds of thousands of samples, simulating
+such data requires breaking these assumptions. While the coalescent remains a
+good approximation to the Wright-Fisher model in the distant past, it can
+exhibit large biases in the recent history. Identity-by-descent (IBD),
+long-range linkage disequilibrium (LD), and ancestry patterns deviate from
+Wright-Fisher expectations, and this bias grows considerably when simulating
+larger sample
 sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
-Precisely this problem occurs when simulating modern human datasets,
-and we have implemented a Discrete Time Wright-Fisher (DTWF) model
-in \msprime\ to address the issue. The DTWF simulates backwards in
-time generation-by-generation so that each gamete has a unique
+
+A backwards-in-time Wright-Fisher model is supported in \msprime, which is
+shown to properly capture the expected distributions of IBD and long-range LD,
+as well as ancestry patterns in admixed
+populations~\citep{nelson2020accounting}. In short, the classical coalescent
+draws recombination events from exponential waiting times, which are continuous
+in time, and each recombination then splits a segment into two distinct
+ancestors. For large sequence lengths, this can dramatically overestimate the
+number of genealogical ancestors of a given sample. Instead, the Wright-Fisher
+model iterates through discrete generations so that each gamete has a unique
 diploid parent, and multiple recombinations within a generation results in
-crossover events between the same two parental haploid copies.
-The method is described in more detail by~\cite{nelson2020accounting}.
+crossover events between the same two parental haploid copies.  This method is
+described in more detail in~\cite{nelson2020accounting}.
 
@@
 \end{center}
-\caption{\label{fig-dtwf-perf} Comparison of Discrete Time Wright-Fisher
-(DTWF) simulation performance in \msprime\ and \ARGON\ (Intel Xeon E5-2680 CPU).
-We simulated ancestry for a sample of 1000 haploids from a population of 10000,
-and report the (A) total CPU time and (B) maximum memory usage for varying
-sequence lengths, and a per-base recombination rate of $10^{-8}$. Each
-point is the average over $5$ replicate simulations.
-We show observations for \ARGON, \msprime's DTWF implementation (``DWTF'')
-and a hybrid simulation of 100 generations of the DTWF followed by
-the standard coalescent with recombination model (``DTWF + Hudson'').
-We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
-to ensure we are measuring only ancestry simulation time.}
+\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
+simulations with \ARGON. To ensure the comparison with \ARGON\ is
+as fair as possible we run it with a mutation rate of 0 and with
+the minimum output options (so that it writes out only an empty VCF file).}
 \end{figure}
 
-Fig.~\ref{fig-dtwf-perf} shows that \msprime\ simulates the DTWF
-more quickly and requires substantially less memory than
-\ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
-However, the generation-by-generation approach of the DTWF is less
-efficient than the coalescent with recombination when the
-number of lineages is significantly less than the population size
-(the regime where the coalescent is an accurate approximat),
-which usually happens in the quite recent
-past~\cite{bhaskar2014distortion}.
-We therefore support changing the simulation model during a simulation
-so that we can run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}.
+We support changing the simulation model at specified times so that we can
+easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}.
 Any number of different simulation models can be combined, allowing for the
-flexible choice of simulation scenarios.
+flexible choice of simulation scenarios without severely damaging efficiency.
 As the discrete time Wright-Fisher model improves accuracy of genealogical
@@
 model and then switch to the standard coalescent to more efficiently simulate
-the more ancient history.
+the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the performance of
+this approach with \ARGON~\citep{palamara2016argon}, a specialised DTWF
+simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
+simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
+hybrid approach provides large efficiency gains.
 

### 2021-08-07 evaluation/generate_sweeps_perf_data.py
Update the sweeps perf evaluation code.

@@ -1,8 +1,21 @@
-import numpy as np
-import pandas as pd
-import msprime
-import time
 import subprocess
+import concurrent.futures
+import multiprocessing
+import tempfile
+import os
+import resource
+
+import subprocess
+
+import matplotlib.pyplot as plt
+import statsmodels.api as sm
+import numpy as np
+import click
+import tskit
+import msprime
+import pandas as pd
+import cpuinfo
+import sys
 
 
 def _discoal_str_to_msprime(args):
@@ -93,36 +106,88 @@ def run_msprime(args):
         pass
 
 
-def main():
+def get_process_resources():
+    utime = 0
+    stime = 0
+    mem = 0
+    for who in [resource.RUSAGE_CHILDREN, resource.RUSAGE_SELF]:
+        info = resource.getrusage(who)
+        utime += info.ru_utime
+        stime += info.ru_stime
+        mem += info.ru_maxrss
+    # Memory is returned in KiB on Linux, scale to bytes
+    return {"user_time": utime, "sys_time": stime, "memory": mem * 1024}
+
+
+def run_benchmark_process(tool, cmd, queue):
+    func = tool_map[tool]
+    func(cmd)
+    queue.put(get_process_resources())
+
+
+def run_benchmark(work):
+
+    tool, L, cmd = work
+    queue = multiprocessing.Queue()
+    p = multiprocessing.Process(target=run_benchmark_process, args=(tool, cmd, queue))
+    p.start()
+    perf_metrics = queue.get()
+    p.join()
+    if p.exitcode < 0:
+        raise ValueError("Error occured", p.exitcode)
+    return {"L": L, "tool": tool, **perf_metrics}
+
+
+tool_map = {
+    "msprime": run_msprime,
+    "discoal": run_discoal,
+}
+
+
+@click.command()
+@click.option("--replicates", type=int, default=100)
+@click.option("--processes", type=int, default=None)
+def benchmark(replicates, processes):
+    """
+    Runs the benchmark between msprime and discoal
+    """
+
+    cpu = cpuinfo.get_cpu_info()
+    with open("data/sweeps_perf_cpu.txt", "w") as f:
+        for k, v in cpu.items():
+            print(k, "\t", v, file=f)
+
     r = 1e-9
     # Set theta to 0 to avoid conflating mutation generation
     theta = 0
     refsize = 1e6
-    # seqlen = np.linspace(200,200000, 10)
-    seqlen = np.linspace(200, 200000, 5)
-    data = []
-    tool_map = {
[…]

### 2021-08-07 evaluation/plot.py

@@ -98,17 +98,26 @@ def sweeps_perf():
     Plot the sweeps benchmark.
     """
     df = pd.read_csv("data/sweeps_perf.csv")
+    df = df.groupby(["L", "tool"]).mean().reset_index()
+    print(df)
 
-    fig, ax1 = plt.subplots(1, 1)
-    for tool in set(df.tool):
+    df.memory /= 1024 ** 3
+    df.L /= 1000
+
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
+    for tool in ["msprime", "discoal"]:
         dft = df[df.tool == tool]
-        ax1.plot(dft.L, dft.time, label=tool)
+        ax1.plot(dft.L, dft.user_time, label=tool)
+        ax2.plot(dft.L, dft.memory, label=tool)
 
-    ax1.set_xlabel("Sequence length")
+    ax1.set_xlabel("Sequence length (Kilobases)")
     ax1.set_ylabel("Time (seconds)")
-
+    ax2.set_xlabel("Sequence length (Kilobases)")
+    ax2.set_ylabel("Memory (GiB)")
     ax1.legend()
     save("sweeps-perf")
+
+
 
 
 @click.command()

### 2021-08-07 evaluation/generate_sweeps_perf_data.py
Final data for sweeps plot

@@ -1,21 +1,13 @@
 import subprocess
 import concurrent.futures
 import multiprocessing
-import tempfile
-import os
 import resource
 
-import subprocess
-
-import matplotlib.pyplot as plt
-import statsmodels.api as sm
 import numpy as np
 import click
-import tskit
 import msprime
 import pandas as pd
 import cpuinfo
-import sys
 
 
 def _discoal_str_to_msprime(args):

### 2021-08-07 evaluation/plot.py

@@ -103,21 +103,46 @@ def sweeps_perf():
 
     df.memory /= 1024 ** 3
     df.L /= 1000
+    # discoal has very high systime usage, so we need to
+    # include it here as well.
+    df["time"] = (df.user_time + df.sys_time) / 60
 
     fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
+    lines = {}
     for tool in ["msprime", "discoal"]:
         dft = df[df.tool == tool]
-        ax1.plot(dft.L, dft.user_time, label=tool)
+        (line,) = ax1.plot(dft.L, dft.time, label=tool)
+        lines[tool] = line
         ax2.plot(dft.L, dft.memory, label=tool)
 
     ax1.set_xlabel("Sequence length (Kilobases)")
-    ax1.set_ylabel("Time (seconds)")
+    ax1.set_ylabel("Time (minutes)")
     ax2.set_xlabel("Sequence length (Kilobases)")
     ax2.set_ylabel("Memory (GiB)")
     ax1.legend()
+
+    dfmsp = df[df["tool"] == "msprime"]
+    largest_L = np.array(dfmsp.L)[-1]
+    largest_value = np.array(dfmsp.time)[-1]
+    ax1.plot([largest_L], [largest_value], "o", color=lines["msprime"].get_color())
+    ax1.annotate(
+        f"{round(largest_value * 60)} seconds",
+        textcoords="offset points",
+        xytext=(-30, 15),
+        xy=(largest_L, largest_value),
+        xycoords="data",
+    )
+    largest_value = np.array(dfmsp.memory)[-1]
+    ax2.plot([largest_L], [largest_value], "o", color=lines["msprime"].get_color())
+    ax2.annotate(
+        f"{round(largest_value * 1024)} MiB",
+        textcoords="offset points",
+        xytext=(-30, 15),
+        xy=(largest_L, largest_value),
+        xycoords="data",
+    )
+    plt.tight_layout()
     save("sweeps-perf")
-
-
 
 
 @click.command()

### 2021-08-08 paper.tex

--- 2021-08-06 paper.tex
@@
 \newcommand{\ms}[0]{\texttt{ms}}
+\newcommand{\msms}[0]{\texttt{msms}}
+\newcommand{\cosiTwo}[0]{\texttt{cosi2}}
 \newcommand{\msHOT}[0]{\texttt{msHOT}}
@@
 \begin{center}
-\includegraphics[width=7cm]{figures/gc-perf}
+\includegraphics[width=\textwidth]{figures/gc-perf}
 \end{center}
@@
 Another elaboration of the standard neutral coalescent with recombination
-is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
-In the context of the coalescent this amounts to creating a structured population
+is the addition of selective
+sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
+Sweeps are modelled by creating a structured population
 during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
@@
 recombination. This allows for many selective sweep scenarios to be simulated
-efficiently, including recurrent, partial,
-and soft selective sweeps. However in general this
+efficiently, including recurrent, partial, and soft selective sweeps.
+However in general this
 efficiency comes at the cost of flexibility in comparison to forwards in
 time simulation.
-
-Currently a small number of other simulators generate
-coalescent samples with sweeps, including SelSim~\citep{spencer2004selsim},
-mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
-Cosi2~\citep{shlyakhter2014cosi2}, and \discoal~\citep{kern2016discoal}.
-Each of these packages has their own benefits and possible drawbacks,
-however none of them at present will output tree sequence formatted data.
+Several specialised simulators have been developed to simulate
+sweeps in the coalescent,
+including \texttt{SelSim}~\citep{spencer2004selsim},
+\texttt{mbs}~\citep{teshima2009mbs}, \msms~\citep{ewing2010msms},
+\cosiTwo~\citep{shlyakhter2014cosi2}
+and \discoal~\citep{kern2016discoal}.
 
@@
 \begin{center}
-\includegraphics[width=6cm]{figures/sweeps-perf}
+\includegraphics[width=\textwidth]{figures/sweeps-perf}
 \end{center}
-\caption{\label{fig-selection-perf}Comparison of the run time for
-\msprime\ and \discoal\ for simulating selective sweeps.
-}
+\caption{\label{fig-selection-perf}Comparison of selective sweep simulation
+performance in \msprime\ and \discoal\ (Intel Xeon E5-2680 CPU).
+We report the total CPU time and maximum memory usage when simulating 100
+replicates for 10 samples in a model
+with \jkcomment{[ADK: fill in the details here please]} and a
+per-base recombination rate of $10^{-9}$.}
 \end{figure}
 
-We have implemented selective sweeps in \msprime\ similarly to
-the \discoal\ implementation---sweep trajectories are
-generated from a jump process approximation to the conditional diffusion
-of an allele bound for fixation~\citep{coop2004ancestral}.
-Details of our implementation are available in the Appendix.
+Following the approach taken in \discoal, we
+have implemented selective sweeps in \msprime\ by generating
+sweep trajectories from a jump process approximation to the conditional diffusion
+of an allele bound for
+fixation~\citep{coop2004ancestral}---see the appendix for more details.
 This implementation facilitates future changes to generalise sweep models to
@@
 coalescent, adding support for similar situations,
-such as inversions, is straightforward~\citep{peischl2013sequential}.
+such as inversions~\citep{peischl2013sequential}, is straightforward.
+Fig.~\ref{fig-selection-perf} shows that \msprime\ has far better
+CPU time and memory performance than \discoal. We have previously shown
+that \msprime\ is much more efficient than \msms\ and \cosiTwo\
+when simulating without selection~\citep{kelleher2016efficient}.
+% JK: do we need an extra proviso like this?
+% and this is unlikely to change unless sweeps dominate the ancestral process.
 
@@
 number of lineages is significantly less than the population size
-(the regime where the coalescent is an accurate approximat),
+(the regime where the coalescent is an accurate approximation),
 which usually happens in the quite recent

### 2021-08-08 evaluation/generate_gc_perf_data.py
Update GC section

@@ -225,6 +225,10 @@ def benchmark_ecoli(replicates, processes):
         for k, v in cpu.items():
             print(k, "\t", v, file=f)
 
+    # NOTE: if we're doing this again it would be simpler to use the *exact*
+    # estimates from Lapierre et al rather than these rough values we using
+    # now. Just easier to explain in the paper.
+
     sample_sizes = np.linspace(10, 500, 20).astype(int)
     work = []
     for sample_size in sample_sizes:

### 2021-08-08 evaluation/plot.py

@@ -68,6 +68,8 @@ def gc_perf():
         ax2.plot(dft.sample_size, dft.memory, label=tool, color=line.get_color())
         lines[tool] = line
 
+    ax1.set_title("(A)")
+    ax2.set_title("(B)")
     ax1.set_xlabel("Sample size")
     ax1.set_ylabel("Time (hours)")
     ax1.set_ylim(0, None)
@@ -115,6 +117,8 @@ def sweeps_perf():
         lines[tool] = line
         ax2.plot(dft.L, dft.memory, label=tool)
 
+    ax1.set_title("(A)")
+    ax2.set_title("(B)")
     ax1.set_xlabel("Sequence length (Kilobases)")
     ax1.set_ylabel("Time (minutes)")
     ax2.set_xlabel("Sequence length (Kilobases)")

### 2021-08-08 paper.tex

@@
 \newcommand{\ARGON}[0]{\texttt{ARGON}}
+\newcommand{\SimBac}[0]{\texttt{SimBac}}
+\newcommand{\FastSimBac}[0]{\texttt{fastSimBac}}
 
@@
 \subsection*{Gene conversion}
-\grgcomment{TODO: discuss Fig~\ref{fig-gc-perf} in text of GC section.}
-
-%introduction
-Gene conversion is a form of recombination that results in unidirectional transfer
+Gene conversion is a form of recombination that results in the transfer
 of a short segment of genetic material,
 for example between homologous chromosomes~\citep{chen2007gene}.
-%description of ancestral process
-Just as in crossover recombination, gene conversion events split how genetic material
-is inherited along some lineage, creating two independent lineages.
-In contrast to the previously described recombination process, where the ancestral
-material to the left and right of a recombination breakpoint follows two
-different ancestral lineages, gene conversion events distribute the genetic material
-of a short segment to one ancestral lineage and the part to the left and right of this
-segment to another lineage.
-%Franz Baumdicker: Not sure if this next sentence is necessary,
-% but it might be of interest for some people?
-The resulting distribution of ancestral material after a gene conversion event
-equals the distribution after two recombination events at the left and right position
-of the gene conversion tract, followed by a common ancestor event.
-%Impact on LD
-However, gene conversions impact much shorter segments than crossover recombinations,
-typically below 1kb, and thus affect linkage patterns differently to
-recombinations~\citep{korunes2017gene}.
-
-% other simulators
-Hudson's \ms\ optionally includes gene conversion and so does
-MaCS~\citep{chen2009fast}, while scrm~\citep{staab2015scrm} does not
-support gene conversions.
-As bacterial recombination can be modelled analogously to eukaryotic allelic
-gene conversion, simulators for microbial evolution often include gene conversion
-instead of recombination.
-This includes the backward simulator SimBac~\citep{brown2016simbac},
-the SMC based approach used in FastSimBac~\citep{demaio2017the},
-and forward-in-time tools within the SLiM framework~\citep{cury2020simulation}.
-
-In \msprime\ 1.0 we implemented the coalescent with gene conversion as defined by
-\cite{wiuf2000coalescent}, where gene conversion events are initiated between any
-two sites in the sequence at a constant rate $\gamma$ and
-the length of the transferred segment is given by a geometric distributed
-random variable with mean length $L$.
-% Above I used the discrete case, but we could easily switch to a description for the
-% continuous case.
+Since gene conversion impacts much shorter segments than
+crossover recombination (typically below 1kb) it
+affects patterns of linkage disequilibrium differently~\citep{korunes2017gene}.
+\cite{wiuf2000coalescent} modelled gene conversion in the coalescent
+via a rate at which gene conversion events are initiated
+along the genome and a geometrically distributed tract length.
+In terms of the ancestral process, gene conversion differs from
+crossover recombination (as described in the previous section)
+in that it extracts a short tract of ancestry into
+an independent lineage, rather than splitting ancestry
+to the left and right of a given breakpoint.
 We have implemented this model of gene
@@
 \ms\ and analytical results~\citep{wiuf2000coalescent}.
-
-% In terms of file size, as expected from the discussion above, we observe that gene
-% conversion is equivalent to double the rate of
-% crossover recombination, so that a tree sequence produced with a `recombination_rate` of $r$
-% and a `gene_conversion_rate` of $g$ is roughly the same size as a tree sequence with no gene
-% conversion and a `recombination_rate` of $r+2g$. However, because gene conversion tends to
-% affect only short regions of the genome, each gene conversion event usually creates
-% less trapped material, hence simulations using mammalian-like gene conversion parameters
-% run faster than simulations in which an equivalent amount of crossover recombination is
-% imposed.
 
@@
 \end{center}
-\caption{\label{fig-gc-perf}Comparison of the running time
-for E-coli simulations using \msprime\ and SimBac for varying
-sample sizes. The parameters used are chosen to mimic the simulation
-of \textit{E. coli} genomes. The genome length was set to 4.5 Mb, the scaled per site
-gene conversion rate was set to $N_e g = \gamma = 0.015$ and
-the mean gene conversion tract length was set to $L = 500$, which is close to
-estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
-$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
-with mean tract length $L = 542$~\citep{lapierre2016the}.}
+\caption{\label{fig-gc-perf}Comparison of simulation performance
+using \msprime, \SimBac, and \FastSimBac\ for varying sample sizes,
+with parameters roughly equivalent to the current estimate
+for \textit{E.~coli}~\citep{lapierre2016the}:
+a 4.5Mb genome with scaled gene conversion rate of 0.015 and
+a mean tract length of 500.
+We report (A) the total CPU time and (B) maximum memory usage
+averaged over 5 replicates (Intel Xeon E5-2680 CPU).
+We did not simulate \SimBac\ after the first two sample sizes because
+of the very long running times.}
+% The genome length was set to 4.5 Mb, the scaled per site
+% gene conversion rate was set to $N_e g = \gamma = 0.015$ and
+% the mean gene conversion tract length was set to $L = 500$, which is close to
+% estimated parameters for \textit{E.~coli}, i.e.\ an effective populations size
+% $N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
+% with mean tract length $L = 542$~\citep{lapierre2016the}.}
 \end{figure}
@@
 bacterial evolution,
-so we compare the performance of \msprime\ with gene conversion to
-SimBac~\citep{brown2016simbac}, a simulator developed specifically to
-simulate bacterial evolution. Although \msprime\ is missing some features
-relevant to this case, e.g.\ the simulation of circular genomes, it is far
-more efficient at simulating the basic processes.
-Furthermore, gene conversion can be considered in combination
-with other current and future features of \msprime\ 1.0,
-including varying recombination rates and complex demographies.
-
-An important avenue for future work
-will be to add further features to model bacterial
-evolution. In particular, we aim to include bacterial gene transfer and utilise
-the succinct tree sequence structure to represent the resulting ancestral gene
-transfer graph~\citep{baumdicker2014AGTG}.
+and so we compare the performance of \msprime\ with gene conversion to
+two specialised bacterial simulators,
+\SimBac~\citep{brown2016simbac} and \FastSimBac~\citep{demaio2017the}.
+Figure~\ref{fig-gc-perf}A shows that \msprime\ is far more efficient than
+both \SimBac\ and the SMC-based approximation \FastSimBac.
+Figure~\ref{fig-gc-perf}B shows that \msprime\ requires somewhat more memory
+than \FastSimBac, (as expected since \FastSimBac\ uses a left-to-right
+SMC approximation) but is still reasonably modest at around 1GiB for a simulation
+of 500 whole \emph{E.~coli} genomes.
+However, \msprime\ is currently lacking many of the specialised features
+required to model bacteria, and so an important avenue for future work
+is to add features such as circular genomes
+and bacterial gene transfer~\citep{baumdicker2014AGTG}.
+
+In terms of predicting the run time for a simulation including gene conversion,
+we recommend following the same approach as discussed in the previous section:
+run a number of simulations for short genome lengths, fit a quadratic to the
+observed CPU times, and use this to predict run time for larger simulations.
+Depending on the relative contributions of gene conversion and crossover
+recombination, this may be an over-estimate since gene conversion events
+tend to generate less trapped ancestral material than crossovers.
+Thus, simulations using mammalian-like gene conversion parameters
+may run faster than simulations in which an equivalent amount of
+crossover recombination is imposed.
+% TODO check this again!
+Since each gene conversion creates two breakpoints and a crossover
+creates only one, we expect the output tree sequence for a given rate
+of gene conversion to be roughly twice the size of the output from a simulation
+with the same rate of crossovers.
 

### 2021-08-09 paper.tex
ARG pass

@@
 \subsection*{Recombination}
+\label{sec-recombination}
 
@@
 averaged over 5 replicates (Intel Xeon E5-2680 CPU).
-We did not simulate \SimBac\ after the first two sample sizes because
+We did not run \SimBac\ beyond first two data points because
 of the very long running times.}
@@
 Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
-as a representation of the coalescent with recombination in terms of a graph.
-This formulation was complementary to Hudson's earlier
+as a representation of the coalescent with recombination
+stochastic process as a graph.
+This formulation is complementary to Hudson's earlier
 work~\citep{hudson1983properties}, and substantially increased our theoretical
@@
 vertices represent common ancestor or recombination events, and edges represent
-lineages through which ancestral material flows.
-We may consider the big ARG, in which we track lineages arising out of
-recombinations regardless of whether they carry ancestral material and
-consequently wait an exponentially large time until all sampled lineages
-coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
-which we only track genetic ancestors, and wait until all marginal trees
-have fully coalesced (like Hudson's algorithm).
+lineages. There is the ``big'' ARG, in which we track lineages arising out of
+recombinations regardless of whether they carry ancestral
+material~\citep{ethier1990two},
+% and
+% consequently wait an exponentially large time until all sampled lineages
+% coalesce into a single ancestor~\citep{ethier1990two},
+and the ``little'' ARG in which we only track genetic ancestors.
+Over time, the interpretation of the term
+has shifted away from the original definition as a stochastic
+process, to a representation of a \emph{particular} genetic ancestry
+using common ancestor and recombination
+events~\citep[e.g.][]{minichiello2006mapping}.
+More recently, ARGs have essentially come to mean any graph-like representation
+of genetic ancestry; for example,
+\cite{mathieson2020ancestry} informally define an ARG as the subset of an
+individual's pedigree from which they have inherited ancestral material.
+% Do we include this sentence? It's true, I'm sure: someone new to the field reads
+% Iain and Aylwyn's paper, and goes, hmm, this ARG thing sounds interesting,
+% let's follow up on that Griffiths reference... Maybe this is part of the
+% reason why ARGs have such a fearsome reputation for being difficult:
+% there's a basic confusion about what they actually *are*.
+Such usage is entirely reasonable, but referring readers back to the
+original Griffiths papers does little to clarify.
 
@@
 \includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
-\includegraphics[width=3.8cm]{figures/arg.pdf}
+% \includegraphics[width=3.8cm]{figures/arg.pdf}
 
@@
 sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation;
-(D) the fraction of nodes in a tree sequence that are concerned with
-extra ARG information in a tree sequence returned using \msprime's
-\texttt{record\_full\_arg} mode}
+topology ``simplified'' down to its minimal tree sequence representation.
+% (D) the fraction of nodes in a tree sequence that are concerned with
+% extra ARG information in a tree sequence returned using \msprime's
+% \texttt{record\_full\_arg} mode}
+}
 \end{figure}
 
-However, this original definition of the ARG in terms of a stochastic process
-is rarely what is meant when the term is currently used. The term has come to
-essentially mean any graph-like representation of genetic ancestry in the presence
-of recombination; for example,
-\cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material.
-Others state that an ARG is equivalent to the sequence of marginal
-% Also SARGE paper
-trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}
-
-Such
-imprecision is unfortunate, since there are many ways to encode genetic
-ancestry as a graph, and referring to the original Griffiths papers does little to
-clarify. As we see below, there are situations in which the precise information
-stored in the graph is important, and conflating a high-level
-discussion of genetic ancestry with this mathematical object is not
-helpful. The point of technical terms is to provide clarity and
-precision on what is being discussed, and it is now essentially
-impossible to know what is meant by the term ``ARG'';
-we therefore use the term ``genetic ancestry'' in high-level
-discussions.
-
 For our purposes, an ARG is a realisation of the coalescent with
-recombination, in the Griffiths (little ARG) sense.
+recombination, in the Griffiths sense.
 As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
@@
 present through the extant lineages and the ancestral material they carry
-over time. Since the ARG is a large and unwieldy object
-(Fig.~\ref{fig-arg}) we do not use it as output, but
+over time. We do not output the graph directly, but
 rather store the information required to recover the genealogical
 history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
-or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
+or ``coalescence records'' in original terminology~\citep{kelleher2016efficient}.
+This is far more efficient than outputting the simulated ARG in its entirety:
+we know that the number of ARG nodes is bounded by
+$\rho^2 \log^2 n$ for
+scaled recombination rate $\rho$ and sample size $n$
+(see the \nameref{sec-recombination} section),
+whereas the size of the tree sequence encoding is
+$O(n + \rho \log n)$~\citep{kelleher2016efficient}. This difference between a
+quadratic and a linear dependency on $\rho$ is profound, and shows why ARG
+based approaches are not suitable for large scale simulation.
 
@@
 This is done by storing extra nodes (marked with specific flags, so they
-can be easily identified later) and edges in the tree sequence.
-Fig.~\ref{fig-arg} shows that this extra ARG information
-quickly comes to dominate the output as we increase the sequence length,
-and the resulting tree sequence files are many times larger
-than the standard minimal tree sequence representation.
-
+can be easily identified later) and edges in the tree sequence
+(Fig.~\ref{fig-arg}).
 One situation in which a record of the full ARG is necessary is when we
@@
 We provide functions to compute the likelihood of ARG realisations
-(encoded using the representation of the \texttt{record\_full\_arg} option)
 and mutational patterns under
 the standard coalescent and infinite sites mutation model.
-See the Appendix for details.
+See the Appendix for details on these likelihood calculations.
 

### 2021-08-09 paper.tex
Handwaving.

@@
 or ``coalescence records'' in original terminology~\citep{kelleher2016efficient}.
-This is far more efficient than outputting the simulated ARG in its entirety:
-we know that the number of ARG nodes is bounded by
-$\rho^2 \log^2 n$ for
-scaled recombination rate $\rho$ and sample size $n$
-(see the \nameref{sec-recombination} section),
+This is far more efficient than outputting the simulated ARG in its entirety.
+For a given scaled recombination rate $\rho$
+(setting aside the dependency on the sample size $n$)
+we know from Eq.~\eqref{eqn-hsw-bound} that the number of ARG
+nodes is $O(\rho^2)$,
 whereas the size of the tree sequence encoding is
-$O(n + \rho \log n)$~\citep{kelleher2016efficient}. This difference between a
-quadratic and a linear dependency on $\rho$ is profound, and shows why ARG
-based approaches are not suitable for large scale simulation.
+$O(\rho)$~\citep{kelleher2016efficient}.
+This difference between a
+quadratic and a linear dependency on $\rho$ is profound, and shows why
+large simulations cannot output an ARG in practice.
 

### 2021-08-09 paper.tex
Updates from review

@@
 For our purposes, an ARG is a realisation of the coalescent with
-recombination, in the Griffiths sense.
+recombination, in the Griffiths (little ARG) sense.
 As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
@@
 (setting aside the dependency on the sample size $n$)
-we know from Eq.~\eqref{eqn-hsw-bound} that the number of ARG
-nodes is $O(\rho^2)$,
+we know from Eq.~\eqref{eqn-hsw-bound} that the number nodes
+in an ARG is $O(\rho^2)$,
 whereas the size of the tree sequence encoding is

### 2021-08-09 paper.tex
Update paper.tex
added my affiliation

@@
 \usepackage{graphicx}
+\usepackage{authblk}
 \usepackage{dsfont}
@@
 
-\title{Msprime 1.0: an efficient and extensible coalescent simulation framework}
-\author{Author list to be filled in
-}
-% \address{
-
-%%Affiliations:
-% Franz Baumdicker:
-% Cluster of Excellence "Controlling Microbes to Fight Infections", Mathematical and Computational Population Genetics, University of Tübingen, Germany
-
-% Consuelo D. Quinto-Cortés
-%National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico
-
-% Peter L. Ralph
-% Institute of Ecology and Evolution, Departments of Mathematics and Biology, University of Oregon, USA
-
-% \author[n]{Graham Gower}
-% \affil[n]{Lundbeck GeoGenetics Centre, Globe Institute, Faculty of Health and Medical Sciences, University of Copenhagen, Denmark}
+\title{Efficient ancestry and mutation simulation with msprime 1.0}
+% First authors
+%% FIXME affiliations are all broken
+\author[1,$\star$]{Franz Baumdicker}
+\author[2,$\star$]{Gertjan Bisschop}
+\author[3,$\star$]{Daniel Goldstein}
+\author[4,$\star$]{Graham Gower}
+\author[5,$\star$]{Aaron P. Ragsdale}
+\author[6,$\star$]{Georgia Tsambos}
+\author[7,$\star$]{Sha Zhu}
+
+% Middle Authors
+\author[8]{Bjarki Eldon}
+\author[9]{Castedo E. Ellerman}
+\author[10]{Jared G. Galloway}
+\author[11]{Ariella L. Gladstein}
+\author[12]{Gregor Gorjanc}
+\author[13]{Guo, Bing}
+\author[7]{Ben Jeffery}
+\author[14]{Warren W. Kretzschmar}
+\author[15]{Ivan Krukov}
+\author[2]{Konrad Lohse}
+\author[16]{Michael Matschiner}
+\author[15]{Dominic Nelson}
+\author[17]{Nathaniel Pope}
+\author[18]{Consuelo D. Quinto-Cort\'es}
+\author[10]{Murillo F. Rodrigues}
+\author[19]{Kumar Suanack}
+\author[20]{Thibaut Sellinger}
+\author[21]{Kevin Thornton}
+\author[22]{Hugo van Kemenade}
+\author[7]{Anthony W. Wohns}
+\author[7]{H. Yan Wong}
+
+% Senior Authors
+\author[15,$\dagger$]{Simon Gravel}
+\author[10,$\dagger$]{Andrew Kern}
+\author[23,$\dagger$]{Jere Koskela}
+\author[10,24,$\dagger$]{Peter L. Ralph}
+
+% Corresponding
+\author[7,$\ddagger$]{Jerome Kelleher}
+
+% Franz
+\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
+% Gertjan, Konrad
+\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
+% Daniel
+\affil[3]{FIXME: DANIEL'S AFFILIATION}
+% Graham
+\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
+% Simon, Dom, Ivan
+\affil[5]{FIXME AARON's AFFILIATION}
+% Georgia
+\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
+    University of Melbourne}
+% Jerome, Yan, Wilder, Joe
+\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
+    University of Oxford}
+
+% Bjarki
+\affil[8]{FIXME: BJARKI's AFFILIATION}
+% Castedo
+\affil[9]{FIXME: CASTEDO's AFFILIATION}
+% Jared, Murillo, Peter, Andy
+\affil[10]{Department of Biology and Institute of Ecology and Evolution,
+    University of Oregon}
+% Ariella
+\affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
+% Gregor
+\affil[12]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
+    University of Edinburgh}
+% Bing
+\affil[13]{FIXME: BING's AFFILIATION}
+% Winni
+\affil[14]{FIXME: WINNI's AFFILIATION}
+% Simon, Dom, Ivan
+\affil[15]{Department of Human Genetics, McGill University}
+% Michael
+\affil[16]{FIXME: MICHAEL's AFFILIATION}
+% Nathaniel
+\affil[17]{FIXME: NATHANIEL's AFFILIATION}
+% Consuelo
+\affil[18]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
+    Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
+% Saunack
+\affil[19]{FIXME: SAUNACK's AFFILIATION}
+% Thibaut
+\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+% Kevin
+\affil[21]{Ecology and Evolutionary Biology, University of California at Irvine}
+% Hugo
+\affil[22]{FIXME HUGO's AFFILIATION}
+% Jere
+\affil[23]{FIXME: JERE'S AFFILIATION}
+% Peter (second affil)
+\affil[24]{Department of Mathematics, University of Oregon}
+
+\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
+\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
+\affil[$\ddagger$]{Denotes corresponding author}
 
@@
 \subsection*{Recombination}
-\label{sec-recombination}
 
@@
 averaged over 5 replicates (Intel Xeon E5-2680 CPU).
-We did not run \SimBac\ beyond first two data points because
+We did not simulate \SimBac\ after the first two sample sizes because
 of the very long running times.}
@@
 Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
-as a representation of the coalescent with recombination
-stochastic process as a graph.
-This formulation is complementary to Hudson's earlier
+as a representation of the coalescent with recombination in terms of a graph.
+This formulation was complementary to Hudson's earlier
 work~\citep{hudson1983properties}, and substantially increased our theoretical
@@
 vertices represent common ancestor or recombination events, and edges represent
-lineages. There is the ``big'' ARG, in which we track lineages arising out of
-recombinations regardless of whether they carry ancestral
-material~\citep{ethier1990two},
-% and
-% consequently wait an exponentially large time until all sampled lineages
-% coalesce into a single ancestor~\citep{ethier1990two},
-and the ``little'' ARG in which we only track genetic ancestors.
-Over time, the interpretation of the term
-has shifted away from the original definition as a stochastic
-process, to a representation of a \emph{particular} genetic ancestry
-using common ancestor and recombination
-events~\citep[e.g.][]{minichiello2006mapping}.
-More recently, ARGs have essentially come to mean any graph-like representation
-of genetic ancestry; for example,
-\cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material.
-% Do we include this sentence? It's true, I'm sure: someone new to the field reads
-% Iain and Aylwyn's paper, and goes, hmm, this ARG thing sounds interesting,
-% let's follow up on that Griffiths reference... Maybe this is part of the
-% reason why ARGs have such a fearsome reputation for being difficult:
-% there's a basic confusion about what they actually *are*.
-Such usage is entirely reasonable, but referring readers back to the
-original Griffiths papers does little to clarify.
+lineages through which ancestral material flows.
+We may consider the big ARG, in which we track lineages arising out of
+recombinations regardless of whether they carry ancestral material and
+consequently wait an exponentially large time until all sampled lineages
+coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
+which we only track genetic ancestors, and wait until all marginal trees
+have fully coalesced (like Hudson's algorithm).
 
@@
 \includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
-% \includegraphics[width=3.8cm]{figures/arg.pdf}
+\includegraphics[width=3.8cm]{figures/arg.pdf}
 
@@
 sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation.
-% (D) the fraction of nodes in a tree sequence that are concerned with
-% extra ARG information in a tree sequence returned using \msprime's
-% \texttt{record\_full\_arg} mode}
-}
+topology ``simplified'' down to its minimal tree sequence representation;
+(D) the fraction of nodes in a tree sequence that are concerned with
+extra ARG information in a tree sequence returned using \msprime's
+\texttt{record\_full\_arg} mode}
 \end{figure}
+
+However, this original definition of the ARG in terms of a stochastic process
+is rarely what is meant when the term is currently used. The term has come to
+essentially mean any graph-like representation of genetic ancestry in the presence
+of recombination; for example,
+\cite{mathieson2020ancestry} informally define an ARG as the subset of an
+individual's pedigree from which they have inherited ancestral material.
+Others state that an ARG is equivalent to the sequence of marginal
+% Also SARGE paper
+trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}
+
+Such
+imprecision is unfortunate, since there are many ways to encode genetic
+ancestry as a graph, and referring to the original Griffiths papers does little to
+clarify. As we see below, there are situations in which the precise information
+stored in the graph is important, and conflating a high-level
+discussion of genetic ancestry with this mathematical object is not
+helpful. The point of technical terms is to provide clarity and
+precision on what is being discussed, and it is now essentially
+impossible to know what is meant by the term ``ARG'';
+we therefore use the term ``genetic ancestry'' in high-level
+discussions.
 
@@
 present through the extant lineages and the ancestral material they carry
-over time. We do not output the graph directly, but
+over time. Since the ARG is a large and unwieldy object
+(Fig.~\ref{fig-arg}) we do not use it as output, but
 rather store the information required to recover the genealogical
 history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
-or ``coalescence records'' in original terminology~\citep{kelleher2016efficient}.
-This is far more efficient than outputting the simulated ARG in its entirety.
-For a given scaled recombination rate $\rho$
-(setting aside the dependency on the sample size $n$)
-we know from Eq.~\eqref{eqn-hsw-bound} that the number nodes
-in an ARG is $O(\rho^2)$,
-whereas the size of the tree sequence encoding is
-$O(\rho)$~\citep{kelleher2016efficient}.
-This difference between a
-quadratic and a linear dependency on $\rho$ is profound, and shows why
-large simulations cannot output an ARG in practice.
+or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
 
@@
 This is done by storing extra nodes (marked with specific flags, so they
-can be easily identified later) and edges in the tree sequence
-(Fig.~\ref{fig-arg}).
+can be easily identified later) and edges in the tree sequence.
+Fig.~\ref{fig-arg} shows that this extra ARG information
+quickly comes to dominate the output as we increase the sequence length,
+and the resulting tree sequence files are many times larger
+than the standard minimal tree sequence representation.
+
 One situation in which a record of the full ARG is necessary is when we
@@
 We provide functions to compute the likelihood of ARG realisations
+(encoded using the representation of the \texttt{record\_full\_arg} option)
 and mutational patterns under
 the standard coalescent and infinite sites mutation model.
-See the Appendix for details on these likelihood calculations.
+See the Appendix for details.
 

### 2021-08-09 paper.tex
Remove itemize in likelihood appendix

--- 2021-08-09 paper.tex
@@
 \usepackage{nameref}
+\usepackage{caption,subcaption}
 \usepackage{tikz}
+\usetikzlibrary{calc,positioning}
 %\usepackage{showkeys}
@@
 
-\title{Msprime 1.0: an efficient and extensible coalescent simulation framework}
-\author{Author list to be filled in
-}
-% \address{
-
-%%Affiliations:
-% Franz Baumdicker:
-% Cluster of Excellence "Controlling Microbes to Fight Infections", Mathematical and Computational Population Genetics, University of Tübingen, Germany
-
-% Consuelo D. Quinto-Cortés
-%National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico
-
-% Peter L. Ralph
-% Institute of Ecology and Evolution, Departments of Mathematics and Biology, University of Oregon, USA
-
-% \author[n]{Graham Gower}
-% \affil[n]{Lundbeck GeoGenetics Centre, Globe Institute, Faculty of Health and Medical Sciences, University of Copenhagen, Denmark}
+\title{Efficient ancestry and mutation simulation with msprime 1.0}
+% First authors
+%% FIXME affiliations are all broken
+\author[1,$\star$]{Franz Baumdicker}
+\author[2,$\star$]{Gertjan Bisschop}
+\author[3,$\star$]{Daniel Goldstein}
+\author[4,$\star$]{Graham Gower}
+\author[5,$\star$]{Aaron P. Ragsdale}
+\author[6,$\star$]{Georgia Tsambos}
+\author[7,$\star$]{Sha Zhu}
+
+% Middle Authors
+\author[8]{Bjarki Eldon}
+\author[9]{Castedo E. Ellerman}
+\author[10]{Jared G. Galloway}
+\author[11]{Ariella L. Gladstein}
+\author[12]{Gregor Gorjanc}
+\author[13]{Guo, Bing}
+\author[7]{Ben Jeffery}
+\author[14]{Warren W. Kretzschmar}
+\author[15]{Ivan Krukov}
+\author[2]{Konrad Lohse}
+\author[16]{Michael Matschiner}
+\author[15]{Dominic Nelson}
+\author[17]{Nathaniel Pope}
+\author[18]{Consuelo D. Quinto-Cort\'es}
+\author[10]{Murillo F. Rodrigues}
+\author[19]{Kumar Suanack}
+\author[20]{Thibaut Sellinger}
+\author[21]{Kevin Thornton}
+\author[22]{Hugo van Kemenade}
+\author[7]{Anthony W. Wohns}
+\author[7]{H. Yan Wong}
+
+% Senior Authors
+\author[15,$\dagger$]{Simon Gravel}
+\author[10,$\dagger$]{Andrew D. Kern}
+\author[23,$\dagger$]{Jere Koskela}
+\author[10,24,$\dagger$]{Peter L. Ralph}
+
+% Corresponding
+\author[7,$\ddagger$]{Jerome Kelleher}
+
+% Franz
+\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
+% Gertjan, Konrad
+\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
+% Daniel
+\affil[3]{Khoury College of Computer Sciences, Northeastern University}
+% Graham
+\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
+% Simon, Dom, Ivan
+\affil[5]{FIXME AARON's AFFILIATION}
+% Georgia
+\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
+    University of Melbourne}
+% Jerome, Yan, Wilder, Joe
+\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
+    University of Oxford}
+
+% Bjarki
+\affil[8]{Leibniz Institute for Evolution and Biodiversity
+Science, Museum f\"ur Naturkunde Berlin }
+% Castedo
+\affil[9]{FIXME: CASTEDO's AFFILIATION}
+% Jared, Murillo, Peter, Andy
+\affil[10]{Institute of Ecology and Evolution, Department of Biology,
+    University of Oregon}
+% Ariella
+\affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
+% Gregor
+\affil[12]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
+    University of Edinburgh}
+% Bing
+\affil[13]{FIXME: BING's AFFILIATION}
+% Winni
+\affil[14]{FIXME: WINNI's AFFILIATION}
+% Simon, Dom, Ivan
+\affil[15]{Department of Human Genetics, McGill University}
+% Michael
+\affil[16]{FIXME: MICHAEL's AFFILIATION}
+% Nathaniel
+\affil[17]{FIXME: NATHANIEL's AFFILIATION}
+% Consuelo
+\affil[18]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
+    Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
+% Saunack
+\affil[19]{FIXME: SAUNACK's AFFILIATION}
+% Thibaut
+\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+% Kevin
+\affil[21]{Ecology and Evolutionary Biology, University of California at Irvine}
+% Hugo
+\affil[22]{FIXME HUGO's AFFILIATION}
+% Jere
+\affil[23]{Department of Statistics, University of Warwick}
+% Peter (second affil)
+\affil[24]{Department of Mathematics, University of Oregon}
+
+\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
+\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
+\affil[$\ddagger$]{Denotes corresponding author}
 
@@
 
-\textbf{Keywords:} Coalescent simulation, Python
+\textbf{Keywords:} Coalescent simulation, Python, ancestral recombination graphs
 
@@
 \begin{figure}
-
-\begin{center}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
     \begin{tikzpicture}[scale=0.45]
@@
     \end{tikzpicture}
-% \begin{center}
-% $\equiv$
-% \end{center}
-% \end{column}
-% \begin{column}{5cm}
-\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
-\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
-% \includegraphics[width=3.8cm]{figures/arg.pdf}
-
-\end{center}
+\caption{}\label{fig-arg-a}
+\end{subfigure}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
+\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.4]
+
+\tikzset{greynode/.style={circle,fill,inner sep=0.7},
+nodelabel/.style={font=\scriptsize}}
+
+% Left sample nodes
+\node (s0) [greynode] {};
+\node (s1) [right=of s0,greynode] {};
+\node (s2) [right=of s1,greynode] {};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Right sample nodes
+\node (rightTree) at (4, 0) {};
+\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
+\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
+\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
+\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Non-sample nodes
+\node [greynode] (s3) at ($(s1) + (0,1)$) {};
+\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
+\node [greynode] (s4) at ($(s2) + (0,2)$) {};
+\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
+\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
+\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
+\node [greynode] (s6) at ($0.5*(s5) + 0.5*(s4) + (0,2)$) {};
+\node [greynode] (s6r) at ($0.5*(s5r) + 0.5*(s4r) + (0,2)$) {};
+\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
+\foreach \u/\lab in {s4/4, s4r/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5, s5r/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s6/6, s6r/6} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges 
+\foreach \child/\parent in {
+	s5/s0, s5/s1, s6/s5, s6/s2,
+	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
+}
+	\draw (\child) -| (\parent);
+
+% Axes
+\draw[very thick] ($(s0) + (0,5.8)$) -- ($(s2r) + (0,5.8)$);
+\node (topAx) at (0,5.8) {};
+\node (topLeft) at ($(s0) + (0,5.8)$) {};
+\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
+\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
+\node[anchor=south] at ($(topLeft)$) {\scriptsize 0.0}; 
+\node[anchor=south] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
+\node[anchor=south] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+
+% Interval endpoints
+\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, -5.5);
+
+% Axis titles
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,1.3)$) {$\scriptsize\textrm{Genome position}$};
+\end{tikzpicture}
+\caption{}\label{fig-arg-b}
+\end{subfigure}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
+\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.4]
+
+\tikzset{greynode/.style={circle,fill,inner sep=0.7},
+nodelabel/.style={font=\scriptsize}}
+
+% Left sample nodes
+\node (s0) [greynode] {};
+\node (s1) [right=of s0,greynode] {};
+\node (s2) [right=of s1,greynode] {};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Right sample nodes
+\node (rightTree) at (4, 0) {};
+\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
+\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
+\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
+\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Non-sample nodes
+%\node [greynode] (s3) at ($(s1) + (0,1)$) {};
+%\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
+%\node [greynode] (s4) at ($(s2) + (0,2)$) {};
+\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
+\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
+%\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
+\node [greynode] (s6) at ($0.5*(s5) + 0.5*(s4) + (0,2)$) {};
+\node [greynode] (s6r) at ($0.5*(s5r) + 0.5*(s4r) + (0,2)$) {};
+%\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
+\foreach \u/\lab in {s4r/3} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/4} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s6/5, s6r/5} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges 
+\foreach \child/\parent in {
+	s5/s0, s5/s1, s6/s5, s6/s2,
+	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
+}
+	\draw (\child) -| (\parent);
+
+% Axes
+\draw[very thick] ($(s0) + (0,5.8)$) -- ($(s2r) + (0,5.8)$);
+\node (topAx) at (0,5.8) {};
+\node (topLeft) at ($(s0) + (0,5.8)$) {};
+\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
+\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
+\node[anchor=south] at ($(topLeft)$) {\scriptsize 0.0}; 
+\node[anchor=south] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
+\node[anchor=south] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+
+% Interval endpoints
+\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, -5.5);
+
+% Axis titles
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,1.3)$) {$\scriptsize\textrm{Genome position}$};
+\end{tikzpicture}
+\caption{}\label{fig-arg-c}
+\end{subfigure}
 \caption{\label{fig-arg} (A) A simple ARG in which a recombination
@@
 sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation.
-% (D) the fraction of nodes in a tree sequence that are concerned with
-% extra ARG information in a tree sequence returned using \msprime's
-% \texttt{record\_full\_arg} mode}
-}
+topology ``simplified'' down to its minimal tree sequence representation.}
 \end{figure}
@@
 (setting aside the dependency on the sample size $n$)
-we know from Eq.~\eqref{eqn-hsw-bound} that the number nodes
+we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
 in an ARG is $O(\rho^2)$,
@@
 model does break down is the combination of large sample size
-and with long recombining genomes, where the large
+and long recombining genomes, where the large
 number of recombination events in the recent past results in
@@
 with mentorship by experienced developers,
-is a hugely useful experience for many users,
+is a hugely useful experience for many users.
 The skills that contributors learn
@@
 Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
+BE was supported by  DFG grant  273887127 through Priority
+Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research
+Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
+Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of
+Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
 
@@
 \texttt{record\_full\_arg} option as their arguments.
-\begin{itemize}
-\item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
+
+The \texttt{msprime.log\_arg\_likelihood(ts, r, N)} function
+returns the natural logarithm of
 the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
@@
 we set $g_i = 1$ and interpret the result as a density.
-\item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
+
+The \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} function returns the
 natural logarithm of the probability of the mutations recorded in the tree sequence
@@
 edges.
-\end{itemize}
+% \end{itemize}
 

### 2021-08-09 paper.tex
added affiliation and funding support

--- 2021-08-09 paper.tex
@@
 % Bjarki
-\affil[8]{FIXME: BJARKI's AFFILIATION}
+\affil[8]{Leibniz Institute for Evolution and Biodiversity
+Science, Museum f\"ur Naturkunde Berlin }
 % Castedo
@@
 % Thibaut
-\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+\affil[20]{FIXME: THIBAUT's AFFILIATION}
 % Kevin
@@
 Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
+BE was supported by  DFG grant  273887127 through Priority
+Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2); BE would also like to acknowledge  funding through The Icelandic Research
+Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051.
 

### 2021-08-09 paper.tex
added affiliation and funding support
Jeres affiliation
Update paper.tex
updadet acknowls

@@
 BE was supported by  DFG grant  273887127 through Priority
-Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2); BE would also like to acknowledge  funding through The Icelandic Research
-Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051.
+Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research
+Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
 

### 2021-08-09 paper.tex
updadet acknowls

@@
 % Thibaut
-\affil[20]{FIXME: THIBAUT's AFFILIATION}
+\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
 % Kevin
@@
 % Jere
-\affil[23]{FIXME: JERE'S AFFILIATION}
+\affil[23]{Department of Statistics, University of Warwick}
 % Peter (second affil)
@@
 Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
+Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of
+Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
 

### 2021-08-09 paper.tex
checked my name / affiliation / grant acks
minor fix to UO affiliation
Tree sequence description
This pull request attempted to resolve the following concerns with this paragraph:

Diploid individuals can have *at most* two nodes *at a given locus*. 
Try to avoid using concepts not yet defined (birth time of a lineage, "individual table")
Remove redundancy on defining the node to individual correspondance
Clarify that parent - child nodes may not correspond to parent - child individuals (unless things have changed).

These all mere suggestions.

@@
 \author[15,$\dagger$]{Simon Gravel}
-\author[10,$\dagger$]{Andrew Kern}
+\author[10,$\dagger$]{Andrew D. Kern}
 \author[23,$\dagger$]{Jere Koskela}
@@
 and \emph{edges} (Fig.~\ref{fig-ts-example}).
-A node in tree sequence terminology corresponds
-to a single chromosome strand, associated with its birth time
-and other optional information. A node may be
-associated with an \emph{individual}, so that, for example, a diploid
-individual is associated with two separate nodes (we omit
-the individual table in Fig.~\ref{fig-ts-example} as it is optional).
-Nodes define haplotypes that exist at a specific time within
-some individual, and edges define genetic inheritance
-between these nodes. An \emph{edge} consists of a
+A node in tree sequence terminology corresponds to a 
+haplotype that existed within a past individual, 
+and is ancestral to the sample. 
+A diploid individual, for example, is associated with at 
+most two separate nodes spanning a given locus. 
+The birth time of the individual is recorded as a node 
+property. By contrast, recording the identity of the ancestral 
+individual is optional because it does not affect sample 
+relatedness. Edges define genetic inheritance
+between nodes. An \emph{edge} consists of a
 parent node, a child node and the left and right coordinates
 of the contiguous genomic interval over which
-the child inherited genetic material from the parent.
+the child node inherited genetic material from the parent node.
+Parent and child nodes may correspond to ancestor and descendant 
+individuals separated by 
+many generations.
+
 Critically, edges can span multiple marginal trees: for

### 2021-08-09 paper.tex
small typo fixes

--- 2021-08-09 paper.tex
@@
 Integration (CI) methodology. Other CI services check code formatting
-and for common errors, as well as producing reports on the level of
+and for common errors, as well as produce reports on the level of
 test coverage, e.g., pinpointing areas where additional unit tests may
@@
 % THERE IS A BETTER WAY!!!
-Scientific software is vital apparatus, and must be engineered
+Scientific software is a vital apparatus, and must be engineered
 to a high quality if we are to trust its results. There is a growing
@@
 Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
-Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of
-Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
 

### 2021-08-09 paper.tex
Add Daniel affiliation
edited selection section

@@
 \author[10]{Jared G. Galloway}
-\author[11]{Ariella L. Gladstein}
-\author[12]{Gregor Gorjanc}
-\author[13]{Guo, Bing}
+\author[11,12]{Ariella L. Gladstein}
+\author[13]{Gregor Gorjanc}
+\author[14]{Guo, Bing}
 \author[7]{Ben Jeffery}
-\author[14]{Warren W. Kretzschmar}
-\author[15]{Ivan Krukov}
+\author[15]{Warren W. Kretzschmar}
+\author[16]{Ivan Krukov}
 \author[2]{Konrad Lohse}
-\author[16]{Michael Matschiner}
-\author[15]{Dominic Nelson}
-\author[17]{Nathaniel Pope}
-\author[18]{Consuelo D. Quinto-Cort\'es}
+\author[17]{Michael Matschiner}
+\author[16]{Dominic Nelson}
+\author[18]{Nathaniel S. Pope}
+\author[19]{Consuelo D. Quinto-Cort\'es}
 \author[10]{Murillo F. Rodrigues}
-\author[19]{Kumar Suanack}
-\author[20]{Thibaut Sellinger}
-\author[21]{Kevin Thornton}
-\author[22]{Hugo van Kemenade}
+\author[20]{Kumar Suanack}
+\author[21]{Thibaut Sellinger}
+\author[22]{Kevin Thornton}
+\author[23]{Hugo van Kemenade}
 \author[7]{Anthony W. Wohns}
@@
 \author[15,$\dagger$]{Simon Gravel}
-\author[10,$\dagger$]{Andrew Kern}
-\author[23,$\dagger$]{Jere Koskela}
-\author[10,24,$\dagger$]{Peter L. Ralph}
+\author[10,$\dagger$]{Andrew D. Kern}
+\author[24,$\dagger$]{Jere Koskela}
+\author[10,25,$\dagger$]{Peter L. Ralph}
 
@@
 % Daniel
-\affil[3]{FIXME: DANIEL'S AFFILIATION}
+\affil[3]{Khoury College of Computer Sciences, Northeastern University}
 % Graham
@@
 % Jared, Murillo, Peter, Andy
-\affil[10]{Department of Biology and Institute of Ecology and Evolution,
+\affil[10]{Institute of Ecology and Evolution, Department of Biology,
     University of Oregon}
@@
 \affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
+\affil[12]{Embark Veterinary, Inc., Boston}
 % Gregor
-\affil[12]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
+\affil[13]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
     University of Edinburgh}
 % Bing
-\affil[13]{FIXME: BING's AFFILIATION}
+\affil[14]{FIXME: BING's AFFILIATION}
 % Winni
-\affil[14]{FIXME: WINNI's AFFILIATION}
+\affil[15]{FIXME: WINNI's AFFILIATION}
 % Simon, Dom, Ivan
-\affil[15]{Department of Human Genetics, McGill University}
+\affil[16]{Department of Human Genetics, McGill University}
 % Michael
-\affil[16]{FIXME: MICHAEL's AFFILIATION}
+\affil[17]{Natural History Museum, University of Oslo}
 % Nathaniel
-\affil[17]{FIXME: NATHANIEL's AFFILIATION}
+\affil[18]{Department of Entomology, Pennsylvania State University}
 % Consuelo
-\affil[18]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
+\affil[19]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
     Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
 % Saunack
-\affil[19]{FIXME: SAUNACK's AFFILIATION}
+\affil[20]{FIXME: SAUNACK's AFFILIATION}
 % Thibaut
-\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+\affil[21]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
 % Kevin
-\affil[21]{Ecology and Evolutionary Biology, University of California at Irvine}
+\affil[22]{Ecology and Evolutionary Biology, University of California Irvine}
 % Hugo
-\affil[22]{FIXME HUGO's AFFILIATION}
+\affil[23]{FIXME HUGO's AFFILIATION}
 % Jere
-\affil[23]{Department of Statistics, University of Warwick}
+\affil[24]{Department of Statistics, University of Warwick}
 % Peter (second affil)
-\affil[24]{Department of Mathematics, University of Oregon}
+\affil[25]{Department of Mathematics, University of Oregon}
 
@@
     for various values of effective population size ($N_e$),
-    genome length in Morgans ($L$), and number of samples ($n$).
+    chromosome length in Morgans ($L$), and number of samples ($n$).
     Run time scales quadratically with the product of $N_e$ and $L$,
@@
     would take about 2 seconds, and (equivalently) that the same number of
-    0.01 Morgan segments with $N_e= 200,000$ would take the same time.
+    0.01 Morgan segments with $N_e=200,000$ would take the same time.
     Since recombination rate in these simulations was $10^{-8}$,
@@
 $\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
-(the sum is a harmonic number, and approximated by $\log{n}$)
-is not well supported by the data except possibly at very
-large values of $\rho$, and so it seems a different dependence
+(the sum is a harmonic number and can be approximated by $\log{n}$)
+is not well supported by observed runtimes except possibly at very
+large values of $\rho$. So it seems a different dependence
 on $n$ would be required to accurately predict simulation time for a
@@
 the \emph{Drosophila melanogaster} chromosome 2L
-is about 23.5Mb long and an average recombination rate of around $2.4 \times 10^{-8}$,
+is about 23.5Mb long with an average recombination rate of around $2.4 \times 10^{-8}$,
 so $L \approx 0.57$,
@@
 vertices represent common ancestor or recombination events, and edges represent
-lineages through which ancestral material flows.
-We may consider the big ARG, in which we track lineages arising out of
-recombinations regardless of whether they carry ancestral material and
-consequently wait an exponentially large time until all sampled lineages
-coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
-which we only track genetic ancestors, and wait until all marginal trees
-have fully coalesced (like Hudson's algorithm).
+lineages. There is the ``big'' ARG, in which we track lineages arising out of
+recombinations regardless of whether they carry ancestral
+material~\citep{ethier1990two},
+% and
+% consequently wait an exponentially large time until all sampled lineages
+% coalesce into a single ancestor~\citep{ethier1990two},
+and the ``little'' ARG in which we only track genetic ancestors.
+Over time, the interpretation of the term
+has shifted away from the original definition as a stochastic
+process, to a representation of a \emph{particular} genetic ancestry
+using common ancestor and recombination
+events~\citep[e.g.][]{minichiello2006mapping}.
+More recently, ARGs have essentially come to mean any graph-like representation
+of genetic ancestry; for example,
+\cite{mathieson2020ancestry} informally define an ARG as the subset of an
+individual's pedigree from which they have inherited ancestral material.
+% Do we include this sentence? It's true, I'm sure: someone new to the field reads
+% Iain and Aylwyn's paper, and goes, hmm, this ARG thing sounds interesting,
+% let's follow up on that Griffiths reference... Maybe this is part of the
+% reason why ARGs have such a fearsome reputation for being difficult:
+% there's a basic confusion about what they actually *are*.
+Such usage is entirely reasonable, but referring readers back to the
+original Griffiths papers does little to clarify.
 
 \begin{figure}
-
-\begin{center}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
     \begin{tikzpicture}[scale=0.45]
@@
     \end{tikzpicture}
-% \begin{center}
-% $\equiv$
-% \end{center}
-% \end{column}
-% \begin{column}{5cm}
-\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
-\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
-\includegraphics[width=3.8cm]{figures/arg.pdf}
-
-\end{center}
+\caption{}\label{fig-arg-a}
+\end{subfigure}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
+\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.4]
+
+\tikzset{greynode/.style={circle,fill,inner sep=0.7},
+nodelabel/.style={font=\scriptsize}}
+
+% Left sample nodes
+\node (s0) [greynode] {};
+\node (s1) [right=of s0,greynode] {};
+\node (s2) [right=of s1,greynode] {};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Right sample nodes
+\node (rightTree) at (4, 0) {};
+\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
+\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
+\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
+\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Non-sample nodes
+\node [greynode] (s3) at ($(s1) + (0,1)$) {};
+\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
+\node [greynode] (s4) at ($(s2) + (0,2)$) {};
+\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
+\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
+\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
+\node [greynode] (s6) at ($0.5*(s5) + 0.5*(s4) + (0,2)$) {};
+\node [greynode] (s6r) at ($0.5*(s5r) + 0.5*(s4r) + (0,2)$) {};
+\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
+\foreach \u/\lab in {s4/4, s4r/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5, s5r/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s6/6, s6r/6} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\foreach \child/\parent in {
+	s5/s0, s5/s1, s6/s5, s6/s2,
+	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
+}
+	\draw (\child) -| (\parent);
+
+% Axes
+\draw[very thick] ($(s0) + (0,5.8)$) -- ($(s2r) + (0,5.8)$);
+\node (topAx) at (0,5.8) {};
+\node (topLeft) at ($(s0) + (0,5.8)$) {};
+\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
+\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
+\node[anchor=south] at ($(topLeft)$) {\scriptsize 0.0};
+\node[anchor=south] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
+\node[anchor=south] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+
+% Interval endpoints
+\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, -5.5);
+
+% Axis titles
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,1.3)$) {$\scriptsize\textrm{Genome position}$};
+\end{tikzpicture}
+\caption{}\label{fig-arg-b}
+\end{subfigure}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
+\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.4]
+
+\tikzset{greynode/.style={circle,fill,inner sep=0.7},
+nodelabel/.style={font=\scriptsize}}
+
+% Left sample nodes
+\node (s0) [greynode] {};
+\node (s1) [right=of s0,greynode] {};
+\node (s2) [right=of s1,greynode] {};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Right sample nodes
+\node (rightTree) at (4, 0) {};
+\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
+\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
+\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
+\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Non-sample nodes
+%\node [greynode] (s3) at ($(s1) + (0,1)$) {};
+%\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
+%\node [greynode] (s4) at ($(s2) + (0,2)$) {};
+\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
+\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
+%\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
+\node [greynode] (s6) at ($0.5*(s5) + 0.5*(s4) + (0,2)$) {};
+\node [greynode] (s6r) at ($0.5*(s5r) + 0.5*(s4r) + (0,2)$) {};
+%\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
+\foreach \u/\lab in {s4r/3} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/4} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s6/5, s6r/5} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\foreach \child/\parent in {
+	s5/s0, s5/s1, s6/s5, s6/s2,
+	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
+}
+	\draw (\child) -| (\parent);
+
+% Axes
+\draw[very thick] ($(s0) + (0,5.8)$) -- ($(s2r) + (0,5.8)$);
+\node (topAx) at (0,5.8) {};
+\node (topLeft) at ($(s0) + (0,5.8)$) {};
+\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
+\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
+\node[anchor=south] at ($(topLeft)$) {\scriptsize 0.0};
+\node[anchor=south] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
+\node[anchor=south] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+
+% Interval endpoints
+\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, -5.5);
+
+% Axis titles
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,1.3)$) {$\scriptsize\textrm{Genome position}$};
+\end{tikzpicture}
+\caption{}\label{fig-arg-c}
+\end{subfigure}
 \caption{\label{fig-arg} (A) A simple ARG in which a recombination
@@
 sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation;
-(D) the fraction of nodes in a tree sequence that are concerned with
-extra ARG information in a tree sequence returned using \msprime's
-\texttt{record\_full\_arg} mode}
+topology ``simplified'' down to its minimal tree sequence representation.}
 \end{figure}
-
-However, this original definition of the ARG in terms of a stochastic process
-is rarely what is meant when the term is currently used. The term has come to
-essentially mean any graph-like representation of genetic ancestry in the presence
-of recombination; for example,
-\cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material.
-Others state that an ARG is equivalent to the sequence of marginal
-% Also SARGE paper
-trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}
-
-Such
-imprecision is unfortunate, since there are many ways to encode genetic
-ancestry as a graph, and referring to the original Griffiths papers does little to
-clarify. As we see below, there are situations in which the precise information
-stored in the graph is important, and conflating a high-level
-discussion of genetic ancestry with this mathematical object is not
-helpful. The point of technical terms is to provide clarity and
-precision on what is being discussed, and it is now essentially
-impossible to know what is meant by the term ``ARG'';
-we therefore use the term ``genetic ancestry'' in high-level
-discussions.
 
@@
 present through the extant lineages and the ancestral material they carry
-over time. Since the ARG is a large and unwieldy object
-(Fig.~\ref{fig-arg}) we do not use it as output, but
+over time. We do not output the graph directly, but
 rather store the information required to recover the genealogical
 history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
-or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
+or ``coalescence records'' in the original terminology~\citep{kelleher2016efficient}.
+This is far more efficient than outputting the simulated ARG in its entirety.
+For a given scaled recombination rate $\rho$
+(setting aside the dependency on the sample size $n$)
+we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
+in an ARG is $O(\rho^2)$,
+whereas the size of the tree sequence encoding is
+$O(\rho)$~\citep{kelleher2016efficient}.
+This difference between a
+quadratic and a linear dependency on $\rho$ is profound, and shows why
+large simulations cannot output an ARG in practice.
 
@@
 replicates for 10 samples in a model
-with \jkcomment{[ADK: fill in the details here please]} and a
-per-base recombination rate of $10^{-9}$.}
+with a single selective sweep in its history where the benefical
+allele had a scaled selection coefficient of $2Ns=1000$,
+the per-base recombination rate of $10^{-9}$, and we allowed
+the sequence length to vary between 1kb-200kb.}
 \end{figure}
@@
 that \msprime\ is much more efficient than \msms\ and \cosiTwo\
-when simulating without selection~\citep{kelleher2016efficient}.
+when simulating without selection~\citep{kelleher2016efficient},
+thus we believe \msprime\ to be the highest performing
+coalescent simulator that allows for sweeps currently.
 % JK: do we need an extra proviso like this?
@@
 Integration (CI) methodology. Other CI services check code formatting
-and for common errors, as well as produce reports on the level of
+and for common errors, as well as producing reports on the level of
 test coverage, e.g., pinpointing areas where additional unit tests may
@@
 % THERE IS A BETTER WAY!!!
-Scientific software is a vital apparatus, and must be engineered
+Scientific software is vital apparatus, and must be engineered
 to a high quality if we are to trust its results. There is a growing

### 2021-08-09 paper.tex
edited selection section

--- 2021-08-09 paper.tex
@@
 and \emph{edges} (Fig.~\ref{fig-ts-example}).
-A node in tree sequence terminology corresponds to a 
-haplotype that existed within a past individual, 
-and is ancestral to the sample. 
-A diploid individual, for example, is associated with at 
-most two separate nodes spanning a given locus. 
-The birth time of the individual is recorded as a node 
-property. By contrast, recording the identity of the ancestral 
-individual is optional because it does not affect sample 
-relatedness. Edges define genetic inheritance
-between nodes. An \emph{edge} consists of a
+A node in tree sequence terminology corresponds
+to a single chromosome strand, associated with its birth time
+and other optional information. A node may be
+associated with an \emph{individual}, so that, for example, a diploid
+individual is associated with two separate nodes (we omit
+the individual table in Fig.~\ref{fig-ts-example} as it is optional).
+Nodes define haplotypes that exist at a specific time within
+some individual, and edges define genetic inheritance
+between these nodes. An \emph{edge} consists of a
 parent node, a child node and the left and right coordinates
 of the contiguous genomic interval over which
-the child node inherited genetic material from the parent node.
-Parent and child nodes may correspond to ancestor and descendant 
-individuals separated by 
-many generations.
-
+the child inherited genetic material from the parent.
 Critically, edges can span multiple marginal trees: for

### 2021-08-10 paper.tex
some further edits to first half of paper

--- 2021-08-09 paper.tex
@@
 \newcommand{\grgcomment}[1]{\textcolor{yellow!60!red}{#1}}
+\newcommand{\adkcomment}[1]{\textcolor{blue}{#1}}
 
@@
 % First authors
-%% FIXME affiliations are all broken
 \author[1,$\star$]{Franz Baumdicker}
@@
 over other simulators~\citep{kelleher2016efficient}
-is its usage of the ``succinct tree sequence''
+is its use of the ``succinct tree sequence''
  data structure to represent simulation results.
 The succinct tree sequence (usually abbreviated to ``tree sequence'')
-is a data structure for concisely encoding
-genetic ancestry and sequence variation.
-The basic concepts were introduced by \cite{kelleher2016efficient}
-and originally developed as part of \msprime. We subsequently
+was introduced by \cite{kelleher2016efficient}
+to concisely encode genetic ancestry and sequence variation
+and was originally implemented as part of \msprime.
+We subsequently
 extracted the tree sequence code that was independent of simulations
@@
 A node in tree sequence terminology corresponds
-to a single chromosome strand, associated with its birth time
-and other optional information. A node may be
-associated with an \emph{individual}, so that, for example, a diploid
-individual is associated with two separate nodes (we omit
+\adkcomment{i found using 'single chromosome strand' here
+and then later 'haplotype' confusing, so edited}
+to a single chromosome, associated with its birth time
+and other optional information. For instance, a node may be
+associated with an \emph{individual}, so that, a diploid
+individual is composed of two separate nodes (we omit
 the individual table in Fig.~\ref{fig-ts-example} as it is optional).
-Nodes define haplotypes that exist at a specific time within
-some individual, and edges define genetic inheritance
-between these nodes. An \emph{edge} consists of a
-parent node, a child node and the left and right coordinates
+While nodes define chromosomal sequences that exist at a specific time,
+edges define genetic inheritance between these nodes. 
+An \emph{edge} contains information about the
+parent node, the child node and the left and right coordinates
 of the contiguous genomic interval over which
@@
 when simulating without selection~\citep{kelleher2016efficient},
-thus we believe \msprime\ to be the highest performing
-coalescent simulator that allows for sweeps currently.
-% JK: do we need an extra proviso like this?
-% and this is unlikely to change unless sweeps dominate the ancestral process.
+thus we believe \msprime\ to be the most efficient
+coalescent simulator when simulating sweeps.
 

### 2021-08-10 paper.tex
review fixes.

--- 2021-08-09 paper.tex
@@
 \usepackage{graphicx}
-\usepackage{authblk}
 \usepackage{dsfont}
@@
 
-\title{Efficient ancestry and mutation simulation with msprime 1.0}
-% First authors
-%% FIXME affiliations are all broken
-\author[1,$\star$]{Franz Baumdicker}
-\author[2,$\star$]{Gertjan Bisschop}
-\author[3,$\star$]{Daniel Goldstein}
-\author[4,$\star$]{Graham Gower}
-\author[5,$\star$]{Aaron P. Ragsdale}
-\author[6,$\star$]{Georgia Tsambos}
-\author[7,$\star$]{Sha Zhu}
-
-% Middle Authors
-\author[8]{Bjarki Eldon}
-\author[9]{Castedo E. Ellerman}
-\author[10]{Jared G. Galloway}
-\author[11]{Ariella L. Gladstein}
-\author[12]{Gregor Gorjanc}
-\author[13]{Guo, Bing}
-\author[7]{Ben Jeffery}
-\author[14]{Warren W. Kretzschmar}
-\author[15]{Ivan Krukov}
-\author[2]{Konrad Lohse}
-\author[16]{Michael Matschiner}
-\author[15]{Dominic Nelson}
-\author[17]{Nathaniel Pope}
-\author[18]{Consuelo D. Quinto-Cort\'es}
-\author[10]{Murillo F. Rodrigues}
-\author[19]{Kumar Suanack}
-\author[20]{Thibaut Sellinger}
-\author[21]{Kevin Thornton}
-\author[22]{Hugo van Kemenade}
-\author[7]{Anthony W. Wohns}
-\author[7]{H. Yan Wong}
-
-% Senior Authors
-\author[15,$\dagger$]{Simon Gravel}
-\author[10,$\dagger$]{Andrew D. Kern}
-\author[23,$\dagger$]{Jere Koskela}
-\author[10,24,$\dagger$]{Peter L. Ralph}
-
-% Corresponding
-\author[7,$\ddagger$]{Jerome Kelleher}
-
-% Franz
-\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
-% Gertjan, Konrad
-\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
-% Daniel
-\affil[3]{FIXME: DANIEL'S AFFILIATION}
-% Graham
-\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
-% Simon, Dom, Ivan
-\affil[5]{FIXME AARON's AFFILIATION}
-% Georgia
-\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
-    University of Melbourne}
-% Jerome, Yan, Wilder, Joe
-\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
-    University of Oxford}
-
-% Bjarki
-\affil[8]{Leibniz Institute for Evolution and Biodiversity
-Science, Museum f\"ur Naturkunde Berlin }
-% Castedo
-\affil[9]{FIXME: CASTEDO's AFFILIATION}
-% Jared, Murillo, Peter, Andy
-\affil[10]{Department of Biology and Institute of Ecology and Evolution,
-    University of Oregon}
-% Ariella
-\affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
-% Gregor
-\affil[12]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
-    University of Edinburgh}
-% Bing
-\affil[13]{FIXME: BING's AFFILIATION}
-% Winni
-\affil[14]{FIXME: WINNI's AFFILIATION}
-% Simon, Dom, Ivan
-\affil[15]{Department of Human Genetics, McGill University}
-% Michael
-\affil[16]{FIXME: MICHAEL's AFFILIATION}
-% Nathaniel
-\affil[17]{FIXME: NATHANIEL's AFFILIATION}
-% Consuelo
-\affil[18]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
-    Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
-% Saunack
-\affil[19]{FIXME: SAUNACK's AFFILIATION}
-% Thibaut
-\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
-% Kevin
-\affil[21]{Ecology and Evolutionary Biology, University of California at Irvine}
-% Hugo
-\affil[22]{FIXME HUGO's AFFILIATION}
-% Jere
-\affil[23]{Department of Statistics, University of Warwick}
-% Peter (second affil)
-\affil[24]{Department of Mathematics, University of Oregon}
-
-\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
-\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
-\affil[$\ddagger$]{Denotes corresponding author}
+\title{Msprime 1.0: an efficient and extensible coalescent simulation framework}
+\author{Author list to be filled in
+}
+% \address{
+
+%%Affiliations:
+% Franz Baumdicker:
+% Cluster of Excellence "Controlling Microbes to Fight Infections", Mathematical and Computational Population Genetics, University of Tübingen, Germany
+
+% Consuelo D. Quinto-Cortés
+%National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico
+
+% Peter L. Ralph
+% Institute of Ecology and Evolution, Departments of Mathematics and Biology, University of Oregon, USA
+
+% \author[n]{Graham Gower}
+% \affil[n]{Lundbeck GeoGenetics Centre, Globe Institute, Faculty of Health and Medical Sciences, University of Copenhagen, Denmark}
 
@@
 
-\textbf{Keywords:} Coalescent simulation, Python, ancestral recombination graphs
+\textbf{Keywords:} Coalescent simulation, Python
 
@@
 sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation;
-(D) the fraction of nodes in a tree sequence that are concerned with
-extra ARG information in a tree sequence returned using \msprime's
-\texttt{record\_full\_arg} mode}
+topology ``simplified'' down to its minimal tree sequence representation.
+% (D) the fraction of nodes in a tree sequence that are concerned with
+% extra ARG information in a tree sequence returned using \msprime's
+% \texttt{record\_full\_arg} mode}
+}
 \end{figure}
-
-However, this original definition of the ARG in terms of a stochastic process
-is rarely what is meant when the term is currently used. The term has come to
-essentially mean any graph-like representation of genetic ancestry in the presence
-of recombination; for example,
-\cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material.
-Others state that an ARG is equivalent to the sequence of marginal
-% Also SARGE paper
-trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}
-
-Such
-imprecision is unfortunate, since there are many ways to encode genetic
-ancestry as a graph, and referring to the original Griffiths papers does little to
-clarify. As we see below, there are situations in which the precise information
-stored in the graph is important, and conflating a high-level
-discussion of genetic ancestry with this mathematical object is not
-helpful. The point of technical terms is to provide clarity and
-precision on what is being discussed, and it is now essentially
-impossible to know what is meant by the term ``ARG'';
-we therefore use the term ``genetic ancestry'' in high-level
-discussions.
 
@@
 present through the extant lineages and the ancestral material they carry
-over time. Since the ARG is a large and unwieldy object
-(Fig.~\ref{fig-arg}) we do not use it as output, but
+over time. We do not output the graph directly, but
 rather store the information required to recover the genealogical
 history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
-or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
+or ``coalescence records'' in original terminology~\citep{kelleher2016efficient}.
+This is far more efficient than outputting the simulated ARG in its entirety.
+For a given scaled recombination rate $\rho$
+(setting aside the dependency on the sample size $n$)
+we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
+in an ARG is $O(\rho^2)$,
+whereas the size of the tree sequence encoding is
+$O(\rho)$~\citep{kelleher2016efficient}.
+This difference between a
+quadratic and a linear dependency on $\rho$ is profound, and shows why
+large simulations cannot output an ARG in practice.
 
@@
 replicates for 10 samples in a model
-with a single selective sweep in its history where the benefical
-allele had a scaled selection coefficient of $2Ns=1000$,
-the per-base recombination rate of $10^{-9}$, and we allowed
-the sequence length to vary between 1kb-200kb.}
+with \jkcomment{[ADK: fill in the details here please]} and a
+per-base recombination rate of $10^{-9}$.}
 \end{figure}
@@
 that \msprime\ is much more efficient than \msms\ and \cosiTwo\
-when simulating without selection~\citep{kelleher2016efficient},
-thus we believe \msprime\ to be the highest performing
-coalescent simulator that allows for sweeps currently.
+when simulating without selection~\citep{kelleher2016efficient}.
 % JK: do we need an extra proviso like this?
@@
 Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
-BE was supported by  DFG grant  273887127 through Priority
-Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research
-Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
-Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of
-Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
 

### 2021-08-10 paper.tex

@@
 
-\title{Msprime 1.0: an efficient and extensible coalescent simulation framework}
-\author{Author list to be filled in
-}
-% \address{
-
-%%Affiliations:
-% Franz Baumdicker:
-% Cluster of Excellence "Controlling Microbes to Fight Infections", Mathematical and Computational Population Genetics, University of Tübingen, Germany
-
-% Consuelo D. Quinto-Cortés
-%National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico
-
-% Peter L. Ralph
-% Institute of Ecology and Evolution, Departments of Mathematics and Biology, University of Oregon, USA
-
-% \author[n]{Graham Gower}
-% \affil[n]{Lundbeck GeoGenetics Centre, Globe Institute, Faculty of Health and Medical Sciences, University of Copenhagen, Denmark}
+\title{Efficient ancestry and mutation simulation with msprime 1.0}
+% First authors
+%% FIXME affiliations are all broken
+\author[1,$\star$]{Franz Baumdicker}
+\author[2,$\star$]{Gertjan Bisschop}
+\author[3,$\star$]{Daniel Goldstein}
+\author[4,$\star$]{Graham Gower}
+\author[5,$\star$]{Aaron P. Ragsdale}
+\author[6,$\star$]{Georgia Tsambos}
+\author[7,$\star$]{Sha Zhu}
+
+% Middle Authors
+\author[8]{Bjarki Eldon}
+\author[9]{Castedo E. Ellerman}
+\author[10]{Jared G. Galloway}
+\author[11]{Ariella L. Gladstein}
+\author[12]{Gregor Gorjanc}
+\author[13]{Guo, Bing}
+\author[7]{Ben Jeffery}
+\author[14]{Warren W. Kretzschmar}
+\author[15]{Ivan Krukov}
+\author[2]{Konrad Lohse}
+\author[16]{Michael Matschiner}
+\author[15]{Dominic Nelson}
+\author[17]{Nathaniel Pope}
+\author[18]{Consuelo D. Quinto-Cort\'es}
+\author[10]{Murillo F. Rodrigues}
+\author[19]{Kumar Suanack}
+\author[20]{Thibaut Sellinger}
+\author[21]{Kevin Thornton}
+\author[22]{Hugo van Kemenade}
+\author[7]{Anthony W. Wohns}
+\author[7]{H. Yan Wong}
+
+% Senior Authors
+\author[15,$\dagger$]{Simon Gravel}
+\author[10,$\dagger$]{Andrew D. Kern}
+\author[23,$\dagger$]{Jere Koskela}
+\author[10,24,$\dagger$]{Peter L. Ralph}
+
+% Corresponding
+\author[7,$\ddagger$]{Jerome Kelleher}
+
+% Franz
+\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
+% Gertjan, Konrad
+\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
+% Daniel
+\affil[3]{FIXME: DANIEL'S AFFILIATION}
+% Graham
+\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
+% Simon, Dom, Ivan
+\affil[5]{FIXME AARON's AFFILIATION}
+% Georgia
+\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
+    University of Melbourne}
+% Jerome, Yan, Wilder, Joe
+\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
+    University of Oxford}
+
+% Bjarki
+\affil[8]{Leibniz Institute for Evolution and Biodiversity
+Science, Museum f\"ur Naturkunde Berlin }
+% Castedo
+\affil[9]{FIXME: CASTEDO's AFFILIATION}
+% Jared, Murillo, Peter, Andy
+\affil[10]{Department of Biology and Institute of Ecology and Evolution,
+    University of Oregon}
+% Ariella
+\affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
+% Gregor
+\affil[12]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
+    University of Edinburgh}
+% Bing
+\affil[13]{FIXME: BING's AFFILIATION}
+% Winni
+\affil[14]{FIXME: WINNI's AFFILIATION}
+% Simon, Dom, Ivan
+\affil[15]{Department of Human Genetics, McGill University}
+% Michael
+\affil[16]{FIXME: MICHAEL's AFFILIATION}
+% Nathaniel
+\affil[17]{FIXME: NATHANIEL's AFFILIATION}
+% Consuelo
+\affil[18]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
+    Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
+% Saunack
+\affil[19]{FIXME: SAUNACK's AFFILIATION}
+% Thibaut
+\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+% Kevin
+\affil[21]{Ecology and Evolutionary Biology, University of California at Irvine}
+% Hugo
+\affil[22]{FIXME HUGO's AFFILIATION}
+% Jere
+\affil[23]{Department of Statistics, University of Warwick}
+% Peter (second affil)
+\affil[24]{Department of Mathematics, University of Oregon}
+
+\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
+\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
+\affil[$\ddagger$]{Denotes corresponding author}
 

### 2021-08-10 paper.tex
Apply suggestions from code review

--- 2021-08-10 paper.tex
@@
 \author[15,$\dagger$]{Simon Gravel}
-\author[10,$\dagger$]{Andrew D. Kern}
+\author[10,$\dagger$]{Andrew Kern}
 \author[23,$\dagger$]{Jere Koskela}
@@
 present through the extant lineages and the ancestral material they carry
-over time. We do not output the graph directly, but
+over time. Since the ARG is a large and unwieldy object
+(Fig.~\ref{fig-arg}) we do not use it as output, but
 rather store the information required to recover the genealogical
 history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
-or ``coalescence records'' in original terminology~\citep{kelleher2016efficient}.
-This is far more efficient than outputting the simulated ARG in its entirety.
-For a given scaled recombination rate $\rho$
-(setting aside the dependency on the sample size $n$)
-we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
-in an ARG is $O(\rho^2)$,
-whereas the size of the tree sequence encoding is
-$O(\rho)$~\citep{kelleher2016efficient}.
-This difference between a
-quadratic and a linear dependency on $\rho$ is profound, and shows why
-large simulations cannot output an ARG in practice.
+or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
 

### 2021-08-10 paper.tex

@@
 \author[10,$\dagger$]{Andrew D. Kern}
-\author[23,$\dagger$]{Jere Koskela}
-\author[10,24,$\dagger$]{Peter L. Ralph}
+\author[24,$\dagger$]{Jere Koskela}
+\author[10,25,$\dagger$]{Peter L. Ralph}
 

### 2021-08-10 paper.tex
ARG section typo

@@
 history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
-or ``coalescence records'' in original terminology~\citep{kelleher2016efficient}.
+or ``coalescence records'' in the original terminology~\citep{kelleher2016efficient}.
 This is far more efficient than outputting the simulated ARG in its entirety.

### 2021-08-10 paper.tex
Tweak to selection section

@@
 that \msprime\ is much more efficient than \msms\ and \cosiTwo\
-when simulating without selection~\citep{kelleher2016efficient}.
-% JK: do we need an extra proviso like this?
-% and this is unlikely to change unless sweeps dominate the ancestral process.
+when simulating without selection~\citep{kelleher2016efficient},
+thus we believe \msprime\ to be the most efficient
+coalescent simulator when simulating sweeps.
 

### 2021-08-10 paper.tex

@@
 over other simulators~\citep{kelleher2016efficient}
-is its usage of the ``succinct tree sequence''
+is its use of the ``succinct tree sequence''
 data structure to represent simulation results.
 The succinct tree sequence (usually abbreviated to ``tree sequence'')
-is a data structure for concisely encoding
-genetic ancestry and sequence variation.
-The basic concepts were introduced by \cite{kelleher2016efficient}
-and originally developed as part of \msprime. We subsequently
+was introduced by \cite{kelleher2016efficient}
+to concisely encode genetic ancestry and sequence variation
+and was originally implemented as part of \msprime.
+We subsequently
 extracted the tree sequence code that was independent of simulations
@@
 % Encoding ancestry
-Given a set of sample chromosomes, the topology
+Before we define tree sequences we require so additional terminology.
+Let a ``ploid'' be the
+% JK this isn't right either, but it's a start.
+% We should add a sentence clarifying what an individual is here.
+% FIXME
+complete set of genetic material that a child inherits from a parent,
+so that, e.g., each human has two ploids~\citep{jouganous2017inferring}.
+Then, given a set of sample ploids
+the structure
 of their genetic ancestry is described by a set of \emph{nodes}
 and \emph{edges} (Fig.~\ref{fig-ts-example}).
-A node in tree sequence terminology corresponds
-to a single chromosome strand, associated with its birth time
-and other optional information. A node may be
-associated with an \emph{individual}, so that, for example, a diploid
-individual is associated with two separate nodes (we omit
-the individual table in Fig.~\ref{fig-ts-example} as it is optional).
-Nodes define haplotypes that exist at a specific time within
-some individual, and edges define genetic inheritance
-between these nodes. An \emph{edge} consists of a
+A node in tree sequence terminology corresponds to a ploid,
+associated with its birth time
+and other optional information.
+% JK: Now that we have ploids, maybe it doesn't help to talk about
+% the individual table?
+% Nodes may be
+% associated with an \emph{individual}, so that, a diploid
+% individual is composed of two separate nodes (we omit
+% the individual table in Fig.~\ref{fig-ts-example} as it is optional).
+% While nodes define chromosomal sequences that exist at a specific time,
+% edges define genetic inheritance between these nodes.
+Edges then define the genetic inheritence between nodes.
+An \emph{edge} consists of a
 parent node, a child node and the left and right coordinates
 of the contiguous genomic interval over which
-the child inherited genetic material from the parent.
+the child node inherited genetic material from the parent node.
+Parent and child nodes may correspond to ancestor and descendant
+ploids separated by many generations.
+
 Critically, edges can span multiple marginal trees: for

### 2021-08-11 paper.tex
aaron's affiliation
Minor edits in recombination section

@@
 \affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
-% Simon, Dom, Ivan
-\affil[5]{FIXME AARON's AFFILIATION}
+% Aaron
+\affil[5]{Department of Integrative Biology, University of Wisconsin--Madison}
 % Georgia
@@
 the increase and decrease in numbers of lineages due to
-common ancestor and recombination events balance out;
-to get an idea of run time we can estimate when this occurs.
+common ancestor and recombination events balance out.
+To get an idea of run time we can estimate when this balance occurs.
 Suppose that the maximum number of lineages is $M$;
@@
 
-What does this tell us about time-varying population sizes?
+What does this tell us about run time for simulating time-varying population sizes?
 The argument above implies that the work is spread out relatively
@@
 Suppose that population size today is $N_1$,
-while longer ago than $T$ generations ago it was $N_2$:
-does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
+while $T$ generations ago it was $N_2$.
+Does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
 The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,

### 2021-08-11 paper.tex
Adding AlphaSimR link/citation

@@
 an existing file, and because of its efficient data interchange mechanisms,
-there is no performance penalty for performing additional operations
+there is no performance penalty for additional operations
 in a different Python package.
@@
 retained---at scale---is unprecedented.
+This interoperability is indicating an opportunity for
+other forward genetic simulators ~\citep[e.g.][]{Gaynor2021AlphaSimR}
+to leverage tree sequence data format and associated tools.
 

### 2021-08-11 paper.tex
Ploids -> genomes

@@
 \author[13]{Gregor Gorjanc}
-\author[14]{Guo, Bing}
+\author[14]{Bing Guo}
 \author[7]{Ben Jeffery}
@@
 % Bing
-\affil[14]{FIXME: BING's AFFILIATION}
+\affil[14]{Institute for Genome Sciences, University of Maryland School of
+Medicine, Baltimore, MD}
 % Winni
-\affil[15]{FIXME: WINNI's AFFILIATION}
+\affil[15]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
 % Simon, Dom, Ivan
@@
 and was originally implemented as part of \msprime.
-We subsequently
-extracted the tree sequence code that was independent of simulations
-from \msprime\ into
-a standalone Python and C library, called \tskit~\citep{tskit2021tskit}.
+We subsequently extracted the core tree sequence functionality
+from \msprime\ to create the \tskit\ library,
+which provides a large suite of tools for processing genetic ancestry
+and variation data via Application Programming Interfaces (APIs)
+in the Python and C languages~\citep{tskit2021tskit}.
 While a full discussion of tree sequences and the capabilities
@@
 % Encoding ancestry
-Before we define tree sequences we require so additional terminology.
-Let a ``ploid'' be the
-% JK this isn't right either, but it's a start.
-% We should add a sentence clarifying what an individual is here.
-% FIXME
-complete set of genetic material that a child inherits from a parent,
-so that, e.g., each human has two ploids~\citep{jouganous2017inferring}.
-Then, given a set of sample ploids
-the structure
-of their genetic ancestry is described by a set of \emph{nodes}
-and \emph{edges} (Fig.~\ref{fig-ts-example}).
-A node in tree sequence terminology corresponds to a ploid,
-associated with its birth time
-and other optional information.
-% JK: Now that we have ploids, maybe it doesn't help to talk about
-% the individual table?
-% Nodes may be
-% associated with an \emph{individual}, so that, a diploid
-% individual is composed of two separate nodes (we omit
-% the individual table in Fig.~\ref{fig-ts-example} as it is optional).
-% While nodes define chromosomal sequences that exist at a specific time,
-% edges define genetic inheritance between these nodes.
-Edges then define the genetic inheritence between nodes.
+% JK: we can say this some other way, this just seemed like the
+% most concise way. People may be irritated by being told what
+% the definition of a "genome" is, but I think we need to be explicit.
+Let us define a genome as the complete set of genetic material
+that a child inherits from one parent. Thus, a diploid individual
+has two monoploid genomes, one inherited from each parent. Since
+each diploid individual lies at the end of two distinct lineages
+of descent, they will be represented by \emph{two} places (nodes)
+in any genealogical tree. In the tree sequence encoding a
+\emph{node} therefore corresponds to a single genome, which
+is associated with its creation time (and other optional information),
+and recorded in a simple tabular format (Fig.~\ref{fig-ts-example}).
+Genetic inheritance between genomes (nodes) % delete?
+is defined by edges.
 An \emph{edge} consists of a
 parent node, a child node and the left and right coordinates
-of the contiguous genomic interval over which
-the child node inherited genetic material from the parent node.
+% JK: This is potentially confusing. Changed "genomic interval"
+% to "chromosome" interval to avoid talking about "the" genome.
+of the contiguous chromosomal segment over which
+the child genome inherited genetic material from the parent genome.
 Parent and child nodes may correspond to ancestor and descendant
-ploids separated by many generations.
-
-Critically, edges can span multiple marginal trees: for
-example, in Fig.~\ref{fig-ts-example} the branch from node
+genome separated by many generations.
+Critically, edges can span multiple marginal trees along the
+genome, and identical node IDs across different trees corresponds
+to the same ancestral genome.
+For example, in Fig.~\ref{fig-ts-example} the branch from node
 0 to 4 is present in both marginal trees, and represented
@@
 This simple device, of explicitly associating tree nodes
-with specific ancestral chromosomes and recording the
-genomic intervals over which parent/child relationships exist,
+with specific ancestral genomes and recording the
+contiguous segments over which parent/child relationships exist,
 generalises the original
@@
 kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
-
-The tree sequence encoding of ancestry as a set of nodes and edges
-as just described, is equivalent to a graph in which edges
-are annotated with a set of disjoint genomic intervals.
-This is closely related to the Ancestral Recombination Graph (ARG),
-a term that is often loosely understood
-to mean some representation of genetic
-ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
-In this sense a succinct tree sequence \emph{is} an ARG.
-More technically, an ARG is a particular
-representation of genetic ancestry as a graph based on the
-flow of ancestral material through recombination and common
-ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
-In this sense, tree sequences can be used to \emph{encode}
-ARGs without loss of information (although
-the default minimal tree sequence encoding is
-much more concise; see the \nameref{sec-arg} section).
-It should be noted that, despite standardisation
-efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
-no shared format for interchanging ARG data, and no supporting software
-libraries.
+See the \nameref{sec-arg} section for a discussion of this
+closely related concept.
 
@@
 
+The tree sequence encoding of ancestry as a set of nodes and edges
+as just described, is equivalent to a graph in which edges
+are annotated with a set of disjoint genomic intervals.
+This is closely related to the Ancestral Recombination Graph (ARG),
+a term that is often loosely understood
+to mean some representation of genetic
+ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
+In this sense a succinct tree sequence \emph{is} an ARG.
+More technically, an ARG is a particular
+Fepresentation of genetic ancestry as a graph based on the
+flow of ancestral material through recombination and common
+ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
+In this sense, tree sequences can be used to \emph{encode}
+ARGs without loss of information (although
+the default minimal tree sequence encoding is
+much more concise; see the \nameref{sec-arg} section).
+It should be noted that, despite standardisation
+efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
+no shared format for interchanging ARG data, and no supporting software
+libraries.
+
 \begin{figure}
@@
 \section*{Acknowledgements}
-JK is supported by the Robertson Foundation.
+We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
+on the manuscript.
+
+% TODO alphabetise this by author I guess
+Jerome Kelleher is supported by the Robertson Foundation.
 Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
@@
 Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
+SG acknowledges funding from the Canada Research Chairs Program, from the
+Canadian Institutes of Health Research PJT 173300, and from the Canadian
+Foundation for Innovation.
 

### 2021-08-11 paper.tex
Caption for fig 1

@@
 \end{center}
-\caption{\label{fig-ts-example} Example tree sequence.
-An example tree sequence describing genealogies and genotypes
-for four samples at ten sites.
+\caption{\label{fig-ts-example} An example tree sequence describing
+genealogies and sequence variation for four samples at ten sites
+on a chromosome of twenty bases long.
+Information is stored in a set of tables
+(the tables shown here include only essential columns, and
+much more information can be associated with the various entities).
+The node table
+stores information about sampled and ancestral genomes. The
+edge table describes how these genomes are related along a
+chromosome, and defines the marginal trees at each position.
+The site and mutation tables together describe sequence variation
+% JK: also amongst the ancestral genomes, but only in positions where
+% they inherit from the sample. I thought it would be less confusing
+% to leave this out?
+among the samples.
 }
@@
 kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
-See the \nameref{sec-arg} section for a discussion of this
+See the \nameref{sec-arg} section below for a discussion of this
 closely related concept.

### 2021-08-11 paper.tex
Move some text around

@@
 
+\jkcomment{This paragraph has been copied from the previous
+version of the Tree Sequences section. Update and merge in here
+as appropriate.}
 The tree sequence encoding of ancestry as a set of nodes and edges

### 2021-08-11 paper.tex
Trim down ARG section

--- 2021-08-11 paper.tex
@@
 that a child inherits from one parent. Thus, a diploid individual
-has two monoploid genomes, one inherited from each parent. Since
+has two (monoploid) genomes, one inherited from each parent. Since
 each diploid individual lies at the end of two distinct lineages
@@
 Parent and child nodes may correspond to ancestor and descendant
-genome separated by many generations.
+genomes separated by many generations.
 Critically, edges can span multiple marginal trees along the
@@
 with specific ancestral genomes and recording the
-contiguous segments over which parent/child relationships exist,
+contiguous segments over which parent-child relationships exist,
 generalises the original
@@
 has occurred on the marginal tree.
-Crucially, the genetic sequences themselves are never stored, or indeed directly
+Crucially, the genome sequences themselves are never stored, or indeed directly
 represented in memory (unless explicitly decoded using \tskit\ APIs).
@@
 has shifted away from the original definition as a stochastic
-process, to a representation of a \emph{particular} genetic ancestry
+process, to a representation of a particular genetic ancestry
 using common ancestor and recombination
-events~\citep[e.g.][]{minichiello2006mapping}.
-More recently, ARGs have essentially come to mean any graph-like representation
-of genetic ancestry; for example,
-\cite{mathieson2020ancestry} informally define an ARG as the subset of an
-individual's pedigree from which they have inherited ancestral material.
-% Do we include this sentence? It's true, I'm sure: someone new to the field reads
-% Iain and Aylwyn's paper, and goes, hmm, this ARG thing sounds interesting,
-% let's follow up on that Griffiths reference... Maybe this is part of the
-% reason why ARGs have such a fearsome reputation for being difficult:
-% there's a basic confusion about what they actually *are*.
-Such usage is entirely reasonable, but referring readers back to the
-original Griffiths papers does little to clarify.
-
-The tree sequence encoding of ancestry as a set of nodes and edges
-as just described, is equivalent to a graph in which edges
-are annotated with a set of disjoint genomic intervals.
-This is closely related to the Ancestral Recombination Graph (ARG),
-a term that is often loosely understood
-to mean some representation of genetic
-ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
-In this sense a succinct tree sequence \emph{is} an ARG.
-More technically, an ARG is a particular
-Fepresentation of genetic ancestry as a graph based on the
-flow of ancestral material through recombination and common
-ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
-In this sense, tree sequences can be used to \emph{encode}
-ARGs without loss of information (although
-the default minimal tree sequence encoding is
-much more concise; see the \nameref{sec-arg} section).
-It should be noted that, despite standardisation
-efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
-no shared format for interchanging ARG data, and no supporting software
-libraries.
+events~\citep[e.g.][]{minichiello2006mapping},
+to being synonymous with the genetic
+ancestry of a sample, without any specific details of how genetic
+material disperses~\citep{mathieson2020ancestry}.
+Under the latter interpretation,
+the tree sequence encoding of genetic ancestry (described above)
+clearly \emph{is} an ARG: the nodes and edges
+define a graph in which edges are annotated with the
+set of disjoint genomic intervals
+through which ancestry flows.
+
+% If we interpret an ARG as being a graph describing genetic
+% ancestry, then clearly a
+
+% \jkcomment{This paragraph has been copied from the previous
+% version of the Tree Sequences section. Update and merge in here
+% as appropriate.}
+% The tree sequence encoding of ancestry as a set of nodes and edges
+% as just described, is equivalent to a graph in which edges
+% are annotated with a set of disjoint genomic intervals.
+% This is closely related to the Ancestral Recombination Graph (ARG),
+% a term that is often loosely understood
+% to mean some representation of genetic
+% ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
+% In this sense a succinct tree sequence \emph{is} an ARG.
+% More technically, an ARG is a particular
+% Fepresentation of genetic ancestry as a graph based on the
+% flow of ancestral material through recombination and common
+% ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
+% In this sense, tree sequences can be used to \emph{encode}
+% ARGs without loss of information (although
+% the default minimal tree sequence encoding is
+% much more concise; see the \nameref{sec-arg} section).
+% It should be noted that, despite standardisation
+% efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
+% no shared format for interchanging ARG data, and no supporting software
+% libraries.
 
@@
 rather store the information required to recover the genealogical
-history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
-or ``coalescence records'' in the original terminology~\citep{kelleher2016efficient}.
+history as nodes and edges in a tree sequence.
 This is far more efficient than outputting the simulated ARG in its entirety.
@@
 Although by default \msprime\ outputs tree sequences that contain full information
-about the genealogical trees, their correlation structure along the genome,
-and the ancestral haplotypes on which coalescences occurred, some information
+about the genealogical trees, their correlation structure along the chromosome,
+and the ancestral genomes on which coalescences occurred, some information
 is lost in this mapping down from ARG space to the minimal tree sequence
 form. In particular, we lose
-information about ancestral haplotypes that were common ancestors but
+information about ancestral genomes that were common ancestors but
 in which no coalescences occurred, and also information about the precise time
-and genomic location of recombination events. In most cases, such
+and chromosomal location of recombination events. In most cases, such
 information is of little relevance as it is in principle unknowable,

### 2021-08-12 paper.tex
Updates to mutation illustration

--- 2021-08-11 paper.tex
@@
 % Franz
-\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
+\affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'',
+Mathematical and Computational Population Genetics, University of T\"ubingen}
 % Gertjan, Konrad
@@
 
-\textbf{Keywords:} Coalescent simulation, Python, ancestral recombination graphs
+\textbf{Keywords:} Coalescent simulation, Python, Ancestral Recombination Graphs
 
@@
 \begin{figure}
-    \includegraphics{illustrations/unmutated_tree}
     \includegraphics{illustrations/mutated_tree}
@@
 
-% If we interpret an ARG as being a graph describing genetic
-% ancestry, then clearly a
-
-% \jkcomment{This paragraph has been copied from the previous
-% version of the Tree Sequences section. Update and merge in here
-% as appropriate.}
-% The tree sequence encoding of ancestry as a set of nodes and edges
-% as just described, is equivalent to a graph in which edges
-% are annotated with a set of disjoint genomic intervals.
-% This is closely related to the Ancestral Recombination Graph (ARG),
-% a term that is often loosely understood
-% to mean some representation of genetic
-% ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
-% In this sense a succinct tree sequence \emph{is} an ARG.
-% More technically, an ARG is a particular
-% Fepresentation of genetic ancestry as a graph based on the
-% flow of ancestral material through recombination and common
-% ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
-% In this sense, tree sequences can be used to \emph{encode}
-% ARGs without loss of information (although
-% the default minimal tree sequence encoding is
-% much more concise; see the \nameref{sec-arg} section).
-% It should be noted that, despite standardisation
-% efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
-% no shared format for interchanging ARG data, and no supporting software
-% libraries.
-
 \begin{figure}

### 2021-08-12 pretty_pictures.py

@@ -25,23 +25,42 @@ def mutated_tree():
         # Sizes are probably all wrong here
         style = """\
             @media print {
-              @page { margin: 0; size: 3in 2in}
+              @page { margin: 0; size: 6in 2.5in}
               body { margin: 1.6cm; }
             }
         """
-        ts.draw_svg(
+        return ts.draw_svg(
             size=(300, 200),
             node_labels={},
             mutation_labels={m.id: m.derived_state for m in ts.mutations()},
             symbol_size=5,
-            force_root_branch=True,
+            # Not sure why this was in here: makes it look like there's a root
+            # branch for mutation to hit?
+            # force_root_branch=True,
             style=style,
             **kwargs
         )
-
-
-    do_svg(ts, path="illustrations/unmutated_tree.svg")
-    do_svg(mts, path="illustrations/mutated_tree.svg")
+    svg1 = do_svg(ts)
+    svg2 = do_svg(mts)
+    height = 280
+    width = 700
+    top = 40
+    fig = (
+        f'<svg baseProfile="full" height="{height+top}" version="1.1" width="{width}" '
+         'xmlns="http://www.w3.org/2000/svg" xmlns:ev="http://www.w3.org/2001/xml-events" '
+         'xmlns:xlink="http://www.w3.org/1999/xlink">'
+        f'<g transform="translate(0 {top})">'
+        '<text x="2" y="-10" font-size="20">a</text>'
+        + svg1 +
+        '</g>'
+        f'<g transform="translate({width/2} {top})">'
+        '<text x="2" y="-10" font-size="20">b</text>'
+        + svg2 +
+        '</g>'
+        '</svg>'
+    )
+    with open("illustrations/mutated_tree.svg", "w") as f:
+        f.write(fig)
 
 
 @click.command()

### 2021-08-12 paper.tex

@@
 % Franz
-\affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'',
-Mathematical and Computational Population Genetics, University of T\"ubingen}
+\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
 % Gertjan, Konrad
@@
 
-\textbf{Keywords:} Coalescent simulation, Python, Ancestral Recombination Graphs
+\textbf{Keywords:} Coalescent simulation, Python, ancestral recombination graphs
 
@@
 \begin{figure}
+    \includegraphics{illustrations/unmutated_tree}
     \includegraphics{illustrations/mutated_tree}

### 2021-08-12 pretty_pictures.py
Update mutation fig to join SVGs

@@ -19,6 +19,10 @@ def mutated_tree():
     model = msprime.F84(kappa=2)
     mts = msprime.sim_mutations(ts, rate=5e-8, model=model, random_seed=5)
 
+    height = 280
+    width = 700
+    top = 50
+    hpad = 0
 
     def do_svg(ts, **kwargs):
 
@@ -30,7 +34,7 @@ def mutated_tree():
             }
         """
         return ts.draw_svg(
-            size=(300, 200),
+            size=(width / 2 - hpad, height - top),
             node_labels={},
             mutation_labels={m.id: m.derived_state for m in ts.mutations()},
             symbol_size=5,
@@ -40,21 +44,31 @@ def mutated_tree():
             style=style,
             **kwargs
         )
+
+    font_size = 15
+
+    # I think serif is the default, and matches what we're using for labels?
+    def make_text(text, y, font_family="serif"):
+        return (
+            f'<text x="{width / 4}" y="{y}" font-size="{font_size}" '
+            f'font-family="{font_family}" text-anchor="middle">'
+            f'{text}</text>'
+        )
+
     svg1 = do_svg(ts)
     svg2 = do_svg(mts)
-    height = 280
-    width = 700
-    top = 40
     fig = (
         f'<svg baseProfile="full" height="{height+top}" version="1.1" width="{width}" '
          'xmlns="http://www.w3.org/2000/svg" xmlns:ev="http://www.w3.org/2001/xml-events" '
          'xmlns:xlink="http://www.w3.org/1999/xlink">'
         f'<g transform="translate(0 {top})">'
-        '<text x="2" y="-10" font-size="20">a</text>'
+        + make_text("(A)", y=-20)
+        + make_text("sim_ancestry", y=-5, font_family="monospace")
         + svg1 +
         '</g>'
         f'<g transform="translate({width/2} {top})">'
-        '<text x="2" y="-10" font-size="20">b</text>'
+        + make_text("(B)", y=-20)
+        + make_text("sim_mutations", y=-5, font_family="monospace")
         + svg2 +
         '</g>'
         '</svg>'

### 2021-08-12 pretty_pictures.py
Change the mutation simulation to two diploid samples

@@ -9,11 +9,11 @@ def mutated_tree():
     Make a figure with (a) a tree and (b) some mutations added to it.
     """
     ts = msprime.sim_ancestry(
-        50,
+        2,
         population_size=1e4,
         recombination_rate=0.5e-8,
         sequence_length=1000,
-        random_seed=14,
+        random_seed=46,
     )
 
     model = msprime.F84(kappa=2)
@@ -22,7 +22,6 @@ def mutated_tree():
     height = 280
     width = 700
     top = 50
-    hpad = 0
 
     def do_svg(ts, **kwargs):
 
@@ -34,7 +33,7 @@ def mutated_tree():
             }
         """
         return ts.draw_svg(
-            size=(width / 2 - hpad, height - top),
+            size=(width / 2, height - top),
             node_labels={},
             mutation_labels={m.id: m.derived_state for m in ts.mutations()},
             symbol_size=5,
@@ -42,7 +41,7 @@ def mutated_tree():
             # branch for mutation to hit?
             # force_root_branch=True,
             style=style,
-            **kwargs
+            **kwargs,
         )
 
     font_size = 15
@@ -52,26 +51,26 @@ def mutated_tree():
         return (
             f'<text x="{width / 4}" y="{y}" font-size="{font_size}" '
             f'font-family="{font_family}" text-anchor="middle">'
-            f'{text}</text>'
+            f"{text}</text>"
         )
 
     svg1 = do_svg(ts)
     svg2 = do_svg(mts)
     fig = (
         f'<svg baseProfile="full" height="{height+top}" version="1.1" width="{width}" '
-         'xmlns="http://www.w3.org/2000/svg" xmlns:ev="http://www.w3.org/2001/xml-events" '
-         'xmlns:xlink="http://www.w3.org/1999/xlink">'
+        'xmlns="http://www.w3.org/2000/svg" xmlns:ev="http://www.w3.org/2001/xml-events" '
+        'xmlns:xlink="http://www.w3.org/1999/xlink">'
         f'<g transform="translate(0 {top})">'
         + make_text("(A)", y=-20)
         + make_text("sim_ancestry", y=-5, font_family="monospace")
-        + svg1 +
-        '</g>'
+        + svg1
+        + "</g>"
         f'<g transform="translate({width/2} {top})">'
         + make_text("(B)", y=-20)
         + make_text("sim_mutations", y=-5, font_family="monospace")
-        + svg2 +
-        '</g>'
[…]

### 2021-08-12 paper.tex
Update mutation fig to join SVGs

--- 2021-08-12 paper.tex
@@
         \label{fig-mutated-trees}
-        Visualisation of the separation between ancestry and mutation simulation:
-        \texttt{sim\_ancestry} produces ancestry
-        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
-        relating 200 samples),
-        and \texttt{sim\_mutations} adds mutations
-        (\textbf{right;} fourteen nucleotide mutations have been added).
-        Graphics produced with \tskit's \texttt{draw\_svg} method.
-        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
-        \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
-        \grgcomment{Add \texttt{sim\_ancestry} and \texttt{sim\_mutations} as subfigure titles?}
+        Visualisation of the separation between ancestry and mutation
+        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
+        is two trees along a 1Kb chunk of genome relating two diploid samples.
+        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
+        which adds mutations.
+        Graphics produced using \tskit's \texttt{draw\_svg} method.
     }

### 2021-08-12 paper.tex
Update figure plots to use absolute sizes.
Also systematise lines so that msprime's main result is the same colour

Closes #105

@@
         simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
-        is two trees along a 1Kb chunk of genome relating two diploid samples.
+        is two trees along a 1kb chunk of genome relating two diploid samples.
         (B) This ancestry is provided as the input to \texttt{sim\_mutations},
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{figures/mutations-perf}
+    \includegraphics{figures/mutations-perf}
 \caption{\label{fig-mutations-perf} Time required to run
@@
 \begin{center}
-    \includegraphics[width=\textwidth]{figures/ancestry-perf}
+    \includegraphics{figures/ancestry-perf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/gc-perf}
+\includegraphics{figures/gc-perf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/sweeps-perf}
+\includegraphics{figures/sweeps-perf}
 \end{center}
@@
 allele had a scaled selection coefficient of $2Ns=1000$,
-the per-base recombination rate of $10^{-9}$, and we allowed
-the sequence length to vary between 1kb-200kb.}
+a per-base recombination rate of $10^{-9}$ and sequence
+length varying from 1kb--200kb.}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/dtwf-perf}
+\includegraphics{figures/dtwf-perf}
 \end{center}
@@
 the standard coalescent with recombination model (``DTWF + Hudson'').
+Memory usage for \msprime's DTWF and hybrid simulations are very similar.
 We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,

### 2021-08-12 evaluation/plot.py

@@ -4,6 +4,7 @@ import pandas as pd
 import numpy as np
 import click
 
+plt.rcParams.update({'font.size': 8})
 
 @click.group()
 def cli():
@@ -15,6 +16,12 @@ def save(name):
     plt.savefig(f"figures/{name}.png")
     plt.savefig(f"figures/{name}.pdf")
 
+def two_panel_fig(**kwargs):
+    fig, (ax1, ax2)  = plt.subplots(1, 2, figsize=(6, 3), **kwargs)
+    ax1.set_title("(A)")
+    ax2.set_title("(B)")
+    return fig, (ax1, ax2)
+
 
 @click.command()
 def mutations_perf():
@@ -25,15 +32,13 @@ def mutations_perf():
     df.L /= 1e6
     rates = np.unique(df.rate)
 
-    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6), sharey=True)
-    ax1.set_title("(A)")
+    fig, (ax1, ax2) = two_panel_fig(sharey=True)
     ax1.set_xlabel("Sample size")
     dfL = df[df.L == 10]
     for rate in rates:
         dfr = dfL[dfL.rate == rate].sort_values("n")
         ax1.plot(dfr.n, dfr.time, label=f"rate={rate}")
 
-    ax2.set_title("(B)")
     ax2.set_xlabel("Sequence length (Megabases)")
     ax1.set_ylabel("Time (seconds)")
 
@@ -60,19 +65,15 @@ def gc_perf():
     df.user_time /= 3600
 
     lines = {}
-    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 6))
-    for tool in ["SimBac", "fastSimBac", "msprime"]:
+    fig, (ax1, ax2) = two_panel_fig()
+    for tool in ["msprime", "SimBac", "fastSimBac"]:
         dft = df[df.tool == tool]
         (line,) = ax1.plot(dft.sample_size, dft.user_time, label=tool)
         ax2.plot(dft.sample_size, dft.memory, label=tool, color=line.get_color())
         lines[tool] = line
 
-    ax1.set_title("(A)")
-    ax2.set_title("(B)")
     ax1.set_xlabel("Sample size")
     ax1.set_ylabel("Time (hours)")
-    ax1.set_ylim(0, None)
-    ax2.set_ylim(0, None)
     ax2.set_xlabel("Sample size")
[…]

### 2021-08-12 evaluation/plot.py
Fixup some plotting details.

@@ -4,7 +4,8 @@ import pandas as pd
 import numpy as np
 import click
 
-plt.rcParams.update({'font.size': 8})
+plt.rcParams.update({"font.size": 8})
+
 
 @click.group()
 def cli():
@@ -16,8 +17,9 @@ def save(name):
     plt.savefig(f"figures/{name}.png")
     plt.savefig(f"figures/{name}.pdf")
 
+
 def two_panel_fig(**kwargs):
-    fig, (ax1, ax2)  = plt.subplots(1, 2, figsize=(6, 3), **kwargs)
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(6, 3), **kwargs)
     ax1.set_title("(A)")
     ax2.set_title("(B)")
     return fig, (ax1, ax2)
@@ -90,7 +92,6 @@ def gc_perf():
         xycoords="data",
     )
     ax2.set_ylim(bottom=0)
-    plt.tight_layout()
 
     save("gc-perf")
 
@@ -144,7 +145,6 @@ def sweeps_perf():
         xy=(largest_L, largest_value),
         xycoords="data",
     )
-    plt.tight_layout()
     save("sweeps-perf")
 
 
@@ -166,7 +166,7 @@ def dtwf_perf():
     for tool in ["msprime", "hybrid", "ARGON"]:
         dft = df[df.tool == tool]
         ax1.plot(dft.L, dft.user_time, label=label_map[tool])
-        line, = ax2.plot(dft.L, dft.memory, label=label_map[tool])
+        (line,) = ax2.plot(dft.L, dft.memory, label=label_map[tool])
         lines[tool] = line
 
     ax1.set_xlabel("Sequence length (Megabases)")
@@ -212,13 +212,13 @@ def ancestry_perf():
 
     fig, axes = two_panel_fig()
 
-    def annotate_rho(ax, rho, x_offset, text):
+    def annotate_rho(ax, rho, x_offset, y_offset, text):
         ax.axvline(rho / 4, color="0.8", linestyle="-.")
-        ax.text(rho / 4 + x_offset, 0, text, fontstyle="italic")
+        ax.text(rho / 4 + x_offset, y_offset, text, fontstyle="italic")
 
-    annotate_rho(axes[0], aratha_chr1, 200, "Arabidopsis thaliana")
-    annotate_rho(axes[1], human_chr1, 1000, "Homo sapiens")
[…]

### 2021-08-12 evaluation/plot.py
axis fix

@@ -289,6 +289,16 @@ def ancestry_perf():
                 fitted_quadratic(dromel_chr2l) / 3600,
                 "hours",
             )
+        max_x = np.max(rho[df["time"] <= tl] / 4)
+        ticks = np.arange(
+                0,
+                1.1 * max_x,
+                10 ** np.floor(np.log10(max_x)),
+        )
+        ax.set_xticks(ticks)
+        ax.set_xticklabels(
+                [f"{x:.0f}"for x in ticks]
+        )
 
     prop = {"size": 7}
     axes[0].legend(handles=legend_adds, prop=prop)

### 2021-08-13 paper.tex
Fix Saunack's details
Move interface section up to top
Text tweaks.

@@
 \author[10]{Murillo F. Rodrigues}
-\author[20]{Kumar Suanack}
+\author[20]{Kumar Saunack}
 \author[21]{Thibaut Sellinger}
@@
 % Saunack
-\affil[20]{FIXME: SAUNACK's AFFILIATION}
+\affil[20]{IIT Bombay, India}
 % Thibaut
@@
 
+% The input - how do we run simulations?
+\subsection*{User interface}
+\label{sec-sim-interface}
+
+The majority of simulation packages are controlled either through
+a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
+a text-based input file
+format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
+or a mixture of both.
+Command line interfaces make it easy to run simple
+simulations, but as model complexity and the number of parameters increase,
+they become difficult to understand and
+error-prone~\citep{ragsdale2020lessons,gower2021demes}.
+Specifying parameters through a text file alleviates this problem to a degree,
+but lacks flexibility, for example, when running simulations with parameters
+drawn from a distribution. In practice, for any reproducible simulation
+project users will write a script
+to generate the required command lines or input parameter files,
+invoke the simulation engine, and process the results in some way.
+This process is cumbersome and labour intensive, and
+a number of packages have been developed
+to allow simulations to be run directly in a high-level
+scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
+
+The more recent trend has been to move away from this file and command-line
+driven approach and to instead provide direct interfaces to the simulation
+engines via an Application Programming Interface (API)~\citep[e.g.][]{
+thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
+The primary interface for \msprime\ is through a thoroughly documented and
+stable Python
+API, which has encouraged the development of an ecosystem of downstream
+% Found these through the Dependent packages on GitHub
+tools~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
+adrion2020community,adrion2020predicting, kamm2020efficiently,
+mckenzie2020ipcoal, montinaro2020revisiting,
+de2021geonomics,rivera2021simulation}.
+As well as providing a stable and efficient platform for building
+downstream applications, \msprime's Python API makes it much easier to
+build reproducible simulation pipelines, as the entire workflow can
+be encapsulated in a single script, and package and version
+dependencies explicitly stated using the \texttt{pip}
+or \texttt{conda} package managers.
+For example, the errors made
+in the influential simulation analysis of
+\cite{martin2017human} were only detected because the pipeline
+could be easily run and reanalysed~\citep{ragsdale2020lessons,martin2020erratum}.
+
+\begin{figure}
+    \includegraphics{illustrations/mutated_tree}
+    \caption{
+        \label{fig-mutated-trees}
+        Visualisation of the separation between ancestry and mutation
+        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
+        is two trees along a 1kb chunk of genome relating two diploid samples.
+        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
+        which adds mutations.
+        Graphics produced using \tskit's \texttt{draw\_svg} method.
+    }
+\end{figure}
+
+A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
+designed in part to avoid sources of error (see the Demography section) but
+also to provide more appropriate defaults while keeping compatibility with
+existing code. In the new APIs, ancestry and mutation simulation are fully
+separated (see Fig.~\ref{fig-mutated-trees}),
+with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
+functions replacing the legacy \texttt{simulate} function. Among other changes,
+the new APIs default to discrete genome coordinates and finite sites mutations,
+making the default settings more realistic and resolving a major source of confusion and error.
+The previous APIs are fully
+supported and tested, and will be maintained for the foreseeable future.
+The \texttt{msp} program has been extended to include new commands
+for simulating ancestry and mutations separately. A particularly useful
+feature is the ability to specify demographic models in
+Demes format~\citep{gower2021demes} from the command line,
+making simulation of complex demographies straightforward.
+We also provide an \ms\ compatible
+command line interface to support existing workflows.
+
 % The data model
@@
 \begin{center}
-\includegraphics[width=6cm]{illustrations/example_tree_sequence}
+\includegraphics{illustrations/example_tree_sequence}
 \end{center}
@@
 
-% The input - how do we run simulations?
-\subsection*{User interface}
-\label{sec-sim-interface}
-
-The majority of simulation packages are controlled either through
-a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
-a text-based input file
-format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
-or a mixture of both.
-Command line interfaces make it very easy to run simple
-simulations, but as model complexity and the number of parameters specified increase,
-they become difficult to understand and
-error-prone~\citep{ragsdale2020lessons,gower2021demes}.
-Specifying parameters through a text file alleviates this problem to a degree,
-but lacks flexibility, for example, when running simulations with parameters
-drawn from some distribution. In practice, for any reproducible simulation
-project users will write a script
-to generate the required command lines or input parameter files,
-invoke the simulation engine, and process the results in some way.
-This process is cumbersome and labour intensive, and
-a number of packages have been developed
-to allow simulations to be run directly in a high-level
-scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.
-
-The more recent trend has been to move away from this file and command-line
-driven approach and to instead provide direct interfaces to the simulation
-engines via an Application Programming Interface (API)~\citep[e.g.][]{
-thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
-The primary interface for \msprime\ is through a thoroughly documented and
-stable Python
-API, which has encouraged the development of an ecosystem of downstream
-% Found these through the Dependent packages on GitHub
-tools~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
-adrion2020community,adrion2020predicting, kamm2020efficiently,
-mckenzie2020ipcoal, montinaro2020revisiting,
-de2021geonomics,rivera2021simulation}.
-As well as providing a stable and efficient platform for building
-downstream applications, \msprime's Python API makes it much easier to
-build reproducible simulation pipelines, as the entire workflow can
-be encapsulated in a single script, and package and version
-dependencies explicitly stated using the \texttt{pip}
-or \texttt{conda} package managers.
-For example, the errors made
-in the influential simulation analysis of
-\cite{martin2017human} were only detected because the pipeline
-could be easily run and reanalysed~\citep{ragsdale2020lessons,martin2020erratum}.
-
-A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
-designed in part to avoid sources of error (see the Demography section) but
-also to provide more appropriate defaults while keeping compatibility with
-existing code. In the new APIs, ancestry and mutation simulation are fully
-separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
-functions replacing the legacy \texttt{simulate} function. Among other changes,
-the new APIs default to discrete genome coordinates and finite sites mutations,
-making the default settings more realistic and resolving a major source of confusion and error.
-The previous APIs are fully
-supported and tested, and will be maintained for the foreseeable future.
-We also provide an \ms\ compatible
-command line interface to support existing workflows. The
-\texttt{msp} program has been extended to include new commands
-for simulating ancestry and mutations separately. A particularly useful
-feature is the ability to specify demographic models in
-Demes format~\citep{gower2021demes} from the command line,
-making simulation of complex demographies straightforward.
-
 % The output - what do we get out?
@@
 the two have been conflated in practice.
-
-\begin{figure}
-    \includegraphics{illustrations/mutated_tree}
-    \caption{
-        \label{fig-mutated-trees}
-        Visualisation of the separation between ancestry and mutation
-        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
-        is two trees along a 1kb chunk of genome relating two diploid samples.
-        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
-        which adds mutations.
-        Graphics produced using \tskit's \texttt{draw\_svg} method.
-    }
-\end{figure}
 
@@
 and the ``little'' ARG in which we only track genetic ancestors.
-Over time, the interpretation of the term
-has shifted away from the original definition as a stochastic
-process, to a representation of a particular genetic ancestry
-using common ancestor and recombination
-events~\citep[e.g.][]{minichiello2006mapping},
-to being synonymous with the genetic
-ancestry of a sample, without any specific details of how genetic
-material disperses~\citep{mathieson2020ancestry}.
+Over time, usage of the term
+has shifted away from its original definition as a stochastic process,
+to being interpreted as a representation of a particular genetic ancestry
+as a graph, without necessarily following the specific details of the Griffiths
+formulation~\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
 Under the latter interpretation,
@@
 \centering
-    \begin{tikzpicture}[scale=0.45]
+\caption*{(A)}
+    \begin{tikzpicture}[scale=0.45,nodestyle/.style={font=\scriptsize}]
 
@@
 
-    \node [below] at (-2,0) {0};
-    \node [below] at (1,0) {1};
-    \node [below] at (4,0) {2};
-    \node [above] at (1,1) {3};
-    \node [below right] at (1,1) {$x = 0.3$};
-    \node [above right] at (3,2) {4};
-    \node [above left] at (-1,3) {5};
-    \node [above] at (1,5) {6};
+    \node[nodestyle] [below] at (-2,0) {0};
+    \node[nodestyle] [below] at (1,0) {1};
+    \node[nodestyle] [below] at (4,0) {2};
+    \node[nodestyle] [above] at (1,1) {3};
+    \node[nodestyle] [below right] at (1,1) {$x = 0.3$};
+    \node[nodestyle] [above right] at (3,2) {4};
+    \node[nodestyle] [above left] at (-1,3) {5};
+    \node[nodestyle] [above] at (1,5) {6};
 
     \end{tikzpicture}
-\caption{}\label{fig-arg-a}
 \end{subfigure}
@@
 \centering
-\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.4]
+\caption*{(B)}
+\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
 
@@
 \node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
-\node [greynode] (s6) at ($0.5*(s5) + 0.5*(s4) + (0,2)$) {};
-\node [greynode] (s6r) at ($0.5*(s5r) + 0.5*(s4r) + (0,2)$) {};
+\node [greynode] (s6) at ($0.25*(s0) + 0.25*(s1) + 0.5*(s2) + (0,5)$) {};
+\node [greynode] (s6r) at ($0.5*(s0r) + 0.25*(s1r) + 0.25*(s2r) + (0,5)$) {};
 \foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
@@
 % Axes
-\draw[very thick] ($(s0) + (0,5.8)$) -- ($(s2r) + (0,5.8)$);
-\node (topAx) at (0,5.8) {};
-\node (topLeft) at ($(s0) + (0,5.8)$) {};
+\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
+\node (topAx) at (0,-1.0) {};
+\node (topLeft) at ($(s0) + (0,-1.0)$) {};
 \node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
 \foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
-\node[anchor=south] at ($(topLeft)$) {\scriptsize 0.0};
-\node[anchor=south] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
-\node[anchor=south] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
+\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
+\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
 
 % Interval endpoints
-\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, -5.5);
+\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, 5.5);
 
 % Axis titles
-\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,1.3)$) {$\scriptsize\textrm{Genome position}$};
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
 \end{tikzpicture}
-\caption{}\label{fig-arg-b}
 \end{subfigure}
@@
 \centering
-\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.4]
+\caption*{(C)}
+\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
 
@@
 %\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
-\node [greynode] (s6) at ($0.5*(s5) + 0.5*(s4) + (0,2)$) {};
-\node [greynode] (s6r) at ($0.5*(s5r) + 0.5*(s4r) + (0,2)$) {};
+\node [greynode] (s6) at ($0.25*(s0) + 0.25*(s1) + 0.5*(s2) + (0,5)$) {};
+\node [greynode] (s6r) at ($0.5*(s0r) + 0.25*(s1r) + 0.25*(s2r) + (0,5)$) {};
 %\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
@@
 % Axes
-\draw[very thick] ($(s0) + (0,5.8)$) -- ($(s2r) + (0,5.8)$);
-\node (topAx) at (0,5.8) {};
-\node (topLeft) at ($(s0) + (0,5.8)$) {};
+\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
+\node (topAx) at (0,-1.0) {};
+\node (topLeft) at ($(s0) + (0,-1.0)$) {};
 \node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
 \foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
-\node[anchor=south] at ($(topLeft)$) {\scriptsize 0.0};
-\node[anchor=south] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
-\node[anchor=south] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
+\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
+\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
 
 % Interval endpoints
-\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, -5.5);
+\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, 5.5);
 
 % Axis titles
-\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,1.3)$) {$\scriptsize\textrm{Genome position}$};
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
 \end{tikzpicture}
-\caption{}\label{fig-arg-c}
 \end{subfigure}

### 2021-08-13 paper.tex
Revise data analysis section

@@
 which provides a large suite of tools for processing genetic ancestry
-and variation data via Application Programming Interfaces (APIs)
+and variation data via APIs
 in the Python and C languages~\citep{tskit2021tskit}.
@@
 mutations (see the \nameref{sec-mutations} section); and (3) output the
-resulting sequences by percolating the effects of the mutations through
+resulting nucleotide sequences by percolating the effects of the mutations through
 the trees. Information about the mutations
@@
 Crucially, the genome sequences themselves are never stored, or indeed directly
-represented in memory (unless explicitly decoded using \tskit\ APIs).
+represented in memory (although \tskit\ can output the variant matrix
+in various formats, if required).
 It may at first seem inconvenient to have only this indirect representation
@@
 \label{sec-data-analysis}
-The point of running simulations is to generate data, and there
-are two important ways in which the \tskit\ tree sequences output by
-\msprime\ are more useful than classical methods:
-they are more efficient and they contain more information.
-We will consider these points in turn.
+% We could drop this first sentence?
+% The point of running simulations is to generate data, and there
+% are two important ways in which the \tskit\ tree sequences output by
+% \msprime\ are more useful than classical methods:
+% they are more efficient and they contain more information.
+% We will consider these points in turn.
 
 The standard way of representing simulation data is
-to render the results in a text format. For example,
-\ms~\citep{hudson2002generating} outputs a set of sequences
+to render the results in a text format, which must subsequently
+be parsed and processed as part of some analysis pipeline. For example,
+\ms\ outputs a set of sequences
 and can also optionally output the marginal trees along
@@
 and variants of this approach are used by many simulators.
-Text files have many advantages, but are slow to process
-at scale.
+Text files have many advantages, but are slow to process at scale.
 The ability to efficiently process simulation results is
@@
 results all the more acute.
+By using the stable APIs and efficient data interchange mechanisms
+provided by \tskit,
+the results of an \msprime\ simulation can be immediately processed,
+without format conversion overhead.
+The \tskit\ library has a rich suite of population
+genetic statistics and other utilities, and is in many cases
+orders of magnitude faster than matrix-based methods for
+large sample sizes~\citep{ralph2020efficiently}. Thus, the combination
+of \msprime\ and \tskit\ substantially increases the overall
+efficiency of many simulation analysis pipelines.
 
 Classical text based output formats like \ms\ are inefficient to process,
-but also lack a great deal of important information.
+but also lack a great deal of important information about the simulated
+process.
 The tree-by-tree topology information output by simulators in Newick
@@
 from the output. Because Newick stores branch lengths rather
-than node times, numerical precision issues also arise when summing
-over many short branches~\citep{mcgill2013graphml}.
+than node times, numerical precision issues also arise for
+large trees~\citep{mcgill2013graphml}.
 Numerous forks  of simulators have been created to access information not provided
@@
 output information about migrating segments~\citep{rosenzweig2016powerful},
-information about ancestral lineages~\citep{chen2013asymptotic},
-an array of population genetic statistics~\citep{ramos2007mlcoalsim}
+ancestral lineages~\citep{chen2013asymptotic},
+% leaving this out because it's not about hidden information
+% an array of population genetic statistics~\citep{ramos2007mlcoalsim}
 % Any more?
@@
 
-These problems of inefficient and limited access to the simulated
-data are comprehensively solved by \msprime\ and \tskit. Complete
-information about ancestry and mutation in simulations is immediately
-available to the user in a single integrated data structure
-(rather than multiple independent files), with no intermediate
-processing required.
-Users can subsequently compute precisely the information that they
-require using the powerful suite of population genetic statistics
-and other methods available in \tskit.
+% These problems of inefficient and limited access to the simulated
+% data are comprehensively solved by \msprime\ and \tskit. Complete
+% information about ancestry and mutation in simulations is immediately
+% available to the user in a single integrated data structure
+% (rather than multiple independent files), with no intermediate
+% processing required.
+% Users can subsequently compute precisely the information that they
+% require using the powerful suite of population genetic statistics
+% and other methods available in \tskit.
 

### 2021-08-13 paper.tex
Edits up to Recombination

@@
 Because coalescent simulations are usually concerned with
-neutral evolution (however, see the \nameref{sec-selection} section)
+neutral evolution (see the \nameref{sec-selection} section, however)
 the problem of generating synthetic genetic variation can be decomposed into
@@
 there is no performance penalty for additional operations
-in a different Python package.
+in a different software package.
 Thanks to this interoperability, \msprime's
@@
 As well as providing a new API
-that emphasises the logical split between ancestry and mutation simulation
-(see the \nameref{sec-sim-interface} section),
+that emphasises the logical split between ancestry and mutation simulation,
 we have greatly extended the sophistication of
@@
 can be defined easily and efficiently in
-user code by specifying an arbitrary transition matrix between
+user code by specifying a transition matrix between
 any number of alleles (which can be arbitrary unicode strings).
@@
 
-\subsection*{Selection}
+\subsection*{Selective sweeps}
 \label{sec-selection}

### 2021-08-13 paper.tex

--- 2021-08-13 paper.tex
@@
 or a mixture of both.
-Command line interfaces make it easy to run simple
-simulations, but as model complexity and the number of parameters increase,
+Command line interfaces make it very easy to run simple
+simulations, but as model complexity and the number of parameters specified increase,
 they become difficult to understand and
@@
 but lacks flexibility, for example, when running simulations with parameters
-drawn from a distribution. In practice, for any reproducible simulation
+drawn from some distribution. In practice, for any reproducible simulation
 project users will write a script
@@
 
-\begin{figure}
-    \includegraphics{illustrations/mutated_tree}
-    \caption{
-        \label{fig-mutated-trees}
-        Visualisation of the separation between ancestry and mutation
-        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
-        is two trees along a 1kb chunk of genome relating two diploid samples.
-        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
-        which adds mutations.
-        Graphics produced using \tskit's \texttt{draw\_svg} method.
-    }
-\end{figure}
-
 A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
@@
 existing code. In the new APIs, ancestry and mutation simulation are fully
-separated (see Fig.~\ref{fig-mutated-trees}),
-with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
+separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
 functions replacing the legacy \texttt{simulate} function. Among other changes,
@@
 the two have been conflated in practice.
+
+\begin{figure}
+    \includegraphics{illustrations/mutated_tree}
+    \caption{
+        \label{fig-mutated-trees}
+        Visualisation of the separation between ancestry and mutation
+        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
+        is two trees along a 1kb chunk of genome relating two diploid samples.
+        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
+        which adds mutations.
+        Graphics produced using \tskit's \texttt{draw\_svg} method.
+    }
+\end{figure}
 
@@
 and the ``little'' ARG in which we only track genetic ancestors.
-Over time, usage of the term
-has shifted away from its original definition as a stochastic process,
-to being interpreted as a representation of a particular genetic ancestry
-as a graph, without necessarily following the specific details of the Griffiths
-formulation~\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+Over time, the interpretation of the term
+has shifted away from the original definition as a stochastic
+process, to a representation of a particular genetic ancestry
+using common ancestor and recombination
+events~\citep[e.g.][]{minichiello2006mapping},
+to being synonymous with the genetic
+ancestry of a sample, without any specific details of how genetic
+material disperses~\citep{mathieson2020ancestry}.
 Under the latter interpretation,

### 2021-08-13 paper.tex
Update demography section
Closes #155

--- 2021-08-13 paper.tex
@@
 Recombination events split the ancestral material for some lineage
-at a given breakpoint, creating two independent
+at a breakpoint, creating two independent
 lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
-this process is much more efficient to simulate than either
-methods based on the Ancestral Recombination Graph
-formulation~\citep{griffiths1991two,griffiths1997ancestral}
-or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
+this process is much more efficient to simulate
+than the equivalent left-to-right
+approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
 In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
@@
 with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
-The SMC and SMC' models are available
+The SMC and SMC' models are supported
 in \msprime\ 1.0. However, they are currently implemented using a
@@
 (the sum is a harmonic number and can be approximated by $\log{n}$)
-is not well supported by observed runtimes except possibly at very
-large values of $\rho$. So it seems a different dependence
-on $n$ would be required to accurately predict simulation time for a
+is not well supported by observed run times (or numbers of events)
+except possibly at very
+large values of $\rho$. It therefore appears that a different dependence
+on $n$ is required to accurately predict simulation time for a
 given $\rho$ and $n$.
@@
 $N_e L \approx 10^6$, so extrapolating the curve in Fig.~\ref{fig-ancestry-perf}B
-predicts that simulation would require around 177 hours.
+predicts that simulation would require around 177 hours for 1000 samples.
 For such large values of $\rho$  we recommend
@@
 have efficient implementations of approximate coalescent with recombination
-models, we recommend using SMC based methods such as \scrm\,
-particularly for moderate sample sizes.
+models, we recommend using SMC based methods such as \scrm,
+particularly for small sample sizes.
 In practice, to predict the running time of a given
@@
 predicting the run times of days-long simulations
-using the timing of minutes-long runs is unlikely to be very accurate.
+using the timing of seconds-long runs is unlikely to be very accurate.
 
@@
 migration can be recorded.
+Large demographic models can be simulated efficiently in version \msprime\ 1.0,
+since we only consider populations that contain lineages and
+have non-zero migration rates when generating migration event
+waiting times. This is a considerable improvement
+over version 0.x, which scaled quadratically with the number of populations.
 
@@
 Another improvement over the
-0.X APIs is the introduction of explicit population split and admixture events,
+0.x APIs is the introduction of explicit population split and admixture events,
 and a population state machine that ensures that lineages cannot migrate into
@@
 the \texttt{Demography} class, and we also support importing demographic
-models in the form of Newick species tree descriptions, as well as those
-produced by programs such as *BEAST~\citep{heled2009bayesian}.
-
-The \texttt{DemographyDebugger} can output detailed information for debugging
-demographic models, but also provides
-methods to compute a suite of numerical predictions about the location of
-lineages over time, for example the coalescence rates for two or more lineages
+models from Newick species trees and the output of programs
+like *BEAST~\citep{heled2009bayesian}.
+
+The \texttt{DemographyDebugger} provides detailed information about
+demographic models as well as numerical methods to make predictions
+about these models. For example, we can compute
+the coalescence rates for two or more lineages
 drawn from populations at specified times in the past,
-which can be inverted to obtain the ``inverse instantaneous coalescence rate'',
-or IICR of \citet{chikhi2018iicr}.
+which can inverted to obtain the
+``inverse instantaneous coalescence rate'' of \citet{chikhi2018iicr}.
 Many popular approaches in population genetics use the distribution of
@@
 populations~\citep{wang2020tracking}.
-The \texttt{DemographyDebugger} provides a valuable
+These numerical methods provide a valuable
 ground-truth when evaluating such inference methods, as illustrated by

### 2021-08-13 paper.tex
last few sections

@@
 \subsection*{Demography}
+\label{sec-demography}
 % TODO should we put in some citations here for the models? Hardly seems
@@
 algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
-by adding events to the Demography object (see the Demography section)
+by adding events to the \texttt{Demography} class
+(see the \nameref{sec-demography} section)
 and can be used in combination with any other demographic modelling
@@
 have also been shown to be relevant for modelling the effects of selection on
-gene genealogies \citep{Gillespie909,DS04}.
+gene genealogies
+\citep{Gillespie909,DS04,desai2013genetic,neher2013genealogies,schweinsberg2017rigorous}.
 
@@
 The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
-includes a number of extensions to the Beta-coalescent.
+also includes a number of extensions to the Beta-coalescent.
 None of these methods work with large genomes, and very little work
@@
  as well as more general properties of coalescent processes.
-Please see the Appendix for more details and model derivations.
+See the Appendix for more details and model derivations.
 
@@
 sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation.}
+topology ``simplified'' down to its minimal tree sequence representation.
+Note that the internal nodes have been renumbered in the simplified
+representation, so that, e.g., node 5 in (C) corresponds to node 6 in
+(A) and (B).}
 \end{figure}
@@
 As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
-works by dynamically generating and traversing a little ARG.
+works by dynamically traversing a little ARG.
 The graph is not explicitly represented in memory, but is partially
@@
 of an allele bound for
-fixation~\citep{coop2004ancestral}---see the appendix for more details.
+fixation~\citep{coop2004ancestral}---see the Appendix for more details.
 This implementation facilitates future changes to generalise sweep models to
-changing population sizes as well as selection from standing variation,
-both of which are features that are planned for the near future.
+changing population sizes as well as selection from standing variation.
 Moreover because this implements the basic infrastructure for the structured
@@
 that \msprime\ is much more efficient than \msms\ and \cosiTwo\
-when simulating without selection~\citep{kelleher2016efficient},
-thus we believe \msprime\ to be the most efficient
-coalescent simulator when simulating sweeps.
+when simulating without selection~\citep{kelleher2016efficient}.
+% jk - this just seems clumsy to me: surely the conclusion is clear?
+% and we therefore believe \msprime\ to be the most efficient
+% coalescent simulator when simulating sweeps.
 
@@
 This pathological behaviour results in
-Identity-by-descent,
-long-range linkage disequilibrium (LD), and ancestry patterns deviating from
+identity-by-descent,
+long-range linkage disequilibrium and ancestry patterns deviating from
 Wright-Fisher expectations, and the bias grows with larger sample
@@
 genetic ancestry. Given a tree sequence that
-is complete up until time $t$ as input
+is complete up until time $t$ ago as input
 (where marginal trees may or may not have fully coalesced),
@@
 retained---at scale---is unprecedented.
-This interoperability is indicating an opportunity for
+This interoperability indicates an opportunity for
 other forward genetic simulators ~\citep[e.g.][]{Gaynor2021AlphaSimR}
-to leverage tree sequence data format and associated tools.
+to leverage the tree sequence data format and associated tools.
 
@@
 (where a contributor proposes a code change), using standard Continuous
-Integration (CI) methodology. Other CI services check code formatting
-and for common errors, as well as producing reports on the level of
-test coverage, e.g., pinpointing areas where additional unit tests may
-be required.
+Integration (CI) methodology. Other CI services
+check for common errors, code formatting issues, and produce reports on
+the level of test coverage for the proposed change.
 
@@
 % We can simulate so much more now
-
 This release makes \msprime\ able to produce simulations
@@
 to a high quality if we are to trust its results. There is a growing
-realisation across the sciences \citet[OTHER CITATIONS]{siepel2019challenges}
-% Maybe cite: https://www.biorxiv.org/content/10.1101/092205v3
+realisation across the sciences
+% More citations?
+\citep{siepel2019challenges,harris2020array,gardner2021sustained}
 that investing in
@@
 established development practices and infrastructure.
-% I'm going out on a limb here - do folks agree?
+% JK: I'm going out on a limb here - do folks agree?
 % PLR- personally I have learned SO MUCH, and I'm passing it on to others
@@
 with mentorship by experienced developers,
-is a hugely useful experience for many users.
+is a useful experience for many users.
 The skills that contributors learn

### 2021-08-13 paper.tex
Update affiliations
Update caption

@@
 \author[22]{Kevin Thornton}
-\author[23]{Hugo van Kemenade}
+\author[9]{Hugo van Kemenade}
 \author[7]{Anthony W. Wohns}
@@
 \author[10,$\dagger$]{Andrew D. Kern}
-\author[24,$\dagger$]{Jere Koskela}
-\author[10,25,$\dagger$]{Peter L. Ralph}
+\author[23,$\dagger$]{Jere Koskela}
+\author[10,24,$\dagger$]{Peter L. Ralph}
 
@@
 Science, Museum f\"ur Naturkunde Berlin }
-% Castedo
-\affil[9]{FIXME: CASTEDO's AFFILIATION}
+% Castedo, Hugo
+\affil[9]{No affiliation}
 % Jared, Murillo, Peter, Andy
@@
 \affil[22]{Ecology and Evolutionary Biology, University of California Irvine}
-% Hugo
-\affil[23]{FIXME HUGO's AFFILIATION}
 % Jere
-\affil[24]{Department of Statistics, University of Warwick}
+\affil[23]{Department of Statistics, University of Warwick}
 % Peter (second affil)
-\affil[25]{Department of Mathematics, University of Oregon}
+\affil[24]{Department of Mathematics, University of Oregon}
 
@@
         simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
-        is two trees along a 1kb chunk of genome relating two diploid samples.
+        is two trees along a 1kb chunk of genome relating three diploid samples.
+        Each diploid individual consists of two genomes (or nodes), indicated
+        by colour.
         (B) This ancestry is provided as the input to \texttt{sim\_mutations},

### 2021-08-13 paper.tex
Abstract pass

@@
 \begin{abstract}
-Coalescent simulation is a key tool in population genetics and
-has been integral to theoretical developments since its beginnings.
-Because of the ease with which the basic coalescent model can be simulated,
-a large number of different simulators have been developed. However,
-the coalescent with recombination is far more challenging to simulate,
-and, until recently, it was not possible to simulate efficiently.
-The \msprime\ software has revolutionised
-coalescent simulation, making it possible to simulate millions
-of whole genomes for the first time. We summarise the many features
-of \msprime\ 1.0, which is built around a core model of efficiently
-implementing recombination via the succinct tree sequence data
-structure. Moreover, development of \msprime\ is collaborative
-and open source, which reduces duplication of effort and promotes
+Stochastic simulation is a key tool in population genetics,
+since the models involved are often analytically intractable
+and simulation is usually the only way of obtaining ground-truth data to
+evaluate inferences.
+Because of this necessity, a large number
+of specialised simulation programs have been developed, each
+filling a particular niche, but with largely overlapping functionality
+and a substantial duplication of effort.
+Since its introduction, the \msprime\ simulator has substantially
+expanded the scope of coalescent simulation, enabling the
+simulation and efficient analysis of datasets on a far
+larger scale than previously possible.
+Here, we introduce \msprime\ version 1.0,
+which is built around a core model of efficiently
+implementing ancestry and mutation simulations using
+the succinct tree sequence data structure and \tskit\ library.
+We summarise \msprime's many features, and show that
+its performance is excellent, often many times faster
+and more memory efficient than specialised alternatives.
+These high-performance features have been built using a collaborative,
+open source development model,
+which reduces duplication of effort and promotes
 software quality via community engagement.
@@
 
-\textbf{Keywords:} Coalescent simulation, Python, Ancestral Recombination Graphs
+\textbf{Keywords:} Simulation, Coalescent, Mutations, Ancestral Recombination
+Graphs
 

### 2021-08-14 pretty_pictures.py
Try to standardise fonts

@@ -25,14 +25,20 @@ def mutated_tree():
     width = 700
     top = 50
 
-    def do_svg(ts, **kwargs):
-        colours = plt.rcParams['axes.prop_cycle'].by_key()['color']
+    colours = plt.rcParams['axes.prop_cycle'].by_key()['color']
+    print(colours)
 
-        # Sizes are probably all wrong here
+    def do_svg(ts, **kwargs):
+
+        # The page style here is just for Chromium. We shouldn't
+        # need it for other vonersion options.
         style = """\
             @media print {
               @page { margin: 0; size: 6in 2.5in}
               body { margin: 1.6cm; }
+            }
+            text {
+                font-family:DejaVuSans;
             }
             """
         for j in range(ts.num_individuals):
@@ -56,7 +62,7 @@ def mutated_tree():
     font_size = 15
 
     # I think serif is the default, and matches what we're using for labels?
-    def make_text(text, y, font_family="serif"):
+    def make_text(text, y, font_family="sans"):
         return (
             f'<text x="{width / 4}" y="{y}" font-size="{font_size}" '
             f'font-family="{font_family}" text-anchor="middle">'
@@ -83,6 +89,10 @@ def mutated_tree():
     )
     with open("illustrations/mutated_tree.svg", "w") as f:
         f.write(fig)
+
+    # When we use SVG 1.1 in the tskit output we can convert directly
+    # import cairosvg
+    # cairosvg.svg2pdf(url='illustrations/mutated_tree.svg', write_to='image.pdf')
 
 
 @click.command()

### 2021-08-14 paper.tex
Update abstract

@@
 and a substantial duplication of effort.
-Since its introduction, the \msprime\ simulator has substantially
-expanded the scope of coalescent simulation, enabling the
+Since its introduction, the \msprime\ simulator has significantly
+extended the scope of coalescent simulation, enabling the
 simulation and efficient analysis of datasets on a far
 larger scale than previously possible.
-Here, we introduce \msprime\ version 1.0,
-which is built around a core model of efficiently
-implementing ancestry and mutation simulations using
+Here, we introduce \msprime\ version 1.0, which
+efficiently implements ancestry and mutation simulations using
 the succinct tree sequence data structure and \tskit\ library.
@@
 and more memory efficient than specialised alternatives.
-These high-performance features have been built using a collaborative,
+These high-performance features have been thoroughly
+tested and validated, and built using a collaborative,
 open source development model,

### 2021-08-14 paper.tex
intro pass

@@
 process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
-models the ancestry of a set of sampled genomes. Specifically,
-the coalescent provides a mathematical description of the
+models the ancestry of a set of sampled genomes,
+providing a mathematical description of the
 genealogical tree that relates sampled genomes to one another.
-The coalescent has proved to be a powerful model,
-and it is now central to
+It has proved to be a powerful model,
+and is now central to
 population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
-Among the most impactful contributions of the coalescent since its origins
-~\citep{hudson1983testing} is that it provides a path towards more efficient
-simulation of population genetic samples than is possible using traditional,
-forwards in time simulation.
-As a result of this great utility, dozens of different simulation tools have been
-developed over the decades~\citep{carvajal2008simulation,liu2008survey,
+The coalescent is a very efficient framework for population genetic
+simulation, because it allows us to simulate the genetic ancestry for
+a sample from a Wright-Fisher population, without explicitly representing
+the population in memory or stepping through the generations.
+Indeed, \cite{hudson1983testing} independently derived the
+% Anyone disagree here? That's the impression I get from the paper.
+coalescent \emph{in order to} efficiently simulate data.
+This efficiency, and the adaptability of the model to different
+applications has lead to dozens of different simulation tools
+being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
 arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
@@
 can be simulated in $O(n)$ time~\citep{hudson1990gene},
-the coalescent with recombination, which generates multiple correlated trees, is
-far more challenging.
-% JK will need to revise this sentence, in light of what we'll be saying
-% down further about needing the SMC for very large rho.
-Programs such as Hudson's classical \ms~\citep{hudson2002generating}
-can only simulate short segments under the influence of recombination,
-and, until recently, approximate models of
-recombination~\citep{mcvean2005approximating,staab2015scrm}
-were necessary to simulate whole chromosomes.
-
+the coalescent with recombination is far more challenging,
+and methods such as
+Hudson's classical \ms~\citep{hudson2002generating}
+can only simulate short segments under the influence of recombination.
 The second major challenge to existing coalescent simulators is the
@@
 Classical simulators such as \ms\ and even fast approximate methods
-such as \scrm\ simply cannot cope with such a large number of samples.
+such as \scrm~\citep{staab2015scrm} simply cannot
+cope with such a large number of samples.
 
@@
 genetic simulation.
-This paper marks the release of \msprime\ 1.0, a state-of-the-art
-coalescent simulator both with respect to efficiency as well as feature-richness.
-Here we provide an overview of the extensive abilities of \msprime\ 1.0,
+This paper marks the release of \msprime\ 1.0.
+We provide an overview of its extensive features,
 demonstrate its performance advantages over alternative software,
-and finally lay out future plans for
-further open-source development of what we hope will become an industry
-standard piece of computational infrastructure for the genetics community.
+and discuss opportunites for ongoing
+open-source community-based development.
+
+% and finally lay out future plans for
+% further open-source development of what we hope will become an industry
+% standard piece of computational infrastructure for the genetics community.
 

### 2021-08-14 paper.tex
Tweaks

@@
 providing a mathematical description of the
-genealogical tree that relates sampled genomes to one another.
+genealogical tree that relates the samples to one another.
 It has proved to be a powerful model,
@@
 This efficiency, and the adaptability of the model to different
-applications has lead to dozens of different simulation tools
+applications has led to dozens of different simulation tools
 being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
@@
 of magnitude smaller than commonly used formats.
-The succinct tree sequence has also lead to major advances in forwards-time
+The succinct tree sequence has also led to major advances in forwards-time
 simulation~\citep{kelleher2018efficient,haller2018tree},
@@
 A major change for \msprime\ 1.0 is the introduction of the new Demography API,
-designed to address a design flaw in the \msprime\ 0.x interface which lead to
+designed to address a design flaw in the \msprime\ 0.x interface which led to
 a number of avoidable errors in downstream

### 2021-08-15 paper.tex
Tweak intro

@@
 % Anyone disagree here? That's the impression I get from the paper.
-coalescent \emph{in order to} efficiently simulate data.
+coalescent \emph{in order to} efficiently simulate data under a neutral model.
 This efficiency, and the adaptability of the model to different
@@
 peng2015genetic}.
-Two technological developments of recent years, however,
-pose major challenges to the majority of these methods.
+
+Two technological developments of recent years
+pose major challenges to the majority of existing simulation methods.
+Firstly, fourth-generation sequencing technologies have made
+complete chromosome-level assemblies possible~\citep{miga2020telomere},
 % Some citations here? Darwin tree of life maybe?
-Firstly, whole genome sequencing---and in particular fourth-generation
-sequencing technologies---have made chromosome-level assemblies
-commonplace. Thus, modelling genetic variation data as a series of unlinked loci,
-a previously reasonable assumption, is no longer defensible and
+and high quality assemblies are now available for many species.
+Thus, modelling genetic variation data as a series of unlinked loci
+is no longer a reasonable approximation, and
 we must fully account for recombination.
@@
 can be simulated in $O(n)$ time~\citep{hudson1990gene},
-the coalescent with recombination is far more challenging,
-and methods such as
+the coalescent with recombination is far more complex,
+and programs such as
 Hudson's classical \ms~\citep{hudson2002generating}
 can only simulate short segments under the influence of recombination.
-The second major challenge to existing coalescent simulators is the
-exponential increase in sample size of genetic studies. Human datasets like the
+The second challenge facing simulation methods is that
+sample sizes in genetic studies have grown very quickly
+in recent years, enabled by the precipitous fall in genome sequencing costs.
+Human datasets like the
 UK Biobank~\citep{bycroft2018genome} and
@@
 wohns2021unified},
-introduced as part of \msprime\, makes it possible to store
-such huge simulations in a few gigabytes, several orders
+originally introduced as part of \msprime, makes it possible to store
+such large simulations in a few gigabytes, several orders
 of magnitude smaller than commonly used formats.

### 2021-08-15 paper.tex
results head

@@
 \section*{Results}
-
-\jkcomment{Give a quick roadmap,
-like, we discuss the underlying tree sequence object,
-then mutations (because it's quite separate from the rest)
-and then a bunch of new features for ancestry simulation. There's no
-point in listing it out section by section.}
-
-% JK: commented out this text as it's misleading and would confuse
-% people. We should adapt it when writing the roadmap when the structure
-% has settled down.
-
-% In this section we describe the main features of \msprime\ 1.0. Methodologically,
-% there exists a logical separation between simulating coalescent genealogies and
-% adding neutral mutations onto these genealogies. Simulating genetic ancestry
-% requires methods that model the underlying reproductive process (e.g. Moran,
-% Wright-Fisher, etc.) as well as accounting for recombination of various sorts
-% (including gene conversion), demography, and selection. These methods are
-% discussed in the first N subsections below. One of the strengths of the
-% approach taken in msprime is that neutral mutations can be retrospectively
-% added to simulated genealogies. Methods to generate mutations, and thereby
-% generate genetic variation, are therefore discussed in a separate and subsequent
-% subsection. Finally, we describe the \msprime\ API, which is designed to enforce
-% the distinction between simulating genealogies and mutations; discuss the
-% interoperability provided by the use of \tskit\, the tree sequence toolkit;
-% and highlight the open source development process which allows for \msprime\
-% to be extended by addition of new features while ensuring robustness and
-% maintainability of the core codebase.
+In the following sections we describe the main features of \msprime\ 1.0,
+focusing on the aspects that are either new for this version, or in which
+our approach differs significantly from classical methods. Where appropriate,
+we benchmark \msprime\ against other simulators, but the comparisons are
+illustrative and not intended to be systematic or exhaustive. Please
+see~\cite{kelleher2016efficient} for a performance comparison of
+\msprime\ against simulators such as \ms, \msms, and \scrm.
+
+% jk: I started writing a roadmap here, but it's quite hard to do concisely
+% and you either end up leaving sections out, or just summarising them
+% one-by-one which makes it long and tedious. I think readers can either read
+% through fully if they're interested, or just skim the section titles which
+% should give them a good idea of whether they want to read it or not.
 

### 2021-08-15 paper.tex
spell check

@@
 demonstrate its performance advantages over alternative software,
-and discuss opportunites for ongoing
+and discuss opportunities for ongoing
 open-source community-based development.
@@
 replicates for 10 samples in a model
-with a single selective sweep in its history where the benefical
+with a single selective sweep in its history where the beneficial
 allele had a scaled selection coefficient of $2Ns=1000$,

### 2021-08-15 paper.tex
Tweak discussion

@@
 % Anyone disagree here? That's the impression I get from the paper.
-coalescent \emph{in order to} efficiently simulate data under a neutral model.
-This efficiency, and the adaptability of the model to different
-applications has led to dozens of different simulation tools
+coalescent \emph{in order to} efficiently simulate data,
+and he used these simulations to characterise an analytically
+intractable distribution.
+This efficiency and the great utility of simulations for a wide
+range of purposes has led to dozens of different tools
 being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
@@
 
-Two technological developments of recent years
+Two technological developments of recent years, however,
 pose major challenges to the majority of existing simulation methods.
@@
 and high quality assemblies are now available for many species.
-Thus, modelling genetic variation data as a series of unlinked loci
+Thus, modelling genetic variation data as a series of unlinked
+non-recombining loci
 is no longer a reasonable approximation, and
@@
 has greatly increased the scope of coalescent simulations,
-and it is now straightforward to simulate millions of whole chromosomes.
+and it is now straightforward to simulate millions of whole chromosomes
+for a wide range of organisms.
 The ``succinct tree sequence'' data
@@
 Fig.~\ref{fig-selection-perf} shows that \msprime\ has far better
-CPU time and memory performance than \discoal. We have previously shown
-that \msprime\ is much more efficient than \msms\ and \cosiTwo\
-when simulating without selection~\citep{kelleher2016efficient}.
-% jk - this just seems clumsy to me: surely the conclusion is clear?
-% and we therefore believe \msprime\ to be the most efficient
-% coalescent simulator when simulating sweeps.
+CPU time and memory performance than \discoal.
 
@@
 % We can simulate so much more now
-This release makes \msprime\ able to produce simulations
-with a great deal more biological realism, without a loss in speed and efficiency.
-This ability will allow researchers to perform more robust
-power analyses, more reliably test new methods,
-and carry out more reliable inferences,
-without worrying that gene conversion or back mutation might be sabotaging
-their applications to real data.
+The 1.0 release of \msprime\ marks a major increase in the
+biological realism of simulations
+and the breadth of available features, without a loss of efficiency.
+These abilities will allow researchers to perform more robust power analyses,
+more reliably test new methods,
+carry out more reliable inferences,
+and explore the properties of more theoretical models.
 These advances have only been possible thanks to a distributed,
@@
 and the work of many people.
-We hope the tool continues to be useful to many.
+% We hope the tool continues to be useful to many.
 
@@
 % More citations?
-\citep{siepel2019challenges,harris2020array,gardner2021sustained}
+\citep[e.g.][]{siepel2019challenges,harris2020array,gardner2021sustained}
 that investing in
@@
 
-The tree sequence developed for \msprime\ allows users a view of not only
+The succinct tree sequence data structure developed
+for \msprime\ allows users a view of not only
 genetic variation but also the genetic ancestry that produced that variation.

### 2021-08-16 paper.tex
Editing pass

@@
 and a substantial duplication of effort.
-Since its introduction, the \msprime\ simulator has significantly
-extended the scope of coalescent simulation, enabling the
-simulation and efficient analysis of datasets on a far
-larger scale than previously possible.
 Here, we introduce \msprime\ version 1.0, which
-efficiently implements ancestry and mutation simulations using
+efficiently implements ancestry and mutation simulations based on
 the succinct tree sequence data structure and \tskit\ library.
@@
 simulation, because it allows us to simulate the genetic ancestry for
-a sample from a Wright-Fisher population, without explicitly representing
+a sample from an idealised population model, without explicitly representing
 the population in memory or stepping through the generations.
@@
 coalescent \emph{in order to} efficiently simulate data,
-and he used these simulations to characterise an analytically
+and used these simulations to characterise an analytically
 intractable distribution.
-This efficiency and the great utility of simulations for a wide
-range of purposes has led to dozens of different tools
+This inherent efficiency, and the great utility of simulations for a wide
+range of purposes, has led to dozens of different tools
 being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
@@
 Two technological developments of recent years, however,
-pose major challenges to the majority of existing simulation methods.
+pose major challenges to most existing simulation methods.
 Firstly, fourth-generation sequencing technologies have made
@@
 we must fully account for recombination.
-While a genealogical tree relating $n$ samples in the single-locus coalescent
+However, while a genealogical tree relating $n$ samples in the single-locus coalescent
 can be simulated in $O(n)$ time~\citep{hudson1990gene},
@@
 The 1.0 release of \msprime\ marks a major increase in the
-biological realism of simulations
-and the breadth of available features, without a loss of efficiency.
+breadth of available features and the
+potential biological realism of simulations.
 These abilities will allow researchers to perform more robust power analyses,
@@
 carry out more reliable inferences,
-and explore the properties of more theoretical models.
+and more throughly explore the properties of theoretical models.
+Despite this complexity and generality,
+\msprime's performance is state-of-the-art
+and all features are thoroughly tested and statistically validated.
 These advances have only been possible thanks to a distributed,
@@
 significant software quality issues at all stages. This is unsurprising, since
-the majority of simulation software development is performed by students, who
-are themselves learning to code, often without formal training in software
-development.
+the majority of simulation software development is performed by students,
+often without formal training in software development.
 Although this ``Haldane's sieve'' of software has produced many good tools
@@
 established development practices and infrastructure.
-% JK: I'm going out on a limb here - do folks agree?
-% PLR- personally I have learned SO MUCH, and I'm passing it on to others
 Software development in a welcoming community,
@@
 The succinct tree sequence data structure developed
-for \msprime\ allows users a view of not only
-genetic variation but also the genetic ancestry that produced that variation.
+for \msprime\ provides a view of not only
+genetic variation, but also the genetic ancestry that produced that variation.
 Recent breakthroughs in methods to infer genetic ancestry in recombining
-organisms~\citep{kelleher2019inferring,
+organisms~\citep{rasmussen2014genome,kelleher2019inferring,
 speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
@@
 on the manuscript.
-
-% TODO alphabetise this by author I guess
+% alphabetised by author. Names spelled out if there's an initial collision
+ADK was supported by NIH awards R01GM117241 and R01HG010774.
+BE was supported by  DFG grant  273887127 through Priority Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
+FB is funded by the Deutsche Forschungsgemeinschaft EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
+Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
+Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
+Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
 Jerome Kelleher is supported by the Robertson Foundation.
-Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
-ADK was supported by NIH awards R01GM117241 and R01HG010774.
-Franz Baumdicker is funded by the Deutsche Forschungsgemeinschaft
-EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
-Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
-BE was supported by  DFG grant  273887127 through Priority
-Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research
-Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
-Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of
-Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
-SG acknowledges funding from the Canada Research Chairs Program, from the
-Canadian Institutes of Health Research PJT 173300, and from the Canadian
-Foundation for Innovation.
+SG acknowledges funding from the Canada Research Chairs Program, from the Canadian Institutes of Health Research PJT 173300, and from the Canadian Foundation for Innovation.
 

### 2021-08-16 paper.tex
Some comments on the mutation appendix
I have provided comments as inline LaTex comments. I think this is mostly for @petrelharp

@@
 this is done separately for each distinct region.
+% I am a bit confused with the following lines - as the section in the main paper says, algo is done on edges,
+% but this is skipped here and the way its written it makes it confusing - #edges are propto #marginal_trees
+% so there should be no difference (the way I read this part), but in fact the difference on working with edges
+% or each tree and its branches can be large
 The algorithm scales with the number of edges in the tree sequence.
@@
 Then, each mutation on the tree is considered in turn,
+% I don't understand what is meant by "older mutations are considered first"?
+% Aren't past/older mutations irrelevant given the current state (=parental allele) and a transition matrix?
 in order so that older mutations are considered first,
@@
 (which may be the ancestral allele or the derived allele of a previous mutation)
+% What is meant by "some randomness"?
 and some randomness.
@@
 
+% which algo do you refer above? InfiniteAlleles or the general version above the above paragraph? Better to spell it out.
 Since the algorithm above applies mutations at a single rate
@@
 work in this way: mutations are placed at the maximum mutation rate,
+% I think I understand what you want to say here, but it would be clearer if an example would have been given - similar to the next section.
 and then some silent mutations will result.

### 2021-08-16 paper.tex
edits to mutation appendix

--- 2021-08-16 paper.tex
@@
 The algorithm scales with the number of edges in the tree sequence.
+
 Since the number of edges is proportional to the number of marginal trees,
-an algorithm that applies mutations separately to each marginal tree
-would take at least as many steps,
-and would require additional bookkeeping necessary to place mutations uniformly on a tree.
+this algorithm is $O(T)$,
+where $T$ is the number of trees.
+An alternative algorithm would apply mutations separately to each tree.
+Since the number of branches per tree is proportional to the number of leaves,
+the simplest implementation would be $O(T S)$,
+where $S$ is the number of samples.
+Any more efficient tree-by-tree algorithm would have to make use of the fact
+that structure is shared between adjacent trees---effectively, therefore,
+working with shared edges.
 
@@
 Then, each mutation on the tree is considered in turn,
-in order so that older mutations are considered first,
-and a derived allele is chosen based on the parental allele
-(which may be the ancestral allele or the derived allele of a previous mutation)
-and some randomness.
+and a derived allele is randomly chosen based on the parental allele
+(which may be the ancestral allele or the derived allele of a previous mutation).
 Finally, information about the mutations are recorded in the site and mutation tables
@@
 
-Since the algorithm above applies mutations at a single rate
+The distribution of ancestral alleles
+is used to choose the allele present at the root of the tree at each mutated site,
+i.e., the \texttt{root\_distribution}.
+Mutation models with a finite possible set of alleles have a natural choice
+for this distribution---the \emph{stationary distribution} of the mutation process.
+(All mutation models are Markovian,
+so this may be found as the top left eigenvector of the mutation matrix.)
+This is the default in most models,
+except, e.g., the \texttt{BinaryMutationModel}, whose alleles are 0 and 1
+and always labels the ancestral allele ``0''.
+However, mutational processes are not in general stationary,
+so we often allow different root distribution to be specified.
+
+Since the general algorithm above applies mutations at a single rate
 independent of ancestral state,
@@
 if not all equilibrium frequencies are the same.
-Although the presence of silent mutations may at first be surprising
+The presence of silent mutations may at first be surprising
 (although they don't affect calculations of statistics in any way),
-there is a good reason to leave them in:
+but there is a good reason to leave them in:
 to allow layering of different mutation models.

### 2021-08-17 paper.tex
nucleotides in example

@@
 i.e., mutations in which the derived allele is equal to the parental allele.
-To illustrate this, consider a mutation model on the binary alphabet
-$\{X, Y\}$ in which $X$ mutates to $Y$ at rate $\alpha$
-and $Y$ mutates to $X$ at rate $\beta$, with $\beta < \alpha$.
+To illustrate this, consider a mutation model
+in which $A$ or $T$ mutates to a randomly chosen different nucleotide at rate $\alpha$
+and $C$ or $G$ mutates at rate $\beta$, with $\beta < \alpha$.
 To implement this, first place mutations at the largest total rate, which is $\alpha$.
-Then, at each site, choose an ancestral allele from the ``root distribution''
-(a natural choice would be to choose from the stationary distribution,
-$X$ with probability $\beta / (\alpha + \beta)$),
+Then, at each site, choose an ancestral allele from the root distribution,
 and for each mutation, choose a derived allele as follows:
-if the parental allele is $X$, then choose $Y$;
-if the parental allele is $Y$, then choose $X$ with probability $\beta/(\alpha + \beta)$,
-and choose $Y$ otherwise.
+if the parental allele is $A$ or $T$, then choose a random derived allele
+different than the parental allele;
+if the parental allele is $C$ or $G$,
+then choose the derived allele to be equal to the parent allele
+with probability $\beta/(\alpha + \beta)$,
+and randomly choose a different nucleotide otherwise.
 This produces the correct distribution by Poisson thinning:
@@
 work in this way: mutations are placed at the maximum mutation rate,
-% I think I understand what you want to say here, but it would be clearer if an example would have been given - similar to the next section.
 and then some silent mutations will result.

### 2021-08-17 paper.tex
Updates from Overleaf

--- 2021-08-16 paper.tex
@@
 population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
-The coalescent is a very efficient framework for population genetic
+The coalescent is an efficient framework for population genetic
 simulation, because it allows us to simulate the genetic ancestry for
@@
 data. Although \msprime\ also supports simulating both ancestry and mutations,
-the two aspects are functionally independent within the software.
-Both ancestry and mutation simulators are present in \msprime\
+the two aspects are functionally independent within the software; both ancestry and mutation simulators are present in \msprime\
 for reasons of convenience and history,
@@
 output formats from the mutation generation algorithms.
-Given these timings, it seems quite unlikely
+Given these timings, it seems unlikely
 that generating mutations with \msprime\ would be a bottleneck in any
@@
 approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
-In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
+In \msprime\ 1.0, recombination rates can vary along a chromosome, allowing
 us to simulate recombination hotspots and patterns
@@
 have efficient implementations of approximate coalescent with recombination
-models, we recommend using SMC based methods such as \scrm,
-particularly for small sample sizes.
+models, in these cases we recommend using SMC based methods such as \scrm,
+particularly if sample sizes are small.
 In practice, to predict the running time of a given
@@
 and recombination events increase their number.
-However, with long genomes
+However, with long genomes,
 only a small fraction of the common ancestor events
-will involve overlapping segments of ancestry,
+will involve overlapping segments of ancestry
 and lead to coalescence in the marginal trees.
@@
 which usually happens in the quite recent
-past~\cite{bhaskar2014distortion}.
+past~\citep{bhaskar2014distortion}.
 We therefore support changing the simulation model during a simulation
@@
 We also view contributions to documentation as equally important to the project
-as writing code or designing methods---what use would it be
+as writing code or designing methods: what use would it be
 to write reliable, stable software if no-one used it?
@@
 this is done separately for each distinct region.
-% I am a bit confused with the following lines - as the section in the main paper says, algo is done on edges,
-% but this is skipped here and the way its written it makes it confusing - #edges are propto #marginal_trees
-% so there should be no difference (the way I read this part), but in fact the difference on working with edges
-% or each tree and its branches can be large
 The algorithm scales with the number of edges in the tree sequence.
@@
 Then, each mutation on the tree is considered in turn,
-% I don't understand what is meant by "older mutations are considered first"?
-% Aren't past/older mutations irrelevant given the current state (=parental allele) and a transition matrix?
 in order so that older mutations are considered first,
@@
 (which may be the ancestral allele or the derived allele of a previous mutation)
-% What is meant by "some randomness"?
 and some randomness.
@@
 
-% which algo do you refer above? InfiniteAlleles or the general version above the above paragraph? Better to spell it out.
 Since the algorithm above applies mutations at a single rate

### 2021-08-17 paper.tex
Tweaks to mutations appendix

@@
 but there are occasions such as visualisation or computing likelihoods (see
-below) in which is useful. We therefore provide
+below) in which it is useful. We therefore provide
 the \texttt{record\_full\_arg} option in \msprime\
@@
 If an edge spans a region of the genome with more than one mutation rate,
-this is done separately for each distinct region.
-The algorithm scales with the number of edges in the tree sequence.
-
-Since the number of edges is proportional to the number of marginal trees,
-this algorithm is $O(T)$,
-where $T$ is the number of trees.
-An alternative algorithm would apply mutations separately to each tree.
-Since the number of branches per tree is proportional to the number of leaves,
-the simplest implementation would be $O(T S)$,
-where $S$ is the number of samples.
-Any more efficient tree-by-tree algorithm would have to make use of the fact
-that structure is shared between adjacent trees---effectively, therefore,
-working with shared edges.
+this is done separately for each sub-region on which the mutation rate is constant.
+Since each edge is processed independently, the algorithm
+scales linearly with the number of edges in the tree sequence.
 
@@
 if the parental allele is $A$ or $T$, then choose a random derived allele
-different than the parental allele;
+different to the parental allele;
 if the parental allele is $C$ or $G$,
@@
 The presence of silent mutations may at first be surprising
-(although they don't affect calculations of statistics in any way),
+(although they are fully accounted for in downstream calculations of statistics
+by \tskit),
 but there is a good reason to leave them in:
@@
 \emph{independently} of the parent (so that 1/4 of mutations will be silent).
-If so -- and, more generally, if the first mutational process put down is state-independent --
-then the result of sequentially applying the two mutation models
+If so---and, more generally, if the first mutational process put down is
+state-independent---then the result of sequentially applying the two mutation models
 is equivalent to the simultaneous model.
@@
 \end{eqnarray*}
-where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$). Further
-$\mu(p)$ is
+where
 \begin{eqnarray*}
-	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
+	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}.
 \end{eqnarray*}
+Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$).
 

### 2021-08-17 paper.tex
Add line numbers
Apply @mmatschiner's comments.
Closes #168

@@
 \usepackage{tikz}
+\usepackage[right]{lineno}
 \usetikzlibrary{calc,positioning}
@@
 
+\linenumbers
 \section*{Introduction}
@@
 % Any more?
-and \ms's fork \msHOT~\citep{hellenthal2007mshot},
+and \ms's fork \msHOT~\citep{hellenthal2007mshot}
 has in turn been forked to output information on local
@@
 samples from a population of thousands of individuals takes seconds,
-while samples from a population of tens of thousands takes minutes.
+while samples from a population of tens of thousands take minutes.
 Simulating whole chromosomes for many species is very fast,

### 2021-08-17 evaluation/plot.py

@@ -39,7 +39,7 @@ def mutations_perf():
     dfL = df[df.L == 10]
     for rate in rates:
         dfr = dfL[dfL.rate == rate].sort_values("n")
-        ax1.plot(dfr.n, dfr.time, label=f"rate={rate}")
+        ax1.plot(dfr.n, dfr.time, label=f"mutation rate={rate}")
 
     ax2.set_xlabel("Sequence length (Megabases)")
     ax1.set_ylabel("Time (seconds)")
@@ -47,7 +47,7 @@ def mutations_perf():
     dfn = df[df.n == 1000].sort_values("L")
     for rate in rates:
         dfr = dfn[dfn.rate == rate]
-        ax2.plot(dfr.L, dfr.time, label=f"rate={rate}")
+        ax2.plot(dfr.L, dfr.time, label=f"mutation rate={rate}")
     print(dfn[dfn.L == 100])
 
     ax1.legend()

### 2021-08-17 paper.tex
Jared's second affil

@@
 \author[9]{Castedo E. Ellerman}
-\author[10]{Jared G. Galloway}
-\author[11,12]{Ariella L. Gladstein}
-\author[13]{Gregor Gorjanc}
-\author[14]{Bing Guo}
+\author[10,11]{Jared G. Galloway}
+\author[12,13]{Ariella L. Gladstein}
+\author[14]{Gregor Gorjanc}
+\author[15]{Bing Guo}
 \author[7]{Ben Jeffery}
-\author[15]{Warren W. Kretzschmar}
-\author[16]{Ivan Krukov}
+\author[16]{Warren W. Kretzschmar}
+\author[17]{Ivan Krukov}
 \author[2]{Konrad Lohse}
-\author[17]{Michael Matschiner}
-\author[16]{Dominic Nelson}
-\author[18]{Nathaniel S. Pope}
-\author[19]{Consuelo D. Quinto-Cort\'es}
+\author[18]{Michael Matschiner}
+\author[17]{Dominic Nelson}
+\author[19]{Nathaniel S. Pope}
+\author[20]{Consuelo D. Quinto-Cort\'es}
 \author[10]{Murillo F. Rodrigues}
-\author[20]{Kumar Saunack}
-\author[21]{Thibaut Sellinger}
-\author[22]{Kevin Thornton}
+\author[21]{Kumar Saunack}
+\author[22]{Thibaut Sellinger}
+\author[23]{Kevin Thornton}
 \author[9]{Hugo van Kemenade}
@@
 % Senior Authors
-\author[15,$\dagger$]{Simon Gravel}
+\author[16,$\dagger$]{Simon Gravel}
 \author[10,$\dagger$]{Andrew D. Kern}
-\author[23,$\dagger$]{Jere Koskela}
-\author[10,24,$\dagger$]{Peter L. Ralph}
+\author[24,$\dagger$]{Jere Koskela}
+\author[10,25,$\dagger$]{Peter L. Ralph}
 
@@
     University of Oregon}
+% Jared
+\affil[11]{Computational Biology Program, Fred Hutchinson Cancer Research
+Center, Seattle, WA 98102, USA}
 % Ariella
-\affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
-\affil[12]{Embark Veterinary, Inc., Boston}
+\affil[12]{Department of Genetics, University of North Carolina at Chapel Hill}
+\affil[13]{Embark Veterinary, Inc., Boston}
 % Gregor
-\affil[13]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
+\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
     University of Edinburgh}
 % Bing
-\affil[14]{Institute for Genome Sciences, University of Maryland School of
+\affil[15]{Institute for Genome Sciences, University of Maryland School of
 Medicine, Baltimore, MD}
 % Winni
-\affil[15]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
+\affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
 % Simon, Dom, Ivan
-\affil[16]{Department of Human Genetics, McGill University}
+\affil[17]{Department of Human Genetics, McGill University}
 % Michael
-\affil[17]{Natural History Museum, University of Oslo}
+\affil[18]{Natural History Museum, University of Oslo}
 % Nathaniel
-\affil[18]{Department of Entomology, Pennsylvania State University}
+\affil[19]{Department of Entomology, Pennsylvania State University}
 % Consuelo
-\affil[19]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
+\affil[20]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
     Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
 % Saunack
-\affil[20]{IIT Bombay, India}
+\affil[21]{IIT Bombay, India}
 % Thibaut
-\affil[21]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+\affil[22]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
 % Kevin
-\affil[22]{Ecology and Evolutionary Biology, University of California Irvine}
+\affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
 % Jere
-\affil[23]{Department of Statistics, University of Warwick}
+\affil[24]{Department of Statistics, University of Warwick}
 % Peter (second affil)
-\affil[24]{Department of Mathematics, University of Oregon}
+\affil[25]{Department of Mathematics, University of Oregon}
 

### 2021-08-17 paper.tex
Bjarki's comments.
Closes #166

@@
 
+\linenumbers
 \title{Efficient ancestry and mutation simulation with msprime 1.0}
@@
 
-\linenumbers
 \section*{Introduction}
@@
 The coalescent
-process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
+process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
+tajima1983evolutionary}
 models the ancestry of a set of sampled genomes,
@@
 and could be split into separate packages.
-The mature and efficient C and Python interfaces for
+The efficient C and Python interfaces for
 \tskit\ make it straightforward to add further information to
@@
 We have added two multiple merger coalescent models in \msprime\ 1.0, the
-Beta-coalescent and ``Dirac''-coalescent, allowing us to simulate
-such models with recombination efficiently for the first time.
+Beta-coalescent~\citep{schweinsberg03} and
+``Dirac''-coalescent~\citep{BBE13}, allowing us to efficiently simulate
+such models with recombination for the first time.
 These simulation models have been extensively validated against
@@
 \appendix
-
-\section*{Mutation generation}
+\section*{Appendix}
+
+\subsection*{Mutation generation}
 \label{app-mutation-algorithm}
@@
 
-\section*{Likelihood calculations}
+\subsection*{Likelihood calculations}
 \label{app-likelihoods}
@@
 \label{app-multiple-mergers}
-\section*{Multiple merger coalescent model}
+\subsection*{Multiple merger coalescent model}
 
@@
 \label{app-sweeps}
-\section*{Selective sweep trajectories}
+\subsection*{Selective sweep trajectories}
 We have implemented the coalescent with selective sweeps as a structured

### 2021-08-17 paper.tex

--- 2021-08-17 paper.tex
@@
 \usepackage{tikz}
-\usepackage[right]{lineno}
 \usetikzlibrary{calc,positioning}
@@
 % Any more?
-and \ms's fork \msHOT~\citep{hellenthal2007mshot}
+and \ms's fork \msHOT~\citep{hellenthal2007mshot},
 has in turn been forked to output information on local
@@
 samples from a population of thousands of individuals takes seconds,
-while samples from a population of tens of thousands take minutes.
+while samples from a population of tens of thousands takes minutes.
 Simulating whole chromosomes for many species is very fast,
@@
 If an edge spans a region of the genome with more than one mutation rate,
-this is done separately for each sub-region on which the mutation rate is constant.
-Since each edge is processed independently, the algorithm
-scales linearly with the number of edges in the tree sequence.
+this is done separately for each distinct region.
+The algorithm scales with the number of edges in the tree sequence.
+
+Since the number of edges is proportional to the number of marginal trees,
+this algorithm is $O(T)$,
+where $T$ is the number of trees.
+An alternative algorithm would apply mutations separately to each tree.
+Since the number of branches per tree is proportional to the number of leaves,
+the simplest implementation would be $O(T S)$,
+where $S$ is the number of samples.
+Any more efficient tree-by-tree algorithm would have to make use of the fact
+that structure is shared between adjacent trees---effectively, therefore,
+working with shared edges.
 
@@
 if the parental allele is $A$ or $T$, then choose a random derived allele
-different to the parental allele;
+different than the parental allele;
 if the parental allele is $C$ or $G$,
@@
 The presence of silent mutations may at first be surprising
-(although they are fully accounted for in downstream calculations of statistics
-by \tskit),
+(although they don't affect calculations of statistics in any way),
 but there is a good reason to leave them in:
@@
 \emph{independently} of the parent (so that 1/4 of mutations will be silent).
-If so---and, more generally, if the first mutational process put down is
-state-independent---then the result of sequentially applying the two mutation models
+If so -- and, more generally, if the first mutational process put down is state-independent --
+then the result of sequentially applying the two mutation models
 is equivalent to the simultaneous model.
@@
 \end{eqnarray*}
-where
+where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$). Further
+$\mu(p)$ is
 \begin{eqnarray*}
-	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}.
+	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
 \end{eqnarray*}
-Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$).
 

### 2021-08-17 evaluation/plot.py
Tweak plots
Closes #171

@@ -39,7 +39,7 @@ def mutations_perf():
     dfL = df[df.L == 10]
     for rate in rates:
         dfr = dfL[dfL.rate == rate].sort_values("n")
-        ax1.plot(dfr.n, dfr.time, label=f"mutation rate={rate}")
+        ax1.plot(dfr.n, dfr.time, label=f"Mutation rate={rate}")
 
     ax2.set_xlabel("Sequence length (Megabases)")
     ax1.set_ylabel("Time (seconds)")
@@ -47,7 +47,7 @@ def mutations_perf():
     dfn = df[df.n == 1000].sort_values("L")
     for rate in rates:
         dfr = dfn[dfn.rate == rate]
-        ax2.plot(dfr.L, dfr.time, label=f"mutation rate={rate}")
+        ax2.plot(dfr.L, dfr.time, label=f"Mutation rate={rate}")
     print(dfn[dfn.L == 100])
 
     ax1.legend()
@@ -132,7 +132,7 @@ def sweeps_perf():
     ax1.annotate(
         f"{round(largest_value * 60)} seconds",
         textcoords="offset points",
-        xytext=(-30, 5),
+        xytext=(-38, 5),
         xy=(largest_L, largest_value),
         xycoords="data",
     )

### 2021-08-18 paper.tex
Update paper.tex
I think some previous reordering must have messed up definitions: we were pointing to alpha in an equation where it could not be found. Reordered for a more logical flow.
Update paper.tex
move the period (with github interface)

@@
 If an edge spans a region of the genome with more than one mutation rate,
-this is done separately for each distinct region.
+this is done separately for each sub-region on which the mutation rate is constant.
 The algorithm scales with the number of edges in the tree sequence.
@@
 \end{eqnarray*}
-where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$). Further
-$\mu(p)$ is
+where 
 \begin{eqnarray*}
-	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
+	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}.
 \end{eqnarray*}
+Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$).
 

### 2021-08-18 paper.tex
Address comment re silent mutations

--- 2021-08-17 paper.tex
@@
 The presence of silent mutations may at first be surprising
-(although they are fully accounted for in downstream calculations of statistics
-by \tskit),
 but there is a good reason to leave them in:
@@
 and makes the model closer to state-independent.
+
+Silent mutations are fully supported by \tskit, which correctly accounts for
+their presence when computing statistics and performing other operations.
+For example, silent mutations have no effect on calculations of nucleotide
+site diversity.
 

### 2021-08-18 paper.tex
Add Wiuf & Hein ref for trapped material.

@@
 When this happens, the merged lineage carries ``trapped'' genetic material
-that is not ancestral to any samples, but where recombinations can still occur.
+that is not ancestral to any samples, but where recombinations can still
+occur~\citep{wiuf1999recombination}.
 The SMC approximations work by disallowing common ancestor events

### 2021-08-18 paper.tex

--- 2021-08-18 paper.tex
@@
 the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
-per-generation recombination probability \texttt{r} and population size \texttt{N}.
+per-generation recombination probability \texttt{r} and population size \texttt{N} \citep[e.g.][equation (1)]{kuhner2000maximum}.
 Specifically, the function returns the logarithm of
@@
 $q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
-of time intervals which end in a recombination, and $g_i$ is the corresponding
-\emph{gap}:  the length of material in which the recombination at the end of the
-$i$th period could have taken place without affecting the resulting genealogy.
-A recombination in ancestral material will always have $g_i = 1$, while a recombination
-in trapped non-ancestral material can result in other values of $g_i$.
+of time intervals which end in a recombination, and $g_i$ is the corresponding \emph{gap}:
+the length of contiguous non-ancestral material around the link at which
+the recombination in question took place. The gap indicates the number of links
+(or length of genome in a continuous model) at which a recombination would result
+in exactly the observed pattern of ancestral material in the ARG.
 For a continuous model of the genome and a recombination in ancestral material,

### 2021-08-18 paper.tex
Rework sweeps section
Closes #170

--- 2021-08-18 paper.tex
@@
 in which lineages may transit between favoured and unfavoured backgrounds through
-recombination. This allows for many selective sweep scenarios to be simulated
+recombination. This approach allows for many selective sweep scenarios to be simulated
 efficiently, including recurrent, partial, and soft selective sweeps.
-However in general this
-efficiency comes at the cost of flexibility in comparison to forwards in
+However this efficiency comes at the cost of flexibility in comparison to forwards in
 time simulation.
@@
 
-Following the approach taken in \discoal, we
-have implemented selective sweeps in \msprime\ by generating
-sweep trajectories from a jump process approximation to the conditional diffusion
+We have implemented selective sweeps in \msprime\ as a structured
+coalescent process~\citep{braverman1995hitchhiking},
+following the approach used in \discoal. The first
+step in simulating a sweep is to generate
+an allele frequency trajectory
+from a jump process approximation to the conditional diffusion
 of an allele bound for
-fixation~\citep{coop2004ancestral}---see the Appendix for more details.
-This implementation facilitates future changes to generalise sweep models to
-changing population sizes as well as selection from standing variation.
-Moreover because this implements the basic infrastructure for the structured
+fixation~\citep{coop2004ancestral}.
+The jump process moves back in time following the beneficial
+allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
+the origination of the allele at $p=1/(2N)$, tracking time in small
+increments $\delta t$. Then, given the frequency $p$ at time $t$,
+the frequency $p^\prime$ at time $t + \delta t$ is given by
+\begin{equation*}
+p^\prime =
+\begin{cases}
+    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
+    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
+\end{cases}
+\end{equation*}
+where
+\begin{eqnarray*}
+	\mu(p)=\frac{\alpha p (1 - p)}{\tanh(\alpha(1-p))}.
+\end{eqnarray*}
+Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
+can be generalised straightforwardly to include arbitrary dominance.
+We can also define trajectories to model nuetral alleles and soft
+selective sweeps, which we plan as future additions to \msprime.
+
+Then, given a randomly generated allele frequency trajectory under
+the above model, the simulation of a sweep works by assigning lineages to
+two different structured coalescent ``labels'', based on
+whether they carry the beneficial allele.
+The allele frequency trajectory determines the relative sizes
+of the ``populations'' in these labels over time, and therefore
+the rates at which various events occur.
+Common ancestor events can then only merge lineages from \emph{within}
+a label, but lineages can transfer from one label to the other
+(i.e., from the advantageous to disadvantageous backgrounds,
+and vice versa) as a result of recombination events.
+Once we have reached the end of the simulated trajectory the sweep
+is complete, and we remove the structured coalescent labels.
+Simulation may then resume under any other ancestry model.
+
+Fig.~\ref{fig-selection-perf} compares the performance of \msprime\
+and \discoal\ under a simple sweep model, and shows
+that \msprime\ has far better
+CPU time and memory performance.
+Since our implementation includes the basic infrastructure for the structured
 coalescent, adding support for similar situations,
-such as inversions~\citep{peischl2013sequential}, is straightforward.
-Fig.~\ref{fig-selection-perf} shows that \msprime\ has far better
-CPU time and memory performance than \discoal.
+such as inversions~\citep{peischl2013sequential}, should be straightforward.
 
@@
 
-\label{app-sweeps}
-\subsection*{Selective sweep trajectories}
-We have implemented the coalescent with selective sweeps as a structured
-coalescent process (as in~\citep{braverman1995hitchhiking}). During the
-``sweep phase'' of the simulation the rate of coalescence and recombination
-is tracked separately in lineages linked to the beneficial allele or not,
-with those rates depending on the frequency of the beneficial allele in
-the population at a given time. The allele frequency trajectory of the
-sweeping allele is modelled using a jump process that approximates
-the conditional diffusion of the allele, conditioning on its eventual
-fixation~\citep{coop2004ancestral}.
-The jump process moves back in time following the beneficial
-allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
-the origination of the allele at $p=1/2N$, and we track time in small
-increments $\delta t$. This works such that the allele frequency at
-the next time step back in time, call it $p^\prime$ is
-
-\begin{eqnarray*}
-	p'=\left\{ \begin{array}{ll}
-	    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
-	    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
-	\end{array} \right.
-\end{eqnarray*}
-where
-\begin{eqnarray*}
-	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}.
-\end{eqnarray*}
-Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$).
-
-While our current implementation only deals with genic selection,
-we plan in the future to add sweeps with arbitrary dominance. Further planned
-additions include trajectories of purely neutral alleles, as well as trajectories
-that model soft selective sweeps.
-
 \end{document}

### 2021-08-18 paper.tex
Updates from review

@@
 
-We have implemented selective sweeps in \msprime\ as a structured
-coalescent process~\citep{braverman1995hitchhiking},
-following the approach used in \discoal. The first
-step in simulating a sweep is to generate
-an allele frequency trajectory
-from a jump process approximation to the conditional diffusion
-of an allele bound for
+Selective sweeps are implemented in the coalescent as a two
+step-process: first generating an allele frequency trajectory,
+and then simulating a structured coalescent process conditioned
+on that trajectory.
+Following \discoal, we generate sweep trajectories in \msprime\ using a
+jump process approximation to the conditional
+diffusion of an allele bound for
 fixation~\citep{coop2004ancestral}.
@@
 CPU time and memory performance.
-Since our implementation includes the basic infrastructure for the structured
-coalescent, adding support for similar situations,
+Since our implementation uses the abstract label system mentioned above,
+adding support for similar situations,
 such as inversions~\citep{peischl2013sequential}, should be straightforward.

### 2021-08-19 paper.tex

@@
 in the Python and C languages~\citep{tskit2021tskit}.
+The availability of \tskit\ as a liberally licensed (MIT) open source
+toolkit has enabled several other
+projects~\citep[e.g.][]{kelleher2019inferring,haller2019slim,wohns2021unified,
+de2021geonomics} to take advantage
+of the same efficient data structures used in \msprime,
+and we hope that many more will follow.
 While a full discussion of tree sequences and the capabilities
@@
 retained---at scale---is unprecedented.
-This interoperability indicates an opportunity for
-other forward genetic simulators ~\citep[e.g.][]{Gaynor2021AlphaSimR}
+There may be an opportunity for
+other forward genetic simulators~\citep[e.g.][]{gaynor2021alphasimr}
 to leverage the tree sequence data format and associated tools.

### 2021-08-19 paper.tex
sieve fixup, closes #162

@@
 often without formal training in software development.
-Although this ``Haldane's sieve'' of software has produced many good tools
+The result resembles Haldane's sieve for new mutations:
+many new pieces of software stay permanently on a dusty shelf of supplementary materials,
+while some of those that prove particularly useful when new (like dominant alleles)
+are quickly adopted.
+Although this has produced many good tools
 and enabled decades of research,

### 2021-08-20 paper.tex
Konrad's ARG clarifations.

@@
 Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
-as a representation of the coalescent with recombination
-stochastic process as a graph.
+to represent the stochastic process of the coalescent with recombination as a graph.
 This formulation is complementary to Hudson's earlier

### 2021-08-21 paper.tex
Bjarki's comments
Closes #181

@@
 attractive and both analytic results and empirical
-inference~\citep{griffiths1994sampling, galtier2000detecting,
+inference~\citep{griffiths1994sampling, birkner2009modified,galtier2000detecting,
 bunnefeld2015inferring} have been developed under this model, there has
@@
 can be generalised straightforwardly to include arbitrary dominance.
-We can also define trajectories to model nuetral alleles and soft
+We can also define trajectories to model neutral alleles and soft
 selective sweeps, which we plan as future additions to \msprime.
@@
 \end{equation}
-where $a \geq 0$ is a constant and $\Lambda$ is a finite measure on the unit interval without an atom at zero \citep{DK99,P99,S99}.
+where $\one{A} := 1$ if $A$ holds, and zero otherwise,
+$a \geq 0$ is a constant, and $\Lambda$ is a finite measure on the
+unit interval without an atom at zero \citep{DK99,P99,S99}.
 There is also a larger class of simultaneous multiple merger coalescents involving

### 2021-08-21 paper.tex
Wilder's comments
Closes #182

@@
 \author[9]{Hugo van Kemenade}
-\author[7]{Anthony W. Wohns}
+\author[7,24]{Anthony W. Wohns}
 \author[7]{H. Yan Wong}
@@
 \author[10,$\dagger$]{Andrew D. Kern}
-\author[24,$\dagger$]{Jere Koskela}
-\author[10,25,$\dagger$]{Peter L. Ralph}
+\author[25,$\dagger$]{Jere Koskela}
+\author[10,26,$\dagger$]{Peter L. Ralph}
 
@@
 \affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
+% Wilder 2
+\affil[24]{Broad Institute of MIT and Harvard}
 % Jere
-\affil[24]{Department of Statistics, University of Warwick}
+\affil[25]{Department of Statistics, University of Warwick}
 % Peter (second affil)
-\affil[25]{Department of Mathematics, University of Oregon}
+\affil[26]{Department of Mathematics, University of Oregon}
 
@@
 
-The software ecosystem built around \tskit\ is stable and mature;
-simulators such as
+The software ecosystem built around \tskit\ is stable, mature and
+rapidly growing.
+Simulators such as
 \fwdpy~\citep{thornton2014cpp},
@@
 carry out more reliable inferences,
-and more throughly explore the properties of theoretical models.
+and more thoroughly explore the properties of theoretical models.
 Despite this complexity and generality,
 \msprime's performance is state-of-the-art
-and all features are thoroughly tested and statistically validated.
+and all features are extensively tested and statistically validated.
 These advances have only been possible thanks to a distributed,

### 2021-08-23 paper.tex

@@
 edge table describes how these genomes are related along a
-chromosome, and defines the marginal trees at each position.
+chromosome, and defines the genealogical tree at each position.
 The site and mutation tables together describe sequence variation
@@
 genomes separated by many generations.
-Critically, edges can span multiple marginal trees along the
-genome, and identical node IDs across different trees corresponds
+Critically, edges can span multiple trees  along the
+genome (usually referred to as ``marginal'' trees),
+and identical node IDs across different trees corresponds
 to the same ancestral genome.

### 2021-08-23 paper.tex

@@
 drawn from populations at specified times in the past,
-which can inverted to obtain the
+which can be inverted to obtain the
 ``inverse instantaneous coalescence rate'' of \citet{chikhi2018iicr}.

### 2021-09-10 paper.tex
Made affils easier to copy
Drop Ivan from author list for preprint
Update dev section to acknowledge gender skew
Text mainly developed by @cdquinto, @agladstein, @gtsambos and
@petrelharp

Closes #191

@@
 \author[16]{Warren W. Kretzschmar}
-\author[17]{Ivan Krukov}
+% Haven't been able to contact Ivan, so forced to drop for preprint
+% \author[17]{Ivan Krukov}
 \author[2]{Konrad Lohse}
-\author[18]{Michael Matschiner}
-\author[17]{Dominic Nelson}
+\author[17]{Michael Matschiner}
+\author[18]{Dominic Nelson}
 \author[19]{Nathaniel S. Pope}
@@
 % Senior Authors
-\author[16,$\dagger$]{Simon Gravel}
+\author[18,$\dagger$]{Simon Gravel}
 \author[10,$\dagger$]{Andrew D. Kern}
@@
 % Franz
-\affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'',
-Mathematical and Computational Population Genetics, University of T\"ubingen}
+\affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'', Mathematical and Computational Population Genetics, University of T\"ubingen}
 % Gertjan, Konrad
@@
 % Jerome, Yan, Wilder, Joe
-\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
-    University of Oxford}
+\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}
 
 % Bjarki
-\affil[8]{Leibniz Institute for Evolution and Biodiversity
-Science, Museum f\"ur Naturkunde Berlin }
+\affil[8]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin }
 % Castedo, Hugo
@@
 % Jared, Murillo, Peter, Andy
-\affil[10]{Institute of Ecology and Evolution, Department of Biology,
-    University of Oregon}
+\affil[10]{Institute of Ecology and Evolution, Department of Biology, University of Oregon}
 % Jared
@@
 % Gregor
-\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
-    University of Edinburgh}
+\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh}
 % Bing
-\affil[15]{Institute for Genome Sciences, University of Maryland School of
-Medicine, Baltimore, MD}
+\affil[15]{Institute for Genome Sciences, University of Maryland School of Medicine, Baltimore, MD}
 % Winni
 \affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
+% Michael
+\affil[17]{Natural History Museum, University of Oslo}
 % Simon, Dom, Ivan
-\affil[17]{Department of Human Genetics, McGill University}
-% Michael
-\affil[18]{Natural History Museum, University of Oslo}
+\affil[18]{Department of Human Genetics, McGill University}
 % Nathaniel
@@
 % Consuelo
-\affil[20]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
-    Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
+\affil[20]{National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
 % Saunack
@@
 
+A important goal of the msprime development model is to maximise
+accessibility for prospective users and contributors. However, it is important
+to acknowledge that under-representation persists in population genetics,
+especially in methodology and open-source software
+development~\citep{trinkenreich2021women}. The authorship
+of our paper reflects this with a skew towards men and affiliations in the USA
+and Europe. We know the importance of creating and strengthening networks to
+develop and maintain a diverse community of contributors, and we are committed
+to fostering a supportive collaborative environment that helps to address these
+inequalities in our field.
+
 \section*{Discussion}

### 2021-09-10 paper.tex
Comments from review

@@
 
-A important goal of the msprime development model is to maximise
+An important goal of the msprime development model is to maximise
 accessibility for prospective users and contributors. However, it is important
@@
 develop and maintain a diverse community of contributors, and we are committed
-to fostering a supportive collaborative environment that helps to address these
+to fostering a supportive and collaborative environment that helps to address these
 inequalities in our field.

### 2021-09-14 paper.tex
Add Broad affiliation for Daniel Goldstein

--- 2021-08-23 paper.tex
@@
 \usepackage{nameref}
-\usepackage{caption,subcaption}
 \usepackage{tikz}
-\usepackage[right]{lineno}
-\usetikzlibrary{calc,positioning}
 %\usepackage{showkeys}
@@
 \newcommand{\grgcomment}[1]{\textcolor{yellow!60!red}{#1}}
-\newcommand{\adkcomment}[1]{\textcolor{blue}{#1}}
 
@@
 
-\linenumbers
 \title{Efficient ancestry and mutation simulation with msprime 1.0}
 % First authors
+%% FIXME affiliations are all broken
 \author[1,$\star$]{Franz Baumdicker}
 \author[2,$\star$]{Gertjan Bisschop}
-\author[3,$\star$]{Daniel Goldstein}
+\author[3,24,$\star$]{Daniel Goldstein}
 \author[4,$\star$]{Graham Gower}
@@
 \author[9]{Castedo E. Ellerman}
-\author[10,11]{Jared G. Galloway}
-\author[12,13]{Ariella L. Gladstein}
-\author[14]{Gregor Gorjanc}
-\author[15]{Bing Guo}
+\author[10]{Jared G. Galloway}
+\author[11]{Ariella L. Gladstein}
+\author[12]{Gregor Gorjanc}
+\author[13]{Guo, Bing}
 \author[7]{Ben Jeffery}
-\author[16]{Warren W. Kretzschmar}
-\author[17]{Ivan Krukov}
+\author[14]{Warren W. Kretzschmar}
+\author[15]{Ivan Krukov}
 \author[2]{Konrad Lohse}
-\author[18]{Michael Matschiner}
-\author[17]{Dominic Nelson}
-\author[19]{Nathaniel S. Pope}
-\author[20]{Consuelo D. Quinto-Cort\'es}
+\author[16]{Michael Matschiner}
+\author[15]{Dominic Nelson}
+\author[17]{Nathaniel Pope}
+\author[18]{Consuelo D. Quinto-Cort\'es}
 \author[10]{Murillo F. Rodrigues}
-\author[21]{Kumar Saunack}
-\author[22]{Thibaut Sellinger}
-\author[23]{Kevin Thornton}
-\author[9]{Hugo van Kemenade}
-\author[7,24]{Anthony W. Wohns}
+\author[19]{Kumar Suanack}
+\author[20]{Thibaut Sellinger}
+\author[21]{Kevin Thornton}
+\author[22]{Hugo van Kemenade}
+\author[7]{Anthony W. Wohns}
 \author[7]{H. Yan Wong}
@@
 % Senior Authors
-\author[16,$\dagger$]{Simon Gravel}
-\author[10,$\dagger$]{Andrew D. Kern}
-\author[25,$\dagger$]{Jere Koskela}
-\author[10,26,$\dagger$]{Peter L. Ralph}
+\author[15,$\dagger$]{Simon Gravel}
+\author[10,$\dagger$]{Andrew Kern}
+\author[23,$\dagger$]{Jere Koskela}
+\author[10,24,$\dagger$]{Peter L. Ralph}
 
@@
 \affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
-% Aaron
-\affil[5]{Department of Integrative Biology, University of Wisconsin--Madison}
+% Simon, Dom, Ivan
+\affil[5]{FIXME AARON's AFFILIATION}
 % Georgia
@@
 Science, Museum f\"ur Naturkunde Berlin }
-% Castedo, Hugo
-\affil[9]{No affiliation}
+% Castedo
+\affil[9]{FIXME: CASTEDO's AFFILIATION}
 % Jared, Murillo, Peter, Andy
-\affil[10]{Institute of Ecology and Evolution, Department of Biology,
+\affil[10]{Department of Biology and Institute of Ecology and Evolution,
     University of Oregon}
-% Jared
-\affil[11]{Computational Biology Program, Fred Hutchinson Cancer Research
-Center, Seattle, WA 98102, USA}
 % Ariella
-\affil[12]{Department of Genetics, University of North Carolina at Chapel Hill}
-\affil[13]{Embark Veterinary, Inc., Boston}
+\affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
 % Gregor
-\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
+\affil[12]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
     University of Edinburgh}
 % Bing
-\affil[15]{Institute for Genome Sciences, University of Maryland School of
-Medicine, Baltimore, MD}
+\affil[13]{FIXME: BING's AFFILIATION}
 % Winni
-\affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
+\affil[14]{FIXME: WINNI's AFFILIATION}
 % Simon, Dom, Ivan
-\affil[17]{Department of Human Genetics, McGill University}
+\affil[15]{Department of Human Genetics, McGill University}
 % Michael
-\affil[18]{Natural History Museum, University of Oslo}
+\affil[16]{FIXME: MICHAEL's AFFILIATION}
 % Nathaniel
-\affil[19]{Department of Entomology, Pennsylvania State University}
+\affil[17]{FIXME: NATHANIEL's AFFILIATION}
 % Consuelo
-\affil[20]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
+\affil[18]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
     Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
 % Saunack
-\affil[21]{IIT Bombay, India}
+\affil[19]{FIXME: SAUNACK's AFFILIATION}
 % Thibaut
-\affil[22]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
 % Kevin
-\affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
-% Wilder 2
-\affil[24]{Broad Institute of MIT and Harvard}
+\affil[21]{Ecology and Evolutionary Biology, University of California at Irvine}
+% Hugo
+\affil[22]{FIXME HUGO's AFFILIATION}
 % Jere
-\affil[25]{Department of Statistics, University of Warwick}
+\affil[23]{Department of Statistics, University of Warwick}
 % Peter (second affil)
-\affil[26]{Department of Mathematics, University of Oregon}
+\affil[24]{Department of Mathematics, University of Oregon}
 
@@
 \begin{abstract}
-Stochastic simulation is a key tool in population genetics,
-since the models involved are often analytically intractable
-and simulation is usually the only way of obtaining ground-truth data to
-evaluate inferences.
-Because of this necessity, a large number
-of specialised simulation programs have been developed, each
-filling a particular niche, but with largely overlapping functionality
-and a substantial duplication of effort.
-Here, we introduce \msprime\ version 1.0, which
-efficiently implements ancestry and mutation simulations based on
-the succinct tree sequence data structure and \tskit\ library.
-We summarise \msprime's many features, and show that
-its performance is excellent, often many times faster
-and more memory efficient than specialised alternatives.
-These high-performance features have been thoroughly
-tested and validated, and built using a collaborative,
-open source development model,
-which reduces duplication of effort and promotes
+Coalescent simulation is a key tool in population genetics and
+has been integral to theoretical developments since its beginnings.
+Because of the ease with which the basic coalescent model can be simulated,
+a large number of different simulators have been developed. However,
+the coalescent with recombination is far more challenging to simulate,
+and, until recently, it was not possible to simulate efficiently.
+The \msprime\ software has revolutionised
+coalescent simulation, making it possible to simulate millions
+of whole genomes for the first time. We summarise the many features
+of \msprime\ 1.0, which is built around a core model of efficiently
+implementing recombination via the succinct tree sequence data
+structure. Moreover, development of \msprime\ is collaborative
+and open source, which reduces duplication of effort and promotes
 software quality via community engagement.
@@
 
-\textbf{Keywords:} Simulation, Coalescent, Mutations, Ancestral Recombination
-Graphs
+\textbf{Keywords:} Coalescent simulation, Python, ancestral recombination graphs
 
@@
 The coalescent
-process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
-tajima1983evolutionary}
-models the ancestry of a set of sampled genomes,
-providing a mathematical description of the
-genealogical tree that relates the samples to one another.
-It has proved to be a powerful model,
-and is now central to
+process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
+models the ancestry of a set of sampled genomes. Specifically,
+the coalescent provides a mathematical description of the
+genealogical tree that relates sampled genomes to one another.
+The coalescent has proved to be a powerful model,
+and it is now central to
 population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
-The coalescent is an efficient framework for population genetic
-simulation, because it allows us to simulate the genetic ancestry for
-a sample from an idealised population model, without explicitly representing
-the population in memory or stepping through the generations.
-Indeed, \cite{hudson1983testing} independently derived the
-% Anyone disagree here? That's the impression I get from the paper.
-coalescent \emph{in order to} efficiently simulate data,
-and used these simulations to characterise an analytically
-intractable distribution.
-This inherent efficiency, and the great utility of simulations for a wide
-range of purposes, has led to dozens of different tools
-being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
+Among the most impactful contributions of the coalescent since its origins
+~\citep{hudson1983testing} is that it provides a path towards more efficient
+simulation of population genetic samples than is possible using traditional,
+forwards in time simulation.
+As a result of this great utility, dozens of different simulation tools have been
+developed over the decades~\citep{carvajal2008simulation,liu2008survey,
 arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
 peng2015genetic}.
-
 Two technological developments of recent years, however,
-pose major challenges to most existing simulation methods.
-Firstly, fourth-generation sequencing technologies have made
-complete chromosome-level assemblies possible~\citep{miga2020telomere},
+pose major challenges to the majority of these methods.
 % Some citations here? Darwin tree of life maybe?
-and high quality assemblies are now available for many species.
-Thus, modelling genetic variation data as a series of unlinked
-non-recombining loci
-is no longer a reasonable approximation, and
+Firstly, whole genome sequencing---and in particular fourth-generation
+sequencing technologies---have made chromosome-level assemblies
+commonplace. Thus, modelling genetic variation data as a series of unlinked loci,
+a previously reasonable assumption, is no longer defensible and
 we must fully account for recombination.
-However, while a genealogical tree relating $n$ samples in the single-locus coalescent
+While a genealogical tree relating $n$ samples in the single-locus coalescent
 can be simulated in $O(n)$ time~\citep{hudson1990gene},
-the coalescent with recombination is far more complex,
-and programs such as
-Hudson's classical \ms~\citep{hudson2002generating}
-can only simulate short segments under the influence of recombination.
-The second challenge facing simulation methods is that
-sample sizes in genetic studies have grown very quickly
-in recent years, enabled by the precipitous fall in genome sequencing costs.
-Human datasets like the
+the coalescent with recombination, which generates multiple correlated trees, is
+far more challenging.
+% JK will need to revise this sentence, in light of what we'll be saying
+% down further about needing the SMC for very large rho.
+Programs such as Hudson's classical \ms~\citep{hudson2002generating}
+can only simulate short segments under the influence of recombination,
+and, until recently, approximate models of
+recombination~\citep{mcvean2005approximating,staab2015scrm}
+were necessary to simulate whole chromosomes.
+
+The second major challenge to existing coalescent simulators is the
+exponential increase in sample size of genetic studies. Human datasets like the
 UK Biobank~\citep{bycroft2018genome} and
@@
 Classical simulators such as \ms\ and even fast approximate methods
-such as \scrm~\citep{staab2015scrm} simply cannot
-cope with such a large number of samples.
+such as \scrm\ simply cannot cope with such a large number of samples.
 
@@
 has greatly increased the scope of coalescent simulations,
-and it is now straightforward to simulate millions of whole chromosomes
-for a wide range of organisms.
+and it is now straightforward to simulate millions of whole chromosomes.
 The ``succinct tree sequence'' data
@@
 wohns2021unified},
-originally introduced as part of \msprime, makes it possible to store
-such large simulations in a few gigabytes, several orders
+introduced as part of \msprime\, makes it possible to store
+such huge simulations in a few gigabytes, several orders
 of magnitude smaller than commonly used formats.
-The succinct tree sequence has also led to major advances in forwards-time
+The succinct tree sequence has also lead to major advances in forwards-time
 simulation~\citep{kelleher2018efficient,haller2018tree},
@@
 genetic simulation.
-This paper marks the release of \msprime\ 1.0.
-We provide an overview of its extensive features,
+This paper marks the release of \msprime\ 1.0, a state-of-the-art
+coalescent simulator both with respect to efficiency as well as feature-richness.
+Here we provide an overview of the extensive abilities of \msprime\ 1.0,
 demonstrate its performance advantages over alternative software,
-and discuss opportunities for ongoing
-open-source community-based development.
-
-% and finally lay out future plans for
-% further open-source development of what we hope will become an industry
-% standard piece of computational infrastructure for the genetics community.
+and finally lay out future plans for
+further open-source development of what we hope will become an industry
+standard piece of computational infrastructure for the genetics community.
 
 \section*{Results}
-In the following sections we describe the main features of \msprime\ 1.0,
-focusing on the aspects that are either new for this version, or in which
-our approach differs significantly from classical methods. Where appropriate,
-we benchmark \msprime\ against other simulators, but the comparisons are
-illustrative and not intended to be systematic or exhaustive. Please
-see~\cite{kelleher2016efficient} for a performance comparison of
-\msprime\ against simulators such as \ms, \msms, and \scrm.
-
-% jk: I started writing a roadmap here, but it's quite hard to do concisely
-% and you either end up leaving sections out, or just summarising them
-% one-by-one which makes it long and tedious. I think readers can either read
-% through fully if they're interested, or just skim the section titles which
-% should give them a good idea of whether they want to read it or not.
+
+\jkcomment{Give a quick roadmap,
+like, we discuss the underlying tree sequence object,
+then mutations (because it's quite separate from the rest)
+and then a bunch of new features for ancestry simulation. There's no
+point in listing it out section by section.}
+
+% JK: commented out this text as it's misleading and would confuse
+% people. We should adapt it when writing the roadmap when the structure
+% has settled down.
+
+% In this section we describe the main features of \msprime\ 1.0. Methodologically,
+% there exists a logical separation between simulating coalescent genealogies and
+% adding neutral mutations onto these genealogies. Simulating genetic ancestry
+% requires methods that model the underlying reproductive process (e.g. Moran,
+% Wright-Fisher, etc.) as well as accounting for recombination of various sorts
+% (including gene conversion), demography, and selection. These methods are
+% discussed in the first N subsections below. One of the strengths of the
+% approach taken in msprime is that neutral mutations can be retrospectively
+% added to simulated genealogies. Methods to generate mutations, and thereby
+% generate genetic variation, are therefore discussed in a separate and subsequent
+% subsection. Finally, we describe the \msprime\ API, which is designed to enforce
+% the distinction between simulating genealogies and mutations; discuss the
+% interoperability provided by the use of \tskit\, the tree sequence toolkit;
+% and highlight the open source development process which allows for \msprime\
+% to be extended by addition of new features while ensuring robustness and
+% maintainability of the core codebase.
+
+% The data model
+\subsection*{Tree sequences}
+\label{sec-ts}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=6cm]{illustrations/example_tree_sequence}
+\end{center}
+\caption{\label{fig-ts-example} Example tree sequence.
+An example tree sequence describing genealogies and genotypes
+for four samples at ten sites.
+}
+\end{figure}
+
+One of the key reasons for \msprime's substantial performance advantage
+over other simulators~\citep{kelleher2016efficient}
+is its usage of the ``succinct tree sequence''
+ data structure to represent simulation results.
+The succinct tree sequence (usually abbreviated to ``tree sequence'')
+is a data structure for concisely encoding
+genetic ancestry and sequence variation.
+The basic concepts were introduced by \cite{kelleher2016efficient}
+and originally developed as part of \msprime. We subsequently
+extracted the tree sequence code that was independent of simulations
+from \msprime\ into
+a standalone Python and C library, called \tskit~\citep{tskit2021tskit}.
+While a full discussion of tree sequences and the capabilities
+of \tskit\ is beyond the scope of this article, we summarise
+some aspects that are important for simulation.
+
+% Encoding ancestry
+Given a set of sample chromosomes, the topology
+of their genetic ancestry is described by a set of \emph{nodes}
+and \emph{edges} (Fig.~\ref{fig-ts-example}).
+A node in tree sequence terminology corresponds
+to a single chromosome strand, associated with its birth time
+and other optional information. A node may be
+associated with an \emph{individual}, so that, for example, a diploid
+individual is associated with two separate nodes (we omit
+the individual table in Fig.~\ref{fig-ts-example} as it is optional).
+Nodes define haplotypes that exist at a specific time within
+some individual, and edges define genetic inheritance
+between these nodes. An \emph{edge} consists of a
+parent node, a child node and the left and right coordinates
+of the contiguous genomic interval over which
+the child inherited genetic material from the parent.
+Critically, edges can span multiple marginal trees: for
+example, in Fig.~\ref{fig-ts-example} the branch from node
+0 to 4 is present in both marginal trees, and represented
+by a single edge (the first row in the edge table).
+This simple device, of explicitly associating tree nodes
+with specific ancestral chromosomes and recording the
+genomic intervals over which parent/child relationships exist,
+generalises the original
+``coalescence records'' concept~\citep{kelleher2016efficient},
+and is the key to the efficiency of tree
+sequences~\citep{
+kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
+
+The tree sequence encoding of ancestry as a set of nodes and edges
+as just described, is equivalent to a graph in which edges
+are annotated with a set of disjoint genomic intervals.
+This is closely related to the Ancestral Recombination Graph (ARG),
+a term that is often loosely understood
+to mean some representation of genetic
+ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
+In this sense a succinct tree sequence \emph{is} an ARG.
+More technically, an ARG is a particular
+representation of genetic ancestry as a graph based on the
+flow of ancestral material through recombination and common
+ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
+In this sense, tree sequences can be used to \emph{encode}
+ARGs without loss of information (although
+the default minimal tree sequence encoding is
+much more concise; see the \nameref{sec-arg} section).
+It should be noted that, despite standardisation
+efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
+no shared format for interchanging ARG data, and no supporting software
+libraries.
+
+% OK, back to tree sequences: what about sequence variation?
+The final output of most population genetic simulations is some representation
+of sequence variation among the specified samples. For coalescent
+simulations, we usually have three steps: (1) simulate
+the genetic ancestry, and optionally output the resulting marginal trees;
+(2) simulate sequence evolution conditioned on this ancestry by generating
+mutations (see the \nameref{sec-mutations} section); and (3) output the
+resulting sequences by percolating the effects of the mutations through
+the trees. Information about the mutations
+themselves---e.g., where they have occurred on the
+trees---is usually not retained or made available for subsequent analysis.
+In \msprime, however, we skip step (3), instead using \tskit's
+combined data model of ancestry and mutations
+to represent the simulated sequences.
+As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
+fully integrated part of \tskit's tree sequence data model, and genetic
+variation is encoded by recording sites at which mutations
+have occurred, and where each mutation at those sites
+has occurred on the marginal tree.
+Crucially, the genetic sequences themselves are never stored, or indeed directly
+represented in memory (unless explicitly decoded using \tskit\ APIs).
+It may at first seem inconvenient to have only this indirect representation
+of the genome sequences, but it is extremely powerful.
+Firstly, the storage space required for simulations is dramatically
+reduced. For a simulation of $n$ samples with $m$ variant sites,
+we would require $O(nm)$ space to store the sequence data as a variant
+matrix. However, if this simulation was of a recombining genome
+with $t$ trees, then the \tskit\ tree sequence encoding
+requires $O(n + t + m)$ space,
+assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
+For large sample sizes, this difference is profound, making it
+conceivable, for example, to store the genetic ancestry
+and variation data for the entire human population on a
+laptop~\citep{kelleher2019inferring}. As well as the huge
+difference in storage efficiency, it is often far more efficient
+to compute statistics of the sequence data from the
+trees and mutations than it is to work with the sequences themselves.
+For example, computing Tajima's $D$ from simulated data stored in the \tskit\ format
+is several orders of magnitude faster than efficient variant matrix
+libraries for large sample sizes~\citep{ralph2020efficiently}.
+
+The vast genomic datasets produced during the SARS-CoV-2 pandemic
+have highlighted the advantages of storing genetic variation data using the underlying trees.
+\cite{turakhia2021ultrafast} propose the
+Mutation Annotated Tree (MAT)
+format (consisting of a Newick tree
+and associated mutations in a binary format) and
+the \texttt{matUtils} program as an efficient way to store
+and process large viral datasets~\citep{mcbroome2021daily},
+achieving excellent compression and processing performance.
+Similarly, \phastsim~\citep{demaio2021phastsim}
+was developed to simulate sequence evolution on such large SARS-CoV-2
+phylogenies, and also outputs a Newick tree annotated with mutations (not
+in MAT format) to avoid the bottleneck of generating
+and storing the simulated sequences.
+While these methods illustrate the advantages of the general
+approach of storing ancestry and mutations rather than sequences,
+they do not generalise beyond their immediate settings, and
+no software library support is available.
+% \msprime, on the other hand, provides a fast, powerful, and well-tested
+% generic interface for generating and storing mutations on trees.
+
+The software ecosystem built around \tskit\ is stable and mature;
+simulators such as
+\fwdpy~\citep{thornton2014cpp},
+\SLiM~\citep{haller2019slim},
+\stdpopsim~\citep{adrion2020community},
+\texttt{Geonomics}~\citep{de2021geonomics}
+and
+\texttt{GSpace}~\citep{virgoulay2021gspace},
+and inference methods such as
+\texttt{tsinfer}~\citep{kelleher2019inferring},
+\texttt{tsdate}~\citep{,wohns2021unified}
+and \texttt{Relate}~\citep{speidel2019method}
+use either the Python or C
+APIs to support outputting results in tree sequence format.
+Tree sequences are stored in an efficient binary file format, and
+are fully portable across operating systems and processor
+architectures. The \tskit\ library ensures interoperability
+between programs by having strict definitions of how the
+information in each of the tables is interpreted, and
+stringent checks for the internal consistency of the data model.
 
@@
 
-\begin{figure}
-    \includegraphics{illustrations/mutated_tree}
-    \caption{
-        \label{fig-mutated-trees}
-        Visualisation of the separation between ancestry and mutation
-        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
-        is two trees along a 1kb chunk of genome relating three diploid samples.
-        Each diploid individual consists of two genomes (or nodes), indicated
-        by colour.
-        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
-        which adds mutations.
-        Graphics produced using \tskit's \texttt{draw\_svg} method.
-    }
-\end{figure}
-
 A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
@@
 supported and tested, and will be maintained for the foreseeable future.
-The \texttt{msp} program has been extended to include new commands
+We also provide an \ms\ compatible
+command line interface to support existing workflows. The
+\texttt{msp} program has been extended to include new commands
 for simulating ancestry and mutations separately. A particularly useful
@@
 making simulation of complex demographies straightforward.
-We also provide an \ms\ compatible
-command line interface to support existing workflows.
-
-% The data model
-\subsection*{Tree sequences}
-\label{sec-ts}
-
-\begin{figure}
-\begin{center}
-\includegraphics{illustrations/example_tree_sequence}
-\end{center}
-\caption{\label{fig-ts-example} An example tree sequence describing
-genealogies and sequence variation for four samples at ten sites
-on a chromosome of twenty bases long.
-Information is stored in a set of tables
-(the tables shown here include only essential columns, and
-much more information can be associated with the various entities).
-The node table
-stores information about sampled and ancestral genomes. The
-edge table describes how these genomes are related along a
-chromosome, and defines the genealogical tree at each position.
-The site and mutation tables together describe sequence variation
-% JK: also amongst the ancestral genomes, but only in positions where
-% they inherit from the sample. I thought it would be less confusing
-% to leave this out?
-among the samples.
-}
-\end{figure}
-
-One of the key reasons for \msprime's substantial performance advantage
-over other simulators~\citep{kelleher2016efficient}
-is its use of the ``succinct tree sequence''
-data structure to represent simulation results.
-The succinct tree sequence (usually abbreviated to ``tree sequence'')
-was introduced by \cite{kelleher2016efficient}
-to concisely encode genetic ancestry and sequence variation
-and was originally implemented as part of \msprime.
-We subsequently extracted the core tree sequence functionality
-from \msprime\ to create the \tskit\ library,
-which provides a large suite of tools for processing genetic ancestry
-and variation data via APIs
-in the Python and C languages~\citep{tskit2021tskit}.
-The availability of \tskit\ as a liberally licensed (MIT) open source
-toolkit has enabled several other
-projects~\citep[e.g.][]{kelleher2019inferring,haller2019slim,wohns2021unified,
-de2021geonomics} to take advantage
-of the same efficient data structures used in \msprime,
-and we hope that many more will follow.
-While a full discussion of tree sequences and the capabilities
-of \tskit\ is beyond the scope of this article, we summarise
-some aspects that are important for simulation.
-
-% Encoding ancestry
-% JK: we can say this some other way, this just seemed like the
-% most concise way. People may be irritated by being told what
-% the definition of a "genome" is, but I think we need to be explicit.
-Let us define a genome as the complete set of genetic material
-that a child inherits from one parent. Thus, a diploid individual
-has two (monoploid) genomes, one inherited from each parent. Since
-each diploid individual lies at the end of two distinct lineages
-of descent, they will be represented by \emph{two} places (nodes)
-in any genealogical tree. In the tree sequence encoding a
-\emph{node} therefore corresponds to a single genome, which
-is associated with its creation time (and other optional information),
-and recorded in a simple tabular format (Fig.~\ref{fig-ts-example}).
-Genetic inheritance between genomes (nodes) % delete?
-is defined by edges.
-An \emph{edge} consists of a
-parent node, a child node and the left and right coordinates
-% JK: This is potentially confusing. Changed "genomic interval"
-% to "chromosome" interval to avoid talking about "the" genome.
-of the contiguous chromosomal segment over which
-the child genome inherited genetic material from the parent genome.
-Parent and child nodes may correspond to ancestor and descendant
-genomes separated by many generations.
-Critically, edges can span multiple trees  along the
-genome (usually referred to as ``marginal'' trees),
-and identical node IDs across different trees corresponds
-to the same ancestral genome.
-For example, in Fig.~\ref{fig-ts-example} the branch from node
-0 to 4 is present in both marginal trees, and represented
-by a single edge (the first row in the edge table).
-This simple device, of explicitly associating tree nodes
-with specific ancestral genomes and recording the
-contiguous segments over which parent-child relationships exist,
-generalises the original
-``coalescence records'' concept~\citep{kelleher2016efficient},
-and is the key to the efficiency of tree
-sequences~\citep{
-kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
-See the \nameref{sec-arg} section below for a discussion of this
-closely related concept.
-
-% OK, back to tree sequences: what about sequence variation?
-The final output of most population genetic simulations is some representation
-of sequence variation among the specified samples. For coalescent
-simulations, we usually have three steps: (1) simulate
-the genetic ancestry, and optionally output the resulting marginal trees;
-(2) simulate sequence evolution conditioned on this ancestry by generating
-mutations (see the \nameref{sec-mutations} section); and (3) output the
-resulting nucleotide sequences by percolating the effects of the mutations through
-the trees. Information about the mutations
-themselves---e.g., where they have occurred on the
-trees---is usually not retained or made available for subsequent analysis.
-In \msprime, however, we skip step (3), instead using \tskit's
-combined data model of ancestry and mutations
-to represent the simulated sequences.
-As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
-fully integrated part of \tskit's tree sequence data model, and genetic
-variation is encoded by recording sites at which mutations
-have occurred, and where each mutation at those sites
-has occurred on the marginal tree.
-Crucially, the genome sequences themselves are never stored, or indeed directly
-represented in memory (although \tskit\ can output the variant matrix
-in various formats, if required).
-It may at first seem inconvenient to have only this indirect representation
-of the genome sequences, but it is extremely powerful.
-Firstly, the storage space required for simulations is dramatically
-reduced. For a simulation of $n$ samples with $m$ variant sites,
-we would require $O(nm)$ space to store the sequence data as a variant
-matrix. However, if this simulation was of a recombining genome
-with $t$ trees, then the \tskit\ tree sequence encoding
-requires $O(n + t + m)$ space,
-assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
-For large sample sizes, this difference is profound, making it
-conceivable, for example, to store the genetic ancestry
-and variation data for the entire human population on a
-laptop~\citep{kelleher2019inferring}. As well as the huge
-difference in storage efficiency, it is often far more efficient
-to compute statistics of the sequence data from the
-trees and mutations than it is to work with the sequences themselves.
-For example, computing Tajima's $D$ from simulated data stored in the \tskit\ format
-is several orders of magnitude faster than efficient variant matrix
-libraries for large sample sizes~\citep{ralph2020efficiently}.
-
-The vast genomic datasets produced during the SARS-CoV-2 pandemic
-have highlighted the advantages of storing genetic variation data using the underlying trees.
-\cite{turakhia2021ultrafast} propose the
-Mutation Annotated Tree (MAT)
-format (consisting of a Newick tree
-and associated mutations in a binary format) and
-the \texttt{matUtils} program as an efficient way to store
-and process large viral datasets~\citep{mcbroome2021daily},
-achieving excellent compression and processing performance.
-Similarly, \phastsim~\citep{demaio2021phastsim}
-was developed to simulate sequence evolution on such large SARS-CoV-2
-phylogenies, and also outputs a Newick tree annotated with mutations (not
-in MAT format) to avoid the bottleneck of generating
-and storing the simulated sequences.
-While these methods illustrate the advantages of the general
-approach of storing ancestry and mutations rather than sequences,
-they do not generalise beyond their immediate settings, and
-no software library support is available.
-% \msprime, on the other hand, provides a fast, powerful, and well-tested
-% generic interface for generating and storing mutations on trees.
-
-The software ecosystem built around \tskit\ is stable, mature and
-rapidly growing.
-Simulators such as
-\fwdpy~\citep{thornton2014cpp},
-\SLiM~\citep{haller2019slim},
-\stdpopsim~\citep{adrion2020community},
-\texttt{Geonomics}~\citep{de2021geonomics}
-and
-\texttt{GSpace}~\citep{virgoulay2021gspace},
-and inference methods such as
-\texttt{tsinfer}~\citep{kelleher2019inferring},
-\texttt{tsdate}~\citep{,wohns2021unified}
-and \texttt{Relate}~\citep{speidel2019method}
-use either the Python or C
-APIs to support outputting results in tree sequence format.
-Tree sequences are stored in an efficient binary file format, and
-are fully portable across operating systems and processor
-architectures. The \tskit\ library ensures interoperability
-between programs by having strict definitions of how the
-information in each of the tables is interpreted, and
-stringent checks for the internal consistency of the data model.
 
@@
 \label{sec-data-analysis}
-% We could drop this first sentence?
-% The point of running simulations is to generate data, and there
-% are two important ways in which the \tskit\ tree sequences output by
-% \msprime\ are more useful than classical methods:
-% they are more efficient and they contain more information.
-% We will consider these points in turn.
+The point of running simulations is to generate data, and there
+are two important ways in which the \tskit\ tree sequences output by
+\msprime\ are more useful than classical methods:
+they are more efficient and they contain more information.
+We will consider these points in turn.
 
 The standard way of representing simulation data is
-to render the results in a text format, which must subsequently
-be parsed and processed as part of some analysis pipeline. For example,
-\ms\ outputs a set of sequences
+to render the results in a text format. For example,
+\ms~\citep{hudson2002generating} outputs a set of sequences
 and can also optionally output the marginal trees along
@@
 and variants of this approach are used by many simulators.
-Text files have many advantages, but are slow to process at scale.
+Text files have many advantages, but are slow to process
+at scale.
 The ability to efficiently process simulation results is
@@
 results all the more acute.
-By using the stable APIs and efficient data interchange mechanisms
-provided by \tskit,
-the results of an \msprime\ simulation can be immediately processed,
-without format conversion overhead.
-The \tskit\ library has a rich suite of population
-genetic statistics and other utilities, and is in many cases
-orders of magnitude faster than matrix-based methods for
-large sample sizes~\citep{ralph2020efficiently}. Thus, the combination
-of \msprime\ and \tskit\ substantially increases the overall
-efficiency of many simulation analysis pipelines.
 
 Classical text based output formats like \ms\ are inefficient to process,
-but also lack a great deal of important information about the simulated
-process.
+but also lack a great deal of important information.
 The tree-by-tree topology information output by simulators in Newick
@@
 from the output. Because Newick stores branch lengths rather
-than node times, numerical precision issues also arise for
-large trees~\citep{mcgill2013graphml}.
+than node times, numerical precision issues also arise when summing
+over many short branches~\citep{mcgill2013graphml}.
 Numerous forks  of simulators have been created to access information not provided
@@
 output information about migrating segments~\citep{rosenzweig2016powerful},
-ancestral lineages~\citep{chen2013asymptotic},
-% leaving this out because it's not about hidden information
-% an array of population genetic statistics~\citep{ramos2007mlcoalsim}
+information about ancestral lineages~\citep{chen2013asymptotic},
+an array of population genetic statistics~\citep{ramos2007mlcoalsim}
 % Any more?
-and \ms's fork \msHOT~\citep{hellenthal2007mshot}
+and \ms's fork \msHOT~\citep{hellenthal2007mshot},
 has in turn been forked to output information on local
@@
 
-% These problems of inefficient and limited access to the simulated
-% data are comprehensively solved by \msprime\ and \tskit. Complete
-% information about ancestry and mutation in simulations is immediately
-% available to the user in a single integrated data structure
-% (rather than multiple independent files), with no intermediate
-% processing required.
-% Users can subsequently compute precisely the information that they
-% require using the powerful suite of population genetic statistics
-% and other methods available in \tskit.
+These problems of inefficient and limited access to the simulated
+data are comprehensively solved by \msprime\ and \tskit. Complete
+information about ancestry and mutation in simulations is immediately
+available to the user in a single integrated data structure
+(rather than multiple independent files), with no intermediate
+processing required.
+Users can subsequently compute precisely the information that they
+require using the powerful suite of population genetic statistics
+and other methods available in \tskit.
 
@@
 Because coalescent simulations are usually concerned with
-neutral evolution (see the \nameref{sec-selection} section, however)
+neutral evolution (however, see the \nameref{sec-selection} section)
 the problem of generating synthetic genetic variation can be decomposed into
@@
 
+\begin{figure}
+    \includegraphics{illustrations/unmutated_tree}
+    \includegraphics{illustrations/mutated_tree}
+    \caption{
+        \label{fig-mutated-trees}
+        Visualisation of the separation between ancestry and mutation simulation:
+        \texttt{sim\_ancestry} produces ancestry
+        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
+        relating 200 samples),
+        and \texttt{sim\_mutations} adds mutations
+        (\textbf{right;} fourteen nucleotide mutations have been added).
+        Graphics produced with \tskit's \texttt{draw\_svg} method.
+        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
+        \jkcomment{Why 200 samples? Seems like far too many to me, how about 5?}
+        \grgcomment{Add \texttt{sim\_ancestry} and \texttt{sim\_mutations} as subfigure titles?}
+    }
+\end{figure}
+
 Part of the reason for this poor record of software reuse and modularity is the
@@
 data. Although \msprime\ also supports simulating both ancestry and mutations,
-the two aspects are functionally independent within the software; both ancestry and mutation simulators are present in \msprime\
+the two aspects are functionally independent within the software.
+Both ancestry and mutation simulators are present in \msprime\
 for reasons of convenience and history,
 and could be split into separate packages.
-The efficient C and Python interfaces for
+The mature and efficient C and Python interfaces for
 \tskit\ make it straightforward to add further information to
 an existing file, and because of its efficient data interchange mechanisms,
-there is no performance penalty for additional operations
-in a different software package.
+there is no performance penalty for performing additional operations
+in a different Python package.
 Thanks to this interoperability, \msprime's
@@
 As well as providing a new API
-that emphasises the logical split between ancestry and mutation simulation,
+that emphasises the logical split between ancestry and mutation simulation
+(see the \nameref{sec-sim-interface} section),
 we have greatly extended the sophistication of
@@
 can be defined easily and efficiently in
-user code by specifying a transition matrix between
+user code by specifying an arbitrary transition matrix between
 any number of alleles (which can be arbitrary unicode strings).
@@
 \begin{figure}
-    \includegraphics{figures/mutations-perf}
+    \includegraphics[width=\textwidth]{figures/mutations-perf}
 \caption{\label{fig-mutations-perf} Time required to run
@@
 output formats from the mutation generation algorithms.
-Given these timings, it seems unlikely
+Given these timings, it seems quite unlikely
 that generating mutations with \msprime\ would be a bottleneck in any
@@
 Recombination events split the ancestral material for some lineage
-at a breakpoint, creating two independent
+at a given breakpoint, creating two independent
 lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
-this process is much more efficient to simulate
-than the equivalent left-to-right
-approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
-In \msprime\ 1.0, recombination rates can vary along a chromosome, allowing
+this process is much more efficient to simulate than either
+methods based on the Ancestral Recombination Graph
+formulation~\citep{griffiths1991two,griffiths1997ancestral}
+or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
+In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
 us to simulate recombination hotspots and patterns
@@
 with recombination~\citep{chen2009fast,wang2014new,staab2015scrm}.
-The SMC and SMC' models are supported
+The SMC and SMC' models are available
 in \msprime\ 1.0. However, they are currently implemented using a
@@
 \begin{center}
-    \includegraphics{figures/ancestry-perf}
+    \includegraphics[width=\textwidth]{figures/ancestry-perf}
 \end{center}
@@
     for various values of effective population size ($N_e$),
-    chromosome length in Morgans ($L$), and number of samples ($n$).
+    genome length in Morgans ($L$), and number of samples ($n$).
     Run time scales quadratically with the product of $N_e$ and $L$,
@@
     would take about 2 seconds, and (equivalently) that the same number of
-    0.01 Morgan segments with $N_e=200,000$ would take the same time.
+    0.01 Morgan segments with $N_e= 200,000$ would take the same time.
     Since recombination rate in these simulations was $10^{-8}$,
@@
 When this happens, the merged lineage carries ``trapped'' genetic material
-that is not ancestral to any samples, but where recombinations can still
-occur~\citep{wiuf1999recombination}.
+that is not ancestral to any samples, but where recombinations can still occur.
 The SMC approximations work by disallowing common ancestor events
@@
 $\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
-(the sum is a harmonic number and can be approximated by $\log{n}$)
-is not well supported by observed run times (or numbers of events)
-except possibly at very
-large values of $\rho$. It therefore appears that a different dependence
-on $n$ is required to accurately predict simulation time for a
+(the sum is a harmonic number, and approximated by $\log{n}$)
+is not well supported by the data except possibly at very
+large values of $\rho$, and so it seems a different dependence
+on $n$ would be required to accurately predict simulation time for a
 given $\rho$ and $n$.
@@
 the \emph{Drosophila melanogaster} chromosome 2L
-is about 23.5Mb long with an average recombination rate of around $2.4 \times 10^{-8}$,
+is about 23.5Mb long and an average recombination rate of around $2.4 \times 10^{-8}$,
 so $L \approx 0.57$,
@@
 $N_e L \approx 10^6$, so extrapolating the curve in Fig.~\ref{fig-ancestry-perf}B
-predicts that simulation would require around 177 hours for 1000 samples.
+predicts that simulation would require around 177 hours.
 For such large values of $\rho$  we recommend
@@
 have efficient implementations of approximate coalescent with recombination
-models, in these cases we recommend using SMC based methods such as \scrm,
-particularly if sample sizes are small.
+models, we recommend using SMC based methods such as \scrm\,
+particularly for moderate sample sizes.
 In practice, to predict the running time of a given
@@
 predicting the run times of days-long simulations
-using the timing of seconds-long runs is unlikely to be very accurate.
+using the timing of minutes-long runs is unlikely to be very accurate.
 
@@
 and recombination events increase their number.
-However, with long genomes,
+However, with long genomes
 only a small fraction of the common ancestor events
-will involve overlapping segments of ancestry
+will involve overlapping segments of ancestry,
 and lead to coalescence in the marginal trees.
@@
 the increase and decrease in numbers of lineages due to
-common ancestor and recombination events balance out.
-To get an idea of run time we can estimate when this balance occurs.
+common ancestor and recombination events balance out;
+to get an idea of run time we can estimate when this occurs.
 Suppose that the maximum number of lineages is $M$;
@@
 
-What does this tell us about run time for simulating time-varying population sizes?
+What does this tell us about time-varying population sizes?
 The argument above implies that the work is spread out relatively
@@
 Suppose that population size today is $N_1$,
-while $T$ generations ago it was $N_2$.
-Does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
+while longer ago than $T$ generations ago it was $N_2$:
+does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
 The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
@@
 \begin{center}
-\includegraphics{figures/gc-perf}
+\includegraphics[width=\textwidth]{figures/gc-perf}
 \end{center}
@@
 \subsection*{Demography}
-\label{sec-demography}
 % TODO should we put in some citations here for the models? Hardly seems
@@
 migration can be recorded.
-Large demographic models can be simulated efficiently in version \msprime\ 1.0,
-since we only consider populations that contain lineages and
-have non-zero migration rates when generating migration event
-waiting times. This is a considerable improvement
-over version 0.x, which scaled quadratically with the number of populations.
 
 A major change for \msprime\ 1.0 is the introduction of the new Demography API,
-designed to address a design flaw in the \msprime\ 0.x interface which led to
+designed to address a design flaw in the \msprime\ 0.x interface which lead to
 a number of avoidable errors in downstream
@@
 Another improvement over the
-0.x APIs is the introduction of explicit population split and admixture events,
+0.X APIs is the introduction of explicit population split and admixture events,
 and a population state machine that ensures that lineages cannot migrate into
@@
 the \texttt{Demography} class, and we also support importing demographic
-models from Newick species trees and the output of programs
-like *BEAST~\citep{heled2009bayesian}.
-
-The \texttt{DemographyDebugger} provides detailed information about
-demographic models as well as numerical methods to make predictions
-about these models. For example, we can compute
-the coalescence rates for two or more lineages
+models in the form of Newick species tree descriptions, as well as those
+produced by programs such as *BEAST~\citep{heled2009bayesian}.
+
+The \texttt{DemographyDebugger} can output detailed information for debugging
+demographic models, but also provides
+methods to compute a suite of numerical predictions about the location of
+lineages over time, for example the coalescence rates for two or more lineages
 drawn from populations at specified times in the past,
-which can be inverted to obtain the
-``inverse instantaneous coalescence rate'' of \citet{chikhi2018iicr}.
+which can be inverted to obtain the ``inverse instantaneous coalescence rate'',
+or IICR of \citet{chikhi2018iicr}.
 Many popular approaches in population genetics use the distribution of
@@
 populations~\citep{wang2020tracking}.
-These numerical methods provide a valuable
+The \texttt{DemographyDebugger} provides a valuable
 ground-truth when evaluating such inference methods, as illustrated by
@@
 attractive and both analytic results and empirical
-inference~\citep{griffiths1994sampling, birkner2009modified,galtier2000detecting,
+inference~\citep{griffiths1994sampling, galtier2000detecting,
 bunnefeld2015inferring} have been developed under this model, there has
@@
 algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
-by adding events to the \texttt{Demography} class
-(see the \nameref{sec-demography} section)
+by adding events to the Demography object (see the Demography section)
 and can be used in combination with any other demographic modelling
@@
 have also been shown to be relevant for modelling the effects of selection on
-gene genealogies
-\citep{Gillespie909,DS04,desai2013genetic,neher2013genealogies,schweinsberg2017rigorous}.
+gene genealogies \citep{Gillespie909,DS04}.
 
@@
 The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
-also includes a number of extensions to the Beta-coalescent.
+includes a number of extensions to the Beta-coalescent.
 None of these methods work with large genomes, and very little work
@@
 We have added two multiple merger coalescent models in \msprime\ 1.0, the
-Beta-coalescent~\citep{schweinsberg03} and
-``Dirac''-coalescent~\citep{BBE13}, allowing us to efficiently simulate
-such models with recombination for the first time.
+Beta-coalescent and ``Dirac''-coalescent, allowing us to simulate
+such models with recombination efficiently for the first time.
 These simulation models have been extensively validated against
@@
  as well as more general properties of coalescent processes.
-See the Appendix for more details and model derivations.
+Please see the Appendix for more details and model derivations.
 
@@
 Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
-to represent the stochastic process of the coalescent with recombination as a graph.
-This formulation is complementary to Hudson's earlier
+as a representation of the coalescent with recombination in terms of a graph.
+This formulation was complementary to Hudson's earlier
 work~\citep{hudson1983properties}, and substantially increased our theoretical
@@
 vertices represent common ancestor or recombination events, and edges represent
-lineages. There is the ``big'' ARG, in which we track lineages arising out of
-recombinations regardless of whether they carry ancestral
-material~\citep{ethier1990two},
-% and
-% consequently wait an exponentially large time until all sampled lineages
-% coalesce into a single ancestor~\citep{ethier1990two},
-and the ``little'' ARG in which we only track genetic ancestors.
-Over time, usage of the term
-has shifted away from its original definition as a stochastic process,
-to being interpreted as a representation of a particular genetic ancestry
-as a graph, without necessarily following the specific details of the Griffiths
-formulation~\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
-Under the latter interpretation,
-the tree sequence encoding of genetic ancestry (described above)
-clearly \emph{is} an ARG: the nodes and edges
-define a graph in which edges are annotated with the
-set of disjoint genomic intervals
-through which ancestry flows.
+lineages through which ancestral material flows.
+We may consider the big ARG, in which we track lineages arising out of
+recombinations regardless of whether they carry ancestral material and
+consequently wait an exponentially large time until all sampled lineages
+coalesce into a single ancestor~\citep{ethier1990two}, or the little ARG in
+which we only track genetic ancestors, and wait until all marginal trees
+have fully coalesced (like Hudson's algorithm).
 
 \begin{figure}
-\begin{subfigure}[t]{0.33\textwidth}
-\centering
-\caption*{(A)}
-    \begin{tikzpicture}[scale=0.45,nodestyle/.style={font=\scriptsize}]
+
+\begin{center}
+    \begin{tikzpicture}[scale=0.45]
 
@@
 
-    \node[nodestyle] [below] at (-2,0) {0};
-    \node[nodestyle] [below] at (1,0) {1};
-    \node[nodestyle] [below] at (4,0) {2};
-    \node[nodestyle] [above] at (1,1) {3};
-    \node[nodestyle] [below right] at (1,1) {$x = 0.3$};
-    \node[nodestyle] [above right] at (3,2) {4};
-    \node[nodestyle] [above left] at (-1,3) {5};
-    \node[nodestyle] [above] at (1,5) {6};
+    \node [below] at (-2,0) {0};
+    \node [below] at (1,0) {1};
+    \node [below] at (4,0) {2};
+    \node [above] at (1,1) {3};
+    \node [below right] at (1,1) {$x = 0.3$};
+    \node [above right] at (3,2) {4};
+    \node [above left] at (-1,3) {5};
+    \node [above] at (1,5) {6};
 
     \end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.33\textwidth}
-\centering
-\caption*{(B)}
-\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
-
-\tikzset{greynode/.style={circle,fill,inner sep=0.7},
-nodelabel/.style={font=\scriptsize}}
-
-% Left sample nodes
-\node (s0) [greynode] {};
-\node (s1) [right=of s0,greynode] {};
-\node (s2) [right=of s1,greynode] {};
-\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-
-% Right sample nodes
-\node (rightTree) at (4, 0) {};
-\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
-\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
-\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
-\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-
-% Non-sample nodes
-\node [greynode] (s3) at ($(s1) + (0,1)$) {};
-\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
-\node [greynode] (s4) at ($(s2) + (0,2)$) {};
-\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
-\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
-\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
-\node [greynode] (s6) at ($0.25*(s0) + 0.25*(s1) + 0.5*(s2) + (0,5)$) {};
-\node [greynode] (s6r) at ($0.5*(s0r) + 0.25*(s1r) + 0.25*(s2r) + (0,5)$) {};
-\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
-\foreach \u/\lab in {s4/4, s4r/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/5, s5r/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s6/6, s6r/6} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\foreach \child/\parent in {
-	s5/s0, s5/s1, s6/s5, s6/s2,
-	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
-}
-	\draw (\child) -| (\parent);
-
-% Axes
-\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
-\node (topAx) at (0,-1.0) {};
-\node (topLeft) at ($(s0) + (0,-1.0)$) {};
-\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
-\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
-\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
-\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
-\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
-
-% Interval endpoints
-\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, 5.5);
-
-% Axis titles
-\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.33\textwidth}
-\centering
-\caption*{(C)}
-\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
-
-\tikzset{greynode/.style={circle,fill,inner sep=0.7},
-nodelabel/.style={font=\scriptsize}}
-
-% Left sample nodes
-\node (s0) [greynode] {};
-\node (s1) [right=of s0,greynode] {};
-\node (s2) [right=of s1,greynode] {};
-\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-
-% Right sample nodes
-\node (rightTree) at (4, 0) {};
-\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
-\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
-\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
-\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-
-% Non-sample nodes
-%\node [greynode] (s3) at ($(s1) + (0,1)$) {};
-%\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
-%\node [greynode] (s4) at ($(s2) + (0,2)$) {};
-\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
-\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
-%\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
-\node [greynode] (s6) at ($0.25*(s0) + 0.25*(s1) + 0.5*(s2) + (0,5)$) {};
-\node [greynode] (s6r) at ($0.5*(s0r) + 0.25*(s1r) + 0.25*(s2r) + (0,5)$) {};
-%\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
-\foreach \u/\lab in {s4r/3} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/4} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s6/5, s6r/5} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\foreach \child/\parent in {
-	s5/s0, s5/s1, s6/s5, s6/s2,
-	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
-}
-	\draw (\child) -| (\parent);
-
-% Axes
-\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
-\node (topAx) at (0,-1.0) {};
-\node (topLeft) at ($(s0) + (0,-1.0)$) {};
-\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
-\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
-\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
-\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
-\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
-
-% Interval endpoints
-\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, 5.5);
-
-% Axis titles
-\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
-\end{tikzpicture}
-\end{subfigure}
+% \begin{center}
+% $\equiv$
+% \end{center}
+% \end{column}
+% \begin{column}{5cm}
+\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
+\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
+\includegraphics[width=3.8cm]{figures/arg.pdf}
+
+\end{center}
 \caption{\label{fig-arg} (A) A simple ARG in which a recombination
@@
 sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation.
-Note that the internal nodes have been renumbered in the simplified
-representation, so that, e.g., node 5 in (C) corresponds to node 6 in
-(A) and (B).}
+topology ``simplified'' down to its minimal tree sequence representation;
+(D) the fraction of nodes in a tree sequence that are concerned with
+extra ARG information in a tree sequence returned using \msprime's
+\texttt{record\_full\_arg} mode}
 \end{figure}
+
+However, this original definition of the ARG in terms of a stochastic process
+is rarely what is meant when the term is currently used. The term has come to
+essentially mean any graph-like representation of genetic ancestry in the presence
+of recombination; for example,
+\cite{mathieson2020ancestry} informally define an ARG as the subset of an
+individual's pedigree from which they have inherited ancestral material.
+Others state that an ARG is equivalent to the sequence of marginal
+% Also SARGE paper
+trees along the genome~\citep{yang2016rapidly,mirzaei2016rent}
+
+Such
+imprecision is unfortunate, since there are many ways to encode genetic
+ancestry as a graph, and referring to the original Griffiths papers does little to
+clarify. As we see below, there are situations in which the precise information
+stored in the graph is important, and conflating a high-level
+discussion of genetic ancestry with this mathematical object is not
+helpful. The point of technical terms is to provide clarity and
+precision on what is being discussed, and it is now essentially
+impossible to know what is meant by the term ``ARG'';
+we therefore use the term ``genetic ancestry'' in high-level
+discussions.
 
@@
 As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
-works by dynamically traversing a little ARG.
+works by dynamically generating and traversing a little ARG.
 The graph is not explicitly represented in memory, but is partially
 present through the extant lineages and the ancestral material they carry
-over time. We do not output the graph directly, but
+over time. Since the ARG is a large and unwieldy object
+(Fig.~\ref{fig-arg}) we do not use it as output, but
 rather store the information required to recover the genealogical
-history as nodes and edges in a tree sequence.
-This is far more efficient than outputting the simulated ARG in its entirety.
-For a given scaled recombination rate $\rho$
-(setting aside the dependency on the sample size $n$)
-we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
-in an ARG is $O(\rho^2)$,
-whereas the size of the tree sequence encoding is
-$O(\rho)$~\citep{kelleher2016efficient}.
-This difference between a
-quadratic and a linear dependency on $\rho$ is profound, and shows why
-large simulations cannot output an ARG in practice.
+history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
+or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
 
 Although by default \msprime\ outputs tree sequences that contain full information
-about the genealogical trees, their correlation structure along the chromosome,
-and the ancestral genomes on which coalescences occurred, some information
+about the genealogical trees, their correlation structure along the genome,
+and the ancestral haplotypes on which coalescences occurred, some information
 is lost in this mapping down from ARG space to the minimal tree sequence
 form. In particular, we lose
-information about ancestral genomes that were common ancestors but
+information about ancestral haplotypes that were common ancestors but
 in which no coalescences occurred, and also information about the precise time
-and chromosomal location of recombination events. In most cases, such
+and genomic location of recombination events. In most cases, such
 information is of little relevance as it is in principle unknowable,
 but there are occasions such as visualisation or computing likelihoods (see
-below) in which it is useful. We therefore provide
+below) in which is useful. We therefore provide
 the \texttt{record\_full\_arg} option in \msprime\
@@
 We provide functions to compute the likelihood of ARG realisations
+(encoded using the representation of the \texttt{record\_full\_arg} option)
 and mutational patterns under
 the standard coalescent and infinite sites mutation model.
-See the Appendix for details on these likelihood calculations.
-
-\subsection*{Selective sweeps}
+See the Appendix for details.
+
+\subsection*{Selection}
 \label{sec-selection}
@@
 in which lineages may transit between favoured and unfavoured backgrounds through
-recombination. This approach allows for many selective sweep scenarios to be simulated
+recombination. This allows for many selective sweep scenarios to be simulated
 efficiently, including recurrent, partial, and soft selective sweeps.
-However this efficiency comes at the cost of flexibility in comparison to forwards in
+However in general this
+efficiency comes at the cost of flexibility in comparison to forwards in
 time simulation.
@@
 \begin{center}
-\includegraphics{figures/sweeps-perf}
+\includegraphics[width=\textwidth]{figures/sweeps-perf}
 \end{center}
@@
 replicates for 10 samples in a model
-with a single selective sweep in its history where the beneficial
-allele had a scaled selection coefficient of $2Ns=1000$,
-a per-base recombination rate of $10^{-9}$ and sequence
-length varying from 1kb--200kb.}
+with \jkcomment{[ADK: fill in the details here please]} and a
+per-base recombination rate of $10^{-9}$.}
 \end{figure}
 
-Selective sweeps are implemented in the coalescent as a two
-step-process: first generating an allele frequency trajectory,
-and then simulating a structured coalescent process conditioned
-on that trajectory.
-Following \discoal, we generate sweep trajectories in \msprime\ using a
-jump process approximation to the conditional
-diffusion of an allele bound for
-fixation~\citep{coop2004ancestral}.
-The jump process moves back in time following the beneficial
-allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
-the origination of the allele at $p=1/(2N)$, tracking time in small
-increments $\delta t$. Then, given the frequency $p$ at time $t$,
-the frequency $p^\prime$ at time $t + \delta t$ is given by
-\begin{equation*}
-p^\prime =
-\begin{cases}
-    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
-    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
-\end{cases}
-\end{equation*}
-where
-\begin{eqnarray*}
-	\mu(p)=\frac{\alpha p (1 - p)}{\tanh(\alpha(1-p))}.
-\end{eqnarray*}
-Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
-can be generalised straightforwardly to include arbitrary dominance.
-We can also define trajectories to model neutral alleles and soft
-selective sweeps, which we plan as future additions to \msprime.
-
-Then, given a randomly generated allele frequency trajectory under
-the above model, the simulation of a sweep works by assigning lineages to
-two different structured coalescent ``labels'', based on
-whether they carry the beneficial allele.
-The allele frequency trajectory determines the relative sizes
-of the ``populations'' in these labels over time, and therefore
-the rates at which various events occur.
-Common ancestor events can then only merge lineages from \emph{within}
-a label, but lineages can transfer from one label to the other
-(i.e., from the advantageous to disadvantageous backgrounds,
-and vice versa) as a result of recombination events.
-Once we have reached the end of the simulated trajectory the sweep
-is complete, and we remove the structured coalescent labels.
-Simulation may then resume under any other ancestry model.
-
-Fig.~\ref{fig-selection-perf} compares the performance of \msprime\
-and \discoal\ under a simple sweep model, and shows
-that \msprime\ has far better
-CPU time and memory performance.
-Since our implementation uses the abstract label system mentioned above,
-adding support for similar situations,
-such as inversions~\citep{peischl2013sequential}, should be straightforward.
+Following the approach taken in \discoal, we
+have implemented selective sweeps in \msprime\ by generating
+sweep trajectories from a jump process approximation to the conditional diffusion
+of an allele bound for
+fixation~\citep{coop2004ancestral}---see the appendix for more details.
+This implementation facilitates future changes to generalise sweep models to
+changing population sizes as well as selection from standing variation,
+both of which are features that are planned for the near future.
+Moreover because this implements the basic infrastructure for the structured
+coalescent, adding support for similar situations,
+such as inversions~\citep{peischl2013sequential}, is straightforward.
+Fig.~\ref{fig-selection-perf} shows that \msprime\ has far better
+CPU time and memory performance than \discoal. We have previously shown
+that \msprime\ is much more efficient than \msms\ and \cosiTwo\
+when simulating without selection~\citep{kelleher2016efficient}.
+% JK: do we need an extra proviso like this?
+% and this is unlikely to change unless sweeps dominate the ancestral process.
 
@@
 model does break down is the combination of large sample size
-and long recombining genomes, where the large
+and with long recombining genomes, where the large
 number of recombination events in the recent past results in
@@
 This pathological behaviour results in
-identity-by-descent,
-long-range linkage disequilibrium and ancestry patterns deviating from
+Identity-by-descent,
+long-range linkage disequilibrium (LD), and ancestry patterns deviating from
 Wright-Fisher expectations, and the bias grows with larger sample
@@
 \begin{center}
-\includegraphics{figures/dtwf-perf}
+\includegraphics[width=\textwidth]{figures/dtwf-perf}
 \end{center}
@@
 the standard coalescent with recombination model (``DTWF + Hudson'').
-Memory usage for \msprime's DTWF and hybrid simulations are very similar.
 We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
@@
 which usually happens in the quite recent
-past~\citep{bhaskar2014distortion}.
+past~\cite{bhaskar2014distortion}.
 We therefore support changing the simulation model during a simulation
@@
 genetic ancestry. Given a tree sequence that
-is complete up until time $t$ ago as input
+is complete up until time $t$ as input
 (where marginal trees may or may not have fully coalesced),
@@
 retained---at scale---is unprecedented.
-There may be an opportunity for
-other forward genetic simulators~\citep[e.g.][]{gaynor2021alphasimr}
-to leverage the tree sequence data format and associated tools.
 
@@
 (where a contributor proposes a code change), using standard Continuous
-Integration (CI) methodology. Other CI services
-check for common errors, code formatting issues, and produce reports on
-the level of test coverage for the proposed change.
+Integration (CI) methodology. Other CI services check code formatting
+and for common errors, as well as producing reports on the level of
+test coverage, e.g., pinpointing areas where additional unit tests may
+be required.
 
@@
 We also view contributions to documentation as equally important to the project
-as writing code or designing methods: what use would it be
+as writing code or designing methods---what use would it be
 to write reliable, stable software if no-one used it?
@@
 % We can simulate so much more now
-The 1.0 release of \msprime\ marks a major increase in the
-breadth of available features and the
-potential biological realism of simulations.
-These abilities will allow researchers to perform more robust power analyses,
-more reliably test new methods,
-carry out more reliable inferences,
-and more thoroughly explore the properties of theoretical models.
-Despite this complexity and generality,
-\msprime's performance is state-of-the-art
-and all features are extensively tested and statistically validated.
+
+This release makes \msprime\ able to produce simulations
+with a great deal more biological realism, without a loss in speed and efficiency.
+This ability will allow researchers to perform more robust
+power analyses, more reliably test new methods,
+and carry out more reliable inferences,
+without worrying that gene conversion or back mutation might be sabotaging
+their applications to real data.
 These advances have only been possible thanks to a distributed,
@@
 and the work of many people.
-% We hope the tool continues to be useful to many.
+We hope the tool continues to be useful to many.
 
@@
 significant software quality issues at all stages. This is unsurprising, since
-the majority of simulation software development is performed by students,
-often without formal training in software development.
-The result resembles Haldane's sieve for new mutations:
-many new pieces of software stay permanently on a dusty shelf of supplementary materials,
-while some of those that prove particularly useful when new (like dominant alleles)
-are quickly adopted.
-Although this has produced many good tools
+the majority of simulation software development is performed by students, who
+are themselves learning to code, often without formal training in software
+development.
+Although this ``Haldane's sieve'' of software has produced many good tools
 and enabled decades of research,
@@
 to a high quality if we are to trust its results. There is a growing
-realisation across the sciences
-% More citations?
-\citep[e.g.][]{siepel2019challenges,harris2020array,gardner2021sustained}
+realisation across the sciences \citet[OTHER CITATIONS]{siepel2019challenges}
+% Maybe cite: https://www.biorxiv.org/content/10.1101/092205v3
 that investing in
@@
 established development practices and infrastructure.
+% I'm going out on a limb here - do folks agree?
+% PLR- personally I have learned SO MUCH, and I'm passing it on to others
 Software development in a welcoming community,
 with mentorship by experienced developers,
-is a useful experience for many users.
+is a hugely useful experience for many users,
 The skills that contributors learn
@@
 
-The succinct tree sequence data structure developed
-for \msprime\ provides a view of not only
-genetic variation, but also the genetic ancestry that produced that variation.
+The tree sequence developed for \msprime\ allows users a view of not only
+genetic variation but also the genetic ancestry that produced that variation.
 Recent breakthroughs in methods to infer genetic ancestry in recombining
-organisms~\citep{rasmussen2014genome,kelleher2019inferring,
+organisms~\citep{kelleher2019inferring,
 speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
@@
 \section*{Acknowledgements}
-We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
-on the manuscript.
-% alphabetised by author. Names spelled out if there's an initial collision
+JK is supported by the Robertson Foundation.
+Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
 ADK was supported by NIH awards R01GM117241 and R01HG010774.
-BE was supported by  DFG grant  273887127 through Priority Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
-FB is funded by the Deutsche Forschungsgemeinschaft EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
+Franz Baumdicker is funded by the Deutsche Forschungsgemeinschaft
+EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
 Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
-Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
-Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
-Jerome Kelleher is supported by the Robertson Foundation.
-SG acknowledges funding from the Canada Research Chairs Program, from the Canadian Institutes of Health Research PJT 173300, and from the Canadian Foundation for Innovation.
+BE was supported by  DFG grant  273887127 through Priority
+Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research
+Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
 
@@
 \appendix
-\section*{Appendix}
-
-\subsection*{Mutation generation}
+
+\section*{Mutation generation}
 \label{app-mutation-algorithm}
@@
 If an edge spans a region of the genome with more than one mutation rate,
-this is done separately for each sub-region on which the mutation rate is constant.
-Since each edge is processed independently, the algorithm
-scales linearly with the number of edges in the tree sequence.
+this is done separately for each distinct region.
+The algorithm scales with the number of edges in the tree sequence.
+Since the number of edges is proportional to the number of marginal trees,
+an algorithm that applies mutations separately to each marginal tree
+would take at least as many steps,
+and would require additional bookkeeping necessary to place mutations uniformly on a tree.
 
@@
 Then, each mutation on the tree is considered in turn,
-and a derived allele is randomly chosen based on the parental allele
-(which may be the ancestral allele or the derived allele of a previous mutation).
+in order so that older mutations are considered first,
+and a derived allele is chosen based on the parental allele
+(which may be the ancestral allele or the derived allele of a previous mutation)
+and some randomness.
 Finally, information about the mutations are recorded in the site and mutation tables
@@
 
-The distribution of ancestral alleles
-is used to choose the allele present at the root of the tree at each mutated site,
-i.e., the \texttt{root\_distribution}.
-Mutation models with a finite possible set of alleles have a natural choice
-for this distribution---the \emph{stationary distribution} of the mutation process.
-(All mutation models are Markovian,
-so this may be found as the top left eigenvector of the mutation matrix.)
-This is the default in most models,
-except, e.g., the \texttt{BinaryMutationModel}, whose alleles are 0 and 1
-and always labels the ancestral allele ``0''.
-However, mutational processes are not in general stationary,
-so we often allow different root distribution to be specified.
-
-Since the general algorithm above applies mutations at a single rate
+Since the algorithm above applies mutations at a single rate
 independent of ancestral state,
@@
 i.e., mutations in which the derived allele is equal to the parental allele.
-To illustrate this, consider a mutation model
-in which $A$ or $T$ mutates to a randomly chosen different nucleotide at rate $\alpha$
-and $C$ or $G$ mutates at rate $\beta$, with $\beta < \alpha$.
+To illustrate this, consider a mutation model on the binary alphabet
+$\{X, Y\}$ in which $X$ mutates to $Y$ at rate $\alpha$
+and $Y$ mutates to $X$ at rate $\beta$, with $\beta < \alpha$.
 To implement this, first place mutations at the largest total rate, which is $\alpha$.
-Then, at each site, choose an ancestral allele from the root distribution,
+Then, at each site, choose an ancestral allele from the ``root distribution''
+(a natural choice would be to choose from the stationary distribution,
+$X$ with probability $\beta / (\alpha + \beta)$),
 and for each mutation, choose a derived allele as follows:
-if the parental allele is $A$ or $T$, then choose a random derived allele
-different to the parental allele;
-if the parental allele is $C$ or $G$,
-then choose the derived allele to be equal to the parent allele
-with probability $\beta/(\alpha + \beta)$,
-and randomly choose a different nucleotide otherwise.
+if the parental allele is $X$, then choose $Y$;
+if the parental allele is $Y$, then choose $X$ with probability $\beta/(\alpha + \beta)$,
+and choose $Y$ otherwise.
 This produces the correct distribution by Poisson thinning:
@@
 if not all equilibrium frequencies are the same.
-The presence of silent mutations may at first be surprising
-but there is a good reason to leave them in:
+Although the presence of silent mutations may at first be surprising
+(although they don't affect calculations of statistics in any way),
+there is a good reason to leave them in:
 to allow layering of different mutation models.
@@
 
-Silent mutations are fully supported by \tskit, which correctly accounts for
-their presence when computing statistics and performing other operations.
-For example, silent mutations have no effect on calculations of nucleotide
-site diversity.
-
-\subsection*{Likelihood calculations}
+\section*{Likelihood calculations}
 \label{app-likelihoods}
@@
 \texttt{record\_full\_arg} option as their arguments.
-
-The \texttt{msprime.log\_arg\_likelihood(ts, r, N)} function
-returns the natural logarithm of
+\begin{itemize}
+\item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
 the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
-per-generation recombination probability \texttt{r} and population size \texttt{N} \citep[e.g.][equation (1)]{kuhner2000maximum}.
+per-generation recombination probability \texttt{r} and population size \texttt{N}.
 Specifically, the function returns the logarithm of
@@
 $q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
-of time intervals which end in a recombination, and $g_i$ is the corresponding \emph{gap}:
-the length of contiguous non-ancestral material around the link at which
-the recombination in question took place. The gap indicates the number of links
-(or length of genome in a continuous model) at which a recombination would result
-in exactly the observed pattern of ancestral material in the ARG.
+of time intervals which end in a recombination, and $g_i$ is the corresponding
+\emph{gap}:  the length of material in which the recombination at the end of the
+$i$th period could have taken place without affecting the resulting genealogy.
+A recombination in ancestral material will always have $g_i = 1$, while a recombination
+in trapped non-ancestral material can result in other values of $g_i$.
 For a continuous model of the genome and a recombination in ancestral material,
 we set $g_i = 1$ and interpret the result as a density.
-
-The \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} function returns the
+\item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
 natural logarithm of the probability of the mutations recorded in the tree sequence
@@
 edges.
-% \end{itemize}
+\end{itemize}
 
 \label{app-multiple-mergers}
-\subsection*{Multiple merger coalescent model}
+\section*{Multiple merger coalescent model}
 
@@
 \end{equation}
-where $\one{A} := 1$ if $A$ holds, and zero otherwise,
-$a \geq 0$ is a constant, and $\Lambda$ is a finite measure on the
-unit interval without an atom at zero \citep{DK99,P99,S99}.
+where $a \geq 0$ is a constant and $\Lambda$ is a finite measure on the unit interval without an atom at zero \citep{DK99,P99,S99}.
 There is also a larger class of simultaneous multiple merger coalescents involving
@@
 
+\label{app-sweeps}
+\section*{Selective sweep trajectories}
+We have implemented the coalescent with selective sweeps as a structured
+coalescent process (as in~\citep{braverman1995hitchhiking}). During the
+``sweep phase'' of the simulation the rate of coalescence and recombination
+is tracked separately in lineages linked to the beneficial allele or not,
+with those rates depending on the frequency of the beneficial allele in
+the population at a given time. The allele frequency trajectory of the
+sweeping allele is modelled using a jump process that approximates
+the conditional diffusion of the allele, conditioning on its eventual
+fixation~\citep{coop2004ancestral}.
+The jump process moves back in time following the beneficial
+allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
+the origination of the allele at $p=1/2N$, and we track time in small
+increments $\delta t$. This works such that the allele frequency at
+the next time step back in time, call it $p^\prime$ is
+
+\begin{eqnarray*}
+	p'=\left\{ \begin{array}{ll}
+	    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
+	    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \quad \mbox{with probability $1/2$}\\
+	\end{array} \right.
+\end{eqnarray*}
+where $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$). Further
+$\mu(p)$ is
+\begin{eqnarray*}
+	\mu(p)=\frac{\alpha p (1 - p)}{tanh(\alpha(1-p))}
+\end{eqnarray*}
+
+While our current implementation only deals with genic selection,
+we plan in the future to add sweeps with arbitrary dominance. Further planned
+additions include trajectories of purely neutral alleles, as well as trajectories
+that model soft selective sweeps.
+
 \end{document}

### 2021-09-19 paper.tex
Rework paragraph addressing inequalities.

--- 2021-09-10 paper.tex
@@
 
-An important goal of the msprime development model is to maximise
-accessibility for prospective users and contributors. However, it is important
-to acknowledge that under-representation persists in population genetics,
-especially in methodology and open-source software
-development~\citep{trinkenreich2021women}. The authorship
-of our paper reflects this with a skew towards men and affiliations in the USA
+An important goal of \msprime's development model is to maximise
+accessibility for prospective users and contributors, and
+to encourage diversity in our community.
+Gender inequality is a major problem across
+the sciences~\citep{shannon2019gender}
+and in open source software development~\citep{trinkenreich2021women}.
+Within our field, the contribution of women to early computational methods
+in population genetics was marginalised~\citep{dung2019illuminating},
+and women continue to be under-represented
+in computational biology~\citep{bonham2017women}.
+The authorship of our paper reflects these trends,
+with a skew towards men and affiliations in the USA
 and Europe. We know the importance of creating and strengthening networks to

### 2021-09-19 cover_letter/cover_letter.tex
Cover letter.

\documentclass{letter}

\signature{Jerome Kelleher}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{GENETICS}

\opening{Dear Editors,}

I am writing on behalf of my coauthors to submit our manuscript entitled
\emph{Efficient ancestry and mutation simulation with msprime 1.0}
for consideration for publication in GENETICS
as a ``Methods, Technology, and Resources'' full length research article.

Since its introduction in 2016, \texttt{msprime} has grown in popularity
and is now challenging \texttt{ms} as the widely used coalescent simulator
(according to Google Scholar, \texttt{ms} has had 100 citations so far
in 2021 and \texttt{msprime} 107). This manuscript marks the important
1.0 release milestone and represents the work
of a large community. We illustrate the large range of features
that \texttt{msprime} has accumulated since the 2016 publication,
and show that performance is excellent,
often many times faster and more memory efficient than specialised
alternative simulators.

We believe that this manuscript is a summary of the state-of-the-art
in coalescent simulation, and it will be of substantial
interest to the readers of GENETICS. We hope that you agree.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2021-09-20 paper.tex
Update paper.tex
Co-authored-by: Peter Ralph <petrel.harp@gmail.com>

@@
 to encourage diversity in our community.
-Gender inequality is a major problem across
+Gender and racial inequality caused by discrimination and marginalization is a major problem across
 the sciences~\citep{shannon2019gender}

### 2021-09-20 paper.tex
Cite paper

--- 2021-09-19 paper.tex
@@
 to encourage diversity in our community.
-Gender inequality is a major problem across
-the sciences~\citep{shannon2019gender}
+Gender and racial inequality caused by discrimination and marginalisation
+is a major problem across
+the sciences~\citep{wellenreuther2016women,shannon2019gender}
 and in open source software development~\citep{trinkenreich2021women}.

### 2021-09-20 paper.tex
typos

--- 2021-09-19 paper.tex
@@
 point is the average over $5$ replicate simulations.
-We show observations for \ARGON, \msprime's DTWF implementation (``DWTF'')
+We show observations for \ARGON, \msprime's DTWF implementation (``DTWF'')
 and a hybrid simulation of 100 generations of the DTWF followed by
@@
 
-An important goal of \msprime's development model is to maximise
-accessibility for prospective users and contributors, and
-to encourage diversity in our community.
-Gender inequality is a major problem across
-the sciences~\citep{shannon2019gender}
-and in open source software development~\citep{trinkenreich2021women}.
-Within our field, the contribution of women to early computational methods
-in population genetics was marginalised~\citep{dung2019illuminating},
-and women continue to be under-represented
-in computational biology~\citep{bonham2017women}.
-The authorship of our paper reflects these trends,
-with a skew towards men and affiliations in the USA
-and Europe. We know the importance of creating and strengthening networks to
-develop and maintain a diverse community of contributors, and we are committed
-to fostering a supportive and collaborative environment that helps to address these
-inequalities in our field.
-
 \section*{Discussion}
@@
 However, mutational processes are not in general stationary,
-so we often allow different root distribution to be specified.
+so we often allow a different root distribution to be specified.
 

### 2021-09-20 paper.tex

--- 2021-09-20 paper.tex
@@
 \author[2,$\star$]{Gertjan Bisschop}
-\author[3,$\star$]{Daniel Goldstein}
+\author[3,24,$\star$]{Daniel Goldstein}
 \author[4,$\star$]{Graham Gower}

### 2021-09-21 cover_letter/cover_letter.tex

@@
 Since its introduction in 2016, \texttt{msprime} has grown in popularity
-and is now challenging \texttt{ms} as the widely used coalescent simulator
+and is now challenging \texttt{ms} as the most widely used coalescent simulator
 (according to Google Scholar, \texttt{ms} has had 100 citations so far
@@
 1.0 release milestone and represents the work
-of a large community. We illustrate the large range of features
+of a large community--the list of 32 authors represents substantial efforts from a large consortium. We illustrate the large range of features
 that \texttt{msprime} has accumulated since the 2016 publication,
@@
 We believe that this manuscript is a summary of the state-of-the-art
-in coalescent simulation, and it will be of substantial
+in coalescent simulation, presents many substantial advances in this fundamental tool for popgen work, and it will be of substantial
 interest to the readers of GENETICS. We hope that you agree.

### 2021-11-14 paper.tex

@@
 \author[8]{Bjarki Eldon}
-\author[9]{Castedo E. Ellerman}
+\author[9]{E. Castedo Ellerman}
 \author[10,11]{Jared G. Galloway}
@@
 
-\begin{figure}
-\begin{center}
-    \includegraphics{figures/ancestry-perf}
-\end{center}
-\caption{\label{fig-ancestry-perf}
-    Running time for \msprime\
-    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
-    Each point is the run time of one simulation,
-    for various values of effective population size ($N_e$),
-    chromosome length in Morgans ($L$), and number of samples ($n$).
-    Run time scales quadratically with the product of $N_e$ and $L$,
-    shown on the horizontal axis. For example, (A) shows that
-    1,000 samples of 1 Morgan-length chromosomes
-    from a population of $N_e=2,000$ diploids
-    would take about 2 seconds, and (equivalently) that the same number of
-    0.01 Morgan segments with $N_e=200,000$ would take the same time.
-    Since recombination rate in these simulations was $10^{-8}$,
-    $L$ is the number of base pairs divided by $10^8$.
-    The black lines are quadratic fits separately in each panel
-    and sample size. Vertical grey lines show the approximate values of
-    $N_e L$ for chromosome 1 in three species, using values from
-    the \stdpopsim\ catalogue~\citep{adrion2020community}.
-}
-\end{figure}
-
-% Output from  python3 evaluation/plot.py ancestry-perf
-% Predicted time for DroMel chr2L with n = 1000 = 107.85106995938845 hours
-% Predicted time for DroMel chr2L with n = 100000 = -511.74273443828514 hours
-% Predicted time for DroMel chr2L with n = 1000 = 177.40184421122058 hours
-% Predicted time for DroMel chr2L with n = 100000 = 361.90415260982536 hours
-
 In human-like parameter regimes and for large sample sizes,
@@
 
-For large $\rho$, recombination events in trapped ancestral material will
-dominate, and so we can use this as a proxy for the overall number
-of events in Hudson's algorithm.
+Recombination events have a dominant effect on the execution time of
+simulations. For details on the time complexity of the simulation algorithm,
+see the appendix: \nameref{app-time-complexity}.
+
+\subsection*{Gene conversion}
+Gene conversion is a form of recombination that results in the transfer
+of a short segment of genetic material,
+for example between homologous chromosomes~\citep{chen2007gene}.
+Since gene conversion impacts much shorter segments than
+crossover recombination (typically below 1kb) it
+affects patterns of linkage disequilibrium differently~\citep{korunes2017gene}.
+\cite{wiuf2000coalescent} modelled gene conversion in the coalescent
+via a rate at which gene conversion events are initiated
+along the genome and a geometrically distributed tract length.
+In terms of the ancestral process, gene conversion differs from
+crossover recombination (as described in the previous section)
+in that it extracts a short tract of ancestry into
+an independent lineage, rather than splitting ancestry
+to the left and right of a given breakpoint.
+We have implemented this model of gene
+conversion in \msprime\ 1.0, and validated the output against
+\ms\ and analytical results~\citep{wiuf2000coalescent}.
+
+\begin{figure}
+\begin{center}
+\includegraphics{figures/gc-perf}
+\end{center}
+\caption{\label{fig-gc-perf}Comparison of simulation performance
+using \msprime, \SimBac, and \FastSimBac\ for varying sample sizes,
+with parameters roughly equivalent to the current estimate
+for \textit{E.~coli}~\citep{lapierre2016the}:
+a 4.5Mb genome with scaled gene conversion rate of 0.015 and
+a mean tract length of 500.
+We report (A) the total CPU time and (B) maximum memory usage
+averaged over 5 replicates (Intel Xeon E5-2680 CPU).
+We did not run \SimBac\ beyond first two data points because
+of the very long running times.}
+% The genome length was set to 4.5 Mb, the scaled per site
+% gene conversion rate was set to $N_e g = \gamma = 0.015$ and
+% the mean gene conversion tract length was set to $L = 500$, which is close to
+% estimated parameters for \textit{E.~coli}, i.e.\ an effective populations size
+% $N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
+% with mean tract length $L = 542$~\citep{lapierre2016the}.}
+\end{figure}
+
+Gene conversion is particularly useful to model homologous recombination in
+bacterial evolution,
+and so we compare the performance of \msprime\ with gene conversion to
+two specialised bacterial simulators,
+\SimBac~\citep{brown2016simbac} and \FastSimBac~\citep{demaio2017the}.
+Figure~\ref{fig-gc-perf}A shows that \msprime\ is far more efficient than
+both \SimBac\ and the SMC-based approximation \FastSimBac.
+Figure~\ref{fig-gc-perf}B shows that \msprime\ requires somewhat more memory
+than \FastSimBac, (as expected since \FastSimBac\ uses a left-to-right
+SMC approximation) but is still reasonably modest at around 1GiB for a simulation
+of 500 whole \emph{E.~coli} genomes.
+However, \msprime\ is currently lacking many of the specialised features
+required to model bacteria, and so an important avenue for future work
+is to add features such as circular genomes
+and bacterial gene transfer~\citep{baumdicker2014AGTG}.
+
+In terms of predicting the run time for a simulation including gene conversion,
+we recommend following the same approach as discussed in the previous section:
+run a number of simulations for short genome lengths, fit a quadratic to the
+observed CPU times, and use this to predict run time for larger simulations.
+Depending on the relative contributions of gene conversion and crossover
+recombination, this may be an over-estimate since gene conversion events
+tend to generate less trapped ancestral material than crossovers.
+Thus, simulations using mammalian-like gene conversion parameters
+may run faster than simulations in which an equivalent amount of
+crossover recombination is imposed.
+% TODO check this again!
+Since each gene conversion creates two breakpoints and a crossover
+creates only one, we expect the output tree sequence for a given rate
+of gene conversion to be roughly twice the size of the output from a simulation
+with the same rate of crossovers.
+
+\subsection*{Demography}
+\label{sec-demography}
+% TODO should we put in some citations here for the models? Hardly seems
+% necessary.
+One of the key applications of population genetic simulations is to generate
+data for complex demographies. Beyond idealised cases such as stepping-stone or
+island models, or specialised cases such as isolation-with-migration models,
+analytical results are rarely possible. Simulation is therefore integral to the
+development and evaluation of methods for demographic inference. The demography
+model in \msprime\ is directly derived from the approach used in \ms, and
+supports an arbitrary number of randomly mating populations exchanging migrants
+at specified rates. A range of demographic events are supported, which allow for
+varying population sizes and growth rates, changing migration rates over time,
+as well as population splits, admixtures and pulse migrations.
+The location of sampled lineages
+can be tracked through time in as much detail as required: each tree node is
+automatically associated with the population in which it arose, the location of
+lineages can be recorded at any given time via census events, or every lineage
+migration can be recorded.
+Large demographic models can be simulated efficiently in version \msprime\ 1.0,
+since we only consider populations that contain lineages and
+have non-zero migration rates when generating migration event
+waiting times. This is a considerable improvement
+over version 0.x, which scaled quadratically with the number of populations.
+
+A major change for \msprime\ 1.0 is the introduction of the new Demography API,
+designed to address a design flaw in the \msprime\ 0.x interface which led to
+a number of avoidable errors in downstream
+simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
+separate parameters be provided to the \texttt{simulate} function to describe a
+demographic model, making it easy to accidentally omit information. The 1.0 API
+resolves this issue by creating a new \texttt{Demography} class, which
+encapsulates all information about the demographic model, and fully decouples
+the definition from other simulation details. An instance of this class is then
+provided as a parameter to the new \texttt{sim\_ancestry} function,
+substantially reducing the potential for error.
+Another improvement over the
+0.x APIs is the introduction of explicit population split and admixture events,
+and a population state machine that ensures that lineages cannot migrate into
+(or be sampled from) inactive populations. This demography model is compatible
+with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
+class supports importing and exporting Demes models.
+Models previously constructed using the 0.x API can be seamlessly imported into
+the \texttt{Demography} class, and we also support importing demographic
+models from Newick species trees and the output of programs
+like *BEAST~\citep{heled2009bayesian}.
+
+The \texttt{DemographyDebugger} provides detailed information about
+demographic models as well as numerical methods to make predictions
+about these models. For example, we can compute
+the coalescence rates for two or more lineages
+drawn from populations at specified times in the past,
+which can be inverted to obtain the
+``inverse instantaneous coalescence rate'' of \citet{chikhi2018iicr}.
+Many popular approaches in population genetics use the distribution of
+coalescence rates between pairs of lineages in one or more populations to infer
+effective population sizes over
+time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
+split times and subsequent migration rates between
+populations~\citep{wang2020tracking}.
+These numerical methods provide a valuable
+ground-truth when evaluating such inference methods, as illustrated by
+\cite{adrion2020community}.
+
+\subsection*{Instantaneous bottlenecks}
+
+A common approach to modelling the effect of demographic history on genealogies
+is to assume that effective population size ($N_e$) changes in discrete steps
+which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
+keightley2007joint,li2011inference}. In this setting of piecewise constant
+$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
+reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
+recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
+bottlenecks are short both on the timescale of coalescence and mutations,
+there may be little information about the duration of a bottleneck
+($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
+assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
+and generate a sudden burst of coalescence events (a multiple merger event) in
+the genealogy. The strength of the bottleneck $B$ can be thought of as an
+(imaginary) time period during which coalescence events are collapsed,
+i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
+single pair of lineages entering the bottleneck coalesce during the bottleneck
+is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
+attractive and both analytic results and empirical
+inference~\citep{griffiths1994sampling, birkner2009modified,galtier2000detecting,
+bunnefeld2015inferring} have been developed under this model, there has
+been no software available to simulate data under instantaneous
+bottleneck histories.
+
+We have implemented instantaneous bottlenecks in \msprime~1.0
+using a variant of Hudson's linear time single-locus coalescent
+algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
+by adding events to the \texttt{Demography} class
+(see the \nameref{sec-demography} section)
+and can be used in combination with any other demographic modelling
+features. We have validated the results of these simulations by comparing
+against analytical expectations for coalescence times and the
+site frequency spectrum~\citep{bunnefeld2015inferring}.
+
+\subsection*{Multiple merger coalescents}
+
+Kingman's coalescent assumes that only two ancestral lineages can merge at
+each merger event. Although this is generally a reasonable approximation, there
+are certain situations in which the underlying mathematical assumptions are
+violated. For example in certain highly fecund organisms
+\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
+to produce numbers of offspring on the order of the population size and
+therefore a few individuals may produce the bulk of the offspring in any given
+generation~\citep{hedgecock_94}. These population dynamics violate basic
+assumptions of the Kingman coalescent, and are better modelled by
+`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
+two lineages can merge in a given event. Multiple-merger coalescent processes
+have also been shown to be relevant for modelling the effects of selection on
+gene genealogies
+\citep{Gillespie909,DS04,desai2013genetic,neher2013genealogies,schweinsberg2017rigorous}.
+
+Although multiple merger coalescents have been of significant theoretical
+interest for around two decades, there has been little practical software
+available to simulate these models.
+\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
+simulate a related spatial continuum model~\citep{barton2010new},
+\cite{zhu2015hybrid} simulate genealogies within a species tree
+based on a multiple-merger model, and
+\cite{becheler2020occupancy} provide a general method for simulating
+multiple merger processes
+as part of the Quetzal framework~\citep{becheler2019quetzal}.
+The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
+also includes a number of extensions to the Beta-coalescent.
+None of these methods work with large genomes, and very little work
+has been performed on simulating multiple merger processes with recombination.
+
+We have added two multiple merger coalescent models in \msprime\ 1.0, the
+Beta-coalescent~\citep{schweinsberg03} and
+``Dirac''-coalescent~\citep{BBE13}, allowing us to efficiently simulate
+such models with recombination for the first time.
+These simulation models have been extensively validated against
+analytical results from the site frequency
+spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
+ as well as more general properties of coalescent processes.
+See the \nameref{app-multiple-mergers} appendix for more details and model derivations.
+
+\subsection*{Ancestral Recombination Graphs}
+\label{sec-arg}
+
+The Ancestral Recombination Graph (ARG) was introduced by
+Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
+to represent the stochastic process of the coalescent with recombination as a graph.
+This formulation is complementary to Hudson's earlier
+work~\citep{hudson1983properties}, and substantially increased our theoretical
+understanding of recombination. In Griffiths' ARG formulation,
+a realisation of the coalescent with recombination is a graph in which
+vertices represent common ancestor or recombination events, and edges represent
+lineages. There is the ``big'' ARG, in which we track lineages arising out of
+recombinations regardless of whether they carry ancestral
+material~\citep{ethier1990two},
+% and
+% consequently wait an exponentially large time until all sampled lineages
+% coalesce into a single ancestor~\citep{ethier1990two},
+and the ``little'' ARG in which we only track genetic ancestors.
+Over time, usage of the term
+has shifted away from its original definition as a stochastic process,
+to being interpreted as a representation of a particular genetic ancestry
+as a graph, without necessarily following the specific details of the Griffiths
+formulation~\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
+Under the latter interpretation,
+the tree sequence encoding of genetic ancestry (described above)
+clearly \emph{is} an ARG: the nodes and edges
+define a graph in which edges are annotated with the
+set of disjoint genomic intervals
+through which ancestry flows.
+
+\begin{figure}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
+\caption*{(A)}
+    \begin{tikzpicture}[scale=0.45,nodestyle/.style={font=\scriptsize}]
+
+    \draw (-2,0) -- (-2,3) -- (0, 3);
+    \draw (1,0) -- (1,1);
+    \draw (0,1) -- (2,1) -- (2,2) -- (4,2) -- (4,0);
+    \draw (3,2) -- (3,5);
+    \draw (0,1) -- (0,3);
+    \draw (-1, 3) -- (-1, 5) -- (3,5);
+
+    \draw[fill] (-2,0) circle [radius=0.05];
+    \draw[fill] (-1,3) circle [radius=0.05];
+    \draw[fill] (1,0) circle [radius=0.05];
+    \draw[fill] (4,0) circle [radius=0.05];
+    \draw[fill] (1,1) circle [radius=0.05];
+    \draw[fill] (3,2) circle [radius=0.05];
+    \draw[fill] (1,5) circle [radius=0.05];
+
+    \node[nodestyle] [below] at (-2,0) {0};
+    \node[nodestyle] [below] at (1,0) {1};
+    \node[nodestyle] [below] at (4,0) {2};
+    \node[nodestyle] [above] at (1,1) {3};
+    \node[nodestyle] [below right] at (1,1) {$x = 0.3$};
+    \node[nodestyle] [above right] at (3,2) {4};
+    \node[nodestyle] [above left] at (-1,3) {5};
+    \node[nodestyle] [above] at (1,5) {6};
+
+    \end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
+\caption*{(B)}
+\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
+
+\tikzset{greynode/.style={circle,fill,inner sep=0.7},
+nodelabel/.style={font=\scriptsize}}
+
+% Left sample nodes
+\node (s0) [greynode] {};
+\node (s1) [right=of s0,greynode] {};
+\node (s2) [right=of s1,greynode] {};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Right sample nodes
+\node (rightTree) at (4, 0) {};
+\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
+\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
+\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
+\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Non-sample nodes
+\node [greynode] (s3) at ($(s1) + (0,1)$) {};
+\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
+\node [greynode] (s4) at ($(s2) + (0,2)$) {};
+\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
+\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
+\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
+\node [greynode] (s6) at ($0.25*(s0) + 0.25*(s1) + 0.5*(s2) + (0,5)$) {};
+\node [greynode] (s6r) at ($0.5*(s0r) + 0.25*(s1r) + 0.25*(s2r) + (0,5)$) {};
+\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
+\foreach \u/\lab in {s4/4, s4r/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5, s5r/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s6/6, s6r/6} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\foreach \child/\parent in {
+	s5/s0, s5/s1, s6/s5, s6/s2,
+	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
+}
+	\draw (\child) -| (\parent);
+
+% Axes
+\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
+\node (topAx) at (0,-1.0) {};
+\node (topLeft) at ($(s0) + (0,-1.0)$) {};
+\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
+\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
+\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
+\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
+\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+
+% Interval endpoints
+\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, 5.5);
+
+% Axis titles
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.33\textwidth}
+\centering
+\caption*{(C)}
+\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
+
+\tikzset{greynode/.style={circle,fill,inner sep=0.7},
+nodelabel/.style={font=\scriptsize}}
+
+% Left sample nodes
+\node (s0) [greynode] {};
+\node (s1) [right=of s0,greynode] {};
+\node (s2) [right=of s1,greynode] {};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Right sample nodes
+\node (rightTree) at (4, 0) {};
+\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
+\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
+\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
+\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+
+% Non-sample nodes
+%\node [greynode] (s3) at ($(s1) + (0,1)$) {};
+%\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
+%\node [greynode] (s4) at ($(s2) + (0,2)$) {};
+\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
+\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
+%\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
+\node [greynode] (s6) at ($0.25*(s0) + 0.25*(s1) + 0.5*(s2) + (0,5)$) {};
+\node [greynode] (s6r) at ($0.5*(s0r) + 0.25*(s1r) + 0.25*(s2r) + (0,5)$) {};
+%\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
+\foreach \u/\lab in {s4r/3} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/4} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s6/5, s6r/5} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\foreach \child/\parent in {
+	s5/s0, s5/s1, s6/s5, s6/s2,
+	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
+}
+	\draw (\child) -| (\parent);
+
+% Axes
+\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
+\node (topAx) at (0,-1.0) {};
+\node (topLeft) at ($(s0) + (0,-1.0)$) {};
+\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
+\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
+\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
+\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
+\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+
+% Interval endpoints
+\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, 5.5);
+
+% Axis titles
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
+\end{tikzpicture}
+\end{subfigure}
+\caption{\label{fig-arg} (A) A simple ARG in which a recombination
+occurs at position 0.3; (B) the equivalent topology depicted as a tree
+sequence, including the recombination node; (C) the same tree sequence
+topology ``simplified'' down to its minimal tree sequence representation.
+Note that the internal nodes have been renumbered in the simplified
+representation, so that, e.g., node 5 in (C) corresponds to node 6 in
+(A) and (B).}
+\end{figure}
+
+For our purposes, an ARG is a realisation of the coalescent with
+recombination, in the Griffiths (little ARG) sense.
+As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
+works by dynamically traversing a little ARG.
+The graph is not explicitly represented in memory, but is partially
+present through the extant lineages and the ancestral material they carry
+over time. We do not output the graph directly, but
+rather store the information required to recover the genealogical
+history as nodes and edges in a tree sequence.
+This is far more efficient than outputting the simulated ARG in its entirety.
+For a given scaled recombination rate $\rho$
+(setting aside the dependency on the sample size $n$)
+we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
+in an ARG is $O(\rho^2)$,
+whereas the size of the tree sequence encoding is
+$O(\rho)$~\citep{kelleher2016efficient}.
+This difference between a
+quadratic and a linear dependency on $\rho$ is profound, and shows why
+large simulations cannot output an ARG in practice.
+
+Although by default \msprime\ outputs tree sequences that contain full information
+about the genealogical trees, their correlation structure along the chromosome,
+and the ancestral genomes on which coalescences occurred, some information
+is lost in this mapping down from ARG space to the minimal tree sequence
+form. In particular, we lose
+information about ancestral genomes that were common ancestors but
+in which no coalescences occurred, and also information about the precise time
+and chromosomal location of recombination events. In most cases, such
+information is of little relevance as it is in principle unknowable,
+but there are occasions such as visualisation or computing likelihoods (see
+below) in which it is useful. We therefore provide
+the \texttt{record\_full\_arg} option in \msprime\
+to store a representation of the complete ARG traversed during simulation.
+This is done by storing extra nodes (marked with specific flags, so they
+can be easily identified later) and edges in the tree sequence
+(Fig.~\ref{fig-arg}).
+One situation in which a record of the full ARG is necessary is when we
+wish to compute likelihoods during inference.
+The likelihood is a central quantity in evaluating the plausibility of a putative
+ancestry as an explanation of DNA sequence data, both directly through
+e.g.~approaches based on maximum likelihood, and as an ingredient of
+methods such as
+Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
+We provide functions to compute the likelihood of ARG realisations
+and mutational patterns under
+the standard coalescent and infinite sites mutation model.
+For details, see the appendix: \nameref{app-likelihoods}.
+
+\subsection*{Selective sweeps}
+\label{sec-selection}
+Another elaboration of the standard neutral coalescent with recombination
+is the addition of selective
+sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
+Sweeps are modelled by creating a structured population
+during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
+in which lineages may transit between favoured and unfavoured backgrounds through
+recombination. This approach allows for many selective sweep scenarios to be simulated
+efficiently, including recurrent, partial, and soft selective sweeps.
+However this efficiency comes at the cost of flexibility in comparison to forwards in
+time simulation.
+Several specialised simulators have been developed to simulate
+sweeps in the coalescent,
+including \texttt{SelSim}~\citep{spencer2004selsim},
+\texttt{mbs}~\citep{teshima2009mbs}, \msms~\citep{ewing2010msms},
+\cosiTwo~\citep{shlyakhter2014cosi2}
+and \discoal~\citep{kern2016discoal}.
+
+\begin{figure}
+\begin{center}
+\includegraphics{figures/sweeps-perf}
+\end{center}
+\caption{\label{fig-selection-perf}Comparison of selective sweep simulation
+performance in \msprime\ and \discoal\ (Intel Xeon E5-2680 CPU).
+We report the total CPU time and maximum memory usage when simulating 100
+replicates for 10 samples in a model
+with a single selective sweep in its history where the beneficial
+allele had a scaled selection coefficient of $2Ns=1000$,
+a per-base recombination rate of $10^{-9}$ and sequence
+length varying from 1kb--200kb.}
+\end{figure}
+
+Selective sweeps are implemented in the coalescent as a two
+step-process: first generating an allele frequency trajectory,
+and then simulating a structured coalescent process conditioned
+on that trajectory.
+Following \discoal, we generate sweep trajectories in \msprime\ using a
+jump process approximation to the conditional
+diffusion of an allele bound for
+fixation~\citep{coop2004ancestral}.
+The jump process moves back in time following the beneficial
+allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
+the origination of the allele at $p=1/(2N)$, tracking time in small
+increments $\delta t$. Then, given the frequency $p$ at time $t$,
+the frequency $p^\prime$ at time $t + \delta t$ is given by
+\begin{equation*}
+p^\prime =
+\begin{cases}
+    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
+    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
+\end{cases}
+\end{equation*}
+where
+\begin{eqnarray*}
+	\mu(p)=\frac{\alpha p (1 - p)}{\tanh(\alpha(1-p))}.
+\end{eqnarray*}
+Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
+can be generalised straightforwardly to include arbitrary dominance.
+We can also define trajectories to model neutral alleles and soft
+selective sweeps, which we plan as future additions to \msprime.
+
+Then, given a randomly generated allele frequency trajectory under
+the above model, the simulation of a sweep works by assigning lineages to
+two different structured coalescent ``labels'', based on
+whether they carry the beneficial allele.
+The allele frequency trajectory determines the relative sizes
+of the ``populations'' in these labels over time, and therefore
+the rates at which various events occur.
+Common ancestor events can then only merge lineages from \emph{within}
+a label, but lineages can transfer from one label to the other
+(i.e., from the advantageous to disadvantageous backgrounds,
+and vice versa) as a result of recombination events.
+Once we have reached the end of the simulated trajectory the sweep
+is complete, and we remove the structured coalescent labels.
+Simulation may then resume under any other ancestry model.
+
+Fig.~\ref{fig-selection-perf} compares the performance of \msprime\
+and \discoal\ under a simple sweep model, and shows
+that \msprime\ has far better
+CPU time and memory performance.
+Since our implementation uses the abstract label system mentioned above,
+adding support for similar situations,
+such as inversions~\citep{peischl2013sequential}, should be straightforward.
+
+\subsection*{Discrete time Wright-Fisher}
+The coalescent is an idealised model and makes many simplifying assumptions,
+but it is often surprisingly robust to violations of these
+assumptions~\citep{wakeley2012gene}. One situation in which the
+model does break down is the combination of large sample size
+and long recombining genomes, where the large
+number of recombination events in the recent past results in
+more than the biologically possible $2^t$ ancestors in
+$t$ diploid generations~\citep{nelson2020accounting}.
+This pathological behaviour results in
+identity-by-descent,
+long-range linkage disequilibrium and ancestry patterns deviating from
+Wright-Fisher expectations, and the bias grows with larger sample
+sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
+Precisely this problem occurs when simulating modern human datasets,
+and we have implemented a Discrete Time Wright-Fisher (DTWF) model
+in \msprime\ to address the issue. The DTWF simulates backwards in
+time generation-by-generation so that each gamete has a unique
+diploid parent, and multiple recombinations within a generation results in
+crossover events between the same two parental haploid copies.
+The method is described in more detail by~\cite{nelson2020accounting}.
+
+\begin{figure}
+\begin{center}
+\includegraphics{figures/dtwf-perf}
+\end{center}
+\caption{\label{fig-dtwf-perf} Comparison of Discrete Time Wright-Fisher
+(DTWF) simulation performance in \msprime\ and \ARGON\ (Intel Xeon E5-2680 CPU).
+We simulated ancestry for a sample of 1000 haploids from a population of 10000,
+and report the (A) total CPU time and (B) maximum memory usage for varying
+sequence lengths, and a per-base recombination rate of $10^{-8}$. Each
+point is the average over $5$ replicate simulations.
+We show observations for \ARGON, \msprime's DTWF implementation (``DTWF'')
+and a hybrid simulation of 100 generations of the DTWF followed by
+the standard coalescent with recombination model (``DTWF + Hudson'').
+Memory usage for \msprime's DTWF and hybrid simulations are very similar.
+We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
+to ensure we are measuring only ancestry simulation time.}
+\end{figure}
+
+Fig.~\ref{fig-dtwf-perf} shows that \msprime\ simulates the DTWF
+more quickly and requires substantially less memory than
+\ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
+However, the generation-by-generation approach of the DTWF is less
+efficient than the coalescent with recombination when the
+number of lineages is significantly less than the population size
+(the regime where the coalescent is an accurate approximation),
+which usually happens in the quite recent
+past~\citep{bhaskar2014distortion}.
+We therefore support changing the simulation model during a simulation
+so that we can run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}.
+Any number of different simulation models can be combined, allowing for the
+flexible choice of simulation scenarios.
+As the discrete time Wright-Fisher model improves accuracy of genealogical
+patterns in the recent past, we can simulate the recent history using this
+model and then switch to the standard coalescent to more efficiently simulate
+the more ancient history.
+
+\subsection*{Integration with forward simulators}
+A unique feature of \msprime\ is its ability to simulate
+genetic ancestries by extending an existing partial
+genetic ancestry. Given a tree sequence that
+is complete up until time $t$ ago as input
+(where marginal trees may or may not have fully coalesced),
+\msprime\ can efficiently obtain the segments of ancestral material
+present at this time, and then run the simulation backwards in time from there.
+This allows a simulated ancestry to be produced by any
+number of different processes across disjoint time slices.
+In practice this feature is used to ``complete''
+forwards-time ancestry simulations~\citep{kelleher2018efficient}
+that may have not fully coalesced. This process
+(``recapitation'') can be orders of magnitude faster than
+the standard approach of neutral burn-in; see
+\cite{haller2018tree} for more details and examples.
+This interoperability between simulators, where a partial ancestry
+simulation produced by \SLiM~\citep{haller2019slim}
+or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
+by another simulator, with complete information
+retained---at scale---is unprecedented.
+There may be an opportunity for
+other forward genetic simulators~\citep[e.g.][]{gaynor2021alphasimr}
+to leverage the tree sequence data format and associated tools.
+
+\subsection*{Development model}
+\texttt{Msprime} has a large number of features, encompassing
+the functionality of several more specialised simulators
+while maintaining excellent performance.
+It is developed by a geographically distributed team of volunteers under an
+open source community development model, with a strong emphasis
+on code quality, correctness, good documentation, and inclusive development.
+As in any large code base,
+unit tests play a key role in ensuring that new additions behave
+as expected and \msprime\ has an extensive suite.
+As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
+lines of Python, with suites of 122 C tests (7K lines of code) and
+1350 Python tests (22K lines of code). These tests are run
+automatically on different operating systems on each pull request
+(where a contributor proposes a code change), using standard Continuous
+Integration (CI) methodology. Other CI services
+check for common errors, code formatting issues, and produce reports on
+the level of test coverage for the proposed change.
+
+Unit tests are vital for ensuring software quality and correctness, but
+they are usually of little value in assessing the statistical properties
+of simulations. To validate the correctness of simulation output
+we maintain a suite of statistical tests (as of 1.0.0,
+217 validation tests, in 6K lines of code). These consist of running many
+replicate simulations to check the properties of the output
+against other simulators, and where possible against analytical results.
+For example, simulations of complex demography are validated against \ms,
+selective sweeps against \discoal, and Wright-Fisher simulations
+against forwards in time simulations in \SLiM.
+This suite of tests is run before every release, to ensure
+that statistical errors have not been introduced.
+
+More visibly to the end user, we also have a high standard for documentation,
+with precise, comprehensive, and cross-linked documentation
+that is automatically built from the code base
+and served through the website \texttt{https://tskit.dev}.
+With the goal of lowering the entry barrier to new users,
+we have invested significant effort in writing examples and introductions,
+and making common tasks discoverable.
+We also view contributions to documentation as equally important to the project
+as writing code or designing methods: what use would it be
+to write reliable, stable software if no-one used it?
+
+An important goal of \msprime's development model is to maximise
+accessibility for prospective users and contributors, and
+to encourage diversity in our community.
+Gender and racial inequality caused by discrimination and marginalisation
+is a major problem across
+the sciences~\citep{wellenreuther2016women,shannon2019gender}
+and in open source software development~\citep{trinkenreich2021women}.
+Within our field, the contribution of women to early computational methods
+in population genetics was marginalised~\citep{dung2019illuminating},
+and women continue to be under-represented
+in computational biology~\citep{bonham2017women}.
+The authorship of our paper reflects these trends,
+with a skew towards men and affiliations in the USA
+and Europe. We know the importance of creating and strengthening networks to
+develop and maintain a diverse community of contributors, and we are committed
+to fostering a supportive and collaborative environment that helps to address these
+inequalities in our field.
+
+\section*{Discussion}
+
+% We can simulate so much more now
+The 1.0 release of \msprime\ marks a major increase in the
+breadth of available features and the
+potential biological realism of simulations.
+These abilities will allow researchers to perform more robust power analyses,
+more reliably test new methods,
+carry out more reliable inferences,
+and more thoroughly explore the properties of theoretical models.
+Despite this complexity and generality,
+\msprime's performance is state-of-the-art
+and all features are extensively tested and statistically validated.
+These advances have only been possible thanks to a distributed,
+collaborative model of software development,
+and the work of many people.
+% We hope the tool continues to be useful to many.
+
+% Simulation software is historically a beautiful shitshow
+
+Even though simulation has long been a vital tool in population genetics,
+such collaborative software development has historically been uncommon.
+A huge proliferation of tools have been published
+(the references here are not exhaustive)
+and only a small minority of these are actively developed and
+maintained today. The ecosystem is highly fragmented, with numerous different
+ways of specifying parameters and representing results, and there are
+significant software quality issues at all stages. This is unsurprising, since
+the majority of simulation software development is performed by students,
+often without formal training in software development.
+The result resembles Haldane's sieve for new mutations:
+many new pieces of software stay permanently on a dusty shelf of supplementary materials,
+while some of those that prove particularly useful when new (like dominant alleles)
+are quickly adopted.
+Although this has produced many good tools
+and enabled decades of research,
+it also represents a missed opportunity to invest as a community
+in shared infrastructure and mentorship in good software development practice.
+
+% THERE IS A BETTER WAY!!!
+Scientific software is vital apparatus, and must be engineered
+to a high quality if we are to trust its results. There is a growing
+realisation across the sciences
+% More citations?
+\citep[e.g.][]{siepel2019challenges,harris2020array,gardner2021sustained}
+that investing in
+shared community infrastructure produces better results than a
+proliferation of individually maintained tools, allowing scientists
+to focus on their specific questions rather than software engineering.
+\texttt{Msprime} 1.0 is the result of such a community process,
+with features added by motivated users, taking advantage of the
+established development practices and infrastructure.
+Software development in a welcoming community,
+with mentorship by experienced developers,
+is a useful experience for many users.
+The skills that contributors learn
+can lead to greatly increased productivity in subsequent
+work (e.g., through more reliable code and better debugging skills).
+We hope that users who find that
+features they require are missing will continue to contribute to
+\msprime, leading to a community project that is both high quality
+and sustainable in the long term.
+
+The succinct tree sequence data structure developed
+for \msprime\ provides a view of not only
+genetic variation, but also the genetic ancestry that produced that variation.
+Recent breakthroughs in methods to infer genetic ancestry in recombining
+organisms~\citep{rasmussen2014genome,kelleher2019inferring,
+speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
+have made it possible to estimate such ancestry from real data at scale for
+the first time~\citep{harris2019database,tang2019genealogy}.
+Given such inferred ancestry, many exciting applications
+become possible. For example, \cite{osmond2021estimating}
+developed a method to estimate the location of genetic ancestors
+based on inferred trees, and other uses are sure to follow.
+Since the inferred genetic ancestry
+becomes the input for other downstream inferences, it is vitally
+important that these primary inferences are thoroughly validated,
+with the detailed properties of the inferred ancestries
+catalogued and understood.
+\texttt{Msprime} will continue to be an important tool for these inferences and validations,
+and in this context the ability to interoperate
+with other methods---particularly forwards simulators---through
+the succinct tree sequence data structure and \tskit\ library
+will be essential.
+% this final sentence is hard to parse
+
+\section*{Availability}
+\texttt{Msprime} is freely available under the terms of the
+GNU General Public License v3.0, and can be installed from the Python
+Package Index (PyPI) or the \texttt{conda-forge} conda channel.
+Development is conducted openly on GitHub at
+\texttt{https://github.com/tskit-dev/msprime/}.
+The documentation for \texttt{msprime} is available at
+\texttt{https://tskit.dev/msprime/docs/}.
+The source code for all the evaluations and figures in this manuscript
+is available at \texttt{https://github.com/tskit-dev/msprime-1.0-paper/}.
+
+\section*{Acknowledgements}
+We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
+on the manuscript.
+% alphabetised by author. Names spelled out if there's an initial collision
+ADK was supported by NIH awards R01GM117241 and R01HG010774.
+BE was supported by  DFG grant  273887127 through Priority Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
+FB is funded by the Deutsche Forschungsgemeinschaft EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
+Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
+Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
+Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
+Jerome Kelleher is supported by the Robertson Foundation.
+SG acknowledges funding from the Canada Research Chairs Program, from the Canadian Institutes of Health Research PJT 173300, and from the Canadian Foundation for Innovation.
+
+\bibliographystyle{plainnat}
+\bibliography{paper}
+
+%% local definitions for section multiple merger coalescents
+ \newcommand{\be}{\begin{equation}}
+ \newcommand{\ee}{\end{equation}}
+ \newcommand{\bd}{\begin{displaymath}}
+ \newcommand{\ed}{\end{displaymath}}
+\newcommand{\IN}{\ensuremath{\mathds{N}}}%
+\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
+\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
+\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%
+
+\NewEnviron{esplit}[1]{%
+\begin{equation}
+\label{#1}
+\begin{split}
+  \BODY
+\end{split}\end{equation}
+}
+
+\setcounter{secnumdepth}{2} % Print out appendix section numbers
+
+\appendix
+\section*{Appendix}
+
+\subsection*{Mutation generation}
+\label{app-mutation-algorithm}
+
+The algorithm that \msprime\ uses to simulate mutations on a tree sequence
+proceeds in two steps:
+first, mutations are ``placed'' on the tree sequence
+(i.e., sampling their locations in time, along the genome, and on the marginal tree),
+and then the ancestral and derived alleles of each mutation are generated.
+All mutation models share the code to place mutations,
+but choose alleles in different ways.
+
+First, mutations are placed on the tree sequence under an inhomogeneous Poisson model
+by applying them independently to each edge.
+If an edge spans a region $[a, b)$ of the genome
+and connected parent and child nodes with times $s < t$,
+and the mutation rate locally is $\mu$,
+then the number of mutations on the edge is Poisson with mean $\mu (t-s) (b-a)$,
+and each mutation is placed independently at a position chosen uniformly in $[a,b)$
+and a time uniformly in $[s,t)$.
+In a discrete genome, all positions are integers and so more than one mutation
+may occur at the same position on the same edge.
+Otherwise (i.e., for an infinite-sites model),
+positions are rejection sampled to obtain a unique floating-point number.
+If an edge spans a region of the genome with more than one mutation rate,
+this is done separately for each sub-region on which the mutation rate is constant.
+Since each edge is processed independently, the algorithm
+scales linearly with the number of edges in the tree sequence.
+
+Next, alleles are chosen for each mutation.
+If the site was not previously mutated, then a new ancestral allele is chosen for the site,
+according to an input distribution of ancestral state allele probabilities.
+Then, each mutation on the tree is considered in turn,
+and a derived allele is randomly chosen based on the parental allele
+(which may be the ancestral allele or the derived allele of a previous mutation).
+Finally, information about the mutations are recorded in the site and mutation tables
+of the tree sequence.
+
+A mutation model must, therefore, provide two things:
+a way of choosing an ancestral allele for each new variant site,
+and a way of choosing a derived allele given the parental allele at each mutation.
+Perhaps the simplest mutation model implemented in \msprime\ is the
+\texttt{InfiniteAlleles} mutation model,
+which keeps an internal counter so that the requested alleles are
+assigned subsequent (and therefore unique) integers.
+
+The distribution of ancestral alleles
+is used to choose the allele present at the root of the tree at each mutated site,
+i.e., the \texttt{root\_distribution}.
+Mutation models with a finite possible set of alleles have a natural choice
+for this distribution---the \emph{stationary distribution} of the mutation process.
+(All mutation models are Markovian,
+so this may be found as the top left eigenvector of the mutation matrix.)
+This is the default in most models,
+except, e.g., the \texttt{BinaryMutationModel}, whose alleles are 0 and 1
+and always labels the ancestral allele ``0''.
+However, mutational processes are not in general stationary,
+so we often allow a different root distribution to be specified.
+
+Since the general algorithm above applies mutations at a single rate
+independent of ancestral state,
+a model in which different alleles mutate at different rates
+must necessarily produce some silent mutations,
+i.e., mutations in which the derived allele is equal to the parental allele.
+To illustrate this, consider a mutation model
+in which $A$ or $T$ mutates to a randomly chosen different nucleotide at rate $\alpha$
+and $C$ or $G$ mutates at rate $\beta$, with $\beta < \alpha$.
+To implement this, first place mutations at the largest total rate, which is $\alpha$.
+Then, at each site, choose an ancestral allele from the root distribution,
+and for each mutation, choose a derived allele as follows:
+if the parental allele is $A$ or $T$, then choose a random derived allele
+different to the parental allele;
+if the parental allele is $C$ or $G$,
+then choose the derived allele to be equal to the parent allele
+with probability $\beta/(\alpha + \beta)$,
+and randomly choose a different nucleotide otherwise.
+This produces the correct distribution by Poisson thinning:
+a Poisson process with rate $\alpha$ in which each point is discarded independently
+with probability $\beta / (\alpha + \beta)$ is equivalent to a Poisson
+process with rate $\beta$.
+All finite-state models (implemented under the generic \texttt{MatrixMutationModel} class)
+work in this way: mutations are placed at the maximum mutation rate,
+and then some silent mutations will result.
+
+In previous versions of \msprime, silent mutations were disallowed,
+and we could have removed them from the output entirely.
+However, we have chosen to leave them in, so that for instance
+simulating with the \texttt{HKY} mutation model will result in silent mutations
+if not all equilibrium frequencies are the same.
+The presence of silent mutations may at first be surprising
+but there is a good reason to leave them in:
+to allow layering of different mutation models.
+Suppose that we wanted to model the mutation process as a mixture of more than one model,
+e.g., Jukes-Cantor mutations at rate $\mu_1$, and
+HKY mutations occur at rate $\mu_2$.
+Layering multiple calls to \texttt{sim\_mutations} is allowed,
+so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
+and then add more with the \texttt{HKY} model at rate $\mu_2$.
+However, there is a small statistical problem:
+suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
+but then the HKY mutations inserts another mutation in the middle,
+resulting in $A \to C \to C$.
+If neither mutation model allows silent transitions,
+then this is clearly not correct,
+i.e., it is not equivalent to a model that simultaneously applies the two models.
+(The impact is small, however, as it only affects sites with more than one mutation.)
+The solution is to make the Jukes-Cantor model \emph{state-independent}
+(also called ``parent-independent''),
+by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
+\emph{independently} of the parent (so that 1/4 of mutations will be silent).
+If so---and, more generally, if the first mutational process put down is
+state-independent---then the result of sequentially applying the two mutation models
+is equivalent to the simultaneous model.
+To facilitate this, many mutation models have a \texttt{state\_independent} option
+that increases the number of silent mutations
+and makes the model closer to state-independent.
+
+Silent mutations are fully supported by \tskit, which correctly accounts for
+their presence when computing statistics and performing other operations.
+For example, silent mutations have no effect on calculations of nucleotide
+site diversity.
+
+\subsection*{Time complexity of the algorithm}
+\label{app-time-complexity}
+
+As discussed in the \nameref{sec-recombination} section, the time complexity of
+Hudson's algorithm is quadratic in the population scaled recombination rate
+$\rho = 4 N_e L$, where $L$ is the length of the genome in units of
+recombination distance. For large $\rho$, recombination events in trapped
+ancestral material will dominate, and so we can use this as a proxy for the
+overall number of events in Hudson's algorithm.
 \citet[Eq.~5.10]{hein2004gene} gave
@@
 given $\rho$ and $n$.
+
+\subsubsection*{Practical examples}
+
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/ancestry-perf}
+\end{center}
+\caption{\label{fig-ancestry-perf}
+    Running time for \msprime\
+    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
+    Each point is the run time of one simulation,
+    for various values of effective population size ($N_e$),
+    chromosome length in Morgans ($L$), and number of samples ($n$).
+    Run time scales quadratically with the product of $N_e$ and $L$,
+    shown on the horizontal axis. For example, (A) shows that
+    1,000 samples of 1 Morgan-length chromosomes
+    from a population of $N_e=2,000$ diploids
+    would take about 2 seconds, and (equivalently) that the same number of
+    0.01 Morgan segments with $N_e=200,000$ would take the same time.
+    Since recombination rate in these simulations was $10^{-8}$,
+    $L$ is the number of base pairs divided by $10^8$.
+    The black lines are quadratic fits separately in each panel
+    and sample size. Vertical grey lines show the approximate values of
+    $N_e L$ for chromosome 1 in three species, using values from
+    the \stdpopsim\ catalogue~\citep{adrion2020community}.
+}
+\end{figure}
+
+% Output from  python3 evaluation/plot.py ancestry-perf
+% Predicted time for DroMel chr2L with n = 1000 = 107.85106995938845 hours
+% Predicted time for DroMel chr2L with n = 100000 = -511.74273443828514 hours
+% Predicted time for DroMel chr2L with n = 1000 = 177.40184421122058 hours
+% Predicted time for DroMel chr2L with n = 100000 = 361.90415260982536 hours
 
@@
 using the timing of seconds-long runs is unlikely to be very accurate.
+
+\subsubsection*{Changing population size}
 
@@
 
-\subsection*{Gene conversion}
-Gene conversion is a form of recombination that results in the transfer
-of a short segment of genetic material,
-for example between homologous chromosomes~\citep{chen2007gene}.
-Since gene conversion impacts much shorter segments than
-crossover recombination (typically below 1kb) it
-affects patterns of linkage disequilibrium differently~\citep{korunes2017gene}.
-\cite{wiuf2000coalescent} modelled gene conversion in the coalescent
-via a rate at which gene conversion events are initiated
-along the genome and a geometrically distributed tract length.
-In terms of the ancestral process, gene conversion differs from
-crossover recombination (as described in the previous section)
-in that it extracts a short tract of ancestry into
-an independent lineage, rather than splitting ancestry
-to the left and right of a given breakpoint.
-We have implemented this model of gene
-conversion in \msprime\ 1.0, and validated the output against
-\ms\ and analytical results~\citep{wiuf2000coalescent}.
-
-\begin{figure}
-\begin{center}
-\includegraphics{figures/gc-perf}
-\end{center}
-\caption{\label{fig-gc-perf}Comparison of simulation performance
-using \msprime, \SimBac, and \FastSimBac\ for varying sample sizes,
-with parameters roughly equivalent to the current estimate
-for \textit{E.~coli}~\citep{lapierre2016the}:
-a 4.5Mb genome with scaled gene conversion rate of 0.015 and
-a mean tract length of 500.
-We report (A) the total CPU time and (B) maximum memory usage
-averaged over 5 replicates (Intel Xeon E5-2680 CPU).
-We did not run \SimBac\ beyond first two data points because
-of the very long running times.}
-% The genome length was set to 4.5 Mb, the scaled per site
-% gene conversion rate was set to $N_e g = \gamma = 0.015$ and
-% the mean gene conversion tract length was set to $L = 500$, which is close to
-% estimated parameters for \textit{E.~coli}, i.e.\ an effective populations size
-% $N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
-% with mean tract length $L = 542$~\citep{lapierre2016the}.}
-\end{figure}
-
-Gene conversion is particularly useful to model homologous recombination in
-bacterial evolution,
-and so we compare the performance of \msprime\ with gene conversion to
-two specialised bacterial simulators,
-\SimBac~\citep{brown2016simbac} and \FastSimBac~\citep{demaio2017the}.
-Figure~\ref{fig-gc-perf}A shows that \msprime\ is far more efficient than
-both \SimBac\ and the SMC-based approximation \FastSimBac.
-Figure~\ref{fig-gc-perf}B shows that \msprime\ requires somewhat more memory
-than \FastSimBac, (as expected since \FastSimBac\ uses a left-to-right
-SMC approximation) but is still reasonably modest at around 1GiB for a simulation
-of 500 whole \emph{E.~coli} genomes.
-However, \msprime\ is currently lacking many of the specialised features
-required to model bacteria, and so an important avenue for future work
-is to add features such as circular genomes
-and bacterial gene transfer~\citep{baumdicker2014AGTG}.
-
-In terms of predicting the run time for a simulation including gene conversion,
-we recommend following the same approach as discussed in the previous section:
-run a number of simulations for short genome lengths, fit a quadratic to the
-observed CPU times, and use this to predict run time for larger simulations.
-Depending on the relative contributions of gene conversion and crossover
-recombination, this may be an over-estimate since gene conversion events
-tend to generate less trapped ancestral material than crossovers.
-Thus, simulations using mammalian-like gene conversion parameters
-may run faster than simulations in which an equivalent amount of
-crossover recombination is imposed.
-% TODO check this again!
-Since each gene conversion creates two breakpoints and a crossover
-creates only one, we expect the output tree sequence for a given rate
-of gene conversion to be roughly twice the size of the output from a simulation
-with the same rate of crossovers.
-
-\subsection*{Demography}
-\label{sec-demography}
-% TODO should we put in some citations here for the models? Hardly seems
-% necessary.
-One of the key applications of population genetic simulations is to generate
-data for complex demographies. Beyond idealised cases such as stepping-stone or
-island models, or specialised cases such as isolation-with-migration models,
-analytical results are rarely possible. Simulation is therefore integral to the
-development and evaluation of methods for demographic inference. The demography
-model in \msprime\ is directly derived from the approach used in \ms, and
-supports an arbitrary number of randomly mating populations exchanging migrants
-at specified rates. A range of demographic events are supported, which allow for
-varying population sizes and growth rates, changing migration rates over time,
-as well as population splits, admixtures and pulse migrations.
-The location of sampled lineages
-can be tracked through time in as much detail as required: each tree node is
-automatically associated with the population in which it arose, the location of
-lineages can be recorded at any given time via census events, or every lineage
-migration can be recorded.
-Large demographic models can be simulated efficiently in version \msprime\ 1.0,
-since we only consider populations that contain lineages and
-have non-zero migration rates when generating migration event
-waiting times. This is a considerable improvement
-over version 0.x, which scaled quadratically with the number of populations.
-
-A major change for \msprime\ 1.0 is the introduction of the new Demography API,
-designed to address a design flaw in the \msprime\ 0.x interface which led to
-a number of avoidable errors in downstream
-simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
-separate parameters be provided to the \texttt{simulate} function to describe a
-demographic model, making it easy to accidentally omit information. The 1.0 API
-resolves this issue by creating a new \texttt{Demography} class, which
-encapsulates all information about the demographic model, and fully decouples
-the definition from other simulation details. An instance of this class is then
-provided as a parameter to the new \texttt{sim\_ancestry} function,
-substantially reducing the potential for error.
-Another improvement over the
-0.x APIs is the introduction of explicit population split and admixture events,
-and a population state machine that ensures that lineages cannot migrate into
-(or be sampled from) inactive populations. This demography model is compatible
-with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
-class supports importing and exporting Demes models.
-Models previously constructed using the 0.x API can be seamlessly imported into
-the \texttt{Demography} class, and we also support importing demographic
-models from Newick species trees and the output of programs
-like *BEAST~\citep{heled2009bayesian}.
-
-The \texttt{DemographyDebugger} provides detailed information about
-demographic models as well as numerical methods to make predictions
-about these models. For example, we can compute
-the coalescence rates for two or more lineages
-drawn from populations at specified times in the past,
-which can be inverted to obtain the
-``inverse instantaneous coalescence rate'' of \citet{chikhi2018iicr}.
-Many popular approaches in population genetics use the distribution of
-coalescence rates between pairs of lineages in one or more populations to infer
-effective population sizes over
-time~\citep{li2011inference,sheehan2013estimating,schiffels2014inferring} or
-split times and subsequent migration rates between
-populations~\citep{wang2020tracking}.
-These numerical methods provide a valuable
-ground-truth when evaluating such inference methods, as illustrated by
-\cite{adrion2020community}.
-
-\subsection*{Instantaneous bottlenecks}
-
-A common approach to modelling the effect of demographic history on genealogies
-is to assume that effective population size ($N_e$) changes in discrete steps
-which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
-keightley2007joint,li2011inference}. In this setting of piecewise constant
-$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
-reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
-recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
-bottlenecks are short both on the timescale of coalescence and mutations,
-there may be little information about the duration of a bottleneck
-($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
-assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
-and generate a sudden burst of coalescence events (a multiple merger event) in
-the genealogy. The strength of the bottleneck $B$ can be thought of as an
-(imaginary) time period during which coalescence events are collapsed,
-i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
-single pair of lineages entering the bottleneck coalesce during the bottleneck
-is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
-attractive and both analytic results and empirical
-inference~\citep{griffiths1994sampling, birkner2009modified,galtier2000detecting,
-bunnefeld2015inferring} have been developed under this model, there has
-been no software available to simulate data under instantaneous
-bottleneck histories.
-
-We have implemented instantaneous bottlenecks in \msprime~1.0
-using a variant of Hudson's linear time single-locus coalescent
-algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
-by adding events to the \texttt{Demography} class
-(see the \nameref{sec-demography} section)
-and can be used in combination with any other demographic modelling
-features. We have validated the results of these simulations by comparing
-against analytical expectations for coalescence times and the
-site frequency spectrum~\citep{bunnefeld2015inferring}.
-
-\subsection*{Multiple merger coalescents}
-
-Kingman's coalescent assumes that only two ancestral lineages can merge at
-each merger event. Although this is generally a reasonable approximation, there
-are certain situations in which the underlying mathematical assumptions are
-violated. For example in certain highly fecund organisms
-\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
-to produce numbers of offspring on the order of the population size and
-therefore a few individuals may produce the bulk of the offspring in any given
-generation~\citep{hedgecock_94}. These population dynamics violate basic
-assumptions of the Kingman coalescent, and are better modelled by
-`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
-two lineages can merge in a given event. Multiple-merger coalescent processes
-have also been shown to be relevant for modelling the effects of selection on
-gene genealogies
-\citep{Gillespie909,DS04,desai2013genetic,neher2013genealogies,schweinsberg2017rigorous}.
-
-Although multiple merger coalescents have been of significant theoretical
-interest for around two decades, there has been little practical software
-available to simulate these models.
-\cite{kelleher2013coalescent,kelleher2014coalescent} developed packages to
-simulate a related spatial continuum model~\citep{barton2010new},
-\cite{zhu2015hybrid} simulate genealogies within a species tree
-based on a multiple-merger model, and
-\cite{becheler2020occupancy} provide a general method for simulating
-multiple merger processes
-as part of the Quetzal framework~\citep{becheler2019quetzal}.
-The \texttt{Beta-Xi-Sim} simulator~\citep{koskela2018multi,koskela2019robust}
-also includes a number of extensions to the Beta-coalescent.
-None of these methods work with large genomes, and very little work
-has been performed on simulating multiple merger processes with recombination.
-
-We have added two multiple merger coalescent models in \msprime\ 1.0, the
-Beta-coalescent~\citep{schweinsberg03} and
-``Dirac''-coalescent~\citep{BBE13}, allowing us to efficiently simulate
-such models with recombination for the first time.
-These simulation models have been extensively validated against
-analytical results from the site frequency
-spectrum~\citep{birkner2013statistical,blath2016site,hobolth2019phase}
- as well as more general properties of coalescent processes.
-See the Appendix for more details and model derivations.
-
-\subsection*{Ancestral Recombination Graphs}
-\label{sec-arg}
-
-The Ancestral Recombination Graph (ARG) was introduced by
-Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
-to represent the stochastic process of the coalescent with recombination as a graph.
-This formulation is complementary to Hudson's earlier
-work~\citep{hudson1983properties}, and substantially increased our theoretical
-understanding of recombination. In Griffiths' ARG formulation,
-a realisation of the coalescent with recombination is a graph in which
-vertices represent common ancestor or recombination events, and edges represent
-lineages. There is the ``big'' ARG, in which we track lineages arising out of
-recombinations regardless of whether they carry ancestral
-material~\citep{ethier1990two},
-% and
-% consequently wait an exponentially large time until all sampled lineages
-% coalesce into a single ancestor~\citep{ethier1990two},
-and the ``little'' ARG in which we only track genetic ancestors.
-Over time, usage of the term
-has shifted away from its original definition as a stochastic process,
-to being interpreted as a representation of a particular genetic ancestry
-as a graph, without necessarily following the specific details of the Griffiths
-formulation~\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
-Under the latter interpretation,
-the tree sequence encoding of genetic ancestry (described above)
-clearly \emph{is} an ARG: the nodes and edges
-define a graph in which edges are annotated with the
-set of disjoint genomic intervals
-through which ancestry flows.
-
-\begin{figure}
-\begin{subfigure}[t]{0.33\textwidth}
-\centering
-\caption*{(A)}
-    \begin{tikzpicture}[scale=0.45,nodestyle/.style={font=\scriptsize}]
-
-    \draw (-2,0) -- (-2,3) -- (0, 3);
-    \draw (1,0) -- (1,1);
-    \draw (0,1) -- (2,1) -- (2,2) -- (4,2) -- (4,0);
-    \draw (3,2) -- (3,5);
-    \draw (0,1) -- (0,3);
-    \draw (-1, 3) -- (-1, 5) -- (3,5);
-
-    \draw[fill] (-2,0) circle [radius=0.05];
-    \draw[fill] (-1,3) circle [radius=0.05];
-    \draw[fill] (1,0) circle [radius=0.05];
-    \draw[fill] (4,0) circle [radius=0.05];
-    \draw[fill] (1,1) circle [radius=0.05];
-    \draw[fill] (3,2) circle [radius=0.05];
-    \draw[fill] (1,5) circle [radius=0.05];
-
-    \node[nodestyle] [below] at (-2,0) {0};
-    \node[nodestyle] [below] at (1,0) {1};
-    \node[nodestyle] [below] at (4,0) {2};
-    \node[nodestyle] [above] at (1,1) {3};
-    \node[nodestyle] [below right] at (1,1) {$x = 0.3$};
-    \node[nodestyle] [above right] at (3,2) {4};
-    \node[nodestyle] [above left] at (-1,3) {5};
-    \node[nodestyle] [above] at (1,5) {6};
-
-    \end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.33\textwidth}
-\centering
-\caption*{(B)}
-\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
-
-\tikzset{greynode/.style={circle,fill,inner sep=0.7},
-nodelabel/.style={font=\scriptsize}}
-
-% Left sample nodes
-\node (s0) [greynode] {};
-\node (s1) [right=of s0,greynode] {};
-\node (s2) [right=of s1,greynode] {};
-\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-
-% Right sample nodes
-\node (rightTree) at (4, 0) {};
-\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
-\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
-\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
-\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-
-% Non-sample nodes
-\node [greynode] (s3) at ($(s1) + (0,1)$) {};
-\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
-\node [greynode] (s4) at ($(s2) + (0,2)$) {};
-\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
-\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
-\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
-\node [greynode] (s6) at ($0.25*(s0) + 0.25*(s1) + 0.5*(s2) + (0,5)$) {};
-\node [greynode] (s6r) at ($0.5*(s0r) + 0.25*(s1r) + 0.25*(s2r) + (0,5)$) {};
-\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
-\foreach \u/\lab in {s4/4, s4r/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/5, s5r/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s6/6, s6r/6} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\foreach \child/\parent in {
-	s5/s0, s5/s1, s6/s5, s6/s2,
-	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
-}
-	\draw (\child) -| (\parent);
-
-% Axes
-\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
-\node (topAx) at (0,-1.0) {};
-\node (topLeft) at ($(s0) + (0,-1.0)$) {};
-\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
-\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
-\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
-\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
-\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
-
-% Interval endpoints
-\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, 5.5);
-
-% Axis titles
-\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.33\textwidth}
-\centering
-\caption*{(C)}
-\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
-
-\tikzset{greynode/.style={circle,fill,inner sep=0.7},
-nodelabel/.style={font=\scriptsize}}
-
-% Left sample nodes
-\node (s0) [greynode] {};
-\node (s1) [right=of s0,greynode] {};
-\node (s2) [right=of s1,greynode] {};
-\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-
-% Right sample nodes
-\node (rightTree) at (4, 0) {};
-\node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
-\node [greynode] (s1r) at ($(s1) + (rightTree)$) {};
-\node [greynode] (s2r) at ($(s2) + (rightTree)$) {};
-\foreach \u/\lab in {s0r/0, s1r/1, s2r/2} \node[nodelabel,anchor=north] at (\u) {\lab};
-
-% Non-sample nodes
-%\node [greynode] (s3) at ($(s1) + (0,1)$) {};
-%\node [greynode] (s3r) at ($(s1r) + (0,1)$) {};
-%\node [greynode] (s4) at ($(s2) + (0,2)$) {};
-\node [greynode] (s4r) at ($0.5*(s1r) + 0.5*(s2r) + (0,2)$) {};
-\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
-%\node [greynode] (s5r) at ($(s0r) + (0,3)$) {};
-\node [greynode] (s6) at ($0.25*(s0) + 0.25*(s1) + 0.5*(s2) + (0,5)$) {};
-\node [greynode] (s6r) at ($0.5*(s0r) + 0.25*(s1r) + 0.25*(s2r) + (0,5)$) {};
-%\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
-\foreach \u/\lab in {s4r/3} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/4} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s6/5, s6r/5} \node[nodelabel,anchor=south] at (\u) {\lab};
-
-%% Edges
-\foreach \child/\parent in {
-	s5/s0, s5/s1, s6/s5, s6/s2,
-	s4r/s1r, s4r/s2r, s6r/s0r, s6r/s4r%
-}
-	\draw (\child) -| (\parent);
-
-% Axes
-\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
-\node (topAx) at (0,-1.0) {};
-\node (topLeft) at ($(s0) + (0,-1.0)$) {};
-\node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
-\foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
-\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
-\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
-\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
-
-% Interval endpoints
-\draw[thin,color=black!50,dashed] ($(topLeft) + 5*(genUnit)$) -- +(0, 5.5);
-
-% Axis titles
-\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
-\end{tikzpicture}
-\end{subfigure}
-\caption{\label{fig-arg} (A) A simple ARG in which a recombination
-occurs at position 0.3; (B) the equivalent topology depicted as a tree
-sequence, including the recombination node; (C) the same tree sequence
-topology ``simplified'' down to its minimal tree sequence representation.
-Note that the internal nodes have been renumbered in the simplified
-representation, so that, e.g., node 5 in (C) corresponds to node 6 in
-(A) and (B).}
-\end{figure}
-
-For our purposes, an ARG is a realisation of the coalescent with
-recombination, in the Griffiths (little ARG) sense.
-As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
-works by dynamically traversing a little ARG.
-The graph is not explicitly represented in memory, but is partially
-present through the extant lineages and the ancestral material they carry
-over time. We do not output the graph directly, but
-rather store the information required to recover the genealogical
-history as nodes and edges in a tree sequence.
-This is far more efficient than outputting the simulated ARG in its entirety.
-For a given scaled recombination rate $\rho$
-(setting aside the dependency on the sample size $n$)
-we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
-in an ARG is $O(\rho^2)$,
-whereas the size of the tree sequence encoding is
-$O(\rho)$~\citep{kelleher2016efficient}.
-This difference between a
-quadratic and a linear dependency on $\rho$ is profound, and shows why
-large simulations cannot output an ARG in practice.
-
-Although by default \msprime\ outputs tree sequences that contain full information
-about the genealogical trees, their correlation structure along the chromosome,
-and the ancestral genomes on which coalescences occurred, some information
-is lost in this mapping down from ARG space to the minimal tree sequence
-form. In particular, we lose
-information about ancestral genomes that were common ancestors but
-in which no coalescences occurred, and also information about the precise time
-and chromosomal location of recombination events. In most cases, such
-information is of little relevance as it is in principle unknowable,
-but there are occasions such as visualisation or computing likelihoods (see
-below) in which it is useful. We therefore provide
-the \texttt{record\_full\_arg} option in \msprime\
-to store a representation of the complete ARG traversed during simulation.
-This is done by storing extra nodes (marked with specific flags, so they
-can be easily identified later) and edges in the tree sequence
-(Fig.~\ref{fig-arg}).
-One situation in which a record of the full ARG is necessary is when we
-wish to compute likelihoods during inference.
-The likelihood is a central quantity in evaluating the plausibility of a putative
-ancestry as an explanation of DNA sequence data, both directly through
-e.g.~approaches based on maximum likelihood, and as an ingredient of
-methods such as
-Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
-We provide functions to compute the likelihood of ARG realisations
-and mutational patterns under
-the standard coalescent and infinite sites mutation model.
-See the Appendix for details on these likelihood calculations.
-
-\subsection*{Selective sweeps}
-\label{sec-selection}
-Another elaboration of the standard neutral coalescent with recombination
-is the addition of selective
-sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
-Sweeps are modelled by creating a structured population
-during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
-in which lineages may transit between favoured and unfavoured backgrounds through
-recombination. This approach allows for many selective sweep scenarios to be simulated
-efficiently, including recurrent, partial, and soft selective sweeps.
-However this efficiency comes at the cost of flexibility in comparison to forwards in
-time simulation.
-Several specialised simulators have been developed to simulate
-sweeps in the coalescent,
-including \texttt{SelSim}~\citep{spencer2004selsim},
-\texttt{mbs}~\citep{teshima2009mbs}, \msms~\citep{ewing2010msms},
-\cosiTwo~\citep{shlyakhter2014cosi2}
-and \discoal~\citep{kern2016discoal}.
-
-\begin{figure}
-\begin{center}
-\includegraphics{figures/sweeps-perf}
-\end{center}
-\caption{\label{fig-selection-perf}Comparison of selective sweep simulation
-performance in \msprime\ and \discoal\ (Intel Xeon E5-2680 CPU).
-We report the total CPU time and maximum memory usage when simulating 100
-replicates for 10 samples in a model
-with a single selective sweep in its history where the beneficial
-allele had a scaled selection coefficient of $2Ns=1000$,
-a per-base recombination rate of $10^{-9}$ and sequence
-length varying from 1kb--200kb.}
-\end{figure}
-
-Selective sweeps are implemented in the coalescent as a two
-step-process: first generating an allele frequency trajectory,
-and then simulating a structured coalescent process conditioned
-on that trajectory.
-Following \discoal, we generate sweep trajectories in \msprime\ using a
-jump process approximation to the conditional
-diffusion of an allele bound for
-fixation~\citep{coop2004ancestral}.
-The jump process moves back in time following the beneficial
-allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
-the origination of the allele at $p=1/(2N)$, tracking time in small
-increments $\delta t$. Then, given the frequency $p$ at time $t$,
-the frequency $p^\prime$ at time $t + \delta t$ is given by
-\begin{equation*}
-p^\prime =
-\begin{cases}
-    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
-    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
-\end{cases}
-\end{equation*}
-where
-\begin{eqnarray*}
-	\mu(p)=\frac{\alpha p (1 - p)}{\tanh(\alpha(1-p))}.
-\end{eqnarray*}
-Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
-can be generalised straightforwardly to include arbitrary dominance.
-We can also define trajectories to model neutral alleles and soft
-selective sweeps, which we plan as future additions to \msprime.
-
-Then, given a randomly generated allele frequency trajectory under
-the above model, the simulation of a sweep works by assigning lineages to
-two different structured coalescent ``labels'', based on
-whether they carry the beneficial allele.
-The allele frequency trajectory determines the relative sizes
-of the ``populations'' in these labels over time, and therefore
-the rates at which various events occur.
-Common ancestor events can then only merge lineages from \emph{within}
-a label, but lineages can transfer from one label to the other
-(i.e., from the advantageous to disadvantageous backgrounds,
-and vice versa) as a result of recombination events.
-Once we have reached the end of the simulated trajectory the sweep
-is complete, and we remove the structured coalescent labels.
-Simulation may then resume under any other ancestry model.
-
-Fig.~\ref{fig-selection-perf} compares the performance of \msprime\
-and \discoal\ under a simple sweep model, and shows
-that \msprime\ has far better
-CPU time and memory performance.
-Since our implementation uses the abstract label system mentioned above,
-adding support for similar situations,
-such as inversions~\citep{peischl2013sequential}, should be straightforward.
-
-\subsection*{Discrete time Wright-Fisher}
-The coalescent is an idealised model and makes many simplifying assumptions,
-but it is often surprisingly robust to violations of these
-assumptions~\citep{wakeley2012gene}. One situation in which the
-model does break down is the combination of large sample size
-and long recombining genomes, where the large
-number of recombination events in the recent past results in
-more than the biologically possible $2^t$ ancestors in
-$t$ diploid generations~\citep{nelson2020accounting}.
-This pathological behaviour results in
-identity-by-descent,
-long-range linkage disequilibrium and ancestry patterns deviating from
-Wright-Fisher expectations, and the bias grows with larger sample
-sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
-Precisely this problem occurs when simulating modern human datasets,
-and we have implemented a Discrete Time Wright-Fisher (DTWF) model
-in \msprime\ to address the issue. The DTWF simulates backwards in
-time generation-by-generation so that each gamete has a unique
-diploid parent, and multiple recombinations within a generation results in
-crossover events between the same two parental haploid copies.
-The method is described in more detail by~\cite{nelson2020accounting}.
-
-\begin{figure}
-\begin{center}
-\includegraphics{figures/dtwf-perf}
-\end{center}
-\caption{\label{fig-dtwf-perf} Comparison of Discrete Time Wright-Fisher
-(DTWF) simulation performance in \msprime\ and \ARGON\ (Intel Xeon E5-2680 CPU).
-We simulated ancestry for a sample of 1000 haploids from a population of 10000,
-and report the (A) total CPU time and (B) maximum memory usage for varying
-sequence lengths, and a per-base recombination rate of $10^{-8}$. Each
-point is the average over $5$ replicate simulations.
-We show observations for \ARGON, \msprime's DTWF implementation (``DTWF'')
-and a hybrid simulation of 100 generations of the DTWF followed by
-the standard coalescent with recombination model (``DTWF + Hudson'').
-Memory usage for \msprime's DTWF and hybrid simulations are very similar.
-We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
-to ensure we are measuring only ancestry simulation time.}
-\end{figure}
-
-Fig.~\ref{fig-dtwf-perf} shows that \msprime\ simulates the DTWF
-more quickly and requires substantially less memory than
-\ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
-However, the generation-by-generation approach of the DTWF is less
-efficient than the coalescent with recombination when the
-number of lineages is significantly less than the population size
-(the regime where the coalescent is an accurate approximation),
-which usually happens in the quite recent
-past~\citep{bhaskar2014distortion}.
-We therefore support changing the simulation model during a simulation
-so that we can run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}.
-Any number of different simulation models can be combined, allowing for the
-flexible choice of simulation scenarios.
-As the discrete time Wright-Fisher model improves accuracy of genealogical
-patterns in the recent past, we can simulate the recent history using this
-model and then switch to the standard coalescent to more efficiently simulate
-the more ancient history.
-
-\subsection*{Integration with forward simulators}
-A unique feature of \msprime\ is its ability to simulate
-genetic ancestries by extending an existing partial
-genetic ancestry. Given a tree sequence that
-is complete up until time $t$ ago as input
-(where marginal trees may or may not have fully coalesced),
-\msprime\ can efficiently obtain the segments of ancestral material
-present at this time, and then run the simulation backwards in time from there.
-This allows a simulated ancestry to be produced by any
-number of different processes across disjoint time slices.
-In practice this feature is used to ``complete''
-forwards-time ancestry simulations~\citep{kelleher2018efficient}
-that may have not fully coalesced. This process
-(``recapitation'') can be orders of magnitude faster than
-the standard approach of neutral burn-in; see
-\cite{haller2018tree} for more details and examples.
-This interoperability between simulators, where a partial ancestry
-simulation produced by \SLiM~\citep{haller2019slim}
-or \fwdpy~\citep{thornton2014cpp} can be picked up and completed
-by another simulator, with complete information
-retained---at scale---is unprecedented.
-There may be an opportunity for
-other forward genetic simulators~\citep[e.g.][]{gaynor2021alphasimr}
-to leverage the tree sequence data format and associated tools.
-
-\subsection*{Development model}
-\texttt{Msprime} has a large number of features, encompassing
-the functionality of several more specialised simulators
-while maintaining excellent performance.
-It is developed by a geographically distributed team of volunteers under an
-open source community development model, with a strong emphasis
-on code quality, correctness, good documentation, and inclusive development.
-As in any large code base,
-unit tests play a key role in ensuring that new additions behave
-as expected and \msprime\ has an extensive suite.
-As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
-lines of Python, with suites of 122 C tests (7K lines of code) and
-1350 Python tests (22K lines of code). These tests are run
-automatically on different operating systems on each pull request
-(where a contributor proposes a code change), using standard Continuous
-Integration (CI) methodology. Other CI services
-check for common errors, code formatting issues, and produce reports on
-the level of test coverage for the proposed change.
-
-Unit tests are vital for ensuring software quality and correctness, but
-they are usually of little value in assessing the statistical properties
-of simulations. To validate the correctness of simulation output
-we maintain a suite of statistical tests (as of 1.0.0,
-217 validation tests, in 6K lines of code). These consist of running many
-replicate simulations to check the properties of the output
-against other simulators, and where possible against analytical results.
-For example, simulations of complex demography are validated against \ms,
-selective sweeps against \discoal, and Wright-Fisher simulations
-against forwards in time simulations in \SLiM.
-This suite of tests is run before every release, to ensure
-that statistical errors have not been introduced.
-
-More visibly to the end user, we also have a high standard for documentation,
-with precise, comprehensive, and cross-linked documentation
-that is automatically built from the code base
-and served through the website \texttt{https://tskit.dev}.
-With the goal of lowering the entry barrier to new users,
-we have invested significant effort in writing examples and introductions,
-and making common tasks discoverable.
-We also view contributions to documentation as equally important to the project
-as writing code or designing methods: what use would it be
-to write reliable, stable software if no-one used it?
-
-An important goal of \msprime's development model is to maximise
-accessibility for prospective users and contributors, and
-to encourage diversity in our community.
-Gender and racial inequality caused by discrimination and marginalisation
-is a major problem across
-the sciences~\citep{wellenreuther2016women,shannon2019gender}
-and in open source software development~\citep{trinkenreich2021women}.
-Within our field, the contribution of women to early computational methods
-in population genetics was marginalised~\citep{dung2019illuminating},
-and women continue to be under-represented
-in computational biology~\citep{bonham2017women}.
-The authorship of our paper reflects these trends,
-with a skew towards men and affiliations in the USA
-and Europe. We know the importance of creating and strengthening networks to
-develop and maintain a diverse community of contributors, and we are committed
-to fostering a supportive and collaborative environment that helps to address these
-inequalities in our field.
-
-\section*{Discussion}
-
-% We can simulate so much more now
-The 1.0 release of \msprime\ marks a major increase in the
-breadth of available features and the
-potential biological realism of simulations.
-These abilities will allow researchers to perform more robust power analyses,
-more reliably test new methods,
-carry out more reliable inferences,
-and more thoroughly explore the properties of theoretical models.
-Despite this complexity and generality,
-\msprime's performance is state-of-the-art
-and all features are extensively tested and statistically validated.
-These advances have only been possible thanks to a distributed,
-collaborative model of software development,
-and the work of many people.
-% We hope the tool continues to be useful to many.
-
-% Simulation software is historically a beautiful shitshow
-
-Even though simulation has long been a vital tool in population genetics,
-such collaborative software development has historically been uncommon.
-A huge proliferation of tools have been published
-(the references here are not exhaustive)
-and only a small minority of these are actively developed and
-maintained today. The ecosystem is highly fragmented, with numerous different
-ways of specifying parameters and representing results, and there are
-significant software quality issues at all stages. This is unsurprising, since
-the majority of simulation software development is performed by students,
-often without formal training in software development.
-The result resembles Haldane's sieve for new mutations:
-many new pieces of software stay permanently on a dusty shelf of supplementary materials,
-while some of those that prove particularly useful when new (like dominant alleles)
-are quickly adopted.
-Although this has produced many good tools
-and enabled decades of research,
-it also represents a missed opportunity to invest as a community
-in shared infrastructure and mentorship in good software development practice.
-
-% THERE IS A BETTER WAY!!!
-Scientific software is vital apparatus, and must be engineered
-to a high quality if we are to trust its results. There is a growing
-realisation across the sciences
-% More citations?
-\citep[e.g.][]{siepel2019challenges,harris2020array,gardner2021sustained}
-that investing in
-shared community infrastructure produces better results than a
-proliferation of individually maintained tools, allowing scientists
-to focus on their specific questions rather than software engineering.
-\texttt{Msprime} 1.0 is the result of such a community process,
-with features added by motivated users, taking advantage of the
-established development practices and infrastructure.
-Software development in a welcoming community,
-with mentorship by experienced developers,
-is a useful experience for many users.
-The skills that contributors learn
-can lead to greatly increased productivity in subsequent
-work (e.g., through more reliable code and better debugging skills).
-We hope that users who find that
-features they require are missing will continue to contribute to
-\msprime, leading to a community project that is both high quality
-and sustainable in the long term.
-
-The succinct tree sequence data structure developed
-for \msprime\ provides a view of not only
-genetic variation, but also the genetic ancestry that produced that variation.
-Recent breakthroughs in methods to infer genetic ancestry in recombining
-organisms~\citep{rasmussen2014genome,kelleher2019inferring,
-speidel2019method,wohns2021unified,schaefer2021ancestral,speidel2021inferring}
-have made it possible to estimate such ancestry from real data at scale for
-the first time~\citep{harris2019database,tang2019genealogy}.
-Given such inferred ancestry, many exciting applications
-become possible. For example, \cite{osmond2021estimating}
-developed a method to estimate the location of genetic ancestors
-based on inferred trees, and other uses are sure to follow.
-Since the inferred genetic ancestry
-becomes the input for other downstream inferences, it is vitally
-important that these primary inferences are thoroughly validated,
-with the detailed properties of the inferred ancestries
-catalogued and understood.
-\texttt{Msprime} will continue to be an important tool for these inferences and validations,
-and in this context the ability to interoperate
-with other methods---particularly forwards simulators---through
-the succinct tree sequence data structure and \tskit\ library
-will be essential.
-% this final sentence is hard to parse
-
-\section*{Acknowledgements}
-We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
-on the manuscript.
-% alphabetised by author. Names spelled out if there's an initial collision
-ADK was supported by NIH awards R01GM117241 and R01HG010774.
-BE was supported by  DFG grant  273887127 through Priority Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
-FB is funded by the Deutsche Forschungsgemeinschaft EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
-Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
-Gregor Gorjanc is supported by the Chancellor's Fellowship of the University of Edinburgh and the BBSRC grant to The Roslin Institute BBS/E/D/30002275.
-Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
-Jerome Kelleher is supported by the Robertson Foundation.
-SG acknowledges funding from the Canada Research Chairs Program, from the Canadian Institutes of Health Research PJT 173300, and from the Canadian Foundation for Innovation.
-
-\bibliographystyle{plainnat}
-\bibliography{paper}
-
-%% local definitions for section multiple merger coalescents
- \newcommand{\be}{\begin{equation}}
- \newcommand{\ee}{\end{equation}}
- \newcommand{\bd}{\begin{displaymath}}
- \newcommand{\ed}{\end{displaymath}}
-\newcommand{\IN}{\ensuremath{\mathds{N}}}%
-\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
-\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
-\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%
-
-\NewEnviron{esplit}[1]{%
-\begin{equation}
-\label{#1}
-\begin{split}
-  \BODY
-\end{split}\end{equation}
-}
-
-\setcounter{secnumdepth}{2} % Print out appendix section numbers
-
-\appendix
-\section*{Appendix}
-
-\subsection*{Mutation generation}
-\label{app-mutation-algorithm}
-
-The algorithm that \msprime\ uses to simulate mutations on a tree sequence
-proceeds in two steps:
-first, mutations are ``placed'' on the tree sequence
-(i.e., sampling their locations in time, along the genome, and on the marginal tree),
-and then the ancestral and derived alleles of each mutation are generated.
-All mutation models share the code to place mutations,
-but choose alleles in different ways.
-
-First, mutations are placed on the tree sequence under an inhomogeneous Poisson model
-by applying them independently to each edge.
-If an edge spans a region $[a, b)$ of the genome
-and connected parent and child nodes with times $s < t$,
-and the mutation rate locally is $\mu$,
-then the number of mutations on the edge is Poisson with mean $\mu (t-s) (b-a)$,
-and each mutation is placed independently at a position chosen uniformly in $[a,b)$
-and a time uniformly in $[s,t)$.
-In a discrete genome, all positions are integers and so more than one mutation
-may occur at the same position on the same edge.
-Otherwise (i.e., for an infinite-sites model),
-positions are rejection sampled to obtain a unique floating-point number.
-If an edge spans a region of the genome with more than one mutation rate,
-this is done separately for each sub-region on which the mutation rate is constant.
-Since each edge is processed independently, the algorithm
-scales linearly with the number of edges in the tree sequence.
-
-Next, alleles are chosen for each mutation.
-If the site was not previously mutated, then a new ancestral allele is chosen for the site,
-according to an input distribution of ancestral state allele probabilities.
-Then, each mutation on the tree is considered in turn,
-and a derived allele is randomly chosen based on the parental allele
-(which may be the ancestral allele or the derived allele of a previous mutation).
-Finally, information about the mutations are recorded in the site and mutation tables
-of the tree sequence.
-
-A mutation model must, therefore, provide two things:
-a way of choosing an ancestral allele for each new variant site,
-and a way of choosing a derived allele given the parental allele at each mutation.
-Perhaps the simplest mutation model implemented in \msprime\ is the
-\texttt{InfiniteAlleles} mutation model,
-which keeps an internal counter so that the requested alleles are
-assigned subsequent (and therefore unique) integers.
-
-The distribution of ancestral alleles
-is used to choose the allele present at the root of the tree at each mutated site,
-i.e., the \texttt{root\_distribution}.
-Mutation models with a finite possible set of alleles have a natural choice
-for this distribution---the \emph{stationary distribution} of the mutation process.
-(All mutation models are Markovian,
-so this may be found as the top left eigenvector of the mutation matrix.)
-This is the default in most models,
-except, e.g., the \texttt{BinaryMutationModel}, whose alleles are 0 and 1
-and always labels the ancestral allele ``0''.
-However, mutational processes are not in general stationary,
-so we often allow a different root distribution to be specified.
-
-Since the general algorithm above applies mutations at a single rate
-independent of ancestral state,
-a model in which different alleles mutate at different rates
-must necessarily produce some silent mutations,
-i.e., mutations in which the derived allele is equal to the parental allele.
-To illustrate this, consider a mutation model
-in which $A$ or $T$ mutates to a randomly chosen different nucleotide at rate $\alpha$
-and $C$ or $G$ mutates at rate $\beta$, with $\beta < \alpha$.
-To implement this, first place mutations at the largest total rate, which is $\alpha$.
-Then, at each site, choose an ancestral allele from the root distribution,
-and for each mutation, choose a derived allele as follows:
-if the parental allele is $A$ or $T$, then choose a random derived allele
-different to the parental allele;
-if the parental allele is $C$ or $G$,
-then choose the derived allele to be equal to the parent allele
-with probability $\beta/(\alpha + \beta)$,
-and randomly choose a different nucleotide otherwise.
-This produces the correct distribution by Poisson thinning:
-a Poisson process with rate $\alpha$ in which each point is discarded independently
-with probability $\beta / (\alpha + \beta)$ is equivalent to a Poisson
-process with rate $\beta$.
-All finite-state models (implemented under the generic \texttt{MatrixMutationModel} class)
-work in this way: mutations are placed at the maximum mutation rate,
-and then some silent mutations will result.
-
-In previous versions of \msprime, silent mutations were disallowed,
-and we could have removed them from the output entirely.
-However, we have chosen to leave them in, so that for instance
-simulating with the \texttt{HKY} mutation model will result in silent mutations
-if not all equilibrium frequencies are the same.
-The presence of silent mutations may at first be surprising
-but there is a good reason to leave them in:
-to allow layering of different mutation models.
-Suppose that we wanted to model the mutation process as a mixture of more than one model,
-e.g., Jukes-Cantor mutations at rate $\mu_1$, and
-HKY mutations occur at rate $\mu_2$.
-Layering multiple calls to \texttt{sim\_mutations} is allowed,
-so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
-and then add more with the \texttt{HKY} model at rate $\mu_2$.
-However, there is a small statistical problem:
-suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
-but then the HKY mutations inserts another mutation in the middle,
-resulting in $A \to C \to C$.
-If neither mutation model allows silent transitions,
-then this is clearly not correct,
-i.e., it is not equivalent to a model that simultaneously applies the two models.
-(The impact is small, however, as it only affects sites with more than one mutation.)
-The solution is to make the Jukes-Cantor model \emph{state-independent}
-(also called ``parent-independent''),
-by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
-\emph{independently} of the parent (so that 1/4 of mutations will be silent).
-If so---and, more generally, if the first mutational process put down is
-state-independent---then the result of sequentially applying the two mutation models
-is equivalent to the simultaneous model.
-To facilitate this, many mutation models have a \texttt{state\_independent} option
-that increases the number of silent mutations
-and makes the model closer to state-independent.
-
-Silent mutations are fully supported by \tskit, which correctly accounts for
-their presence when computing statistics and performing other operations.
-For example, silent mutations have no effect on calculations of nucleotide
-site diversity.
-
 \subsection*{Likelihood calculations}
@@
 
+\subsection*{Multiple merger coalescent model}
 \label{app-multiple-mergers}
-\subsection*{Multiple merger coalescent model}
 

### 2021-11-14 response-to-reviewers.tex
Some infrastructure for responses

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the reviewers}
% General intro text goes here
We thank the reviewers for their close reading of our manuscript and
insightful comments. In the following we address the points raised
in turn.

% Let's start point-by-point with Reviewer 1
\reviewersection

% Point one description
\begin{point}
stuff
\end{point}

\begin{reply}
responds
\end{reply}

\reviewersection

\begin{point}
stuff
\end{point}

\begin{reply}
responds
\end{reply}

\end{document}

### 2021-11-14 reviewed-paper.tex
Infrastructure for latexdiff

--- 2021-09-20 paper.tex
@@
 
+\section*{Availability}
+\texttt{Msprime} is freely available under the terms of the
+GNU General Public License v3.0, and can be installed from the Python
+Package Index (PyPI) or the \texttt{conda-forge} conda channel.
+Development is conducted openly on GitHub at
+\texttt{https://github.com/tskit-dev/msprime/}.
+The documentation for \texttt{msprime} is available at
+\texttt{https://tskit.dev/msprime/docs/}.
+The source code for all the evaluations and figures in this manuscript
+is available at \texttt{https://github.com/tskit-dev/msprime-1.0-paper/}.
+
 \section*{Acknowledgements}

### 2021-11-14 response-to-reviewers.tex
Outline for response to reviewers

@@
 \begin{point}
-stuff
+Simulation is a vital part of population genetics, and msprime has quickly
+become a crucial tool for many researchers. msprime was first presented in a
+publication in 2016, where the main focus was the use of "sparse trees" and
+"coalescence records" to greatly increase the efficiency and practicality of
+large coalescent simulations. In this new manuscript, the authors introduce a
+new version of the software, with this manuscript ""focusing on the aspects
+that are either new for this version, or in which our approach differs
+significantly from classical methods."
+
+Since 2016, msprime has gained popularity not only due to its computational
+efficiency, but also for its software development style with well-documented
+features, development guidelines, and ever-growing capabilities. The extensive
+author list shows how the community around msprime and associated programs has
+grown over the last five years. The current manuscript has a huge number of
+citations (I count ~178!), generally representing a willingness to engage with
+a broad spectrum of the adjacent scientific community.
+
+The manuscript covers many features of the 1.0 version of msprime including at
+least (my list and organization):
+
+Software / API / innovations:
+Clear separations of ancestry and mutation simulations
+Improved methods for downstream analyses
+Integration with other software including other simulators
+Estimating run times of sim\_ancestry
+ARG recording
+Specific API for specifying demographic histories
+Software development model
+
+Implementations of external methods:
+Selective sweeps
+Instantaneous bottlenecks
+Multiple merger coalescents
+Discrete time Wright-Fisher
+Gene conversion
+Mutation models
+
+As well as benchmarks against other software implementations.
+
+The manuscript is well written and comprehensive - I do not find any major
+deficiencies. This manuscript handles a wide range of topics with a high degree
+of clarity. And I find the manuscript certainly reaches the criteria "... tools
+of interest to a wide range of geneticists" and so is a good fit as an
+investigation within the "Methods, Technology, \& Resources" subject.
+
+However, I do think it would be possible to increase the clarity and usefulness
+of the manuscript with some relatively small changes in organization.
 \end{point}
@@
 \begin{reply}
-responds
+Thanks Reviewer 1, TODO
+\end{reply}
+
+\begin{point}
+As, is, the manuscript lacks a simple overview of the new capabilities of
+version 1.0. Instead these are described in separate sections. The manuscript
+might benefit from a short summary and or table of the software's basic
+capabilities, API innovations, new functionality, etc. This may also provide a
+way to structure the larger flow of the paper, which currently seems to touch
+on the different topics in a somewhat random order.
+\end{point}
+
+\begin{reply}
+TODO Issue 207
+\end{reply}
+
+\begin{point}
+Generally, I found the figures and examples appropriate and well done. However,
+I think they could benefit from more cohesive presentations and descriptions. I
+will provide a number of examples - figure 3 and figure 4 present running times
+of various aspects of msprime, but present similar data in different ways. In
+Figure 3, the data points are not presented directly, while they are present on
+Figure 4. And based on the descriptions sometimes replicates are summed over
+(Figure 7) or averaged over (Figure 8?) to determine the values on the y-axis.
+Figure 3 clearly specifies sim\_mutations as the input for the benchmark, while
+for most other figures (e.g. 4) it frequently just says something like "running
+time for msprime." Sometimes only population-size scaled rates are provided in
+descriptions (e.g Figure 5 and 7) while in others both scaled and absolute
+sizes are provided. Sometimes "samples" are described to be diploid (figure 1)
+or sometime haploid (Figure 3), but other times the ploidy is left implied (e.g
+Figure 4). Sometimes the population size is referred to "effective population
+size" (e.g Figure 4) and other times just "population size" (Figure 3). More
+consistency in these simulations, descriptions, and figures could add clarity
+to the presentation. I did find the lack of a disk storage benchmark an
+interesting choice, as this aspect of the software is highlighted in the
+introduction but is not present in the manuscript. The algorithmic details of
+the simulations and methods are generally not included in the current
+manuscript (with some presented in the supplement, and some available in the
+2016 paper), and are left for citations. I think this decision serves the
+manuscript well otherwise it would likely become unwieldy.
+\end{point}
+
+\begin{reply}
+TODO Issue 202
+\end{reply}
+
+\begin{point}
+Figure comments
+Figure 2
+I think it could be made more clear that the genotype matrix is not part of the set of tables.
+\end{point}
+
+\begin{reply}
+We agree, and have revised Figure 2 and the caption to clarify the distinction between
+stored and derived data.
+\end{reply}
+
+\begin{point}
+Figure 4
+It is not clear if these two plots present all the data used for the quadratic
+fits, or if these data continue outside of the plots. Samples are described,
+but not their ploidy, are they haploids or diploid? From the plots here, it is
+not clear if the quadratic relationship holds for sample sizes << 1000. I
+suggest this figure should also include a smaller sample size, (eg n=10)
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+Figure 5
+I suggest including the absolute Ne and gc rates be included in the legend, in
+addition to the scaled rate.
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+Figure 7
+Why is an order of magnitude lower recombination rate used here than in other
+examples? This choice seems arbitrary and makes it difficult to relate these
+benchmarks results to any others in the paper. Samples are diploid or haploid?
+I would suggest the absolute Ne and s rates be included in the legend, in
+addition to the scaled rate.
+\end{point}
+
+\begin{reply}
+TODO
+
+The difference is out of necessity to run discoal (note the y scale).
+\end{reply}
+
+\begin{point}
+
+Figure 8
+Averaged points are referenced but not shown "Each point...". The plot could be
+more clear where the data points are and where the lines are interpolating
+between data points. Because of how the simulations were described and plotted,
+I found it awkward to try to compare the running time of the DTWF and
+coalescent models, even if this was not the purpose of the current figure. The
+figure says "to ensure we are measuring" I would suggest changing the text to
+"with a goal of measuring".
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+Individual line comments
+Line 142 - what is the "msp" program? Is this a command line version?
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+Line 370 - the text here makes it sound like Kelleher et al (2016) makes a
+statement relating eq. 1, but the Hein et al (2004) is not cited in the Keleher
+et al (2016).
+
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+Line 370 - population size(s)
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+
+Line 420 - Despite the long description, I was not able to follow how the
+previous paragraph implies that "work is spread out relatively evenly on the
+coalescent time scale"
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+
+Line 422 - There is no context to what is meant by "large" or "small" here - is
+2 or 10 or 10000 "large".
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+Line 606 - something goes wrong with the line numbering here.
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+
+Line 607 - The equations for the allele frequencies during sweeps seem slightly
+out of place as (I assume) these are not novel to msprime. There are many
+places where more details could be provided about methods discussed here, but
+in other cases the equations are left for the supplement or other papers.
+\end{point}
+
+\begin{reply}
+TODO
 \end{reply}
@@
 \begin{point}
-stuff
+Kelleher and colleagues present the 1.0 version of 'msprime', the leading
+coalescent simulator which is built upon a highly memory- and time-efficient
+data structure, introduced by the lead author, which supports exact simulation
+under the coalescent model with recombination.
+
+In addition to being highly efficient, the software engineering of msprime is
+of the highest quality. The program includes extensive unit and validation
+tests, and offers an API that simplifies worksflows and reduces the
+inefficiency and sources of error associated with large intermediate text
+files.
+
+Kelleher has made every effort to engage with the community, with the result
+that msprime now support many specialized features that have been contributed
+by other developers. This has the important benefit of reducing the chances
+that this software will lack support in future.
+
+In short, this is an admirable project, and a shining example of successful
+academic software development.
 \end{point}
@@
 \begin{reply}
-responds
+Thanks reviewer 2, TODO.
 \end{reply}
 
+\begin{point}
+
+The paper itself however is less polished. It is very long (37 pages / 1200
+lines without appendix), mostly because it is in places overly detailed, and
+seems to err on the side of completeness rather than conciseness or clarity.
+For instance, lines 160-177 describe the data structure in some detail, while
+this is published information. Line 192 remarks that "storage space is
+dramatically reduced", and goes on to make this more precise and point out some
+(fairly obvious) advantages in lines 193-202.
+As another example, line 301-4
+states that "Simulating mutations ... is efficient" and refers to Fig 3 for
+evidence. The rest of the section (304-319) details a number of examples shown
+in Fig 3 that do not add much to the story. As a third example, in the
+recombination section after useful comparisons with other approaches, line 354
+states that the proposed algorithm is still quadratic in the recombination
+rate. The long section from line 354 to 426 details the algorithmic reasons,
+and use various analytic approximations and simulations to support this. This
+material is of interest only to a few algorithm developers, and I would imagine
+it would be better placed in an appendix, with the key observations summarized
+in the main text.
+
+\end{point}
+
+\begin{reply}
+TODO.
+\end{reply}
+
+\begin{point}
+Another feature of the current paper is that a number of topics return several
+times with minor variations. This is particularly true for algorithmic
+efficiency, and the API, both of which are discussed multiple times. It is of
+course true that algorithmic efficiency is the major distinguishing feature of
+msprime; nevertheless, the focus of the current paper seems to be the rich
+feature set, and the software development model, as well as the various APIs
+that allow for a more integrated and less error prone workflow. I would suggest
+to introduce separate sections for algorithmic efficiency and the API, and
+discuss the various features (as you do now) in their own separate section, but
+focus on these features from a user's rather than a developer's perspective.
+
+\end{point}
+\begin{reply}
+TODO.
+\end{reply}
+
+\begin{point}
+Occasionally the paper reads more like documentation than a paper, such as when
+the Demography class is introduced and several associated functions are
+mentioned - this is not relevant for a paper in my opinion.
+
+\end{point}
+\begin{reply}
+TODO.
+\end{reply}
+
+\begin{point}
+The section about the development model is interesting - very few papers
+include such a section, and indeed I think this is one of the stand-out
+features of this project that it is so successfully supported by many
+contributors. My only suggestion here is to remove the listing of number of
+lines of code - this is a very questionable proxy for quality or quantity of
+work, and the work is impressive as it is.
+
+\end{point}
+\begin{reply}
+TODO.
+\end{reply}
+
+\reviewersection
+
+\begin{point}
+This manuscript introduces a simulation tool, msprime version 1.0, and
+describes features, additions, and performance of the simulator relative to the
+earlier msprime software. I will begin by saying that I believe all researchers
+in population genomics who use simulations have appreciated the quality work
+that has been previously done in the msprime sphere. The contributions proposed
+here clearly improve on a number of features, and that of course has value.
+\end{point}
+\begin{reply}
+TODO.
+\end{reply}
+
+\begin{point}
+However, I did not find this work to be appropriate for Genetics, and would
+rather view it as more appropriate for a journal like Bioinformatics. The
+fundamental reason is that this basically appears to perform old
+functionalities faster, and the added functionalities are modest. Again, both
+are valuable, but this work doesn't address much larger biological /
+statistical inference problems that are in need of development in this area,
+which could justify publication in a journal like Genetics.
+\end{point}
+\begin{reply}
+TODO.
+\end{reply}
+
+\begin{point}
+
+Specifically, the need of the field is really to simulate more realistic models
+in a coalescent framework, rather than the slower forward-in-time choice that
+stands as the only option presently. For example, the practical use of a tool
+that only simulates strictly neutral histories, or histories with the addition
+of positive selection, is likely rather limited empirically. By which I mean,
+the functional regions of a genome of course experience purifying selection ,
+but there is growing appreciation of the fact that this may result in
+background selection effects that may be widespread across the genomes of many
+species. In order for the strictly neutral demographic simulations to be of
+value, one must identify genomic regions that are not only neutral but are also
+unaffected by selection at linked sites. In a great many organisms, such
+regions may well not exist at all; in terms of commonly studied species, these
+regions may only exist in a handful of large, coding-sparse vertebrate and
+plant genomes (though, even if these regions exist, they are not necessarily
+easy to identify, or sufficiently large to perform neutral demographic
+estimation). I note that the manuscript avoids any mention or discussion of
+this problem, and avoids any citations that quantify this issue, but the
+neglect doesn't alleviate the problem. The authors should really consider,
+discuss, and utilize the following work carefully in this regard:
+
+Lohmueller et al. 2011. Natural selection affects multiple aspects of genetic
+variation at putatively neutral sites across the human genome. PLOS Genetics.
+
+Comeron. 2014. Background selection as baseline for nucleotide variation across
+the Drosophila genome. PLOS Genetics.
+
+Elyashiv et al. 2016. A genomic map of the effects of linked selection in
+Drosophila. PLOS Genetics.
+
+Comeron. 2017. Background selection as a null hypothesis in population
+genomics: insights and challenges from Drosophila studies. PTRSB.
+
+Pouyet et al. 2018. Background selection and biased gene conversion affect more
+than 95\% of the human genome and bias demographic inference. eLife.
+
+Torres et al. 2018. Human demographic history has amplified the effects of
+background selection across the genome. PLOS Genetics.
+
+Campos \& Charlesworth. 2019. The effects on neutral variability of recurrent
+selective sweeps and background selection. Genetics.
+
+Jensen et al. 2019. The importance of the neutral theory in 1968 and 50 years
+on: a response to Kern and Hahn 2018. Evolution.
+
+Castellano et al. 2020. Impact of mutation rate and selection at linked sites
+on DNA variation across the genomes of humans and other Homininae. GBE.
+
+Torres et al. 2020. The temporal dynamics of background selection in
+nonequilibrium populations. Genetics.
+
+Johri et al. 2021. The impact of purifying and background selection on the
+inference of population history: problems and prospects. MBE.
+
+Murphy et al. 2021. Broad-scale variation in human genetic diversity levels is
+predicted by purifying selection on coding and non-coding elements. bioRxiv.
+
+\end{point}
+\begin{reply}
+TODO.
+\end{reply}
+
+\begin{point}
+
+Relatedly, for the simulations of positive selection, while it has certainly
+been common practice in the field for a few decades, it is rather odd to model
+a single positively selected site on a background of all other neutral sites.
+Positive selection may act on mutations in functional regions, and functional
+regions will be largely shaped by purifying selection. Furthermore, the
+Hill-Robertson effects between the modeled positively selected mutation and the
+neglected array of negatively selected mutations is often non-trivial. In that
+sense, the positive selection model is probably only of empirical relevance in
+organisms with something approaching free recombination (to eliminate such
+linkage effects), as in perhaps HIV or the like. But of course, these sorts of
+organisms tend to have small, coding-dense genomes and experience very strong
+selective pressures, thus returning to the first problem mentioned above that
+regions free of background selection effects may not exist in the first place.
+\end{point}
+\begin{reply}
+TODO.
+\end{reply}
+
+\begin{point}
+
+In summary, I believe this to be a nice software addition that belongs as a
+Note in a bioinformatics journal. The software solution that would represent an
+important biological advance for Genetics would involve simulating more
+realistic models as discussed above (as forward simulators are currently
+capable of doing), but at a fast enough rate to be applicable to large-scale
+simulation-based genomic inference.
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
 \end{document}

### 2021-11-15 reviewed-paper.tex
answered comments about the number of lines of code

@@
 Fig.~\ref{fig-ancestry-perf} shows the observed run time for
-simulations with a variety of population size, chromosome
+simulations with a variety of population sizes, chromosome
 length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
@@
 as expected and \msprime\ has an extensive suite.
-As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
-lines of Python, with suites of 122 C tests (7K lines of code) and
-1350 Python tests (22K lines of code). These tests are run
-automatically on different operating systems on each pull request
+These tests are run automatically on different operating systems on each pull request
 (where a contributor proposes a code change), using standard Continuous
@@
 we maintain a suite of statistical tests (as of 1.0.0,
-217 validation tests, in 6K lines of code). These consist of running many
+217 validation tests). These consist of running many
 replicate simulations to check the properties of the output

### 2021-11-17 response-to-reviewers.tex
removed number of lines of code and corrected one typo
made changes to address issue: 
https://github.com/tskit-dev/msprime-1.0-paper/issues/205
Respond to GC figure point.
Closes #213

@@
 \begin{reply}
-TODO
+We avoided included the absolute values here in order to make the comparison
+between the simulators as simple as possible (since SimBac and FastSimBac
+required scaled inputs). Since the values we use aren't precisely the estimates
+for E.~coli (we rounded some values for simplicity), it would probably
+be misleading to include some Ne and gc rates that are consistent with the
+scaled value we used, but not equal to the actual E.~coli estimates, and
+we would therefore prefer not to include them.
 \end{reply}
@@
 \begin{reply}
-TODO
+Thank you, we have corrected this sentence in the revised manuscript.
 \end{reply}
@@
 \begin{reply}
-TODO.
+Thank you for your suggestion, we have removed the number of lines of code in the revised manuscript.
 \end{reply}

### 2021-11-17 response-to-reviewers.tex
Update figures and DTWF caption

@@
 between the simulators as simple as possible (since SimBac and FastSimBac
-required scaled inputs). Since the values we use aren't precisely the estimates
+require scaled inputs). Since the values we use aren't precisely the estimates
 for E.~coli (we rounded some values for simplicity), it would probably
@@
 \begin{reply}
-TODO
-
-The difference is out of necessity to run discoal (note the y scale).
+TODO Issue 214
 \end{reply}
@@
 \begin{point}
-
 Figure 8
@@
 \begin{reply}
-TODO
+We have update this figure (and all others in the main text) to include
+data markers. We have also rephrased the caption to try to clarify the
+parameters simulated and make them consistent with other figures.
 \end{reply}

### 2021-11-17 paper.tex
Outline for response to reviewers
Update figures and DTWF caption

@@
 \newcommand{\adkcomment}[1]{\textcolor{blue}{#1}}
+
+%% local definitions for section multiple merger coalescents
+\newcommand{\be}{\begin{equation}}
+\newcommand{\ee}{\end{equation}}
+\newcommand{\bd}{\begin{displaymath}}
+\newcommand{\ed}{\end{displaymath}}
+\newcommand{\IN}{\ensuremath{\mathds{N}}}%
+\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
+\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
+\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%
+
+\NewEnviron{esplit}[1]{%
+\begin{equation}
+\label{#1}
+\begin{split}
+  \BODY
+\end{split}\end{equation}
+}
 
@@
 The site and mutation tables together describe sequence variation
-% JK: also amongst the ancestral genomes, but only in positions where
-% they inherit from the sample. I thought it would be less confusing
-% to leave this out?
-among the samples.
+among the samples. The genotype matrix and tree topologies shown
+on the left are derived from these tables.
 }
@@
 performance in \msprime\ and \discoal\ (Intel Xeon E5-2680 CPU).
-We report the total CPU time and maximum memory usage when simulating 100
-replicates for 10 samples in a model
+We report the average CPU time and maximum memory usage when simulating 3
+replicates for 200 samples in a model
 with a single selective sweep in its history where the beneficial
-allele had a scaled selection coefficient of $2Ns=1000$,
-a per-base recombination rate of $10^{-9}$ and sequence
-length varying from 1kb--200kb.}
+allele had a selection coefficient of $s=0.05$,
+a per-base recombination rate of $10^{-8}$, population size of $N=10^4$,
+and sequence length varying from 100kb--3000kb.}
 \end{figure}
@@
 (DTWF) simulation performance in \msprime\ and \ARGON\ (Intel Xeon E5-2680 CPU).
-We simulated ancestry for a sample of 1000 haploids from a population of 10000,
-and report the (A) total CPU time and (B) maximum memory usage for varying
-sequence lengths, and a per-base recombination rate of $10^{-8}$. Each
-point is the average over $5$ replicate simulations.
+We ran simulations with a population size of $10^4$ and
+recombination rate of $10^{-8}$, with 500 diploid samples
+and varying sequence length.
+We report (A) total CPU time and (B) maximum memory usage;
+each point is the average over $5$ replicate simulations.
 We show observations for \ARGON, \msprime's DTWF implementation (``DTWF'')
 and a hybrid simulation of 100 generations of the DTWF followed by
-the standard coalescent with recombination model (``DTWF + Hudson'').
+the standard coalescent with recombination (``DTWF + Hudson'').
+We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
+with a goal of measuring only ancestry simulation time.
 Memory usage for \msprime's DTWF and hybrid simulations are very similar.
-We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
-to ensure we are measuring only ancestry simulation time.}
+}
 \end{figure}
@@
 \bibliography{paper}
-
-%% local definitions for section multiple merger coalescents
- \newcommand{\be}{\begin{equation}}
- \newcommand{\ee}{\end{equation}}
- \newcommand{\bd}{\begin{displaymath}}
- \newcommand{\ed}{\end{displaymath}}
-\newcommand{\IN}{\ensuremath{\mathds{N}}}%
-\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
-\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
-\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%
-
-\NewEnviron{esplit}[1]{%
-\begin{equation}
-\label{#1}
-\begin{split}
-  \BODY
-\end{split}\end{equation}
-}
 

### 2021-11-17 evaluation/plot.py

@@ -37,17 +37,18 @@ def mutations_perf():
     fig, (ax1, ax2) = two_panel_fig(sharey=True)
     ax1.set_xlabel("Sample size")
     dfL = df[df.L == 10]
-    for rate in rates:
+    markers= ["+", "x", "1"]
+    for marker, rate in zip(markers, rates):
         dfr = dfL[dfL.rate == rate].sort_values("n")
-        ax1.plot(dfr.n, dfr.time, label=f"Mutation rate={rate}")
+        ax1.plot(dfr.n, dfr.time, marker=marker, label=f"Mutation rate={rate}")
 
     ax2.set_xlabel("Sequence length (Megabases)")
     ax1.set_ylabel("Time (seconds)")
 
     dfn = df[df.n == 1000].sort_values("L")
-    for rate in rates:
+    for marker, rate in zip(markers, rates):
         dfr = dfn[dfn.rate == rate]
-        ax2.plot(dfr.L, dfr.time, label=f"Mutation rate={rate}")
+        ax2.plot(dfr.L, dfr.time, marker=marker, label=f"Mutation rate={rate}")
     print(dfn[dfn.L == 100])
 
     ax1.legend()
@@ -68,10 +69,18 @@ def gc_perf():
 
     lines = {}
     fig, (ax1, ax2) = two_panel_fig()
+    marker_map = {
+        "msprime": ".",
+        "SimBac": "+",
+        "fastSimBac": "x",
+    }
+
     for tool in ["msprime", "SimBac", "fastSimBac"]:
         dft = df[df.tool == tool]
-        (line,) = ax1.plot(dft.sample_size, dft.user_time, label=tool)
-        ax2.plot(dft.sample_size, dft.memory, label=tool, color=line.get_color())
+        (line,) = ax1.plot(dft.sample_size, dft.user_time,
+                marker=marker_map[tool], label=tool)
+        ax2.plot(dft.sample_size, dft.memory, label=tool,
+                marker=marker_map[tool], color=line.get_color())
         lines[tool] = line
 
     ax1.set_xlabel("Sample size")
@@ -83,7 +92,7 @@ def gc_perf():
     dfmsp = df[df["tool"] == "msprime"]
     largest_n = np.array(dfmsp.sample_size)[-1]
     largest_value = np.array(dfmsp.user_time)[-1]
-    ax1.plot([largest_n], [largest_value], "o", color=lines["msprime"].get_color())
[…]

### 2021-11-17 response-to-reviewers.tex
Tidy up figures.
Closes #202

@@
 consistency in these simulations, descriptions, and figures could add clarity
-to the presentation. I did find the lack of a disk storage benchmark an
-interesting choice, as this aspect of the software is highlighted in the
-introduction but is not present in the manuscript. The algorithmic details of
-the simulations and methods are generally not included in the current
-manuscript (with some presented in the supplement, and some available in the
-2016 paper), and are left for citations. I think this decision serves the
-manuscript well otherwise it would likely become unwieldy.
+to the presentation.
 \end{point}
@@
 \begin{reply}
-TODO Issue 202
+Thank you for examining the figures in this detail and providing such
+helpful feedback. We have tried to improve the presentation by taking
+the following steps for the main text figures:
+
+\begin{itemize}
+
+\item All plots now contain data markers to help show where the data
+points are and to distinguish the lines from each other.
+
+\item All plots state they are running ``sim\_ancestry`` or ``sim\_mutations``.
+
+\item Stated explicitly whether samples are haploid or diploid in the
+figure captions, and updated axis labels also.
+
+\item TODO discoal figure showing total time over 100 reps. (Issue 214)
+
+\end{itemize}
+
+In addition figure 4 has been moved to the appendix, making the remaining
+figures much more uniform in appearance and purpose (it is a special
+case because it is showing how well our theoretical predictions map
+to observations, as well as those observations).
 \end{reply}
@@
 \begin{point}
-Figure comments
+I did find the lack of a disk storage benchmark an
+interesting choice, as this aspect of the software is highlighted in the
+introduction but is not present in the manuscript.
+\end{point}
+
+\begin{reply}
+We did not include any benchmarks of storage space as this was explored
+in detail in the 2019 Nature Genetics paper, and we directly reference
+this paper when discussing the storage efficiency of tree sequences.
+\end{reply}
+
+\begin{point} Figure comments
 Figure 2

### 2021-11-17 paper.tex
Tidy up figures.
Closes #202

@@
 \caption{\label{fig-gc-perf}Comparison of simulation performance
-using \msprime, \SimBac, and \FastSimBac\ for varying sample sizes,
+using \msprime\ (\texttt{sim\_ancestry}), \SimBac, and \FastSimBac\ for varying
+(haploid) sample sizes,
 with parameters roughly equivalent to the current estimate
@@
 We report the average CPU time and maximum memory usage when simulating 3
-replicates for 200 samples in a model
+replicates for 200 haploid samples in a model
 with a single selective sweep in its history where the beneficial
@@
 \caption{\label{fig-dtwf-perf} Comparison of Discrete Time Wright-Fisher
-(DTWF) simulation performance in \msprime\ and \ARGON\ (Intel Xeon E5-2680 CPU).
+(DTWF) simulation performance in \msprime\ (\texttt{sim\_ancestry})
+and \ARGON\ (Intel Xeon E5-2680 CPU).
 We ran simulations with a population size of $10^4$ and
@@
 \caption{\label{fig-ancestry-perf}
-    Running time for \msprime\
+    Running time of \texttt{sim\_ancestry}
     for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
@@
     for various values of effective population size ($N_e$),
-    chromosome length in Morgans ($L$), and number of samples ($n$).
+    chromosome length in Morgans ($L$), and number of diploid samples ($n$).
     Run time scales quadratically with the product of $N_e$ and $L$,

### 2021-11-17 evaluation/plot.py

@@ -35,7 +35,7 @@ def mutations_perf():
     rates = np.unique(df.rate)
 
     fig, (ax1, ax2) = two_panel_fig(sharey=True)
-    ax1.set_xlabel("Sample size")
+    ax1.set_xlabel("Sample size (haploid)")
     dfL = df[df.L == 10]
     markers= ["+", "x", "1"]
     for marker, rate in zip(markers, rates):
@@ -83,9 +83,9 @@ def gc_perf():
                 marker=marker_map[tool], color=line.get_color())
         lines[tool] = line
 
-    ax1.set_xlabel("Sample size")
+    ax1.set_xlabel("Sample size (haploid)")
     ax1.set_ylabel("Time (hours)")
-    ax2.set_xlabel("Sample size")
+    ax2.set_xlabel("Sample size (haploid)")
     ax2.set_ylabel("Memory (GiB)")
     ax1.legend()
 

### 2021-11-17 response-to-reviewers.tex
Response to points about performance figure

@@
 \begin{reply}
-TODO
+We have moved Figure 4 to the appendix, as it is of more specialised interest
+than the other main text figures. We have truncated the x-axis on the
+quadratic fits to just the available data, to clarify that we are showing
+all of the data. We have clarified that samples are diploid.
+
+The plot is already quite complex and visually busy, and it is difficult to
+see how we could add a third line without major changes. The figure is
+primarily intended as a useful yardstick for users who wish to run simulations
+and would like a rough idea of how long it should take, and therefore adding
+a third line for very small sample sizes is likely to make it less useful
+for this purpose.
 \end{reply}

### 2021-11-17 evaluation/plot.py

@@ -288,7 +288,8 @@ def ancestry_perf():
                         rho[utN] / 4, df["time"][utN], color=rgb[k], marker=m, **sargs
                     )
 
-            xx = np.linspace(0, 1.05 * max(X[:, 0]), 51)
+            # xx = np.linspace(0, 1.05 * max(X[:, 0]), 51)
+            xx = np.linspace(0, max(X[:, 0]), 51)
 
             # Note the two quadratic curves are not the same!
             pargs = {}

### 2021-11-17 paper.tex
Respond to Hein et al point

@@
 supported and tested, and will be maintained for the foreseeable future.
-The \texttt{msp} program has been extended to include new commands
+The \texttt{msp} program (a command line interface to the library)
+has been extended to include new commands
 for simulating ancestry and mutations separately. A particularly useful
@@
 length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
-correctly predicts the quadratic dependence on $\rho$,
-as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
+correctly predicts the quadratic dependence on $\rho$
+\citep[see also][Fig.~2]{kelleher2016efficient}.
 We also see that the dependence on $n$ is quite weak,

### 2021-11-17 response-to-reviewers.tex
Respond to Hein et al point
Update figures

@@
 
-The plot is already quite complex and visually busy, and it is difficult to
+The plot is already quite complex and it is difficult to
 see how we could add a third line without major changes. The figure is
@@
 \begin{reply}
-TODO
+We have clarified this point with a parenthetical comment.
 \end{reply}
@@
 et al (2016).
-
 \end{point}
@@
 \begin{reply}
-TODO
+We have rephrased this to say "(see also Kelleher et al., 2016, Fig. 2)".
 \end{reply}

### 2021-11-17 evaluation/generate_sweeps_perf_data.py
revising sweep figure

@@ -149,18 +149,18 @@ def benchmark(replicates, processes):
         for k, v in cpu.items():
             print(k, "\t", v, file=f)
 
-    r = 1e-9
+    r = 1e-8
     # Set theta to 0 to avoid conflating mutation generation
     theta = 0
-    refsize = 1e6
-    # 1kb up to 200kb
-    Ls = np.linspace(1_000, 200_000, 20).astype(int)
+    refsize = 1e4
+    # 1kb up to 500kb
+    Ls = np.linspace(1_000, 500_000, 20).astype(int)
     work = []
     for L in Ls:
         for name in tool_map.keys():
             rho = 4 * refsize * (L - 1) * r
             cmd = (
-                f"10 {replicates} {L} -t {theta} -r {rho} -ws 0 -a 1000 -x 0.5 -N 10000"
+                f"10 {replicates} {L} -t {theta} -r {rho} -ws 0 -a 1000 -x 0.5 -N {refsize}"
             )
             work.extend([(name, L, cmd)])
 
@@ -171,6 +171,8 @@ def benchmark(replicates, processes):
             data.append(future.result())
             print(data[-1])
             df = pd.DataFrame(data)
+            df['user_time'] = df.user_time / replicates
+            df['sys_time'] = df.sys_time / replicates
             df.to_csv("data/sweeps_perf.csv")
 
 

### 2021-11-17 evaluation/generate_sweeps_perf_data.py
messing with figure params

@@ -97,7 +97,6 @@ def run_msprime(args):
     for _ in replicates:
         pass
 
-
 def get_process_resources():
     utime = 0
     stime = 0
@@ -137,8 +136,8 @@ tool_map = {
 
 
 @click.command()
-@click.option("--replicates", type=int, default=100)
-@click.option("--processes", type=int, default=None)
+@click.option("--replicates", type=int, default=3)
+@click.option("--processes", type=int, default=1)
 def benchmark(replicates, processes):
     """
     Runs the benchmark between msprime and discoal
@@ -153,14 +152,14 @@ def benchmark(replicates, processes):
     # Set theta to 0 to avoid conflating mutation generation
     theta = 0
     refsize = 1e4
-    # 1kb up to 500kb
-    Ls = np.linspace(1_000, 500_000, 20).astype(int)
+    # 100kb up to 3000kb
+    Ls = np.linspace(100_000, 3_000_000, 20).astype(int)
     work = []
     for L in Ls:
         for name in tool_map.keys():
             rho = 4 * refsize * (L - 1) * r
             cmd = (
-                f"10 {replicates} {L} -t {theta} -r {rho} -ws 0 -a 1000 -x 0.5 -N {refsize}"
+                f"200 {replicates} {L} -t {theta} -r {rho} -ws 0 -a 50 -x 0.5 -N {refsize}"
             )
             work.extend([(name, L, cmd)])
 

### 2021-11-18 paper.tex

--- 2021-11-17 paper.tex
@@
 (DTWF) simulation performance in \msprime\ and \ARGON\ (Intel Xeon E5-2680 CPU).
-We ran simulations with a population size of $10^4$ and
-recombination rate of $10^{-8}$, with 500 diploid samples
-and varying sequence length.
-We report (A) total CPU time and (B) maximum memory usage;
-each point is the average over $5$ replicate simulations.
+We simulated ancestry for a sample of 1000 haploids from a population of 10000,
+and report the (A) total CPU time and (B) maximum memory usage for varying
+sequence lengths, and a per-base recombination rate of $10^{-8}$. Each
+point is the average over $5$ replicate simulations.
 We show observations for \ARGON, \msprime's DTWF implementation (``DTWF'')
 and a hybrid simulation of 100 generations of the DTWF followed by
-the standard coalescent with recombination (``DTWF + Hudson'').
+the standard coalescent with recombination model (``DTWF + Hudson'').
+Memory usage for \msprime's DTWF and hybrid simulations are very similar.
 We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
-with a goal of measuring only ancestry simulation time.
-Memory usage for \msprime's DTWF and hybrid simulations are very similar.
-}
+to ensure we are measuring only ancestry simulation time.}
 \end{figure}

### 2021-11-18 response-to-reviewers.tex
Tweak sweeps plot and caption

@@
 
-\item TODO discoal figure showing total time over 100 reps. (Issue 214)
+\item All figures now show the time averaged across replicates.
 
@@
 \begin{reply}
-TODO Issue 214
+We have redone this figure to use the same recombination rate and population
+sizes as the rest of the paper, clarified that samples are diploid, given
+results in terms of the average per replicate and generally made the plot
+as close to the others as possible.
 \end{reply}

### 2021-11-18 evaluation/plot.py

@@ -115,7 +115,7 @@ def sweeps_perf():
     print(df)
 
     df.memory /= 1024 ** 3
-    df.L /= 1000
+    df.L /= 1000**2
     # discoal has very high systime usage, so we need to
     # include it here as well.
     df["time"] = (df.user_time + df.sys_time) / 60
@@ -133,9 +133,9 @@ def sweeps_perf():
         lines[tool] = line
         ax2.plot(dft.L, dft.memory, marker=marker_map[tool], label=tool)
 
-    ax1.set_xlabel("Sequence length (Kilobases)")
+    ax1.set_xlabel("Sequence length (Megabases)")
     ax1.set_ylabel("Time (minutes)")
-    ax2.set_xlabel("Sequence length (Kilobases)")
+    ax2.set_xlabel("Sequence length (Megabases)")
     ax2.set_ylabel("Memory (GiB)")
     ax1.legend()
 
@@ -144,7 +144,7 @@ def sweeps_perf():
     largest_value = np.array(dfmsp.time)[-1]
     ax1.plot([largest_L], [largest_value], "*", color="black")
     ax1.annotate(
-        f"{round(largest_value * 60)} seconds",
+        f"{largest_value * 60:.2f} seconds",
         textcoords="offset points",
         xytext=(-38, 5),
         xy=(largest_L, largest_value),

### 2021-11-18 paper.tex

--- 2021-11-17 paper.tex
@@
 \caption{\label{fig-selection-perf}Comparison of selective sweep simulation
-performance in \msprime\ and \discoal\ (Intel Xeon E5-2680 CPU).
+performance in \msprime\ (\texttt{sim\_ancestry}) and \discoal\ (Intel Xeon Gold 6148 CPU).
 We report the average CPU time and maximum memory usage when simulating 3
-replicates for 200 haploid samples in a model
-with a single selective sweep in its history where the beneficial
+replicates for 100 diploid samples in a model
+with a single selective sweep in its history, where the beneficial
 allele had a selection coefficient of $s=0.05$,

### 2021-11-18 response-to-reviewers.tex
Move sweeps equations to appendix # Please enter the commit message for your changes. Lines starting

@@
 but not their ploidy, are they haploids or diploid? From the plots here, it is
-not clear if the quadratic relationship holds for sample sizes << 1000. I
+not clear if the quadratic relationship holds for sample sizes $\ll$ 1000. I
 suggest this figure should also include a smaller sample size, (eg n=10)
@@
 \begin{point}
-
 Line 607 - The equations for the allele frequencies during sweeps seem slightly
@@
 \begin{reply}
-TODO
+We have moved the sweep trajectory equations to an appendix.
 \end{reply}

### 2021-11-18 response-to-reviewers.tex
Add outline response to editor

@@
 \begin{document}
+
+\section*{Response to the editor}
+% General intro text goes here
+TODO thanks editor
+
+% \reviewersection
+
+\textit{
+It is most important that you address the following in your resubmission:
+Improve the structure and focus of the main body of the manuscript. This should
+include moving a significant portion of the material to appendices or
+supplemental material, and adding an up-front summary of the main new features
+of msprime (relative to the 2016 paper).}
+
+TODO
+
+We have moved  Hudson's alg and selective sweeps details to appendixes.
 

### 2021-11-18 evaluation/generate_gc_perf_data.py
Update GC values with actual estimates

@@ -19,6 +19,7 @@ def run_simbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=False):
     R = gc_rate * 2
     # Set theta to 0 to remove mutations (defaults to 0.01)
     args = f"-N {sample_size} -B {int(L)} -R {R} -D {gc_tract_length} -T 0"
+    print(args)
     if count_trees:
         # This is cleaned up automatically when the variable goes out of scope
         tempdir = tempfile.TemporaryDirectory()
@@ -53,7 +54,7 @@ def run_fastsimbac(
     if set_seed > 0:
         args += " -s " + str(set_seed)
 
-    # print("running", args)
+    print("running", args)
     output = subprocess.run(
         "./tools/fastSimBac " + args, shell=True, check=True, capture_output=True
     )
@@ -108,9 +109,16 @@ def get_process_resources():
 
 
 def run_benchmark_process(tool, sample_size, queue):
-    L = 4_500_000
-    gc_rate = 0.015
-    gc_tract_length = 500
+    # Approximate values we used previously
+    # L = 4_500_000
+    # gc_rate = 0.015
+    # gc_tract_length = 500
+    L = 4641652
+    N = 180000000.0
+    g = 8.9e-11
+    gc_tract_length = 542
+    # exact values used by Lapierre et al, with scaled rate = 0.01602
+    gc_rate = N * g
     func = tool_map[tool]
     func(
         sample_size=sample_size,

### 2021-11-18 response-to-reviewers.tex
Respond to some reviewer 2 points.

@@
 \begin{point}
-
 The paper itself however is less polished. It is very long (37 pages / 1200
@@
 seems to err on the side of completeness rather than conciseness or clarity.
+\end{point}
+
+\begin{reply}
+Thank you for this criticism --- we have reduced the length of the manuscript
+by moving some material to appendices, and tried to address specific points
+raised below.
+\end{reply}
+
+\begin{point}
 For instance, lines 160-177 describe the data structure in some detail, while
@@
 (fairly obvious) advantages in lines 193-202.
+\end{point}
+
+\begin{reply}
+It is true that the details of the succinct tree sequence have been laid out
+in earlier publications, and it is reassuring that the advantages are
+obvious to you. However, for many of the intended audience these methods will
+be new, and we feel it's important that this paper is a self-contained,
+thorough, and convincing argument for the advantages of tree sequences and the
+tskit library for performing simulations.
+Very many papers are still published using ms and other
+simulators that use inefficient data formats, and it is these users that
+we want to convince.
+\end{reply}
+
+\begin{point}
 As another example, line 301-4
@@
 evidence. The rest of the section (304-319) details a number of examples shown
-in Fig 3 that do not add much to the story. As a third example, in the
+in Fig 3 that do not add much to the story.
+\end{point}
+
+\begin{reply}
+TODO
+\end{reply}
+
+\begin{point}
+As a third example, in the
 recombination section after useful comparisons with other approaches, line 354
@@
 in the main text.
-
 \end{point}
@@
 \begin{reply}
-TODO.
+We have moved this detailed discussion to an appendix.
 \end{reply}
@@
 focus on these features from a user's rather than a developer's perspective.
-
 \end{point}
@@
 mentioned - this is not relevant for a paper in my opinion.
-
 \end{point}

### 2021-11-18 paper.tex
Draft of new features table

@@
 \usetikzlibrary{calc,positioning}
+\usepackage{booktabs}
 %\usepackage{showkeys}
@@
 
+\begin{table}
+\begin{tabular}{lp{12cm}}
+Interface
+& Separation of ancestry and mutation simulations. Integration
+with forward simulators such as \SLiM\ and \fwdpy.
+Rich suite of analytical and visualisation methods via \tskit\ library. \\
+\midrule
+
+Ancestry
+& SMC, SMC', Beta- and Dirac-coalescent, discrete time Wright-Fisher,
+  and selective sweep models. Instantaneous bottlenecks.
+  Discrete or continuous coordinates, arbitary ploidy,
+  gene conversion.
+  Recording full ARG, full migration history, census
+  events.  More efficient for large numbers of populations.\\
+
+Demography
+& Improved inferface, import from Newick species tree and
+*BEAST~\citep{heled2009bayesian},
+Demes support~\citep{gower2022demes}, numerical methods to compute
+coalescence rates.\\
+
+Mutations
+& JC69, HKY, F84, GTR, BLOSUM62, PAM, infinite alleles, SLiM and
+general matrix mutation models.
+Varying rates along the genome, recurrent/back mutations,
+discrete or continuous coordinates, overlaying multiple layers
+of mutations, exact times associated with mutations,
+restricted time interval.\\
+
+\end{tabular}
+\caption{Major features of \msprime\ 1.0 new since version
+0.3.0~\citep{kelleher2016efficient}.}
+\end{table}
+
 % jk: I started writing a roadmap here, but it's quite hard to do concisely
@@
 they become difficult to understand and
-error-prone~\citep{ragsdale2020lessons,gower2021demes}.
+error-prone~\citep{ragsdale2020lessons,gower2022demes}.
 Specifying parameters through a text file alleviates this problem to a degree,
@@
 feature is the ability to specify demographic models in
-Demes format~\citep{gower2021demes} from the command line,
+Demes format~\citep{gower2022demes} from the command line,
 making simulation of complex demographies straightforward.
@@
 and variation data via APIs
-in the Python and C languages~\citep{tskit2021tskit}.
+in the Python and C languages~\citep{tskit2022tskit}.
 The availability of \tskit\ as a liberally licensed (MIT) open source
@@
 an existing file, and because of its efficient data interchange mechanisms,
-there is no performance penalty for additional operations
+there is no performance penalty for operations being performed
 in a different software package.
@@
 
-As well as providing a new API
-that emphasises the logical split between ancestry and mutation simulation,
-we have greatly extended the sophistication of
+We have greatly extended the sophistication of
 \msprime's mutation generation engine for version 1.0,
@@
 user code by specifying a transition matrix between
-any number of alleles (which can be arbitrary unicode strings).
+any number of alleles.
 Mutation rates can vary along the genome, and multiple mutation
@@
 rather than tree-by-tree and branch-by-branch as would otherwise be required.
-In the above example from Fig.~\ref{fig-mutations-perf}B,
-if we generated mutations tree-by-tree, we would have to iterate over 273,887,838 branches
-(since there are 137,081 trees and 1,998 branches in each
-tree) rather than 508,125 edges, resulting in $\sim$500 times more work.
-% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
-Even if we have a tree sequence consisting of a single tree
-(negating the advantage of working edge-by-edge),
-\msprime's mutation generator is still very efficient.
-For example, we simulated mutations under the BLOSUM62 amino
+Simulating mutations on a single tree is also very efficient;
+for example,
+we simulated mutations under the BLOSUM62 amino
 acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
@@
 output formats from the mutation generation algorithms.
-Given these timings, it seems unlikely
+Given the above timings, it seems unlikely
 that generating mutations with \msprime\ would be a bottleneck in any
@@
 (haploid) sample sizes,
-with parameters roughly equivalent to the current estimate
-for \textit{E.~coli}~\citep{lapierre2016the}:
-a 4.5Mb genome with scaled gene conversion rate of 0.015 and
-a mean tract length of 500.
+and the current estimates
+for \textit{E.~coli} parameters~\citep{lapierre2016the}:
+a 4.6Mb genome, $N_e = 1.8\times 10^8$, gene conversion rate of $8.9\times 10^{-11}$
+per base and tract length of $542$.
 We report (A) the total CPU time and (B) maximum memory usage
@@
 of the very long running times.}
-% The genome length was set to 4.5 Mb, the scaled per site
-% gene conversion rate was set to $N_e g = \gamma = 0.015$ and
-% the mean gene conversion tract length was set to $L = 500$, which is close to
-% estimated parameters for \textit{E.~coli}, i.e.\ an effective populations size
-% $N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
-% with mean tract length $L = 542$~\citep{lapierre2016the}.}
 \end{figure}
@@
 (or be sampled from) inactive populations. This demography model is compatible
-with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
+with the Demes standard~\citep{gower2022demes}, and the \texttt{Demography}
 class supports importing and exporting Demes models.
@@
 diffusion of an allele bound for
-fixation~\citep{coop2004ancestral}.
-The jump process moves back in time following the beneficial
-allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
-the origination of the allele at $p=1/(2N)$, tracking time in small
-increments $\delta t$. Then, given the frequency $p$ at time $t$,
-the frequency $p^\prime$ at time $t + \delta t$ is given by
-\begin{equation*}
-p^\prime =
-\begin{cases}
-    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
-    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
-\end{cases}
-\end{equation*}
-where
-\begin{eqnarray*}
-	\mu(p)=\frac{\alpha p (1 - p)}{\tanh(\alpha(1-p))}.
-\end{eqnarray*}
-Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
-assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
-can be generalised straightforwardly to include arbitrary dominance.
-We can also define trajectories to model neutral alleles and soft
-selective sweeps, which we plan as future additions to \msprime.
-
-Then, given a randomly generated allele frequency trajectory under
-the above model, the simulation of a sweep works by assigning lineages to
+fixation~\citep{coop2004ancestral}, as detailed in the \nameref{app-sweeps}
+appendix. Given a randomly generated allele frequency trajectory,
+the simulation of a sweep works by assigning lineages to
 two different structured coalescent ``labels'', based on
@@
 
+\subsection*{Selective sweeps model}
+\label{app-sweeps}
+Sweep trajectories are generated in \msprime\ using a
+jump process approximation to the conditional
+diffusion of an allele bound for
+fixation~\citep{coop2004ancestral}.
+The jump process moves back in time following the beneficial
+allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
+the origination of the allele at $p=1/(2N)$, tracking time in small
+increments $\delta t$. Then, given the frequency $p$ at time $t$,
+the frequency $p^\prime$ at time $t + \delta t$ is given by
+\begin{equation*}
+p^\prime =
+\begin{cases}
+    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
+    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
+\end{cases}
+\end{equation*}
+where
+\begin{eqnarray*}
+	\mu(p)=\frac{\alpha p (1 - p)}{\tanh(\alpha(1-p))}.
+\end{eqnarray*}
+Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
+assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
+can be generalised straightforwardly to include arbitrary dominance.
+We can also define trajectories to model neutral alleles and soft
+selective sweeps, which we plan as future additions to \msprime.
+
 \subsection*{Likelihood calculations}

### 2021-11-18 paper.tex
Trim down mutations section
Closes #203

@@
 in a tree sequence with 508,125 edges.
-This efficiency for large numbers of trees is possible because
-the tree sequence encoding allows us to generate mutations
-on an edge-by-edge basis
-(see Fig.~\ref{fig-ts-example} and the~\nameref{app-mutation-algorithm}
-appendix),
-rather than tree-by-tree and branch-by-branch as would otherwise be required.
-In the above example from Fig.~\ref{fig-mutations-perf}B,
-if we generated mutations tree-by-tree, we would have to iterate over 273,887,838 branches
-(since there are 137,081 trees and 1,998 branches in each
-tree) rather than 508,125 edges, resulting in $\sim$500 times more work.
-% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
-Even if we have a tree sequence consisting of a single tree
-(negating the advantage of working edge-by-edge),
-\msprime's mutation generator is still very efficient.
-For example, we simulated mutations under the BLOSUM62 amino
+Simulating mutations on a single tree is also very efficient;
+for example,
+we simulated mutations under the BLOSUM62 amino
 acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting

### 2021-11-19 evaluation/generate_gc_perf_data.py
Updated data for gc eval

@@ -19,7 +19,7 @@ def run_simbac(*, sample_size, L, gc_rate, gc_tract_length, count_trees=False):
     R = gc_rate * 2
     # Set theta to 0 to remove mutations (defaults to 0.01)
     args = f"-N {sample_size} -B {int(L)} -R {R} -D {gc_tract_length} -T 0"
-    print(args)
+    # print(args)
     if count_trees:
         # This is cleaned up automatically when the variable goes out of scope
         tempdir = tempfile.TemporaryDirectory()
@@ -46,15 +46,15 @@ def run_fastsimbac(
     # using R=2*gc_rate as gene conversion/recombination rate
     R = gc_rate * 2
     # Set theta to 0 to remove mutations (defaults to 0.01)
-    # We need the -b 500 to set the burn-in so we get the correct distribution
+    # We need the -b {L} to set the burn-in so we get the correct distribution
     # of trees.
-    args = f"{sample_size} {int(L)} -r {R} {gc_tract_length} -t 0 -b 500"
+    args = f"{sample_size} {int(L)} -r {R} {gc_tract_length} -t 0 -b {int(L)}"
     if count_trees:
         args += " -T "
     if set_seed > 0:
         args += " -s " + str(set_seed)
 
-    print("running", args)
+    # print("running", args)
     output = subprocess.run(
         "./tools/fastSimBac " + args, shell=True, check=True, capture_output=True
     )

### 2021-11-19 paper.tex
Tweaks

--- 2021-11-18 paper.tex
@@
 \begin{table}
+{ % begin box to localize effect of arraystretch change
+\renewcommand{\arraystretch}{1.5}
 \begin{tabular}{lp{12cm}}
@@
 with forward simulators such as \SLiM\ and \fwdpy.
+Ability to store arbitrary metadata along with simulation
+results, and automatic recording of provenance information
+for reproducibility.
 Rich suite of analytical and visualisation methods via \tskit\ library. \\
-\midrule
 
@@
   and selective sweep models. Instantaneous bottlenecks.
-  Discrete or continuous coordinates, arbitary ploidy,
+  Discrete or continuous genomic coordinates, arbitary ploidy,
   gene conversion.
@@
 Varying rates along the genome, recurrent/back mutations,
-discrete or continuous coordinates, overlaying multiple layers
+discrete or continuous genomic coordinates, overlaying multiple layers
 of mutations, exact times associated with mutations,
@@
 \end{tabular}
+}
 \caption{Major features of \msprime\ 1.0 new since version

### 2021-11-19 response-to-reviewers.tex

@@
 \begin{reply}
-We avoided included the absolute values here in order to make the comparison
-between the simulators as simple as possible (since SimBac and FastSimBac
-require scaled inputs). Since the values we use aren't precisely the estimates
-for E.~coli (we rounded some values for simplicity), it would probably
-be misleading to include some Ne and gc rates that are consistent with the
-scaled value we used, but not equal to the actual E.~coli estimates, and
-we would therefore prefer not to include them.
+We have redone this figure with the exact estimates from Lapierre et al
+and quoted the absolute Ne and gc values.
 \end{reply}

### 2021-11-19 response-to-reviewers.tex

@@
 simulators that use inefficient data formats, and it is these users that
-we want to convince.
+we want to convince. We also hope to convince those that might be developing
+their own simulators to take advantage of tskit and its extensive
+functionality, rather then developing everything from scratch, as is the
+classical and still dominant approach.
 \end{reply}
@@
 \begin{reply}
-TODO
+We have removed the some of the examples and shortened the section.
 \end{reply}

### 2021-11-19 paper.tex

--- 2021-11-18 paper.tex
@@
 in a tree sequence with 508,125 edges.
+This efficiency for large numbers of trees is possible because
+the tree sequence encoding allows us to generate mutations
+on an edge-by-edge basis
+(see Fig.~\ref{fig-ts-example} and the~\nameref{app-mutation-algorithm}
+appendix),
+rather than tree-by-tree and branch-by-branch as would otherwise be required.
 Simulating mutations on a single tree is also very efficient;

### 2021-11-19 paper.tex
Update table

--- 2021-11-19 paper.tex
@@
 Interface
-& Separation of ancestry and mutation simulations. Integration
-with forward simulators such as \SLiM\ and \fwdpy.
+& Separation of ancestry and mutation simulations.
 Ability to store arbitrary metadata along with simulation
@@
 for reproducibility.
+Jupyter notebook~\citep{kluyver2016jupyter} integration.
 Rich suite of analytical and visualisation methods via \tskit\ library. \\
@@
 gene conversion.
-  Recording full ARG, full migration history, census
-  events.  More efficient for large numbers of populations.\\
+Output full ARG with recombination nodes, ARG likelihood calculations.
+Record full migration history and census
+events.  Improved performance for large numbers of populations.
+Integration with forward simulators such as \SLiM\ and \fwdpy\
+(``recapitation'').
+\\
 

### 2021-11-19 paper.tex
Update

@@
 discrete or continuous genomic coordinates, overlaying multiple layers
-of mutations, exact times associated with mutations,
-restricted time interval.\\
+of mutations, exact times associated with mutations.
 

### 2021-11-19 response-to-reviewers.tex
Responses to reviewer 2 & 3

@@
 \end{point}
+
 \begin{reply}
-TODO.
+The large number and diversity of features in msprime has made the
+organisation of this
+manuscript a major challenge, and the present layout of a series of
+relatively self-contained topics arrived at after much experimentation.
+Issues like interface design and algorithmic efficiency necessarily
+cut across most of these topics. Separate sections on efficiency
+and API design would need to refer to these other sections for context,
+and would lead, we believe, to a longer and less cohesive document.
+
+However, we do agree that some sections are too long and this criticism
+has been very helpful to allow us see where content could be cut
+without loss of information.
+For example, we have cut out references to the API design in the
+Simulation Mutations section, and substantially cut back the Demography
+section (see next point).
 \end{reply}
@@
 \begin{reply}
-TODO.
+This is a very helpful observation, and helped us clarify where
+text can be cut. We have substantially but back the Demography section.
 \end{reply}
@@
 \begin{reply}
-Thank you for your suggestion, we have removed the number of lines of code in the revised manuscript.
+Thank you for your suggestion, we have removed the number of lines of code
+in the revised manuscript.
 \end{reply}
@@
 \begin{reply}
-TODO.
+Thank you, we appreciate the kind words.
 \end{reply}
@@
 \begin{reply}
-TODO.
+We respectfully disagree, and believe that this paper is both of substantial
+interest to the readership of Genetics, and presents a number of significant
+innovations that will enable years of research to follow.
+These innovations either
+allow simulation of models for which there is no other available simulator
+(for example, the $\Lambda$-coalescent models and instantaneous bottlenecks)
+or at a genuinely transformative performance differential from existing
+methods. For example, msprime can simulate E-coli genomes tens of
+times faster than specialised bacterial simulators, and selective
+sweeps multiple orders of magnitude faster than discoal.
+These are not minor incremental improvements, and fundamentally shift
+the realms of possibility for population genetic simulations.
 \end{reply}
@@
 \begin{point}
-
 Specifically, the need of the field is really to simulate more realistic models
@@
 predicted by purifying selection on coding and non-coding elements. bioRxiv.
+\end{point}
 
-\end{point}
 \begin{reply}
-TODO.
+It is, of course, absolutely correct to point out that neutral simulations
+are an approximation, and that this approximation will be highly inaccurate
+and inappropriate in many situations. However, it is reasonable to say that
+it is both a widely used approximation, and that the
+shortcomings of the neutral approximation have been thoroughly discussed
+elsewhere. It is surely beyond the scope of this paper
+to settle a decades long debate.
 \end{reply}
@@
 \begin{point}
-
 Relatedly, for the simulations of positive selection, while it has certainly
@@
 \begin{reply}
-TODO.
+This is a fair point and absolutely correct. However, we have simply
+implemented the standard approach that has been used and studied extensively
+for decades, and have not made any claims beyond this.
 \end{reply}
@@
 \begin{point}
-
 In summary, I believe this to be a nice software addition that belongs as a
@@
 \begin{reply}
-TODO
+Thank you for taking the time to explain your perspective. While we disagree on
+the value of this manuscript and its potential interest to the Genetics
+readership, we do agree that there is a large gap in present-day simulation
+methods and that devising more sophisticated approaches to selection in
+particular is a vital avenue for future work.
 \end{reply}

### 2021-11-19 paper.tex
Tweak
Update table

@@
 Demography
-& Improved inferface, import from Newick species tree and
-*BEAST~\citep{heled2009bayesian},
-Demes support~\citep{gower2022demes}, numerical methods to compute
+&
+Improved interface with integrated metadata and referencing populations by
+name. Import from Newick species tree, *BEAST~\citep{heled2009bayesian},
+and Demes~\citep{gower2022demes}. Numerical methods to compute
 coalescence rates.\\
@@
 }
-\caption{Major features of \msprime\ 1.0 new since version
+\caption{Major features of \msprime\ 1.0 added since version
 0.3.0~\citep{kelleher2016efficient}.}

### 2021-11-19 paper.tex

--- 2021-11-19 paper.tex
@@
 \usetikzlibrary{calc,positioning}
-\usepackage{booktabs}
 %\usepackage{showkeys}
@@
 
-\begin{table}
-{ % begin box to localize effect of arraystretch change
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{lp{12cm}}
-Interface
-& Separation of ancestry and mutation simulations.
-Ability to store arbitrary metadata along with simulation
-results, and automatic recording of provenance information
-for reproducibility.
-Jupyter notebook~\citep{kluyver2016jupyter} integration.
-Rich suite of analytical and visualisation methods via \tskit\ library. \\
-
-Ancestry
-& SMC, SMC', Beta- and Dirac-coalescent, discrete time Wright-Fisher,
-and selective sweep models. Instantaneous bottlenecks.
-Discrete or continuous genomic coordinates, arbitary ploidy,
-gene conversion.
-Output full ARG with recombination nodes, ARG likelihood calculations.
-Record full migration history and census
-events.  Improved performance for large numbers of populations.
-Integration with forward simulators such as \SLiM\ and \fwdpy\
-(``recapitation'').
-\\
-
-Demography
-& Improved inferface, import from Newick species tree and
-*BEAST~\citep{heled2009bayesian},
-Demes support~\citep{gower2022demes}, numerical methods to compute
-coalescence rates.\\
-
-Mutations
-& JC69, HKY, F84, GTR, BLOSUM62, PAM, infinite alleles, SLiM and
-general matrix mutation models.
-Varying rates along the genome, recurrent/back mutations,
-discrete or continuous genomic coordinates, overlaying multiple layers
-of mutations, exact times associated with mutations.
-
-\end{tabular}
-}
-\caption{Major features of \msprime\ 1.0 new since version
-0.3.0~\citep{kelleher2016efficient}.}
-\end{table}
-
 % jk: I started writing a roadmap here, but it's quite hard to do concisely
@@
 they become difficult to understand and
-error-prone~\citep{ragsdale2020lessons,gower2022demes}.
+error-prone~\citep{ragsdale2020lessons,gower2021demes}.
 Specifying parameters through a text file alleviates this problem to a degree,
@@
 feature is the ability to specify demographic models in
-Demes format~\citep{gower2022demes} from the command line,
+Demes format~\citep{gower2021demes} from the command line,
 making simulation of complex demographies straightforward.
@@
 and variation data via APIs
-in the Python and C languages~\citep{tskit2022tskit}.
+in the Python and C languages~\citep{tskit2021tskit}.
 The availability of \tskit\ as a liberally licensed (MIT) open source
@@
 a 4.6Mb genome, $N_e = 1.8\times 10^8$, gene conversion rate of $8.9\times 10^{-11}$
-per base and tract length of $542$.
+per base and mean tract length of $542$.
 We report (A) the total CPU time and (B) maximum memory usage
@@
 (or be sampled from) inactive populations. This demography model is compatible
-with the Demes standard~\citep{gower2022demes}, and the \texttt{Demography}
+with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
 class supports importing and exporting Demes models.

### 2021-11-19 paper.tex
Cut down Demography section

--- 2021-11-19 paper.tex
@@
 \label{sec-demography}
-% TODO should we put in some citations here for the models? Hardly seems
-% necessary.
 One of the key applications of population genetic simulations is to generate
@@
 as well as population splits, admixtures and pulse migrations.
-The location of sampled lineages
-can be tracked through time in as much detail as required: each tree node is
-automatically associated with the population in which it arose, the location of
-lineages can be recorded at any given time via census events, or every lineage
-migration can be recorded.
-Large demographic models can be simulated efficiently in version \msprime\ 1.0,
-since we only consider populations that contain lineages and
-have non-zero migration rates when generating migration event
-waiting times. This is a considerable improvement
-over version 0.x, which scaled quadratically with the number of populations.
 
@@
 designed to address a design flaw in the \msprime\ 0.x interface which led to
-a number of avoidable errors in downstream
-simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
-separate parameters be provided to the \texttt{simulate} function to describe a
-demographic model, making it easy to accidentally omit information. The 1.0 API
-resolves this issue by creating a new \texttt{Demography} class, which
-encapsulates all information about the demographic model, and fully decouples
-the definition from other simulation details. An instance of this class is then
-provided as a parameter to the new \texttt{sim\_ancestry} function,
-substantially reducing the potential for error.
-Another improvement over the
-0.x APIs is the introduction of explicit population split and admixture events,
-and a population state machine that ensures that lineages cannot migrate into
-(or be sampled from) inactive populations. This demography model is compatible
-with the Demes standard~\citep{gower2022demes}, and the \texttt{Demography}
-class supports importing and exporting Demes models.
-Models previously constructed using the 0.x API can be seamlessly imported into
-the \texttt{Demography} class, and we also support importing demographic
-models from Newick species trees and the output of programs
-like *BEAST~\citep{heled2009bayesian}.
-
-The \texttt{DemographyDebugger} provides detailed information about
-demographic models as well as numerical methods to make predictions
-about these models. For example, we can compute
+avoidable errors in downstream simulations~\citep{ragsdale2020lessons}.
+The new API is more user-friendly, providing the ability, for example,
+to refer to populations by name rather than their integer identifiers.
+We also provide numerical methods to compute
 the coalescence rates for two or more lineages
-drawn from populations at specified times in the past,
 which can be inverted to obtain the
@@
 Many popular approaches in population genetics use the distribution of
-coalescence rates between pairs of lineages in one or more populations to infer
+coalescence rates between pairs of lineages to infer
 effective population sizes over

### 2021-11-19 paper.tex
Cut down bottlenecks section

@@
 using a variant of Hudson's linear time single-locus coalescent
-algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
-by adding events to the \texttt{Demography} class
-(see the \nameref{sec-demography} section)
-and can be used in combination with any other demographic modelling
-features. We have validated the results of these simulations by comparing
-against analytical expectations for coalescence times and the
-site frequency spectrum~\citep{bunnefeld2015inferring}.
+algorithm~\citep{hudson1990gene}, and
+validated the results by comparing
+against analytical expectations~\citep{bunnefeld2015inferring}.
 

### 2021-11-19 paper.tex
More misc cuts

@@
 focusing on the aspects that are either new for this version, or in which
-our approach differs significantly from classical methods. Where appropriate,
+our approach differs significantly from classical methods
+(summarised in Table~\ref{tab-v1-summary}).
+Where appropriate,
 we benchmark \msprime\ against other simulators, but the comparisons are
@@
 }
-\caption{Major features of \msprime\ 1.0 added since version
-0.3.0~\citep{kelleher2016efficient}.}
+\caption{\label{tab-v1-summary}Major features
+of \msprime\ 1.0 added since version 0.3.0~\citep{kelleher2016efficient}.}
 \end{table}
@@
 thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
-The primary interface for \msprime\ is through a thoroughly documented and
-stable Python
+The primary interface for \msprime\ is through a thoroughly documented
+Python
 API, which has encouraged the development of an ecosystem of downstream
@@
 crossover events between the same two parental haploid copies.
-The method is described in more detail by~\cite{nelson2020accounting}.
+The method is described in detail by~\cite{nelson2020accounting}.
 
@@
 flexible choice of simulation scenarios.
-As the discrete time Wright-Fisher model improves accuracy of genealogical
+As the DTWF improves accuracy of genealogical
 patterns in the recent past, we can simulate the recent history using this

### 2021-11-19 paper.tex
Tidy up recombination material

@@
 kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
-See the \nameref{sec-arg} section below for a discussion of this
-closely related concept.
+Note that this formulation is fully compatible with
+the concept of an Ancestral Recombination Graph (ARG)
+and any ARG topology can be fully and efficiently encoded
+in the node and edge tables illustrated in Fig.~\ref{fig-ts-example};
+see the \nameref{sec-arg} section below for
+more details.
 
@@
 occur~\citep{wiuf1999recombination}.
-The SMC approximations work by disallowing common ancestor events
-that generate trapped material, greatly simplifying the process.
-However, this also removes subtle long-range
-correlations in the trees since there are many ways in which ancestry
-segments can merge without overlapping.
-
-Recombination events have a dominant effect on the execution time of
-simulations. For details on the time complexity of the simulation algorithm,
-see the appendix: \nameref{app-time-complexity}.
+For large $\rho$, recombination events in trapped
+ancestral material will dominate, and so we can use this as a proxy for the
+overall number of events in Hudson's algorithm.
+\citet[Eq.~5.10]{hein2004gene} gave
+\begin{equation}\label{eqn-hsw-bound}
+\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
+\end{equation}
+as an upper bound
+on the number of recombination events within trapped ancestral material
+for $n$ samples.
+As discussed in the \nameref{app-time-complexity} appendix, the quadratic
+dependence of simulation running time on $\rho$ implied
+by \eqref{eqn-hsw-bound} is well supported
+by observations, and provides a useful means of predicting how long
+a particular simulation might require.
 
@@
 
-\subsection*{Time complexity of the algorithm}
+\subsection*{Time complexity of Hudson's algorithm}
 \label{app-time-complexity}
@@
 As discussed in the \nameref{sec-recombination} section, the time complexity of
-Hudson's algorithm is quadratic in the population scaled recombination rate
-$\rho = 4 N_e L$, where $L$ is the length of the genome in units of
-recombination distance. For large $\rho$, recombination events in trapped
-ancestral material will dominate, and so we can use this as a proxy for the
-overall number of events in Hudson's algorithm.
-\citet[Eq.~5.10]{hein2004gene} gave
-\begin{equation}\label{eqn-hsw-bound}
-\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
-\end{equation}
-as an upper bound
-on the number of recombination events within trapped ancestral material
-in Hudson's algorithm for $n$ samples.
-% Since the sum is well approximated by $\log{n}$,
-% a natural conjecture is that the time complexity of Hudson's
-% algorithm is $O(\rho^2 \log^2 n)$.
-Fig.~\ref{fig-ancestry-perf} shows the observed run time for
+Hudson's algorithm is predicted to be
+quadratic in the population scaled recombination rate
+$\rho = 4 N_e L$ (where $L$ is the length of the genome in units of
+recombination distance) by Eq.~\eqref{eqn-hsw-bound}.
+Fig.~\ref{fig-ancestry-perf} shows the running time for
 simulations with a variety of population size, chromosome
-length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
-correctly predicts the quadratic dependence on $\rho$
-\citep[see also][Fig.~2]{kelleher2016efficient}.
+length and sample sizes, and shows this quadratic
+prediction is well supported by
+observations~\citep[see also][Fig.~2]{kelleher2016efficient}.
 We also see that the dependence on $n$ is quite weak,
@@
 given $\rho$ and $n$.
-
-\subsubsection*{Practical examples}
 
@@
 using the timing of seconds-long runs is unlikely to be very accurate.
-
-\subsubsection*{Changing population size}
 

### 2021-11-19 paper.tex
Cut prediction para in GC section

@@
 
-In terms of predicting the run time for a simulation including gene conversion,
-we recommend following the same approach as discussed in the previous section:
-run a number of simulations for short genome lengths, fit a quadratic to the
-observed CPU times, and use this to predict run time for larger simulations.
-Depending on the relative contributions of gene conversion and crossover
-recombination, this may be an over-estimate since gene conversion events
-tend to generate less trapped ancestral material than crossovers.
-Thus, simulations using mammalian-like gene conversion parameters
-may run faster than simulations in which an equivalent amount of
-crossover recombination is imposed.
-% TODO check this again!
-Since each gene conversion creates two breakpoints and a crossover
-creates only one, we expect the output tree sequence for a given rate
-of gene conversion to be roughly twice the size of the output from a simulation
-with the same rate of crossovers.
-
 \subsection*{Demography}

### 2021-11-19 paper.tex
add affiliation for Castedo

--- 2021-11-19 paper.tex
@@
 \author[23]{Kevin Thornton}
-\author[9]{Hugo van Kemenade}
-\author[7,24]{Anthony W. Wohns}
+\author[24]{Hugo van Kemenade}
+\author[7,25]{Anthony W. Wohns}
 \author[7]{H. Yan Wong}
@@
 \author[10,$\dagger$]{Andrew D. Kern}
-\author[25,$\dagger$]{Jere Koskela}
-\author[10,26,$\dagger$]{Peter L. Ralph}
+\author[26,$\dagger$]{Jere Koskela}
+\author[10,27,$\dagger$]{Peter L. Ralph}
 
@@
 \affil[8]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin }
-% Castedo, Hugo
-\affil[9]{No affiliation}
+% Castedo
+\affil[9]{Fresh Pond Research Institute}
 % Jared, Murillo, Peter, Andy
@@
 \affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
+% Hugo
+\affil[24]{No affiliation}
 % Wilder 2
-\affil[24]{Broad Institute of MIT and Harvard}
+\affil[25]{Broad Institute of MIT and Harvard}
 % Jere
-\affil[25]{Department of Statistics, University of Warwick}
+\affil[26]{Department of Statistics, University of Warwick}
 % Peter (second affil)
-\affil[26]{Department of Mathematics, University of Oregon}
+\affil[27]{Department of Mathematics, University of Oregon}
 
@@
 focusing on the aspects that are either new for this version, or in which
-our approach differs significantly from classical methods
-(summarised in Table~\ref{tab-v1-summary}).
-Where appropriate,
+our approach differs significantly from classical methods. Where appropriate,
 we benchmark \msprime\ against other simulators, but the comparisons are
@@
 \msprime\ against simulators such as \ms, \msms, and \scrm.
-
-\begin{table}
-{ % begin box to localize effect of arraystretch change
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{lp{12cm}}
-Interface
-& Separation of ancestry and mutation simulations.
-Ability to store arbitrary metadata along with simulation
-results, and automatic recording of provenance information
-for reproducibility.
-Jupyter notebook~\citep{kluyver2016jupyter} integration.
-Rich suite of analytical and visualisation methods via \tskit\ library. \\
-
-Ancestry
-& SMC, SMC', Beta- and Dirac-coalescent, discrete time Wright-Fisher,
-and selective sweep models. Instantaneous bottlenecks.
-Discrete or continuous genomic coordinates, arbitary ploidy,
-gene conversion.
-Output full ARG with recombination nodes, ARG likelihood calculations.
-Record full migration history and census
-events.  Improved performance for large numbers of populations.
-Integration with forward simulators such as \SLiM\ and \fwdpy\
-(``recapitation'').
-\\
-
-Demography
-&
-Improved interface with integrated metadata and referencing populations by
-name. Import from Newick species tree, *BEAST~\citep{heled2009bayesian},
-and Demes~\citep{gower2022demes}. Numerical methods to compute
-coalescence rates.\\
-
-Mutations
-& JC69, HKY, F84, GTR, BLOSUM62, PAM, infinite alleles, SLiM and
-general matrix mutation models.
-Varying rates along the genome, recurrent/back mutations,
-discrete or continuous genomic coordinates, overlaying multiple layers
-of mutations, exact times associated with mutations.
-
-\end{tabular}
-}
-\caption{\label{tab-v1-summary}Major features
-of \msprime\ 1.0 added since version 0.3.0~\citep{kelleher2016efficient}.}
-\end{table}
 
@@
 thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
-The primary interface for \msprime\ is through a thoroughly documented
-Python
+The primary interface for \msprime\ is through a thoroughly documented and
+stable Python
 API, which has encouraged the development of an ecosystem of downstream
@@
 \label{sec-demography}
+% TODO should we put in some citations here for the models? Hardly seems
+% necessary.
 One of the key applications of population genetic simulations is to generate
@@
 as well as population splits, admixtures and pulse migrations.
+The location of sampled lineages
+can be tracked through time in as much detail as required: each tree node is
+automatically associated with the population in which it arose, the location of
+lineages can be recorded at any given time via census events, or every lineage
+migration can be recorded.
+Large demographic models can be simulated efficiently in version \msprime\ 1.0,
+since we only consider populations that contain lineages and
+have non-zero migration rates when generating migration event
+waiting times. This is a considerable improvement
+over version 0.x, which scaled quadratically with the number of populations.
 
@@
 designed to address a design flaw in the \msprime\ 0.x interface which led to
-avoidable errors in downstream simulations~\citep{ragsdale2020lessons}.
-The new API is more user-friendly, providing the ability, for example,
-to refer to populations by name rather than their integer identifiers.
-We also provide numerical methods to compute
+a number of avoidable errors in downstream
+simulations~\citep{ragsdale2020lessons}. Briefly, the 0.x API required three
+separate parameters be provided to the \texttt{simulate} function to describe a
+demographic model, making it easy to accidentally omit information. The 1.0 API
+resolves this issue by creating a new \texttt{Demography} class, which
+encapsulates all information about the demographic model, and fully decouples
+the definition from other simulation details. An instance of this class is then
+provided as a parameter to the new \texttt{sim\_ancestry} function,
+substantially reducing the potential for error.
+Another improvement over the
+0.x APIs is the introduction of explicit population split and admixture events,
+and a population state machine that ensures that lineages cannot migrate into
+(or be sampled from) inactive populations. This demography model is compatible
+with the Demes standard~\citep{gower2021demes}, and the \texttt{Demography}
+class supports importing and exporting Demes models.
+Models previously constructed using the 0.x API can be seamlessly imported into
+the \texttt{Demography} class, and we also support importing demographic
+models from Newick species trees and the output of programs
+like *BEAST~\citep{heled2009bayesian}.
+
+The \texttt{DemographyDebugger} provides detailed information about
+demographic models as well as numerical methods to make predictions
+about these models. For example, we can compute
 the coalescence rates for two or more lineages
+drawn from populations at specified times in the past,
 which can be inverted to obtain the
@@
 Many popular approaches in population genetics use the distribution of
-coalescence rates between pairs of lineages to infer
+coalescence rates between pairs of lineages in one or more populations to infer
 effective population sizes over
@@
 using a variant of Hudson's linear time single-locus coalescent
-algorithm~\citep{hudson1990gene}, and
-validated the results by comparing
-against analytical expectations~\citep{bunnefeld2015inferring}.
+algorithm~\citep{hudson1990gene}. Instantaneous bottlenecks are specified
+by adding events to the \texttt{Demography} class
+(see the \nameref{sec-demography} section)
+and can be used in combination with any other demographic modelling
+features. We have validated the results of these simulations by comparing
+against analytical expectations for coalescence times and the
+site frequency spectrum~\citep{bunnefeld2015inferring}.
 
@@
 crossover events between the same two parental haploid copies.
-The method is described in detail by~\cite{nelson2020accounting}.
+The method is described in more detail by~\cite{nelson2020accounting}.
 
@@
 flexible choice of simulation scenarios.
-As the DTWF improves accuracy of genealogical
+As the discrete time Wright-Fisher model improves accuracy of genealogical
 patterns in the recent past, we can simulate the recent history using this

### 2021-11-19 paper.tex

@@
 \author[23]{Kevin Thornton}
-\author[24]{Hugo van Kemenade}
-\author[7,25]{Anthony W. Wohns}
+\author[9]{Hugo van Kemenade}
+\author[7,24]{Anthony W. Wohns}
 \author[7]{H. Yan Wong}
@@
 \author[10,$\dagger$]{Andrew D. Kern}
-\author[26,$\dagger$]{Jere Koskela}
-\author[10,27,$\dagger$]{Peter L. Ralph}
+\author[25,$\dagger$]{Jere Koskela}
+\author[10,26,$\dagger$]{Peter L. Ralph}
 
@@
 \affil[8]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin }
-% Castedo
-\affil[9]{Fresh Pond Research Institute}
+% Castedo, Hugo
+\affil[9]{No affiliation}
 % Jared, Murillo, Peter, Andy
@@
 \affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
-% Hugo
-\affil[24]{No affiliation}
 % Wilder 2
-\affil[25]{Broad Institute of MIT and Harvard}
+\affil[24]{Broad Institute of MIT and Harvard}
 % Jere
-\affil[26]{Department of Statistics, University of Warwick}
+\affil[25]{Department of Statistics, University of Warwick}
 % Peter (second affil)
-\affil[27]{Department of Mathematics, University of Oregon}
+\affil[26]{Department of Mathematics, University of Oregon}
 
@@
 
+\begin{table}
+{ % begin box to localize effect of arraystretch change
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{lp{12cm}}
+Interface
+& Separation of ancestry and mutation simulations.
+Ability to store arbitrary metadata along with simulation
+results, and automatic recording of provenance information
+for reproducibility.
+Jupyter notebook~\citep{kluyver2016jupyter} integration.
+Rich suite of analytical and visualisation methods via \tskit\ library. \\
+
+Ancestry
+& SMC, SMC', Beta- and Dirac-coalescent, discrete time Wright-Fisher,
+and selective sweep models. Instantaneous bottlenecks.
+Discrete or continuous genomic coordinates, arbitary ploidy,
+gene conversion.
+Output full ARG with recombination nodes, ARG likelihood calculations.
+Record full migration history and census
+events.  Improved performance for large numbers of populations.
+Integration with forward simulators such as \SLiM\ and \fwdpy\
+(``recapitation'').
+\\
+
+Demography
+&
+Improved interface with integrated metadata and referencing populations by
+name. Import from Newick species tree, *BEAST~\citep{heled2009bayesian},
+and Demes~\citep{gower2022demes}. Numerical methods to compute
+coalescence rates.\\
+
+Mutations
+& JC69, HKY, F84, GTR, BLOSUM62, PAM, infinite alleles, SLiM and
+general matrix mutation models.
+Varying rates along the genome, recurrent/back mutations,
+discrete or continuous genomic coordinates, overlaying multiple layers
+of mutations, exact times associated with mutations.
+
+\end{tabular}
+}
+\caption{Major features of \msprime\ 1.0 added since version
+0.3.0~\citep{kelleher2016efficient}.}
+\end{table}
+
 % jk: I started writing a roadmap here, but it's quite hard to do concisely

### 2021-11-19 paper.tex
Redo LOC cuts.
Don't know how  #212 got backed out

--- 2021-11-19 paper.tex
@@
 as expected and \msprime\ has an extensive suite.
-As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
-lines of Python, with suites of 122 C tests (7K lines of code) and
-1350 Python tests (22K lines of code). These tests are run
+These tests are run
 automatically on different operating systems on each pull request
@@
 Fig.~\ref{fig-ancestry-perf} shows the running time for
-simulations with a variety of population size, chromosome
+simulations with a variety of population sizes, chromosome
 length and sample sizes, and shows this quadratic

### 2021-11-23 paper.tex

--- 2021-11-19 paper.tex
@@
 \author[23]{Kevin Thornton}
-\author[9]{Hugo van Kemenade}
-\author[7,24]{Anthony W. Wohns}
-\author[7]{H. Yan Wong}
+\author[24]{Hugo van Kemenade}
+\author[7,25]{Anthony W. Wohns}
+\author[7]{Yan Wong}
 

### 2021-11-24 response-to-reviewers.tex
Remove sentence about work being spread out

@@
 \begin{point}
-
 Line 420 - Despite the long description, I was not able to follow how the
@@
 \begin{reply}
-TODO
+This is implied by the fact that the number of lineages decreases slowly
+on the coalescent timescale. However, the point is not crucial to the
+main goal of the paragraph, and we have therefore deleted the sentence
+for clarity.
 \end{reply}
@@
 \begin{point}
-
 Line 422 - There is no context to what is meant by "large" or "small" here - is

### 2021-11-24 response-to-reviewers.tex
Fill out responses to editor and reviewers 1 and 2

@@
 % General intro text goes here
-TODO thanks editor
+Thank you for considering this manuscript for publication. We are glad that
+Reviewers 1 and 2 are enthusiastic about this community effort, and their reviews
+have helped to resolve a number of issues with the manuscript. We are
+disappointed that although Reviewer 3 agrees in the value of the software,
+they do not believe the manuscript
+is appropriate for publication in Genetics. However, as we point out in the
+response, it is unclear whether they were reviewing according to the
+``Methods, technology, and resources`` section scope.
 
-% \reviewersection
+\subsection*{Specific comments}
+% We address your specific comments below:
+% % \reviewersection
 
@@
 
-TODO
-
-We have moved  Hudson's alg and selective sweeps details to appendixes.
+We have shortened the manuscript by moving the details of the analysis
+of Hudson's algorithm and selective sweeps to appendices, and deleting
+unnecessary text in several sections. We have added Table 1, which
+summarises the features added to msprime since the 2016 paper.
 
@@
 investigation within the "Methods, Technology, \& Resources" subject.
+\end{point}
 
+\begin{reply}
+Thank you for the comprehensive summary. We are delighted that these community
+efforts are so well received.
+\end{reply}
+
+\begin{point}
 However, I do think it would be possible to increase the clarity and usefulness
@@
 \begin{reply}
-Thanks Reviewer 1, TODO
+Thank you for the detailed review and suggestions, which we hope have resulted
+in an improved manuscript.
 \end{reply}
@@
 \begin{reply}
-TODO Issue 207
+We have added table 1 to summarise the new features.
 \end{reply}
@@
 figures much more uniform in appearance and purpose (it is a special
-case because it is showing how well our theoretical predictions map
+case because it is showing how well theoretical predictions map
 to observations, as well as those observations).
@@
 \begin{point}
-Line 606 - something goes wrong with the line numbering here.
-\end{point}
-
-\begin{reply}
-TODO
-\end{reply}
-
-\begin{point}
 Line 607 - The equations for the allele frequencies during sweeps seem slightly
@@
 \begin{reply}
-Thanks reviewer 2, TODO.
+Thank you for the kind words, we are delighted that these community efforts are so well
+received.
 \end{reply}
@@
 \begin{reply}
-Thank you for this criticism --- we have reduced the length of the manuscript
-by moving some material to appendices, and tried to address specific points
-raised below.
+Thank you for this well-founded criticism. We have reduced the length of the manuscript
+by moving some material to appendices, cutting ``documentation-like'' text,
+and tried to address specific points raised below. We hope that this has
+resulted in an improved manuscript.
 \end{reply}
@@
 thorough, and convincing argument for the advantages of tree sequences and the
-tskit library for performing simulations.
+tskit library in simulation workflows.
 Very many papers are still published using ms and other
@@
 For example, we have cut out references to the API design in the
-Simulation Mutations section, and substantially cut back the Demography
+Simulating Mutations section, and substantially cut back the Demography
 section (see next point).
@@
 This is a very helpful observation, and helped us clarify where
-text can be cut. We have substantially but back the Demography section.
+text can be cut. We have substantially cut back the Demography section
+and other areas where we were veering towards documentation.
 \end{reply}

### 2021-11-24 response-to-reviewers.tex

@@
 \end{point}
+
 \begin{reply}
-We respectfully disagree, and believe that this paper is both of substantial
-interest to the readership of Genetics, and presents a number of significant
-innovations that will enable years of research to follow.
-These innovations either
-allow simulation of models for which there is no other available simulator
-(for example, the $\Lambda$-coalescent models and instantaneous bottlenecks)
-or at a genuinely transformative performance differential from existing
-methods. For example, msprime can simulate E-coli genomes tens of
-times faster than specialised bacterial simulators, and selective
-sweeps multiple orders of magnitude faster than discoal.
-These are not minor incremental improvements, and fundamentally shift
-the realms of possibility for population genetic simulations.
+We respectfully disagree, and wonder if there is perhaps a simple
+misunderstanding. While we absolutely agree that if this
+manuscript were submitted as an ``Investigation'' to GENETICS,
+it would need to make substantial methodological
+advances of the type that you describe. However, we are submitting
+the manuscript for consideration in the ``Methods, Technology, and Resources``
+section, which has the following description:
+
+\begin{quotation}
+GENETICS welcomes manuscripts that describe genetic or analytic methods or
+resources that are likely to have broad impact. They can be full length
+research articles or Communications. The method or resource needs to be novel,
+or be a significant advance in an existing method or resource. It should be of
+considerable interest to a wide range of geneticists or of extraordinary
+interest to a smaller group of geneticists. The method, resource or technology
+should enable experiments that will allow investigators to address significant
+biological questions and should be described well enough so others can
+implement the method. The necessary reagents or resources need to be available
+upon request.
+\end{quotation}
+
+We believe that this manuscript amply meets these requirements:
+
+\begin{itemize}
+\item The software is likely to have broad impact---the 2016 paper has
+been cited 123 times so far in 2021 according to Google Scholar.
+\item This is a significant advance over the 2016 paper, with a great deal of
+new functionality (see the newly-added Table 1).
+\item It is of interest to population geneticists who use simulations.
+\item The functionality enables novel simulation types (e.g.,
+$\Lambda$-coalescents) and greatly increases the scale over which
+existing methods can be applied (e.g., bacterial simulations).
+\end{itemize}
+
 \end{reply}
@@
 \begin{reply}
-It is, of course, absolutely correct to point out that neutral simulations
-are an approximation, and that this approximation will be highly inaccurate
-and inappropriate in many situations. However, it is reasonable to say that
-it is both a widely used approximation, and that the
-shortcomings of the neutral approximation have been thoroughly discussed
-elsewhere. It is surely beyond the scope of this paper
-to settle a decades long debate.
+We agree that neutral simulations are an approximation, and care needs to
+be exercised in their use. We have added a sentence to introduction to
+emphasise this point. However, as we have argued above, major methodological
+advances of the type that you discuss (which we agree are extremely important!)
+are beyond the scope of the current manuscript.
 \end{reply}
@@
 \begin{reply}
-This is a fair point and absolutely correct. However, we have simply
-implemented the standard approach that has been used and studied extensively
-for decades, and have not made any claims beyond this.
+We agree that this is an important point, but again this is a deep
+methodological issue that is beyond the scope of the current manuscript.
+We have simply implemented methodology that has been used and
+studied decades. Even if the goal is to demonstrate that such
+approximations are deeply flawed, it is still necessary to be
+able to simulate them efficiently and conveniently.
 \end{reply}
@@
 \begin{reply}
-Thank you for taking the time to explain your perspective. While we disagree on
-the value of this manuscript and its potential interest to the Genetics
-readership, we do agree that there is a large gap in present-day simulation
-methods and that devising more sophisticated approaches to selection in
-particular is a vital avenue for future work.
+Thank you for the detailed and interesting review---we agree that there
+is much work to be done in the field of population genetic simulation!
+As we have argued above, we believe the manuscript is entirely suitable
+for publication in Genetics as a
+``Methods, Technology, and Resources`` article.
 \end{reply}

### 2021-11-24 paper.tex
Remove sentence about work being spread out

@@
 \section*{Acknowledgements}
+We acknowledge the contributions of Ivan Krukov who we consider
+eligible for authorship, but were unable to contact for approval.
 We would like to thank Iain Mathieson and Alywyn Scally for helpful comments
@@
 What does this tell us about run time for simulating time-varying population sizes?
-The argument above implies that the work is spread out relatively
-evenly on the coalescent time scale.
 Suppose that population size today is $N_1$,

### 2021-11-24 paper.tex
added a sentence to appease rev3

@@
 catalogued and understood.
-\texttt{Msprime} will continue to be an important tool for these inferences and validations,
-and in this context the ability to interoperate
+While population genetics, as a field, has come to recognize
+the limitations of purely neutral simulations~\citep[e.g.,][]{comeron2014background}, 
+\texttt{msprime} will continue to be an important tool,
+particularly in its ability to interoperate
 with other methods---particularly forwards simulators---through

### 2021-11-25 paper.tex
Caution about neutral sims.
Closes #228

--- 2021-11-24 paper.tex
@@
 
-% and finally lay out future plans for
-% further open-source development of what we hope will become an industry
-% standard piece of computational infrastructure for the genetics community.
+The efficiency of coalescent simulations depends
+crucially on the assumption of neutrality, and it is important
+to note that there are many situations in which this will be a poor approximation
+of biological reality~\citep{johri2021statistical}.
+In particular the existence of pervasive background
+selection in many
+species~\citep{charlesworth1993effect,charlesworth1995pattern,charlesworth2021effects}
+means that care must be taken to ensure that the results of
+neutral simulations are appropriate for the study in question.
 
@@
 name. Import from Newick species tree, *BEAST~\citep{heled2009bayesian},
-and Demes~\citep{gower2022demes}. Numerical methods to compute
+and Demes~\citep{gower2021demes}. Numerical methods to compute
 coalescence rates.\\
@@
 Does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
-The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
-then run time will be similar to a population of size $N_1$;
-while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
-For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
+The answer depends on how $T$ compares to $N_1$:
+if $T / N_1 \ll 1$ then the number of extant lineages remaining after $T$
+generations is likely to be substantial, and the algorithm runtime is primarily
+determined by $N_2$. Conversely, if $T / N_1 \gg 1$, then few extant
+lineages are likely to remain by time $T$ and runtime depends mainly on $N_1$.
+For instance, in many agricultural species $N_1 \approx 100$, while $N_2
+\approx 10^5$,
 and the run time will depend critically on $T$---in other words,

### 2021-11-25 response-to-reviewers.tex
Clarify algorithm timings.
Closes #219

Also change propto to approx - doesn't make any sense to say something
is proportional to 100, right?

@@
 \begin{reply}
-TODO
+We have clarified and explained this point.
 \end{reply}
@@
 \begin{reply}
-We agree that neutral simulations are an approximation, and care needs to
-be exercised in their use. We have added a sentence to introduction to
-emphasise this point. However, as we have argued above, major methodological
+We have added a cautionary paragraph at the end of the introduction, noting that
+neutral simulations are an approximation, which often do not accurately
+reflect biological reality. However, as we have argued above, major methodological
 advances of the type that you discuss (which we agree are extremely important!)

### 2021-11-25 paper.tex
Clarify algorithm timings.
Closes #219

Also change propto to approx - doesn't make any sense to say something
is proportional to 100, right?

@@
 name. Import from Newick species tree, *BEAST~\citep{heled2009bayesian},
-and Demes~\citep{gower2021demes}. Numerical methods to compute
+and Demes~\citep{gower2022demes}. Numerical methods to compute
 coalescence rates.\\

### 2021-11-26 response-to-reviewers.tex
Konrad's edites

@@
 \begin{reply}
-We have added a cautionary paragraph at the end of the introduction, noting that
-neutral simulations are an approximation, which often do not accurately
-reflect biological reality. However, as we have argued above, major methodological
-advances of the type that you discuss (which we agree are extremely important!)
-are beyond the scope of the current manuscript.
+We agree that neutral simulations are drastic approximations, and (as with all
+null models) care needs to be exercised in their interpretation and use. We
+have added text to the introduction to emphasise this point. However, as we
+have argued above, major methodological advances of the type that you discuss
+(which we agree are extremely important!) are beyond the scope of the current
+manuscript. While the question of how to best approximate background selection
+in a coalescent framework is a topic of ongoing research (e.g. Nicolaisen \&
+Desai 2013, Zeng \& Charlesworth 2011), we would argue that the long term
+ambition of simulating ancestry under more realistic non-neutral null models
+(e.g. condition on maps of background selection) only makes sense if such
+simulations are both efficient and can be embedded in realistic models of other
+biological processes, many of which we have implemented in this version of
+msprime.
 \end{reply}

### 2021-11-26 paper.tex
Konrad's edites

--- 2021-11-25 paper.tex
@@
 of biological reality~\citep{johri2021statistical}.
-In particular the existence of pervasive background
-selection in many
-species~\citep{charlesworth1993effect,charlesworth1995pattern,charlesworth2021effects}
-means that care must be taken to ensure that the results of
-neutral simulations are appropriate for the study in question.
+In particular, background selection has been shown to affect genomewide
+sequence variation in a wide range of
+species~\citep{charlesworth1993effect,charlesworth1995pattern,charlesworth2021effects}.
+Thus care must be taken to ensure that the results of neutral simulations are
+appropriate for the question and genomic partition under study.
 

### 2021-11-27 paper.tex
Add/tweak some citations.

@@
 violated. For example in certain highly fecund organisms
-\citep{hedgecock_94,B94,HP11,A04,irwin16}, where individuals have the ability
+\citep{hedgecock_94,B94,HP11,A04,irwin16,vendrami2021sweepstake},
+where individuals have the ability
 to produce numbers of offspring on the order of the population size and
@@
 FB is funded by the Deutsche Forschungsgemeinschaft EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
+GB and KL are supported by an ERC starting grant (ModelGenomLand 757648) to KL.
 Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).

### 2021-11-27 paper.tex
Update paper.tex
line 1986:  'in which no more than one group of a random number of ... '

--- 2021-11-26 paper.tex
@@
 
-Multiple merger coalescents, in which a random number of ancestral
+Multiple merger coalescents, in which no more than one group of a random number of ancestral
 lineages may merge into a common ancestor at a given time,

### 2021-11-27 paper.tex
Update paper.tex
@jeromekelleher :
 the line numbers refer to the github version of paper.tex 

line 217: added  ',and because the mathematical construction separates the mutation process from the ancestral process.'

line 873: 'human-like' sounds a bit funny 
how about 'For values of parameters (recombination rate, mutation rate)  that have been estimated for humans ...' 

line 1452: 'Community-based approach' sounds more appropriate 

these are only  suggestions, either way is fine

--- 2021-11-26 paper.tex
@@
 a sample from an idealised population model, without explicitly representing
-the population in memory or stepping through the generations.
+the population in memory or stepping through the generations, and because the mathematical construction separates the mutation process from the ancestral process.
 Indeed, \cite{hudson1983testing} independently derived the
@@
 
+%%'human-like' sounds a bit funny
 In human-like  parameter regimes and for large sample sizes,
@@
 to leverage the tree sequence data format and associated tools.
-
+%% maybe 'Community-based approach'  
 \subsection*{Development model}

### 2021-11-27 paper.tex
Update paper.tex
@jeromekelleher  put in the  proposed changes for lines  873  and lines  1452;

by the way, is it clear what a 'unit test' is (line  1461) ?

--- 2021-11-27 paper.tex
@@
 
-In human-like parameter regimes and for large sample sizes,
+%%In human-like parameter regimes and for large sample sizes,
+For values of parameters (recombination rate, mutation rate) that have been estimated for humans  
 \msprime's implementation of the
@@
 
-\subsection*{Development model}
+%%\subsection*{Development model}
+\subsection*{Community-based approach}
 \texttt{Msprime} has a large number of features, encompassing
@@
 
-Multiple merger coalescents, in which no more than one group of a random number of ancestral
+Multiple merger coalescents, in which a random number of ancestral
 lineages may merge into a common ancestor at a given time,

### 2021-11-30 paper.tex
End warning on a more positive note

@@
 species~\citep{charlesworth1993effect,charlesworth1995pattern,charlesworth2021effects}.
-Thus care must be taken to ensure that the results of neutral simulations are
+Thus care must be taken to ensure that the results of purely neutral simulations are
 appropriate for the question and genomic partition under study.
+A major strength of \msprime\ is that it can be used in conjunction
+with forwards-time simulators, enabling the simulation of much
+more realistic models than otherwise
+possible~\citep{kelleher2018efficient,haller2018tree}.
 

### 2021-12-01 paper.tex

@@
 appropriate for the question and genomic partition under study.
-A major strength of \msprime\ is that it can be used in conjunction
-with forwards-time simulators, enabling the simulation of much
+A major strength of \msprime, however, is that it can be used in conjunction
+with forwards-time simulators, enabling the simulation of
 more realistic models than otherwise

### 2021-12-01 paper.tex
Remove unnecessary packages and add genetics files.
Compiling version for genetics format

@@
-\documentclass{article}
-\usepackage[round]{natbib}
-\usepackage{listings}
-\usepackage[english]{babel}%
-\usepackage[T1]{fontenc}%
-\usepackage[utf8]{inputenc}%
-\usepackage{amsmath,amssymb,amsfonts}%
-\usepackage{geometry}%
-\usepackage{color}
+\documentclass[9pt,twocolumn,twoside,lineno]{gsajnl}
+\articletype{inv} % article type
+
+\usepackage{amsmath,amssymb}%
 \usepackage{graphicx}
@@
 \usepackage{dsfont}
-\usepackage{verbatim}%
 \usepackage{environ}%
-\usepackage[right]{lineno}%
 \usepackage{nameref}
-\usepackage{caption,subcaption}
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{tikz}
-\usepackage[right]{lineno}
 \usetikzlibrary{calc,positioning}
 \usepackage{booktabs}
-%\usepackage{showkeys}
 
@@
 
-\begin{document}
-
-\linenumbers
 \title{Efficient ancestry and mutation simulation with msprime 1.0}
-% First authors
-\author[1,$\star$]{Franz Baumdicker}
-\author[2,$\star$]{Gertjan Bisschop}
-\author[3,24,$\star$]{Daniel Goldstein}
-\author[4,$\star$]{Graham Gower}
-\author[5,$\star$]{Aaron P. Ragsdale}
-\author[6,$\star$]{Georgia Tsambos}
-\author[7,$\star$]{Sha Zhu}
-
-% Middle Authors
-\author[8]{Bjarki Eldon}
-\author[9]{E. Castedo Ellerman}
-\author[10,11]{Jared G. Galloway}
-\author[12,13]{Ariella L. Gladstein}
-\author[14]{Gregor Gorjanc}
-\author[15]{Bing Guo}
-\author[7]{Ben Jeffery}
-\author[16]{Warren W. Kretzschmar}
-% Haven't been able to contact Ivan, so forced to drop for preprint
-% \author[17]{Ivan Krukov}
-\author[2]{Konrad Lohse}
-\author[17]{Michael Matschiner}
-\author[18]{Dominic Nelson}
-\author[19]{Nathaniel S. Pope}
-\author[20]{Consuelo D. Quinto-Cort\'es}
-\author[10]{Murillo F. Rodrigues}
-\author[21]{Kumar Saunack}
-\author[22]{Thibaut Sellinger}
-\author[23]{Kevin Thornton}
-\author[24]{Hugo van Kemenade}
-\author[7,25]{Anthony W. Wohns}
-\author[7]{Yan Wong}
-
-% Senior Authors
-\author[18,$\dagger$]{Simon Gravel}
-\author[10,$\dagger$]{Andrew D. Kern}
-\author[26,$\dagger$]{Jere Koskela}
-\author[10,27,$\dagger$]{Peter L. Ralph}
-
-% Corresponding
-\author[7,$\ddagger$]{Jerome Kelleher}
-
-% Franz
-\affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'', Mathematical and Computational Population Genetics, University of T\"ubingen}
-% Gertjan, Konrad
-\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
-% Daniel
-\affil[3]{Khoury College of Computer Sciences, Northeastern University}
-% Graham
-\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
-% Aaron
-\affil[5]{Department of Integrative Biology, University of Wisconsin--Madison}
-% Georgia
-\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
-    University of Melbourne}
-% Jerome, Yan, Wilder, Joe
-\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}
-
-% Bjarki
-\affil[8]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin }
-% Castedo
-\affil[9]{Fresh Pond Research Institute}
-% Jared, Murillo, Peter, Andy
-\affil[10]{Institute of Ecology and Evolution, Department of Biology, University of Oregon}
-% Jared
-\affil[11]{Computational Biology Program, Fred Hutchinson Cancer Research
-Center, Seattle, WA 98102, USA}
-% Ariella
-\affil[12]{Department of Genetics, University of North Carolina at Chapel Hill}
-\affil[13]{Embark Veterinary, Inc., Boston}
-% Gregor
-\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh}
-% Bing
-\affil[15]{Institute for Genome Sciences, University of Maryland School of Medicine, Baltimore, MD}
-% Winni
-\affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
-% Michael
-\affil[17]{Natural History Museum, University of Oslo}
-% Simon, Dom, Ivan
-\affil[18]{Department of Human Genetics, McGill University}
-% Nathaniel
-\affil[19]{Department of Entomology, Pennsylvania State University}
-% Consuelo
-\affil[20]{National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
-% Saunack
-\affil[21]{IIT Bombay, India}
-% Thibaut
-\affil[22]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
-% Kevin
-\affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
-% Hugo
-\affil[24]{No affiliation}
-% Wilder 2
-\affil[25]{Broad Institute of MIT and Harvard}
-% Jere
-\affil[26]{Department of Statistics, University of Warwick}
-% Peter (second affil)
-\affil[27]{Department of Mathematics, University of Oregon}
-
-\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
-\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
-\affil[$\ddagger$]{Denotes corresponding author}
-
-% \section*{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}
-
-\maketitle
+
+\author[$\ast$,1]{Author One}
+\author[$\dagger$,1]{Author Two}
+\author[$\ddagger$,2]{Author Three}
+\author[$\S$]{Author Four}
+\author[$\ast\ast$,2,3]{Author Five}
+
+\affil[$\ast$]{Author one affiliation}
+\affil[$\dagger$]{Author two affiliation}
+\affil[$\ddagger$]{Author three affiliation}
+\affil[$\S$]{Author four affiliation}
+\affil[$\ast\ast$]{Author five affiliation}
+
+\keywords{Simulation, Coalescent, Mutations, Ancestral Recombination Graphs}
+
+\runningtitle{Efficient ancestry and mutation simulation with msprime 1.0}
+\runningauthor{Baumdicker \textit{et al.}}
+
+% % First authors
+% \author[1,$\star$]{Franz Baumdicker}
+% \author[2,$\star$]{Gertjan Bisschop}
+% \author[3,24,$\star$]{Daniel Goldstein}
+% \author[4,$\star$]{Graham Gower}
+% \author[5,$\star$]{Aaron P. Ragsdale}
+% \author[6,$\star$]{Georgia Tsambos}
+% \author[7,$\star$]{Sha Zhu}
+
+% % Middle Authors
+% \author[8]{Bjarki Eldon}
+% \author[9]{E. Castedo Ellerman}
+% \author[10,11]{Jared G. Galloway}
+% \author[12,13]{Ariella L. Gladstein}
+% \author[14]{Gregor Gorjanc}
+% \author[15]{Bing Guo}
+% \author[7]{Ben Jeffery}
+% \author[16]{Warren W. Kretzschmar}
+% % Haven't been able to contact Ivan, so forced to drop for preprint
+% % \author[17]{Ivan Krukov}
+% \author[2]{Konrad Lohse}
+% \author[17]{Michael Matschiner}
+% \author[18]{Dominic Nelson}
+% \author[19]{Nathaniel S. Pope}
+% \author[20]{Consuelo D. Quinto-Cort\'es}
+% \author[10]{Murillo F. Rodrigues}
+% \author[21]{Kumar Saunack}
+% \author[22]{Thibaut Sellinger}
+% \author[23]{Kevin Thornton}
+% \author[24]{Hugo van Kemenade}
+% \author[7,25]{Anthony W. Wohns}
+% \author[7]{Yan Wong}
+
+% % Senior Authors
+% \author[18,$\dagger$]{Simon Gravel}
+% \author[10,$\dagger$]{Andrew D. Kern}
+% \author[26,$\dagger$]{Jere Koskela}
+% \author[10,27,$\dagger$]{Peter L. Ralph}
+
+% % Corresponding
+% \author[7,$\ddagger$]{Jerome Kelleher}
+
+% % Franz
+% \affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'', Mathematical and Computational Population Genetics, University of T\"ubingen}
+% % Gertjan, Konrad
+% \affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
+% % Daniel
+% \affil[3]{Khoury College of Computer Sciences, Northeastern University}
+% % Graham
+% \affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
+% % Aaron
+% \affil[5]{Department of Integrative Biology, University of Wisconsin--Madison}
+% % Georgia
+% \affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
+%     University of Melbourne}
+% % Jerome, Yan, Wilder, Joe
+% \affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}
+
+% % Bjarki
+% \affil[8]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin }
+% % Castedo
+% \affil[9]{Fresh Pond Research Institute}
+% % Jared, Murillo, Peter, Andy
+% \affil[10]{Institute of Ecology and Evolution, Department of Biology, University of Oregon}
+% % Jared
+% \affil[11]{Computational Biology Program, Fred Hutchinson Cancer Research
+% Center, Seattle, WA 98102, USA}
+% % Ariella
+% \affil[12]{Department of Genetics, University of North Carolina at Chapel Hill}
+% \affil[13]{Embark Veterinary, Inc., Boston}
+% % Gregor
+% \affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh}
+% % Bing
+% \affil[15]{Institute for Genome Sciences, University of Maryland School of Medicine, Baltimore, MD}
+% % Winni
+% \affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
+% % Michael
+% \affil[17]{Natural History Museum, University of Oslo}
+% % Simon, Dom, Ivan
+% \affil[18]{Department of Human Genetics, McGill University}
+% % Nathaniel
+% \affil[19]{Department of Entomology, Pennsylvania State University}
+% % Consuelo
+% \affil[20]{National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
+% % Saunack
+% \affil[21]{IIT Bombay, India}
+% % Thibaut
+% \affil[22]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+% % Kevin
+% \affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
+% % Hugo
+% \affil[24]{No affiliation}
+% % Wilder 2
+% \affil[25]{Broad Institute of MIT and Harvard}
+% % Jere
+% \affil[26]{Department of Statistics, University of Warwick}
+% % Peter (second affil)
+% \affil[27]{Department of Mathematics, University of Oregon}
+
+% \affil[$\star$]{Denotes shared first authorship, listed alphabetically}
+% \affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
+% \affil[$\ddagger$]{Denotes corresponding author}
+
+% % \section*{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}
+
+% \maketitle
 
@@
 
-\textbf{Keywords:} Simulation, Coalescent, Mutations, Ancestral Recombination
-Graphs
-
-\section*{Introduction}
+\begin{document}
+
+\maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+
+% Use the \equalcontrib command to mark authors with equal
+% contributions, using the relevant superscript numbers
+\equalcontrib{1}
+\equalcontrib{2}
+
+\correspondingauthoraffiliation{3}{Corresponding author: Please insert the affiliation correspondence address and email for the corresponding author.
+The corresponding author should be marked with the relevant number in the author list, as shown in the example.}
+\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
+
+\section{Introduction}
 
@@
 
-\bibliographystyle{plainnat}
 \bibliography{paper}

### 2021-12-01 paper.tex
Fixup authors

--- 2021-12-01 paper.tex
@@
 
-\begin{document}
-
-\linenumbers
 \title{Efficient ancestry and mutation simulation with msprime 1.0}
+
 % First authors
@@
 \author[16]{Warren W. Kretzschmar}
-% Haven't been able to contact Ivan, so forced to drop for preprint
-% \author[17]{Ivan Krukov}
 \author[2]{Konrad Lohse}
@@
 
-\affil[$\star$]{Denotes shared first authorship, listed alphabetically}
-\affil[$\dagger$]{Denotes shared senior authorship, listed alphabetically}
-\affil[$\ddagger$]{Denotes corresponding author}
-
-% \section*{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}
-
-\maketitle
+% % \section*{Contact:} \href{jerome.kelleher@bdi.ox.ac.uk}{jerome.kelleher@bdi.ox.ac.uk}
+
+% \maketitle
+
+\keywords{Simulation, Coalescent, Mutations, Ancestral Recombination Graphs}
+
+\runningtitle{Efficient ancestry and mutation simulation with msprime 1.0}
+\runningauthor{Baumdicker \textit{et al.}}
 
@@
 
-\textbf{Keywords:} Simulation, Coalescent, Mutations, Ancestral Recombination
-Graphs
-
-\section*{Introduction}
+\begin{document}
+
+\maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+
+% Use the \equalcontrib command to mark authors with equal
+% contributions, using the relevant superscript numbers
+% \equalcontrib{1}
+% \equalcontrib{2}
+\equalcontrib{$\star$} % {Denotes shared first authorship, listed alphabetically}
+\equalcontrib{$\dagger$} % {Denotes shared senior authorship, listed alphabetically}
+% \equalcontrib{$\ddagger$}{Denotes corresponding author}
+
+\correspondingauthoraffiliation{$\ddagger$}{\url{jerome.kelleher@bdi.ox.ac.uk}}
+\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
+
+\section{Introduction}
 

### 2021-12-01 paper.tex
Fixup table 1

@@
 \begin{table}
+\small{
 { % begin box to localize effect of arraystretch change
 \renewcommand{\arraystretch}{1.5}
-\begin{tabular}{lp{12cm}}
+\begin{tabular}{lp{6.5cm}}
 Interface
@@
 }
+}
 \caption{\label{tab-v1-summary}Major features
-of \msprime\ 1.0 added since version 0.3.0~\citep{kelleher2016efficient}.}
+of \msprime\ 1.0 added since version 0.3.0 \citep{kelleher2016efficient}.}
 \end{table}

### 2021-12-01 paper.tex
Convert figures to two-column directly
Split some equations in appendix

@@
 
-\runningtitle{Efficient ancestry and mutation simulation with msprime 1.0}
+\runningtitle{Efficient simulation with msprime 1.0}
 \runningauthor{Baumdicker \textit{et al.}}
@@
 
-\begin{figure}
+\begin{figure*}
     \includegraphics{illustrations/mutated_tree}
@@
     }
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
 }
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
     \includegraphics{figures/mutations-perf}
@@
 length from 1 to 100 megabases.}
-\end{figure}
+\end{figure*}
 
@@
 of the very long running times.}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{subfigure}[t]{0.33\textwidth}
@@
 (A) and (B).}
-\end{figure}
+\end{figure*}
 
@@
 and sequence length varying from 100kb--3000kb.}
-\end{figure}
+\end{figure*}
 
@@
 \begin{align*}
-  \lambda_{n; k_1, \ldots, k_r; s}  = {}& \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
-_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
+  \lambda_{n; k_1, \ldots, k_r; s}  = {}&
+\int_\Delta  \sum_{\ell = 0}^s
+\sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty
+    \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}} \\
+    & \times \left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx) \\
     & +  a\one{r=1, k_1 = 2},
@@
 up to four simultaneous mergers, giving rise to a $\Xi$-coalescent with merger rate
-\begin{esplit}{Cconst} \lambda^{\text{Dirac}}_{b; k_1, \ldots,
-k_r;s } &= \frac{c \psi^2 / 4}{1 + c \psi^2 / 4} \frac{4}{ \psi^2}\sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell}
-(4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots +
-k_r + \ell} + \frac{\one{r=1, k_1 = 2}}{1 + c \psi^2 / 4},  \\ \end{esplit}
+\begin{esplit}{Cconst}
+\lambda^{\text{Dirac}}_{b; k_1, \ldots, k_r;s } &=
+    \frac{c \psi^2 / 4}{1 + c \psi^2 / 4}
+    \frac{4}{ \psi^2}\sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell}
+    (4)_{r+\ell} (1-\psi)^{s-\ell } \\
+    & \times \left( \frac{\psi}{4} \right)^{k_1 + \cdots + k_r + \ell} + \frac{\one{r=1, k_1 = 2}}{1 + c \psi^2 / 4},
+\end{esplit}
 The interpretation of \eqref{Cconst} is that `small' reproduction events in which

### 2021-12-01 paper.tex

@@
 Because of this necessity, a large number
-of specialised simulation programs have been developed, each
+of specialized simulation programs have been developed, each
 filling a particular niche, but with largely overlapping functionality
@@
 the succinct tree sequence data structure and \tskit\ library.
-We summarise \msprime's many features, and show that
+We summarize \msprime's many features, and show that
 its performance is excellent, often many times faster
-and more memory efficient than specialised alternatives.
+and more memory efficient than specialized alternatives.
 These high-performance features have been thoroughly
@@
 simulation, because it allows us to simulate the genetic ancestry for
-a sample from an idealised population model, without explicitly representing
+a sample from an idealized population model, without explicitly representing
 the population in memory or stepping through the generations.
@@
 coalescent \emph{in order to} efficiently simulate data,
-and used these simulations to characterise an analytically
+and used these simulations to characterize an analytically
 intractable distribution.
@@
 and high quality assemblies are now available for many species.
-Thus, modelling genetic variation data as a series of unlinked
+Thus, modeling genetic variation data as a series of unlinked
 non-recombining loci
@@
 our approach differs significantly from classical methods
-(summarised in Table~\ref{tab-v1-summary}).
+(summarized in Table~\ref{tab-v1-summary}).
 Where appropriate,
@@
 Jupyter notebook~\citep{kluyver2016jupyter} integration.
-Rich suite of analytical and visualisation methods via \tskit\ library. \\
+Rich suite of analytical and visualization methods via \tskit\ library. \\
 
@@
 and selective sweep models. Instantaneous bottlenecks.
-Discrete or continuous genomic coordinates, arbitary ploidy,
+Discrete or continuous genomic coordinates, arbitrary ploidy,
 gene conversion.
@@
 invoke the simulation engine, and process the results in some way.
-This process is cumbersome and labour intensive, and
+This process is cumbersome and labor intensive, and
 a number of packages have been developed
@@
 \cite{martin2017human} were only detected because the pipeline
-could be easily run and reanalysed~\citep{ragsdale2020lessons,martin2020erratum}.
+could be easily run and reanalyzed~\citep{ragsdale2020lessons,martin2020erratum}.
 
@@
         \label{fig-mutated-trees}
-        Visualisation of the separation between ancestry and mutation
+        Visualization of the separation between ancestry and mutation
         simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
@@
         Each diploid individual consists of two genomes (or nodes), indicated
-        by colour.
+        by color.
         (B) This ancestry is provided as the input to \texttt{sim\_mutations},
@@
 While a full discussion of tree sequences and the capabilities
-of \tskit\ is beyond the scope of this article, we summarise
+of \tskit\ is beyond the scope of this article, we summarize
 some aspects that are important for simulation.
@@
 contiguous segments over which parent-child relationships exist,
-generalises the original
+generalizes the original
 ``coalescence records'' concept~\citep{kelleher2016efficient},
@@
 approach of storing ancestry and mutations rather than sequences,
-they do not generalise beyond their immediate settings, and
+they do not generalize beyond their immediate settings, and
 no software library support is available.
@@
 Part of the reason for this poor record of software reuse and modularity is the
-lack of standardised file formats, and in particular, the absence of common
+lack of standardized file formats, and in particular, the absence of common
 library infrastructure to abstract the details of interchanging simulation
@@
 affects patterns of linkage disequilibrium differently~\citep{korunes2017gene}.
-\cite{wiuf2000coalescent} modelled gene conversion in the coalescent
+\cite{wiuf2000coalescent} modeled gene conversion in the coalescent
 via a rate at which gene conversion events are initiated
@@
 and so we compare the performance of \msprime\ with gene conversion to
-two specialised bacterial simulators,
+two specialized bacterial simulators,
 \SimBac~\citep{brown2016simbac} and \FastSimBac~\citep{demaio2017the}.
@@
 of 500 whole \emph{E.~coli} genomes.
-However, \msprime\ is currently lacking many of the specialised features
+However, \msprime\ is currently lacking many of the specialized features
 required to model bacteria, and so an important avenue for future work
@@
 One of the key applications of population genetic simulations is to generate
-data for complex demographies. Beyond idealised cases such as stepping-stone or
-island models, or specialised cases such as isolation-with-migration models,
+data for complex demographies. Beyond idealized cases such as stepping-stone or
+island models, or specialized cases such as isolation-with-migration models,
 analytical results are rarely possible. Simulation is therefore integral to the
@@
 
-A common approach to modelling the effect of demographic history on genealogies
+A common approach to modeling the effect of demographic history on genealogies
 is to assume that effective population size ($N_e$) changes in discrete steps
@@
 generation~\citep{hedgecock_94}. These population dynamics violate basic
-assumptions of the Kingman coalescent, and are better modelled by
+assumptions of the Kingman coalescent, and are better modeled by
 `multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
 two lineages can merge in a given event. Multiple-merger coalescent processes
-have also been shown to be relevant for modelling the effects of selection on
+have also been shown to be relevant for modeling the effects of selection on
 gene genealogies
@@
 understanding of recombination. In Griffiths' ARG formulation,
-a realisation of the coalescent with recombination is a graph in which
+a realization of the coalescent with recombination is a graph in which
 vertices represent common ancestor or recombination events, and edges represent
@@
 
-For our purposes, an ARG is a realisation of the coalescent with
+For our purposes, an ARG is a realization of the coalescent with
 recombination, in the Griffiths (little ARG) sense.
@@
 information is of little relevance as it is in principle unknowable,
-but there are occasions such as visualisation or computing likelihoods (see
+but there are occasions such as visualization or computing likelihoods (see
 below) in which it is useful. We therefore provide
@@
 Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
-We provide functions to compute the likelihood of ARG realisations
+We provide functions to compute the likelihood of ARG realizations
 and mutational patterns under
@@
 sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
-Sweeps are modelled by creating a structured population
+Sweeps are modeled by creating a structured population
 during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
-in which lineages may transit between favoured and unfavoured backgrounds through
+in which lineages may transit between favored and unfavoured backgrounds through
 recombination. This approach allows for many selective sweep scenarios to be simulated
@@
 time simulation.
-Several specialised simulators have been developed to simulate
+Several specialized simulators have been developed to simulate
 sweeps in the coalescent,
@@
 \subsection*{Discrete time Wright-Fisher}
-The coalescent is an idealised model and makes many simplifying assumptions,
+The coalescent is an idealized model and makes many simplifying assumptions,
 but it is often surprisingly robust to violations of these
@@
 $t$ diploid generations~\citep{nelson2020accounting}.
-This pathological behaviour results in
+This pathological behavior results in
 identity-by-descent,
@@
 more quickly and requires substantially less memory than
-\ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
+\ARGON~\citep{palamara2016argon}, a specialized DTWF simulator.
 However, the generation-by-generation approach of the DTWF is less
@@
 \texttt{Msprime} has a large number of features, encompassing
-the functionality of several more specialised simulators
+the functionality of several more specialized simulators
 while maintaining excellent performance.
@@
 
-An important goal of \msprime's development model is to maximise
+An important goal of \msprime's development model is to maximize
 accessibility for prospective users and contributors, and
 to encourage diversity in our community.
-Gender and racial inequality caused by discrimination and marginalisation
+Gender and racial inequality caused by discrimination and marginalization
 is a major problem across
@@
 Within our field, the contribution of women to early computational methods
-in population genetics was marginalised~\citep{dung2019illuminating},
+in population genetics was marginalized~\citep{dung2019illuminating},
 and women continue to be under-represented
@@
 to a high quality if we are to trust its results. There is a growing
-realisation across the sciences
+realization across the sciences
 % More citations?
@@
 with the detailed properties of the inferred ancestries
-catalogued and understood.
+cataloged and understood.
 \texttt{Msprime} will continue to be an important tool for these inferences and validations,
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 We acknowledge the contributions of Ivan Krukov who we consider
@@
     The black lines are quadratic fits separately in each panel
-    and sample size. Vertical grey lines show the approximate values of
+    and sample size. Vertical gray lines show the approximate values of
     $N_e L$ for chromosome 1 in three species, using values from
-    the \stdpopsim\ catalogue~\citep{adrion2020community}.
+    the \stdpopsim\ catalog~\citep{adrion2020community}.
 }
@@
 assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
-can be generalised straightforwardly to include arbitrary dominance.
+can be generalized straightforwardly to include arbitrary dominance.
 We can also define trajectories to model neutral alleles and soft
@@
 \texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
-a normalising constant which depends on the pattern of mutations,
+a normalizing constant which depends on the pattern of mutations,
 but not on the tree sequence or the per-site, per-generation mutation
@@
 \be\label{jX} \lim_{k\to \infty} C k^\alpha \prb{X \ge k} = 1 \ee
-with index $\alpha > 0$, and where $C > 0$ is a normalising constant.
+with index $\alpha > 0$, and where $C > 0$ is a normalizing constant.
 If the  total number of juveniles $S_N := X_1 + \ldots + X_N$ produced
@@
 The stable law \eqref{jX} assumes that individuals can produce arbitrarily
-large numbers of juveniles. Since juveniles are at least fertilised eggs,
+large numbers of juveniles. Since juveniles are at least fertilized eggs,
 it may be desirable to suppose that the number of juveniles

### 2021-12-01 cover_letter/cover_letter2.tex

\documentclass{letter}

\signature{Jerome Kelleher}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{GENETICS}

\opening{Dear Editors,}

I am writing on behalf of my coauthors to resubmit our manuscript entitled
\emph{Efficient ancestry and mutation simulation with msprime 1.0}
for consideration for publication in GENETICS
as a ``Methods, Technology, and Resources'' full length research article.

We thank the reviewers for their thoughtful and thorough reviews. We
believe that we have answered all questions raised, leading to a more
concise and consistently presented manuscript. We have attached
a detailed point-by-point response in the \texttt{response-to-reviewers.pdf}
file.

We have also attached a file showing the changes made to the manuscript
in response to reviewers in \texttt{review-diff.pdf}.

Finally, we attach \texttt{ms-genetics-format.pdf}, which is the manuscript
reformatted using the GENETICS latex template, and converted to US English.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2021-12-04 paper.tex
Citation for conda-forge
Closes #245

@@
 GNU General Public License v3.0, and can be installed from the Python
-Package Index (PyPI) or the \texttt{conda-forge} conda channel.
+Package Index (PyPI) or the \texttt{conda-forge}~\citep{condaforge2015} conda channel.
 Development is conducted openly on GitHub at

### 2021-12-05 paper.tex
Spell check in overleaf

@@
 of biological reality~\citep{johri2021statistical}.
-In particular, background selection has been shown to affect genomewide
+In particular, background selection has been shown to affect genome wide
 sequence variation in a wide range of
@@
 \end{table}
-
-% jk: I started writing a roadmap here, but it's quite hard to do concisely
-% and you either end up leaving sections out, or just summarising them
-% one-by-one which makes it long and tedious. I think readers can either read
-% through fully if they're interested, or just skim the section titles which
-% should give them a good idea of whether they want to read it or not.
 

### 2021-12-05 evaluation/plot.py
Batch of tweaks to plots

@@ -4,7 +4,10 @@ import pandas as pd
 import numpy as np
 import click
 
-plt.rcParams.update({"font.size": 8})
+# Main text size is 9pt
+plt.rcParams.update({"font.size": 7})
+plt.rcParams.update({"legend.fontsize": 6})
+plt.rcParams.update({"lines.markersize": 4})
 
 
 @click.group()
@@ -19,7 +22,10 @@ def save(name):
 
 
 def two_panel_fig(**kwargs):
-    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(6, 3), **kwargs)
+    # The columnwidth of the genetics format is ~250pt, which is
+    # 3 15/32 inch, = 3.46
+    width = 3.46
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(width, width / 2), **kwargs)
     ax1.set_title("(A)")
     ax2.set_title("(B)")
     return fig, (ax1, ax2)
@@ -40,7 +46,9 @@ def mutations_perf():
     markers= ["+", "x", "1"]
     for marker, rate in zip(markers, rates):
         dfr = dfL[dfL.rate == rate].sort_values("n")
-        ax1.plot(dfr.n, dfr.time, marker=marker, label=f"Mutation rate={rate}")
+        rate_exp = int(np.log10(rate))
+        ax1.plot(dfr.n, dfr.time, marker=marker,
+        label=f"Mutation rate=$10^{{{rate_exp}}}$")
 
     ax2.set_xlabel("Sequence length (Megabases)")
     ax1.set_ylabel("Time (seconds)")
@@ -87,7 +95,6 @@ def gc_perf():
     ax1.set_ylabel("Time (hours)")
     ax2.set_xlabel("Sample size (haploid)")
     ax2.set_ylabel("Memory (GiB)")
-    ax1.legend()
 
     dfmsp = df[df["tool"] == "msprime"]
     largest_n = np.array(dfmsp.sample_size)[-1]
@@ -96,11 +103,12 @@ def gc_perf():
     ax1.annotate(
         f"{round(largest_value * 60)} mins",
         textcoords="offset points",
-        xytext=(-30, 5),
+        xytext=(-25, 5),
         xy=(largest_n, largest_value),
         xycoords="data",
     )
     ax2.set_ylim(bottom=0)
+    ax2.legend()
 
     save("gc-perf")
 
@@ -144,9 +152,9 @@ def sweeps_perf():
     largest_value = np.array(dfmsp.time)[-1]
[…]

### 2021-12-05 paper.tex
Batch of tweaks to plots
Update plots to two-column format

@@
 
-\begin{figure*}
+\begin{figure}
     \includegraphics{figures/mutations-perf}
@@
 (B) Holding number of samples fixed at 1000, and varying the sequence
-length from 1 to 100 megabases.}
-\end{figure*}
+length from 1 to 100 megabases.
+\the\columnwidth}
+\end{figure}
 
@@
 
-\begin{figure*}
+\begin{figure}
 \begin{center}
@@
 of the very long running times.}
-\end{figure*}
+\end{figure}
 
@@
 and sequence length varying from 100kb--3000kb.}
-\end{figure*}
+\end{figure}
 
@@
 and a hybrid simulation of 100 generations of the DTWF followed by
-the standard coalescent with recombination (``DTWF + Hudson'').
+the standard coalescent with recombination (``Hybrid'').
 We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
@@
 }
-\end{figure*}
+\end{figure}
 

### 2021-12-05 evaluation/plot.py

@@ -181,7 +181,7 @@ def dtwf_perf():
     # print(df)
 
     df.memory /= 1024 ** 3
-    label_map = {"ARGON": "ARGON", "msprime": "DTWF", "hybrid": "DTWF + Hudson"}
+    label_map = {"ARGON": "ARGON", "msprime": "DTWF", "hybrid": "Hybrid"}
 
     fig, (ax1, ax2) = two_panel_fig()
     lines = {}
@@ -206,7 +206,7 @@ def dtwf_perf():
     ax2.annotate(
         f"{round(largest_value * 1024)} MiB",
         textcoords="offset points",
-        xytext=(-30, 5),
+        xytext=(-26, 5),
         xy=(largest_L, largest_value),
         xycoords="data",
     )

### 2021-12-05 evaluation/plot.py
Revert ancestry perf figure to one-col

@@ -30,6 +30,13 @@ def two_panel_fig(**kwargs):
     ax2.set_title("(B)")
     return fig, (ax1, ax2)
 
+def two_panel_fig_single_col(**kwargs):
+    width = 6
+    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(width, width / 2), **kwargs)
+    ax1.set_title("(A)")
+    ax2.set_title("(B)")
+    return fig, (ax1, ax2)
+
 
 @click.command()
 def mutations_perf():
@@ -235,7 +242,7 @@ def ancestry_perf():
     dromel_chr2l = 4 * 1720600 * 23513712 * 2.40462600791e-08
     print(f"Dromel rho {dromel_chr2l:.2g}")
 
-    fig, axes = two_panel_fig()
+    fig, axes = two_panel_fig_single_col()
 
     def annotate_rho(ax, rho, x_offset, y_offset, text):
         ax.axvline(rho / 4, color="0.8", linestyle="-.")
@@ -337,7 +344,9 @@ cli.add_command(mutations_perf)
 cli.add_command(gc_perf)
 cli.add_command(sweeps_perf)
 cli.add_command(dtwf_perf)
-cli.add_command(ancestry_perf)
+
+with matplotlib.rc_context({"font.size": 7}):
+    cli.add_command(ancestry_perf)
 
 if __name__ == "__main__":
     cli()

### 2021-12-05 paper.tex
Arrange Figure 5 vertically.

@@
 \begin{figure*}
+    \begin{center}
     \includegraphics{illustrations/mutated_tree}
+    \end{center}
     \caption{
@@
 
-\begin{figure*}
-\begin{subfigure}[t]{0.33\textwidth}
+\begin{figure}
+
+\newcommand\subfigtextsize{\small}
+\begin{subfigure}{\columnwidth}
 \centering
 \caption*{(A)}
-    \begin{tikzpicture}[scale=0.45,nodestyle/.style={font=\scriptsize}]
-
-    \draw (-2,0) -- (-2,3) -- (0, 3);
-    \draw (1,0) -- (1,1);
-    \draw (0,1) -- (2,1) -- (2,2) -- (4,2) -- (4,0);
-    \draw (3,2) -- (3,5);
-    \draw (0,1) -- (0,3);
-    \draw (-1, 3) -- (-1, 5) -- (3,5);
-
-    \draw[fill] (-2,0) circle [radius=0.05];
-    \draw[fill] (-1,3) circle [radius=0.05];
-    \draw[fill] (1,0) circle [radius=0.05];
-    \draw[fill] (4,0) circle [radius=0.05];
-    \draw[fill] (1,1) circle [radius=0.05];
-    \draw[fill] (3,2) circle [radius=0.05];
-    \draw[fill] (1,5) circle [radius=0.05];
-
-    \node[nodestyle] [below] at (-2,0) {0};
-    \node[nodestyle] [below] at (1,0) {1};
-    \node[nodestyle] [below] at (4,0) {2};
-    \node[nodestyle] [above] at (1,1) {3};
-    \node[nodestyle] [below right] at (1,1) {$x = 0.3$};
-    \node[nodestyle] [above right] at (3,2) {4};
-    \node[nodestyle] [above left] at (-1,3) {5};
-    \node[nodestyle] [above] at (1,5) {6};
+    \begin{tikzpicture}[xscale=0.8, yscale=0.4, node distance=2mm and 30mm]
+
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\subfigtextsize}}
+
+\node (s0) [greynode] {};
+\node (s1) [right=of s0,greynode] {};
+\node (s2) [right=of s1,greynode] {};
+\node [greynode] (s3) at ($(s1) + (0,1)$) {};
+\node [greynode] (s4) at ($0.5*(s1) + 0.5*(s2) + (0,2)$) {};
+\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
+\node [greynode] (s6) at ($(s1) + (0,5)$) {};
+
+\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x = 0.3$};
+\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
+\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s3/3, s6/6} \node[nodelabel,anchor=south] at (\u) {\lab};
+
+%% Edges
+\draw (s1) -- (s3);
+\foreach \child/\parent in {s6/s5, s6/s4, s5/s0, s4/s2}
+	\draw (\child) -| (\parent);
+
+\draw (s5) -| ($0.5*(s3) + 0.5*(s3-|s5)$) |- (s3);
+\draw (s4) -| ($0.5*(s3) + 0.5*(s3-|s4)$) |- (s3);
 
 \end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.33\textwidth}
+\end{subfigure}\\
+\begin{subfigure}{\columnwidth}
 \centering
 \caption*{(B)}
-\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
-
-\tikzset{greynode/.style={circle,fill,inner sep=0.7},
-nodelabel/.style={font=\scriptsize}}
+\begin{tikzpicture}[xscale=0.8, yscale=0.2, node distance=2mm and 10mm]
+
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\subfigtextsize}}
 
@@
 % Right sample nodes
-\node (rightTree) at (4, 0) {};
+\node (rightTree) at (5, 0) {};
 \node [greynode] (s0r) at ($(s0) + (rightTree)$) {};
@@
 % Axes
-\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
+\draw[very thick] ($(s0) + (0,-2.5)$) -- ($(s2r) + (0,-2.5)$);
 \node (topAx) at (0,-1.0) {};
-\node (topLeft) at ($(s0) + (0,-1.0)$) {};
+\node (topLeft) at ($(s0) + (0,-2.5)$) {};
 \node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
 \foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
-\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
-\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
-\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+\node[anchor=north] at ($(topLeft)$) {\subfigtextsize 0.0};
+\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\subfigtextsize 0.3};
+\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\subfigtextsize 1.0};
 
@@
 % Axis titles
-\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-3.0)$) {$\subfigtextsize\textrm{Genome position}$};
 \end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.33\textwidth}
+\end{subfigure}\\
+\begin{subfigure}{\columnwidth}
 \centering
 \caption*{(C)}
-\begin{tikzpicture}[node distance=2mm and 4mm,scale=0.45]
-
-\tikzset{greynode/.style={circle,fill,inner sep=0.7},
-nodelabel/.style={font=\scriptsize}}
+\begin{tikzpicture}[xscale=0.8, yscale=0.2, node distance=2mm and 10mm]
+
+\tikzset{greynode/.style={circle,fill,inner sep=1},
+nodelabel/.style={font=\subfigtextsize}}
 
@@
 % Axes
-\draw[very thick] ($(s0) + (0,-1.0)$) -- ($(s2r) + (0,-1.0)$);
+\draw[very thick] ($(s0) + (0,-2.0)$) -- ($(s2r) + (0,-2.0)$);
 \node (topAx) at (0,-1.0) {};
-\node (topLeft) at ($(s0) + (0,-1.0)$) {};
+\node (topLeft) at ($(s0) + (0,-2.0)$) {};
 \node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
 \foreach \x in {0, 5, 10} \draw ($(topLeft) + \x*(genUnit) + (0,0.1)$) -- +(0, -0.2); % tick marks
-\node[anchor=north] at ($(topLeft)$) {\scriptsize 0.0};
-\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\scriptsize 0.3};
-\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\scriptsize 1.0};
+\node[anchor=north] at ($(topLeft)$) {\subfigtextsize 0.0};
+\node[anchor=north] at ($(topLeft) + 5*(genUnit)$) {\subfigtextsize 0.3};
+\node[anchor=north] at ($(topLeft) + 10*(genUnit)$) {\subfigtextsize 1.0};
 
@@
 % Axis titles
-\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-1.3)$) {$\scriptsize\textrm{Genome position}$};
+\node (topLabel) at ($(topLeft) + 5*(genUnit) + (0,-3.0)$) {$\subfigtextsize\textrm{Genome position}$};
 \end{tikzpicture}
@@
 (A) and (B).}
-\end{figure*}
+\end{figure}
 

### 2021-12-05 pretty_pictures.py
Make single col fig1

@@ -19,26 +19,26 @@ def mutated_tree():
     )
 
     model = msprime.F84(kappa=2)
-    mts = msprime.sim_mutations(ts, rate=1e-7, model=model, random_seed=4)
+    mts = msprime.sim_mutations(ts, rate=1e-7, model=model, random_seed=45)
 
-    height = 280
-    width = 700
+    height = 210  # height of the plotting box for each TS
+    width = 370
     top = 50
 
     colours = plt.rcParams['axes.prop_cycle'].by_key()['color']
-    print(colours)
+    # print(colours)
 
     def do_svg(ts, **kwargs):
 
         # The page style here is just for Chromium. We shouldn't
-        # need it for other vonersion options.
+        # need it for other output options.
         style = """\
             @media print {
-              @page { margin: 0; size: 6in 2.5in}
-              body { margin: 1.6cm; }
+              @page { margin: 0; size: 3.5in 5in}
+              body { margin: 0cm; }
             }
             text {
-                font-family:DejaVuSans;
+                font-family: "Dejavu Sans", sans-serif;
             }
             """
         for j in range(ts.num_individuals):
@@ -51,7 +51,7 @@ def mutated_tree():
         )
 
         return ts.draw_svg(
-            size=(width / 2, height - top),
+            size=(width, height),
             node_labels={},
             mutation_labels={m.id: m.derived_state for m in ts.mutations()},
             symbol_size=5,
@@ -62,27 +62,29 @@ def mutated_tree():
     font_size = 15
 
     # I think serif is the default, and matches what we're using for labels?
-    def make_text(text, y, font_family="sans"):
-        return (
-            f'<text x="{width / 4}" y="{y}" font-size="{font_size}" '
-            f'font-family="{font_family}" text-anchor="middle">'
-            f"{text}</text>"
-        )
+    def make_text(text, y, font_family=None):
+        html = f'<text x="{width/2}" y="{y}" text-anchor="middle" font-size="{font_size}"'
+        if font_family is not None:
+            html += f' style="font-family: {font_family}"'
+        html += f">{text}</text>"
+        return html
 
-    svg1 = do_svg(ts)
-    svg2 = do_svg(mts)
+    params = {}
+    # params = {'y_axis': True, 'y_ticks': {float(x): x for x in ["0", "1e4", "2e4", "3e4"]}}
[…]

### 2021-12-05 paper.tex
Make single col fig1
Some minor nitpick suggestions

--- 2021-12-05 paper.tex
@@
 evaluate inferences.
-Because of this necessity, a large number
+Because of this, a large number
 of specialized simulation programs have been developed, each
@@
 efficiently implements ancestry and mutation simulations based on
-the succinct tree sequence data structure and \tskit\ library.
+the succinct tree sequence data structure and the \tskit\ library.
 We summarize \msprime's many features, and show that
@@
 Jupyter notebook~\citep{kluyver2016jupyter} integration.
-Rich suite of analytical and visualization methods via \tskit\ library. \\
+Rich suite of analytical and visualization methods via the \tskit\ library. \\
 
@@
 
-\begin{figure*}
-\begin{subfigure}[t]{0.33\textwidth}
+\begin{figure}
+\begin{subfigure}[t]{0.33\columnwidth}
 \centering
@@
 \end{subfigure}
-\begin{subfigure}[t]{0.33\textwidth}
+\begin{subfigure}[t]{0.33\columnwidth}
 \centering

### 2021-12-05 paper.tex

--- 2021-12-05 paper.tex
@@
 \caption*{(A)}
-    \begin{tikzpicture}[xscale=0.8, yscale=0.4, node distance=2mm and 30mm]
+    \begin{tikzpicture}[xscale=0.8, yscale=0.3, node distance=2mm and 20mm]
 
@@
 %\foreach \u/\lab in {s3/3, s3r/3} \node[nodelabel,anchor=west] at (\u) {\lab};
-\foreach \u/\lab in {s4r/3} \node[nodelabel,anchor=south west] at (\u) {\lab};
-\foreach \u/\lab in {s5/4} \node[nodelabel,anchor=south east] at (\u) {\lab};
-\foreach \u/\lab in {s6/5, s6r/5} \node[nodelabel,anchor=south] at (\u) {\lab};
+\foreach \u/\lab in {s4r/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
+\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
+\foreach \u/\lab in {s6/6, s6r/6} \node[nodelabel,anchor=south] at (\u) {\lab};
 
@@
 % Axes
-\draw[very thick] ($(s0) + (0,-2.0)$) -- ($(s2r) + (0,-2.0)$);
+\draw[very thick] ($(s0) + (0,-2.5)$) -- ($(s2r) + (0,-2.5)$);
 \node (topAx) at (0,-1.0) {};
-\node (topLeft) at ($(s0) + (0,-2.0)$) {};
+\node (topLeft) at ($(s0) + (0,-2.5)$) {};
 \node (genUnit) at ($0.1*(s2r) - 0.1*(s0)$) {};
@@
 topology ``simplified'' down to its minimal tree sequence representation.
-Note that the internal nodes have been renumbered in the simplified
-representation, so that, e.g., node 5 in (C) corresponds to node 6 in
-(A) and (B).}
+Note the original node IDs have been retained for clarity.}
 \end{figure}

### 2021-12-05 paper.tex

@@
 
-\begin{figure*}
+\begin{figure}
     \begin{center}
@@
     }
-\end{figure*}
+\end{figure}
 

### 2021-12-06 paper.tex

@@
 The coalescent
-process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
+process~\citep{kingman1982genealogy,kingman1982coalescent,hudson1983testing,
 tajima1983evolutionary}

### 2021-12-06 cover_letter/cover_letter3.tex
Cover letter for final submission

\documentclass{letter}

\signature{Jerome Kelleher}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{GENETICS}

\opening{Dear Editors,}

Please see enclosed the final version of our manuscript in PDF format.
I am unclear what the requirements for the Latex source and image
files are, and will therefore await instructions.

Unfortunately, I will need to revise the grant data provided in the
online form, as it is not complete. Can you advise on how to do this please?

\closing{Sincerely,}

\end{letter}
\end{document}

### 2021-12-06 paper.tex
Funding section

@@
 % Franz
-\affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'', Mathematical and Computational Population Genetics, University of T\"ubingen}
+\affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'', Mathematical and Computational Population Genetics, University of T\"ubingen, 72076 T\"ubingen, Germany}
 % Gertjan, Konrad
-\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
+\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh, EH9 3FL, UK}
 % Daniel
-\affil[3]{Khoury College of Computer Sciences, Northeastern University}
+\affil[3]{Khoury College of Computer Sciences, Northeastern University, MA 02115, USA}
 % Graham
-\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
+\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen, 1350 Copenhagen K, Denmark}
 % Aaron
-\affil[5]{Department of Integrative Biology, University of Wisconsin--Madison}
+\affil[5]{Department of Integrative Biology, University of Wisconsin--Madison, WI 53706, USA}
 % Georgia
-\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
-    University of Melbourne}
+\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics, University of Melbourne, Victoria, 3010, Australia}
 % Jerome, Yan, Wilder, Joe
-\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}
+\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
 
 % Bjarki
-\affil[8]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin }
+\affil[8]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin, 10115, Germany }
 % Castedo
-\affil[9]{Fresh Pond Research Institute}
+\affil[9]{Fresh Pond Research Institute, Cambridge, MA 02140, USA}
 % Jared, Murillo, Peter, Andy
-\affil[10]{Institute of Ecology and Evolution, Department of Biology, University of Oregon}
+\affil[10]{Institute of Ecology and Evolution, Department of Biology, University of Oregon, OR 97405, USA}
 % Jared
-\affil[11]{Computational Biology Program, Fred Hutchinson Cancer Research
-Center, Seattle, WA 98102, USA}
+\affil[11]{Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98102, USA}
 % Ariella
-\affil[12]{Department of Genetics, University of North Carolina at Chapel Hill}
-\affil[13]{Embark Veterinary, Inc., Boston}
+\affil[12]{Department of Genetics, University of North Carolina at Chapel Hill, NC 27599-7264, USA}
+\affil[13]{Embark Veterinary, Inc., Boston, MA 02111, USA}
 % Gregor
-\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh}
+\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
 % Bing
-\affil[15]{Institute for Genome Sciences, University of Maryland School of Medicine, Baltimore, MD}
+\affil[15]{Institute for Genome Sciences, University of Maryland School of Medicine, Baltimore, MD, 21201, USA}
 % Winni
-\affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
+\affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute, 141 83 Huddinge, Sweden}
 % Michael
-\affil[17]{Natural History Museum, University of Oslo}
+\affil[17]{Natural History Museum, University of Oslo, Blindern 0318 Oslo, Norway}
 % Simon, Dom, Ivan
-\affil[18]{Department of Human Genetics, McGill University}
+\affil[18]{Department of Human Genetics, McGill University, Montr\'eal, QC H3A 0C7, Canada}
 % Nathaniel
-\affil[19]{Department of Entomology, Pennsylvania State University}
+\affil[19]{Department of Entomology, Pennsylvania State University, PA 16802, USA}
 % Consuelo
@@
 % Saunack
-\affil[21]{IIT Bombay, India}
+\affil[21]{IIT Bombay, Powai, Mumbai 400 076, Maharashtra, India}
 % Thibaut
-\affil[22]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
+\affil[22]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich, 85354 Freising, Germany}
 % Kevin
-\affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
+\affil[23]{Ecology and Evolutionary Biology, University of California, Irvine, CA 92697, USA}
 % Hugo
@@
 % Wilder 2
-\affil[25]{Broad Institute of MIT and Harvard}
+\affil[25]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
 % Jere
-\affil[26]{Department of Statistics, University of Warwick}
+\affil[26]{Department of Statistics, University of Warwick, CV4 7AL, UK}
 % Peter (second affil)
-\affil[27]{Department of Mathematics, University of Oregon}
+\affil[27]{Department of Mathematics, University of Oregon, OR 97405 USA}
 
@@
 on the manuscript.
+
+\section*{Funding}
 % alphabetised by author. Names spelled out if there's an initial collision

### 2021-12-06 paper.tex

@@
 % Jared, Murillo, Peter, Andy
-\affil[10]{Institute of Ecology and Evolution, Department of Biology, University of Oregon, OR 97405, USA}
+\affil[10]{Institute of Ecology and Evolution, Department of Biology, University of Oregon, OR 97403-5289, USA}
 % Jared
@@
 % Peter (second affil)
-\affil[27]{Department of Mathematics, University of Oregon, OR 97405 USA}
+\affil[27]{Department of Mathematics, University of Oregon, OR 97403-5289 USA}
 

### 2021-12-22 paper.tex
Put PLR in alphabetised list
Add Ariella's funding
fix Goldstein affiliation

@@
 \author[2,$\star$]{Gertjan Bisschop}
-\author[3,24,$\star$]{Daniel Goldstein}
-\author[4,$\star$]{Graham Gower}
-\author[5,$\star$]{Aaron P. Ragsdale}
-\author[6,$\star$]{Georgia Tsambos}
-\author[7,$\star$]{Sha Zhu}
+\author[3,4,$\star$]{Daniel Goldstein}
+\author[5,$\star$]{Graham Gower}
+\author[6,$\star$]{Aaron P. Ragsdale}
+\author[7,$\star$]{Georgia Tsambos}
+\author[8,$\star$]{Sha Zhu}
 
 % Middle Authors
-\author[8]{Bjarki Eldon}
-\author[9]{E. Castedo Ellerman}
-\author[10,11]{Jared G. Galloway}
-\author[12,13]{Ariella L. Gladstein}
-\author[14]{Gregor Gorjanc}
-\author[15]{Bing Guo}
-\author[7]{Ben Jeffery}
-\author[16]{Warren W. Kretzschmar}
+\author[9]{Bjarki Eldon}
+\author[10]{E. Castedo Ellerman}
+\author[11,12]{Jared G. Galloway}
+\author[13,14]{Ariella L. Gladstein}
+\author[15]{Gregor Gorjanc}
+\author[16]{Bing Guo}
+\author[8]{Ben Jeffery}
+\author[17]{Warren W. Kretzschmar}
 \author[2]{Konrad Lohse}
-\author[17]{Michael Matschiner}
-\author[18]{Dominic Nelson}
-\author[19]{Nathaniel S. Pope}
-\author[20]{Consuelo D. Quinto-Cort\'es}
-\author[10]{Murillo F. Rodrigues}
-\author[21]{Kumar Saunack}
-\author[22]{Thibaut Sellinger}
-\author[23]{Kevin Thornton}
-\author[24]{Hugo van Kemenade}
-\author[7,25]{Anthony W. Wohns}
-\author[7]{Yan Wong}
+\author[18]{Michael Matschiner}
+\author[19]{Dominic Nelson}
+\author[20]{Nathaniel S. Pope}
+\author[21]{Consuelo D. Quinto-Cort\'es}
+\author[11]{Murillo F. Rodrigues}
+\author[22]{Kumar Saunack}
+\author[23]{Thibaut Sellinger}
+\author[24]{Kevin Thornton}
+\author[25]{Hugo van Kemenade}
+\author[4,8]{Anthony W. Wohns}
+\author[8]{Yan Wong}
 
 % Senior Authors
-\author[18,$\dagger$]{Simon Gravel}
-\author[10,$\dagger$]{Andrew D. Kern}
+\author[19,$\dagger$]{Simon Gravel}
+\author[11,$\dagger$]{Andrew D. Kern}
 \author[26,$\dagger$]{Jere Koskela}
-\author[10,27,$\dagger$]{Peter L. Ralph}
+\author[11,27,$\dagger$]{Peter L. Ralph}
 
 % Corresponding
-\author[7,$\ddagger$]{Jerome Kelleher}
+\author[8,$\ddagger$]{Jerome Kelleher}
 
@@
 \affil[3]{Khoury College of Computer Sciences, Northeastern University, MA 02115, USA}
+% Daniel, Wilder
+\affil[4]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
 % Graham
-\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen, 1350 Copenhagen K, Denmark}
+\affil[5]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen, 1350 Copenhagen K, Denmark}
 % Aaron
-\affil[5]{Department of Integrative Biology, University of Wisconsin--Madison, WI 53706, USA}
+\affil[6]{Department of Integrative Biology, University of Wisconsin--Madison, WI 53706, USA}
 % Georgia
-\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics, University of Melbourne, Victoria, 3010, Australia}
+\affil[7]{Melbourne Integrative Genomics, School of Mathematics and Statistics, University of Melbourne, Victoria, 3010, Australia}
 % Jerome, Yan, Wilder, Joe
-\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
+\affil[8]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
 
 % Bjarki
-\affil[8]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin, 10115, Germany }
+\affil[9]{Leibniz Institute for Evolution and Biodiversity Science, Museum f\"ur Naturkunde Berlin, 10115, Germany }
 % Castedo
-\affil[9]{Fresh Pond Research Institute, Cambridge, MA 02140, USA}
+\affil[10]{Fresh Pond Research Institute, Cambridge, MA 02140, USA}
 % Jared, Murillo, Peter, Andy
-\affil[10]{Institute of Ecology and Evolution, Department of Biology, University of Oregon, OR 97403-5289, USA}
+\affil[11]{Institute of Ecology and Evolution, Department of Biology, University of Oregon, OR 97403-5289, USA}
 % Jared
-\affil[11]{Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98102, USA}
+\affil[12]{Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98102, USA}
 % Ariella
-\affil[12]{Department of Genetics, University of North Carolina at Chapel Hill, NC 27599-7264, USA}
-\affil[13]{Embark Veterinary, Inc., Boston, MA 02111, USA}
+\affil[13]{Department of Genetics, University of North Carolina at Chapel Hill, NC 27599-7264, USA}
+\affil[14]{Embark Veterinary, Inc., Boston, MA 02111, USA}
 % Gregor
-\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
+\affil[15]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
 % Bing
-\affil[15]{Institute for Genome Sciences, University of Maryland School of Medicine, Baltimore, MD, 21201, USA}
+\affil[16]{Institute for Genome Sciences, University of Maryland School of Medicine, Baltimore, MD, 21201, USA}
 % Winni
-\affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute, 141 83 Huddinge, Sweden}
+\affil[17]{Center for Hematology and Regenerative Medicine, Karolinska Institute, 141 83 Huddinge, Sweden}
 % Michael
-\affil[17]{Natural History Museum, University of Oslo, Blindern 0318 Oslo, Norway}
+\affil[18]{Natural History Museum, University of Oslo, Blindern 0318 Oslo, Norway}
 % Simon, Dom, Ivan
-\affil[18]{Department of Human Genetics, McGill University, Montr\'eal, QC H3A 0C7, Canada}
+\affil[19]{Department of Human Genetics, McGill University, Montr\'eal, QC H3A 0C7, Canada}
 % Nathaniel
-\affil[19]{Department of Entomology, Pennsylvania State University, PA 16802, USA}
+\affil[20]{Department of Entomology, Pennsylvania State University, PA 16802, USA}
 % Consuelo
-\affil[20]{National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
+\affil[21]{National Laboratory of Genomics for Biodiversity (LANGEBIO), Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
 % Saunack
-\affil[21]{IIT Bombay, Powai, Mumbai 400 076, Maharashtra, India}
+\affil[22]{IIT Bombay, Powai, Mumbai 400 076, Maharashtra, India}
 % Thibaut
-\affil[22]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich, 85354 Freising, Germany}
+\affil[23]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich, 85354 Freising, Germany}
 % Kevin
-\affil[23]{Ecology and Evolutionary Biology, University of California, Irvine, CA 92697, USA}
+\affil[24]{Ecology and Evolutionary Biology, University of California, Irvine, CA 92697, USA}
 % Hugo
-\affil[24]{No affiliation}
-% Wilder 2
-\affil[25]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
+\affil[25]{No affiliation}
 % Jere
@@
 ADK was supported by NIH awards R01GM117241 and R01HG010774.
+AG was supported by NIH award R00HG008696 to Daniel R Schrider.
 BE was supported by  DFG grant  273887127 through Priority Programme SPP 1819: Rapid Evolutionary Adaptation (grant STE 325/17-2) to Wolfgang Stephan; BE would also like to acknowledge  funding through The Icelandic Research Centre (Rann{\'i}s) through an Icelandic Research Fund Grant of  Excellence nr.\ 185151-051 to Einar \'Arnason,  Katr\'in Halld\'orsd\'ottir,  Alison Etheridge,  Wolfgang Stephan, and BE.
@@
 Jerome Kelleher is supported by the Robertson Foundation.
+PLR was supported by NIH award R01HG010774.
 SG acknowledges funding from the Canada Research Chairs Program, from the Canadian Institutes of Health Research PJT 173300, and from the Canadian Foundation for Innovation.

### 2022-01-06 paper.tex
remove columnwidth; closes #265
No affiliation -> Independent Researcher
different shared author statements

@@
 % Hugo
-\affil[25]{No affiliation}
+\affil[25]{Independent Researcher}
 % Jere
@@
 % \equalcontrib{2}
-\equalcontrib{$\star$} % {Denotes shared first authorship, listed alphabetically}
-\equalcontrib{$\dagger$} % {Denotes shared senior authorship, listed alphabetically}
+\makeatletter
+\newcommand{\equalcontribfirst}[1]{\@authfootnote{#1}{These authors contributed equally to this work, as joint first authors.}}
+\newcommand{\equalcontriblast}[1]{\@authfootnote{#1}{These authors contributed equally to this work, as joint last authors.}}
+\makeatother
+
+\equalcontribfirst{$\star$} % {Denotes shared first authorship, listed alphabetically}
+\equalcontriblast{$\dagger$} % {Denotes shared senior authorship, listed alphabetically}
 % \equalcontrib{$\ddagger$}{Denotes corresponding author}
@@
 length from 1 to 100 megabases.
-\the\columnwidth}
+}
 \end{figure}

### 2022-01-06 paper.tex
error in description of mutation algorithm

@@
 then choose the derived allele to be equal to the parent allele
-with probability $\beta/(\alpha + \beta)$,
+with probability $1-\beta/\alpha$,
 and randomly choose a different nucleotide otherwise.
 This produces the correct distribution by Poisson thinning:
-a Poisson process with rate $\alpha$ in which each point is discarded independently
-with probability $\beta / (\alpha + \beta)$ is equivalent to a Poisson
+a Poisson process with rate $\alpha$ in which each point is kept independently
+with probability $\beta / \alpha$ is equivalent to a Poisson
 process with rate $\beta$.

### 2022-01-06 paper.tex
time ago

@@
 \usepackage{environ}%
-\usepackage{nameref}
 \usepackage{caption}
@@
 in the node and edge tables illustrated in Fig.~\ref{fig-ts-example};
-see the \nameref{sec-arg} section below for
+see the Ancestral Recombination Graphs section below for
 more details.
@@
 (2) simulate sequence evolution conditioned on this ancestry by generating
-mutations (see the \nameref{sec-mutations} section); and (3) output the
+mutations (see the Simulating mutations section); and (3) output the
 resulting nucleotide sequences by percolating the effects of the mutations through
@@
 stored via options such as \texttt{record\_migrations} or
-\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
+\texttt{record\_full\_arg} (see the Ancestral Recombination Graphs section) and
 can be efficiently and conveniently processed via \tskit\ APIs.
@@
 Because coalescent simulations are usually concerned with
-neutral evolution (see the \nameref{sec-selection} section, however)
+neutral evolution (see the Selective Sweeps section, however)
 the problem of generating synthetic genetic variation can be decomposed into
@@
 on an edge-by-edge basis
-(see Fig.~\ref{fig-ts-example} and the~\nameref{app-mutation-algorithm}
+(see Fig.~\ref{fig-ts-example} and the Mutation Generation
 appendix),
@@
 for $n$ samples.
-As discussed in the \nameref{app-time-complexity} appendix, the quadratic
+As discussed in the Time complexity of Hudson's algorithm appendix, the quadratic
 dependence of simulation running time on $\rho$ implied
@@
  as well as more general properties of coalescent processes.
-See the \nameref{app-multiple-mergers} appendix for more details and model derivations.
+See the Multiple Merger Coalescent Model appendix for more details and model derivations.
 
@@
 the standard coalescent and infinite sites mutation model.
-For details, see the appendix: \nameref{app-likelihoods}.
+For details, see the appendix: Likelihood Calculations.
 
@@
 \label{sec-selection}
+
 Another elaboration of the standard neutral coalescent with recombination
@@
 diffusion of an allele bound for
-fixation~\citep{coop2004ancestral}, as detailed in the \nameref{app-sweeps}
+fixation~\citep{coop2004ancestral}, as detailed in the Selective Sweeps Model
 appendix. Given a randomly generated allele frequency trajectory,
@@
 
-As discussed in the \nameref{sec-recombination} section, the time complexity of
+As discussed in the Recombination section, the time complexity of
 Hudson's algorithm is predicted to be
@@
 the origination of the allele at $p=1/(2N)$, tracking time in small
-increments $\delta t$. Then, given the frequency $p$ at time $t$,
-the frequency $p^\prime$ at time $t + \delta t$ is given by
+increments $\delta t$. Then, given the frequency $p$ at time $t$ ago,
+the frequency $p^\prime$ at time $t + \delta t$ ago is given by
 \begin{equation*}

### 2022-01-19 paper.tex
Small addition

@@
 attractive and both analytic results and empirical
-inference~\citep{griffiths1994sampling, birkner2009modified,galtier2000detecting,
+inference methods~\citep{griffiths1994sampling, birkner2009modified,galtier2000detecting,
 bunnefeld2015inferring} have been developed under this model, there has
